### 2013-11-24 scripts/ApTran_assemble.Rmd
Forked Aphaneogaster transcriptome assembly repository from /projects/climate-cascade/projects/ApTranscriptome
New assembly uses an .Rmd script with `system()` calls to run programs such as `flash` and `Trinity`

This has the *major* advantage that I can use a single script, developing each stage as I go
along, and caching the results so I don't have to re-run each individual program which can take
many hours.

Also makes it easier to intergrate R stats in single script.

Aphaenogaster transcriptome assembly
=======================================

**Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)

Requires the following programs installed and available on path:

* [rlsim](https://github.com/sbotond/rlsim) to simulate Illumina RNAseq reads
* [simNGS](https://github.com/timmassingham/simNGS) to simulate reads 
* [Trim_Galore!](http://www.bioinformatics.babraham.ac.uk/projects/trim <- galore/)
  - requires [cutadapt](https://code.google.com/p/cutadapt/) and [FastQC](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
* [khmer](https://github.com/ged-lab/khmer) to perform digital normalization and filtering
* [Trinity](http://trinityrnaseq.sourceforge.net/) for transcriptome assembly

```{r setup}
options(stringAsFactors=FALSE)

library(ggplot2)
library(knitr)
opts_chunk$set(cache=TRUE)

source("assemble_functions.R")
```

## *in silico* spike-in ##

Simulate Illumina reads from known mRNA sequence to "spike-in" during transcriptome assembly.
Use these simluated reads to evaluate assembly at end.

Fasta file of known mRNA transcripts for simulated reads with known expression values.
I use a file with 1000 *Arabidopsis* mRNA transcripts downloaded from [European Nucleotide Archive](http://www.ebi.ac.uk/ena/home)

```{r simulate}
knownfasta <- "../data/ena.fasta"
system("mkdir -p ../data/sim")
simout <- "../data/sim"

# Randomly select X sequences; output "known.fasta"
sample.fasta(knownfasta, 100)

## Move intermediate files
system(paste("mv mean-length-fasta.txt ", simout, sep=""))

## Trim fasta titles to just ID
system("cut -f1 -d' ' known.fasta > known1.fasta")

## Assign expression-level to each sequence using the `sel` tool in [rlsim](https://github.com/sbotond/rlsim)

system("sel -d '1.0:g:(1500, 3)' known1.fasta > known-sim.fasta")
system("rm known.fasta known1.fasta") # Clean-up temp files
message("Done with expression levels")

## Simulate fragments created during library prep using `rlsim`. Based on length of fragments estimated when using FLASH to pair reads, use empirical length distribution of 180 bp (SD: 20bp).

system("rlsim -v -n 150000 -d '1:n:(180, 20, 100, 500)' known-sim.fasta > known-sim-frags.fasta")
message("Done with simulated fragmentation levels")

## Generate simulated Illumina paired-end reads using [simNGS](http://www.ebi.ac.uk/goldman-srv/simNGS/) and a default runfile provided with simNGS
# output is 'reads_end1.fq' and 'reads_end2.fq'
system("cat known-sim-frags.fasta | simNGS -p paired -o fastq -O reads /opt/software/simNGS/data/s_3_4x.runfile")
message("Done with Illumina read simulation")

# move intermediate files
system(paste("mv rlsim_report.json ", simout, sep=""))
system(paste("mv sel_report.pdf ", simout, sep=""))

# move files to data directory
system("mv reads_end1.fq ../data/A22-si-R1.fastq")
system("mv reads_end2.fq ../data/A22-si-R2.fastq")
system("mv known-sim.fasta ../data/sim")
system("mv known-sim-frags.fasta ../data/sim")
```

## QC ##

Clip adapter sequence from reads and trim poor quality reads using Trim_Galore

```{r trimclip}
# Data directory 
datadir <- "/home/data/Aphaeno_transcriptome/130509_SN1073_0326_BD25DAACXX/Project_Stanton-Geddes_Project_001"
# Directory to store results
system("mkdir -p ../data/trimclip")
clipdir <- "../data/trimclip"

# List of samples
samplist <- list.files(datadir)
(reads1 <- samplist[grep("A[r2]{1,2}-..-R1.fastq", samplist)])
(reads2 <- samplist[grep("A[r2]{1,2}-..-R2.fastq", samplist)])

# Loop across all the data files and trimclip

for (i in 1:length(reads1)) {
    # Run `trim_galore`.
    r1 <- paste(datadir, reads1[i], sep="/")
    r2 <- paste(datadir, reads2[i], sep="/")
    system(paste("trim_galore --quality 20 --phred33 --fastqc_args '--o", clipdir, "' --length 20 --paired --output_dir", clipdir, r1, r2, sep=" "))
    message("Done with QC for sample: ", reads1[i])
}

# Trimclip simulated spike-in reads
system(paste("trim_galore --quality 20 --phred33 --fastqc_args '--o", clipdir, "' --length 20 --paired --output_dir", clipdir, "../data/A22-si-R1.fastq ../data/A22-si-R2.fastq", sep=" "))
```

## Merge overlapping paired-end reads

With standard insert size of 200 bp and 2 x 100bp paired-end sequencing, many paired-reads are overlapping. FLASH attempts to merge paired-end reads when possible.

```{r FLASH}
# make new directory for merged reads
system("mkdir -p ../data/merged")
mergedir <- "../data/merged/"

# list of trimclipped samples
clippedlist <- list.files(clipdir)
(clippedR1 <- clippedlist[grep("A[r2]{1,2}-..-R1_val_1.fq$", clippedlist)])
(clippedR2 <- clippedlist[grep("A[r2]{1,2}-..-R2_val_2.fq$", clippedlist)])

# loop across data files and merge overlapping reads

for (j in 1:length(clippedR1)) {
    outpre <- strsplit(clippedR1[j], split="-R")[[1]][1]
    sampR1 <- paste(clipdir, clippedR1[j], sep="/")
    sampR2 <- paste(clipdir, clippedR2[j], sep="/")

    system(paste("flash --phred-offset 33 --output-directory", mergedir, "--output-prefix", outpre, sampR1, sampR2, sep=" "))
    message("Done with merging reads for sample ", clippedR1[j], ": ", Sys.time())
}
```

## Digital normalization

Normalize reads using khmer

## Transcriptome assemly

Use Trinity for transcriptome assembly.
Requires reads split into 'left' and 'right' files.
Combine all *.notCombined into reads 1 and 2

cat *.1 > Ap-r1.fq
cat *.2 > Ap-r2.fq

Add *.extendedFrags to left reads

Run Trinity

Trinity.pl --seqType fq --JM 50G --left Ap-r1.fq Ap-r2.fq --output

### 2013-11-24 scripts/assemble_functions.R

@@ -1,0 +1,23 @@
+## Functions required by assembleR script
+
+
+sample.fasta <- function(filein, X) {
+    ##-------------Load libraries--------------------------------
+    require(Biostrings)
+
+    ## Read known fasta file
+    file <- readDNAStringSet(filein)
+    length(file)
+    mean(width(file))
+
+    # Sample X sequences
+    samp.file <- sample(file, X)
+    length(samp.file)
+    head(samp.file)
+
+    # Write subset of fasta sequences to file
+    writeXStringSet(samp.file, file="known.fasta")
+
+    # Write mean length to file
+    cat(mean(width(samp.file)), '\n', file="mean-length-fasta.txt")
+}

### 2013-11-24 scripts/expression-analysis.R

@@ -1,0 +1,120 @@
+#################################################################
+## Script to determine which transcripts show a significant norm
+## of reaction across a thermal gradient
+##
+## John Stanton-Geddes
+## 2013-10-24
+#################################################################
+
+# Settings
+
+# Load libraries
+library(plyr)
+library(reshape2)
+
+# Command-line arguments
+
+setwd("../data/A22-merge-diginorm-oases-assembly/sailfish-out")
+
+#read.tables <- function(file.names, ...) {
+#        require(plyr)
+#        ldply(file.names, function(fn) data.frame(Filename=fn, read.table(fn, sep="\t", ...)))
+#    }
+
+# Read all data files
+#data <- read.tables(c("A22-00_quant/quant_bias_corrected.sf", "A22-03_quant/quant_bias_corrected.sf"))
+#colnames(data) <- c("Filename", "transcript", "length", "TPM", "RPKM")
+
+# Cast data so that each transcript in one row
+# test <- head(data)
+#colnames(test) <- c("Filename", "transcript", "length", "TPM", "RPKM")
+#head(test)
+#dcastest <- dcast(melt(test, id.vars="transcript", ID~variable + TPM))
+
+quant00 <- read.table("A22-00_quant/quant_bias_corrected.sf", sep="\t")
+colnames(quant00) <- c("transcript", "length", "TPM00", "RPKM")
+head(quant00)
+dim(quant00)
+
+quant03 <- read.table("A22-03_quant/quant_bias_corrected.sf", sep="\t")
+colnames(quant03) <- c("transcript", "length", "TPM03", "RPKM")
+
+quant07 <- read.table("A22-07_quant/quant_bias_corrected.sf", sep="\t")
+colnames(quant07) <- c("transcript", "length", "TPM07", "RPKM")
+
+quant10 <- read.table("A22-10_quant/quant_bias_corrected.sf", sep="\t")
+colnames(quant10) <- c("transcript", "length", "TPM10", "RPKM")
+
+quant14 <- read.table("A22-14_quant/quant_bias_corrected.sf", sep="\t")
+colnames(quant14) <- c("transcript", "length", "TPM14", "RPKM")
+
[…]

### 2013-11-25 scripts/ApTran_assemble.R
First test of complete script for assembly
  --> Trim_galore for QC (WORKS)
  --> FLASH to merge overlapping reads (WORKS)
  --> diginorm for digital normalization
  --> Trinity for transcriptome assembly

Added supplemental python scripts needed for processing...

@@ -1,0 +1,225 @@
+Aphaenogaster transcriptome assembly
+=======================================
+
+**Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
+
+Requires the following programs installed and available on path:
+
+* [rlsim](https://github.com/sbotond/rlsim) to simulate Illumina RNAseq reads
+* [simNGS](https://github.com/timmassingham/simNGS) to simulate reads 
+* [Trim_Galore!](http://www.bioinformatics.babraham.ac.uk/projects/trim <- galore/)
+  - requires [cutadapt](https://code.google.com/p/cutadapt/) and [FastQC](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
+* [khmer](https://github.com/ged-lab/khmer) to perform digital normalization and filtering
+* [Trinity](http://trinityrnaseq.sourceforge.net/) for transcriptome assembly
+
+
+```{r setup}
+options(stringAsFactors=FALSE)
+
+library(ggplot2)
+library(knitr)
+opts_chunk$set(cache=TRUE)
+
+source("assemble_functions.R")
+```
+
+## *in silico* spike-in ##
+
+Simulate Illumina reads from known mRNA sequence to "spike-in" during transcriptome assembly.
+Use these simluated reads to evaluate assembly at end.
+
+Fasta file of known mRNA transcripts for simulated reads with known expression values.
+I use a file with 1000 *Arabidopsis* mRNA transcripts downloaded from [European Nucleotide Archive](http://www.ebi.ac.uk/ena/home) 
+
+```{r simulate}
+knownfasta <- "../data/ena.fasta"
+system("mkdir -p ../data/sim")
+simout <- "../data/sim"
+
+# Randomly select X sequences; output "known.fasta"
+sample.fasta(knownfasta, 100)
+
+## Move intermediate files
+system(paste("mv mean-length-fasta.txt ", simout, sep=""))
+
+## Trim fasta titles to just ID
+system("cut -f1 -d' ' known.fasta > known1.fasta")
+
+## Assign expression-level to each sequence using the `sel` tool in [rlsim](https://github.com/sbotond/rlsim)
+
+system("sel -d '1.0:g:(1500, 3)' known1.fasta > known-sim.fasta")
+system("rm known.fasta known1.fasta") # Clean-up temp files
+message("Done with expression levels")
+
[…]

### 2013-11-25 scripts/assemstats2.py

@@ -1,0 +1,136 @@
+#! /usr/bin/env python
+#
+# This file is part of khmer, http://github.com/ged-lab/khmer/, and is
+# Copyright (C) Michigan State University, 2009-2013. It is licensed under
+# the three-clause BSD license; see doc/LICENSE.txt. Contact: ctb@msu.edu
+#
+'''
+assemstats.py
+
+Uses screed to calculate various assembly statistics.
+
+You can obtain screed at github by running
+   git clone git://github.com/acr/screed.git
+
+Then, install by running
+   python setup.py install
+in the newly created screed directory.
+
+Once completed, you should be able to run this script as is.
+
+Author: Jason Pell (pelljaso@cse.msu.edu)
+'''
+
+import screed.fasta
+import sys
+import glob
+
+
+def trimLens(lens, minLen):
+    '''
+    Eliminates any reads below a certain threshold.
+    Function assumes that input list lens is sorted smallest to largest.
+    '''
+
+    index = 0
+
+    for i in range(len(lens)):
+        if lens[i] < minLen:
+            index += 1
+        else:
+            break
+
+    return lens[index:len(lens)]
+
+
+def getLens(filename):
+    '''
+    Parses FASTA file using screed to create a sorted list of contig lengths.
+    '''
+    lens = []
+    fd = open(filename, 'r')
+
+    fa_instance = screed.fasta.fasta_iter(fd)
+    for record in fa_instance:
+        lens.append(len(record['sequence']))
+
+    fd.close()
+
+    return sorted(lens)
+
+
+def calcNXX(lens, percent):
+    '''
+    Calculates any NXX (e.g. N50, N90) statistic.
+    '''
+
+    lenSum = sum(lens)
+    threshold = (float(percent) / 100) * lenSum
+
+    runningSum = 0
+    nxx = 0
+    nxxLen = 0
+
+    for i in range(len(lens) - 1, -1, -1):
+        myLen = lens[i]
+        nxx += 1
+        runningSum += myLen
+
+        if runningSum >= threshold:
+            nxxLen = myLen
+            break
+
+    return nxx, nxxLen
+
+
+def main():
+    '''
+    Outputs assembly statistics for provided FASTA files.
+    '''
+    if len(sys.argv) < 3:
+        print "Usage: python assemstats.py <min contig length> [ FASTA files ]"
+        return
+
+    try:
+        minLen = int(sys.argv[1])
+    except ValueError:
+        print "Minimum contig length must be an integer."
+        return
+
[…]

### 2013-11-26 scripts/ApTran_assemble.R
Updates to assembly script - debugging digital normalization

@@ -147,8 +147,7 @@ message("Start diginorm on extendedFrags: ", Sys.time())
 system("/opt/software/khmer/scripts/normalize-by-median.py -R diginorm.out -C 20 -k 20 -N 4 -x 3e9 --savehash ../data/merged/Ap-diginorm-C20k20.kh ../data/merged/*.extendedFrags.fastq")
 message("Done with diginorm on extendedFrags: ", Sys.time())
 
-# next, run `normalize-by-median` for notCombined reads. first need to add Illumina 1.3 style tags
-# /1 or /2 to reads
+# to run `normalize-by-median` for notCombined reads, need to add Illumina 1.3 style tags /1 or /2 to reads
 
 # list of trimclipped samples
 mergedlist <- list.files(mergedir)
@@ -166,24 +165,27 @@ message("Done with diginorm on notCombined reads: ", Sys.time())
 
 # Trim likely erroneous k-mers. Note that this will orphan some reads with poor quality partners
 message("Trim errorneous k-mers: ", Sys.time())
-system("/opt/software/khmer/scripts/filter-abund.py -V Ap-diginorm-C20k20.kh *.keep")
+system("/opt/software/khmer/scripts/filter-abund.py -V ../data/merged/Ap-diginorm-C20k20.kh *.keep")
 
 # Separate orphaned from still-paired reads in .notCombined.fastq.out.keep.abundfilt
 message("Separate orphaned from still-paired reads: ", Sys.time())
-dnlist <- list.files(mergedir)
+dnlist <- list.files()
 (dn <- dnlist[grep("A[r2]{1,2}-...notCombined.fastq.out.keep.abundfilt", dnlist)])
 
 for (n in dn) {
     system(paste("/opt/software/khmer/scripts/extract-paired-reads.py ", n, sep=""))
 }
 
-# Move files to diginorm directory and remove intermediate files
-message("Move final files and clean-up: ", Sys.time())
-system("mv *.abundfilt ../diginorm/)
-system("mv *.abundfilt.pe ../diginorm/")
-system("mv *.abundfilt.se ../diginorm/")
+```
 
-system("rm *.abundfilt")
+ Move final normalized fastq files to diginorm directory and remove intermediate files
+
+```{r diginorm_cleanup}
+message("Move final files and clean-up: ", Sys.time())
[…]

### 2013-12-11 scripts/ApTran_assemble.R
Updated script to evaluate assembly based on in silico spike-in reads. Added function 'assembly.spikein.eval' to assemble_functions.R to do this. Started to add uclust and CAP3 transcriptome filtering to assembly script.

@@ -14,12 +14,18 @@ Requires the following programs installed and available on path:
 
 
 ```{r setup}
+# Global settings
 options(stringAsFactors=FALSE)
 
+# Load libraries
 library(ggplot2)
 library(knitr)
+library(Biostrings)
+
+# knitr options
 opts_chunk$set(cache=TRUE)
 
+# Load personal functions
 source("assemble_functions.R")
 ```
 
@@ -285,7 +291,7 @@ for (e in extendedFrags) {
 
 Run Trinity
 
-```{r A22-trinity, eval=TRUE}
+```{r A22-trinity, eval=FALSE}
 diginormdir <- "../data/diginorm/"
 A22trinitydir <- paste("../results/A22-trinity", Sys.Date(), sep="-")
 system(paste("mkdir -p ", A22trinitydir, sep=""))
@@ -333,14 +339,37 @@ system(paste("python assemstats2.py 100 ", Artrinitydir, "Trinity.fasta", sep=""
 
 Yay! Trinity assembly completed for Ar!
 
+Remove intermediate files and compress final fastq files to save disk space
+
+```{r assembly_cleanup, eval=FALSE}
+# Keep only files directly needed for Trinity assembly
+setwd(mergedir)
+system("ls | grep -P 'A.*.notCombined.fastq$' | xargs -d'\n' rm")
+
+setwd("../diginorm")
+system("ls | grep -P 'A.*.abundfilt$' | xargs -d'\n' rm")
+system("ls | grep -P 'A.*.extendedFrags.fastq.keep$' | xargs -d'\n' rm")
+system("ls | grep -P 'A.*.extendedFrags.fastq.keep.abundfilt$' | xargs -d'\n' rm")
+
+system("ls | grep -P 'A.*.notCombined.fastq.out.keep.abundfilt.pe$' | xargs -d'\n' rm")
+
+
+# compress diginorm directory
+setwd("../")
+system("tar -czf diginorm_fastq.tar.gz diginorm")
+
+# remove directories
+system("rm -rf diginorm merged trimclip")
+
+setwd("../scripts")
+```
 
 ## Evaluate assembly
 
 To evaluate assembly, blast transcripts against known spike-in reads
 
-```{r evaluate_assembly, eval=FALSE}
+```{r evaluate_assembly, eval=TRUE}
 # load spike-in fasta file
-library(Biostrings)
 spikein <- "../data/sim/known-sim.fasta"
[…]

### 2013-12-11 scripts/assemble_functions.R

@@ -1,4 +1,19 @@
-## Functions required by assembleR script
+## Functions for transcriptome assembly, evaluation and evaluation
+## Author: [John Stanton-Geddes](john.stantongeddes.research@uvm.edu)
+## Date: 2013-12-11
+
+######################################################
+## sample.fasta(filein, X)
+##
+## Usage:
+##
+##     sample.fasta(filein, X)
+##
+##     where filein is a fasta file
+##     and X is the number of transcripts to sample
+##     output is a fasta file containing the X randomly
+##     sampled transcripts. Also outputs mean length of the
+##     sampled transcripts
 
 
 sample.fasta <- function(filein, X) {
@@ -20,4 +35,161 @@ sample.fasta <- function(filein, X) {
 
     # Write mean length to file
     cat(mean(width(samp.file)), '\n', file="mean-length-fasta.txt")
+}
+
+
+#####################################################
+## assembly.spikein.eval(fasta, blastout)
+##
+## Usage: 
+##
+##     assembly.spikein.eval(fasta, blastout)
+##
+##     where fasta is the 'spikein' fasta file and
+##     blastout is the result of running blastn of the
+##     transcriptome assembly against the known seqeunces
+##     specifying `-outfmt 6` to allow reading with read.blast() 
+##
+## Description:
+##     Script to evaluate simulated transcriptome assembly
+##     Takes in list of known 'spikein' fasta transcripts
+##     genes and results of BLAST output of transcriptome assembly
+##     against these spikein transcripts 
+##
+##     Reports the starting number of genes, number of genes
+##     captured in assembly, number of genes missing in assembly
+##     and the average length of each of genes in these groups.
+##     Also produces histogram showing distribution of captured
+##     versus missing genes.
+##
+#####################################################
+
+assembly.spikein.eval <- function(fasta, blastout) {
+    ##-------------Load libraries--------------------------------
+                                        #source("http://bioconductor.org/biocLite.R")
+                                        #biocLite("ShortRead")
+    require(Biostrings)
+    require(stringr)
+    require(RFLPtools)
+    require(ggplot2)
+
+                                        # read fasta file
+    known <- readDNAStringSet(fasta)
+                                        # extract names
+    known.names <- names(known)
+                                        # strip names
+                                        #kn.split <- str_split_fixed(known.names, "\\$", 2)
+                                        #known.names <- kn.split[,1]
+    head(known.names)
+
[…]

### 2013-12-12 scripts/evaluate-spikein.Rmd
Continuing work on assembly script
  Bugs fixed
  - mistake to remove `trimclip` reads - need these for expression quantification!!!

  Additions
  - merge chimeric and redundant transcript using CAP3
  - remove *in silico* contigs from A22 assembly
  - gene expression quantification using `sailfish`

## Check 'spike-in'

Evaluate whether any BLAST hits to 'spike-in' reads from Trinity transcriptome assembly using all reads, but *not* including *in silico spike-in reads`

```{r evaluate_assembly, eval=TRUE}
# load spike-in fasta file
spikein <- "../data/sim/known-sim.fasta"
spikeinfa <- readDNAStringSet(spikein)
length(names(spikeinfa))

# make BLAST database from spike-in fasta
system(paste("makeblastdb -dbtype nucl -in ", spikein, sep=""))

# BLAST A22 assembled transcripts (including spike-in reads) against database
transcripts <- "../results/trinity-full/Trinity.fasta"
system(paste("blastn -query ", transcripts, " -db ", spikein, " -outfmt 6 -out blast_full.txt",sep=""))

# evaluate
blastfull <- "blast_full.txt"
try(assembly.spikein.eval(spikein, blastfull))
```

No hits!

### 2013-12-12 scripts/ApTran_assemble.R

@@ -11,9 +11,11 @@ Requires the following programs installed and available on path:
   - requires [cutadapt](https://code.google.com/p/cutadapt/) and [FastQC](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
 * [khmer](https://github.com/ged-lab/khmer) to perform digital normalization and filtering
 * [Trinity](http://trinityrnaseq.sourceforge.net/) for transcriptome assembly
+* [CAP3](http://seq.cs.iastate.edu/cap3.html) for removing chimeric sequences and merging redundant sequences
+* [sailfish](http://www.cs.cmu.edu/~ckingsf/software/sailfish/index.html) for expression Quantification
 
 
-```{r setup}
+```{r setup, echo=FALSE}
 # Global settings
 options(stringAsFactors=FALSE)
 
@@ -21,6 +23,7 @@ options(stringAsFactors=FALSE)
 library(ggplot2)
 library(knitr)
 library(Biostrings)
+library(stringr)
 
 # knitr options
 opts_chunk$set(cache=TRUE)
@@ -140,7 +143,7 @@ for (j in 1:length(clippedR1)) {
 Normalize reads separately for samples from each colony using [khmer](https://github.com/ged-lab/khmer) software
 
 
-```{r A22_diginorm, eval=TRUE}
+```{r A22_diginorm, eval=FALSE}
 # set PYTHONPATH to khmer module
 system("export PYTHONPATH=/opt/software/khmer/python")
 
@@ -176,7 +179,7 @@ message("Done with diginorm on A22 notCombined reads: ", Sys.time())
 Trim low abundance parts of high coverage reads - these are likely erroneous.
 Note that this will orphan some reads with poor quality partners
 
-```{r A22_diginorm_trim, eval=TRUE}
+```{r A22_diginorm_trim, eval=FALSE}
 message("Trim low abundance k-mers: ", Sys.time())
 system("/opt/software/khmer/scripts/filter-abund.py -V ../data/diginorm/A22-diginorm-C20k20.kh *.keep")
 
@@ -206,7 +209,7 @@ system("mv *notCombined.fastq.out.keep.abundfilt ../data/diginorm/")
 Normalize reads separately for samples from each colony using [khmer](https://github.com/ged-lab/khmer) software
 
 
[…]

### 2013-12-12 scripts/assemble_functions.R

@@ -66,24 +66,21 @@ sample.fasta <- function(filein, X) {
 
 assembly.spikein.eval <- function(fasta, blastout) {
     ##-------------Load libraries--------------------------------
-                                        #source("http://bioconductor.org/biocLite.R")
-                                        #biocLite("ShortRead")
+    #source("http://bioconductor.org/biocLite.R")
+    #biocLite("ShortRead")
     require(Biostrings)
     require(stringr)
     require(RFLPtools)
     require(ggplot2)
 
-                                        # read fasta file
+    # read fasta file
     known <- readDNAStringSet(fasta)
-                                        # extract names
+    # extract names
     known.names <- names(known)
-                                        # strip names
-                                        #kn.split <- str_split_fixed(known.names, "\\$", 2)
-                                        #known.names <- kn.split[,1]
     head(known.names)
 
-                                        # read blastn file
-    blast.res <- read.blast(blastout)
+    # read blastn file
+    suppressWarnings(blast.res <- read.blast(blastout))
     head(blast.res)
     dim(blast.res)
     subject.id.names <- unique(blast.res$subject.id)
@@ -94,22 +91,22 @@ assembly.spikein.eval <- function(fasta, blastout) {
 
     ##-----------Directory for results---------------------------------------
 
-                                        # prefix for output files
+    # prefix for output files
     pre <- str_split_fixed(blastout, "\\.", 2)[,1]
 
-                                        # make output directory
+    # make output directory
     system(paste("mkdir -p ", pre,"_results", sep=""))
     outdir <- paste(pre,"_results/", sep="")
 
     ##------------Evaluate assembly against known starting sequences-----------
 
-                                        # get names of known genes that were mapped in assembly
+    # get names of known genes that were mapped in assembly
     captured <- which(known.names %in% subject.id.names)
     length(captured)
     captured.names <- known.names[captured]
 
-                                        # If some query transcripts are not mapped in assembly, report
-                                        # differences in mean length of missing versus assembled transcripts
+    # If some query transcripts are not mapped in assembly, report
+    # differences in mean length of missing versus assembled transcripts
 
     if(length(captured.names) < length(known.names)) {
         known.length <- width(known)
@@ -121,7 +118,7 @@ assembly.spikein.eval <- function(fasta, blastout) {
         ml.miss <- mean(known.length.missing)
         num.miss <- length(known.length.missing)
         
[…]

### 2014-02-10 ApRxN-shinyapp/server.R
added shiny app to interactively display expression levels for selected transcripts

@@ -1,0 +1,43 @@
+library(shiny)
+library(ggplot2)
+library(stringr)
+library(plyr)
+
+# load data
+gxdata <- read.csv("Ap.dt.csv")
+
+# Define server logic required to plot variables 
+shinyServer(function(input, output) {
+  
+  # Caption text
+  captionText <- reactive({
+    paste("Reaction norm for ", input$searchtext, "related genes")
+  })
+  
+  # Return the formula text for printing as a caption
+  output$caption <- renderText({
+    captionText()
+  })
+  
+  # Subset data to transcripts that have a match to 'searchtext' in best.hit.to.nr or GO.Biological.Process
+  datasub <- reactive({
+    sd <- gxdata[union(grep(input$searchtext, gxdata$best.hit.to.nr), grep(input$searchtext, gxdata$GO.Biological.Process)), ]
+  })
+  
+  # Extract GO annotation to display
+  output$GOannotation <- renderPrint({
+    foo <- datasub()
+    unlist(str_split(foo[1, "GO.Biological.Process"], " ", n=2))[2]
+  })
+  
+  # Generate a plot of the requested variable against temperature
+  output$RxNplot <- renderPlot({
+    gdata <- datasub()
+    #cat("gdata", dim(gdata), "\n")
+    g <- ggplot(gdata, aes(x=val, y=exp.scaled, group=Transcript)) + 
+      geom_smooth(aes(colour = exp_type)) + 
+      facet_grid(. ~ colony)
+    print(g)
+  })
+  
+})

### 2014-02-10 ApRxN-shinyapp/ui.R

@@ -1,0 +1,25 @@
+library(shiny)
+
+# Define UI for selecting GO term
+shinyUI(pageWithSidebar(
+    
+  # Application title
+  headerPanel("Aphaenogaster thermally-responsive genes"),
+  
+  # Sidebar to select which GO term to plot transcripts for
+  sidebarPanel(
+    textInput("searchtext", "Search term:", value = "response to stress"),
+    submitButton(text = "Apply Changes"),
+    br(),
+    p("Search terms to try: heat shock protein, behavior, phagocytosis, immune system, GO:0006950")
+  ),
+  
+  # Display caption and plot of requested GO term
+  mainPanel(
+    h3(textOutput("caption")),
+    
+    plotOutput("RxNplot"),
+    
+    verbatimTextOutput("GOannotation")
+    )
+))

### 2014-02-10 ApTranscriptome_TR.R
Moved ApTranscriptome_TR.R to root directory - makes more sense to call scripts, load data and save results from here.

@@ -3,6 +3,8 @@ Thermal reactionome of the common ant species *Aphaenogaster*
   
 **Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
 
+**February 10, 2014**
+
 **Technical Report No. 3**
 
 **Department of Biology**
@@ -27,13 +29,12 @@ library(plyr)
 # Load Bioconductor libraries
 #source("http://bioconductor.org/biocLite.R")
 #biocLite("topGO")
-#biocLite("qvalue")
 #biocLite("Rgraphviz")
 library(topGO) 
 suppressMessages(library(Rgraphviz))
 
 # load custom functions
-source("RxNseq.R")
+source("scripts/RxNseq.R")
 ```
 
 ## Summary ##
@@ -174,9 +175,9 @@ Then, for each sample, run the following command:
 
 Or, with a loop:                                                 
                                                  
-```{r sailfish, eval=TRUE, echo=TRUE, cache=TRUE}
+```{r sailfish, eval=FALSE, echo=TRUE, cache=TRUE}
 # directory containing trimmed reads
-readdir <- "../data/ind_files/" 
+readdir <- "data/ind_files/" 
 
 # list of reads in each of four trimmed classes
 readlist <- list.files(readdir)
@@ -198,20 +199,20 @@ for (j in 1:length(samples)) {
     samp.paired.r <- paste(readdir, paired.right[samp.pos], sep="")
     samp.unpaired.l <- paste(readdir, unpaired.left[samp.pos], sep="")
     samp.unpaired.r <- paste(readdir, unpaired.right[samp.pos], sep="")
-    sailfishcmd <- paste("sailfish quant -i ../results/trinity-full/sailfish-index -o ../results/trinity-full/sailfish-expression/", quantdir, " --reads ", samp.paired.l, " ", samp.paired.r, " ", samp.unpaired.l, " ", samp.unpaired.r, " -p 4", sep="")
+    sailfishcmd <- paste("sailfish quant -i results/trinity-full/sailfish-index -o results/trinity-full/sailfish-expression/", quantdir, " --reads ", samp.paired.l, " ", samp.paired.r, " ", samp.unpaired.l, " ", samp.unpaired.r, " -p 4", sep="")
     print(sailfishcmd)
-#    system(sailfishcmd)
+    system(sailfishcmd)
[…]

### 2014-02-13 ApTranscriptome_TR.R
Updates to script - identify transcripts with bimodal expression

@@ -54,8 +54,23 @@ This script is completely reproducible assuming that R, `knitr` and the other re
 
 The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.
 
+## Data ##
+
+The raw Illumina fastq files are available from [https://minilims1.uvm.edu/BCProject-26-Cahan/_downloads/trimmomatic_output.tar.gz]
+
+We download and unzip the Trimmomatic filtered data available from [https://minilims1.uvm.edu/BCProject-26-Cahan/_downloads/trimmomatic_output.tar.gz]
+
+~~~
+# download
+wget --no-check-certificate https://minilims1.uvm.edu/BCProject-26-Cahan/_downloads/trimmomatic_output.tar.gz
+
+# move and unzip
+mv raw_data.tar.gz data/.
+tar -zxvf raw_data.tar.gz
+~~~
+
 ```{r download, echo = TRUE, results = "hide"}
-### Transcriptome assembly
+### Download 
 
 ### Expression data
 
@@ -330,6 +345,11 @@ Many of these are likely false positives, so I adjusted P-values using false dis
 ```{r fdr}
 RxNout$qval <- p.adjust(RxNout$pval, method = "fdr")
 
+# Plot FDR values against initial pvalues
+par(mfrow = c(2,1))
+hist(RxNout$pval)
+hist(RxNout$qval)
+
 # subset to significant transcripts
 signif.transcripts <- RxNout[which(RxNout$qval < 0.05), ]
 ```
@@ -370,70 +390,59 @@ str(colony.transcripts.ann)
 write.table(colony.transcripts.ann, file = paste(resultsdir, "Ap_colony_transcripts_GO.txt", sep=""), quote = FALSE, sep = "\t", row.names = FALSE)
 ```
 
-Identify shape of curve:
+For responsive transcripts, identify by shape of response:
 
-* increase - exrpression increases with temperature
-* decrease - expression decreases with temperature
-* concave - expression greatest at intermediate temperatures
-* convex - expression lowest at intermediate temperatures
+* High - increase expression with temperature
+* Low - decrease expression with temperature
+* Intermediate - maximum expression at intermediate temperatures (14 - 28C)
[…]

### 2014-02-14 ApTranscriptome_TR.R
Fixed bug with 'select_FDR' function keeping too many transcripts for topGO analysis. Updated expression type function for plotting

@@ -404,7 +404,7 @@ The 'Bimodal' class is of special interest as these transcripts may be those tha
 setkey(TPM.dt.sub, Transcript)
 Ap.dt <- TPM.dt.sub[responsive.transcripts.ann]
 
-Ap.max.min.exp <- ddply(Ap.dt, .(Transcript), function(df1) {
+Ap.exp_type <- ddply(Ap.dt, .(Transcript), function(df1) {
     lmout <- lm(TPM ~ val + I(val^2), data = df1)
     vals <- c(0, 3.5, 10, 14, 17.5, 21, 24.5, 28, 31.5, 35, 38.5)
     newdf <- data.frame(val = vals)
@@ -417,32 +417,35 @@ Ap.max.min.exp <- ddply(Ap.dt, .(Transcript), function(df1) {
 
     # report coefficients
     #coef(lmout)
-    shape = if(coef(lmout)['val'] > 0 & coef(lmout)['I(val^2)'] > 0) "High" else {
+    exp_type = if(coef(lmout)['val'] > 0 & coef(lmout)['I(val^2)'] > 0) "High" else {
         if(coef(lmout)['val'] < 0 & coef(lmout)['I(val^2)'] < 0) "Low" else {
             if(coef(lmout)['val'] > 0 & coef(lmout)['I(val^2)'] < 0) "Intermediate" else {
                 "convex"}}}
 
-    # for transcripts with convex shape, check if expression is truly bimodal
-    if(shape == "convex") {
+    # for transcripts with convex exp_type, check if expression is truly bimodal
+    if(exp_type == "convex") {
         if(max(pout[pout$val <= 10, "exp"]) > 2*sd(pout$exp) &
-           max(pout[pout$val >= 31.5, "exp"]) > 2*sd(pout$exp)) shape = "Bimodal" else {
+           max(pout[pout$val >= 31.5, "exp"]) > 2*sd(pout$exp)) exp_type = "Bimodal" else {
            # linear increase?
-               if(max.val > min.val) shape = "High" else shape = "Low"
+               if(max.val > min.val) exp_type = "High" else exp_type = "Low"
            }
     }
            
     # return values
     return(c(max.val = vals[which(pout$exp == max(pout$exp))],
              min.val = vals[which(pout$exp == min(pout$exp))],
-             shape = shape))
[…]

### 2014-03-25 ApTranscriptome_TR.R
Small changes

@@ -1,6 +1,6 @@
 Thermal reactionome of the common ant species *Aphaenogaster*
 ================================================================
-  
+   
 **Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
 
 **March 20, 2014**

### 2014-03-25 ApTranscriptome_TR.R
Started new branch to examine differences in mean and shape of reaction norm for each transcript between colonies, following method in Murren et al 2014 AmNat

@@ -578,22 +578,6 @@ pandoc.table(exp_type.table, style="rmarkdown", caption = "Number of transcripts
 Table 4 shows the number of transcripts that fall into each expression type for each each colony. The totals for each colony include the `r nrow(temperature.transcripts)` transcripts that have consistent temperature responses between the two colonies. 
 
 
-```{r RxN_by_colony, echo=FALSE, eval=FALSE}
-setkey(TPM.dt.sub, Transcript)
-interaction.transcripts.TPM <- TPM.dt.sub[interaction.transcripts]
-setkey(interaction.transcripts.TPM, colony)
-
-# RxNseq for A22 only - set 'threshold' to 0.99 so all coefficients reported
-model2 <- "TPM ~ val + I(val^2)"
-A22.RxNout <- RxNseq(interaction.transcripts.TPM["A22"], model = model2, threshold = 1)
-str(A22.RxNout)
-
-# RxNseq for Ar only - set 'threshold' to 0.99 so all coefficients reported
-Ar.RxNout <- RxNseq(interaction.transcripts.TPM["Ar"], model = model2, threshold = 1)
-str(Ar.RxNout)
-```
-
-
 ## Shiny interactive web-app
 
 To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package. The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.
@@ -610,6 +594,24 @@ write.csv(responsive.transcripts.TPM, file = "ApRxN-shinyapp/responsive.transcri
 
 
 ## Colony-level comparison ##
+
+In this section, I examine differences in thermal reaction norms of transcripts between *Ar* and *A22* using the categories as defined by `r citep("10.1086/675302")`:
+
+- *Offset, O*: overall difference in the mean expression value across all temperatures
+- *Slope, S*: difference in overall slope
+- *Curvature, C*: average difference in curvature of the reaction norm
+- *Wiggle, W*: variability in shape not captured by the previous three measures
+
+```{r offset, echo=TRUE, eval=TRUE}
+A22offset <- ddply(TPM.dt.sub, .(colony, Transcript), summarise, offset = mean(TPM))
+
+```
+
+
+
+
+
+
 
 ### Plasticity versus constitutive expression
 

### 2014-03-26 ApTranscriptome_TR.R
Finally figured out how to use .Rmd files with ESS by using [polymode](https://github.com/vitoshka/polymode).
With this change, can work directly with .Rmd file and removed .R file

@@ -1,1263 +1,0 @@
-Thermal reactionome of the common ant species *Aphaenogaster*
-================================================================
-   
-**Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
-
-**March 20, 2014**
-
-**Technical Report No. 3**
-
-**Department of Biology**
-
-**University of Vermont**
-
-```{r setup, echo=FALSE, results='hide', message = FALSE}
-# Global settings
-options(stringsAsFactors=FALSE)
-
-# Load libraries
-library(R.utils)
-library(ggplot2)
-library(knitr)
-library(knitcitations)
-library(pander)
-library(stringr)
-library(data.table)
-library(RCurl)
-library(plyr)
-
-# Load Bioconductor libraries
-#source("http://bioconductor.org/biocLite.R")
-#biocLite("topGO")
-#biocLite("Rgraphviz")
-library(topGO) 
-suppressMessages(library(Rgraphviz))
-
-# load custom functions
-source("scripts/RxNseq.R")
-
-# knitr options
-opts_chunk$set(cache=TRUE)
-
-# directory to save results
-resultsdir <- "results/"
-```
-
-
-## Summary ##
-  
-In this technical report, which accompanies the manuscript **Thermal reactionome of a common ant species** (Stanton-Geddes et al., in press), we:
-
-1. describe the *de novo* assembly of the transcriptome for two ant colonies with in the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
-2. identify thermally-responsive genes
-3. perform gene set enrichment analysis
-
-This script is completely reproducible assuming that R, `knitr` and the other required libraries (listed within the source document) are installed on a standard linux system using the following:
-    
-    Rscript -e "library(knitr); knit('ApTranscriptome_TR.Rmd')"
-
-The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.
-
-
-
-## Data ##
-
-The Illumina fastq files are available from [https://minilims1.uvm.edu/BCProject-26-Cahan/downloads.html]. The Trimmomatic filtered fastq files should be downloaded, uncompressed and moved to the appropriate directory using the following code.
-
-~~~
[…]

### 2014-03-27 ApTranscriptome_TR.Rmd

Thermal reactionome of the common ant species *Aphaenogaster*
================================================================

**Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)

**March 20, 2014**

**Technical Report No. 3**

**Department of Biology**

**University of Vermont**

```{r setup, echo=FALSE, results='hide', message = FALSE}
# Global settings
options(stringsAsFactors=FALSE)

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
library(topGO) 
suppressMessages(library(Rgraphviz))

# load custom functions
source("scripts/RxNseq.R")

# knitr options
opts_chunk$set(cache=TRUE)

# directory to save results
resultsdir <- "results/"
```

## Summary ##

In this technical report, which accompanies the manuscript **Thermal reactionome of a common ant species** (Stanton-Geddes et al., in press), we:

1. describe the *de novo* assembly of the transcriptome for two ant colonies with in the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
2. identify thermally-responsive genes
3. perform gene set enrichment analysis

This script is completely reproducible assuming that R, `knitr` and the other required libraries (listed within the source document) are installed on a standard linux system using the following:

    Rscript -e "library(knitr); knit('ApTranscriptome_TR.Rmd')"

The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.

## Data ##

The Illumina fastq files are available from [https://minilims1.uvm.edu/BCProject-26-Cahan/downloads.html]. The Trimmomatic filtered fastq files should be downloaded, uncompressed and moved to the appropriate directory using the following code.

~~~
# download filtered fastq files, uncompress and move
wget --no-check-certificate https://minilims1.uvm.edu/BCProject-26-Cahan/_downloads/trimmomatic_output.tar.gz
tar -zxvf trimmomatic_output.tar.gz
mkdir -p data/
mv ind_files data/.
~~~

## Sample description ##

Two ant colonies were used for the transcriptome sequencing. The first, designated A22, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. This colony was putatively identifed as *A. picea*. The second colony, designated Ar, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection.

For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5C ranging from 0C to 38.5C, for one hour in glass tubes in a water bath. The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen). Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20e6 and 16e6 reads for the A22 and Ar samples, respectively.

## Transcriptome assembly ##

The Illumina reads were filtered using the program [Trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) `r citep("10.1093/nar/gks540")` to remove Ilumina adapter sequences and filter out bases with quality scores less than ??.

~~~
TRIMMOMATIC CODE
~~~

This filtering yielded...

These reads were combined and used in *de novo* transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`. Note that this required ??? GB RAM and ??? hours run-time and was run on the [Vermont Genetics Network](http://vgn.uvm.edu/) computing cluster.

~~~
TRINITY CODE
~~~

This assembly contained 100,381 unique components (roughly genes) in 126,172 total transcripts (Table 1).

As we were assembling two divergent colonies into a single transcriptome, we suspected that this assembly would be susceptible to known problems of errors during assembly (e.g. chimeric transcripts that are fusions of two transcripts) and redundancy `r citep("10.1186/1471-2164-14-328")`. To account for this, we performed two post-assembly processing steps.

First, we ran the program [cap3](http://seq.cs.iastate.edu/) `r citep("10.1101/gr.9.9.868")` setting the maximum gap length and band expansion size to 50 `-f 50 -a 50`, no end clipping as the reads were already filtered `k 0`, requiring 90% identity for assembly, and a minimum overlap length of 100 bp `-o 100`. The percent identity threshold of 90% was chosen to liberally collapse orthologous contigs from the two colonies that may have been assembled separately.

    cap3 Trinity.fasta -f 50 -a 50 -k 0 -p 90 -o 100 > Trinity_cap3.out

The output of `cap3` gives assembled "contigs" and unassembled "singlets" that were concatenated into a single file.

    # check the number of contigs clustered
    grep -c "Contig" Trinity.fasta.cap.contigs
    grep -c "comp" Trinity.fasta.cap.singlets
    # compare to contigs from Trinity output
    grep -c "comp" Trinity.fasta

    # Combine contigs and singlets from CAP3
    cat Trinity.fasta.cap.contigs Trinity.fasta.cap.singlets > Trinity_cap3.fasta

The output file "Trinity.fasta.cap.info" gives specific information on which contigs were collapsed.

Subsequent to running `cap3`, we ran [uclust](http://drive5.com/usearch/manual/uclust_algo.html) to cluster sequences completely contained within longer sequences, again specificing a 90% identity cutoff for clustering.

    # sort
    uclust --sort Trinity_cap3.fasta --output Trinity_cap3_sorted.fasta
    # cluster by 90% similarity threshold
    uclust --input Trinity_cap3_sorted.fasta --uc Trinity_cap3_uclust.out --id 0.90
    # convert uclust to fasta format
    uclust --uc2fasta Trinity_cap3_uclust.out --input Trinity_cap3_uclust.fasta

These post-processing step removed `r round((126172-105536)/126172 * 100, 0)`% of the initial reads (Table 1).

```{r assemstats, results = "asis", echo=FALSE}
# make table
trinity <- c("126,172", "100,389,539", "358", "795", "16,201", "1,631")
reduced <- c("105,536", "62,648,997", "320", "593", "15,491", "895")

assemstats <- rbind(trinity, reduced)
colnames(assemstats) <- c("Total contigs", "Total length", "Median contig size", "Mean contig size", "N50 contig", "N50 Length")

pandoc.table(assemstats, style="rmarkdown", caption = "Statistics for Trinity and cap3+uclust reduced transcriptome assemblies")
```

Running Trinity is time and computationally-intensive, so the final filtered assembly can be downloaded from [http://johnstantongeddes.org/assets/files/Trinity_filtered_assembly.tgz]

~~~
# download filtered Trinity assembly, uncompress and move
wget http://johnstantongeddes.org/assets/files/Trinity_filtered_assembly.tgz
tar -zxvf trimmomatic_output.tar.gz
mkdir -p results/
mkdir -p results/trinity-full/
mv Trinity_cap3_uclust.fasta /results/trinity-full/.
~~~

## Transcriptome annotation ##

Annotation was performed by uploading the reduced assembly "Trinity_cap3_uclust.fasta" to the web-based annotation program [FastAnnotator](http://fastannotator.cgu.edu.tw/index.php) `r citep("10.1186/1471-2164-13-S7-S9")`.

#Results are available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).

This annotation file can be read directly to R:

```{r download, echo = TRUE, results = "hide"}
### Annotation file
# from either AWS or GoogleDrive
annotationURL <- getURL("http://johnstantongeddes.org/assets/files/Aphaeno_transcriptome_AnnotationTable.txt")
#a2 <- getURL("https://googledrive.com/host/0B75IymziRJ_9Tlg1U1Vxbjk1bzg") # GoogleDrive link

annotationfile <- read.csv(textConnection(annotationURL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotationfile)

# Convert to data.table
annotationtable <- data.table(annotationfile)
head(annotationtable)
```

Transcriptome annotations are nearly impossible to visualize in a meaningful way. For lack of better ideas, I made a word cloud:

```{r wordcloud, cache=TRUE, eval=FALSE}
library(tm)
library(wordcloud)
library(RColorBrewer)
library(stringr)

# subset and extract unique best hits to NCBI nr database
ann <- unique(annotationfile$best.hit.to.nr)
annsplit <- str_split_fixed(ann, pattern = " ", n=2)
annvec <- annsplit[,2]

# create Corpus of terms. convert to dataframe for plotting wordcloud
ap.corpus <- Corpus(VectorSource(annvec))
ap.corpus <- tm_map(ap.corpus, removePunctuation)
ap.corpus <- tm_map(ap.corpus, tolower)
ap.corpus <- tm_map(ap.corpus, function(x) removeWords(x, stopwords("english")))
tdm <- TermDocumentMatrix(ap.corpus)
m <- as.matrix(tdm)
v <- sort(rowSums(m),decreasing=TRUE)
d <- data.frame(word = names(v),freq=v)
pal <- brewer.pal(4, "Dark2")
pal <- pal[-(1:2)]

#png(paste(resultsdir, "wordcloud.png", sep=""), width=1280,height=800)
wordcloud(d$word,d$freq, scale=c(8,.3),min.freq=2,max.words=100, random.order=T, rot.per=.15, colors=pal, vfont=c("sans serif","plain"))
#dev.off()
```

This wordcloud shows that we do not know much about the transcriptome - mostly hypothetical or predicted proteins. I removed these uninformative words and generated a new wordcloud.

```{r wordcloud2, cache=TRUE}
# remove 'protein', 'predicted' and 'hypothetical'
rmwords <- c("hypothetical", "protein", "isoform", "subunit", "family", "putative", "partial", "domain", "containing", "predicted", "domaincontaining", "uncharacterized")
ap.corpus.rmwords <- tm_map(ap.corpus, removeWords, rmwords)
tdm <- TermDocumentMatrix(ap.corpus.rmwords)
m <- as.matrix(tdm)
v <- sort(rowSums(m),decreasing=TRUE)
d <- data.frame(word = names(v),freq=v)
pal <- brewer.pal(8, "Dark2")
pal <- pal[-(1:2)]
#png(paste(resultsdir, "wordcloud2.png", sep=""), width=1280,height=800)
wordcloud(d$word,d$freq, scale=c(8,.3),min.freq=2,max.words=100, random.order=T, rot.per=.15, colors=pal, vfont=c("sans serif","plain"))
#dev.off()
```

## Identification of thermally-responsive genes ##

### Quantify gene expression ###

I quantified gene expression using [sailfish](http://www.cs.cmu.edu/~ckingsf/software/sailfish/index.html). To run this program, first make sure that PATHs to the software libraries are set up correctly:

    export LD_LIBRARY_PATH=/opt/software/Sailfish-0.6.2-Linux_x86-64/lib:$LD_LIBRARY_PATH
    export PATH=/opt/software/Sailfish-0.6.2-Linux_x86-64/bin:$PATH

Then build the index of the assembly:

    sailfish index -t Trinity_cap3_uclust.fasta -o sailfish-index -k 20 -p 4

Once this is done, quantify expression for the Trimmomatic filtered reads from each colony-treatment sample separately. Note that for each sample, there are three four filtered read files:

- paired.left.fastq
- paired.right.fastq
- unpaired.left.fastq
- unpaired.right.fastq

Make a directory for the expression values

    mkdir sailfish-expression

Then, for each sample, run the following command:

    sailfish -i sailfish-index -o sailfish-expression/A22-0 --reads A22-0_ATCACG.paired.left.fastq A22-0_ATCACG.paired.right.fastq A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq -p 4

Or, using a loop in R:

```{r sailfish, eval=TRUE, echo=TRUE, cache=TRUE}
# directory containing trimmed reads
readdir <- "data/ind_files/" 
# sailfish index directory
sfindex <- "results/trinity-full/sailfish-index"
# sailfish expression directory
sfexpressionroot <- "results/trinity-full/sailfish-expression/"

# list of reads in each of four trimmed classes
readlist <- list.files(readdir)
(paired.left <- readlist[grep(".\\.paired.left.fastq$", readlist)])
(paired.right <- readlist[grep("\\.paired.right.fastq$", readlist)])
(unpaired.left <- readlist[grep("unpaired.left.fastq$", readlist)])
(unpaired.right <- readlist[grep("unpaired.right.fastq$", readlist)])

# Loop across each sample and quantify expression

# NOTE - samples listed in same order as given by the above lists
samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")

for (j in 1:length(samples)) {
    message("Start expression quantification for sample ", samples[j], ": ", Sys.time())
    quantdir <- paste(sfexpressionroot, samples[j], "_quant", sep="")
    samp.pos <- grep(paste(paste(samples[j], "_", sep="")), paired.left)
    samp.paired.l <- paste(readdir, paired.left[samp.pos], sep="")
    samp.paired.r <- paste(readdir, paired.right[samp.pos], sep="")
    samp.unpaired.l <- paste(readdir, unpaired.left[samp.pos], sep="")
    samp.unpaired.r <- paste(readdir, unpaired.right[samp.pos], sep="")
    sailfishcmd <- paste("sailfish quant -i ", sfindex, " -o ", quantdir, " --reads ", samp.paired.l, " ", samp.paired.r, " ", samp.unpaired.l, " ", samp.unpaired.r, " -p 4", sep="")
#   print(sailfishcmd)
    system(sailfishcmd)
    message("Done with expression quantification for sample ", samples[j], ": ", Sys.time(), "\n")
}
```

This generated a directory for each sample

`r list.files(sfexpressionroot)`

and within each directory there are the following r:

`r list.files(quantdir)`

The file *quant_bias_corrected.sf* contains the following columns, following a number of header lines:

1. Transcript ID
2. Transcript Length
3. Transcripts per Million (TPM): computed as described in `r citep("10.1093/bioinformatics/btp692")`, and is meant as an estimate of the number of transcripts, per million observed transcripts, originating from each isoform.
4. Reads Per Kilobase per Million mapped reads (RPKM): classic measure of relative transcript abundance, and is an estimate of the number of reads per kilobase of transcript (per million mapped reads) originating from each transcript.

The TPM column for each sample was extracted and combined into a matrix for each colony.

```{r load_expression_data, eval=TRUE, echo=FALSE, results='hide'}
# read in each file using loop
samples <- c("A22-0", "A22-3", "A22-7", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "Ar-0", "Ar-3", "Ar-7", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38")

for (j in 1:length(samples)) {
    samp <- samples[j]
    trtval <- as.numeric(str_split_fixed(samp, "-", 2)[2])
    outpre <- gsub("-", "_", samp)
    outname <- paste(outpre, "_quant", sep="")
    read.sailfish.quant(filein=paste(sfexpressionroot, samp, "_quant/quant_bias_corrected.sf", sep=""), outname=outname, samp = samp, trtval = trtval)
}

# combine into long format
A22.TPM <- rbind(A22_0_quant, A22_3_quant, A22_7_quant, A22_10_quant, A22_14_quant, A22_17_quant, A22_21_quant, A22_24_quant, A22_28_quant, A22_31_quant, A22_35_quant, A22_38_quant)
str(A22.TPM)

# convert to data.table
A22.TPM <- data.table(A22.TPM)
setkey(A22.TPM, Transcript, val)

# set "trt" to true values - truncated in file names for convenience

A22.TPM[val==3,val:=3.5]
A22.TPM[val==10,val:=10.5]
A22.TPM[val==17,val:=17.5]
A22.TPM[val==24,val:=24.5]
A22.TPM[val==31,val:=31.5]
A22.TPM[val==38,val:=38.5]
head(A22.TPM)
str(A22.TPM)

# Repeat for Ar
Ar.TPM <- rbind(Ar_0_quant, Ar_3_quant, Ar_7_quant, Ar_10_quant, Ar_14_quant, Ar_17_quant, Ar_21_quant, Ar_24_quant, Ar_28_quant, Ar_31_quant, Ar_35_quant, Ar_38_quant)
Ar.TPM <- data.table(Ar.TPM)
setkey(Ar.TPM, Transcript, val)
Ar.TPM[val==3,val:=3.5]
Ar.TPM[val==10,val:=10.5]
Ar.TPM[val==17,val:=17.5]
Ar.TPM[val==24,val:=24.5]
Ar.TPM[val==31,val:=31.5]
Ar.TPM[val==38,val:=38.5]
head(Ar.TPM)
str(Ar.TPM)

tables()
```

Note that expression levels at each temperature treatment are highly correlated between the two colonies.

```{r exp_correlations, echo=FALSE, results='asis'}
# Make Rmarkdown pandoc table of correlations among expression levels between colonies at each temp

temp <- c(0, 3.5, 7, 10.5, 14, 17.5, 21, 24.5, 28, 31.5, 35, 38.5)
cors <- c(round(cor(Ar_0_quant$TPM, A22_0_quant$TPM), 2), round(cor(Ar_3_quant$TPM, A22_3_quant$TPM), 2), round(cor(Ar_7_quant$TPM, A22_7_quant$TPM), 2), round(cor(Ar_10_quant$TPM, A22_10_quant$TPM), 2), round(cor(Ar_14_quant$TPM, A22_14_quant$TPM), 2), round(cor(Ar_17_quant$TPM, A22_17_quant$TPM), 2), round(cor(Ar_21_quant$TPM, A22_21_quant$TPM), 2), round(cor(Ar_24_quant$TPM, A22_24_quant$TPM), 2), round(cor(Ar_28_quant$TPM, A22_28_quant$TPM), 2), round(cor(Ar_31_quant$TPM, A22_31_quant$TPM), 2), round(cor(Ar_35_quant$TPM, A22_35_quant$TPM), 2), round(cor(Ar_38_quant$TPM, A22_38_quant$TPM), 2))

cortable <- cbind(temp, cors)

pandoc.table(cortable, style="rmarkdown", caption = "correlations between colonies at each temperature treatment")
```

## Identification of thermally-responsive genes

To identify transcripts (roughly equivalent to genes) that show thermal responsiveness, I fit the following linear model to each transcript:

$$ TPM = \beta_0 + \beta_1(colony) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(colony * temp) + \\beta_5(colony * temp^2) + \epsilon $$

where TPM is transcripts per million.

**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the analysis.**

```{r RxN, echo=TRUE, eval=TRUE, cache=TRUE}
A22.TPM[,colony:="A22"]
Ar.TPM[,colony:="Ar"]
TPM.dt <- rbind(A22.TPM, Ar.TPM)
TPM.dt$colony <- as.factor(TPM.dt$colony)
str(TPM.dt)

setkey(TPM.dt, val)
TPM.dt.sub <- TPM.dt[val != 7] 
unique(TPM.dt.sub$val)

# define model for RxN function
model <- "TPM ~ colony + val + I(val^2) + colony:val + colony:I(val^2)"

# identify responsive transcripts
RxNout <- RxNseq(f = TPM.dt.sub, model = model, threshold = 0.05)

save.image("RxN_combined_results.RData")
```

Of the `r nrow(RxNout)` transcripts, `r length(which(RxNout$pval < 0.05))` have models with P < 0.05.

Many of these are likely false positives, so I adjusted P-values using false discovery rate (FDR) to identify only those transcripts with less than 5% FDR as significant.

```{r fdr}
RxNout$padj <- p.adjust(RxNout$pval, method = "fdr")
# Plot FDR values against initial pvalues
par(mfrow = c(2,1))
hist(RxNout$pval)
hist(RxNout$padj)

# subset to significant transcripts
signif.transcripts <- RxNout[which(RxNout$padj < 0.05), ]
```

## Functional annotation ##

In the previous section, I identified transcripts that show significant responses in expression. Next, I add gene annotation and ontology information to these transcripts.

```{r annotation}
# add annotation information
setkey(annotationtable, Sequence.Name)
signif.transcripts <- data.table(signif.transcripts)
setkey(signif.transcripts, Transcript)
signif.transcripts <- annotationtable[signif.transcripts]
setnames(signif.transcripts, "Sequence.Name", "Transcript")
```

## Thermally-responsive transcripts

The set of transcripts with significant expression patterns include those with expression that differs by colony, temperature and the interaction of colony and temperature. In this section, I am specifically interested in the thermally-responsive transcripts (temperature and colony x temperature) so I subset the significant transcripts to examine these.

```{r responsive_transcripts}
# transcripts that differ in expression by colony only
colony.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.colony') & is.na(signif.transcripts$'coef.val') & is.na(signif.transcripts$'coef.I(val^2)') & is.na(signif.transcripts$'coef.colony:val') & is.na(signif.transcripts$'coef.colony:I(val^2)')]
# check that subset correct - no transcripts with significant temperature terms
length(which(!is.na(colony.transcripts$'coef.val') | !is.na(colony.transcripts$'coef.I(val^2)') | !is.na(colony.transcripts$'coef.colony:val') | !is.na(colony.transcripts$'coef.I(val^2)')))
str(colony.transcripts)
# order by adjusted P-value
colony.transcripts <- colony.transcripts[order(colony.transcripts$padj), ]

# remaining transcripts are thermally responsive
responsive.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.val') | !is.na(signif.transcripts$'coef.I(val^2)') | !is.na(signif.transcripts$'coef.colony:val') | !is.na(signif.transcripts$'coef.colony:I(val^2)')]
dim(responsive.transcripts)

# check that these two groups sum to total significant transcripts
stopifnot(all.equal(nrow(signif.transcripts), nrow(colony.transcripts) + nrow(responsive.transcripts)))

# split responsive_transcripts into those that have same temperature response in both colonies and 
# main effect of temperature only
temperature.transcripts <- responsive.transcripts[is.na(responsive.transcripts$'coef.colony:val') & is.na(responsive.transcripts$'coef.colony:I(val^2)')]
# order by adjusted P-value
temperature.transcripts <- temperature.transcripts[order(temperature.transcripts$padj), ]
str(temperature.transcripts)

# transcripts that have colony x temperature interaction
interaction.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.colony:val') | !is.na(signif.transcripts$'coef.colony:I(val^2)')]
# order by adjusted P-value
interaction.transcripts <- interaction.transcripts[order(interaction.transcripts$padj), ]
str(interaction.transcripts)

# check that these two groups sum to total responsive transcripts
stopifnot(all.equal(nrow(responsive.transcripts), nrow(temperature.transcripts) + nrow(interaction.transcripts)))
```

```{r table_results, echo=FALSE, results='asis'}
clist <- c('Total', 'Colony', 'Temperature', 'Temperature:Colony')
siglist <- c(nrow(signif.transcripts),
             nrow(colony.transcripts),
             nrow(temperature.transcripts),
             nrow(interaction.transcripts)
             )

sigtable <- data.frame(Coefficient = clist, "Number significant" = siglist)

pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts (of 105,536 total) at 5% FDR with expression that depends on colony, temperature or their interaction.")
```

Table 3 shows that `r round((nrow(temperature.transcripts) + nrow(interaction.transcripts))/nrow(signif.transcripts), 2)*100`% of the significant transcripts are thermally-responsive, with `r round(nrow(interaction.transcripts)/(nrow(temperature.transcripts) + nrow(interaction.transcripts)), 2)*100`% of these transcripts having a thermal-response that differs by colony.

I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.

```{r GO_contingency_test}
# GO 'response to stress' hits in responsive transcripts
responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)]

resp.stress.responsive.count <- nrow(responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)

GO.stress.Xsq <- chisq.test(GO.stress.table)
GO.stress.Xsq
```

There are also 4 heat shock related genes in the responsive transcripts, out of 130 total.

```{r grep_shock}
hsp_responsive <- responsive.transcripts[grep("shock", responsive.transcripts$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
nrow(hsp_responsive)

hsp_all <- annotationtable[grep("shock", annotationtable$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
nrow(hsp_all)
```

### Thermal-response functional types ###

The previous section simply identified the transcripts with thermally-responsive expression. In this section, I determine the shape of the expression response to temperature for each transcript. Categories of expression response are:

* High - increase expression with temperature
* Low - decrease expression with temperature
* Intermediate - maximum expression at intermediate temperatures (14 - 28C)
* Bimodal - expressed greater than two standard deviations of expression at both low and high temperatures

I do this first for the thermally-responsive transcripts where there is no interaction with colony. For the transcripts where thermal-responsive expression depends on colony, I determine the functional type of the expression response separately for each colony.

```{r expression_shape, eval=TRUE, echo=TRUE}
# merge RxN results with expression values
temperature.transcripts.TPM <- TPM.dt.sub[temperature.transcripts]

# apply the 'RxNply' function to determine maximum and minimum temperatures of expression, as well as the expression level at the optimum temperature of 19.25C
temperature.transcripts.type <- ddply(temperature.transcripts.TPM, .(Transcript), RxNply)
temperature.transcripts.type$max.val <- as.numeric(temperature.transcripts.type$max.val)
temperature.transcripts.type$min.val <- as.numeric(temperature.transcripts.type$min.val)
temperature.transcripts.type$opt.exp <- as.numeric(temperature.transcripts.type$opt.exp)
temperature.transcripts.type$exp_type <- as.factor(temperature.transcripts.type$exp_type)

# repeat for transcripts that have an interaction with colongy
# A22
A22.interaction.transcripts.type <- ddply(interaction.transcripts.TPM["A22"], .(Transcript), RxNply)
A22.interaction.transcripts.type$max.val <- as.numeric(A22.interaction.transcripts.type$max.val)
A22.interaction.transcripts.type$min.val <- as.numeric(A22.interaction.transcripts.type$min.val)
A22.interaction.transcripts.type$opt.exp <- as.numeric(A22.interaction.transcripts.type$opt.exp)
A22.interaction.transcripts.type$exp_type <- as.factor(A22.interaction.transcripts.type$exp_type)
str(A22.interaction.transcripts.type)

# Ar
Ar.interaction.transcripts.type <- ddply(interaction.transcripts.TPM["Ar"], .(Transcript), RxNply)
Ar.interaction.transcripts.type$max.val <- as.numeric(Ar.interaction.transcripts.type$max.val)
Ar.interaction.transcripts.type$min.val <- as.numeric(Ar.interaction.transcripts.type$min.val)
Ar.interaction.transcripts.type$opt.exp <- as.numeric(Ar.interaction.transcripts.type$opt.exp)
Ar.interaction.transcripts.type$exp_type <- as.factor(Ar.interaction.transcripts.type$exp_type)
str(Ar.interaction.transcripts.type)

# combine into single data.table
A22.interaction.transcripts.type$colony <- "A22"
Ar.interaction.transcripts.type$colony <- "Ar"
# first copy of temperature.transcripts.type as colony A22
temperature.transcripts.type$colony <- "A22"
# combine
responsive.transcripts.type <- rbind(A22.interaction.transcripts.type, Ar.interaction.transcripts.type, temperature.transcripts.type)
# add second copy of temperature.transcripts.type as colony Ar
temperature.transcripts.type$colony <- "Ar"
responsive.transcripts.type <- rbind(responsive.transcripts.type, temperature.transcripts.type)
responsive.transcripts.type$colony <- as.factor(responsive.transcripts.type$colony)
str(responsive.transcripts.type)
```

```{r expression_exp_type_table, eval=TRUE, echo=FALSE, results='asis'}
# table type of response
A22.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "A22", 'exp_type'])
Ar.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "Ar", 'exp_type'])

exp_type.table <- rbind(A22.type.table, Ar.type.table)
rownames(exp_type.table) <- c("A22", "Ar")

pandoc.table(exp_type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
```

Table 4 shows the number of transcripts that fall into each expression type for each each colony. The totals for each colony include the `r nrow(temperature.transcripts)` transcripts that have consistent temperature responses between the two colonies.

## Shiny interactive web-app

To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package. The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.

Export data for interactive shiny app.

```{r shiny_file}
# scale expression values 
responsive.transcripts.TPM <- TPM.dt.sub[responsive.transcripts]
responsive.transcripts.TPM[,exp.scaled:=scale(TPM), by = Transcript]
str(responsive.transcripts.TPM)
write.csv(responsive.transcripts.TPM, file = "ApRxN-shinyapp/responsive.transcripts.TPM.csv", quote = TRUE, row.names = FALSE)
```

## Colony-level comparison ##

In this section, I examine differences in thermal reaction norms of transcripts between *Ar* and *A22* using the categories as defined by `r citep("10.1086/675302")`:

- *Offset, O*: overall difference in the mean expression value across all temperatures
- *Slope, S*: difference in overall slope
- *Curvature, C*: average difference in curvature of the reaction norm
- *Wiggle, W*: variability in shape not captured by the previous three measures

```{r offset, echo=TRUE, eval=TRUE}
testset <- TPM.dt.sub[1:220, ]
testoffset <- ddply(testset, .(colony, Transcript), summarise, offset = mean(TPM))
A22meanTPM <- ddply(TPM.dt.sub, .(colony, Transcript), summarise, meanTPM = mean(TPM))

```

### Plasticity versus constitutive expression

Compare the expression levels at optimum (19.25C) between the two colonies for genes in each expression group. Specifically, are genes that are upregulated at high temperatures in A22 more highly expressed at 21C in Ar? Conversely, are genes that are upregulated at low temps more highly expressed at 21C in A22?

```{r optimum_expression_comparison, cache=TRUE, eval=FALSE}
# list of transcripts that are 'high' expressed in A22
A22_high_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$exp_type == "High"), "Transcript"]
# dataframe of transcripts from both colonies that are 'high' expressed in A22
A22_high_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% A22_high_transcripts), ]

# Compare expression at optimum temp (19.25C) between colonies using t-test
A22_high_df$colony <- as.factor(A22_high_df$colony)
boxplot(data = A22_high_df, opt.exp ~ colony)
boxplot(data = A22_high_df, log(opt.exp+10) ~ colony)
t.test(A22_high_df[which(A22_high_df$colony == "Ar"), "opt.exp"],  A22_high_df[which(A22_high_df$colony == "A22"), "opt.exp"])

# remove outlier transcripts
A22_high_df_out <- A22_high_df[-which(A22_high_df$opt.exp > 1000), ]
t.test(A22_high_df_out[which(A22_high_df_out$colony == "Ar"), "opt.exp"],  A22_high_df_out[which(A22_high_df_out$colony == "A22"), "opt.exp"])
```

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099")`. As my goal is simply to determine if expression is typically greater at 19.25C in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test

```{r A22_high_wilcoxon}
wilcox.test(A22_high_df[which(A22_high_df$colony == "A22"), "opt.exp"],  A22_high_df[which(A22_high_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

Yes - transcripts that are up-regulated at high temperatures in *A22* are more highly-expressed at 19.25C in *Ar*. Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.

Next I test the converse, that transcripts that are up-regulated at low temperatures in *Ar* are more highly-expressed at 19.25C temperatures in *A22*.

```{r Ar_low_wilcoxon}
# list of transcripts that are 'Low' expressed in Ar
Ar_low_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$exp_type == "Low"), "Transcript"]
# dataframe of transcripts from both colonies that are 'Low' expressed in Ar
Ar_low_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% Ar_low_transcripts), ]
dim(Ar_low_df)

# boxplots
boxplot(data = Ar_low_df, opt.exp ~ colony)
boxplot(data = Ar_low_df, log(opt.exp+1) ~ colony)

# Wilcoxon signed rank-test
wilcox.test(Ar_low_df[which(Ar_low_df$colony == "A22"), "opt.exp"],  Ar_low_df[which(Ar_low_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

No difference in transcript expression at 19.25C between colonies for transcripts that are up-regulated at low temperatures in *Ar*.

To confirm that there are not sample-level issues, I performed the same comparison using transcripts where I do *not* expect to see a difference in expression.

```{r intermediate_transcripts}
# list of transcripts that are 'Intermediate' expressed in A22
A22_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$exp_type == "Intermediate"), "Transcript"]
A22_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% A22_int_transcripts), ]
A22_int_df$colony <- as.factor(A22_int_df$colony)

# Wilcoxon signed rank-test
wilcox.test(A22_int_df[which(A22_int_df$colony == "A22"), "opt.exp"],  A22_int_df[which(A22_int_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)

# list of transcripts that are 'Intermediate' expressed in Ar
Ar_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$exp_type == "Intermediate"), "Transcript"]
Ar_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% Ar_int_transcripts), ]

# Wilcoxon signed rank-test
wilcox.test(Ar_int_df[which(Ar_int_df$colony == "A22"), "opt.exp"],  Ar_int_df[which(Ar_int_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

As expected, for "Intermediate" expressed transcripts there is no difference in expression at 19.25C between colonies.

### Thermal tolerance indicated by region of constant gene expression ###

The 'Intermediate' expressed transcripts are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermall stress. We hypothesized that if the more southern *Ar* colony was more thermally-tolerant than *A22*, transcripts with 'Intermediate' expression (10-30C) would be active across a wider range of temperatures. To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each colony.

```{r calc_Intermediate_variance}
# extract 'Intermediate' expressed transcripts for A22 colony
setkey(TPM.dt.sub, Transcript)
A22trlist <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$exp_type == "Intermediate"), "Transcript"]
A22.TPM.int.dt <- TPM.dt.sub[A22trlist]
setkey(A22.TPM.int.dt, colony)
A22.TPM.int.dt <- A22.TPM.int.dt["A22"]
str(A22.TPM.int.dt)

# calculate standard deviation of expression for each transcript
A22_int_sd <- ddply(A22.TPM.int.dt, .(Transcript), RxNsd.concave)

# repeat for Ar
Artrlist <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$exp_type == "Intermediate"), "Transcript"]
Ar.TPM.int.dt <- TPM.dt.sub[Artrlist]
setkey(Ar.TPM.int.dt, colony)
Ar.TPM.int.dt <- Ar.TPM.int.dt["Ar"]
str(Ar.TPM.int.dt)

Ar_int_sd <- ddply(Ar.TPM.int.dt, .(Transcript), RxNsd.concave)
```

With these values calculated, I compare the standard deviation of expression between colonies.

```{r compare_Intermediate_variance}
# T-test
t.test(A22_int_sd$exp_sd, Ar_int_sd$exp_sd)

# Plot
# prep data
A22_int_var$colony <- "A22"
Ar_int_var$colony <- "Ar"
comb_int_var <- rbind(A22_int_var, Ar_int_var)
# ggplot
g1 <- ggplot(comb_int_var, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g1 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression"))
```

Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.

### Thermal sensitivity indicated by response of bimodally-expressed transcripts ###

As the converse of the above hypothesis, a colony that is especially thermally-sensitive is likely to activate expression of molecular processes more quickly. We tested this using the same approach as for the 'Intermediate' transcripts, but using the inverse of the 'Bimodal' expressed transcripts.

```{r calc_Bimodal_variance}
# extract 'Bimodal' expressed transcripts for A22 colony
A22trlist.bim <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$exp_type == "Bimodal"), "Transcript"]
A22.TPM.bim.dt <- TPM.dt.sub[A22trlist.bim]
setkey(A22.TPM.bim.dt, colony)
A22.TPM.bim.dt <- A22.TPM.bim.dt["A22"]
str(A22.TPM.bim.dt)

# calculate standard deviation of expression for each transcript
A22_bim_sd <- ddply(A22.TPM.bim.dt, .(Transcript), RxNsd.convex)

# repeat for Ar
Artrlist.bim <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$exp_type == "Bimodal"), "Transcript"]
Ar.TPM.bim.dt <- TPM.dt.sub[Artrlist.bim]
setkey(Ar.TPM.bim.dt, colony)
Ar.TPM.bim.dt <- Ar.TPM.bim.dt["Ar"]
str(Ar.TPM.bim.dt)

Ar_bim_sd <- ddply(Ar.TPM.bim.dt, .(Transcript), RxNsd.convex)
```

Next, I compare the standard deviations of expression for each transcipt between colonies.

```{r compare_Intermediate_variance}
# t-test
t.test(A22_bim_sd$exp_sd, Ar_bim_sd$exp_sd)

# Plot
# prep data
A22_bim_var$colony <- "A22"
Ar_bim_var$colony <- "Ar"
comb_bim_var <- rbind(A22_bim_var, Ar_bim_var)
# ggplot
g2 <- ggplot(comb_bim_var, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g2 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression function"))
```

Both the T-test and plot show that there is no difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.

### Compare peak expression among colonies ###

Cumulative distribution function of peak expression for transcripts that differ in expression between *A22* and *Ar*

```{r max_exp_CDF}
png(paste(resultsdir, "CDF_expression_all.png", sep=""))
g3 <- ggplot(responsive.transcripts.type, aes(x=max.val, fill=colony)) + geom_density(alpha=0.2, position="identity")
g3 + scale_y_continuous(name="Density") +
  scale_x_continuous(name="Temperature of maximum expression")
dev.off()

# merge data.tables of transcripts with colony x temperature interaction
interaction.transcripts.type <- rbind(A22.interaction.transcripts.type, Ar.interaction.transcripts.type)
png(paste(resultsdir, "CDF_expression_interaction.png", sep=""))
g4 <- ggplot(interaction.transcripts.type, aes(x=max.val, fill=colony)) + geom_density(alpha=0.2, position="identity")
g4 + scale_y_continuous(name="Density") +
  scale_x_continuous(name="Temperature of maximum expression")
dev.off()

# faceted by type
png(paste(resultsdir, "CDF_expression_intermediate_interaction.png", sep=""))
g5 <- ggplot(interaction.transcripts.type, aes(x=max.val, group=colony)) + geom_density(alpha=0.2, position="identity")
g5 + facet_grid(. ~ exp_type) + scale_y_continuous(name="Density") +
  scale_x_continuous(name="Temperature of maximum expression")
dev.off()
```

## Gene set enrichment analysis ##

I use [topGO](http://www.bioconductor.org/packages/2.12/bioc/html/topGO.html) to perform gene set enrichment analysis (GSEA) seperately for each expression type (bimodal, intermediate, high, low).

First need to create gene ID to GO term map file

```{r geneid2go, echo=TRUE, eval=TRUE, cache=TRUE}
# create geneid2go.map file from FastAnnotator AnnotationTable.txt
geneid2GOmap(annotationfile)
```

then read map file.

```{r readmap, echo=TRUE, eval=TRUE}
# read mappings file
geneID2GO <- readMappings(file = "geneid2go.map")
str(head(geneID2GO))
```

### GSEA for thermally-responsive transcripts ###

Using this gene2GO map file, perform GSEA for:

**1) all responsive transcripts**

Use `selectFDR` function to select transcripts with adjusted P < 0.05.

```{r gsea_all}
# create geneList. note that NA values cause problems with topGO
# so set any NA to 1 as need to retain for GO analysis
Ap.geneList <- RxNout$padj
Ap.geneList[which(is.na(Ap.geneList))] <- 1
stopifnot(length(which(is.na(Ap.geneList))) == 0)
names(Ap.geneList) <- RxNout$Transcript
str(Ap.geneList)

# Function to select top genes (defined above)
selectFDR <- function(padj) {
    return(padj < 0.05)
}

# create topGOdata object
Ap.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis", ontology = "BP",
                 allGenes = Ap.geneList, geneSel = selectFDR,
                 nodeSize = 10,
                 annot = annFUN.gene2GO, gene2GO = geneID2GO)

Ap.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.BP.resultParentChild <- runTest(Ap.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.BP.resultParentChild

# table results
Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 118)
#Ap.BP.ResTable
write.table(Ap.BP.ResTable, file = paste(resultsdir, "Ap_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.BP.ResTable)

# graph significant nodes

pdf(paste(resultsdir, "Ap.BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.BP.GOdata, score(Ap.BP.resultParentChild), firstSigNodes = 10, useInfo = 'all')
dev.off()
```

**2) High and low expressed transcripts**

Use `selectTranscript` function to select transcripts based on 'exp_type'. As only significant transcripts have an 'exp_type' assigned, this is a small subset of the above.

```{r exp_type_function}
selectTranscript <- function(score) {
    return(score == 1)
}

# To select genes, add 'exp_type' to RxN df
RxNout.t <- merge(RxNout, Ap.exp_type, by="Transcript", all=TRUE)
length(which(!is.na(RxNout.t$exp_type)))
```

```{r bim.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.bim.geneList <- rep(0, length = nrow(RxNout.t))
Ap.bim.geneList[which(RxNout.t$exp_type=="Bimodal")] <- 1
names(Ap.bim.geneList) <- RxNout.t$Transcript
str(Ap.bim.geneList)
table(Ap.bim.geneList)

# create topGOdata object
Ap.bim.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.bim.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.bim.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.bim.BP.resultParentChild <- runTest(Ap.bim.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.bim.BP.resultParentChild
```

Enriched gene sets for genes expressed at both high and low temperatures.

```{r, echo=FALSE, results="asis"}
Ap.bim.BP.ResTable <- GenTable(Ap.bim.BP.GOdata, parentchild = Ap.bim.BP.resultParentChild, topNodes = 52)
#Ap.bim.BP.ResTable
write.table(Ap.bim.BP.ResTable, file = paste(resultsdir, "Ap_bimodal_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.bim.BP.ResTable)
```

**3) Intermediate-temperature expressed transcripts**

```{r int.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.int.geneList <- rep(0, length = nrow(RxNout.t))
Ap.int.geneList[which(RxNout.t$exp_type=="Intermediate")] <- 1
names(Ap.int.geneList) <- RxNout.t$Transcript
str(Ap.int.geneList)
table(Ap.int.geneList)

# create topGOdata object
Ap.int.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.int.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.int.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.int.BP.resultParentChild <- runTest(Ap.int.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.int.BP.resultParentChild
```

Enriched gene sets for genes expressed at intermediate temperatures.

```{r, echo=FALSE, results="asis"}
Ap.int.BP.ResTable <- GenTable(Ap.int.BP.GOdata, parentchild = Ap.int.BP.resultParentChild, topNodes=16)
Ap.int.BP.ResTable
write.table(Ap.int.BP.ResTable, file = paste(resultsdir, "Ap_intermediate_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.int.BP.ResTable)
```

**4) High temperature expressed transcripts**

```{r hig.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.hig.geneList <- rep(0, length = nrow(RxNout.t))
Ap.hig.geneList[which(RxNout.t$exp_type=="High")] <- 1
names(Ap.hig.geneList) <- RxNout.t$Transcript
str(Ap.hig.geneList)
table(Ap.hig.geneList)

# create topGOdata object
Ap.hig.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.hig.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.hig.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.hig.BP.resultParentChild <- runTest(Ap.hig.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.hig.BP.resultParentChild
```

Enriched gene sets for genes expressed at both high temperatures only.

```{r, echo=FALSE, results="asis"}
Ap.hig.BP.ResTable <- GenTable(Ap.hig.BP.GOdata, parentchild = Ap.hig.BP.resultParentChild, topNodes=12)
#Ap.hig.BP.ResTable
write.table(Ap.hig.BP.ResTable, file = paste(resultsdir, "Ap_high_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.hig.BP.ResTable)
```

**5) Low temperature expressed transcripts**

```{r low.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.low.geneList <- rep(0, length = nrow(RxNout.t))
Ap.low.geneList[which(RxNout.t$exp_type=="Low")] <- 1
names(Ap.low.geneList) <- RxNout.t$Transcript
str(Ap.low.geneList)
table(Ap.low.geneList)

# create topGOdata object
Ap.low.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.low.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.low.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.low.BP.resultParentChild <- runTest(Ap.low.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.low.BP.resultParentChild
```

Enriched gene sets for genes expressed at low temperatures only.

```{r, echo=FALSE, results="asis"}
Ap.low.BP.ResTable <- GenTable(Ap.low.BP.GOdata, parentchild = Ap.low.BP.resultParentChild, topNodes=26)
Ap.low.BP.ResTable
write.table(Ap.low.BP.ResTable, file = paste(resultsdir, "Ap_low_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.low.BP.ResTable)
```

```{r, echo=FALSE}
# graph significant nodes
pdf(paste(resultsdir, "Ap_bimodal_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.bim.BP.GOdata, score(Ap.bim.BP.resultParentChild), firstSigNodes = 52, useInfo = 'all')
dev.off()

pdf(paste(resultsdir, "Ap_intermediate_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.int.BP.GOdata, score(Ap.int.BP.resultParentChild), firstSigNodes = 16, useInfo = 'all')
dev.off()

pdf(paste(resultsdir, "Ap_high_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.hig.BP.GOdata, score(Ap.hig.BP.resultParentChild), firstSigNodes = 12, useInfo = 'all')
dev.off()

pdf(paste(resultsdir, "Ap_low_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.low.BP.GOdata, score(Ap.low.BP.resultParentChild), firstSigNodes = 26, useInfo = 'all')
dev.off()
```

Visualize differens in GO terms among expression types using a wordcloud.

```{gsea_wordcloud, cache=TRUE}
# create Corpus for text mining
GO.dfs <- c(paste(Ap.low.BP.ResTable$Term, collapse=" "), paste(Ap.hig.BP.ResTable$Term, collapse=" "),
            paste(Ap.int.BP.ResTable$Term, collapse=" "), paste(Ap.bim.BP.ResTable$Term, collapse=" "))

vs <- VectorSource(GO.dfs)
GO.corp <- Corpus(vs)
GO.corp

GO.corp <- tm_map(GO.corp, removePunctuation)

# create matrix of terms
term.matrix <- TermDocumentMatrix(GO.corp)
term.matrix <- as.matrix(term.matrix)
colnames(term.matrix) <- c('low', 'high', 'intermediate', 'bimodal')

# comparison cloud
comparison.cloud(term.matrix, max.words=200, random.order=FALSE, scale=c(0.25, 0.5))
```

## Comparison of expression patterns between colonies

A set of `r nrow(interaction.transcripts)` transcripts have expression patterns that depend on colony. In this section, I examine how these transcripts differ between colonies. Specifically, I expect genes that are up-regulated at high temperatures in the northern *A22* colony to be at constitutively higher expression in the more southern *Ar* colony. Conversely, I will examine the extent to which genes upregulated at colder temperatures in the *Ar* colony are constitutively expressed in the *A22* colony.

## Visualize responsive transcripts

Make plots for all significant transcripts

```{r plot_responsive, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
p1 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p1

# Smooth plot, expression against temp, faceted by colony
p2 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p2

# Same as p1, faceted by expression type
p3 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(exp_type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p3

# Smooth plot, expression against temp, faceted by colony
p4 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(exp_type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p4
```

Make plots for transcripts with significant temperature by colony interaction

```{r plot_interaction_responsive, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
p1 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p1

# Smooth plot, expression against temp, faceted by colony
p2 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p2

# Same as p1, faceted by expression type
p3 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(exp_type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p3

# Smooth plot, expression against temp, faceted by colony
p4 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(exp_type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p4
```

```{r plot, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_by_colony_line.png", sep=""))
  p1 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p1
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_by_colony_smooth.png", sep=""))
  p2 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p2
dev.off()

# Same as p1, faceted by expression type
png(paste(resultsdir, "Ap_expression_by_colony_exp_line.png", sep=""))
  p3 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(exp_type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p3
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_by_colony_exp_smooth.png", sep=""))
  p4 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(exp_type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p4
dev.off()

```

```{r plot_interaction_responsive_to_file, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_interaction_by_colony_line.png", sep=""))
  p1 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p1
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_interaction_by_colony_smooth.png", sep=""))
  p2 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p2
dev.off()

# Same as p1, faceted by expression type
png(paste(resultsdir, "Ap_expression_interaction_by_colony_exp_line.png", sep=""))
  p3 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(exp_type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p3
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_interaction_by_colony_exp_smooth.png", sep=""))
  p4 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(exp_type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p4
dev.off()
```

## Session information ##

```{r session}
save.image()
sessionInfo()
```

## References

```{r references, results='asis'}
bibliography()
```

### 2014-03-27 ApTranscriptome_TR.Rmd
Changed  function to use robust regression function  from library(robustbase). Currently broken

@@
 # Load libraries
+library(robustbase)
 library(R.utils)
@@
 
-# identify responsive transcripts
-RxNout <- RxNseq(f = TPM.dt.sub, model = model, threshold = 0.05)
+# identify responsive transcripts using robust regression
+testdata <- TPM.dt.sub[23:44,]
+
+RxNout <- RxNseq(f = testdata, model = model, threshold = 0.05)
 

### 2014-03-27 scripts/RxNseq.R

@@ -49,38 +49,35 @@ RxNseq <- function(f, model = "NA", threshold = 0.05) {
 
     # requires
     require(stringr)
+    require(robustbase)
     require(plyr)
     
-    #########################################################################
-    # Function to report overall model P-value
-
-    lmp <- function(modelobject) {
-        if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
-        f <- summary(modelobject)$fstatistic
-        p <- unname(pf(f[1],f[2],f[3],lower.tail=F))
-        attributes(p) <- NULL
-        return(p)
-    }
-    ##########################################################################
-
-    dd <- ddply(f, .(Transcript), function(f) {
+    dd <- ddply(f, .(Transcript), .progress = "text", function(f) {
         # fit model
-        lmout <- eval(parse(text = paste("lm(", model, ", data = f)", sep = "")))
-        Fvals <- anova(lmout)$'Pr(>F)'
+        try(lmrob.out <- eval(parse(text = paste("lmrob(", model, ", setting = 'KS2011', data = f)", sep = ""))))
+        if(inherits(lmrob.out, "try-error"))
+            { # error due to inability fit model. skip this transcript
+                continue
+            }
+            # fit null model
+        try(lmrob.null <- lmrob(TPM ~ 1, data = f))
+        # test of overall model fit
+        lmrob.p <- anova(lmrob.null, lmrob.out)$"Pr(>chisq)"[2]
+
+        # get coefficients and their P(>|t|)
+        coef.vals <- coefficients(summary(lmrob.out))[,1]
+        coef.signif <- coefficients(summary(lmrob.out))[,4]
 
         # assign coefficients only for significant terms, else NA
         coefvec <- vector()
-        coefnames <- unlist(str_split(model, pattern = " "))
-        coefnames <- paste("coef", coefnames[c(TRUE,FALSE)][-1], sep = ".")
-        
-        for(i in 1:(length(Fvals)-1)) { # skip last value which is residuals
-            coefval <- ifelse(Fvals[i] < threshold, Fvals[i], NA)
+                
+        for(i in 1:(length(coef.vals))) {
+            coefval <- ifelse(coef.signif[i] < threshold, coef.vals, 0)
             coefvec <- append(coefvec, coefval)
-            names(coefvec)[i] <- coefnames[i]
         }
         
         # return values
-        return(c(pval = lmp(lmout),
+        return(c(modp = lmrob.p,
                  coefvec))
         } # end function
     ) # end ddply

### 2014-03-28 scripts/RxNseq.R
Changed RxNseq function to perform robust regression using lmrob. Had to use supressWarnings() so messages from lmrob did not prevent ddply from running across all transcripts.

@@ -45,7 +45,6 @@ RxNseq <- function(f, model = "NA", threshold = 0.05) {
     # Returns:
     #  Returns dataframe with p-value and regression coefficients
     #  for each transcript
-      
 
     # requires
     require(stringr)
@@ -53,29 +52,21 @@ RxNseq <- function(f, model = "NA", threshold = 0.05) {
     require(plyr)
     
     dd <- ddply(f, .(Transcript), .progress = "text", function(f) {
-        # fit model
-        try(lmrob.out <- eval(parse(text = paste("lmrob(", model, ", setting = 'KS2011', data = f)", sep = ""))))
-        if(inherits(lmrob.out, "try-error"))
-            { # error due to inability fit model. skip this transcript
-                continue
-            }
+        # robust regression - suppressWarnings so ddply runs
+        lmrob.out <- suppressWarnings(eval(parse(text = paste("lmrob(", model, ", setting = 'KS2011', data = f)", sep = ""))))
+        # check if model converged. if not, modp=NA
+        if(lmrob.out$converged == FALSE) {
+            coefvec <- coefficients(summary(lmrob.out))[,1]
+            lmrob.p <- NA
+        } else {
             # fit null model
-        try(lmrob.null <- lmrob(TPM ~ 1, data = f))
-        # test of overall model fit
-        lmrob.p <- anova(lmrob.null, lmrob.out)$"Pr(>chisq)"[2]
-
-        # get coefficients and their P(>|t|)
-        coef.vals <- coefficients(summary(lmrob.out))[,1]
-        coef.signif <- coefficients(summary(lmrob.out))[,4]
-
-        # assign coefficients only for significant terms, else NA
-        coefvec <- vector()
-                
-        for(i in 1:(length(coef.vals))) {
-            coefval <- ifelse(coef.signif[i] < threshold, coef.vals, 0)
-            coefvec <- append(coefvec, coefval)
-        }
-        
+            lmrob.null <- lmrob(TPM ~ 1, data = f)
+            # test of overall model fit
+            lmrob.p <- anova(lmrob.null, lmrob.out)$"Pr(>chisq)"[2]
+            # report only coefficients with `P(>|t|)` < threshold
+            coefvec <- ifelse(coefficients(summary(lmrob.out))[,4] < threshold, coefficients(summarylmrob.out)[,4], NA)
+        } # end else
+                                      
         # return values
         return(c(modp = lmrob.p,
                  coefvec))

### 2014-03-28 ApTranscriptome_TR.Rmd
Changed RxNseq function to perform robust regression using lmrob. Had to use supressWarnings() so messages from lmrob did not prevent ddply from running across all transcripts.
Updates - trying to get RxNseq to work with lmrob. Problems with using 'anova' on exact model fits.

@@
 # identify responsive transcripts using robust regression
-testdata <- TPM.dt.sub[23:44,]
-
-RxNout <- RxNseq(f = testdata, model = model, threshold = 0.05)
+RxNout <- RxNseq(f = TPM.dt.sub, model = model, threshold = 0.05)
+
+RxNout <- RxNseq(f = test, model = model, threshold = 0.05)
+
+## debug
+        lmrob.out345 <- suppressWarnings(eval(parse(text = paste("lmrob(", model, ", setting = 'KS2011', data = piece345)", sep = ""))))
 

### 2014-03-28 scripts/RxNseq.R

@@ -51,7 +51,7 @@ RxNseq <- function(f, model = "NA", threshold = 0.05) {
     require(robustbase)
     require(plyr)
     
-    dd <- ddply(f, .(Transcript), .progress = "text", function(f) {
+    dd <- ddply(f, .(Transcript), .progress = "text", .inform = TRUE, function(f) {
         # robust regression - suppressWarnings so ddply runs
         lmrob.out <- suppressWarnings(eval(parse(text = paste("lmrob(", model, ", setting = 'KS2011', data = f)", sep = ""))))
         # check if model converged. if not, modp=NA
@@ -64,7 +64,7 @@ RxNseq <- function(f, model = "NA", threshold = 0.05) {
             # test of overall model fit
             lmrob.p <- anova(lmrob.null, lmrob.out)$"Pr(>chisq)"[2]
             # report only coefficients with `P(>|t|)` < threshold
-            coefvec <- ifelse(coefficients(summary(lmrob.out))[,4] < threshold, coefficients(summarylmrob.out)[,4], NA)
+            coefvec <- ifelse(coefficients(summary(lmrob.out))[,4] < threshold, coefficients(summary(lmrob.out))[,4], NA)
         } # end else
                                       
         # return values

### 2014-03-31 ApTranscriptome_TR.Rmd
Changed from robust regression, to simply using log-transformed response to control for outliers
* Reworking analysis script to save results of `lm` for expression of each transcript in a list.
* Fit model using log to lessen effects of outliers with high expression on results. This had little effect, actually slightly increasing the number of significant transcripts.

--- 2014-03-27 ApTranscriptome_TR.Rmd
@@
 
-**March 20, 2014**
+**March 31, 2014**
 
@@
 library(plyr)
+library(MASS)
 
@@
 unique(TPM.dt.sub$val)
-
+```
+
+(1) Identify transcripts with overall significant model fit.
+Adjust *P* values for multiple testing using FDR and retain transcripts with FDR < 0.05.
+Use log-transformed response to account for outliers.
+
+```{r modp}
 # define model for RxN function
-model <- "TPM ~ colony + val + I(val^2) + colony:val + colony:I(val^2)"
-
-# identify responsive transcripts
-RxNout <- RxNseq(f = TPM.dt.sub, model = model, threshold = 0.05)
-
-save.image("RxN_combined_results.RData")
-```
-
-Of the `r nrow(RxNout)` transcripts, `r length(which(RxNout$pval < 0.05))` have models with P < 0.05.
-
-Many of these are likely false positives, so I adjusted P-values using false discovery rate (FDR) to identify only those transcripts with less than 5% FDR as significant.
+model <-  "log(TPM+1) ~ colony + val + I(val^2) + colony:val + colony:I(val^2)"
+
+# define function for `ddply` to calculate overall model P value
+modpFunc <- function(foo) {
+  # fit lm
+  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))
+
+  # calculate overall model significance
+  f <- summary(lmout)$fstatistic
+  pval <- unname(pf(f[1],f[2],f[3],lower.tail=F))
+  attributes(pval) <- NULL
+
+  # return pvalue
+  c(pval = pval) 
+}
+
+RxNpval <- ddply(TPM.dt.sub, .(Transcript), .progress="text", .inform="TRUE", modpFunc)
+```
+
+Of the `r nrow(RxNpval)` transcripts, `r length(which(RxNpval$pval < 0.05))` have models with P < 0.05.
+
+Many of these are likely false positives, so I adjust P-values using false discovery rate (FDR). Only those transcripts with less than 5% FDR are retained as significant.
 
 ```{r fdr}
-RxNout$padj <- p.adjust(RxNout$pval, method = "fdr")
+RxNpval$padj <- p.adjust(RxNpval$pval, method = "fdr")
 # Plot FDR values against initial pvalues
 par(mfrow = c(2,1))
-hist(RxNout$pval)
-hist(RxNout$padj)
+hist(RxNpval$pval)
+hist(RxNpval$padj)
 
 # subset to significant transcripts
-signif.transcripts <- RxNout[which(RxNout$padj < 0.05), ]
-```
-
-## Functional annotation ##
-
-In the previous section, I identified transcripts that show significant responses in expression. Next, I add gene annotation and ontology information to these transcripts.
-
-```{r annotation}
-# add annotation information
-setkey(annotationtable, Sequence.Name)
-signif.transcripts <- data.table(signif.transcripts)
-setkey(signif.transcripts, Transcript)
-signif.transcripts <- annotationtable[signif.transcripts]
-setnames(signif.transcripts, "Sequence.Name", "Transcript")
+signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]
+
+# extract significant transcripts
+sig.TPM.dt.sub <- TPM.dt.sub[signif.transcripts$Transcript]
+```
+
+At the 5% FDR significance threshold, there are `r nrow(signif.transcripts)` transcripts with an overall significant model.
+
+(2) Fit linear model to overall significant transcripts; perform stepAIC to retain only significant terms, and save `lm` output to list
+
+```{r lmFunc}
+# first, define function to use
+lmFunc <- function(bar) {
+  # fit lm
+  lmout <- eval(parse(text = paste("lm(", model, ", data = bar)", sep = "")))
+
+  # stepAIC to drop non-significant terms
+  f.lmout <- stepAIC(lmout)
+
+  # return final model
+  return(f.lmout) 
+}
+
+# apply function to all responsive transcripts
+RxNlmAIC <- dlply(sig.TPM.dt.sub, .(Transcript), .progress="text", lmFunc)
 ```
@@
 # transcripts that differ in expression by colony only
-colony.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.colony') & is.na(signif.transcripts$'coef.val') & is.na(signif.transcripts$'coef.I(val^2)') & is.na(signif.transcripts$'coef.colony:val') & is.na(signif.transcripts$'coef.colony:I(val^2)')]
-# check that subset correct - no transcripts with significant temperature terms
-length(which(!is.na(colony.transcripts$'coef.val') | !is.na(colony.transcripts$'coef.I(val^2)') | !is.na(colony.transcripts$'coef.colony:val') | !is.na(colony.transcripts$'coef.I(val^2)')))
-str(colony.transcripts)
-# order by adjusted P-value
-colony.transcripts <- colony.transcripts[order(colony.transcripts$padj), ]
-
-# remaining transcripts are thermally responsive
-responsive.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.val') | !is.na(signif.transcripts$'coef.I(val^2)') | !is.na(signif.transcripts$'coef.colony:val') | !is.na(signif.transcripts$'coef.colony:I(val^2)')]
-dim(responsive.transcripts)
-
-# check that these two groups sum to total significant transcripts
-stopifnot(all.equal(nrow(signif.transcripts), nrow(colony.transcripts) + nrow(responsive.transcripts)))
+keepthis <- function(lmitem, term = NA) {
+  coefs <- coefficients(lmitem)
+  coef.names <- names(coefs)
+  keep <- ifelse(term %in% coef.names, TRUE, FALSE)
+}
+
+# identify sets of transcripts that include each term in final model
+colony.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr") == TRUE)]
+temp.lin.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "val") == TRUE)]
+temp.quad.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "I(val^2)") == TRUE)]
+col.by.temp.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr:val") == TRUE)]
+col.by.temp.quad.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr:I(val^2)") == TRUE)]
+
+# union of thermally-responsive transcripts
+responsive.transcripts <- union(temp.lin.transcripts, temp.quad.transcripts)
+responsive.transcripts <- union(responsive.transcripts, col.by.temp.transcripts)
+responsive.transcripts <- union(responsive.transcripts, col.by.temp.quad.transcripts)
+length(responsive.transcripts)
+
+# transcripts that have a colony effect only 
+colony.transcripts <- setdiff(RxNlmAIC, responsive.transcripts)
+
+# transcripts that have colony x temperature interaction
+interaction.transcripts <- union(col.by.temp.transcripts, col.by.temp.quad.transcripts)
 
@@
 # main effect of temperature only
-temperature.transcripts <- responsive.transcripts[is.na(responsive.transcripts$'coef.colony:val') & is.na(responsive.transcripts$'coef.colony:I(val^2)')]
-# order by adjusted P-value
-temperature.transcripts <- temperature.transcripts[order(temperature.transcripts$padj), ]
-str(temperature.transcripts)
-
-# transcripts that have colony x temperature interaction
-interaction.transcripts <- signif.transcripts[!is.na(signif.transcripts$'coef.colony:val') | !is.na(signif.transcripts$'coef.colony:I(val^2)')]
-# order by adjusted P-value
-interaction.transcripts <- interaction.transcripts[order(interaction.transcripts$padj), ]
-str(interaction.transcripts)
-
-# check that these two groups sum to total responsive transcripts
-stopifnot(all.equal(nrow(responsive.transcripts), nrow(temperature.transcripts) + nrow(interaction.transcripts)))
+temperature.transcripts <- setdiff(responsive.transcripts, interaction.transcripts)
 ```
@@
 clist <- c('Total', 'Colony', 'Temperature', 'Temperature:Colony')
-siglist <- c(nrow(signif.transcripts),
-             nrow(colony.transcripts),
-             nrow(temperature.transcripts),
-             nrow(interaction.transcripts)
+siglist <- c(length(RxNlmAIC),
+             length(colony.transcripts),
+             length(temperature.transcripts),
+             length(interaction.transcripts)
              )
@@
 
+## Functional annotation ##
+
+In the previous section, I identified transcripts that show significant responses in expression. Next, I add gene annotation and ontology information to these transcripts.
+
+```{r annotation}
+# add annotation information
+setkey(annotationtable, Sequence.Name)
+signif.transcripts <- data.table(signif.transcripts)
+setkey(signif.transcripts, Transcript)
+signif.transcripts <- annotationtable[signif.transcripts]
+setnames(signif.transcripts, "Sequence.Name", "Transcript")
+```
+
 ## Shiny interactive web-app

### 2014-03-31 scripts/RxNseq.R

@@ -45,30 +45,37 @@ RxNseq <- function(f, model = "NA", threshold = 0.05) {
     # Returns:
     #  Returns dataframe with p-value and regression coefficients
     #  for each transcript
+      
 
     # requires
     require(stringr)
-    require(robustbase)
     require(plyr)
     
-    dd <- ddply(f, .(Transcript), .progress = "text", .inform = TRUE, function(f) {
-        # robust regression - suppressWarnings so ddply runs
-        lmrob.out <- suppressWarnings(eval(parse(text = paste("lmrob(", model, ", setting = 'KS2011', data = f)", sep = ""))))
-        # check if model converged. if not, modp=NA
-        if(lmrob.out$converged == FALSE) {
-            coefvec <- coefficients(summary(lmrob.out))[,1]
-            lmrob.p <- NA
-        } else {
-            # fit null model
-            lmrob.null <- lmrob(TPM ~ 1, data = f)
-            # test of overall model fit
-            lmrob.p <- anova(lmrob.null, lmrob.out)$"Pr(>chisq)"[2]
-            # report only coefficients with `P(>|t|)` < threshold
-            coefvec <- ifelse(coefficients(summary(lmrob.out))[,4] < threshold, coefficients(summary(lmrob.out))[,4], NA)
-        } # end else
-                                      
+    #########################################################################
+    # Function to report overall model P-value
+
+    lmp <- function(modelobject) {
+        if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
+        f <- summary(modelobject)$fstatistic
+        p <- unname(pf(f[1],f[2],f[3],lower.tail=F))
+        attributes(p) <- NULL
+        return(p)
+    }
+    ##########################################################################
+
+    dd <- ddply(f, .(Transcript), .progress="text", function(f) {
+        # fit model
+        lmout <- eval(parse(text = paste("lm(", model, ", data = f)", sep = "")))
+        Fvals <- anova(lmout)$'Pr(>F)'
+        
+        # stepAIC to drop non-significant terms
+        f.lmout <- stepAIC(lmout)
+        
+        # report significant coefficients
+        coefvec <- coefficients(f.lmout)
+        
         # return values
-        return(c(modp = lmrob.p,
+        return(c(pval = lmp(lmout),
                  coefvec))
         } # end function
     ) # end ddply

### 2014-04-01 ApTranscriptome_TR.Rmd
Continuing work to rewrite analysis working on `lm` results saved to list.

@@
 ```{r responsive_transcripts}
-# transcripts that differ in expression by colony only
-keepthis <- function(lmitem, term = NA) {
+grepFunc <- function(lmitem, term = NA) {
   coefs <- coefficients(lmitem)
   coef.names <- names(coefs)
-  keep <- ifelse(term %in% coef.names, TRUE, FALSE)
+  keep <- if(length(grep(term, coef.names)) > 0) TRUE else FALSE
 }
 
-# identify sets of transcripts that include each term in final model
-colony.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr") == TRUE)]
-temp.lin.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "val") == TRUE)]
-temp.quad.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "I(val^2)") == TRUE)]
-col.by.temp.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr:val") == TRUE)]
-col.by.temp.quad.transcripts <- RxNlmAIC[which(Map(keepthis, RxNlmAIC, term = "colonyAr:I(val^2)") == TRUE)]
-
-# union of thermally-responsive transcripts
-responsive.transcripts <- union(temp.lin.transcripts, temp.quad.transcripts)
-responsive.transcripts <- union(responsive.transcripts, col.by.temp.transcripts)
-responsive.transcripts <- union(responsive.transcripts, col.by.temp.quad.transcripts)
-length(responsive.transcripts)
-
-# transcripts that have a colony effect only 
-colony.transcripts <- setdiff(RxNlmAIC, responsive.transcripts)
-
-# transcripts that have colony x temperature interaction
-interaction.transcripts <- union(col.by.temp.transcripts, col.by.temp.quad.transcripts)
-
-# split responsive_transcripts into those that have same temperature response in both colonies and 
-# main effect of temperature only
-temperature.transcripts <- setdiff(responsive.transcripts, interaction.transcripts)
+interaction.transcripts <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
+other.transcripts <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.transcripts))]
+temperature.transcripts <- other.transcripts[which(Map(grepFunc, other.transcripts, term = "val") == TRUE)]
+colony.transcripts <- other.transcripts[setdiff(names(other.transcripts), names(temperature.transcripts))]
+rm(other.transcripts)
 ```
@@
 
-Table 3 shows that `r round((nrow(temperature.transcripts) + nrow(interaction.transcripts))/nrow(signif.transcripts), 2)*100`% of the significant transcripts are thermally-responsive, with `r round(nrow(interaction.transcripts)/(nrow(temperature.transcripts) + nrow(interaction.transcripts)), 2)*100`% of these transcripts having a thermal-response that differs by colony.
-
-I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.
-
-```{r GO_contingency_test}
-# GO 'response to stress' hits in responsive transcripts
-responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)]
-
-resp.stress.responsive.count <- nrow(responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
-# GO 'response to stress' hits in all transcripts
-resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
-
-GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
-
-GO.stress.Xsq <- chisq.test(GO.stress.table)
-GO.stress.Xsq
-```
-
-There are also 4 heat shock related genes in the responsive transcripts, out of 130 total.
-
-```{r grep_shock}
-hsp_responsive <- responsive.transcripts[grep("shock", responsive.transcripts$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
-nrow(hsp_responsive)
-
-hsp_all <- annotationtable[grep("shock", annotationtable$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
-nrow(hsp_all)
-```
-
 ### Thermal-response functional types ###
 
-The previous section simply identified the transcripts with thermally-responsive expression.
+The previous section identified the transcripts with thermally-responsive expression.
 In this section, I determine the shape of the expression response to temperature for each transcript.
@@
 ```{r expression_shape, eval=TRUE, echo=TRUE}
+
+test.transcripts <- temperature.transcripts[1:10]
+
+response.type <- ldply(test.transcripts, .progress = "text", RxNtype)
+
+RxNtype <- function(lmitem) {
+  # predict 
+  pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
+  newdf <- data.frame(val = pred.vals, colony = as.factor(rep(unlist(lmitem$xlevels), each = length(pred.vals))))
+  pout <- predict(lmout, newdata=newdf)
+  pout <- cbind(newdf, pTPM = pout)
+
+  max.val
+  # for each colony
+  for(c in levels(pout$colony)) {
+
+    # get max val
+    max.val <- pout[which(pout$pTPM == max(pout$pTPM)), "val"]
+    # get min val
+    min.val <- pout[which(pout$pTPM == min(pout$pTPM)), "val"]
+    # determine expression shape
+    exp_type <- NA
+    # expression type defined as "Bimodal" if max expression below 10C and above 30C is greater than 2 standard deviations above expression at the optimum temperature (19.5C)
+    if(max(pout[pout$val < 10, "pTPM"]) > (pout[pout$val == 19.5, "pTPM"] + 2*sd(pout$pTPM)) & max(pout[pout$val > 10, "pTPM"]) > (pout[pout$val == 19.5, "pTPM"] + 2*sd(pout$pTPM)))   exp_type <- "Bimodal" else {
+      # expression type "High" if max expression at temperature above 30C
+      if(max.val > 30) exp_type <- "High" else {
+        # expression type "Low" if max expression at temperature below 10C
+        if(max.val < 10) exp_type <- "Low" else
+          exp_type <- "Intermediate"
+        } # end else
+      } # end else
+    }
+  }
+
+# plot
+#qplot(val, pTPM, facets = .~colony, data = pout, geom="smooth")
+
 # merge RxN results with expression values
-temperature.transcripts.TPM <- TPM.dt.sub[temperature.transcripts]
+temperature.transcripts.TPM <- TPM.dt.sub[names(temperature.transcripts)]
 
@@
 
+I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.
+
+```{r GO_contingency_test}
+# GO 'response to stress' hits in responsive transcripts
+responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)]
+
+resp.stress.responsive.count <- nrow(responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
+# GO 'response to stress' hits in all transcripts
+resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
+
+GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
+
+GO.stress.Xsq <- chisq.test(GO.stress.table)
+GO.stress.Xsq
+```
+
+There are also 4 heat shock related genes in the responsive transcripts, out of 130 total.
+
+```{r grep_shock}
+hsp_responsive <- responsive.transcripts[grep("shock", responsive.transcripts$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
+nrow(hsp_responsive)
+
+hsp_all <- annotationtable[grep("shock", annotationtable$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
+nrow(hsp_all)
+```
+
 ## Functional annotation ##

### 2014-04-01 scripts/RxNseq.R

@@ -25,64 +25,6 @@ read.sailfish.quant <- function(filein, outname, samp, trtval) {
 }
 
 
-############################################################################################
-## RxNseq
-############################################################################################
-
-RxNseq <- function(f, model = "NA", threshold = 0.05) {
-    # Identify transcripts with significant reaction norms against a continuous variable
-    #
-    # Args:
-    #  f: data.frame in long format. must contain columns specified in model formula.
-    #     column with name "Transcript" must exist for proper split-apply
-    #  model: model formula
-    #     NOTE - for function to work correctly, each model term must be separated
-    #     by a single space. for example:
-    #            TPM ~ colony + val + colony:val
-    #     it is important that interactions are grouped together, no spaces
-    #  threshold: P-value threshold to determine if a coefficient is significant
-    #  
-    # Returns:
-    #  Returns dataframe with p-value and regression coefficients
-    #  for each transcript
-      
-
-    # requires
-    require(stringr)
-    require(plyr)
-    
-    #########################################################################
-    # Function to report overall model P-value
-
-    lmp <- function(modelobject) {
-        if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
-        f <- summary(modelobject)$fstatistic
-        p <- unname(pf(f[1],f[2],f[3],lower.tail=F))
-        attributes(p) <- NULL
-        return(p)
-    }
-    ##########################################################################
-
-    dd <- ddply(f, .(Transcript), .progress="text", function(f) {
-        # fit model
-        lmout <- eval(parse(text = paste("lm(", model, ", data = f)", sep = "")))
-        Fvals <- anova(lmout)$'Pr(>F)'
-        
-        # stepAIC to drop non-significant terms
-        f.lmout <- stepAIC(lmout)
-        
-        # report significant coefficients
-        coefvec <- coefficients(f.lmout)
-        
-        # return values
-        return(c(pval = lmp(lmout),
-                 coefvec))
-        } # end function
-    ) # end ddply
-} # end function
-
-
-
 
 ############################################################################################
 ## RxNply

### 2014-04-01 ApTranscriptome_TR.Rmd
Continuing work. Moved `RxNtype` function (used in `ldply`) to scripts.
Error message about "split_labels" to debug

@@
   lmout <- eval(parse(text = paste("lm(", model, ", data = bar)", sep = "")))  
-
   # stepAIC to drop non-significant terms
   f.lmout <- stepAIC(lmout)
-
   # return final model
@@
 
-test.transcripts <- temperature.transcripts[1:10]
-
-response.type <- ldply(test.transcripts, .progress = "text", RxNtype)
-
-RxNtype <- function(lmitem) {
-  # predict 
-  pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
-  newdf <- data.frame(val = pred.vals, colony = as.factor(rep(unlist(lmitem$xlevels), each = length(pred.vals))))
-  pout <- predict(lmout, newdata=newdf)
-  pout <- cbind(newdf, pTPM = pout)
-
-  max.val
-  # for each colony
-  for(c in levels(pout$colony)) {
-
-    # get max val
-    max.val <- pout[which(pout$pTPM == max(pout$pTPM)), "val"]
-    # get min val
-    min.val <- pout[which(pout$pTPM == min(pout$pTPM)), "val"]
-    # determine expression shape
-    exp_type <- NA
-    # expression type defined as "Bimodal" if max expression below 10C and above 30C is greater than 2 standard deviations above expression at the optimum temperature (19.5C)
-    if(max(pout[pout$val < 10, "pTPM"]) > (pout[pout$val == 19.5, "pTPM"] + 2*sd(pout$pTPM)) & max(pout[pout$val > 10, "pTPM"]) > (pout[pout$val == 19.5, "pTPM"] + 2*sd(pout$pTPM)))   exp_type <- "Bimodal" else {
-      # expression type "High" if max expression at temperature above 30C
-      if(max.val > 30) exp_type <- "High" else {
-        # expression type "Low" if max expression at temperature below 10C
-        if(max.val < 10) exp_type <- "Low" else
-          exp_type <- "Intermediate"
-        } # end else
-      } # end else
-    }
-  }
+response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
 
 # plot
-#qplot(val, pTPM, facets = .~colony, data = pout, geom="smooth")
+#qplot(val, pTPM, facets = .~colony, data = A22.pout, geom="smooth")
 
@@
 temperature.transcripts.type$opt.exp <- as.numeric(temperature.transcripts.type$opt.exp)
-temperature.transcripts.type$exp_type <- as.factor(temperature.transcripts.type$exp_type)
+temperature.transcripts.type$A22.exp.type <- as.factor(temperature.transcripts.type$A22.exp.type)
 
@@
 A22.interaction.transcripts.type$opt.exp <- as.numeric(A22.interaction.transcripts.type$opt.exp)
-A22.interaction.transcripts.type$exp_type <- as.factor(A22.interaction.transcripts.type$exp_type)
+A22.interaction.transcripts.type$A22.exp.type <- as.factor(A22.interaction.transcripts.type$A22.exp.type)
 str(A22.interaction.transcripts.type)
@@
 Ar.interaction.transcripts.type$opt.exp <- as.numeric(Ar.interaction.transcripts.type$opt.exp)
-Ar.interaction.transcripts.type$exp_type <- as.factor(Ar.interaction.transcripts.type$exp_type)
+Ar.interaction.transcripts.type$A22.exp.type <- as.factor(Ar.interaction.transcripts.type$A22.exp.type)
 str(Ar.interaction.transcripts.type)
@@
 
-```{r expression_exp_type_table, eval=TRUE, echo=FALSE, results='asis'}
+```{r expression_A22.exp.type_table, eval=TRUE, echo=FALSE, results='asis'}
 # table type of response
-A22.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "A22", 'exp_type'])
-Ar.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "Ar", 'exp_type'])
-
-exp_type.table <- rbind(A22.type.table, Ar.type.table)
-rownames(exp_type.table) <- c("A22", "Ar")
-
-pandoc.table(exp_type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
+A22.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "A22", 'A22.exp.type'])
+Ar.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "Ar", 'A22.exp.type'])
+
+A22.exp.type.table <- rbind(A22.type.table, Ar.type.table)
+rownames(A22.exp.type.table) <- c("A22", "Ar")
+
+pandoc.table(A22.exp.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
 ```
@@
 # list of transcripts that are 'high' expressed in A22
-A22_high_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$exp_type == "High"), "Transcript"]
+A22_high_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$A22.exp.type == "High"), "Transcript"]
 # dataframe of transcripts from both colonies that are 'high' expressed in A22
@@
 # list of transcripts that are 'Low' expressed in Ar
-Ar_low_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$exp_type == "Low"), "Transcript"]
+Ar_low_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$A22.exp.type == "Low"), "Transcript"]
 # dataframe of transcripts from both colonies that are 'Low' expressed in Ar
@@
 # list of transcripts that are 'Intermediate' expressed in A22
-A22_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$exp_type == "Intermediate"), "Transcript"]
+A22_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
 A22_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% A22_int_transcripts), ]
@@
 # list of transcripts that are 'Intermediate' expressed in Ar
-Ar_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$exp_type == "Intermediate"), "Transcript"]
+Ar_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
 Ar_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% Ar_int_transcripts), ]
@@
 setkey(TPM.dt.sub, Transcript)
-A22trlist <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$exp_type == "Intermediate"), "Transcript"]
+A22trlist <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
 A22.TPM.int.dt <- TPM.dt.sub[A22trlist]
@@
 # repeat for Ar
-Artrlist <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$exp_type == "Intermediate"), "Transcript"]
+Artrlist <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
 Ar.TPM.int.dt <- TPM.dt.sub[Artrlist]
@@
 # extract 'Bimodal' expressed transcripts for A22 colony
-A22trlist.bim <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$exp_type == "Bimodal"), "Transcript"]
+A22trlist.bim <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$A22.exp.type == "Bimodal"), "Transcript"]
 A22.TPM.bim.dt <- TPM.dt.sub[A22trlist.bim]
@@
 # repeat for Ar
-Artrlist.bim <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$exp_type == "Bimodal"), "Transcript"]
+Artrlist.bim <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$A22.exp.type == "Bimodal"), "Transcript"]
 Ar.TPM.bim.dt <- TPM.dt.sub[Artrlist.bim]
@@
 g5 <- ggplot(interaction.transcripts.type, aes(x=max.val, group=colony)) + geom_density(alpha=0.2, position="identity")
-g5 + facet_grid(. ~ exp_type) + scale_y_continuous(name="Density") +
+g5 + facet_grid(. ~ A22.exp.type) + scale_y_continuous(name="Density") +
   scale_x_continuous(name="Temperature of maximum expression")
@@
 
-Use `selectTranscript` function to select transcripts based on 'exp_type'. As only significant transcripts have an 'exp_type' assigned, this is a small subset of the above.
-
-```{r exp_type_function}
+Use `selectTranscript` function to select transcripts based on 'A22.exp.type'. As only significant transcripts have an 'A22.exp.type' assigned, this is a small subset of the above.
+
+```{r A22.exp.type_function}
 selectTranscript <- function(score) {
@@
 
-# To select genes, add 'exp_type' to RxN df
-RxNout.t <- merge(RxNout, Ap.exp_type, by="Transcript", all=TRUE)
-length(which(!is.na(RxNout.t$exp_type)))
+# To select genes, add 'A22.exp.type' to RxN df
+RxNout.t <- merge(RxNout, Ap.A22.exp.type, by="Transcript", all=TRUE)
+length(which(!is.na(RxNout.t$A22.exp.type)))
 ```
@@
 Ap.bim.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.bim.geneList[which(RxNout.t$exp_type=="Bimodal")] <- 1
+Ap.bim.geneList[which(RxNout.t$A22.exp.type=="Bimodal")] <- 1
 names(Ap.bim.geneList) <- RxNout.t$Transcript
@@
 Ap.int.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.int.geneList[which(RxNout.t$exp_type=="Intermediate")] <- 1
+Ap.int.geneList[which(RxNout.t$A22.exp.type=="Intermediate")] <- 1
 names(Ap.int.geneList) <- RxNout.t$Transcript
@@
 Ap.hig.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.hig.geneList[which(RxNout.t$exp_type=="High")] <- 1
+Ap.hig.geneList[which(RxNout.t$A22.exp.type=="High")] <- 1
 names(Ap.hig.geneList) <- RxNout.t$Transcript
@@
 Ap.low.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.low.geneList[which(RxNout.t$exp_type=="Low")] <- 1
+Ap.low.geneList[which(RxNout.t$A22.exp.type=="Low")] <- 1
 names(Ap.low.geneList) <- RxNout.t$Transcript
@@
   geom_line() +
-  facet_grid(exp_type ~ colony) +
+  facet_grid(A22.exp.type ~ colony) +
   scale_y_continuous(name="Expression (scaled)") +
@@
   geom_smooth() +
-  facet_grid(exp_type ~ colony) +
+  facet_grid(A22.exp.type ~ colony) +
   scale_y_continuous(name="Expression (scaled)") +
@@
     geom_line() +
-    facet_grid(exp_type ~ colony) +
+    facet_grid(A22.exp.type ~ colony) +
     scale_y_continuous(name="Expression (scaled)") +
@@
     geom_smooth() +
-    facet_grid(exp_type ~ colony) +
+    facet_grid(A22.exp.type ~ colony) +
     scale_y_continuous(name="Expression (scaled)") +

### 2014-04-01 scripts/RxNseq.R

@@ -25,150 +25,76 @@ read.sailfish.quant <- function(filein, outname, samp, trtval) {
 }
 
 
-
 ############################################################################################
-## RxNply
+## RxNtype
 ############################################################################################
 
-RxNply <- function(df1) {
-  # Function supplied to `ddply` to report the values at which minimum and maximum expression
-  # occur, the expression level at the optimum 'temperature' and the overall shape of the 
-  # expression function
-  #
-  # Args:
-  #  df1: dataframe in long format containing expression information
-  #
-  # Returns:
-  #  dataframe with maximum value of expression (max.val), minimum value of expression (min.val),
-  #  expression at optimum (opt.exp) and the shape of the expression function (exp_type)
-  
-  lmout <- lm(TPM ~ val + I(val^2), data = df1)
-  # NOTE - added a point to predict at 19.25C as this is the mean of the end points (0, 38.5)
-  vals <- c(0, 3.5, 10, 14, 17.5, 19.25, 21, 24.5, 28, 31.5, 35, 38.5)
-  newdf <- data.frame(val = vals)
-  pout <- predict(lmout, newdata=newdf)
-  pout <- data.frame(val = vals, exp = pout)
-  
-  # if all data is zero, set values to NA 
-  if(coef(lmout)['(Intercept)'] == 0 & coef(lmout)['val'] == 0 & coef(lmout)['I(val^2)'] == 0) {
-    max.val = NA
-    min.val = NA
-    opt.exp = NA
-    exp_type = "NotExp"
-  } else { # else set values based on predicted expression levels
+RxNtype <- function(lmitem) {
+    # Function passed to `ldply` that defines transcripts into response type
+    #
+    # Arguments:
+    #  lmitem: object of class `lm`
+    #
+    # Returns
+    #   dataframe with values of max expression, min expression and expression type for each colony
     
-    # get vals of max and min expression, and expression at 
-    max.val <- vals[which(pout$exp == max(pout$exp))]
-    min.val <- vals[which(pout$exp == min(pout$exp))]
-    
-    # report coefficients
-    #coef(lmout)
-    exp_type = if(coef(lmout)['val'] > 0 & coef(lmout)['I(val^2)'] > 0) "High" else {
[…]

### 2014-04-02 ApTranscriptome_TR.Rmd
Functioning(!) analysis that works on `lm` output from each transcript to report max, min temperatures of expression, and expression shape.

@@
 
-The previous section identified the transcripts with thermally-responsive expression.
-In this section, I determine the shape of the expression response to temperature for each transcript.
-Categories of expression response are:
+The previous section identified the transcripts with thermally-responsive expression. In this section, I determine the shape of the expression response to temperature for each transcript. Catego
+ries of expression response are:
 
@@
 
-response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
+interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
+
+test.transcripts <- interaction.transcripts[1000:1009]
+response.type <- ldply(test.transcripts, .progress = "text", .inform = TRUE, RxNtype)
 
 # plot
-#qplot(val, pTPM, facets = .~colony, data = A22.pout, geom="smooth")
+qplot(val, pTPM, facets = .~colony, data = Ar.pout, geom="smooth")
+qplot(val, pTPM, facets = .~colony, data = pout, geom="smooth")
+
+qplot(val, pTPM, facets = .~colony, data = Ar.pout, geom="smooth")
 

### 2014-04-02 scripts/RxNseq.R

@@ -44,55 +44,71 @@ RxNtype <- function(lmitem) {
   pout <- predict(lmitem, newdata=newdf)
   pout <- cbind(newdf, pTPM = pout)
 
-  # Calculate for A22
+  # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
+  pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
+  
+  # calculate for A22
   A22.pout <- pout[pout$colony == "A22", ]
-  # get max val
-  A22.max.val <- A22.pout[which(A22.pout$pTPM == max(A22.pout$pTPM)), "val"]
-  # get min val
-  A22.min.val <- A22.pout[which(A22.pout$pTPM == min(A22.pout$pTPM)), "val"]
-  # expression at optimum temp
-  A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
-  # determine expression shape
-  A22.exp.type <- NA
-  # expression type defined as "Bimodal" if max expression below 10C and above 30C is greater than 2 standard deviations above expression at the optimum temperature (19.5C)
-  if(max(A22.pout[A22.pout$val < 10, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + 2*sd(A22.pout$pTPM)) & max(A22.pout[A22.pout$val > 30, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + 2*sd(A22.pout$pTPM))) A22.exp.type <- "Bimodal" else {
-    # expression type "High" if max expression at temperature above 30C
-    if(A22.max.val > 30) A22.exp.type <- "High" else {
-      # expression type "Low" if max expression at temperature below 10C
-      if(A22.max.val < 10) A22.exp.type <- "Low" else
-        A22.exp.type <- "Intermediate"
-      } # end else
-    } # end else
+  # if pTPM doesn't vary, set max and min values to NA
+  if(sd(A22.pout$pTPM) < 0.01) {
+      A22.max.val <- NA
+      A22.min.val <- NA
+      A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
+      A22.exp.type <- "NotExp"
+  } else {
+          # else, set values from data
+          A22.max.val <- median(A22.pout[which(A22.pout$pTPM == max(A22.pout$pTPM)), "val"])
+          # for min.val, take median point as it may be string of points at zero
[…]

### 2014-04-02 ApTranscriptome_TR.Rmd
Analysis updated through 'thermal breadth'

--- 2014-04-01 ApTranscriptome_TR.Rmd
@@
 ```{r expression_shape, eval=TRUE, echo=TRUE}
-
-response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
-
-# plot
-#qplot(val, pTPM, facets = .~colony, data = A22.pout, geom="smooth")
-
-# merge RxN results with expression values
-temperature.transcripts.TPM <- TPM.dt.sub[names(temperature.transcripts)]
-
-# apply the 'RxNply' function to determine maximum and minimum temperatures of expression, as well as the expression level at the optimum temperature of 19.25C
-temperature.transcripts.type <- ddply(temperature.transcripts.TPM, .(Transcript), RxNply)
-temperature.transcripts.type$max.val <- as.numeric(temperature.transcripts.type$max.val)
-temperature.transcripts.type$min.val <- as.numeric(temperature.transcripts.type$min.val)
-temperature.transcripts.type$opt.exp <- as.numeric(temperature.transcripts.type$opt.exp)
-temperature.transcripts.type$A22.exp.type <- as.factor(temperature.transcripts.type$A22.exp.type)
-
-# repeat for transcripts that have an interaction with colongy
-# A22
-A22.interaction.transcripts.type <- ddply(interaction.transcripts.TPM["A22"], .(Transcript), RxNply)
-A22.interaction.transcripts.type$max.val <- as.numeric(A22.interaction.transcripts.type$max.val)
-A22.interaction.transcripts.type$min.val <- as.numeric(A22.interaction.transcripts.type$min.val)
-A22.interaction.transcripts.type$opt.exp <- as.numeric(A22.interaction.transcripts.type$opt.exp)
-A22.interaction.transcripts.type$A22.exp.type <- as.factor(A22.interaction.transcripts.type$A22.exp.type)
-str(A22.interaction.transcripts.type)
-
-# Ar
-Ar.interaction.transcripts.type <- ddply(interaction.transcripts.TPM["Ar"], .(Transcript), RxNply)
-Ar.interaction.transcripts.type$max.val <- as.numeric(Ar.interaction.transcripts.type$max.val)
-Ar.interaction.transcripts.type$min.val <- as.numeric(Ar.interaction.transcripts.type$min.val)
-Ar.interaction.transcripts.type$opt.exp <- as.numeric(Ar.interaction.transcripts.type$opt.exp)
-Ar.interaction.transcripts.type$A22.exp.type <- as.factor(Ar.interaction.transcripts.type$A22.exp.type)
-str(Ar.interaction.transcripts.type)
-
-# combine into single data.table
-A22.interaction.transcripts.type$colony <- "A22"
-Ar.interaction.transcripts.type$colony <- "Ar"
-# first copy of temperature.transcripts.type as colony A22
-temperature.transcripts.type$colony <- "A22"
-# combine
-responsive.transcripts.type <- rbind(A22.interaction.transcripts.type, Ar.interaction.transcripts.type, temperature.transcripts.type)
-# add second copy of temperature.transcripts.type as colony Ar
-temperature.transcripts.type$colony <- "Ar"
-responsive.transcripts.type <- rbind(responsive.transcripts.type, temperature.transcripts.type)
-responsive.transcripts.type$colony <- as.factor(responsive.transcripts.type$colony)
-str(responsive.transcripts.type)
-```
-
-```{r expression_A22.exp.type_table, eval=TRUE, echo=FALSE, results='asis'}
+# calculate response type for transcripts with interactions
+interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
+stopifnot(nrow(interaction.response.type) == length(interaction.transcripts))
+
+# calculate response types for transcripts without interactions
+temperature.response.type <- ldply(temperature.transcripts, .progress = "text", .inform = TRUE, RxNtype)
+stopifnot(nrow(temperature.response.type) == length(temperature.transcripts))
+
+# merge results
+Ap.response.type <- rbind(interaction.response.type, temperature.response.type)
+colnames(Ap.response.type)[which(colnames(Ap.response.type) == ".id")] <- "Transcript"
+str(Ap.response.type)
+```
+
+Examine transcripts by temperature response. Compare between colonies.
+
+```{r temperature_response, eval=TRUE, echo=FALSE, results='asis'}
 # table type of response
-A22.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "A22", 'A22.exp.type'])
-Ar.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "Ar", 'A22.exp.type'])
-
-A22.exp.type.table <- rbind(A22.type.table, Ar.type.table)
-rownames(A22.exp.type.table) <- c("A22", "Ar")
-
-pandoc.table(A22.exp.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
+A22.type.table <- table(Ap.response.type[ , 'A22.type'])
+Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])
+
+Ap.type.table <- rbind(A22.type.table, Ar.type.table)
+rownames(Ap.type.table) <- c("A22", "Ar")
+
+pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
 ```
@@
 
-I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.
-
-```{r GO_contingency_test}
-# GO 'response to stress' hits in responsive transcripts
-responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)]
-
-resp.stress.responsive.count <- nrow(responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
-# GO 'response to stress' hits in all transcripts
-resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
-
-GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
-
-GO.stress.Xsq <- chisq.test(GO.stress.table)
-GO.stress.Xsq
-```
-
-There are also 4 heat shock related genes in the responsive transcripts, out of 130 total.
-
-```{r grep_shock}
-hsp_responsive <- responsive.transcripts[grep("shock", responsive.transcripts$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
-nrow(hsp_responsive)
-
-hsp_all <- annotationtable[grep("shock", annotationtable$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
-nrow(hsp_all)
-```
-
-## Functional annotation ##
-
-In the previous section, I identified transcripts that show significant responses in expression. Next, I add gene annotation and ontology information to these transcripts.
-
-```{r annotation}
-# add annotation information
-setkey(annotationtable, Sequence.Name)
-signif.transcripts <- data.table(signif.transcripts)
-setkey(signif.transcripts, Transcript)
-signif.transcripts <- annotationtable[signif.transcripts]
-setnames(signif.transcripts, "Sequence.Name", "Transcript")
-```
-
 ## Shiny interactive web-app
+
+**2014-04-02 -- CURRENTLY BROKEN**
 
@@
 
-```{r shiny_file}
+```{r shiny_file, eval=FALSE, echo=FALSE}
+shiny.transcripts <- TPM.dt.sub[Ap.response.type$Transcript]
+
 # scale expression values 
-responsive.transcripts.TPM <- TPM.dt.sub[responsive.transcripts]
-responsive.transcripts.TPM[,exp.scaled:=scale(TPM), by = Transcript]
+shiny.transcripts[,pTPM.scaled:=scale(pTPM), by = Transcript]
 str(responsive.transcripts.TPM)
@@
 
-In this section, I examine differences in thermal reaction norms of transcripts between *Ar* and *A22* using the categories as defined by `r citep("10.1086/675302")`:
-
-- *Offset, O*: overall difference in the mean expression value across all temperatures
-- *Slope, S*: difference in overall slope
-- *Curvature, C*: average difference in curvature of the reaction norm
-- *Wiggle, W*: variability in shape not captured by the previous three measures
-
-```{r offset, echo=TRUE, eval=TRUE}
-testset <- TPM.dt.sub[1:220, ]
-testoffset <- ddply(testset, .(colony, Transcript), summarise, offset = mean(TPM))
-A22meanTPM <- ddply(TPM.dt.sub, .(colony, Transcript), summarise, meanTPM = mean(TPM))
-
-```
+In this section, I compare the thermal reactionome between the *Ar* and *A22* colonies.
 
@@
 
-Compare the expression levels at optimum (19.25C) between the two colonies for genes in each expression group.
-Specifically, are genes that are upregulated at high temperatures in A22 more highly expressed at 21C in Ar?
-Conversely, are genes that are upregulated at low temps more highly expressed at 21C in A22?
+Selection may have acted in response to thermal stress such that some genes are constitutively activated in one colony while plastically-expressed in the other.
+I evaluate this hypothesis by comparing expression levels at the at optimum (19.5C) temperature between the two colonies for genes in that are either in the 'High' or 'Low' expression group in the other colony.
+Specifically, I expect that genes upregulated at high temperatures in *A22* are more highly expressed at 19.5C in *Ar*?
+Conversely, I expect genes that are upregulated at low temps in *Ar* to have greater constitutive expression in *A22*?
 
@@
 # list of transcripts that are 'high' expressed in A22
-A22_high_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$A22.exp.type == "High"), "Transcript"]
-# dataframe of transcripts from both colonies that are 'high' expressed in A22
-A22_high_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% A22_high_transcripts), ]
-
-# Compare expression at optimum temp (19.25C) between colonies using t-test
-A22_high_df$colony <- as.factor(A22_high_df$colony)
-boxplot(data = A22_high_df, opt.exp ~ colony)
-boxplot(data = A22_high_df, log(opt.exp+10) ~ colony)
-t.test(A22_high_df[which(A22_high_df$colony == "Ar"), "opt.exp"],  A22_high_df[which(A22_high_df$colony == "A22"), "opt.exp"])
-
-# remove outlier transcripts
-A22_high_df_out <- A22_high_df[-which(A22_high_df$opt.exp > 1000), ]
-t.test(A22_high_df_out[which(A22_high_df_out$colony == "Ar"), "opt.exp"],  A22_high_df_out[which(A22_high_df_out$colony == "A22"), "opt.exp"])
+A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
+
+# Compare expression at optimum temp (A22.opt) between colonies using t-test
+boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
+boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
+
+# T test
+t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
 ```
@@
 ```{r A22_high_wilcoxon}
-wilcox.test(A22_high_df[which(A22_high_df$colony == "A22"), "opt.exp"],  A22_high_df[which(A22_high_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
 ```
@@
 ```{r Ar_low_wilcoxon}
-# list of transcripts that are 'Low' expressed in Ar
-Ar_low_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$A22.exp.type == "Low"), "Transcript"]
-# dataframe of transcripts from both colonies that are 'Low' expressed in Ar
-Ar_low_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% Ar_low_transcripts), ]
-dim(Ar_low_df)
-
-# boxplots
-boxplot(data = Ar_low_df, opt.exp ~ colony)
-boxplot(data = Ar_low_df, log(opt.exp+1) ~ colony)
+# list of transcripts that are 'high' expressed in Ar
+Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
+
+# Compare expression at optimum temp (Ar.opt) between colonies using t-test
+boxplot(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
+boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
+
+# T test
+t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
 
 # Wilcoxon signed rank-test
-wilcox.test(Ar_low_df[which(Ar_low_df$colony == "A22"), "opt.exp"],  Ar_low_df[which(Ar_low_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-```
-
-No difference in transcript expression at 19.25C between colonies for transcripts that are up-regulated at low temperatures in *Ar*.
+wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+```
+
+Counter to expectations, expression at 19.5C is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*.
 
@@
 ```{r intermediate_transcripts}
+# list of transcripts that are 'Intermediate' expressed in Ar
+Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
+# T test
+t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
+# Wilcoxon signed rank-test
+wilcox.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+
 # list of transcripts that are 'Intermediate' expressed in A22
-A22_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
-A22_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% A22_int_transcripts), ]
-A22_int_df$colony <- as.factor(A22_int_df$colony)
-
+A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
+# T test
+t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
 # Wilcoxon signed rank-test
-wilcox.test(A22_int_df[which(A22_int_df$colony == "A22"), "opt.exp"],  A22_int_df[which(A22_int_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-
-# list of transcripts that are 'Intermediate' expressed in Ar
-Ar_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
-Ar_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% Ar_int_transcripts), ]
-
-# Wilcoxon signed rank-test
-wilcox.test(Ar_int_df[which(Ar_int_df$colony == "A22"), "opt.exp"],  Ar_int_df[which(Ar_int_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-```
-
-As expected, for "Intermediate" expressed transcripts there is no difference in expression at 19.25C between colonies.
+wilcox.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+```
+
+One of two tests, for *Ar* intermediate transcripts also finds greater constitutive expression in *Ar* than *A22*.
 
@@
 
-```{r calc_Intermediate_variance}
+```{r thermal_breadth}
 # extract 'Intermediate' expressed transcripts for A22 colony
-setkey(TPM.dt.sub, Transcript)
-A22trlist <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
-A22.TPM.int.dt <- TPM.dt.sub[A22trlist]
-setkey(A22.TPM.int.dt, colony)
-A22.TPM.int.dt <- A22.TPM.int.dt["A22"]
-str(A22.TPM.int.dt)
+A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
+length(A22.int.lm)
 
 # calculate standard deviation of expression for each transcript
-A22_int_sd <- ddply(A22.TPM.int.dt, .(Transcript), RxNsd.concave)
+transcriptVar <- function(lmitem, colony) {
+    # predict only for A22
+    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
+    newdf <- data.frame(val = pred.vals, colony = rep(colony, length = length(pred.vals)))
+    pout <- predict(lmitem, newdata=newdf)
+    pout <- cbind(newdf, pTPM = pout)
+
+    # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
+    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
+
+    # draw 1000 'temps' weighted by predicted expression
+    random.draw <- sample(size = 1000, pred.vals, prob = pout$pTPM, replace = TRUE)
+
+    # calculate standard deviation of weighted random.draw
+    data.frame(exp_sd = sd(random.draw))
+    } # end function
+
+# apply function to all transcripts
+A22.int.sd <- unlist(Map(transcriptVar, A22.int.lm, colony = "A22"))
 
 # repeat for Ar
-Artrlist <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
-Ar.TPM.int.dt <- TPM.dt.sub[Artrlist]
-setkey(Ar.TPM.int.dt, colony)
-Ar.TPM.int.dt <- Ar.TPM.int.dt["Ar"]
-str(Ar.TPM.int.dt)
-
-Ar_int_sd <- ddply(Ar.TPM.int.dt, .(Transcript), RxNsd.concave)
-```
-
-With these values calculated, I compare the standard deviation of expression between colonies.
-
-```{r compare_Intermediate_variance}
-# T-test
-t.test(A22_int_sd$exp_sd, Ar_int_sd$exp_sd)
-
-# Plot
-# prep data
-A22_int_var$colony <- "A22"
-Ar_int_var$colony <- "Ar"
-comb_int_var <- rbind(A22_int_var, Ar_int_var)
+Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
+Ar.int.sd <- unlist(Map(transcriptVar, Ar.int.lm, colony = "Ar"))
+
+# T-test comparing standard deviation of expression between colonies
+t.test(A22.int.sd, Ar.int.sd)
+
+```{r plotthermalbreadth, echo=FALSE, eval=TRUE)
+# prep data for plotting
+A22.int.sd <- data.frame(colony = "A22", exp_sd = A22.int.sd)
+Ar.int.sd <- data.frame(colony = "Ar", exp_sd = Ar.int.sd)
+Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)
+
 # ggplot
-g1 <- ggplot(comb_int_var, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
+g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
 g1 + scale_y_continuous(name="Density") +
@@
 Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
+
+~~Continue from here~~
 
@@
 
+In this section, I examine differences in thermal reaction norms of transcripts between *Ar* and *A22* using the categories as defined by `r citep("10.1086/675302")`:
+
+- *Offset, O*: overall difference in the mean expression value across all temperatures
+- *Slope, S*: difference in overall slope
+- *Curvature, C*: average difference in curvature of the reaction norm
+- *Wiggle, W*: variability in shape not captured by the previous three measures
+
+```{r offset, echo=TRUE, eval=TRUE}
+testset <- TPM.dt.sub[1:220, ]
+testoffset <- ddply(testset, .(colony, Transcript), summarise, offset = mean(TPM))
+A22meanTPM <- ddply(TPM.dt.sub, .(colony, Transcript), summarise, meanTPM = mean(TPM))
+
+```
+
+## Functional annotation ##
+
+In the previous section, I identified transcripts that show significant responses in expression. Next, I add gene annotation and ontology information to these transcripts.
+
+```{r annotation}
+# add annotation information
+setkey(annotationtable, Sequence.Name)
+signif.transcripts <- data.table(signif.transcripts)
+setkey(signif.transcripts, Transcript)
+signif.transcripts <- annotationtable[signif.transcripts]
+setnames(signif.transcripts, "Sequence.Name", "Transcript")
+```
+
 ## Gene set enrichment analysis ##
+
+I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.
+
+```{r GO_contingency_test}
+# GO 'response to stress' hits in responsive transcripts
+responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)]
+
+resp.stress.responsive.count <- nrow(responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
+# GO 'response to stress' hits in all transcripts
+resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
+
+GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
+
+GO.stress.Xsq <- chisq.test(GO.stress.table)
+GO.stress.Xsq
+```
+
+There are also 4 heat shock related genes in the responsive transcripts, out of 130 total.
+
+```{r grep_shock}
+hsp_responsive <- responsive.transcripts[grep("shock", responsive.transcripts$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
+nrow(hsp_responsive)
+
+hsp_all <- annotationtable[grep("shock", annotationtable$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
+nrow(hsp_all)
+```
 

### 2014-04-02 scripts/RxNseq.R

@@ -40,7 +40,10 @@ RxNtype <- function(lmitem) {
     
   # predict 
   pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
-  newdf <- data.frame(val = pred.vals, colony = as.factor(rep(unlist(lmitem$xlevels), each = length(pred.vals))))
+  colony.levels <- as.factor(rep(unlist(lmitem$xlevels)))
+  # for transcripts with no 'colony' effect in model, no levels, so create dummy
+  if(length(colony.levels) == 0) colony.levels <- c("A22", "Ar")
+  newdf <- data.frame(val = pred.vals, colony = rep(colony.levels, each = length(pred.vals)))
   pout <- predict(lmitem, newdata=newdf)
   pout <- cbind(newdf, pTPM = pout)
 
@@ -59,7 +62,7 @@ RxNtype <- function(lmitem) {
           # else, set values from data
           A22.max.val <- median(A22.pout[which(A22.pout$pTPM == max(A22.pout$pTPM)), "val"])
           # for min.val, take median point as it may be string of points at zero
-          qA22.min.val <- median(A22.pout[which(A22.pout$pTPM == min(A22.pout$pTPM)), "val"])
+          A22.min.val <- median(A22.pout[which(A22.pout$pTPM == min(A22.pout$pTPM)), "val"])
           A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
           # determine expression shape
           A22.exp.type <- NA

### 2014-04-03 ApTranscriptome_TR.Rmd
Updated analysis through comparison of thermal breadth for bimodally-expressed transcripts. Moved  function to scripts file.

@@
 colony.transcripts <- other.transcripts[setdiff(names(other.transcripts), names(temperature.transcripts))]
+responsive.transcripts <- c(temperature.transcripts, interaction.transcripts)
 rm(other.transcripts)
@@
 ```{r expression_shape, eval=TRUE, echo=TRUE}
-# calculate response type for transcripts with interactions
+# calculate response type for responsive transcripts
 interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
@@
 pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
+
+# only transcripts that differ in expression by colony
+A22.interaction.type.table <- table(interaction.response.type[ , 'A22.type'])
+Ar.interaction.type.table <- table(interaction.response.type[ , 'Ar.type'])
+
+Ap.interaction.type.table <- rbind(A22.interaction.type.table, Ar.interaction.type.table)
+rownames(Ap.interaction.type.table) <- c("A22", "Ar")
+
+pandoc.table(Ap.interaction.type.table, style="rmarkdown", caption = "For transcripts that differ in expression pattern between *A22* and *Ar*, the number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
+
 ```
@@
 # Compare expression at optimum temp (A22.opt) between colonies using t-test
+t.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
 boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
+
+# T test on log-transformed values to control for outliers
+t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
 boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
-
-# T test
-t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
 ```
@@
 
-# Compare expression at optimum temp (Ar.opt) between colonies using t-test
+# compare expression at optimum temp (Ar.opt) between colonies using t-test
+t.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
 boxplot(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
+# t-test with log values
+t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
 boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
-
-# T test
-t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
 
@@
 # T test
+t.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt)
 t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
@@
 # T test
+t.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt)
 t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
@@
 
-# calculate standard deviation of expression for each transcript
-transcriptVar <- function(lmitem, colony) {
-    # predict only for A22
-    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
-    newdf <- data.frame(val = pred.vals, colony = rep(colony, length = length(pred.vals)))
-    pout <- predict(lmitem, newdata=newdf)
-    pout <- cbind(newdf, pTPM = pout)
-
-    # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
-    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
-
-    # draw 1000 'temps' weighted by predicted expression
-    random.draw <- sample(size = 1000, pred.vals, prob = pout$pTPM, replace = TRUE)
-
-    # calculate standard deviation of weighted random.draw
-    data.frame(exp_sd = sd(random.draw))
-    } # end function
-
-# apply function to all transcripts
-A22.int.sd <- unlist(Map(transcriptVar, A22.int.lm, colony = "A22"))
+# apply `transcriptSD` function to all transcripts
+A22.int.sd <- unlist(Map(transcriptSD, A22.int.lm, colony = "A22"))
 
@@
 Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
-Ar.int.sd <- unlist(Map(transcriptVar, Ar.int.lm, colony = "Ar"))
+Ar.int.sd <- unlist(Map(transcriptSD, Ar.int.lm, colony = "Ar"))
 
@@
 t.test(A22.int.sd, Ar.int.sd)
+```
+Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
 
@@
 
-Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
-
-~~Continue from here~~
-
 ### Thermal sensitivity indicated by response of bimodally-expressed transcripts ###
@@
 # extract 'Bimodal' expressed transcripts for A22 colony
-A22trlist.bim <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$A22.exp.type == "Bimodal"), "Transcript"]
-A22.TPM.bim.dt <- TPM.dt.sub[A22trlist.bim]
-setkey(A22.TPM.bim.dt, colony)
-A22.TPM.bim.dt <- A22.TPM.bim.dt["A22"]
-str(A22.TPM.bim.dt)
-
-# calculate standard deviation of expression for each transcript
-A22_bim_sd <- ddply(A22.TPM.bim.dt, .(Transcript), RxNsd.convex)
+A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
+length(A22.bim.lm)
+
+# apply `transcriptSD` function to all transcripts
+A22.bim.sd <- unlist(Map(transcriptSD, A22.bim.lm, colony = "A22"))
 
 # repeat for Ar 
-Artrlist.bim <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$A22.exp.type == "Bimodal"), "Transcript"]
-Ar.TPM.bim.dt <- TPM.dt.sub[Artrlist.bim]
-setkey(Ar.TPM.bim.dt, colony)
-Ar.TPM.bim.dt <- Ar.TPM.bim.dt["Ar"]
-str(Ar.TPM.bim.dt)
-
-Ar_bim_sd <- ddply(Ar.TPM.bim.dt, .(Transcript), RxNsd.convex)
-```
-
-Next, I compare the standard deviations of expression for each transcipt between colonies.
-
-```{r compare_Intermediate_variance}
-# t-test
-t.test(A22_bim_sd$exp_sd, Ar_bim_sd$exp_sd)
-
-# Plot
-# prep data
-A22_bim_var$colony <- "A22"
-Ar_bim_var$colony <- "Ar"
-comb_bim_var <- rbind(A22_bim_var, Ar_bim_var)
-# ggplot
-g2 <- ggplot(comb_bim_var, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
-g2 + scale_y_continuous(name="Density") +
-  scale_x_continuous(name=expression("Standard deviation of expression function"))
-```
-
-Both the T-test and plot show that there is no difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.
+Ar.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]]
+length(Ar.bim.lm)
+Ar.bim.sd <- unlist(Map(transcriptSD, Ar.bim.lm, colony = "Ar"))
+
+# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
+t.test(A22.bim.sd, Ar.bim.sd)
+```
+
+No difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.
 

### 2014-04-03 scripts/RxNseq.R

@@ -45,8 +45,11 @@ RxNtype <- function(lmitem) {
   if(length(colony.levels) == 0) colony.levels <- c("A22", "Ar")
   newdf <- data.frame(val = pred.vals, colony = rep(colony.levels, each = length(pred.vals)))
   pout <- predict(lmitem, newdata=newdf)
-  pout <- cbind(newdf, pTPM = pout)
+  pout <- cbind(newdf, logTPM = pout)
 
+  # back-transform from log-scale
+  pout$pTPM <- exp(pout$logTPM)
+  
   # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
   pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
   
@@ -115,6 +118,36 @@ RxNtype <- function(lmitem) {
            
 } # end RxNtype
 
+
+############################################################################################
+## transcriptSD
+############################################################################################
+
+transcriptSD <- function(lmitem, colony) {
+    # function passed to Map() that calculates the standard deviation of expression for each 'Intermediate' transcript
+    #
+    # Arguments:
+    #  lmitem: List item of class `lm`
+    #  colony: Colony that standard deviation of expression is being calculated for
+    #
+    # Returns:
+    #   list of standard deviation of expression for each transcript
+    
+    # predict only for A22
+    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
+    newdf <- data.frame(val = pred.vals, colony = rep(colony, length = length(pred.vals)))
+    pout <- predict(lmitem, newdata=newdf)
+    pout <- cbind(newdf, pTPM = pout) # note this is on log scale
+
+    # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
+    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
+
+    # draw 1000 'temps' weighted by predicted expression
+    random.draw <- sample(size = 1000, pred.vals, prob = pout$pTPM, replace = TRUE)
+
+    # calculate standard deviation of weighted random.draw
+    data.frame(exp_sd = sd(random.draw))
+    } # end function
 
 
 

### 2014-04-03 ApTranscriptome_TR.Rmd
In progress to compare shape of reaction norms for all transcripts between colonies. Function 'RxNvar' added.

@@
 
-Cumulative distribution function of peak expression for transcripts that differ in expression between *A22* and *Ar*
-
-```{r max_exp_CDF}
-png(paste(resultsdir, "CDF_expression_all.png", sep=""))
-g3 <- ggplot(responsive.transcripts.type, aes(x=max.val, fill=colony)) + geom_density(alpha=0.2, position="identity")
+Probability distribution function of peak expression for transcripts that differ in expression between *A22* and *Ar*
+
+```{r max_exp_PDF}
+# reshape data
+Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("A22", "Ar"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
+
+png(paste(resultsdir, "PDF_expression_all.png", sep=""))
+g3 <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + geom_density(alpha=0.2, position="identity")
 g3 + scale_y_continuous(name="Density") +
@@
 dev.off()
-
-# merge data.tables of transcripts with colony x temperature interaction
-interaction.transcripts.type <- rbind(A22.interaction.transcripts.type, Ar.interaction.transcripts.type)
-png(paste(resultsdir, "CDF_expression_interaction.png", sep=""))
-g4 <- ggplot(interaction.transcripts.type, aes(x=max.val, fill=colony)) + geom_density(alpha=0.2, position="identity")
-g4 + scale_y_continuous(name="Density") +
-  scale_x_continuous(name="Temperature of maximum expression")
-dev.off()
-
-# faceted by type
-png(paste(resultsdir, "CDF_expression_intermediate_interaction.png", sep=""))
-g5 <- ggplot(interaction.transcripts.type, aes(x=max.val, group=colony)) + geom_density(alpha=0.2, position="identity")
-g5 + facet_grid(. ~ A22.exp.type) + scale_y_continuous(name="Density") +
-  scale_x_continuous(name="Temperature of maximum expression")
-dev.off()
-```
-
-In this section, I examine differences in thermal reaction norms of transcripts between *Ar* and *A22* using the categories as defined by `r citep("10.1086/675302")`:
+```
+
+For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect?
+To test this, I use the framework of `r citet("10.1086/675302")`:
 

### 2014-04-03 scripts/RxNseq.R

@@ -37,40 +37,43 @@ RxNtype <- function(lmitem) {
     #
     # Returns
     #   dataframe with values of max expression, min expression and expression type for each colony
-    
-  # predict 
-  pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
-  colony.levels <- as.factor(rep(unlist(lmitem$xlevels)))
-  # for transcripts with no 'colony' effect in model, no levels, so create dummy
-  if(length(colony.levels) == 0) colony.levels <- c("A22", "Ar")
-  newdf <- data.frame(val = pred.vals, colony = rep(colony.levels, each = length(pred.vals)))
-  pout <- predict(lmitem, newdata=newdf)
-  pout <- cbind(newdf, logTPM = pout)
 
-  # back-transform from log-scale
-  pout$pTPM <- exp(pout$logTPM)
+    # check class lm
+    if(!"lm" %in% class(lmitem)) stop("Object not of class lm")
+    
+    # predict 
+    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
+    colony.levels <- as.factor(rep(unlist(lmitem$xlevels)))
+    # for transcripts with no 'colony' effect in model, no levels, so create dummy
+    if(length(colony.levels) == 0) colony.levels <- c("A22", "Ar")
+    newdf <- data.frame(val = pred.vals, colony = rep(colony.levels, each = length(pred.vals)))
+    pout <- predict(lmitem, newdata=newdf)
+    pout <- cbind(newdf, logTPM = pout)
+
+    # back-transform from log-scale
+    pout$pTPM <- exp(pout$logTPM)
   
-  # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
-  pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
+    # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
+    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
   
-  # calculate for A22
-  A22.pout <- pout[pout$colony == "A22", ]
-  # if pTPM doesn't vary, set max and min values to NA
-  if(sd(A22.pout$pTPM) < 0.01) {
-      A22.max.val <- NA
-      A22.min.val <- NA
-      A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
-      A22.exp.type <- "NotExp"
-  } else {
[…]

### 2014-04-03 ApTranscriptome_TR.Rmd
Completed analyisis to compare shape of reaction norm for all transcripts using 'RxNvarshape' function

@@
 ```
+
 Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
@@
 ```{r offset, echo=TRUE, eval=TRUE}
-testset <- TPM.dt.sub[1:220, ]
-testoffset <- ddply(testset, .(colony, Transcript), summarise, offset = mean(TPM))
-A22meanTPM <- ddply(TPM.dt.sub, .(colony, Transcript), summarise, meanTPM = mean(TPM))
+testset <- responsive.transcripts[1:10]
+varshape.out <- ldply(responsive.transcripts, .progress = "text", RxNvarshape)
+
+# to compare among transcripts, standardize by overall mean for each transcript
+
+# t test
+t.test(varshape.out$A22.mean, varshape.out$Ar.mean)
+t.test(varshape.out$A22.slope, varshape.out$Ar.slope)
+t.test(varshape.out$A22.curve, varshape.out$Ar.curve)
+t.test(varshape.out$A22.wiggle, varshape.out$Ar.wiggle)
 

### 2014-04-03 scripts/RxNseq.R

@@ -180,27 +180,29 @@ RxNvarshape <- function(lmitem) {
   
     # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
     pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
-  
+    
     # calculate for A22
     A22.pout <- pout[pout$colony == "A22", ]
-    A22.mean <- mean(A22.pout$pTPM)
+    A22.mean <- round(mean(A22.pout$pTPM), 4)
     Si <- diff(A22.pout$pTPM)
-    A22.slope <- abs(sum(Si))/(nrow(A22.pout)-1)
+    A22.slope <- round(abs(sum(Si))/(nrow(A22.pout)-1), 4)
     Ci <- diff(Si)
-    A22.curve <- abs(sum(Ci))/(nrow(A22.pout)-2)
-    A22.wiggle <- (sum(abs(Ci))/(nrow(A22.pout)-2)) - A22.curve
-
+    A22.curve <- round(abs(sum(Ci))/(nrow(A22.pout)-2), 4)
+    A22.wiggle <- round((sum(abs(Ci))/(nrow(A22.pout)-2)) - A22.curve, 4)
+    
     # calculate for Ar
     Ar.pout <- pout[pout$colony == "Ar", ]
-    Ar.mean <- mean(Ar.pout$pTPM)
+    Ar.mean <- round(mean(Ar.pout$pTPM), 4)
     Si <- diff(Ar.pout$pTPM)
-    Ar.slope <- abs(sum(Si))/(nrow(Ar.pout)-1)
+    Ar.slope <- round(abs(sum(Si))/(nrow(Ar.pout)-1), 4)
     Ci <- diff(Si)
-    Ar.curve <- abs(sum(Ci))/(nrow(Ar.pout)-2)
-    Ar.wiggle <- (sum(abs(Ci))/(nrow(Ar.pout)-2)) - Ar.curve
-
+    Ar.curve <- round(abs(sum(Ci))/(nrow(Ar.pout)-2), 4)
+    Ar.wiggle <- round((sum(abs(Ci))/(nrow(Ar.pout)-2)) - Ar.curve, 4)
+    
+    
+    
     # combine into data.frame
-    c(A22.mean, A22.slope, A22.curve, A22.wiggle, Ar.mean, Ar.slope, Ar.curve, Ar.wiggle)
[…]

### 2014-04-04 ApTranscriptome_TR.Rmd
Analysis complete through comparison of differences in reaction norms between colonies using offset/slope/curvature/wiggle.
Started updates to manuscript.

@@
 ```{r offset, echo=TRUE, eval=TRUE}
-testset <- responsive.transcripts[1:10]
-varshape.out <- ldply(responsive.transcripts, .progress = "text", RxNvarshape)
-
-# to compare among transcripts, standardize by overall mean for each transcript
+varshape.out <- ldply(interaction.transcripts, .progress = "text", RxNvarshape)
+
+# plot
+boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
+boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))
+
+# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
+# offset = mean1 - mean2
+varshape.out$offset <- varshape.out$A22.mean - varshape.out$Ar.mean
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.offset <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.offset"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.mean"], varshape.out[i, "Ar.mean"])))
+}
+
+# t-test
+t.test(varshape.out$s.offset, alternative = "two.sided")
+
+# slope = slope1 - slope2
+varshape.out$slope <- varshape.out$A22.slope - varshape.out$Ar.slope
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.slope <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.slope"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.slope"], varshape.out[i, "Ar.slope"])))
+}
+# for transcripts where both samples had slope=0, change s.slope from Inf to 0 as there is truly no difference in slope
+varshape.out[which(varshape.out$Ar.slope == 0 & varshape.out$A22.slope == 0), "s.slope"] <- 0
+
+# t-test
+t.test(varshape.out$s.slope, alternative = "two.sided")
+
+# curvature = curvature1 - curvature2
+varshape.out$curvature <- varshape.out$A22.curve - varshape.out$Ar.curve
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.curvature <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.curvature"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.curve"], varshape.out[i, "Ar.curve"])))
+}
+# change 'Inf' values to 0
+varshape.out[which(varshape.out$Ar.curve == 0 & varshape.out$A22.curve == 0), "s.curvature"] <- 0
+# t-test
+t.test(varshape.out$s.curvature, alternative = "two.sided")
+
+# wiggle
+varshape.out$wiggle <- varshape.out$A22.wiggle - varshape.out$Ar.wiggle
+varshape.out$s.wiggle <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.wiggle"] <- (varshape.out[i, "wiggle"] / mean(c(varshape.out[i, "A22.wiggle"], varshape.out[i, "Ar.wiggle"])))
+}
+# change 'Inf' values to 0
+varshape.out[which(varshape.out$Ar.curve == 0 & varshape.out$A22.curve == 0), "s.wiggle"] <- 0
+# note that only 407 transcripts have any 'wiggle'
+length(which(!is.na(varshape.out$s.wiggle)))
+
+# t-test
+t.test(varshape.out$s.wiggle, alternative = "two.sided")
+```
+
+Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means (offset), slope, curvature and wiggle.
+
+```{r partition_reaction_norm_variation}
+# take absolute value of each value and sum to get total differences between reaction norms
+varshape.out$s.total <- abs(varshape.out$s.offset) + abs(varshape.out$s.slope) + abs(varshape.out$s.curvature) + abs(varshape.out$s.wiggle)
+
+t.test(varshape.out$s.total, alternative = "two.sided")
+
+```
+
+# table
+pandoc.table(varshape.out[which(abs(varshape.out$offset)>1000), c(".id", "A22.mean", "Ar.mean", "offset")])
 
 # t test
-t.test(varshape.out$A22.mean, varshape.out$Ar.mean)
+t.test(varshape.out$A22.mean, varshape.out$Ar.mean, paired = TRUE)
+t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean), paired = TRUE)
+
+wilcox.test(varshape.out$A22.mean, varshape.out$Ar.mean, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+
 t.test(varshape.out$A22.slope, varshape.out$Ar.slope)

### 2014-04-07 ApTranscriptome_TR.Rmd
Working on manuscript

--- 2014-04-03 ApTranscriptome_TR.Rmd
@@
 
-**March 31, 2014**
+**April 4, 2014**
 
@@
 A22.int.sd <- unlist(Map(transcriptSD, A22.int.lm, colony = "A22"))
+A22.int.sd <- data.frame(colony = "A22", exp_sd = A22.int.sd)
 
@@
 Ar.int.sd <- unlist(Map(transcriptSD, Ar.int.lm, colony = "Ar"))
+Ar.int.sd <- data.frame(colony = "Ar", exp_sd = Ar.int.sd)
 
 # T-test comparing standard deviation of expression between colonies
-t.test(A22.int.sd, Ar.int.sd)
+t.test(A22.int.sd$exp_sd, Ar.int.sd$exp_sd)
 ```
@@
 # prep data for plotting
-A22.int.sd <- data.frame(colony = "A22", exp_sd = A22.int.sd)
-Ar.int.sd <- data.frame(colony = "Ar", exp_sd = Ar.int.sd)
 Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)
@@
 ```{r offset, echo=TRUE, eval=TRUE}
-testset <- responsive.transcripts[1:10]
-varshape.out <- ldply(responsive.transcripts, .progress = "text", RxNvarshape)
-
-# to compare among transcripts, standardize by overall mean for each transcript
-
-# t test
-t.test(varshape.out$A22.mean, varshape.out$Ar.mean)
-t.test(varshape.out$A22.slope, varshape.out$Ar.slope)
-t.test(varshape.out$A22.curve, varshape.out$Ar.curve)
-t.test(varshape.out$A22.wiggle, varshape.out$Ar.wiggle)
-
+varshape.out <- ldply(interaction.transcripts, .progress = "text", RxNvarshape)
+
+# plot
+boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
+boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))
+
+# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
+# offset = mean1 - mean2
+varshape.out$offset <- varshape.out$A22.mean - varshape.out$Ar.mean
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.offset <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.offset"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.mean"], varshape.out[i, "Ar.mean"])))
+}
+
+# t-test
+t.test(varshape.out$s.offset, alternative = "two.sided")
+
+# slope = slope1 - slope2
+varshape.out$slope <- varshape.out$A22.slope - varshape.out$Ar.slope
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.slope <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.slope"] <- (varshape.out[i, "slope"] / mean(c(varshape.out[i, "A22.slope"], varshape.out[i, "Ar.slope"])))
+}
+# for transcripts where both samples had slope=0, change s.slope from Inf to 0 as there is truly no difference in slope
+varshape.out[which(varshape.out$Ar.slope == 0 & varshape.out$A22.slope == 0), "s.slope"] <- 0
+
+# t-test
+t.test(varshape.out$s.slope, alternative = "two.sided")
+
+# curvature = curvature1 - curvature2
+varshape.out$curvature <- varshape.out$A22.curve - varshape.out$Ar.curve
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.curvature <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.curvature"] <- (varshape.out[i, "curvature"] / mean(c(varshape.out[i, "A22.curve"], varshape.out[i, "Ar.curve"])))
+}
+# change 'Inf' values to 0
+varshape.out[which(varshape.out$Ar.curve == 0 & varshape.out$A22.curve == 0), "s.curvature"] <- 0
+# t-test
+t.test(varshape.out$s.curvature, alternative = "two.sided")
+
+# wiggle
+varshape.out$wiggle <- varshape.out$A22.wiggle - varshape.out$Ar.wiggle
+varshape.out$s.wiggle <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.wiggle"] <- (varshape.out[i, "wiggle"] / mean(c(varshape.out[i, "A22.wiggle"], varshape.out[i, "Ar.wiggle"])))
+}
+# change 'Inf' values to 0
+varshape.out[which(varshape.out$Ar.wiggle == 0 & varshape.out$A22.wiggle == 0), "s.wiggle"] <- 0
+# note that only 407 transcripts have any 'wiggle'
+length(which(varshape.out$s.wiggle != 0))
+
+# t-test
+t.test(varshape.out$s.wiggle, alternative = "two.sided")
+```
+
+Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means (offset), slope, curvature and wiggle.
+
+```{r partition_reaction_norm_variation}
+# take absolute value of each value and sum to get total differences between reaction norms
+varshape.out$s.total <- abs(varshape.out$s.offset) + abs(varshape.out$s.slope) + abs(varshape.out$s.curvature) + abs(varshape.out$s.wiggle)
+
+t.test(varshape.out$s.total, alternative = "two.sided")
+
+# variation in reaction norms due to differences in mean
+varshape.out$prop.mean <- abs(varshape.out$s.offset) / varshape.out$s.total
+varshape.out$prop.slope <- abs(varshape.out$s.slope) / varshape.out$s.total
+varshape.out$prop.curve <- abs(varshape.out$s.curvature) / varshape.out$s.total
+varshape.out$prop.wiggle <- abs(varshape.out$s.wiggle) / varshape.out$s.total
+
+mean(varshape.out$prop.mean)
+mean(varshape.out$prop.slope)
+mean(varshape.out$prop.curve)
+mean(varshape.out$prop.wiggle)
 ```

### 2014-04-07 ApTranscriptome_TR.Rmd
Writing

@@
 
-$$ TPM = \beta_0 + \beta_1(colony) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(colony * temp) + \\beta_5(colony * temp^2) + \epsilon $$
+$$ log(TPM + 1) = \beta_0 + \beta_1(colony) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(colony * temp) + \beta_5(colony * temp^2) + \epsilon $$
 
@@
 
-pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
-
-# only transcripts that differ in expression by colony
-A22.interaction.type.table <- table(interaction.response.type[ , 'A22.type'])
-Ar.interaction.type.table <- table(interaction.response.type[ , 'Ar.type'])
-
-Ap.interaction.type.table <- rbind(A22.interaction.type.table, Ar.interaction.type.table)
-rownames(Ap.interaction.type.table) <- c("A22", "Ar")
-
-pandoc.table(Ap.interaction.type.table, style="rmarkdown", caption = "For transcripts that differ in expression pattern between *A22* and *Ar*, the number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
-
+# re-orderfor printing
+Ap.type.table <- Ap.type.table[ , c(2,4,1,3,5)]
+
+pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of thermally-responsive transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not expressed for each colony.")
+
+# Pearson Chi-square test
+chisq.test(Ap.type.table)
 ```
@@
 # T-test comparing standard deviation of expression between colonies
-t.test(A22.int.sd$exp_sd, Ar.int.sd$exp_sd)
+t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd)
 ```
@@
 # t-test to compare standard deviation of 'Bimodal' transcripts between colonies
-t.test(A22.bim.sd, Ar.bim.sd)
+t.test(Ar.bim.sd, A22.bim.sd)
 ```
@@
 
-Probability distribution function of peak expression for transcripts that differ in expression between *A22* and *Ar*
+Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*
 
@@
 # reshape data
-Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("A22", "Ar"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
+Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "ApNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
 
@@
 ```{r partition_reaction_norm_variation}
+# weighted mean-standardized values of each measure
+mean(abs((varshape.out$s.offset)))
+mean(abs((varshape.out$s.slope)))
+mean(abs((varshape.out$s.curvature)))
+mean(abs((varshape.out$s.wiggle)))
+
 # take absolute value of each value and sum to get total differences between reaction norms
@@
 
+# Mean proportion and 95% CI of total variation of each measure
 mean(varshape.out$prop.mean)
+quantile(varshape.out$prop.mean, probs = c(0.05, 0.5, 0.95))
+
 mean(varshape.out$prop.slope)
+quantile(varshape.out$prop.slope, probs = c(0.05, 0.5, 0.95))
+
 mean(varshape.out$prop.curve)
+quantile(varshape.out$prop.curve, probs = c(0.05, 0.5, 0.95))
+
 mean(varshape.out$prop.wiggle)
+quantile(varshape.out$prop.wiggle, probs = c(0.05, 0.5, 0.95))
 ```

### 2014-04-09 ApTranscriptome_TR.Rmd
Gene set enrichment analysis for expression type overall and each colony

@@
 
-**April 4, 2014**
+**April 8, 2014**
 
@@
 Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])
-
-Ap.type.table <- rbind(A22.type.table, Ar.type.table)
-rownames(Ap.type.table) <- c("A22", "Ar")
+# overlap
+overlap <- c(length(Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]),
+             length(Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type == "High"), "Transcript"]),
+             length(Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]),
+             length(Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]),
+             length(Ap.response.type[which(Ap.response.type$A22.type == "NotExp" & Ap.response.type$Ar.type == "NotExp"), "Transcript"]))
+
+# make table
+Ap.type.table <- rbind(A22.type.table, Ar.type.table, overlap)
+rownames(Ap.type.table) <- c("A22", "Ar", "Both")
 
@@
 
-pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of thermally-responsive transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not expressed for each colony.")
+pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of thermally-responsive transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not expressed for each colony and their overlap.")
 
@@
 
-```{r plotthermalbreadth, echo=FALSE, eval=TRUE)
+```{r plotthermalbreadth, echo=FALSE, eval=TRUE}
 # prep data for plotting
@@
 ```{r GO_contingency_test}
+responsive.transcripts.ann <- signif.transcripts[names(responsive.transcripts)]
 # GO 'response to stress' hits in responsive transcripts
-responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)]
-
-resp.stress.responsive.count <- nrow(responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
+responsive.transcripts.ann[grep("GO:0006950", responsive.transcripts.ann$GO.Biological.Process), list(Transcript, best.hit.to.nr)]
+
+resp.stress.responsive.count <- nrow(responsive.transcripts.ann[grep("GO:0006950", responsive.transcripts.ann$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
 # GO 'response to stress' hits in all transcripts
@@
 
-GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
+GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts.ann) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
 
@@
 
-There are also 4 heat shock related genes in the responsive transcripts, out of 130 total.
+There are also 6 heat shock related genes in the responsive transcripts, out of 130 total.
 
 ```{r grep_shock}
-hsp_responsive <- responsive.transcripts[grep("shock", responsive.transcripts$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
+hsp_responsive <- responsive.transcripts.ann[grep("shock", responsive.transcripts.ann$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
 nrow(hsp_responsive)
@@
 # so set any NA to 1 as need to retain for GO analysis
-Ap.geneList <- RxNout$padj
+Ap.geneList <- RxNpval$padj
 Ap.geneList[which(is.na(Ap.geneList))] <- 1
 stopifnot(length(which(is.na(Ap.geneList))) == 0)
-names(Ap.geneList) <- RxNout$Transcript
+names(Ap.geneList) <- RxNpval$Transcript
 str(Ap.geneList)
@@
 # table results
-Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 118)
+Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 82)
 #Ap.BP.ResTable
@@
 
-**2) High and low expressed transcripts**
-
-Use `selectTranscript` function to select transcripts based on 'A22.exp.type'. As only significant transcripts have an 'A22.exp.type' assigned, this is a small subset of the above.
-
-```{r A22.exp.type_function}
-selectTranscript <- function(score) {
-    return(score == 1)
-}
-
-# To select genes, add 'A22.exp.type' to RxN df
-RxNout.t <- merge(RxNout, Ap.A22.exp.type, by="Transcript", all=TRUE)
-length(which(!is.na(RxNout.t$A22.exp.type)))
-```
-
-```{r bim.gsea, echo=TRUE, eval=TRUE}
-# create geneList
-Ap.bim.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.bim.geneList[which(RxNout.t$A22.exp.type=="Bimodal")] <- 1
-names(Ap.bim.geneList) <- RxNout.t$Transcript
-str(Ap.bim.geneList)
-table(Ap.bim.geneList)
-
-# create topGOdata object
-Ap.bim.BP.GOdata <- new("topGOdata",
-                 description = "BP gene set analysis",
-                 ontology = "BP",
-                 allGenes = Ap.bim.geneList,
-                 geneSel = selectTranscript,
-                 nodeSize = 10,
-                 annot = annFUN.gene2GO,
-                 gene2GO = geneID2GO)
-
-Ap.bim.BP.GOdata
-
-# perform enrichment analysis using parentchild method
-Ap.bim.BP.resultParentChild <- runTest(Ap.bim.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
-Ap.bim.BP.resultParentChild
-```
-
-Enriched gene sets for genes expressed at both high and low temperatures.
-
-```{r, echo=FALSE, results="asis"}
-Ap.bim.BP.ResTable <- GenTable(Ap.bim.BP.GOdata, parentchild = Ap.bim.BP.resultParentChild, topNodes = 52)
-#Ap.bim.BP.ResTable
-write.table(Ap.bim.BP.ResTable, file = paste(resultsdir, "Ap_bimodal_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
-pandoc.table(Ap.bim.BP.ResTable)
-```
-
-**3) Intermediate-temperature expressed transcripts**
-
-```{r int.gsea, echo=TRUE, eval=TRUE}
-# create geneList
-Ap.int.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.int.geneList[which(RxNout.t$A22.exp.type=="Intermediate")] <- 1
-names(Ap.int.geneList) <- RxNout.t$Transcript
-str(Ap.int.geneList)
-table(Ap.int.geneList)
-
-# create topGOdata object
-Ap.int.BP.GOdata <- new("topGOdata",
-                 description = "BP gene set analysis",
-                 ontology = "BP",
-                 allGenes = Ap.int.geneList,
-                 geneSel = selectTranscript,
-                 nodeSize = 10,
-                 annot = annFUN.gene2GO,
-                 gene2GO = geneID2GO)
-
-Ap.int.BP.GOdata
-
-# perform enrichment analysis using parentchild method
-Ap.int.BP.resultParentChild <- runTest(Ap.int.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
-Ap.int.BP.resultParentChild
-```
-
-Enriched gene sets for genes expressed at intermediate temperatures.
-
-```{r, echo=FALSE, results="asis"}
-Ap.int.BP.ResTable <- GenTable(Ap.int.BP.GOdata, parentchild = Ap.int.BP.resultParentChild, topNodes=16)
-Ap.int.BP.ResTable
-write.table(Ap.int.BP.ResTable, file = paste(resultsdir, "Ap_intermediate_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
-pandoc.table(Ap.int.BP.ResTable)
-```
-
-**4) High temperature expressed transcripts**
-
-```{r hig.gsea, echo=TRUE, eval=TRUE}
-# create geneList
-Ap.hig.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.hig.geneList[which(RxNout.t$A22.exp.type=="High")] <- 1
-names(Ap.hig.geneList) <- RxNout.t$Transcript
-str(Ap.hig.geneList)
-table(Ap.hig.geneList)
-
-# create topGOdata object
-Ap.hig.BP.GOdata <- new("topGOdata",
-                 description = "BP gene set analysis",
-                 ontology = "BP",
-                 allGenes = Ap.hig.geneList,
-                 geneSel = selectTranscript,
-                 nodeSize = 10,
-                 annot = annFUN.gene2GO,
-                 gene2GO = geneID2GO)
-
-Ap.hig.BP.GOdata
-
-# perform enrichment analysis using parentchild method
-Ap.hig.BP.resultParentChild <- runTest(Ap.hig.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
-Ap.hig.BP.resultParentChild
-```
-
-Enriched gene sets for genes expressed at both high temperatures only.
-
-```{r, echo=FALSE, results="asis"}
-Ap.hig.BP.ResTable <- GenTable(Ap.hig.BP.GOdata, parentchild = Ap.hig.BP.resultParentChild, topNodes=12)
-#Ap.hig.BP.ResTable
-write.table(Ap.hig.BP.ResTable, file = paste(resultsdir, "Ap_high_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
-pandoc.table(Ap.hig.BP.ResTable)
-```
-
-**5) Low temperature expressed transcripts**
-
-```{r low.gsea, echo=TRUE, eval=TRUE}
-# create geneList
-Ap.low.geneList <- rep(0, length = nrow(RxNout.t))
-Ap.low.geneList[which(RxNout.t$A22.exp.type=="Low")] <- 1
-names(Ap.low.geneList) <- RxNout.t$Transcript
-str(Ap.low.geneList)
-table(Ap.low.geneList)
-
-# create topGOdata object
-Ap.low.BP.GOdata <- new("topGOdata",
-                 description = "BP gene set analysis",
-                 ontology = "BP",
-                 allGenes = Ap.low.geneList,
-                 geneSel = selectTranscript,
-                 nodeSize = 10,
-                 annot = annFUN.gene2GO,
-                 gene2GO = geneID2GO)
-
-Ap.low.BP.GOdata
-
-# perform enrichment analysis using parentchild method
-Ap.low.BP.resultParentChild <- runTest(Ap.low.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
-Ap.low.BP.resultParentChild
-```
-
-Enriched gene sets for genes expressed at low temperatures only.
-
-```{r, echo=FALSE, results="asis"}
-Ap.low.BP.ResTable <- GenTable(Ap.low.BP.GOdata, parentchild = Ap.low.BP.resultParentChild, topNodes=26)
-Ap.low.BP.ResTable
-write.table(Ap.low.BP.ResTable, file = paste(resultsdir, "Ap_low_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
+**2) High**
+
+Genes with *High* expression in both colonies
+
+Use `selectTranscript` function to select transcripts from 'Ap.response.type' for GSEA.
+
+```{r gsea_high}
+# Select transcripts
+Ap.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High" & Ap.response.type$A22.type == "High"), "Transcript"]
+Ap.geneList.high <- rep(0, length = length(Ap.geneList))
+names(Ap.geneList.high) <- names(Ap.geneList)
+Ap.geneList.high[(which(names(Ap.geneList.high) %in% Ap.high))] <- 1
+# check correct number of values set to 1
+table(Ap.geneList.high)
+
+# Run GSEA
+Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO)
+pandoc.table(Ap.high.gsea)
+```
+
+**3) Low**
+
+Genes with *Low* expression in both colonies
+
+```{r gsea_low}
+# Select transcripts
+Ap.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low" & Ap.response.type$A22.type == "Low"), "Transcript"]
+Ap.geneList.low <- rep(0, length = length(Ap.geneList))
+names(Ap.geneList.low) <- names(Ap.geneList)
+Ap.geneList.low[(which(names(Ap.geneList.low) %in% Ap.low))] <- 1
+# check correct number of values set to 1
+table(Ap.geneList.low)
+
+# Run GSEA
+Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
 pandoc.table(Ap.low.BP.ResTable)
@@
 
-```{r, echo=FALSE}
-# graph significant nodes
-pdf(paste(resultsdir, "Ap_bimodal_BP_topGO_sig_nodes.pdf", sep=""))
-showSigOfNodes(Ap.bim.BP.GOdata, score(Ap.bim.BP.resultParentChild), firstSigNodes = 52, useInfo = 'all')
-dev.off()
-
-pdf(paste(resultsdir, "Ap_intermediate_BP_topGO_sig_nodes.pdf", sep=""))
-showSigOfNodes(Ap.int.BP.GOdata, score(Ap.int.BP.resultParentChild), firstSigNodes = 16, useInfo = 'all')
-dev.off()
-
-pdf(paste(resultsdir, "Ap_high_BP_topGO_sig_nodes.pdf", sep=""))
-showSigOfNodes(Ap.hig.BP.GOdata, score(Ap.hig.BP.resultParentChild), firstSigNodes = 12, useInfo = 'all')
-dev.off()
-
-pdf(paste(resultsdir, "Ap_low_BP_topGO_sig_nodes.pdf", sep=""))
-showSigOfNodes(Ap.low.BP.GOdata, score(Ap.low.BP.resultParentChild), firstSigNodes = 26, useInfo = 'all')
-dev.off()
+**4) Bimodal**
+
+Genes with *Bimodal* expression in both colonies
+
+```{r gsea_bimodal}
+Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
+names(Ap.geneList.bim) <- names(Ap.geneList)
+Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
+# check correct number of values set to 1
+table(Ap.geneList.bim)
+
+# Run GSEA
+Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
+pandoc.table(Ap.bim.gsea)
+```
+
+**5) Intermediate**
+
+Genes with *Intermediate* expression in both colonies
+
+```{r gsea_intermediate}
+Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+# create gene list, setting value to 1 for "int" transcripts
+Ap.geneList.int <- rep(0, length = length(Ap.geneList))
+names(Ap.geneList.int) <- names(Ap.geneList)
+Ap.geneList.int[(which(names(Ap.geneList.int) %in% Ap.int))] <- 1
+# check correct number of values set to 1
+table(Ap.geneList.int)
+
+# Run GSEA
+Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
+pandoc.table(Ap.int.gsea)
+```
+
+Next, I perform GSEA for genes in each functional type in one colony but not the other (e.g. the set difference) to gain insight on differences between the colonies.
+
+**A22 'High' genes not in Ar**
+
+```{r gsea_high_A22}
+A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+A22.geneList.high <- rep(0, length = length(Ap.geneList))
+names(A22.geneList.high) <- names(Ap.geneList)
+A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
+# check correct number of values set to 1
+table(A22.geneList.high)
+
+# Run GSEA
+A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
+pandoc.table(A22.high.gsea)
+```
+
+**Ar 'High' genes not in Ar**
+
+```{r gsea_high_Ar}
+Ar.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+Ar.geneList.high <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.high) <- names(Ap.geneList)
+Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high))] <- 1
+# check correct number of values set to 1
+table(Ar.geneList.high)
+
+# Run GSEA
+Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
+pandoc.table(Ar.high.gsea)
+```
+
+**A22 'Low' genes not in Ar**
+
+```{r gsea_low_A22}
+A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+A22.geneList.low <- rep(0, length = length(Ap.geneList))
+names(A22.geneList.low) <- names(Ap.geneList)
+A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low))] <- 1
+# check correct number of values set to 1
+table(A22.geneList.low)
+
+# Run GSEA
+A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
+pandoc.table(A22.low.gsea)
+```
+
+**Ar 'Low' genes not in Ar**
+
+```{r gsea_low_Ar}
+Ar.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+Ar.geneList.low <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.low) <- names(Ap.geneList)
+Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low))] <- 1
+# check correct number of values set to 1
+table(Ar.geneList.low)
+
+# Run GSEA
+Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
+pandoc.table(Ar.low.gsea)
+```
+
+**A22 'Bimodal' genes not in Ar**
+
+```{r gsea_bim_A22}
+A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+A22.geneList.bim <- rep(0, length = length(Ap.geneList))
+names(A22.geneList.bim) <- names(Ap.geneList)
+A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim))] <- 1
+# check correct number of values set to 1
+table(A22.geneList.bim)
+
+# Run GSEA
+A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
+pandoc.table(A22.bim.gsea)
+```
+
+**Ar 'Bimodal' genes not in Ar**
+
+```{r gsea_bim_Ar}
+Ar.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.bim) <- names(Ap.geneList)
+Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim))] <- 1
+# check correct number of values set to 1
+table(Ar.geneList.bim)
+
+# Run GSEA
+Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
+pandoc.table(Ar.bim.gsea)
+```
+
+**A22 'Intermediate' genes not in Ar**
+
+```{r gsea_int_A22}
+A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+A22.geneList.int <- rep(0, length = length(Ap.geneList))
+names(A22.geneList.int) <- names(Ap.geneList)
+A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int))] <- 1
+# check correct number of values set to 1
+table(A22.geneList.int)
+
+# Run GSEA
+A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
+pandoc.table(A22.int.gsea)
+```
+
+**Ar 'Low' genes not in Ar**
+
+```{r gsea_int_Ar}
+Ar.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+# create gene list, setting value to 1 for "bim" transcripts
+Ar.geneList.int <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.int) <- names(Ap.geneList)
+Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int))] <- 1
+# check correct number of values set to 1
+table(Ar.geneList.int)
+
+# Run GSEA
+Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO, plotpath = "results/Ar_int_BP_topGO_sig_nodes.pdf")
+pandoc.table(Ar.int.gsea)
 ```

### 2014-04-09 scripts/RxNseq.R

@@ -238,3 +238,47 @@ geneid2GOmap <- function(annotmat) {
   } # end for loop
 } # end function
 
+
+############################################################################################
+## geneid2GOmap
+############################################################################################
+
+gsea <- function(genelist, geneID2GO, plotpath=NA) {
+  # Run topGO using parentChild 
+  #
+  # Args:
+  #  genelist: list of genes with 0 for non-significant and 1 for significant genes. 'names' need to correspond to gene names in geneID2GO file
+  # geneID2GO: gene read mappings created using `geneID2GOMap` function from topGO
+  # plotpath: file path to save plot of significant nodes, max of 10 due to size limitations. if NA, no plot is made
+  #
+  # Returns:
+  #  data.frame with enriched GO terms at p < 0.01
+  
+  # create topGOdata object
+  GOdata <- new("topGOdata",
+                description = "BP gene set analysis",
+                ontology = "BP",
+                allGenes = genelist,
+                geneSel = function(genelist) {return(genelist == 1)},
+                nodeSize = 10,
+                annot = annFUN.gene2GO,
+                gene2GO = geneID2GO)
+  
+  # perform enrichment analysis using parentchild method
+  resultParentChild <- runTest(GOdata, statistic = 'fisher', algorithm = 'parentchild')
+  
+  # get number of significant GO terms p < 0.01
+  numsignodes <- length(which(score(resultParentChild) < 0.01))
+  
+  # result table
+  resTable <- GenTable(GOdata, parentchild = resultParentChild, topNodes = numsignodes)
+  
+ 
+  if(!is.na(plotpath)) {
+    # plot nodes
+    pdf(plotpath)
+    showSigOfNodes(GOdata, score(resultParentChild), firstSigNodes = min(numsignodes, 10), useInfo = 'all')
+    dev.off()
+  }
+  
+}

### 2014-04-09 ApTranscriptome_TR.Rmd
Working on manuscript.
Updating analysis script to perform gene set enrichment analysis for

1. Intersection of genes by expression type from each colony
2. Set difference of genes by expression type from each colony

@@
              length(Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]),
-             length(Ap.response.type[which(Ap.response.type$A22.type == "NotExp" & Ap.response.type$Ar.type == "NotExp"), "Transcript"]))
+             length(Ap.response.type[which(Ap.response.type$A22.type == "NotResp" & Ap.response.type$Ar.type == "NotResp"), "Transcript"]))
 
@@
 Ap.type.table <- rbind(A22.type.table, Ar.type.table, overlap)
-rownames(Ap.type.table) <- c("A22", "Ar", "Both")
+rownames(Ap.type.table) <- c("ApVT", "ApNC", "Overlap")
 
@@
 
-pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of thermally-responsive transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not expressed for each colony and their overlap.")
+pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
 
@@
 
-One of two tests, for *Ar* intermediate transcripts also finds greater constitutive expression in *Ar* than *A22*.
+The non-parametric test for both comparisions also finds greater constitutive expression in *Ar* than *A22*.
 
@@
 
+```{r echo=FALSE, results='hide'}
+# combine responsive.transcripts.ann with "type" information
+Ap.response.type.sub <- 
+responsive.transcripts.ann.type <- merge(responsive.transcripts.ann, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
+str(responsive.transcripts.ann.type)
+write.csv(responsive.transcripts.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
+```
+
 There are also 6 heat shock related genes in the responsive transcripts, out of 130 total.
@@
 Ap.BP.resultParentChild <- runTest(Ap.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
-Ap.BP.resultParentChild
+# Ap.BP.resultParentChild
 
@@
 Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 82)
-#Ap.BP.ResTable
-write.table(Ap.BP.ResTable, file = paste(resultsdir, "Ap_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
-pandoc.table(Ap.BP.ResTable)
+dim(Ap.BP.ResTable)
+#pandoc.table(Ap.BP.ResTable)
 
@@
 # Run GSEA
-Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO)
-pandoc.table(Ap.high.gsea)
+Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = "results/Ar_int_BP_topGO_sig_nodes.pdf")
+#pandoc.table(Ap.high.gsea)
 ```
@@
 Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
-pandoc.table(Ap.low.BP.ResTable)
+#pandoc.table(Ap.low.BP.ResTable)
 ```
@@
 Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
-pandoc.table(Ap.bim.gsea)
+#pandoc.table(Ap.bim.gsea)
 ```
@@
 Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
-pandoc.table(Ap.int.gsea)
+#pandoc.table(Ap.int.gsea)
+```
+
+```{r gsea_table, echo=FALSE, eval=TRUE, results='asis'}
+## merge results into single table
+Ap.high.gsea$Type <- "High"
+Ap.low.gsea$Type <- "Low"
+Ap.bim.gsea$Type <- "Bimodal"
+Ap.int.gsea$Type <- "Intermediate"
+# combine
+Ap.gsea <- rbind(Ap.high.gsea, Ap.low.gsea, Ap.bim.gsea, Ap.int.gsea)
+# reorder
+Ap.gsea1 <- Ap.gsea[,c("Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
+colnames(Ap.gsea1)[7] <- "P"
+# display
+pandoc.table(Ap.gsea1, split.tables = "Inf")
 ```
@@
 A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
-pandoc.table(A22.high.gsea)
-```
-
-**Ar 'High' genes not in Ar**
+#pandoc.table(A22.high.gsea)
+```
+
+**Ar 'High' genes not in A22**
 
@@
 Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
-pandoc.table(Ar.high.gsea)
 ```
@@
 A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
-pandoc.table(A22.low.gsea)
-```
-
-**Ar 'Low' genes not in Ar**
+```
+
+**Ar 'Low' genes not in A22**
 
@@
 Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
-pandoc.table(Ar.low.gsea)
 ```
@@
 A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
-pandoc.table(A22.bim.gsea)
-```
-
-**Ar 'Bimodal' genes not in Ar**
+```
+
+**Ar 'Bimodal' genes not in A22**
 
@@
 Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
-pandoc.table(Ar.bim.gsea)
 ```
@@
 A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
-pandoc.table(A22.int.gsea)
-```
-
-**Ar 'Low' genes not in Ar**
+```
+
+**Ar 'Intermediate' genes not in A22**
 
@@
 # Run GSEA
-Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO, plotpath = "results/Ar_int_BP_topGO_sig_nodes.pdf")
-pandoc.table(Ar.int.gsea)
+Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
+```
+
+Make a table with results.
+
+```{r gsea_table, echo=FALSE, eval=TRUE, results='asis'}
+A22.high.gsea$Type <- "High"
+A22.low.gsea$Type <- "Low"
+A22.int.gsea$Type <- "Intermediate"
+A22.bim.gsea$Type <- "Bimodal"
+A22.gsea <- rbind(A22.high.gsea, A22.low.gsea, A22.int.gsea, A22.bim.gsea)
+A22.gsea$Colony <- "ApVT"
+
+Ar.high.gsea$Type <- "High"
+Ar.low.gsea$Type <- "Low"
+Ar.bim.gsea$Type <- "Bimodal"
+Ar.int.gsea$Type <- "Intermediate"
+Ar.gsea <- rbind(Ar.high.gsea, Ar.low.gsea, Ar.int.gsea, Ar.bim.gsea)
+Ar.gsea$Colony <- "ApNC"
+
+# combine
+Ap.setdiff.gsea <- rbind(A22.gsea, Ar.gsea)
+# reorder
+Ap.setdiff.gsea <- Ap.setdiff.gsea[,c("Colony", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
+colnames(Ap.setdiff.gsea)[8] <- "P"
+
+Ap.setdiff.gsea
 ```
@@
 
+```{r rData, echo=FALSE}
+# save RData for manuscript file
+save(Ap.type.table, Ap.df, file = "ms.RData")
+```
+
 ## Session information ##

### 2014-04-09 scripts/RxNseq.R

@@ -59,11 +59,11 @@ RxNtype <- function(lmitem) {
     # calculate for A22
     A22.pout <- pout[pout$colony == "A22", ]
     # if pTPM doesn't vary, set max and min values to NA
-    if(sd(A22.pout$pTPM) < 0.01) {
+    if(sd(A22.pout$pTPM) < 0.00001) {
         A22.max.val <- NA
         A22.min.val <- NA
         A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
-        A22.exp.type <- "NotExp"
+        A22.exp.type <- "NotResp"
     } else {
         # else, set values from data
         A22.max.val <- median(A22.pout[which(A22.pout$pTPM == max(A22.pout$pTPM)), "val"])
@@ -86,11 +86,11 @@ RxNtype <- function(lmitem) {
     # calculate for Ar
     Ar.pout <- pout[pout$colony == "Ar", ]
     # if pTPM doesn't vary, set max and min values to NA
-    if(sd(Ar.pout$pTPM) < 0.01) {
+    if(sd(Ar.pout$pTPM) < 0.00001) {
         Ar.max.val <- NA
         Ar.min.val <- NA
         Ar.opt <- Ar.pout[Ar.pout$val == 19.5, "pTPM"]
-        Ar.exp.type <- "NotExp"
+        Ar.exp.type <- "NotResp"
     } else {
             # else, set values from data
             Ar.max.val <- median(Ar.pout[which(Ar.pout$pTPM == max(Ar.pout$pTPM)), "val"])
@@ -218,8 +218,7 @@ geneid2GOmap <- function(annotmat) {
   #
   # Returns:
   #  geneid2go.map saved in root directory
-  
-  
+    
   # requires
   require(stringr)
   
@@ -270,15 +269,14 @@ gsea <- function(genelist, geneID2GO, plotpath=NA) {
   # get number of significant GO terms p < 0.01
   numsignodes <- length(which(score(resultParentChild) < 0.01))
   
-  # result table
-  resTable <- GenTable(GOdata, parentchild = resultParentChild, topNodes = numsignodes)
-  
- 
+  # if 'plotpath' is set, make plot of top 10 significant nodes 
   if(!is.na(plotpath)) {
     # plot nodes
     pdf(plotpath)
     showSigOfNodes(GOdata, score(resultParentChild), firstSigNodes = min(numsignodes, 10), useInfo = 'all')
     dev.off()
-  }
[…]

### 2014-04-10 doc/ApTranscriptome_manuscript.Rmd
Added R code to manuscript to generate tables rather than copy and pasting. Changed file type to .Rmd

Thermal reactionome of a temperate ant species
============================================

**Authors**: John Stanton-Geddes, Andrew Nguyen, Rob Dunn, Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan

## Introduction

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. At the population level, thermal tolerance has been assessed in dozens of studies often demonstrating adaptive differentiation [@hoffmann2013]. In combination with environmental data, this thermal tolerance data has provided the backbone to dozens of studies examining limits to species geographic ranges [@angert2011; @diamond2013; @gouveia2014], and the impacts of climate change on population viability [@deutsch2008; @kingsolver2013]. At the organismal level, studies of thermal tolerance have largely focused on candidate genes such as heat shock proteins [@dahlgaard1998]. However, estimates of the standing genetic variation [@krebs1997; @krebs1998; @williams2012], studies of differential gene expression [@leemans2000; @sorensen2005; @teets2012], quantitative trait loci (QTL) [@morgan2006] and knock-outs [@takahashi2011] in the model system *Drosophila melanogaster* all implicate dozens of additional genes underlying thermal tolerance. Little is known about the diversity of these genes or their expression patterns, especially in non-model species.

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved sequencing technology (RNAseq). To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012, @pespeni2013, @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. The authors concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant differences in gene expression. To overcome this limitation, we use a novel method to characterize the reaction norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal stress. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach.

We apply this approach in a study of ants in the *Aphaenogaster rudis-picea* species complex [@umphrey1996]. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to litter processing and decomposition [@lyford1963; @folgarait1998, @terbough2001, @holec2005]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant colonies persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food [...]. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to deal tolerate thermal stress.

To explore the molecular toolkit used by *Aphaenogaster* to cope with thermal stress, we performed a whole organism RNA sequencing study using environmentally-divergent colonies from North Carolina and Vermont, USA. Our specific goals were to 1) generate the set of all genes (i.e. transcriptome) expressed by *Aphaenogaster* across physiologically-relevant temperatures, 2) identify genes that have responsive expression across this thermal gradient (i.e. thermal reactionome), 3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures, and 4) examine differences in the thermal reactionome between the two colonies.

## Methods

### Samples

Colonies of the *Aphaenogaster picea-rudis* complex were collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *ApNC* respectively. The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996]. Though preliminary work indicates that the northern part of the distribution (including *ApVT*) to be a distinct clade and possibly species (Bernice Bacon DeMarco, pers. communication), we will refer to these as distinct colonies rather than species. These colonies were maintained in the lab for 5 months prior to experimentation.

From each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath (...) at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* from VT (A. Nguyen, unpublished data) . At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. mRNA was extracted from three pooled ants using a DNeasy kit (Qiagen Inc; Valencia, CA). Ants were homogenized in Buffer ATL with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY) and the standard Qiagen protocol was followed from this point.

### Sequencing, assembly and annotation

For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the ApVT and ArNC samples respectively. Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].

We assembled the sequenced reads into the full set of mRNA transcripts (the `transcriptome`) for the combined data set from both colonies using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts with enough overlap to assemble into a false, or chimeric, third transcript) [@yang2013] To account for this, we first ran program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length, and second ran the program `uclust` which cluster sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental for exact parameters). We used liberal values (90% similarity) to merge transcripts to account for the fact that we were including sequencing data from two colonies of the same species, and wanted to avoid assembling orthologs individually. We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

We performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.

### Thermally-responsive genes

We quantified expression of each gene using the program `Sailfish` [@patro2013]. `Sailfish` reports gene expression as transcripts per million (TPM) which is the preferred measure of gene expression as values are comparable among sequencing runs [@wagner2012]. In addition, `Sailfish` corrects for sequence composition bias and gene length. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between colonies at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both colonies was justified (Supplemental).

To identify genes that have significant changes in expression across the thermal gradient, we fit the regression model

$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$

to each gene. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005]. This method is robust to overdispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.

We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after stepwide model selection by AIC. All analyses were performed in R [@RCoreTeam2014] and are fully reproducible (Supplemental file).

For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each colony. We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each colony; genes that had greatest expression at high (> 31°) temperatures (*High*), low (< 10°C) temperatures (*Low*), intermediate (10 - 30°C) temperatures (*Intermediate*), both high and low temperatures (*Bimodal*) or were not thermally-responsive in that colony (*NotResp*). For the *Bimodal* group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two colonies using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossman2007] implemented in R [@RCoreTeam2014]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.

### Colony-level comparisons

To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two colonies focusing on the thermally-responsive genes that had a significant colony x temperature interaction. For each gene, we predicted expression levels across temperatures for each colony using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Expressed*. To focus on differences in gene expression between the two colonies, we performed GSEA for the set difference of each category.

With the same set of genes, we explored the extent to which differences in the thermal reactionome between the colonies were due to changes in the mean or shape of the reaction norm for each individual gene. Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix). As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the colonies $\Delta M = (Mean_{ApVT} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{ApVT} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively. Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *ApVT* than *ApNC*.

Further, to quantify the relative contribution of changes in the mean, slope, curvature or wiggle on the overall differences in reaction norms, we defined the *Total* difference in the reaction norm for each gene as $\Delta T = \Delta M + \Delta S + \Delta C + \Delta W$. We then partitioned the contribution of each measure by dividing it by $\Delta T$.

For the subset of genes that responded to temperature in one colony (i.e. *High* expression in the *ApVT* colony)  but not the other, we hypothesized that geographically-divergent selection may have favored constitutive expression. To test this idea, we compared levels of constitutive expression between *A22* and *ApNC* near the optimum temperature of 19.5° C . Specifically, we predicted higher constitutive expression in *ApNC* for genes with *High* expression in *A22*, and conversely, higher constitutive expression  in *A22* for *Low* genes in *ApNC* .

The *Intermediate* expressed genes are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermal stress. We hypothesized that if the more southern *ApNC* colony was more thermally-tolerant than *ApVT* colony, genes with 'Intermediate' expression (10-30° C) would be expressed across a wider range of temperatures. To test this with our data, for each *Intermediate* expressed gene in each colony, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between colonies.

In contrast, *Bimodal* expressed genes are those that are activated in response to thermal stress. Following the reasoning for *Intermediate* genes, we hypothesized that a more thermally-sensitive colony would activate these thermally-responsive genes earlier than a more thermally-tolerant colony. We calculated the standard deviation of expression as for *Intermediate* genes. The interpretation is different, as low standard deviation is a gene with nearly flat expression, while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one colony had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.

**Data and analysis availability**

The raw Illumina reads are available at [...], the script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).

## Results

```{r load, echo=FALSE, results='hide'}
load("../ms.RData")
suppressWarnings(library(pander))
suppressWarnings(library(ggplot2))
```

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 96,253 contigs and a total of 105,536 genes. For all genes, the total length was 63 million bp with an N~50~ length of 895 bp and a mean contig size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. Of these, 5,787 genes have best hits to putative genes in the genome sequence of the ant species *Solenopsis invicta* and XXXX are annotated to genes previously identified from organisms within the class Insecta.

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
```

### Thermally-responsive genes

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transripts, 8,044 significantly change expression with temperature including 6,727 that have a colony x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the colonies ($\chi^2 = 170, P < 0.0001$; Table 1). For both colonies, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *ApNC* than in *ApVT*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

![Probability density function of peak expression for all thermally-responsive in *ApNC* (blue) and *ApVT* (red) colonies](../results/PDF_expression_all.png)

As the molecular pathways involved in thermal response to cold and hot temperatures may differ, we performed gene set enrichment analysis on each expression category separately. Out of a total of 2,859 GO terms scored, we found 6 enriched (*P* < 0.01) for *High* transcripts, included categories implicated in stress response such as "gene expression" and "tissue regeneration" (Table S1). For *Low* genes, 60 GO terms were enriched including some implicated in stress response (e.g. detection of external stimulus; Table S1). For *Bimodal* genes, 11 GO terms were enriched including 10 "regulation" terms (e.g. regulation of primary metabolic process) indicating that these are general genes involved in activating responses to change (Table S1). Finally, for *Intermediate* genes 6 GO terms were enriched, all some type of metabolic or biosynthetic process (Table S1) consistent with the expectation that these are non-essential genes expressed in optimal conditions that are shut-down under thermal stress.

In addition, we examined the reaction norms for genes that we *a priori* expected to have significant thermal responses. In particular, we looked at the expression patterns of heat shock proteins including ...

### Colony-level comparisons

We quantified differences in the reaction norms of thermally-responsive genes between colonies as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta W$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly signifcant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *ApVT* than *ApNC*, indicating a steeper response to temperature.

Of the total differences ($\Delta T$) in reaction norms between colonies for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.

Plasticity is a well-characterized mechanism by which organims tolerate environmental change. Given our sampling of a warmer (*ApNC*) and colder (*ApVT*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression of *High* genes from *ApVT* was greater in the more southern *ApNC* colony (Wilcoxon V = 911970, *P* < 0.001, 95% CI location shift -0.11 - 0.03). However, for genes with *Low* expression in *ApNC* we also found that constitutive expression was greater in *ApNC* than *ApVT* (Wilcoxon V = 2041149, *P* < 0.001, 95% CI location shift -0.23 - 0.15).

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the standard deviation of expression for the *Intermediate* expressed genes. Consistent with our expectations, the more southern *ApNC* colony had greater thermal breadth for *Intermediate* genes than the *ApVT* colony (*t*~df=1406~ = -9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the colonies response to thermal shock. For *High* genes, there were 4 enriched GO terms in *ApVT* and 12 in *ApNC*. For *Low* genes, there were 40 enriched GO terms in *ApVT* and 23 in *ApNC*. For *Intermediate* genes, there were 3 enriched GO terms in *ApVT* and 54 in *ApNC*. For *Bimodal* genes, there were 12 enriched GO terms in *ApVT* and 10 in *ApNC*.

## Discussion

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster rudis*. We found that nearly 10% of the genes in this species show significant thermal responsiveness. Moreover, we found ... differences in patterns of expression among these two species.

Two functional groups of *a priori* interest showed patterns consistent with our expectations. Responsive genes annotated to "heat shock proteins" were up-regulated at high temperatures (>31 C) in the northern *A. picea* colony, but not up-regulated in the putatively more heat tolerant *A. rudis* colony (Fig. 3a). Conversely, genes involved in the biological process of dormancy were up-regulated at cold temperatures (< 10 C) in the southern *A. rudis* colony, but not *A. picea* (Fig. 3b). As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

![Figure 3. goes here]

Responsiveness ... a temperate ant species are thermally responsive. The majority of these genes (about 90%) are up-regulated at high or low temperatures. Discuss the 26 "stress response" genes

Previous studies have using *Drosophila* [@sorensen2005] and a temperate flesh fly [@teets2012b] have implicated GO terms .... as responding to heat stress. In this study, we found .... of these GO terms to be enriched at high temperatures for *Aphaenogaster*. However, when we looked at the exact identify of the genes, we only found X overlapping genes.

This result poses a paradox. Why is the heat shock functional response highly conserved at the biological process level, while the exact molecular pathways appear to be highly labile? We suggest two possible explanations. First, regulatory networks controlling gene expression are evolutionarily dynamic. Trade-offs between function in different organisms may constrain which genes are involved in thermal response, thus resulting in different pathways to the same phenotype. Second, variation in molecular pathways may be due to inconsistent selection on thermal stress response. That is, as the environment changes and species shift their distributions, selection may relax or strengthen repeatedly within and across lineages. Different genes may be recruited within each selective episode. Differentiating between these, and alternate, explanations will require greater geographic and population genetic sampling.

A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*. Our comparative results are limited to two populations, and an important future step will be expanding this work to eaexamine differences in gene expression among a larger number of populations. Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance. The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of colonies, and inability of ants to mate in the lab. However, future work could leverage genomic information to infer the relatedness among colonies in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.

In summary, we have generated the first transcriptome of a temperate ant species, and have identified approximately 10% of genes that show thermal responsiveness. In this group, we find many known stress response genes, but also identify novel genetic pathways that respond to temperature. These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.

## References

### 2014-04-10 ApTranscriptome_TR.Rmd
Working on manuscript in Rmd format. Knit to md. Then compile to pdf using pandoc. Added citations to bib file.

@@
 
-**April 8, 2014**
+**April 10, 2014**
 
@@
 # Global settings
-options(stringsAsFactors=FALSE)
+options(stringsAsFactors=FALSE, digits = 3, scipen = 1)
 
@@
 # T-test comparing standard deviation of expression between colonies
-t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd)
+t1 <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd)
 ```
@@
 # t-test to compare standard deviation of 'Bimodal' transcripts between colonies
-t.test(Ar.bim.sd, A22.bim.sd)
+t2 <- t.test(Ar.bim.sd, A22.bim.sd)
 ```
@@
 # offset = mean1 - mean2
-varshape.out$offset <- varshape.out$A22.mean - varshape.out$Ar.mean
+varshape.out$offset
+<- varshape.out$A22.mean - varshape.out$Ar.mean
 # to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
@@
 # t-test
-t.test(varshape.out$s.offset, alternative = "two.sided")
+t3 <- t.test(varshape.out$s.offset, alternative = "two.sided")
 
@@
 # t-test
-t.test(varshape.out$s.slope, alternative = "two.sided")
+t4 <- t.test(varshape.out$s.slope, alternative = "two.sided")
 
@@
 # t-test
-t.test(varshape.out$s.curvature, alternative = "two.sided")
+tt.test(varshape.out$s.curvature, alternative = "two.sided")
 
@@
 
-## Gene set enrichment analysis ##
-
-I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.
-
-```{r GO_contingency_test}
-responsive.transcripts.ann <- signif.transcripts[names(responsive.transcripts)]
-# GO 'response to stress' hits in responsive transcripts
-responsive.transcripts.ann[grep("GO:0006950", responsive.transcripts.ann$GO.Biological.Process), list(Transcript, best.hit.to.nr)]
-
-resp.stress.responsive.count <- nrow(responsive.transcripts.ann[grep("GO:0006950", responsive.transcripts.ann$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
-# GO 'response to stress' hits in all transcripts
-resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
-
-GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts.ann) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
-
-GO.stress.Xsq <- chisq.test(GO.stress.table)
-GO.stress.Xsq
-```
-
 ```{r echo=FALSE, results='hide'}
+responsive.transcripts.ann.type <- signif.transcripts[names(responsive.transcripts)]
 # combine responsive.transcripts.ann with "type" information
-Ap.response.type.sub <- 
 responsive.transcripts.ann.type <- merge(responsive.transcripts.ann, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
@@
 
+## Gene set enrichment analysis ##
+
+I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.
+
+```{r GO_contingency_test}
+# GO 'response to stress' hits in responsive transcripts
+GO0006950.responsive <- responsive.transcripts.ann.type[grep("GO:0006950", responsive.transcripts.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
+
+# in high category
+GO0006950.responsive[union(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
+# in low category
+GO0006950.responsive[union(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
+# in bimodal category
+GO0006950.responsive[union(with(GO0006950.responsive, grep("Bimodal", Ar.type)), with(GO0006950.responsive, grep("Bimodal", A22.type))), ]
+# in intermediate category
+GO0006950.responsive[union(with(GO0006950.responsive, grep("Intermediate", Ar.type)), with(GO0006950.responsive, grep("Intermediate", A22.type))), ]
+
+# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.transcripts compared to full list
+resp.stress.responsive.count <- nrow(responsive.transcripts.ann.type[grep("GO:0006950", responsive.transcripts.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
+# GO 'response to stress' hits in all transcripts
+resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
+
+GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
+
+GO.stress.Xsq <- chisq.test(GO.stress.table)
+GO.stress.Xsq
+```
+
 There are also 6 heat shock related genes in the responsive transcripts, out of 130 total.
@@
 ```{r grep_shock}
-hsp_responsive <- responsive.transcripts.ann[grep("shock", responsive.transcripts.ann$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
-nrow(hsp_responsive)
+hsp_responsive <- responsive.transcripts.ann.type[grep("shock", responsive.transcripts.ann.type$best.hit.to.nr), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
+hsp_responsive
 

### 2014-04-11 ApTranscriptome_TR.Rmd
Updates to manuscript

@@
 
-interaction.transcripts <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
-other.transcripts <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.transcripts))]
-temperature.transcripts <- other.transcripts[which(Map(grepFunc, other.transcripts, term = "val") == TRUE)]
-colony.transcripts <- other.transcripts[setdiff(names(other.transcripts), names(temperature.transcripts))]
-responsive.transcripts <- c(temperature.transcripts, interaction.transcripts)
-rm(other.transcripts)
+interaction.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
+other.lms <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.lms))]
+temperature.lms <- other.lms[which(Map(grepFunc, other.lms, term = "val") == TRUE)]
+colony.lms <- other.lms[setdiff(names(other.lms), names(temperature.lms))]
+responsive.lms <- c(temperature.lms, interaction.lms)
+rm(other.lms)
 ```
@@
 siglist <- c(length(RxNlmAIC),
-             length(colony.transcripts),
-             length(temperature.transcripts),
-             length(interaction.transcripts)
+             length(colony.lms),
+             length(temperature.lms),
+             length(interaction.lms)
              )
@@
 # calculate response type for responsive transcripts
-interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
-stopifnot(nrow(interaction.response.type) == length(interaction.transcripts))
+interaction.response.type <- ldply(interaction.lms, .progress = "text", .inform = TRUE, RxNtype)
+stopifnot(nrow(interaction.response.type) == length(interaction.lms))
 
 # calculate response types for transcripts without interactions
-temperature.response.type <- ldply(temperature.transcripts, .progress = "text", .inform = TRUE, RxNtype)
-stopifnot(nrow(temperature.response.type) == length(temperature.transcripts))
+temperature.response.type <- ldply(temperature.lms, .progress = "text", .inform = TRUE, RxNtype)
+stopifnot(nrow(temperature.response.type) == length(temperature.lms))
 
@@
 Table 4 shows the number of transcripts that fall into each expression type for each each colony.
-The totals for each colony include the `r nrow(temperature.transcripts)` transcripts that have consistent temperature responses between the two colonies.
-
-## Shiny interactive web-app
-
-**2014-04-02 -- CURRENTLY BROKEN**
-
-To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package.
-The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.
-
-Export data for interactive shiny app.
-
-```{r shiny_file, eval=FALSE, echo=FALSE}
-shiny.transcripts <- TPM.dt.sub[Ap.response.type$Transcript]
-
-# scale expression values 
-shiny.transcripts[,pTPM.scaled:=scale(pTPM), by = Transcript]
-str(responsive.transcripts.TPM)
-write.csv(responsive.transcripts.TPM, file = "ApRxN-shinyapp/responsive.transcripts.TPM.csv", quote = TRUE, row.names = FALSE)
-```
+The totals for each colony include the `r nrow(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
 
@@
 png(paste(resultsdir, "PDF_expression_all.png", sep=""))
-g3 <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + geom_density(alpha=0.2, position="identity")
-g3 + scale_y_continuous(name="Density") +
+g3 <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
+  geom_density(alpha=0.2, position="identity") + 
+  #scale_fill_manual(name = "Colony", values = c("white", "black")) + 
+  scale_y_continuous(name="Density") +
   scale_x_continuous(name="Temperature of maximum expression")
+suppressWarnings(print(g3))
 dev.off()
@@
 ```{r offset, echo=TRUE, eval=TRUE}
-varshape.out <- ldply(interaction.transcripts, .progress = "text", RxNvarshape)
+varshape.out <- ldply(interaction.lms, .progress = "text", RxNvarshape)
 
@@
 ```{r echo=FALSE, results='hide'}
-responsive.transcripts.ann.type <- signif.transcripts[names(responsive.transcripts)]
-# combine responsive.transcripts.ann with "type" information
-responsive.transcripts.ann.type <- merge(responsive.transcripts.ann, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
-str(responsive.transcripts.ann.type)
-write.csv(responsive.transcripts.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
+responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
+# combine responsive.lms.ann with "type" information
+responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
+str(responsive.lms.ann.type)
+write.csv(responsive.lms.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
 ```
@@
 # GO 'response to stress' hits in responsive transcripts
-GO0006950.responsive <- responsive.transcripts.ann.type[grep("GO:0006950", responsive.transcripts.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
+GO0006950.responsive <- responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
 
@@
 
-# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.transcripts compared to full list
-resp.stress.responsive.count <- nrow(responsive.transcripts.ann.type[grep("GO:0006950", responsive.transcripts.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
+# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.lms compared to full list
+resp.stress.responsive.count <- nrow(responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
 # GO 'response to stress' hits in all transcripts
@@
 
-GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
+GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.lms.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
 
@@
 ```{r grep_shock}
-hsp_responsive <- responsive.transcripts.ann.type[grep("shock", responsive.transcripts.ann.type$best.hit.to.nr), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
+hsp_responsive <- responsive.lms.ann.type[grep("shock", responsive.lms.ann.type$best.hit.to.nr), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
 hsp_responsive
@@
 
-Visualize differens in GO terms among expression types using a wordcloud.
-
-```{gsea_wordcloud, cache=TRUE}
-# create Corpus for text mining
-GO.dfs <- c(paste(Ap.low.BP.ResTable$Term, collapse=" "), paste(Ap.hig.BP.ResTable$Term, collapse=" "),
-            paste(Ap.int.BP.ResTable$Term, collapse=" "), paste(Ap.bim.BP.ResTable$Term, collapse=" "))
-
-vs <- VectorSource(GO.dfs)
-GO.corp <- Corpus(vs)
-GO.corp
-
-GO.corp <- tm_map(GO.corp, removePunctuation)
-
-# create matrix of terms
-term.matrix <- TermDocumentMatrix(GO.corp)
-term.matrix <- as.matrix(term.matrix)
-colnames(term.matrix) <- c('low', 'high', 'intermediate', 'bimodal')
-
-# comparison cloud
-comparison.cloud(term.matrix, max.words=200, random.order=FALSE, scale=c(0.25, 0.5))
-```
-
-## Comparison of expression patterns between colonies
-
-A set of `r nrow(interaction.transcripts)` transcripts have expression patterns that depend on colony.
-In this section, I examine how these transcripts differ between colonies.
-Specifically, I expect genes that are up-regulated at high temperatures in the northern *A22* colony to be at constitutively higher expression in the more southern *Ar* colony.
-Conversely, I will examine the extent to which genes upregulated at colder temperatures in the *Ar* colony are constitutively expressed in the *A22* colony.
-
 ## Visualize responsive transcripts
 
-Make plots for all significant transcripts
-
-```{r plot_responsive, echo=FALSE, eval=FALSE, cache=TRUE}
+Organize data
+
+```{r plot_data}
+# extract TPM data for thermally-responsive transcripts
+resp.TPM.dt.sub <- TPM.dt.sub[names(responsive.lms)]
+setkey(resp.TPM.dt.sub, Transcript)
+str(resp.TPM.dt.sub)
+length(unique(resp.TPM.dt.sub$Transcript))
+# scale transcripts so can compare
+resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
+```
+
+Make plots for all genes expressed at *High* temps in GO category "GO:0006950: response to stress"
+
+```{r plot_GOstress, echo=FALSE, eval=FALSE, cache=TRUE}
+high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
+
 # Line plot, expression against temp, faceted by colony
-p1 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
+trp1 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
   geom_line() + 
@@
   scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p1
+print(trp1)
 
 # Smooth plot, expression against temp, faceted by colony
-p2 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
-  geom_smooth() +
+trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
+  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
   facet_grid(. ~ colony) + 
@@
   scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p2
-
-# Same as p1, faceted by expression type
-p3 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
-  geom_line() +
-  facet_grid(A22.exp.type ~ colony) +
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p3
-
-# Smooth plot, expression against temp, faceted by colony
-p4 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
-  geom_smooth() +
-  facet_grid(A22.exp.type ~ colony) +
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p4
-```
-
-Make plots for transcripts with significant temperature by colony interaction
-
-```{r plot_interaction_responsive, echo=FALSE, eval=FALSE, cache=TRUE}
+print(trp2)
+```
+
+Make plots for all genes expressed at *Low* temps in GO category "GO:0006950: response to stress"
+
+```{r plot_GOstress, echo=FALSE, eval=FALSE, cache=TRUE}
+low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
+
 # Line plot, expression against temp, faceted by colony
-p1 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
+trp3 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
   geom_line() + 
@@
   scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p1
+print(trp3)
 
 # Smooth plot, expression against temp, faceted by colony
-p2 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
-  geom_smooth() +
+trp4 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
+  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
   facet_grid(. ~ colony) + 
@@
   scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p2
-
-# Same as p1, faceted by expression type
-p3 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
-  geom_line() +
-  facet_grid(A22.exp.type ~ colony) +
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p3
-
-# Smooth plot, expression against temp, faceted by colony
-p4 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
-  geom_smooth() +
-  facet_grid(A22.exp.type ~ colony) +
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-p4
-```
-
-```{r plot, echo=FALSE, eval=FALSE, cache=TRUE}
-# Line plot, expression against temp, faceted by colony
-png(paste(resultsdir, "Ap_expression_by_colony_line.png", sep=""))
-  p1 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_line() +
-    facet_grid(. ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p1
-dev.off()
-
-# Smooth plot, expression against temp, faceted by colony
-png(paste(resultsdir, "Ap_expression_by_colony_smooth.png", sep=""))
-  p2 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_smooth() +
-    facet_grid(. ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p2
-dev.off()
-
-# Same as p1, faceted by expression type
-png(paste(resultsdir, "Ap_expression_by_colony_exp_line.png", sep=""))
-  p3 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_line() +
-    facet_grid(A22.exp.type ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p3
-dev.off()
-
-# Smooth plot, expression against temp, faceted by colony
-png(paste(resultsdir, "Ap_expression_by_colony_exp_smooth.png", sep=""))
-  p4 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_smooth() +
-    facet_grid(A22.exp.type ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p4
-dev.off()
-
-```
-
-```{r plot_interaction_responsive_to_file, echo=FALSE, eval=FALSE, cache=TRUE}
-# Line plot, expression against temp, faceted by colony
-png(paste(resultsdir, "Ap_expression_interaction_by_colony_line.png", sep=""))
-  p1 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_line() +
-    facet_grid(. ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p1
-dev.off()
-
-# Smooth plot, expression against temp, faceted by colony
-png(paste(resultsdir, "Ap_expression_interaction_by_colony_smooth.png", sep=""))
-  p2 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_smooth() +
-    facet_grid(. ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p2
-dev.off()
-
-# Same as p1, faceted by expression type
-png(paste(resultsdir, "Ap_expression_interaction_by_colony_exp_line.png", sep=""))
-  p3 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_line() +
-    facet_grid(A22.exp.type ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p3
-dev.off()
-
-# Smooth plot, expression against temp, faceted by colony
-png(paste(resultsdir, "Ap_expression_interaction_by_colony_exp_smooth.png", sep=""))
-  p4 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
-    geom_smooth() +
-    facet_grid(A22.exp.type ~ colony) +
-    scale_y_continuous(name="Expression (scaled)") +
-    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-  p4
-dev.off()
+print(trp4)
+```
+
+## Shiny interactive web-app
+
+To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package.
+The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.
+
+Export data for interactive shiny app.
+
+```{r shiny_file, eval=FALSE, echo=FALSE}
+shiny.dt <- TPM.dt.sub[Ap.response.type$Transcript]
+
+# scale expression values 
+shiny.dt[,TPM.scaled:=scale(TPM), by = Transcript]
+str(shiny.dt)
+write.csv(shiny.dt, file = "ApRxN-shinyapp/responsive.lms.TPM.csv", quote = TRUE, row.names = FALSE)
 ```

### 2014-04-11 doc/ApTranscriptome_manuscript.Rmd
Updates to manuscript

@@
 transcriptomics) has rapidly increased due to improved sequencing technology (RNAseq).
-To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012, @pespeni2013, @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
+To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
 However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature.
@@
 The authors concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments.
-Similarly, gene expression studies with few environments are likely to miss many biologically-relevant differences in gene expression.
+Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression.
 To overcome this limitation, we use a novel method to characterize the reaction norm of all genes in the transcriptome (e.g.
@@
 
-We apply this approach in a study of ants in the *Aphaenogaster rudis-picea* species complex [@umphrey1996].
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to litter processing and decomposition [@lyford1963; @folgarait1998, @terbough2001, @holec2005].
+We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster rudis* and *A.
+picea* [@umphrey1996], which have more southern and northern distributions, respectively.
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
 As ectothermic species, ants experience a wide range of temperature conditions.
-Throughout the year, ant colonies persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food [...].
-Thus, at the organismal level we expect that ants utilize multiple molecular pathways to deal tolerate thermal stress.
+Throughout the year, ant colonies persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food.
+Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.
 
-To explore the molecular toolkit used by *Aphaenogaster* to cope with thermal stress, we performed a whole organism RNA sequencing study using environmentally-divergent colonies from North Carolina and Vermont, USA.
+To explore the molecular toolkit used by *Aphaenogaster* to cope with thermal stress, we performed a whole organism RNA sequencing study.
 Our specific goals were to 1) generate the set of all genes (i.e.
 transcriptome) expressed by *Aphaenogaster* across physiologically-relevant temperatures, 2) identify genes that have responsive expression across this thermal gradient (i.e.
-thermal reactionome), 3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures, and 4) examine differences in the thermal reactionome between the two colonies.
+thermal reactionome), 3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures.
+We expected to observe candidate genes in the *response to stress* Gene Ontology (GO) category including *heat shock proteins* and ...
+in the thermally-responsive group.
+
+While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991].
+Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance.
+To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*.
+*A.
+picea*, the northern species, experiences less chronic and acute heat stress than *A.
+rudis*.
+Thus, we that *A.
+picea* would have more genes than *A.
+rudis* that are up-regulated to high temperatures, while *A.
+rudis* would have higher constitutive expression of these thermally-tolerant genes.
+Reciprocally, we expected *A.
+rudis* to have greater up-regulation of genes in response to cold temperatures, while *A.
+picea* would have greater constitutive expression of cold-tolerant genes.
+Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures.
+From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A.
+picea* would have a broader thermal breadth of gene expression than *A.
+rudis*.
 
@@
 We used $log(TPM + 1)$ as the response to control for skew in the expression data.
-For a continous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005].
-This method is robust to overdispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
+For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005].
+This method is robust to over-dispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
 To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.
 
-We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after stepwide model selection by AIC.
-All analyses were performed in R [@RCoreTeam2014] and are fully reproducible (Supplemental file).
+We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
 
@@
 
-To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two colonies using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossman2007] implemented in R [@RCoreTeam2014].
+To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two colonies using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
 Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
@@
 ```{r load, echo=FALSE, results='hide'}
+options(scipen = 1, digits = 3, warn = -1)
 load("../ms.RData")
-suppressWarnings(library(pander))
-suppressWarnings(library(ggplot2))
+suppressMessages(library(pander))
+suppressMessages(library(ggplot2))
 ```
@@
 pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
+library(xtable)
+xtable(Ap.type.table)
 ```
@@
 We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR.
-Of these transripts, 8,044 significantly change expression with temperature including 6,727 that have a colony x temperature interaction.
+Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a colony x temperature interaction.
 The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the colonies ($\chi^2 = 170, P < 0.0001$; Table 1).
@@
 We quantified differences in the reaction norms of thermally-responsive genes between colonies as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta W$) and wiggle ($\Delta W$).
-Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly signifcant.
+Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant.
 On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *ApVT* than *ApNC*, indicating a steeper response to temperature.
@@
 
-Plasticity is a well-characterized mechanism by which organims tolerate environmental change.
+Plasticity is a well-characterized mechanism by which organism tolerate environmental change.
 Given our sampling of a warmer (*ApNC*) and colder (*ApVT*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa.
@@
 
-Previous studies have using *Drosophila* [@sorensen2005] and a temperate flesh fly [@teets2012b] have implicated GO terms ....
+Previous studies have using *Drosophila* [@sorensen2005] and a temperate flesh fly [@teets2012-1] have implicated GO terms ....
 as responding to heat stress.
@@
 A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*.
-Our comparative results are limited to two populations, and an important future step will be expanding this work to eaexamine differences in gene expression among a larger number of populations.
+Our comparative results are limited to two populations, and an important future step will be expanding this work to examine differences in gene expression among a larger number of populations.
 Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance.

### 2014-04-11 ApTranscriptome_TR.Rmd
Version 0.2 of manuscript finished and sent to group

@@
 ```{r A22_high_wilcoxon}
-wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+(w1 <- wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE))
 ```
@@
 # Wilcoxon signed rank-test
-wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+(w2 <- wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE))
 ```
@@
 # T-test comparing standard deviation of expression between colonies
-t1 <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd)
+(t1 <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
 ```
@@
 ```{r annotation}
-# add annotation information
 setkey(annotationtable, Sequence.Name)
@@
 Ap.BP.resultParentChild <- runTest(Ap.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
-# Ap.BP.resultParentChild
+Ap.BP.resultParentChild
 
 # table results
-Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 82)
+Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 113)
 dim(Ap.BP.ResTable)
@@
 #pandoc.table(Ap.high.gsea)
+# output for Revigo
+write.table(Ap.high.gsea[,"GO.ID"], file = "results/Ap_high_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
 ```
@@
 #pandoc.table(Ap.low.BP.ResTable)
+write.table(Ap.low.gsea[,"GO.ID"], file = "results/Ap_low_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
 ```
@@
 #pandoc.table(Ap.bim.gsea)
+write.table(Ap.bim.gsea[,"GO.ID"], file = "results/Ap_bim_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
 ```
@@
 #pandoc.table(Ap.int.gsea)
+write.table(Ap.int.gsea[,"GO.ID"], file = "results/Ap_int_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
 ```
@@
 # Smooth plot, expression against temp, faceted by colony
+png("results/GO_stress_High.png")
 trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
@@
 print(trp2)
+dev.off()
 ```
@@
 # save RData for manuscript file
-save(Ap.type.table, Ap.df, file = "ms.RData")
+save(Ap.type.table, Ap.df, w1, w2, file = "ms.RData")
 ```

### 2014-04-11 doc/ApTranscriptome_manuscript.Rmd
Version 0.2 of manuscript finished and sent to group

@@
 As ectothermic species, ants experience a wide range of temperature conditions.
-Throughout the year, ant colonies persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food.
+Throughout the year, ant species persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food.
 Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.
@@
 
-Colonies of the *Aphaenogaster picea-rudis* complex were collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *ApNC* respectively.
-The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996].
-Though preliminary work indicates that the northern part of the distribution (including *ApVT*) to be a distinct clade and possibly species (Bernice Bacon DeMarco, pers.
-communication), we will refer to these as distinct colonies rather than species.
-These colonies were maintained in the lab for 5 months prior to experimentation.
+*Aphaenogaster picea* was collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and *A.
+rudis* was collected from Durham, North Carolina ( 36.037° N -78.874° W), designated as *A.
+picea* and *A.
+rudis* henceforth.
+Though the phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], preliminary work supports that these species are separate species (Bernice Bacon DeMarco, pers.
+communication).
+species of these species were maintained in the lab for 5 months prior to experimentation.
 
-From each species, we haphazardly collected 12 ants at the same time on 12 days.
+From a single species of each species, we haphazardly collected 12 ants at the same time on 12 days.
 Each day, the 12 ants were placed in glass tubes into a water bath (...) at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour.
-The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* from VT (A.
-Nguyen, unpublished data) .
+The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* [@warren2013.
 At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
@@
 
-For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the ApVT and ArNC samples respectively.
+For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the A.
+picea and A.
+rudis samples respectively.
 Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
 
-We assembled the sequenced reads into the full set of mRNA transcripts (the `transcriptome`) for the combined data set from both colonies using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
+We assembled the sequenced reads into the full set of mRNA transcripts (the `transcriptome`) for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
 *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g.
 regions of two separate transcripts with enough overlap to assemble into a false, or chimeric, third transcript) [@yang2013] To account for this, we first ran program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length, and second ran the program `uclust` which cluster sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental for exact parameters).
-We used liberal values (90% similarity) to merge transcripts to account for the fact that we were including sequencing data from two colonies of the same species, and wanted to avoid assembling orthologs individually.
+We used liberal values (90% similarity) to merge transcripts to account for the fact that we were including sequencing data from two species of the same species, and wanted to avoid assembling orthologs individually.
 We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
@@
 As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis.
-The expression values were highly correlated between colonies at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both colonies was justified (Supplemental).
+The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).
 
@@
 
-For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each colony.
-We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each colony; genes that had greatest expression at high (> 31°) temperatures (*High*), low (< 10°C) temperatures (*Low*), intermediate (10 - 30°C) temperatures (*Intermediate*), both high and low temperatures (*Bimodal*) or were not thermally-responsive in that colony (*NotResp*).
+For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (*High*), low (< 10°C) temperatures (*Low*), intermediate (10 - 30°C) temperatures (*Intermediate*), both high and low temperatures (*Bimodal*) or were not thermally-responsive in that species (*NotResp*).
 For the *Bimodal* group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
 
-To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two colonies using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
+To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
 Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
+To reduce redundancy in the enriched GO terms, we used the web program [ReviGO](http://revigo.irb.hr/).
 
-### Colony-level comparisons
+### Species-level comparisons
 
-To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two colonies focusing on the thermally-responsive genes that had a significant colony x temperature interaction.
-For each gene, we predicted expression levels across temperatures for each colony using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Expressed*.
-To focus on differences in gene expression between the two colonies, we performed GSEA for the set difference of each category.
+To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction.
+For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Expressed*.
+To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.
 
-With the same set of genes, we explored the extent to which differences in the thermal reactionome between the colonies were due to changes in the mean or shape of the reaction norm for each individual gene.
+With the same set of genes, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene.
 Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix).
-As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the colonies $\Delta M = (Mean_{ApVT} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{ApVT} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively.
-Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *ApVT* than *ApNC*.
+As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A.
+picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A.
+picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively.
+Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A.
+picea* than *A.
+rudis*.
 
@@
 
-For the subset of genes that responded to temperature in one colony (i.e.
-*High* expression in the *ApVT* colony)  but not the other, we hypothesized that geographically-divergent selection may have favored constitutive expression.
-To test this idea, we compared levels of constitutive expression between *A22* and *ApNC* near the optimum temperature of 19.5° C .
-Specifically, we predicted higher constitutive expression in *ApNC* for genes with *High* expression in *A22*, and conversely, higher constitutive expression  in *A22* for *Low* genes in *ApNC* .
+To test if genes that were responsive to cold temperatures in the southern *A.
+rudis* species or hot temperatures in the northern *A.
+picea* species had higher constitutive expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.
 
-The *Intermediate* expressed genes are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermal stress.
-We hypothesized that if the more southern *ApNC* colony was more thermally-tolerant than *ApVT* colony, genes with 'Intermediate' expression (10-30° C) would be expressed across a wider range of temperatures.
-To test this with our data, for each *Intermediate* expressed gene in each colony, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
+As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A.
+picea* species would have a broader thermal range of stable gene expression than *A.
+rudis*.
+To test this with our data, for each *Intermediate* expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
 point of maximum expression had highest probability of being sampled).
-We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between colonies.
+We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species.
+Conversely, we predicted that the thermal sensitivity of *Bimodal* expressed genes would be greater in *A.
+rudis* than *A.
+picea*.
+We performed the same analysis as with the *Intermediate* transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern.
+Thus, finding one species had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.
 
-In contrast, *Bimodal* expressed genes are those that are activated in response to thermal stress.
-Following the reasoning for *Intermediate* genes, we hypothesized that a more thermally-sensitive colony would activate these thermally-responsive genes earlier than a more thermally-tolerant colony.
-We calculated the standard deviation of expression as for *Intermediate* genes.
-The interpretation is different, as low standard deviation is a gene with nearly flat expression, while the maximum standard deviation of expression is for a highly-convex expression pattern.
-Thus, finding one colony had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.
-
-**Data and analysis availability**
+### Data availability
 
@@
 
-```{r expression_type_table, echo=FALSE, results='asis'}
-pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
-library(xtable)
-xtable(Ap.type.table)
-```
-
 ### Thermally-responsive genes
@@
 We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR.
-Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a colony x temperature interaction.
-The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the colonies ($\chi^2 = 170, P < 0.0001$; Table 1).
-For both colonies, the majority of responsive transcripts were in the *Low* category (Fig.
+Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a species x temperature interaction.
+The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1).
+For both species, the majority of responsive transcripts were in the *Low* category (Fig.
 1).
-More transcripts had *Intermediate* expression in *ApNC* than in *ApVT*, which had nearly twice as many bimodally-expressed transcripts  (Fig.
+More transcripts had *Intermediate* expression in *A.
+rudis* than in *A.
+picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig.
 1, Table 1).
 
-![Probability density function of peak expression for all thermally-responsive in *ApNC* (blue) and *ApVT* (red) colonies](../results/PDF_expression_all.png)
+```{r expression_type_table, echo=FALSE, results='asis'}
+pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
+```
 
@@
 
-In addition, we examined the reaction norms for genes that we *a priori* expected to have significant thermal responses.
-In particular, we looked at the expression patterns of heat shock proteins including ...
+![Probability density function of peak expression for all thermally-responsive in *A. rudis* (blue) and *A. picea* (red) species](../results/PDF_expression_all.png)
 
-### Colony-level comparisons
+As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
 
-We quantified differences in the reaction norms of thermally-responsive genes between colonies as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta W$) and wiggle ($\Delta W$).
+In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses....
+
+### Species-level comparisons
+
+We quantified differences in the reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$).
 Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant.
-On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *ApVT* than *ApNC*, indicating a steeper response to temperature.
+On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A.
+picea* than *A.
+rudis*, indicating a steeper response to temperature.
 
-Of the total differences ($\Delta T$) in reaction norms between colonies for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.
+Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.
 
 Plasticity is a well-characterized mechanism by which organism tolerate environmental change.
-Given our sampling of a warmer (*ApNC*) and colder (*ApVT*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa.
-Consistent with one of our predictions, we found that constitutive expression of *High* genes from *ApVT* was greater in the more southern *ApNC* colony (Wilcoxon V = 911970, *P* < 0.001, 95% CI location shift -0.11 - 0.03).
-However, for genes with *Low* expression in *ApNC* we also found that constitutive expression was greater in *ApNC* than *ApVT* (Wilcoxon V = 2041149, *P* < 0.001, 95% CI location shift -0.23 - 0.15).
+Given our sampling of a warmer (*A.
+rudis*) and colder (*A.
+picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa.
+Consistent with one of our predictions, we found that constitutive expression was greater in the more southern *A.
+rudis* species for *High* genes in *A.
+picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`).
+However, for genes with *Low* expression in *A.
+rudis* we also found that constitutive expression was greater in *A.
+rudis* than *A.
+picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).
 
-We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the standard deviation of expression for the *Intermediate* expressed genes.
-Consistent with our expectations, the more southern *ApNC* colony had greater thermal breadth for *Intermediate* genes than the *ApVT* colony (*t*~df=1406~ = -9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
+We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range) for the *Intermediate* expressed genes.
+Contrary to our expectations based on critical maximum and minimal temperatures, *A.
+rudis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
 In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
 
-We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the colonies response to thermal shock.
-For *High* genes, there were 4 enriched GO terms in *ApVT* and 12 in *ApNC*.
-For *Low* genes, there were 40 enriched GO terms in *ApVT* and 23 in *ApNC*.
-For *Intermediate* genes, there were 3 enriched GO terms in *ApVT* and 54 in *ApNC*.
-For *Bimodal* genes, there were 12 enriched GO terms in *ApVT* and 10 in *ApNC*.
+We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock.
+For *High* genes, there were 4 enriched GO terms in *A.
+picea* and 12 in *A.
+rudis*.
+For *Low* genes, there were 40 enriched GO terms in *A.
+picea* and 23 in *A.
+rudis*.
+For *Intermediate* genes, there were 3 enriched GO terms in *A.
+picea* and 54 in *A.
+rudis*.
+For *Bimodal* genes, there were 12 enriched GO terms in *A.
+picea* and 10 in *A.
+rudis*.
+
+![Expression (scaled to allow comparisons among genes) for genes in GO category "response to stress"](../results/GO_stress_High.png)
 
@@
 We found that nearly 10% of the genes in this species show significant thermal responsiveness.
-Moreover, we found ...
-differences in patterns of expression among these two species.
 
-Two functional groups of *a priori* interest showed patterns consistent with our expectations.
-Responsive genes annotated to "heat shock proteins" were up-regulated at high temperatures (>31 C) in the northern *A.
-picea* colony, but not up-regulated in the putatively more heat tolerant *A.
-rudis* colony (Fig.
-3a).
-Conversely, genes involved in the biological process of dormancy were up-regulated at cold temperatures (< 10 C) in the southern *A.
-rudis* colony, but not *A.
-picea* (Fig.
-3b).
-As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
+### Molecular toolkit of thermal responsiveness
 
-![Figure 3. goes here]
+While the sheer number of thermally-responsive transcripts makes generalizations difficult, groupings such as Gene Ontology (GO) can be used to identify categories associated with thermal stress.
+For heat response, the semantic space of the enriched GO terms included *gene expression*, *regeneration* and *cellular component organization or biogenesis*.
+...previous work...
+For cold response, there were 65 enriched GO terms that grouped by semantic space into *protein modification process*, *sensory perception of mechanical stimulus*, *response to mechanical stimulus*, *folic acid metabolism*, *membrane lipid metabolism* and *protein complex biogenesis*.
+...previous work...
 
-Responsiveness ... a temperate ant species are thermally responsive. The majority of these genes (about 90%) are up-regulated at high or low temperatures. Discuss the 26 "stress response" genes
+*Bimodal* genes are general stress response genes that are activated under both heat and cold stress.
+The semantic space of the enriched GO terms included *regulation of metabolism*, *biological regulation*, *macromolecule biosynthesis*, ...
 
-Previous studies have using *Drosophila* [@sorensen2005] and a temperate flesh fly [@teets2012-1] have implicated GO terms ....
-as responding to heat stress.
-In this study, we found ....
-of these GO terms to be enriched at high temperatures for *Aphaenogaster*.
-However, when we looked at the exact identify of the genes, we only found X overlapping genes.
+*Intermediate* genes represent non-essential molecular functions that are shut-down as conditions become stressful.
+The semantic space for the enriched GO terms included *lipid metabolic process*, *glycoprotein biosynthetic process*, *membrane lipid metabolic process*, *cellular protein metabolic process*, *membrane lipid biosynthetic process*.
+...indicates shutting down of non essential molecule biosynthesis...
+
+**Sara - comparisons of our gene list to other studies**
 
@@
 
+### Transcriptome-wide comparisions
+
+Of the 8,492 genes that were thermally-responsive, 80% differed in their responsiveness by colony.
+The majority of this difference in the reactionomes was due to changes in the slope ($\Delta S$) and curvature ($\Delta C$) which together accounted for nearly 90% of the differences between reaction norms for all transcripts.
+In a meta-analysis of phenotypic reaction norms, Murren et al.
+[-@murren2014] similarly found that differences in the shape of the reaction norms accounted for the majority of the differences.
+Thus, our finding suggests that this phenotypic pattern may be explained by greater genetic lability in the shape than the mean of gene expression.
+The relative role of plasticity versus adaptation in response to rapid environmental change is widely debated [@vellend...?].
+Our data suggest that plasticity through changes in the responsiveness of gene expression are likely to occur more rapidly than changes in mean values.
+
+An important note is that we would have completely missed this pattern if we had performed a differential expression study using only a few temperatures.
+Only by measuring gene expression across a thermal gradient and fitting a function to each transcript were we able to characterize the shape of the reaction norm of expression...
+To our knowledge, this is the first time that a reaction norm approach has been applied to the study of global gene expression.
+
+Given the large differences in number (Table 1) and responsiveness of genes between the two species, a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures.
+Intriguingly, this
+result mirrors the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
+A plausible explanation is that organisms have fewer molecular mechanisms to tolerate heat than cold stress.
+Future studies ...
+implications for climate change as the evolutionary potential of ecototherms to climate change may be limited as suggested by ...
+[kingsolver?]
+
+### Conclusions
+
 A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*.
@@
 Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance.
-The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of colonies, and inability of ants to mate in the lab.
-However, future work could leverage genomic information to infer the relatedness among colonies in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.
+The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of species, and inability of ants to mate in the lab.
+However, future work could leverage genomic information to infer the relatedness among species in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.
 

### 2014-04-25 ApTranscriptome_TR.Rmd
Minor updates to TR and manuscript

@@
 library(topGO) 
-suppressMessages(library(Rgraphviz))
+#suppressMessages(library(Rgraphviz))
 
@@
 
-```{r gsea_table, echo=FALSE, eval=TRUE, results='asis'}
+```{r gsea_table_2, echo=FALSE, eval=TRUE, results='asis'}
 A22.high.gsea$Type <- "High"
@@
 high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
+
+# in low category
+low.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
+# in bimodal category
+bim.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Bimodal", Ar.type)), with(GO0006950.responsive, grep("Bimodal", A22.type))), ]
+# in intermediate category
+int.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Intermediate", Ar.type)), with(GO0006950.responsive, grep("Intermediate", A22.type))), ]
 
@@
 
-```{r plot_GOstress, echo=FALSE, eval=FALSE, cache=TRUE}
+```{r plot_GOstress_low, echo=FALSE, eval=FALSE, cache=TRUE}
 low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]

### 2014-04-25 doc/ApTranscriptome_manuscript.Rmd
Minor updates to TR and manuscript

@@
 ```{r expression_type_table, echo=FALSE, results='asis'}
-pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
+library(vcd)
+
+pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
+
+mosaicplot(tt2, shade = TRUE, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts.
+For reference, the *High* - *High* cell contains 309 transcripts, while the *Low* - *Low* cell contains 2,417 transcripts")
 ```

### 2014-04-25 ApTranscriptome_TR.Rmd
Minor changes

@@
 
+To remove contigs that are likely contaminants from bacterial, archael, virus or human sources, we used the program [DeconSeq](http://deconseq.sourceforge.net/) `r citep("10.1371/journal.pone.0017288")`.
+We downloaded the bacteria, virus, archae and human [databases of contaminants](ftp://edwards.sdsu.edu:7009/deconseq/db), modified the `DeconSeqConfig.pm` file as described [here](http://www.vcru.wisc.edu/simonlab/bioinformatics/programs/install/deconseq.htm) to point to the databases, and ran DeconSeq specifiying 95% identity over 50% the length of contig
+
+    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust -dbs hsref,bast,vir,arch
+
+This resulted in removing 5,675 contigs as contaminants, leaving 99,861 "clean" contigs.
+We spot-checked the contaminants by BLAST and confirmed that they matched bacteria, human or viral sources by greater than 95%.
+
 ```{r assemstats, results = "asis", echo=FALSE}
@@
 
-Running Trinity is time and computationally-intensive, so the final filtered assembly can be downloaded from [http://johnstantongeddes.org/assets/files/Trinity_filtered_assembly.tgz]
+Running Trinity and subsequent programs is time and memory-intensive, so the final filtered clean assembly can be downloaded from [...]
 
@@
 # download filtered Trinity assembly, uncompress and move
-wget http://johnstantongeddes.org/assets/files/Trinity_filtered_assembly.tgz
-tar -zxvf trimmomatic_output.tar.gz
+wget http://johnstantongeddes.org/assets/files/Aphaenogaster_transcriptome.tar
+tar -xvf Aphaenogaster_transcriptome.tar
+# check md5sum
+md5sum Trinity_cap3_uclust_clean.fa
+# c604159ee944dddda7b4e2a556bf2f53
 mkdir -p results/
 mkdir -p results/trinity-full/
-mv Trinity_cap3_uclust.fasta /results/trinity-full/.
+mv Trinity_cap3_uclust_clean.fasta results/trinity-full/.
 ~~~
@@
 
-#Results are available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).
+Results are available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).
 
@@
 
-Transcriptome annotations are nearly impossible to visualize in a meaningful way. For lack of better ideas, I made a word cloud:
-
-```{r wordcloud, cache=TRUE, eval=FALSE}
-library(tm)
-library(wordcloud)
-library(RColorBrewer)
-library(stringr)
-
-# subset and extract unique best hits to NCBI nr database
-ann <- unique(annotationfile$best.hit.to.nr)
-annsplit <- str_split_fixed(ann, pattern = " ", n=2)
-annvec <- annsplit[,2]
-
-# create Corpus of terms. convert to dataframe for plotting wordcloud
-ap.corpus <- Corpus(VectorSource(annvec))
-ap.corpus <- tm_map(ap.corpus, removePunctuation)
-ap.corpus <- tm_map(ap.corpus, tolower)
-ap.corpus <- tm_map(ap.corpus, function(x) removeWords(x, stopwords("english")))
-tdm <- TermDocumentMatrix(ap.corpus)
-m <- as.matrix(tdm)
-v <- sort(rowSums(m),decreasing=TRUE)
-d <- data.frame(word = names(v),freq=v)
-pal <- brewer.pal(4, "Dark2")
-pal <- pal[-(1:2)]
-
-#png(paste(resultsdir, "wordcloud.png", sep=""), width=1280,height=800)
-wordcloud(d$word,d$freq, scale=c(8,.3),min.freq=2,max.words=100, random.order=T, rot.per=.15, colors=pal, vfont=c("sans serif","plain"))
-#dev.off()
-```
-
-This wordcloud shows that we do not know much about the transcriptome - mostly hypothetical or predicted proteins. I removed these uninformative words and generated a new wordcloud.
-
-```{r wordcloud2, cache=TRUE}
-# remove 'protein', 'predicted' and 'hypothetical'
-rmwords <- c("hypothetical", "protein", "isoform", "subunit", "family", "putative", "partial", "domain", "containing", "predicted", "domaincontaining", "uncharacterized")
-ap.corpus.rmwords <- tm_map(ap.corpus, removeWords, rmwords)
-tdm <- TermDocumentMatrix(ap.corpus.rmwords)
-m <- as.matrix(tdm)
-v <- sort(rowSums(m),decreasing=TRUE)
-d <- data.frame(word = names(v),freq=v)
-pal <- brewer.pal(8, "Dark2")
-pal <- pal[-(1:2)]
-#png(paste(resultsdir, "wordcloud2.png", sep=""), width=1280,height=800)
-wordcloud(d$word,d$freq, scale=c(8,.3),min.freq=2,max.words=100, random.order=T, rot.per=.15, colors=pal, vfont=c("sans serif","plain"))
-#dev.off()
-```
+Note that I used the reduced assembly, prior to cleaning out contaminants. Rather than re-running this, I remove contaminants by ...
 
@@
 
-    sailfish index -t Trinity_cap3_uclust.fasta -o sailfish-index -k 20 -p 4
+    sailfish index -t Trinity_cap3_uclust_clean.fasta -o sailfish-index -k 20 -p 4
 
@@
 ```{r temperature_response, eval=TRUE, echo=FALSE, results='asis'}
-# table type of response
 A22.type.table <- table(Ap.response.type[ , 'A22.type'])
 Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])
-# overlap
-overlap <- c(length(Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]),
-             length(Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type == "High"), "Transcript"]),
-             length(Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]),
-             length(Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]),
-             length(Ap.response.type[which(Ap.response.type$A22.type == "NotResp" & Ap.response.type$Ar.type == "NotResp"), "Transcript"]))
-
-# make table
-Ap.type.table <- rbind(A22.type.table, Ar.type.table, overlap)
-rownames(Ap.type.table) <- c("ApVT", "ApNC", "Overlap")
-
-# re-orderfor printing
-Ap.type.table <- Ap.type.table[ , c(2,4,1,3,5)]
-
-pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
+
+type.table <- table(Ap.response.type[ , 'A22.type'], Ap.response.type[ , 'Ar.type'])
+# reorder for plotting
+tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]
+
+png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""))
+mosaicplot(tt2, shade = TRUE, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts")
+dev.off()
+
+pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
 
@@
 # save RData for manuscript file
-save(Ap.type.table, Ap.df, w1, w2, file = "ms.RData")
+save(tt2, Ap.df, w1, w2, file = "ms.RData")
 ```

### 2014-04-29 ApTranscriptome_TR.Rmd
Updates to script - cleaning up file for knitting

@@
 
-**April 10, 2014**
+**April 29, 2014**
 
@@
 
-To remove contigs that are likely contaminants from bacterial, archael, virus or human sources, we used the program [DeconSeq](http://deconseq.sourceforge.net/) `r citep("10.1371/journal.pone.0017288")`.
-We downloaded the bacteria, virus, archae and human [databases of contaminants](ftp://edwards.sdsu.edu:7009/deconseq/db), modified the `DeconSeqConfig.pm` file as described [here](http://www.vcru.wisc.edu/simonlab/bioinformatics/programs/install/deconseq.htm) to point to the databases, and ran DeconSeq specifiying 95% identity over 50% the length of contig
-
-    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust -dbs hsref,bast,vir,arch
-
-This resulted in removing 5,675 contigs as contaminants, leaving 99,861 "clean" contigs.
-We spot-checked the contaminants by BLAST and confirmed that they matched bacteria, human or viral sources by greater than 95%.
-
 ```{r assemstats, results = "asis", echo=FALSE}
@@
 ```
+
+To remove contigs that are likely contaminants from bacterial, archael, virus or human sources, we used the program [DeconSeq](http://deconseq.sourceforge.net/) `r citep("10.1371/journal.pone.0017288")`.
+We downloaded the bacteria, virus, archae and human [databases of contaminants](ftp://edwards.sdsu.edu:7009/deconseq/db), modified the `DeconSeqConfig.pm` file as described [here](http://www.vcru.wisc.edu/simonlab/bioinformatics/programs/install/deconseq.htm) to point to the databases, and ran DeconSeq specifiying 95% identity over 50% the length of contig
+
+    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust -dbs hsref,bast,vir,arch
+
+This resulted in removing 5,675 contigs as contaminants, leaving 99,861 "clean" contigs.
+We spot-checked the contaminants by BLAST and confirmed that they matched bacteria, human or viral sources by greater than 95%.
 
@@
 
-```{r download, echo = TRUE, results = "hide"}
+```{r download, echo = TRUE}
 ### Annotation file
+
 # from either AWS or GoogleDrive
-annotationURL <- getURL("http://johnstantongeddes.org/assets/files/Aphaeno_transcriptome_AnnotationTable.txt")
+annotationURL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
 #a2 <- getURL("https://googledrive.com/host/0B75IymziRJ_9Tlg1U1Vxbjk1bzg") # GoogleDrive link
@@
 str(Ap.response.type)
+
+# save results to file
+write.table(file = paste(resultsdir, "Ap_responsive_transcripts.txt", sep = ""), Ap.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
+```
+
 ```
@@
 
-pdf(paste(resultsdir, "Ap.BP_topGO_sig_nodes.pdf", sep=""))
-showSigOfNodes(Ap.BP.GOdata, score(Ap.BP.resultParentChild), firstSigNodes = 10, useInfo = 'all')
-dev.off()
+#pdf(paste(resultsdir, "Ap.BP_topGO_sig_nodes.pdf", sep=""))
+#showSigOfNodes(Ap.BP.GOdata, score(Ap.BP.resultParentChild), firstSigNodes = 10, useInfo = 'all')
+#dev.off()
 ```
@@
 # Run GSEA
-Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = "results/Ar_int_BP_topGO_sig_nodes.pdf")
+Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)
 #pandoc.table(Ap.high.gsea)

### 2014-04-29 doc/ApTranscriptome_manuscript.Rmd
Updated gitignore. Working on manuscript

--- 2014-04-11 doc/ApTranscriptome_manuscript.Rmd
@@
 
-We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster rudis* and *A.
+We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
 picea* [@umphrey1996], which have more southern and northern distributions, respectively.
@@
 picea*, the northern species, experiences less chronic and acute heat stress than *A.
-rudis*.
-Thus, we that *A.
+carolensis*.
+Thus, we predicted that *A.
 picea* would have more genes than *A.
-rudis* that are up-regulated to high temperatures, while *A.
-rudis* would have higher constitutive expression of these thermally-tolerant genes.
+carolinensis* that are up-regulated to high temperatures, while *A.
+carolinensis* would have higher constitutive expression of these thermally-tolerant genes.
 Reciprocally, we expected *A.
-rudis* to have greater up-regulation of genes in response to cold temperatures, while *A.
+carolinensis* to have greater up-regulation of genes in response to cold temperatures, while *A.
 picea* would have greater constitutive expression of cold-tolerant genes.
@@
 picea* would have a broader thermal breadth of gene expression than *A.
-rudis*.
+carolinensis*.
 
@@
 *Aphaenogaster picea* was collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and *A.
-rudis* was collected from Durham, North Carolina ( 36.037° N -78.874° W), designated as *A.
+carolinensis* was collected from Durham, North Carolina ( 36.037° N -78.874° W), designated as *A.
 picea* and *A.
-rudis* henceforth.
+carolinensis* henceforth.
 Though the phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], preliminary work supports that these species are separate species (Bernice Bacon DeMarco, pers.
@@
 At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
-mRNA was extracted from three pooled ants using a DNeasy kit (Qiagen Inc; Valencia, CA).
+mRNA was extracted from three pooled ants using a RNeasy kit (Qiagen Inc; Valencia, CA).
 Ants were homogenized in Buffer ATL with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY) and the standard Qiagen protocol was followed from this point.
@@
 picea and A.
-rudis samples respectively.
+carolinensis samples respectively.
 Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
@@
 picea* than *A.
-rudis*.
+carolinensis*.
 
@@
 To test if genes that were responsive to cold temperatures in the southern *A.
-rudis* species or hot temperatures in the northern *A.
+carolinensis* species or hot temperatures in the northern *A.
 picea* species had higher constitutive expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.
@@
 picea* species would have a broader thermal range of stable gene expression than *A.
-rudis*.
+carolinensis*.
 To test this with our data, for each *Intermediate* expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
@@
 Conversely, we predicted that the thermal sensitivity of *Bimodal* expressed genes would be greater in *A.
-rudis* than *A.
+carolinensis* than *A.
 picea*.
@@
 More transcripts had *Intermediate* expression in *A.
-rudis* than in *A.
+carolinensis* than in *A.
 picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig.
@@
 ```{r expression_type_table, echo=FALSE, results='asis'}
-pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
+pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
+
+mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
 ```
@@
 
-![Probability density function of peak expression for all thermally-responsive in *A. rudis* (blue) and *A. picea* (red) species](../results/PDF_expression_all.png)
+![Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. picea* (red) species](../results/PDF_expression_all.png)
 
@@
 picea* than *A.
-rudis*, indicating a steeper response to temperature.
+carolinensis*, indicating a steeper response to temperature.
 
@@
 Given our sampling of a warmer (*A.
-rudis*) and colder (*A.
+carolinensis*) and colder (*A.
 picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa.
 Consistent with one of our predictions, we found that constitutive expression was greater in the more southern *A.
-rudis* species for *High* genes in *A.
+carolinensis* species for *High* genes in *A.
 picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`).
 However, for genes with *Low* expression in *A.
-rudis* we also found that constitutive expression was greater in *A.
-rudis* than *A.
+carolinensis* we also found that constitutive expression was greater in *A.
+carolinensis* than *A.
 picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).
@@
 Contrary to our expectations based on critical maximum and minimal temperatures, *A.
-rudis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
+carolinensis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
 In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
@@
 picea* and 12 in *A.
-rudis*.
+carolinensis*.
 For *Low* genes, there were 40 enriched GO terms in *A.
 picea* and 23 in *A.
-rudis*.
+carolinensis*.
 For *Intermediate* genes, there were 3 enriched GO terms in *A.
 picea* and 54 in *A.
-rudis*.
+carolinensis*.
 For *Bimodal* genes, there were 12 enriched GO terms in *A.
 picea* and 10 in *A.
-rudis*.
+carolinensis*.
 
@@
 heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness.
-In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster rudis*.
+In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*.
 We found that nearly 10% of the genes in this species show significant thermal responsiveness.

### 2014-04-29 ApTranscriptome_TR.Rmd
Minor changes

@@
 
-Note that I used the reduced assembly, prior to cleaning out contaminants. Rather than re-running this, I remove contaminants by ...
+Note that I used the reduced assembly, prior to cleaning out contaminants.
+As I use the "clean" assembly for read mapping and identification of responsive genes, the contaminants are dis-regarded in downstream analyses.
 
@@
 
-```{r fdr}
+```{r padjust}
 RxNpval$padj <- p.adjust(RxNpval$pval, method = "fdr")
@@
 
-```
-
 Examine transcripts by temperature response. Compare between colonies.
@@
 # offset = mean1 - mean2
-varshape.out$offset
-<- varshape.out$A22.mean - varshape.out$Ar.mean
+varshape.out$offset <- varshape.out$A22.mean - varshape.out$Ar.mean
 # to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript

### 2014-04-30 ApTranscriptome_TR.Rmd
Updated script to use DeconSeq "cleaned" transcriptome assembly that removed about 5k bacterial, archael, viral and human contigs.
Surprisingly, after re-running Sailfish, this increased the number of responsive
transcripts, after adjusting P-values, to 30k!!!

@@
 #biocLite("topGO")
-#biocLite("Rgraphviz")
+#biocLite("Rgraphviz") # INSTALL BROKEN IN R 3.1.0
 library(topGO) 
@@
 Two ant colonies were used for the transcriptome sequencing.
-The first, designated A22, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen.
-This colony was putatively identifed as *A.
-picea*.
-The second colony, designated Ar, was collected by Lauren Nichols in Raleigh, North Carolina.
-These colonies were maintained in the lab for 6 months prior to sample collection.
+The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen.
+The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina.
+These colonies were identified as *A.
+picea* and *A.
+carolinensis* by Bernice Bacon de Marco.
+The colonies were maintained in the lab for 6 months prior to sample collection.
 
@@
 The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen).
-Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20e6 and 16e6 reads for the A22 and Ar samples, respectively.
+Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20^6^ and 16^6^ reads for the *A22* and *Ar* samples, respectively.
 
@@
 # extract significant transcripts
+setkey(TPM.dt.sub, Transcript)
 sig.TPM.dt.sub <- TPM.dt.sub[signif.transcripts$Transcript]
@@
 
-png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""))
-mosaicplot(tt2, shade = TRUE, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts")
+png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""), width = 600, height = 480)
+mosaicplot(tt2, xlab = "ApVT", ylab = "AcNC", cex = 0.7, las = 1, main = "")
 dev.off()
@@
 
-Yes - transcripts that are up-regulated at high temperatures in *A22* are more highly-expressed at 19.25C in *Ar*.
+Yes - transcripts that are up-regulated at high temperatures in *A22* are more highly-expressed at 19.25°C in *Ar*.
 Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.
@@
 
+The overall pattern that emerges from these 4 tests is that constitutive expression at optimal conditions of 19.5°C is overall greater for *Ar* than *A22*.
+
 ### Thermal tolerance indicated by region of constant gene expression ###
@@
 The 'Intermediate' expressed transcripts are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermall stress.
-We hypothesized that if the more southern *Ar* colony was more thermally-tolerant than *A22*, transcripts with 'Intermediate' expression (10-30C) would be active across a wider range of temperatures.
+We hypothesized that if the more southern *Ar* colony was more thermally-tolerant than *A22*, transcripts with 'Intermediate' expression (10-30°C) would be active across a wider range of temperatures.
 To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each colony.
@@
 
+```{r plot_thermal_breadth2, echo = FALSE, eval = TRUE}
+# predict expression for these transcripts
+int.predicted <- matrix(nrow = 0, ncol = 5)
+
+# for A22
+colony = "A22"
+
+for(i in 1:length(A22.int.lm)) {
+    lmitem <- A22.int.lm[[i]]
+    pred.vals <- seq(from = 0, to = 38.5, by = 2.5)
+    newdf <- data.frame(Transcript = names(A22.int.lm[i]), colony = rep(colony, length = length(pred.vals)), val = pred.vals)
+    pout <- predict(lmitem, newdata=newdf)
+    pout <- cbind(newdf, logTPM = pout) # note this is on log scale
+    # back-transform from log-scale
+    pout$pTPM <- exp(pout$logTPM)
+    # change any negative values to zero
+    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
+    # concatenate
+    int.predicted <- rbind(int.predicted, pout)
+}
+
+# for Ar
+colony = "Ar"
+
+for(i in 1:length(Ar.int.lm)) {
+    lmitem <- Ar.int.lm[[i]]
+    pred.vals <- seq(from = 0, to = 38.5, by = 2.5)
+    newdf <- data.frame(Transcript = names(Ar.int.lm[i]), colony = rep(colony, length = length(pred.vals)), val = pred.vals)
+    pout <- predict(lmitem, newdata=newdf)
+    pout <- cbind(newdf, logTPM = pout) # note this is on log scale
+    # back-transform from log-scale
+    pout$pTPM <- exp(pout$logTPM)
+    # change any negative values to zero
+    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
+    # concatenate
+    int.predicted <- rbind(int.predicted, pout)
+}
+
+# change to data.table
+int.predicted <- data.table(int.predicted)
+# scale transcripts so can compare
+int.predicted[,pTPM.scaled:=scale(pTPM), by = Transcript]
+str(int.predicted)
+
+# plot
+intplot <- ggplot(int.predicted, aes(x=val, y=pTPM.scaled, group=Transcript)) + 
+  geom_line() + 
+  facet_grid(. ~ colony)
+intplot
+```
+
+Plot shows all the intermediate transcripts, though not possible to tell that *Ar* transcripts have greater variance than *A22* transcripts.
+
 ### Thermal sensitivity indicated by response of bimodally-expressed transcripts ###
@@
 # t-test to compare standard deviation of 'Bimodal' transcripts between colonies
-t2 <- t.test(Ar.bim.sd, A22.bim.sd)
+(t2 <- t.test(Ar.bim.sd, A22.bim.sd))
 ```
@@
 # t-test
-t3 <- t.test(varshape.out$s.offset, alternative = "two.sided")
+(t3 <- t.test(varshape.out$s.offset, alternative = "two.sided"))
 
@@
 # t-test
-t4 <- t.test(varshape.out$s.slope, alternative = "two.sided")
+(t4 <- t.test(varshape.out$s.slope, alternative = "two.sided"))
 
@@
 # t-test
-tt.test(varshape.out$s.curvature, alternative = "two.sided")
+(t5 <- t.test(varshape.out$s.curvature, alternative = "two.sided"))
 
@@
 # t-test
-t.test(varshape.out$s.wiggle, alternative = "two.sided")
+(t6 <- t.test(varshape.out$s.wiggle, alternative = "two.sided"))
 ```

### 2014-04-30 ApTranscriptome_TR.Rmd
Working through revisions to analysis using 'cleaned' transcriptome

@@
 
-Of the `r nrow(RxNpval)` transcripts, `r length(which(RxNpval$pval < 0.05))` have models with P < 0.05.
+Of the `r nrow(RxNpval)` transcripts, `r length(which(RxNpval$pval < 0.05))` have models with P < 0.05!
 
@@
 hist(RxNpval$padj)
-
+```
+
+Wow - even after controlling for FDR `r nrow(signif.transcripts)` transcipts are significant!
+
+Next, I use the draconianally-stringent Bonferonni correction.
+
+```{r bonferroni}
+RxNpval$pbonf <- p.adjust(RxNpval$pval, method = "bonferroni")
+```
+
+Wow - `r length(which(RxNpval$pbonf < 0.05))` transcripts significant after Bonferroni correction!
+
+```{r select_bonferroni}
 # subset to significant transcripts
-signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]
+signif.transcripts <- RxNpval[which(RxNpval$pbonf < 0.05), ]
 
@@
 ```
-
-At the 5% FDR significance threshold, there are `r nrow(signif.transcripts)` transcripts with an overall significant model.
 
@@
 
-Table 4 shows the number of transcripts that fall into each expression type for each each colony.
-The totals for each colony include the `r nrow(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
-
 ## Colony-level comparison ##
@@
 I evaluate this hypothesis by comparing expression levels at the at optimum (19.5C) temperature between the two colonies for genes in that are either in the 'High' or 'Low' expression group in the other colony.
-Specifically, I expect that genes upregulated at high temperatures in *A22* are more highly expressed at 19.5C in *Ar*?
-Conversely, I expect genes that are upregulated at low temps in *Ar* to have greater constitutive expression in *A22*?
+Specifically, are genes upregulated at high temperatures in *A22* more highly expressed near optimum temperatures of 19.5°C in *Ar*?
 
@@
 
-Yes - transcripts that are up-regulated at high temperatures in *A22* are more highly-expressed at 19.25°C in *Ar*.
-Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.
-
-Next I test the converse, that transcripts that are up-regulated at low temperatures in *Ar* are more highly-expressed at 19.25C temperatures in *A22*.
+Contrary to expectations, genes upregulated at high temperatures in *A22* also have greater expression at 19.5°C than *Ar*.
+
+Next I test the converse, that transcripts that are up-regulated at low temperatures in *Ar* are more highly-expressed at 19.25°C temperatures in *A22*.
 
@@
 
-Counter to expectations, expression at 19.5C is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*.
+Consistent with expectations, and the above test, expression at 19.5°C is also greater in *A22* than *Ar* for transcripts upregulated at low temperatures in *Ar*.
 
@@
 t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
+boxplot(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
 # Wilcoxon signed rank-test
@@
 t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
+boxplot(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
 # Wilcoxon signed rank-test
@@
 
-The non-parametric test for both comparisions also finds greater constitutive expression in *Ar* than *A22*.
-
-The overall pattern that emerges from these 4 tests is that constitutive expression at optimal conditions of 19.5°C is overall greater for *Ar* than *A22*.
+The non-parametric test for both comparisions also finds greater constitutive expression in *A22* than *Ar*.
+
+The overall pattern that emerges from these 4 tests is that constitutive expression at optimal conditions of 19.5°C is overall greater for *A22* than *Ar*.
 
@@
 Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
+length(Ar.int.lm)
 Ar.int.sd <- unlist(Map(transcriptSD, Ar.int.lm, colony = "Ar"))
@@
 ```
-
-Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
 

### 2014-05-01 ApTranscriptome_TR.Rmd
Updated script to use full Trinity assembly including contaminant reads. Confirmed previous Sailfish mapping results.

--- 2014-04-29 ApTranscriptome_TR.Rmd
@@
 
-    sailfish index -t Trinity_cap3_uclust_clean.fasta -o sailfish-index -k 20 -p 4
+    sailfish index -t Trinity_cap3_uclust_clean.fasta -o sailfish-index-Trinity_cap3_uclust -k 20 -p 4
 
@@
 
-    mkdir sailfish-expression
+    mkdir sailfish-expression-Trinity-cap3-uclust
 
@@
 # sailfish index directory
-sfindex <- "results/trinity-full/sailfish-index"
+sfindex <- "results/trinity-full/sailfish-index-Trinity_cap3_uclust"
 # sailfish expression directory
-sfexpressionroot <- "results/trinity-full/sailfish-expression/"
+sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
 
@@
 # Plot FDR values against initial pvalues
+png(file=paste(resultsdir, "pval_hist.png"))
 par(mfrow = c(2,1))
@@
 hist(RxNpval$padj)
+dev.off()
 

### 2014-05-07 ApTranscriptome_TR.Rmd
After evaluating the effect of removing contaminants prior to quantifying gene expresssion, determined best course of action was to quantify expression for all genes and then remove the contaminants. This strategy ensures that reads associated with contaminants are removed during the quantification step.
Updated script to follow this quantification strategy. In process, updated to use Sailfish v0.6.3
which required some updates to the "sailfish quant" step.

@@
 
-**April 29, 2014**
+**May 6, 2014**
 
@@
 library(topGO) 
-#suppressMessages(library(Rgraphviz))
+library(Rgraphviz)
 
@@
 md5sum Trinity_cap3_uclust_clean.fa
-# c604159ee944dddda7b4e2a556bf2f53
+# 0160a82ec62c6f371c84e4963fed09d1
 mkdir -p results/
 mkdir -p results/trinity-full/
+mv Trinity_cap3_uclust.fasta results/trinity-full/.
 mv Trinity_cap3_uclust_clean.fasta results/trinity-full/.
@@
 I quantified gene expression using [sailfish](http://www.cs.cmu.edu/~ckingsf/software/sailfish/index.html).
-To run this program, first make sure that PATHs to the software libraries are set up correctly:
-
-    export LD_LIBRARY_PATH=/opt/software/Sailfish-0.6.2-Linux_x86-64/lib:$LD_LIBRARY_PATH
-    export PATH=/opt/software/Sailfish-0.6.2-Linux_x86-64/bin:$PATH
+To run this program, first make sure that PATHs to the software libraries are set up correctly as described on the sailfish website.
 
@@
 
-    sailfish index -t Trinity_cap3_uclust_clean.fasta -o sailfish-index-Trinity_cap3_uclust -k 20 -p 4
+```{r sailfish_index}
+system("sailfish index -t results/trinity-full/Trinity_cap3_uclust.fasta -o results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4")
+```
 
@@
 
-    mkdir sailfish-expression-Trinity-cap3-uclust
-
-Then, for each sample, run the following command:
-
-    sailfish -i sailfish-index -o sailfish-expression/A22-0 --reads A22-0_ATCACG.paired.left.fastq A22-0_ATCACG.paired.right.fastq A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq -p 4
+```{r}
+system("mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust")
+```
+
+Then, for each sample, run the following command in the `results/trinity-full/` directory:
+
+    sailfish -i sailfish-index-Trinity-cap3-uclust -o sailfish-expression-Trinity-cap3-uclust/A22-0 -l "T=PE:O=><:S=U" -r A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq -1 A22-0_ATCACG.paired.left.fastq -2 A22-0_ATCACG.paired.right.fastq  -p 4
 
@@
 # sailfish index directory
-sfindex <- "results/trinity-full/sailfish-index-Trinity_cap3_uclust"
+sfindex <- "results/trinity-full/sailfish-index-Trinity-cap3-uclust"
 # sailfish expression directory
@@
     samp.unpaired.r <- paste(readdir, unpaired.right[samp.pos], sep="")
-    sailfishcmd <- paste("sailfish quant -i ", sfindex, " -o ", quantdir, " --reads ", samp.paired.l, " ", samp.paired.r, " ", samp.unpaired.l, " ", samp.unpaired.r, " -p 4", sep="")
-#   print(sailfishcmd)
+    sailfishcmd <- paste("sailfish quant -i ", sfindex, " -o ", quantdir, " -l 'T=SE:S=U' -r ", samp.paired.l, " ", samp.paired.r, " ", samp.unpaired.l, " ", samp.unpaired.r, " -p 4", sep="")
+    #print(sailfishcmd)
     system(sailfishcmd)
@@
 # read in each file using loop
-samples <- c("A22-0", "A22-3", "A22-7", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "Ar-0", "Ar-3", "Ar-7", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38")
-
 for (j in 1:length(samples)) {
@@
 
-```{r exp_correlations, echo=FALSE, results='asis'}
-# Make Rmarkdown pandoc table of correlations among expression levels between colonies at each temp
-
+```{r exp_correlations, echo=FALSE, results='hide'}
 temp <- c(0, 3.5, 7, 10.5, 14, 17.5, 21, 24.5, 28, 31.5, 35, 38.5)
@@
 cortable <- cbind(temp, cors)
-
+```
+
+```{r exp_correlations_table, echo=FALSE, results='asis'}
 pandoc.table(cortable, style="rmarkdown", caption = "correlations between colonies at each temperature treatment")
@@
 
-## Identification of thermally-responsive genes
+### Regression-model to identify thermally-responsive genes
 

### 2014-05-07 ApTranscriptome_TR.Rmd
Updated script to remove contaminated transcripts after expression quantification

@@
 We spot-checked the contaminants by BLAST and confirmed that they matched bacteria, human or viral sources by greater than 95%.
-
-Running Trinity and subsequent programs is time and memory-intensive, so the final filtered clean assembly can be downloaded from [...]
+For expression quantification, we use the full assembly to ensure that "contaminant" reads are assigned to the contaminants.
+After quantification, these transcripts will then be removed from further analyses.
+
+Running Trinity and subsequent programs is time and memory-intensive so the final assembly can be downloaded and used for all further analyses.
+In addition, the compressed archive contains the "clean" and "contaminant" sequences after filtering with DeconSeq.
 
@@
 wget http://johnstantongeddes.org/assets/files/Aphaenogaster_transcriptome.tar
+# check md5sum
+md5sum Aphaenogaster_transcriptome.tar
+# fd6dbb0b3e88e1c797c9e74611b245b2
+
+# uncompress and move
 tar -xvf Aphaenogaster_transcriptome.tar
-# check md5sum
-md5sum Trinity_cap3_uclust_clean.fa
-# 0160a82ec62c6f371c84e4963fed09d1
 mkdir -p results/
@@
 
-### Regression-model to identify thermally-responsive genes
-
-To identify transcripts (roughly equivalent to genes) that show thermal responsiveness, I fit the following linear model to each transcript:
-
-$$ log(TPM + 1) = \beta_0 + \beta_1(colony) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(colony * temp) + \beta_5(colony * temp^2) + \epsilon $$
-
-where TPM is transcripts per million.
-
-**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the analysis.**
-
-```{r RxN, echo=TRUE, eval=TRUE, cache=TRUE}
+**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.**
+
+```{r combined_TPM, echo=TRUE, eval=TRUE, cache=TRUE}
 A22.TPM[,colony:="A22"]
@@
 unique(TPM.dt.sub$val)
-```
+length(unique(TPM.dt.sub$Transcript))
+```
+
+### Remove *contaminant* transcripts
+
+```{r remove_contaminants}
+# extract transcript names from fasta file rather than loading whole file
+system("grep '^>' results/trinity-full/Trinity_cap3_uclust_cont.fa | cut -f 1 -d ' ' | sed -r 's/^.{1}//' > results/trinity-full/cont.list")
+
+# read file
+cont.list <- read.table("results/trinity-full/cont.list")
+str(cont.list)
+
+# remove from TPM.dt.sub
+setkey(TPM.dt.sub, Transcript)
+TPM.dt.sub <- TPM.dt.sub[!cont.list]
+str(TPM.dt.sub)
+length(unique(TPM.dt.sub$Transcript))
+```
+
+### Regression-model to identify thermally-responsive genes
+
+To identify transcripts (roughly equivalent to genes) that show thermal responsiveness, I fit the following linear model to each transcript:
+
+$$ log(TPM + 1) = \beta_0 + \beta_1(colony) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(colony * temp) + \beta_5(colony * temp^2) + \epsilon $$
+
+where TPM is transcripts per million.
 
@@
 # extract significant transcripts
-setkey(TPM.dt.sub, Transcript)
 sig.TPM.dt.sub <- TPM.dt.sub[signif.transcripts$Transcript]

### 2014-05-07 ApTranscriptome_TR.Rmd
More updates to analysis

@@
 
-## Thermally-responsive transcripts
+### Grouping of thermally-responsive transcripts
 
@@
 png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""))
-mosaicplot(tt2, shade = TRUE, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts")
+mosaicplot(tt2, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts")
 dev.off()
@@
 Table 4 shows the number of transcripts that fall into each expression type for each each colony.
-The totals for each colony include the `r nrow(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
+The totals for each colony include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
 
@@
 Specifically, I expect that genes upregulated at high temperatures in *A22* are more highly expressed at 19.5C in *Ar*?
-Conversely, I expect genes that are upregulated at low temps in *Ar* to have greater constitutive expression in *A22*?
 
@@
 
-Yes - transcripts that are up-regulated at high temperatures in *A22* are more highly-expressed at 19.25C in *Ar*.
+Consistent with expectation, there is greater expression at 19.5C for *Ar* than *A22* transcripts for the set of transcripts that are transcripts that are up-regulated at high temperatures in *A22*.
 Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.
@@
 # t-test
-tt.test(varshape.out$s.curvature, alternative = "two.sided")
+(t.test(varshape.out$s.curvature, alternative = "two.sided"))
 
@@
 # t-test
-t.test(varshape.out$s.wiggle, alternative = "two.sided")
+(t5 <- t.test(varshape.out$s.wiggle, alternative = "two.sided"))
 ```
@@
 
-t.test(varshape.out$s.total, alternative = "two.sided")
+(t6 <- t.test(varshape.out$s.total, alternative = "two.sided"))
 
@@
 
-## Functional annotation ##
+## Functional annotation
 
@@
 
-There are also 6 heat shock related genes in the responsive transcripts, out of 130 total.
+There are also 7 heat shock related genes in the responsive transcripts, out of 130 total.
 
@@
 # table results
-Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 113)
+Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 137)
 dim(Ap.BP.ResTable)

### 2014-05-08 ApTranscriptome_TR.Rmd
Adding analysis of temperature expression turned on/off.
Some updates to knit nicely.

@@
-Thermal reactionome of the common ant species *Aphaenogaster*
-================================================================
+Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
+========================================================================================
 
@@
 library(MASS)
+library(xtable)
+library(methods)
 
@@
 opts_chunk$set(cache=TRUE)
+options(width = 80)
 
@@
 
-```{r}
+```{r mk_sailfish_dir}
 system("mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust")
@@
 
-```{r sailfish, eval=TRUE, echo=TRUE, cache=TRUE}
+```{r sailfish_quant, eval=TRUE, echo=TRUE, cache=TRUE}
 # directory containing trimmed reads
@@
 
-# define function for `ddply` to calculate overall model P value
-modpFunc <- function(foo) {
-  # fit lm
-  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))
-
-  # calculate overall model significance
-  f <- summary(lmout)$fstatistic
-  pval <- unname(pf(f[1],f[2],f[3],lower.tail=F))
-  attributes(pval) <- NULL
-
-  # return pvalue
-  c(pval = pval) 
-}
-
-RxNpval <- ddply(TPM.dt.sub, .(Transcript), .progress="text", .inform="TRUE", modpFunc)
+# calculate overall P value for 'model' for each transcript
+RxNpval <- ddply(TPM.dt.sub, .(Transcript), .inform="TRUE", modpFunc)
 ```
@@
 # Plot FDR values against initial pvalues
-png(file=paste(resultsdir, "pval_hist.png"))
 par(mfrow = c(2,1))
@@
 hist(RxNpval$padj)
-dev.off()
 
@@
 
-```{r lmFunc}
-# first, define function to use
-lmFunc <- function(bar) {
-  # fit lm
-  lmout <- eval(parse(text = paste("lm(", model, ", data = bar)", sep = "")))  
-  # stepAIC to drop non-significant terms
-  f.lmout <- stepAIC(lmout)
-  # return final model
-  return(f.lmout) 
-}
-
-# apply function to all responsive transcripts
-RxNlmAIC <- dlply(sig.TPM.dt.sub, .(Transcript), .progress="text", lmFunc)
+```{r RxNlmAIC, echo=TRUE, eval=TRUE, results='hide'}
+# perform model selection for responsive transcripts
+RxNlmAIC <- try(dlply(sig.TPM.dt.sub, .(Transcript), lmFunc))
 ```
@@
 
-pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts (of 105,536 total) at 5% FDR with expression that depends on colony, temperature or their interaction.")
+pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts of 99,861 total with expression that depends on colony, temperature or their interaction at 5% FDR.")
 ```
@@
 # calculate response type for responsive transcripts
-interaction.response.type <- ldply(interaction.lms, .progress = "text", .inform = TRUE, RxNtype)
+interaction.response.type <- ldply(interaction.lms, .progress = "none", .inform = TRUE, RxNtype)
 stopifnot(nrow(interaction.response.type) == length(interaction.lms))
@@
 # calculate response types for transcripts without interactions
-temperature.response.type <- ldply(temperature.lms, .progress = "text", .inform = TRUE, RxNtype)
+temperature.response.type <- ldply(temperature.lms, .progress = "none", .inform = TRUE, RxNtype)
 stopifnot(nrow(temperature.response.type) == length(temperature.lms))
@@
 
-Examine transcripts by temperature response. Compare between colonies.
-
-```{r temperature_response, eval=TRUE, echo=FALSE, results='asis'}
+Next, I compare the number of thermally-responsive in each response category between the two colonies.
+
+```{r temperature_response}
 A22.type.table <- table(Ap.response.type[ , 'A22.type'])
@@
 
-type.table <- table(Ap.response.type[ , 'A22.type'], Ap.response.type[ , 'Ar.type'])
-# reorder for plotting
-tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]
-
-png(file = paste(resultsdir, "responsive_transcripts_mosaicplot.png", sep = ""))
-mosaicplot(tt2, xlab = "ApVT", ylab = "AcNC", main = "Mosaic plot of responsive transcripts")
-dev.off()
-
-pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
+# table
+Ap.type.table <- rbind(A22.type.table, Ar.type.table)
 
@@
 chisq.test(Ap.type.table)
+
+# Reorganize for plotting to show overlap among categories between colonies
+type.table <- table(Ap.response.type[ , 'A22.type'], Ap.response.type[ , 'Ar.type'])
+
+# reorder 
+tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]
+
+# plot
+mosaicplot(tt2, xlab = "AcNC", ylab = "ApVT", main = "Mosaic plot of responsive transcripts")
+```
+
+The number of thermally-responsive in each response category differs between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies.
+For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts.
+For *AcNC*, transcripts expressed at *Intermediate* temperatures are next most common, followed by *High* and *Bimodal*.
+
+Interestingly, nearly half of the *High* genes in *AcNC* are *Low* in *ApVT*, and vice versa. In contrast, most of the *Low* genes in one species are also *Low* in the other species.
+
+```{r temperature_response_table, echo=FALSE, results='asis'}
+pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.", split.tables = "Inf")
 ```
@@
 
-```{r optimum_expression_comparison, cache=TRUE, eval=FALSE}
+```{r optimum_expression_comparison, cache=TRUE, eval=TRUE}
 # list of transcripts that are 'high' expressed in A22
@@
 non-equal variances.
-This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099")`.
+This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`.
 As my goal is simply to determine if expression is typically greater at 19.25C in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test
@@
 # reshape data
-Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "ApNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
+Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
 
@@
 ```{r offset, echo=TRUE, eval=TRUE}
-varshape.out <- ldply(interaction.lms, .progress = "text", RxNvarshape)
+varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)
 
@@
 ```
+
+## Temperature of gene activation
+
+Thermally-responsive genes could also differ in the temperatures at which they have increased or decreased expression in response to temperature changes.
+To examine this, I determine the temperature at which each responsive gene has the greated increase or decrease in expression.
+
+```{r responsive_TPM}
+# extract TPM data for thermally-responsive transcripts
+resp.TPM.dt.sub <- TPM.dt.sub[names(responsive.lms)]
+setkey(resp.TPM.dt.sub, Transcript)
+str(resp.TPM.dt.sub)
+length(unique(resp.TPM.dt.sub$Transcript))
+# scale transcripts so can compare
+resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
+```
+
+Predict for responsive transcripts
+
+```{r responsive_pred}
+
+predFunc <- function(foo) {
+  # Function supplied to `ddply` to predict expression at each temperature
+  #
+  # Args:
+  #  foo: data.frame model is fit to
+  #
+  #  'model' must be specified in environment  
+  #
+  # Returns:
+  #  data.frame of Transcripts with pval
+
+  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))
+
+  # predict values
+  pout <- predict(lmout)
+  # back-transform from log-scale
+  foo$pTPM <- exp(pout)
+
+  # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
+  foo$pTPM <- ifelse(foo$pTPM < 0, 0, round(foo$pTPM, 5))
+  # plot
+  #with(foo, qplot(val, TPM, colour = colony, geom = "smooth"))
+  #with(foo, qplot(val, pTPM, colour = colony, geom = "smooth"))
+
+  pred.out.df <- data.frame(Transcript = foo$Transcript, TPM = foo$TPM, val = foo$val, colony = foo$colony, pTPM = foo$pTPM)
+}
+
+# apply predFunc to responsive transcripts
+testpred <- ddply(resp.TPM.dt.sub[1:220], .(Transcript), .inform="TRUE", predFunc)
+
+# calculate temperature of greatest change
+T.on.df <- ddply(testpred, .(Transcript, colony), summarise, 
+              T_on <- diff(pTPM))
+
+testdiff[(which(diff(testdiff$pTPM) == max(diff(testdiff$pTPM)))+1), ]
+
+```
+
+### Identification of biological-functions associated with temperature responses
 
@@
 # display
-pandoc.table(Ap.gsea1, split.tables = "Inf")
+Ap.gsea1
 ```
@@
 A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
-#pandoc.table(A22.high.gsea)
+#pandoc.table(A22.high.gsea, split.table = Inf)
 ```
@@
 Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
+#pandoc.table(Ar.high.gsea, split.table = Inf)
 ```
@@
 A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
+#pandoc.table(A22.low.gsea, split.table = Inf)
 ```
@@
 Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
+#pandoc.table(Ar.low.gsea, split.table = Inf)
 ```
@@
 A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
+#pandoc.table(A22.bim.gsea, split.table = Inf)
 ```
@@
 Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
+#pandoc.table(Ar.bim.gsea, split.table = Inf)
 ```
@@
 A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
+#pandoc.table(A22.int.gsea, split.table = Inf)
 ```
@@
 Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
-```
-
-Make a table with results.
+#pandoc.table(Ar.int.gsea, split.table = Inf)
+```
+
+Table of overall GSEA results.
 
@@
 Ar.gsea <- rbind(Ar.high.gsea, Ar.low.gsea, Ar.int.gsea, Ar.bim.gsea)
-Ar.gsea$Colony <- "ApNC"
+Ar.gsea$Colony <- "AcNC"
 
@@
 
-Ap.setdiff.gsea
+print(xtable(Ap.setdiff.gsea, comment = getOption("xtable.comment", FALSE)))
 ```
@@
 
-Organize data
-
-```{r plot_data}
-# extract TPM data for thermally-responsive transcripts
-resp.TPM.dt.sub <- TPM.dt.sub[names(responsive.lms)]
-setkey(resp.TPM.dt.sub, Transcript)
-str(resp.TPM.dt.sub)
-length(unique(resp.TPM.dt.sub$Transcript))
-# scale transcripts so can compare
-resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
-```
-
 Make plots for all genes expressed at *High* temps in GO category "GO:0006950: response to stress"
 
-```{r plot_GOstress, echo=FALSE, eval=FALSE, cache=TRUE}
+```{r plot_GOstress, echo=FALSE, eval=TRUE}
 high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
@@
 # Smooth plot, expression against temp, faceted by colony
-png("results/GO_stress_High.png")
 trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
@@
 print(trp2)
-dev.off()
 ```
@@
 
-```{r plot_GOstress_low, echo=FALSE, eval=FALSE, cache=TRUE}
+```{r plot_GOstress_low, echo=FALSE, eval=TRUE, cache=TRUE}
 low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
@@
 
-```{r references, results='asis'}
+```{r references, echo = FALSE, results='asis'}
 bibliography()

### 2014-05-08 scripts/RxNseq.R

@@ -25,6 +25,60 @@ read.sailfish.quant <- function(filein, outname, samp, trtval) {
 }
 
 
+
+############################################################################################
+## modpFunc
+############################################################################################
+
+modpFunc <- function(foo) {
+  # Function supplied to `ddply` to calculate overall model P value for a linear model
+  #
+  # Args:
+  #  foo: data.frame model is fit to
+  #
+  #  'model' must be specified in environment  
+  #
+  # Returns:
+  #  data.frame of Transcripts with pval
+
+  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))
+  
+  # calculate overall model significance
+  f <- summary(lmout)$fstatistic
+  pval <- unname(pf(f[1],f[2],f[3],lower.tail=F))
+  attributes(pval) <- NULL
+  
+  # return pvalue
+  c(pval = pval) 
+}
+
+
+############################################################################################
+## modpFunc
+############################################################################################
+
+lmFunc <- function(bar) {
+  # Function supplied to `dlply` to calculate overall model P value for a linear model
+  #
+  # Args:
+  #  bar: data.frame model model is fit to
+  #
+  #  'model' must be specified in environment  
+  #
+  # Returns:
+  #  list of class lm
+  
+  
+  lmout <- eval(parse(text = paste("lm(", model, ", data = bar)", sep = "")))  
+  # stepAIC to drop non-significant terms
+  f.lmout <- stepAIC(lmout)
+  # return final model
+  return(f.lmout) 
+}
+
+
+
+
 ############################################################################################
 ## RxNtype
 ############################################################################################

### 2014-05-09 ApTranscriptome_TR.Rmd
Completed analysis to check temp at which expression up-regulated for High genes

@@
 
-1. describe the *de novo* assembly of the transcriptome for two ant colonies with in the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
+1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
 2. identify thermally-responsive genes
-3. perform gene set enrichment analysis
+3. evaluate differences in the expression patterns between the two species
+3. perform gene set enrichment analysis of thermally-responsive genes for the two species
 
@@
 Two ant colonies were used for the transcriptome sequencing.
-The first, designated A22, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen.
-This colony was putatively identifed as *A.
-picea*.
-The second colony, designated Ar, was collected by Lauren Nichols in Raleigh, North Carolina.
+The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen.
+The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina.
 These colonies were maintained in the lab for 6 months prior to sample collection.
+Bernice Bacon DeMarco (Michigan State University) identified colony *A22* as *A.
+picea* and *Ar* as *A.
+carolinensis*.
 
@@
 
-# calculate overall P value for 'model' for each transcript
+# calculate overall P value and R^2 for each transcript
 RxNpval <- ddply(TPM.dt.sub, .(Transcript), .inform="TRUE", modpFunc)
@@
 
-```{r plotthermalbreadth, echo=FALSE, eval=TRUE}
+```{r plot_thermal_breadth, echo=FALSE, eval=TRUE}
 # prep data for plotting
@@
 
-# ggplot
+# plot thermal breadth
 g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
@@
 
-png(paste(resultsdir, "PDF_expression_all.png", sep=""))
-g3 <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
+maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
   geom_density(alpha=0.2, position="identity") + 
@@
   scale_x_continuous(name="Temperature of maximum expression")
-suppressWarnings(print(g3))
-dev.off()
+suppressWarnings(print(maxexplot))
 ```
@@
 
-# apply predFunc to responsive transcripts
-testpred <- ddply(resp.TPM.dt.sub[1:220], .(Transcript), .inform="TRUE", predFunc)
-
-# calculate temperature of greatest change
-T.on.df <- ddply(testpred, .(Transcript, colony), summarise, 
-              T_on <- diff(pTPM))
-
-testdiff[(which(diff(testdiff$pTPM) == max(diff(testdiff$pTPM)))+1), ]
-
+# apply predFunc to all responsive transcripts
+resp.TPM.dt.sub.pred <- ddply(resp.TPM.dt.sub, .(Transcript), .inform="TRUE", predFunc)
+
+# setkey to Transcript and colony 
+resp.TPM.dt.sub.pred <- data.table(resp.TPM.dt.sub.pred)
+setkey(resp.TPM.dt.sub.pred, Transcript, colony)
+
+# transcripts expressed at *High* temperatures in A22
+A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.high, "A22")]
+str(A22.high.TPM.dt.sub)
+
+# make data.frame for results
+A22.high.T_on <- data.frame(Transcript = unique(A22.high.TPM.dt.sub$Transcript), colony = rep(A22.high.TPM.dt.sub$colony, length = length(unique(A22.high.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)
+
+# loop across transcripts, calculating T_on
+
+for(i in unique(A22.high.TPM.dt.sub$Transcript)) {
+    subdf <- A22.high.TPM.dt.sub[i]
+    subdf <- subdf[which(subdf$val > 14), ]
+    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
+    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
+    A22.high.T_on[which(A22.high.T_on$Transcript == i), "T_on"] <- T_on
+    A22.high.T_on[which(A22.high.T_on$Transcript == i), "pT_on"] <- pT_on
+}
+
+# repeat for Ar
+Ar.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.high, "Ar")]
+str(Ar.high.TPM.dt.sub)
+
+Ar.high.T_on <- data.frame(Transcript = unique(Ar.high.TPM.dt.sub$Transcript), colony = rep(Ar.high.TPM.dt.sub$colony, length = length(unique(Ar.high.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)
+
+for(i in unique(Ar.high.TPM.dt.sub$Transcript)) {
+    subdf <- Ar.high.TPM.dt.sub[i]
+    subdf <- subdf[which(subdf$val > 14), ]
+    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
+    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
+    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "T_on"] <- T_on
+    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "pT_on"] <- pT_on
+}
+
+cor.test(Ar.high.T_on$T_on, Ar.high.T_on$pT_on)
+cor.test(A22.high.T_on$T_on, A22.high.T_on$pT_on)
+
+t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
+t.test(Ar.high.T_on$pT_on, A22.high.T_on$pT_on)
+```
+
+Genes with increased expression at *High* temperatures are turned on at higher temperatures in *ApVT* than *AcNC*.
+
+```{r plot_high_T_on, echo=FALSE, eval=TRUE}
+# prep data for plotting
+Ap.high.T_on <- rbind(A22.high.T_on, Ar.high.T_on)
+
+# plot
+png("T_on_high.png")
+T_on_plot <- ggplot(Ap.high.T_on, aes(x=T_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
+T_on_plot + scale_y_continuous(name="Density") +
+  scale_x_continuous(name=expression("Temperature"))
+dev.off()
+
+# plot predicted
+png("pT_on_high.png")
+pT_on_plot <- ggplot(Ap.high.T_on, aes(x=pT_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
+pT_on_plot + scale_y_continuous(name="Density") +
+  scale_x_continuous(name=expression("Temperature"))
+dev.off()
 ```
@@
 # combine responsive.lms.ann with "type" information
-responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
+responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "Ar.type", "Ar.type")], by = "Transcript", all = TRUE)
 str(responsive.lms.ann.type)

### 2014-05-09 scripts/RxNseq.R

@@ -49,7 +49,7 @@ modpFunc <- function(foo) {
   attributes(pval) <- NULL
   
   # return pvalue
-  c(pval = pval) 
+  data.frame(Transcript = unique(foo$Transcript), pval = pval, adj.r.squared = summary(lmout)$adj.r.squared) 
 }
 
 

### 2014-05-09 ApTranscriptome_TR.Rmd
Updated analysis of temperature gene activation for *Low* genes

@@
 
-Predict for responsive transcripts
+Predict expression for responsive transcripts.
 
 ```{r responsive_pred}
-
-predFunc <- function(foo) {
-  # Function supplied to `ddply` to predict expression at each temperature
-  #
-  # Args:
-  #  foo: data.frame model is fit to
-  #
-  #  'model' must be specified in environment  
-  #
-  # Returns:
-  #  data.frame of Transcripts with pval
-
-  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))
-
-  # predict values
-  pout <- predict(lmout)
-  # back-transform from log-scale
-  foo$pTPM <- exp(pout)
-
-  # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
-  foo$pTPM <- ifelse(foo$pTPM < 0, 0, round(foo$pTPM, 5))
-  # plot
-  #with(foo, qplot(val, TPM, colour = colony, geom = "smooth"))
-  #with(foo, qplot(val, pTPM, colour = colony, geom = "smooth"))
-
-  pred.out.df <- data.frame(Transcript = foo$Transcript, TPM = foo$TPM, val = foo$val, colony = foo$colony, pTPM = foo$pTPM)
-}
-
 # apply predFunc to all responsive transcripts
@@
 setkey(resp.TPM.dt.sub.pred, Transcript, colony)
-
+```
+
+Calculate `T_on` for *High* genes in each colony.
+
+```{r T_on_high}
 # transcripts expressed at *High* temperatures in A22
@@
 cor.test(A22.high.T_on$T_on, A22.high.T_on$pT_on)
-
+```
+
+Determine if `T_on` is greater in *A22* or *Ar* for *High* genes.
+
+```{r T_on_high_ttest}
 t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
@@
 ```
-
 Genes with increased expression at *High* temperatures are turned on at higher temperatures in *ApVT* than *AcNC*.
+
+Plot `T_on` for *High* genes.
 
@@
 # plot
-png("T_on_high.png")
 T_on_plot <- ggplot(Ap.high.T_on, aes(x=T_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
@@
   scale_x_continuous(name=expression("Temperature"))
-dev.off()
 
 # plot predicted
-png("pT_on_high.png")
 pT_on_plot <- ggplot(Ap.high.T_on, aes(x=pT_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
@@
   scale_x_continuous(name=expression("Temperature"))
-dev.off()
+```
+
+Repeat analysis for *Low* genes.
+
+```{r T_on_low}
+# transcripts expressed at *Low* temperatures in A22
+A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.low, "A22")]
+str(A22.low.TPM.dt.sub)
+
+intersect(A22.low.TPM.dt.sub, A22.high.TPM.dt.sub)
+# Good - no overlap
+
+# make data.frame for results
+A22.low.T_on <- data.frame(Transcript = unique(A22.low.TPM.dt.sub$Transcript), colony = rep(A22.low.TPM.dt.sub$colony, length = length(unique(A22.low.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)
+
+# loop across transcripts, calculating T_on
+
+for(i in unique(A22.low.TPM.dt.sub$Transcript)) {
+    subdf <- A22.low.TPM.dt.sub[i]
+    subdf <- subdf[which(subdf$val < 21), ]
+    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
+    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
+    A22.low.T_on[which(A22.low.T_on$Transcript == i), "T_on"] <- T_on
+    A22.low.T_on[which(A22.low.T_on$Transcript == i), "pT_on"] <- pT_on
+}
+
+# repeat for Ar
+Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.low, "Ar")]
+str(Ar.low.TPM.dt.sub)
+
+intersect(A22.low.TPM.dt.sub, A22.high.TPM.dt.sub)
+# Good - no overlap
+
+Ar.low.T_on <- data.frame(Transcript = unique(Ar.low.TPM.dt.sub$Transcript), colony = rep(Ar.low.TPM.dt.sub$colony, length = length(unique(Ar.low.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)
+
+for(i in unique(Ar.low.TPM.dt.sub$Transcript)) {
+    subdf <- Ar.low.TPM.dt.sub[i]
+    subdf <- subdf[which(subdf$val < 21), ]
+    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
+    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
+    Ar.low.T_on[which(Ar.low.T_on$Transcript == i), "T_on"] <- T_on
+    Ar.low.T_on[which(Ar.low.T_on$Transcript == i), "pT_on"] <- pT_on
+}
+
+t.test(Ar.low.T_on$T_on, A22.low.T_on$T_on)
+t.test(Ar.low.T_on$pT_on, A22.low.T_on$pT_on)
+```
+
+Genes with increased expression at *Low* temperatures are turned on at lower temperatures in *AcNC* than *ApVT*.
+
+Plot `T_on` for *Low* genes.
+
+```{r plot_low_T_on, echo=FALSE, eval=TRUE}
+# prep data for plotting
+Ap.low.T_on <- rbind(A22.low.T_on, Ar.low.T_on)
+
+# plot
+T_on_plot_low <- ggplot(Ap.low.T_on, aes(x=T_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
+T_on_plot_low + scale_y_continuous(name="Density") +
+  scale_x_continuous(name=expression("Temperature"))
+
+# plot predicted
+pT_on_plot_low <- ggplot(Ap.low.T_on, aes(x=pT_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
+pT_on_plot_low + scale_y_continuous(name="Density") +
+  scale_x_continuous(name=expression("Temperature"))
 ```

### 2014-05-09 scripts/RxNseq.R

@@ -261,6 +261,36 @@ RxNvarshape <- function(lmitem) {
 
 
 ############################################################################################
+## predFunc
+############################################################################################
+
+  predFunc <- function(foo) {
+  # Function supplied to `ddply` to predict expression at each temperature
+  #
+  # Args:
+  #  foo: data.frame model is fit to
+  #
+  #  'model' must be specified in environment  
+  #
+  # Returns:
+  #  data.frame of Transcripts with pval
+
+  lmout <- eval(parse(text = paste("lm(", model, ", data = foo)", sep = "")))
+  
+  # predict values
+  pout <- predict(lmout)
+  # back-transform from log-scale
+  foo$pTPM <- exp(pout)
+  
+  # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
+  foo$pTPM <- ifelse(foo$pTPM < 0, 0, round(foo$pTPM, 5))
+
+  pred.out.df <- data.frame(Transcript = foo$Transcript, TPM = foo$TPM, val = foo$val, colony = foo$colony, pTPM = foo$pTPM)
+}
+
+
+
+############################################################################################
 ## geneid2GOmap
 ############################################################################################
 

### 2014-05-09 doc/ApTranscriptome_manuscript.Rmd
Incorporated edits to methods from AN

@@
 
-*Aphaenogaster picea* was collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and *A.
-carolinensis* was collected from Durham, North Carolina ( 36.037° N -78.874° W), designated as *A.
-picea* and *A.
-carolinensis* henceforth.
-Though the phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], preliminary work supports that these species are separate species (Bernice Bacon DeMarco, pers.
+Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *AcNC* respectively.
+The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers.
 communication).
-species of these species were maintained in the lab for 5 months prior to experimentation.
+species of these species were maintained in the lab for 6 months prior to experimentation.
 
-From a single species of each species, we haphazardly collected 12 ants at the same time on 12 days.
-Each day, the 12 ants were placed in glass tubes into a water bath (...) at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour.
-The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* [@warren2013.
+From each species, we haphazardly collected 12 ants at the same time on 12 days.
+Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour.
+The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
 At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
-mRNA was extracted from three pooled ants using a RNeasy kit (Qiagen Inc; Valencia, CA).
-Ants were homogenized in Buffer ATL with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY) and the standard Qiagen protocol was followed from this point.
+We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then a RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
 

### 2014-05-12 ApTranscriptome_TR.Rmd
Updated figures for T_on

@@
 # plot
-T_on_plot <- ggplot(Ap.high.T_on, aes(x=T_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
-T_on_plot + scale_y_continuous(name="Density") +
+T_on_plot_high <- ggplot(Ap.high.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    scale_y_continuous(name="Density") +
     scale_x_continuous(name=expression("Temperature"))
+print(T_on_plot_high)
 
 # plot predicted
-pT_on_plot <- ggplot(Ap.high.T_on, aes(x=pT_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
-pT_on_plot + scale_y_continuous(name="Density") +
+pT_on_plot_high <- ggplot(Ap.high.T_on, aes(x = pT_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    scale_y_continuous(name="Density") +
     scale_x_continuous(name=expression("Temperature"))
+print(pT_on_plot_high)
 ```
@@
 Ap.low.T_on <- rbind(A22.low.T_on, Ar.low.T_on)
-
 # plot
-T_on_plot_low <- ggplot(Ap.low.T_on, aes(x=T_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
-T_on_plot_low + scale_y_continuous(name="Density") +
+T_on_plot_low <- ggplot(Ap.low.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    scale_y_continuous(name="Density") +
     scale_x_continuous(name=expression("Temperature"))
-
-# plot predicted
-pT_on_plot_low <- ggplot(Ap.low.T_on, aes(x=pT_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
-pT_on_plot_low + scale_y_continuous(name="Density") +
+print(T_on_plot_low)
+
+# plot
+pT_on_plot_low <- ggplot(Ap.low.T_on, aes(x = pT_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    scale_y_continuous(name="Density") +
     scale_x_continuous(name=expression("Temperature"))
+print(pT_on_plot_low)
 ```

### 2014-05-13 doc/ApTranscriptome_manuscript.Rmd
Some changes

@@
+---
+output:
+  pdf_document:
+    biblatex: yes
+biblio-files: ApTranscriptome.bib
+---
 Thermal reactionome of a temperate ant species
@@
 
-```{r load, echo=FALSE, results='hide'}
-options(scipen = 1, digits = 3, warn = -1)
+```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
+options(scipen = 1, digits = 3)
 load("../ms.RData")
-suppressMessages(library(pander))
-suppressMessages(library(ggplot2))
+library(pander)
+library(ggplot2)
 ```
@@
 
-![Expression (scaled to allow comparisons among genes) for genes in GO category "response to stress"](../results/GO_stress_High.png)
+#![Expression (scaled to allow comparisons among genes) for genes in GO category "response to stress"](../results/GO_stress_High.png)
 
@@
 Given the large differences in number (Table 1) and responsiveness of genes between the two species, a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures.
-Intriguingly, this
-result mirrors the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
+Intriguingly, this result mirrors the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
 A plausible explanation is that organisms have fewer molecular mechanisms to tolerate heat than cold stress.

### 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
Small edits

@@
 communication).
-species of these species were maintained in the lab for 6 months prior to experimentation.
+Colonies of these species were maintained in the lab for 6 months prior to experimentation.
 
@@
 Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour.
-The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
+The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
 At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
-We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then a RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
+We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
 
@@
 
-For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the A.
-picea and A.
-carolinensis samples respectively.
+For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the *A.
+picea* and *A.
+carolinensis* samples respectively.
 Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
 
-We assembled the sequenced reads into the full set of mRNA transcripts (the `transcriptome`) for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
+We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
 *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g.
-regions of two separate transcripts with enough overlap to assemble into a false, or chimeric, third transcript) [@yang2013] To account for this, we first ran program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length, and second ran the program `uclust` which cluster sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental for exact parameters).
-We used liberal values (90% similarity) to merge transcripts to account for the fact that we were including sequencing data from two species of the same species, and wanted to avoid assembling orthologs individually.
+regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013].
+We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
+Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters).
+We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
 We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
 
-We performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
+To determine the putative function of the transcripts, we performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
 
@@
 
-We quantified expression of each gene using the program `Sailfish` [@patro2013].
-`Sailfish` reports gene expression as transcripts per million (TPM) which is the preferred measure of gene expression as values are comparable among sequencing runs [@wagner2012].
-In addition, `Sailfish` corrects for sequence composition bias and gene length.
+We quantified expression of each gene using the program `Sailfish` [@patro2013] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression.
 As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis.
@@
 To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction.
-For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Expressed*.
+For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Responsive*.
 To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.
@@
 Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage.
-Of these, 5,787 genes have best hits to putative genes in the genome sequence of the ant species *Solenopsis invicta* and XXXX are annotated to genes previously identified from organisms within the class Insecta.
+Of these, 5,787 genes have best hits to putative genes in the genome sequence of the ant species *Solenopsis invicta* and 47% have best hits to genes from the class Insecta.
 

### 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
Edits to ms

@@
 
-**Authors**: John Stanton-Geddes, Andrew Nguyen, Rob Dunn, Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan
+**Authors**: John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn, Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan
 
@@
 We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
-We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
+To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmeider2011] with the provided bacteria, virus, archae and human databases of contaminants.
 
 To determine the putative function of the transcripts, we performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
+
+We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
 
@@
 We quantified expression of each gene using the program `Sailfish` [@patro2013] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression.
+We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses.
 As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis.
@@
 
+Finally, to test whether the temperature at which thremally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression.
+We grouped the *High* genes along with the high temperature end of the *Bimodal* genes, and did the same for *Low* genes.
+We then tested if the temperature at which the *High* or *Low* group differed among the two species.
+
 ### Data availability
@@
 Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage.
-Of these, 5,787 genes have best hits to putative genes in the genome sequence of the ant species *Solenopsis invicta* and 47% have best hits to genes from the class Insecta.
+DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts.
+Of these, 47% have best hits to genes from the class Insecta.
 
@@
 
-We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR.
-Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a species x temperature interaction.
+We identified 22,399 genes with overall model fits of P < 0.05, retaining 10,597 after correcting for FDR.
+Of these transcripts, 9,156 significantly change expression with temperature including 6,908 that have a species x temperature interaction.
 The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1).
@@
 pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
+```
 
+```{r expression_type_plot, echo=FALSE}
 mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
@@
 carolinensis*) and colder (*A.
-picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa.
-Consistent with one of our predictions, we found that constitutive expression was greater in the more southern *A.
+picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater expression near optimal temperatures of 19°C in the warmer-climate species, and vice versa.
+Partially consistent with our prediction, we found that optimal-temperature expression was greater in the more southern *A.
 carolinensis* species for *High* genes in *A.
@@
 However, for genes with *Low* expression in *A.
-carolinensis* we also found that constitutive expression was greater in *A.
+carolinensis* we also found that optimal-temperature expression was greater in *A.
 carolinensis* than *A.
 picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).
+Moreover, we also found that for genes with *Intermediate* expression in either colony, optimal-temperature expression was greater in *A.
+carolinensis*.
+The overall conclusion from these analyses is that for thermally-responsive genes, the more southern species, *A.
+carolinensis* has greater optimal-temperature expression.
 
-We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range) for the *Intermediate* expressed genes.
+We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* expressed genes.
 Contrary to our expectations based on critical maximum and minimal temperatures, *A.
-carolinensis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
+carolinensis* had a transcriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
 In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
+
+We also explored the extent to which the temperature at which thermally-responsive genes are activated differs between species.
+For genes that are upregulated at *High* temperatures, including both *High* and *Bimodal* genes, we found that ...
+In contrast, for genes that are upregulated at *Low* temperatures, including both *Low* and *Bimodal* genes, we found that...
 

### 2014-05-14 ApTranscriptome_TR.Rmd
Updated T~on~ analysis.
Merged both *Low* and *High* onto same plot - better visualization of when genes turned on.

--- 2014-05-09 ApTranscriptome_TR.Rmd
@@
 
-**May 6, 2014**
+**May 14, 2014**
 
@@
 library(xtable)
-library(methods)
+library(methods) # required by `knitcitations` when file is `knit` file from command-line
 
@@
 
-    sailfish -i sailfish-index-Trinity-cap3-uclust -o sailfish-expression-Trinity-cap3-uclust/A22-0 -l "T=PE:O=><:S=U" -r A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq -1 A22-0_ATCACG.paired.left.fastq -2 A22-0_ATCACG.paired.right.fastq  -p 4
+    sailfish quant -i sailfish-index-Trinity-cap3-uclust -o sailfish-expression-Trinity-cap3-uclust/A22-0 -l "T=SE:S=U" -r A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq A22-0_ATCACG.paired.left.fastq A22-0_ATCACG.paired.right.fastq -p 4
+
+While it is possible to separately specify the paired-end and orphaned single-end reads in Sailfish v0.6.3, the results are exactly the same as if they are all entered as SE.
 
@@
 # perform model selection for responsive transcripts
+# need to use `try` to avoid stopping on error for AIC at Infinity
 RxNlmAIC <- try(dlply(sig.TPM.dt.sub, .(Transcript), lmFunc))
@@
 # save results to file
-write.table(file = paste(resultsdir, "Ap_responsive_transcripts.txt", sep = ""), Ap.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
+write.table(file = paste(resultsdir, "Ap-responsive-transcripts", Sys.Date(), ".txt", sep = ""), Ap.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
 ```
@@
 
-```{r optimum_expression_comparison, cache=TRUE, eval=TRUE}
+```{r optimum_expression_comparison}
 # list of transcripts that are 'high' expressed in A22
@@
 ```{r A22_high_wilcoxon}
-(w1 <- wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE))
+w1 <- wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+w1
 ```
@@
 
-# compare expression at optimum temp (Ar.opt) between colonies using t-test
-t.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
-boxplot(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
 # t-test with log values
@@
 # Wilcoxon signed rank-test
-(w2 <- wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE))
+w2 <- wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+w2
 ```
@@
 # T test
-t.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt)
 t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
@@
 # T test
-t.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt)
 t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
@@
 setkey(resp.TPM.dt.sub.pred, Transcript, colony)
-
-# transcripts expressed at *High* temperatures in A22
-A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.high, "A22")]
+```
+
+For next analyses, extract list of gene names by response type.
+
+```{r extractionTranscripts}
+A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
+Ar.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]
+
+A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
+Ar.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
+
+A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
+Ar.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+
+A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
+Ar.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+```
+
+Calculate T~on~ for *High* genes in each colony.
+
+```{r T_on_high}
+# transcripts expressed at *High* and *Bimodal* temperatures in A22
+A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.high, A22.bim), "A22")]
+setkey(A22.high.TPM.dt.sub, Transcript)
 str(A22.high.TPM.dt.sub)
@@
 # make data.frame for results
-A22.high.T_on <- data.frame(Transcript = unique(A22.high.TPM.dt.sub$Transcript), colony = rep(A22.high.TPM.dt.sub$colony, length = length(unique(A22.high.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)
+l1 <- length(unique(A22.high.TPM.dt.sub$Transcript))
+A22.high.T_on <- data.frame(Transcript = unique(A22.high.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l1), type = rep("High", length = l1), T_on = NA)
 
@@
     T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
-    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
     A22.high.T_on[which(A22.high.T_on$Transcript == i), "T_on"] <- T_on
-    A22.high.T_on[which(A22.high.T_on$Transcript == i), "pT_on"] <- pT_on
 }
@@
 # repeat for Ar
-Ar.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.high, "Ar")]
+Ar.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.high, Ar.bim), "Ar")]
+setkey(Ar.high.TPM.dt.sub, Transcript)
 str(Ar.high.TPM.dt.sub)
 
-Ar.high.T_on <- data.frame(Transcript = unique(Ar.high.TPM.dt.sub$Transcript), colony = rep(Ar.high.TPM.dt.sub$colony, length = length(unique(Ar.high.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)
+l2 <- length(unique(Ar.high.TPM.dt.sub$Transcript))
+Ar.high.T_on <- data.frame(Transcript = unique(Ar.high.TPM.dt.sub$Transcript), colony = rep("AcNC", length = l2), type = rep("High", length = l2), T_on = NA)
 
@@
     T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
-    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
     Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "T_on"] <- T_on
-    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "pT_on"] <- pT_on
 }   
-
-cor.test(Ar.high.T_on$T_on, Ar.high.T_on$pT_on)
-cor.test(A22.high.T_on$T_on, A22.high.T_on$pT_on)
-
+```
+
+Determine if T~on~ is greater in *A22* or *Ar* for *High* genes.
+
+```{r T_on_high_ttest}
 t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
-t.test(Ar.high.T_on$pT_on, A22.high.T_on$pT_on)
-```
-
-Genes with increased expression at *High* temperatures are turned on at higher temperatures in *ApVT* than *AcNC*.
-
-```{r plot_high_T_on, echo=FALSE, eval=TRUE}
-# prep data for plotting
-Ap.high.T_on <- rbind(A22.high.T_on, Ar.high.T_on)
+```
+
+Genes with increased expression at *High* temperatures are turned on at, on average, 1°C higher in *AcNC* than *ApVT*.
+
+Repeat analysis for *Low* genes.
+
+```{r T_on_low}
+# transcripts expressed at *Low* temperatures in A22
+A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.low, "A22")]
+str(A22.low.TPM.dt.sub)
+
+intersect(A22.low.TPM.dt.sub, A22.high.TPM.dt.sub)
+# Good - no overlap
+
+# make data.frame for results
+l3 <- length(unique(A22.low.TPM.dt.sub$Transcript))
+A22.low.T_on <- data.frame(Transcript = unique(A22.low.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l3), type = rep("Low", length = l3), T_on = NA)
+
+# loop across transcripts, calculating T_on
+
+for(i in unique(A22.low.TPM.dt.sub$Transcript)) {
+    subdf <- A22.low.TPM.dt.sub[i]
+    subdf <- subdf[which(subdf$val < 21), ]
+    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
+    A22.low.T_on[which(A22.low.T_on$Transcript == i), "T_on"] <- T_on
+}
+
+# repeat for Ar
+Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.low, "Ar")]
+str(Ar.low.TPM.dt.sub)
+
+intersect(A22.low.TPM.dt.sub, A22.high.TPM.dt.sub)
+# Good - no overlap
+
+l4 <- length(unique(Ar.low.TPM.dt.sub$Transcript))
+Ar.low.T_on <- data.frame(Transcript = unique(Ar.low.TPM.dt.sub$Transcript), colony = rep("AcNC", length = l4), type = rep("Low", length = length(unique(Ar.low.TPM.dt.sub$Transcript))), T_on = NA)
+
+for(i in unique(Ar.low.TPM.dt.sub$Transcript)) {
+    subdf <- Ar.low.TPM.dt.sub[i]
+    subdf <- subdf[which(subdf$val < 21), ]
+    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
+    Ar.low.T_on[which(Ar.low.T_on$Transcript == i), "T_on"] <- T_on
+}
+
+t.test(Ar.low.T_on$T_on, A22.low.T_on$T_on)
+```
+
+Genes with increased expression at *Low* temperatures are turned on at lower temperatures in *AcNC* than *ApVT*.
+
+Visualize T~on~ for both *Low* and *High* genes on the same plot.
+
+```{r plot_low_T_on, echo=FALSE, eval=TRUE}
+# merge data for plotting
+Ap.T_on <- rbind(A22.high.T_on, Ar.high.T_on, A22.low.T_on, Ar.low.T_on)
 
 # plot
-png("T_on_high.png")
-T_on_plot <- ggplot(Ap.high.T_on, aes(x=T_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
-T_on_plot + scale_y_continuous(name="Density") +
+T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    scale_y_continuous(name="Density") +
     scale_x_continuous(name=expression("Temperature"))
-dev.off()
-
-# plot predicted
-png("pT_on_high.png")
-pT_on_plot <- ggplot(Ap.high.T_on, aes(x=pT_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
-pT_on_plot + scale_y_continuous(name="Density") +
-  scale_x_continuous(name=expression("Temperature"))
-dev.off()
+print(T_on_plot)
 ```
@@
 str(responsive.lms.ann.type)
-write.csv(responsive.lms.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
+write.csv(responsive.lms.ann.type, file = paste(resultsdir, "Ap_responsive_genes_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
 ```
@@
 
-```{r gsea_all}
+```{r gsea_all, results='hide'}
 # create geneList. note that NA values cause problems with topGO
@@
 # display
-Ap.gsea1
+Ap.gsea1[ , 1:3]
 ```
@@
 ```{r gsea_high_A22}
-A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 ```{r gsea_high_Ar}
-Ar.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 ```{r gsea_low_A22}
-A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 ```{r gsea_low_Ar}
-Ar.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 ```{r gsea_bim_A22}
-A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 ```{r gsea_bim_Ar}
-Ar.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 ```{r gsea_int_A22}
-A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 ```{r gsea_int_Ar}
-Ar.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 # save RData for manuscript file
-save(tt2, Ap.df, w1, w2, file = "ms.RData")
+save(tt2, Ap.df, w1, w2, Ap.T_on, file = "ms.RData")
 ```

### 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
Updated manuscript with YAML header so can render with rmarkdown. Allows direction citation using pandoc rather than having to run separately.
Minor updates to text.

@@
 ---
+title: "Thermal reactionome of a temperate ant species"
+author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
+  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
 output:
   pdf_document:
-    biblatex: yes
-biblio-files: ApTranscriptome.bib
+    fig_caption: yes
+    latex_engine: xelatex
+  html_document: null
+bibliography: ApTranscriptome.bib
+csl: ecology.csl
 ---
-Thermal reactionome of a temperate ant species
-============================================
 
-**Authors**: John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn, Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan
+## Abstract
+
+When ants are exposed to different temperatures, lots of genes change expression.
 
@@
 The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp.
-Filtering to remove redundant or chimeric reads resulted in an assembly with 96,253 contigs and a total of 105,536 genes.
-For all genes, the total length was 63 million bp with an N~50~ length of 895 bp and a mean contig size of 593 bp.
+Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes.
+The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp.
 Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database.
@@
 DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts.
-Of these, 47% have best hits to genes from the class Insecta.
+Of these, 47% have best hits to genes from the class Insecta [Supplementary Krona plot](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
 
@@
 ```{r expression_type_table, echo=FALSE, results='asis'}
-pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
+pandoc.table(tt2, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species.")
 ```
 
-```{r expression_type_plot, echo=FALSE}
+**Table 1 and Figure 1 are redundant - which is preferable??**
+
+```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
 mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
@@
 
-![Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. picea* (red) species](../results/PDF_expression_all.png)
+```{r exp_plot, echo=FALSE, warning = FALSE, fig.cap = "Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
+
+ggplot(Ap.df, aes(x=max.val, fill=colony)) +
+  geom_density(alpha = 0.2, position = "identity") +
+  scale_y_continuous(name = "Density") +
+  scale_x_continuous(name = "Temperature of maximum expression")
+
+```
 
@@
 
+```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures.
+Bars are Probability density function of peak expression for all thermally-responsive in *A.
+carolinensis* (blue) and *A.
+carolinensis* (red) species"}
+T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    scale_y_continuous(name="Density") +
+    scale_x_continuous(name=expression("Temperature"))
+print(T_on_plot)
+```
+
 We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock.
@@
 carolinensis*.
-
-#![Expression (scaled to allow comparisons among genes) for genes in GO category "response to stress"](../results/GO_stress_High.png)
 

### 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
Updated manuscript with YAML header so can render with rmarkdown. Allows direction citation using pandoc rather than having to run separately.
Minor updates to text.

@@
   Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
+csl: ecology.csl
 output:
@@
 bibliography: ApTranscriptome.bib
-csl: ecology.csl
 ---
@@
 The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp.
-Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes.
-The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp.
+Filtering to remove redundant or chimeric reads resulted in an assembly with 96,253 contigs and a total of 105,536 genes.
+For all genes, the total length was 63 million bp with an N~50~ length of 895 bp and a mean contig size of 593 bp.
 Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database.
@@
 DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts.
-Of these, 47% have best hits to genes from the class Insecta [Supplementary Krona plot](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
+Of these, 47% have best hits to genes from the class Insecta.
 

### 2014-05-15 ApTranscriptome_TR.Rmd
Fixing analyses for manuscript

@@
 opts_chunk$set(cache=TRUE)
-options(width = 80)
 
@@
 
+To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) `r citep("doi:10.1186/1471-2105-12-385")`. I ...
+
+~~~
+KRONA code
+~~~
+
+The interactive visualization is available [here]().
+
 ## Transcriptome annotation ##
@@
 
-```{r sailfish_quant, eval=TRUE, echo=TRUE, cache=TRUE}
+```{r sailfish_quant, results = 'hide'}
 # directory containing trimmed reads
@@
 readlist <- list.files(readdir)
-(paired.left <- readlist[grep(".\\.paired.left.fastq$", readlist)])
-(paired.right <- readlist[grep("\\.paired.right.fastq$", readlist)])
-(unpaired.left <- readlist[grep("unpaired.left.fastq$", readlist)])
-(unpaired.right <- readlist[grep("unpaired.right.fastq$", readlist)])
+paired.left <- readlist[grep(".\\.paired.left.fastq$", readlist)]
+paired.right <- readlist[grep("\\.paired.right.fastq$", readlist)]
+unpaired.left <- readlist[grep("unpaired.left.fastq$", readlist)]
+unpaired.right <- readlist[grep("unpaired.right.fastq$", readlist)]
 
@@
 A22.TPM[val==38,val:=38.5]
-head(A22.TPM)
 str(A22.TPM)
@@
 Ar.TPM[val==38,val:=38.5]
-head(Ar.TPM)
 str(Ar.TPM)
-
-tables()
 ```
@@
 
-cortable <- cbind(temp, cors)
+cortable <- cbind(Temperature = temp, Correlation = cors)
 ```
@@
 ```{r exp_correlations_table, echo=FALSE, results='asis'}
-pandoc.table(cortable, style="rmarkdown", caption = "correlations between colonies at each temperature treatment")
+pandoc.table(cortable, style="rmarkdown", caption = "Correlations between species for gene expression at temperature treatment")
 ```
@@
 
-```{r combined_TPM, echo=TRUE, eval=TRUE, cache=TRUE}
+```{r combined_TPM, results = 'hide'}
 A22.TPM[,colony:="A22"]
@@
 ### Remove *contaminant* transcripts
+
+DeconSeq identified contaminant genes. I remove this n
 
@@
 # Pearson Chi-square test
-chisq.test(Ap.type.table)
+chi1 <- chisq.test(Ap.type.table)
+chi1
 
 # Reorganize for plotting to show overlap among categories between colonies
-type.table <- table(Ap.response.type[ , 'A22.type'], Ap.response.type[ , 'Ar.type'])
-
+type.table <- table(Ap.response.type[ , 'Ar.type'], Ap.response.type[ , 'A22.type'])
 # reorder 
@@
 # plot
-mosaicplot(tt2, xlab = "AcNC", ylab = "ApVT", main = "Mosaic plot of responsive transcripts")
+mosaicplot(t(tt2), ylab = "AcNC", xlab = "ApVT", main = "Mosaic plot of responsive transcripts")
 ```
@@
 ```{r temperature_response_table, echo=FALSE, results='asis'}
-pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.", split.tables = "Inf")
+# add row and column totals
+tt3 <- rbind(tt2, colSums(tt2))
+tt3 <- cbind(tt3, rowSums(tt3))
+colnames(tt3)[6] <- "Total"
+rownames(tt3)[6] <- "Total"
+# move names into table
+tt3 <- rbind(rownames(tt3), tt3)
+tt3 <- cbind(rownames(tt3), tt3)
+# name col/row by species
+rownames(tt3) <- c("AcNC", "", "", "", "", "", "")
+colnames(tt3) <- c("ApVT", "", "", "", "", "", "")
+
+pandoc.table(tt3, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.", split.tables = "Inf")
 ```
@@
 
-### Plasticity versus constitutive expression
-
-Selection may have acted in response to thermal stress such that some genes are constitutively activated in one colony while plastically-expressed in the other.
+### Plasticity versus basal expression
+
+Selection may have acted in response to thermal stress such that some genes are basaly activated in one colony while plastically-expressed in the other.
 I evaluate this hypothesis by comparing expression levels at the at optimum (19.5C) temperature between the two colonies for genes in that are either in the 'High' or 'Low' expression group in the other colony.
@@
 
-The non-parametric test for both comparisions also finds greater constitutive expression in *Ar* than *A22*.
+The non-parametric test for both comparisions also finds greater expression in *Ar* than *A22* at the optimal temperature.
 
@@
 
+t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))
+
 # calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
@@
 varshape.out$offset <- varshape.out$A22.mean - varshape.out$Ar.mean
-# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
-varshape.out$s.offset <- NA
-for(i in 1:nrow(varshape.out)) {
-  varshape.out[i, "s.offset"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.mean"], varshape.out[i, "Ar.mean"])))
-}
 
 # t-test
-(t3 <- t.test(varshape.out$s.offset, alternative = "two.sided"))
+(t.offset <- t.test(varshape.out$s.offset, alternative = "two.sided"))
 
@@
 varshape.out$slope <- varshape.out$A22.slope - varshape.out$Ar.slope
-# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
-varshape.out$s.slope <- NA
-for(i in 1:nrow(varshape.out)) {
-  varshape.out[i, "s.slope"] <- (varshape.out[i, "slope"] / mean(c(varshape.out[i, "A22.slope"], varshape.out[i, "Ar.slope"])))
-}
-# for transcripts where both samples had slope=0, change s.slope from Inf to 0 as there is truly no difference in slope
-varshape.out[which(varshape.out$Ar.slope == 0 & varshape.out$A22.slope == 0), "s.slope"] <- 0
 
 # t-test
-(t4 <- t.test(varshape.out$s.slope, alternative = "two.sided"))
+(t.slope <- t.test(varshape.out$s.slope, alternative = "two.sided"))
 
@@
 # t-test
-(t.test(varshape.out$s.curvature, alternative = "two.sided"))
+(t.curvature <- t.test(varshape.out$s.curvature, alternative = "two.sided"))
 
@@
 varshape.out[which(varshape.out$Ar.wiggle == 0 & varshape.out$A22.wiggle == 0), "s.wiggle"] <- 0
-# note that only 407 transcripts have any 'wiggle'
+# note small number of genes with 'wiggle'
 length(which(varshape.out$s.wiggle != 0))
@@
 # t-test
-(t5 <- t.test(varshape.out$s.wiggle, alternative = "two.sided"))
+(t.wiggle <- t.test(varshape.out$s.wiggle, alternative = "two.sided"))
 ```
@@
 ```{r partition_reaction_norm_variation}
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.offset <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.offset"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.mean"], varshape.out[i, "Ar.mean"])))
+}
+# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
+varshape.out$s.slope <- NA
+for(i in 1:nrow(varshape.out)) {
+  varshape.out[i, "s.slope"] <- (varshape.out[i, "slope"] / mean(c(varshape.out[i, "A22.slope"], varshape.out[i, "Ar.slope"])))
+}
+# for transcripts where both samples had slope=0, change s.slope from Inf to 0 as there is truly no difference in slope
+varshape.out[which(varshape.out$Ar.slope == 0 & varshape.out$A22.slope == 0), "s.slope"] <- 0
+
                                         # weighted mean-standardized values of each measure
@@
 mean(varshape.out$prop.wiggle)
-quantile(varshape.out$prop.wiggle, probs = c(0.05, 0.5, 0.95))
+qwiggle <- quantile(varshape.out$prop.wiggle, probs = c(0.05, 0.5, 0.95))
 ```
@@
 resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
+# rename colonies
+resp.TPM.dt.sub$colony2 <- ifelse(resp.TPM.dt.sub$colony == "A22", "ApVT", "AcNC")
 ```
@@
 
-intersect(A22.low.TPM.dt.sub, A22.high.TPM.dt.sub)
-# Good - no overlap
-
 # make data.frame for results
@@
 
-intersect(A22.low.TPM.dt.sub, A22.high.TPM.dt.sub)
-# Good - no overlap
-
 l4 <- length(unique(Ar.low.TPM.dt.sub$Transcript))
@@
 
-Genes with increased expression at *Low* temperatures are turned on at lower temperatures in *AcNC* than *ApVT*.
+Genes with increased expression at *High* temperatures are turned on at, on average, 1°C higher in *ApVT* than *AcNC*.
 
@@
 
-### Identification of biological-functions associated with temperature responses
-
-## Functional annotation
+## Gene set enrichment analysis for thermal-responsive genes
+
+### Functional annotation
 
@@
 
-```{r echo=FALSE, results='hide'}
+```{r ann_responsive, echo=TRUE}
 responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
 # combine responsive.lms.ann with "type" information
-responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "Ar.type", "Ar.type")], by = "Transcript", all = TRUE)
+responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
 str(responsive.lms.ann.type)
@@
 ```
-
-## Gene set enrichment analysis ##
 
@@
 
-Ap.BP.GOdata
+#Ap.BP.GOdata
 
@@
 # table results
-Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 137)
-dim(Ap.BP.ResTable)
-#pandoc.table(Ap.BP.ResTable)
-
-# graph significant nodes
-
-#pdf(paste(resultsdir, "Ap.BP_topGO_sig_nodes.pdf", sep=""))
-#showSigOfNodes(Ap.BP.GOdata, score(Ap.BP.resultParentChild), firstSigNodes = 10, useInfo = 'all')
-#dev.off()
+Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 10)
+Ap.BP.ResTable
 ```
@@
 
-```{r gsea_high}
+```{r gsea_high, results = 'hide'}
 # Select transcripts
@@
 
-```{r gsea_low}
+```{r gsea_low, results = 'hide'}
 # Select transcripts
@@
 
-```{r gsea_bimodal}
+```{r gsea_bimodal, results = 'hide'}
 Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
@@
 
-```{r gsea_intermediate}
+```{r gsea_intermediate, results = 'hide'}
 Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
@@
 
-```{r gsea_high_A22}
+```{r gsea_high_A22, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_high_Ar}
+```{r gsea_high_Ar, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_low_A22}
+```{r gsea_low_A22, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_low_Ar}
+```{r gsea_low_Ar, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_bim_A22}
+```{r gsea_bim_A22, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_bim_Ar}
+```{r gsea_bim_Ar, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_int_A22}
+```{r gsea_int_A22, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_int_Ar}
+```{r gsea_int_Ar, results = 'hide'}
 # create gene list, setting value to 1 for "bim" transcripts
@@
   geom_line() + 
-  facet_grid(. ~ colony) + 
+  facet_grid(. ~ colony2) + 
   scale_y_continuous(name="Expression (scaled)") +
@@
   geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
-  facet_grid(. ~ colony) + 
+  facet_grid(. ~ colony2) + 
   scale_y_continuous(name="Expression (scaled)") +
@@
 print(trp4)
+```
+
+```{r ggplot_high}
+trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high], aes(x=val, y=TPM.scaled, group=Transcript)) + 
+  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
+  facet_grid(. ~ colony2) + 
+  scale_y_continuous(name="Expression (scaled)") +
+  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
+print(trp_A22_high)
+
 ```
@@
 # save RData for manuscript file
-save(tt2, Ap.df, w1, w2, Ap.T_on, file = "ms.RData")
+save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.offset, t.slope, t.curvature, t.wiggle, file = "ms.RData")
 ```

### 2014-05-15 doc/ApTranscriptome_manuscript.Rmd

--- 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
@@
 
-We identified 22,399 genes with overall model fits of P < 0.05, retaining 10,597 after correcting for FDR.
-Of these transcripts, 9,156 significantly change expression with temperature including 6,908 that have a species x temperature interaction.
-The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1).
+We identified `r length(which(RxNpval$pval < 0.05))` genes with overall model fits of *P* < 0.05, retaining `r length(which(RxNpval$padj < 0.05))` after adjusting *P* values for false discovery rate.
+Of these genes, `r sum(sigtable[sigtable$Coefficient %in% c("Temperature, Temperature:Colony"), "Number.significant"])` significantly change expression with temperature including `r sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"])` that have a species x temperature interaction.
+The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.0001)`; Table 1).
 For both species, the majority of responsive transcripts were in the *Low* category (Fig.

### 2014-05-15 doc/ApTranscriptome_manuscript.Rmd
Update to manuscript

--- 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
@@
 carolinensis* that are up-regulated to high temperatures, while *A.
-carolinensis* would have higher constitutive expression of these thermally-tolerant genes.
+carolinensis* would have higher expression of these thermally-tolerant genes near optimal temperatures.
 Reciprocally, we expected *A.
 carolinensis* to have greater up-regulation of genes in response to cold temperatures, while *A.
-picea* would have greater constitutive expression of cold-tolerant genes.
+picea* would have greater basal expression of cold-tolerant genes.
 Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures.
@@
 
-To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction.
+To gain greater insight on the ecological genetics of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction.
 For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Responsive*.
@@
 carolinensis* species or hot temperatures in the northern *A.
-picea* species had higher constitutive expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.
+picea* species had higher basal expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.
 
@@
 load("../ms.RData")
+library(data.table)
 library(pander)
 library(ggplot2)
+library(topGO)
 ```
@@
 DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts.
-Of these, 47% have best hits to genes from the class Insecta [Supplementary Krona plot](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
+Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
 
@@
 
-We identified 22,399 genes with overall model fits of P < 0.05, retaining 10,597 after correcting for FDR.
-Of these transcripts, 9,156 significantly change expression with temperature including 6,908 that have a species x temperature interaction.
-The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1).
+We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate.
+Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction.
+The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1).
 For both species, the majority of responsive transcripts were in the *Low* category (Fig.
@@
 ```{r expression_type_table, echo=FALSE, results='asis'}
-pandoc.table(tt2, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species.")
+pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species. *AcNC* in rows and *ApVT* in columns.")
 ```
@@
 ```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
-mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
+mosaicplot(t(tt2), shade = FALSE, ylab = "AcNC", xlab = "ApVT", main = "", cex = 0.5)
 ```
 
-As the molecular pathways involved in thermal response to cold and hot temperatures may differ, we performed gene set enrichment analysis on each expression category separately.
-Out of a total of 2,859 GO terms scored, we found 6 enriched (*P* < 0.01) for *High* transcripts, included categories implicated in stress response such as "gene expression" and "tissue regeneration" (Table S1).
-For *Low* genes, 60 GO terms were enriched including some implicated in stress response (e.g.
-detection of external stimulus; Table S1).
-For *Bimodal* genes, 11 GO terms were enriched including 10 "regulation" terms (e.g.
-regulation of primary metabolic process) indicating that these are general genes involved in activating responses to change (Table S1).
-Finally, for *Intermediate* genes 6 GO terms were enriched, all some type of metabolic or biosynthetic process (Table S1) consistent with the expectation that these are non-essential genes expressed in optimal conditions that are shut-down under thermal stress.
+As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately.
+Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "High"), ])` enriched (*P* < 0.01) for *High* transcripts, including categories such as `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][1]` and `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][6]` (Table S1).
+For *Low* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Low"), ])` GO terms were enriched including some implicated in stress response (e.g.
+`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][18]`) and cold tolerance (e.g.
+`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][6]`) (Table S1).
+For *Bimodal* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g.
+regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1).
+Finally, for *Intermediate* genes `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).
+
+In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Figure 2).
+Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
+
+```{r rxn_plot, echo = FALSE, warning = FALSE, fig.cap = "Gene expression in transcripts per million (TPM) for thermally-reactive genes in the GO category *response to stress* for each colony"}
+rxnplot <- ggplot(resp.TPM.dt.sub[GO0006950.responsive$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) +
+  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
+  facet_grid(. ~ colony2) +
+  scale_y_continuous(name = "Expression (scaled)") +
+  scale_x_continuous(name = expression(paste("Temperature", degree, "C")))
+print(rxnplot)
+```
+
+As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
+
+### Species-level comparisons
+
+We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$).
+Overall, we found mean expression was greater in *AcNC* (*t*~`r t.offset$parameter`~ = `r t.offset$statistic`, *P* < `r max(t.offset$p.value, 0.001)`),
+
+but changes in the shape due to slope , curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant.
+On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A.
+picea* than *A.
+carolinensis*, indicating a steeper response to temperature.
 
 ```{r exp_plot, echo=FALSE, warning = FALSE, fig.cap = "Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
-
-ggplot(Ap.df, aes(x=max.val, fill=colony)) +
+ggplot(Ap.df, aes(x=max.val, fill=colony, colour = colony)) +
   geom_density(alpha = 0.2, position = "identity") +
@@
   scale_x_continuous(name = "Temperature of maximum expression")
-
 ```
-
-As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
-
-In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses....
-
-### Species-level comparisons
-
-We quantified differences in the reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$).
-Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant.
-On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A.
-picea* than *A.
-carolinensis*, indicating a steeper response to temperature.
 
@@
     geom_histogram(position = "dodge", binwidth = 3) +
-    geom_density(adjust = 3, fill = NA, size = 2) + 
+    geom_density(alpha = 0.2, adjust = 3) + 
     scale_y_continuous(name="Density") +

### 2014-05-16 ApTranscriptome_TR.Rmd
Cleaning up TR

@@
 
-**May 14, 2014**
+**May 16, 2014**
 
@@
 carolinensis*.
+For historical reasons, I refer to these species as colonies at times throughout this technical report.
 
@@
 
-Once this is done, quantify expression for the Trimmomatic filtered reads from each colony-treatment sample separately. Note that for each sample, there are three four filtered read files:
+Once this is done, quantify expression for the Trimmomatic filtered reads from each species-treatment sample separately. Note that for each sample, there are three four filtered read files:
 
@@
 
-The TPM column for each sample was extracted and combined into a matrix for each colony.
+The TPM column for each sample was extracted and combined into a matrix for each species.
 
@@
 
-DeconSeq identified contaminant genes. I remove this n
+With gene expression quantified using all reads, I now remove the transcripts identified as contaminants by DeconSeq.
 
@@
 
-$$ log(TPM + 1) = \beta_0 + \beta_1(colony) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(colony * temp) + \beta_5(colony * temp^2) + \epsilon $$
+$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$
 
@@
 
-The set of transcripts with significant expression patterns include those with expression that differs by colony, temperature and the interaction of colony and temperature.
-In this section, I am specifically interested in the thermally-responsive transcripts (temperature and colony x temperature) so I subset the significant transcripts to examine these.
+The set of transcripts with significant expression patterns include those with expression that differs by species, temperature and the interaction of species and temperature.
+In this section, I am specifically interested in the thermally-responsive transcripts (temperature and species x temperature) so I subset the significant transcripts to examine these.
 
@@
 
-pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts of 99,861 total with expression that depends on colony, temperature or their interaction at 5% FDR.")
+pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts of 99,861 total with expression that depends on species, temperature or their interaction at 5% FDR.")
 ```
@@
 
-I do this first for the thermally-responsive transcripts where there is no interaction with colony.
-For the transcripts where thermal-responsive expression depends on colony, I determine the functional type of the expression response separately for each colony.
+I do this first for the thermally-responsive transcripts where there is no interaction with species.
+For the transcripts where thermal-responsive expression depends on species, I determine the functional type of the expression response separately for each species.
 
@@
 
-pandoc.table(tt3, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.", split.tables = "Inf")
-```
-
-Table 4 shows the number of transcripts that fall into each expression type for each each colony.
-The totals for each colony include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
-
-## Colony-level comparison ##
-
-In this section, I compare the thermal reactionome between the *Ar* and *A22* colonies.
-
-### Plasticity versus basal expression
-
-Selection may have acted in response to thermal stress such that some genes are basaly activated in one colony while plastically-expressed in the other.
-I evaluate this hypothesis by comparing expression levels at the at optimum (19.5C) temperature between the two colonies for genes in that are either in the 'High' or 'Low' expression group in the other colony.
-Specifically, I expect that genes upregulated at high temperatures in *A22* are more highly expressed at 19.5C in *Ar*?
+pandoc.table(tt3, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.", split.tables = "Inf")
+```
+
+Table 4 shows the number of transcripts that fall into each expression type for each each species.
+The totals for each species include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
+
+## Species comparison ##
+
+In this section, I perform a number of comparisons of the thermal reactionomes between the *Ar* and *A22* species. Specifically, I:
+
+- compare profiles of the temperature of maximum expression for thermally-responsive transcripts 
+- compare basal expression at optimal temperature for thermally-responsive genes
+- compare thermal stability of *Intermediate* genes
+- compare thermal sensitivity of *Bimodal* genes
+- compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes
+- evaluate the extent to which thermal reaction norms differ between species by mean and shape
+
+### Compare temperature of maximum expression between species
+
+Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*
+
+```{r max_exp_PDF}
+# reshape data
+Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
+
+maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
+  geom_density(alpha=0.2, position="identity") + 
+  #scale_fill_manual(name = "Colony", values = c("white", "black")) + 
+  scale_y_continuous(name="Density") +
+  scale_x_continuous(name="Temperature of maximum expression")
+suppressWarnings(print(maxexplot))
+```
+
+### Compare basal expression at optimal temperature for thermally-responsive genes
+
+Genes upregulated in response to thermal stress in one species may have greater basal levels of expression in the other species that experiences those stressful conditions more often.
+To test this hypothesis, we compared expression levels near the optimal temperature (19.5C) between the two species for genes in that are either in the 'High' or 'Low' expression group in the other species.
+Specifically, do genes upregulated at high temperatures in *A22* have greater basal expression at optimal temperatures in *Ar*?
 
@@
 This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`.
-As my goal is simply to determine if expression is typically greater at 19.25C in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test
+As my goal is simply to determine if expression is typically greater at optimal temperatures (19.5 C) in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test
 
@@
 
-Next I test the converse, that transcripts that are up-regulated at low temperatures in *Ar* are more highly-expressed at 19.25C temperatures in *A22*.
+Next I test the converse: do genes up-regulated at low temperatures in *Ar* have greater basal expression near optimal temperatures in *A22*?
 
@@
 
-Counter to expectations, expression at 19.5C is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*.
+Counter to expectations, expression at optimal temperatures is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*.
 
@@
 
-### Thermal tolerance indicated by region of constant gene expression ###
-
-The 'Intermediate' expressed transcripts are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermall stress.
-We hypothesized that if the more southern *Ar* colony was more thermally-tolerant than *A22*, transcripts with 'Intermediate' expression (10-30C) would be active across a wider range of temperatures.
-To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each colony.
+### Compare thermal stability of *Intermediate* genes
+
+*Intermediate* genes are those that have expression that is shut-down as conditions become stressful, likely non-essential molecular processes.
+We hypothesized that if the more southern *Ar* species was more thermally-tolerant than *A22*, transcripts with *Intermediate* expression (10-30C) would be active across a wider range of temperatures.
+To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each species.
 
 ```{r thermal_breadth}
-# extract 'Intermediate' expressed transcripts for A22 colony
+# extract 'Intermediate' expressed transcripts for A22
 A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
@@
 # T-test comparing standard deviation of expression between colonies
-(t1 <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
-```
-
-Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
+(t.thermbreadth <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
+```
+
+Consistent with our hypothesis, *Intermediate* transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
 
@@
 
-### Thermal sensitivity indicated by response of bimodally-expressed transcripts ###
-
-As the converse of the above hypothesis, a colony that is especially thermally-sensitive is likely to activate expression of molecular processes more quickly.
-We tested this using the same approach as for the 'Intermediate' transcripts, but using the inverse of the 'Bimodal' expressed transcripts.
+### Compare thermal sensitivity of *Bimodal* genes
+
+As the converse of the above hypothesis, a species that is especially thermally-sensitive is likely to activate expression of molecular processes for thermal tolerance more quickly.
+We tested this using the same approach as for the *Intermediate* transcripts, but using the inverse of the *Bimodal* expressed transcripts.
 
 ```{r calc_Bimodal_variance}
-# extract 'Bimodal' expressed transcripts for A22 colony
+# extract 'Bimodal' expressed transcripts for A22 species
 A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
@@
 # t-test to compare standard deviation of 'Bimodal' transcripts between colonies
-(t2 <- t.test(Ar.bim.sd, A22.bim.sd))
+(t.themsens <- t.test(Ar.bim.sd, A22.bim.sd))
 ```
@@
 
-### Compare peak expression among colonies ###
-
-Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*
-
-```{r max_exp_PDF}
-# reshape data
-Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
-
-maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
-  geom_density(alpha=0.2, position="identity") + 
-  #scale_fill_manual(name = "Colony", values = c("white", "black")) + 
-  scale_y_continuous(name="Density") +
-  scale_x_continuous(name="Temperature of maximum expression")
-suppressWarnings(print(maxexplot))
-```
-
-For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect?
-To test this, I use the framework of `r citet("10.1086/675302")`:
-
-- *Offset, O*: overall difference in the mean expression value across all temperatures
-- *Slope, S*: difference in overall slope
-- *Curvature, C*: average difference in curvature of the reaction norm
-- *Wiggle, W*: variability in shape not captured by the previous three measures
-
-```{r offset, echo=TRUE, eval=TRUE}
-varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)
-
-# plot
-boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
-boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))
-
-t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))
-
-# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
-# offset = mean1 - mean2
-varshape.out$offset <- varshape.out$A22.mean - varshape.out$Ar.mean
-
-# t-test
-(t.offset <- t.test(varshape.out$s.offset, alternative = "two.sided"))
-
-# slope = slope1 - slope2
-varshape.out$slope <- varshape.out$A22.slope - varshape.out$Ar.slope
-
-# t-test
-(t.slope <- t.test(varshape.out$s.slope, alternative = "two.sided"))
-
-# curvature = curvature1 - curvature2
-varshape.out$curvature <- varshape.out$A22.curve - varshape.out$Ar.curve
-# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
-varshape.out$s.curvature <- NA
-for(i in 1:nrow(varshape.out)) {
-  varshape.out[i, "s.curvature"] <- (varshape.out[i, "curvature"] / mean(c(varshape.out[i, "A22.curve"], varshape.out[i, "Ar.curve"])))
-}
-# change 'Inf' values to 0
-varshape.out[which(varshape.out$Ar.curve == 0 & varshape.out$A22.curve == 0), "s.curvature"] <- 0
-# t-test
-(t.curvature <- t.test(varshape.out$s.curvature, alternative = "two.sided"))
-
-# wiggle
-varshape.out$wiggle <- varshape.out$A22.wiggle - varshape.out$Ar.wiggle
-varshape.out$s.wiggle <- NA
-for(i in 1:nrow(varshape.out)) {
-  varshape.out[i, "s.wiggle"] <- (varshape.out[i, "wiggle"] / mean(c(varshape.out[i, "A22.wiggle"], varshape.out[i, "Ar.wiggle"])))
-}
-# change 'Inf' values to 0
-varshape.out[which(varshape.out$Ar.wiggle == 0 & varshape.out$A22.wiggle == 0), "s.wiggle"] <- 0
-# note small number of genes with 'wiggle'
-length(which(varshape.out$s.wiggle != 0))
-
-# t-test
-(t.wiggle <- t.test(varshape.out$s.wiggle, alternative = "two.sided"))
-```
-
-Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means (offset), slope, curvature and wiggle.
-
-```{r partition_reaction_norm_variation}
-# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
-varshape.out$s.offset <- NA
-for(i in 1:nrow(varshape.out)) {
-  varshape.out[i, "s.offset"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.mean"], varshape.out[i, "Ar.mean"])))
-}
-# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
-varshape.out$s.slope <- NA
-for(i in 1:nrow(varshape.out)) {
-  varshape.out[i, "s.slope"] <- (varshape.out[i, "slope"] / mean(c(varshape.out[i, "A22.slope"], varshape.out[i, "Ar.slope"])))
-}
-# for transcripts where both samples had slope=0, change s.slope from Inf to 0 as there is truly no difference in slope
-varshape.out[which(varshape.out$Ar.slope == 0 & varshape.out$A22.slope == 0), "s.slope"] <- 0
-
-                                        # weighted mean-standardized values of each measure
-mean(abs((varshape.out$s.offset)))
-mean(abs((varshape.out$s.slope)))
-mean(abs((varshape.out$s.curvature)))
-mean(abs((varshape.out$s.wiggle)))
-
-# take absolute value of each value and sum to get total differences between reaction norms
-varshape.out$s.total <- abs(varshape.out$s.offset) + abs(varshape.out$s.slope) + abs(varshape.out$s.curvature) + abs(varshape.out$s.wiggle)
-
-(t6 <- t.test(varshape.out$s.total, alternative = "two.sided"))
-
-# variation in reaction norms due to differences in mean
-varshape.out$prop.mean <- abs(varshape.out$s.offset) / varshape.out$s.total
-varshape.out$prop.slope <- abs(varshape.out$s.slope) / varshape.out$s.total
-varshape.out$prop.curve <- abs(varshape.out$s.curvature) / varshape.out$s.total
-varshape.out$prop.wiggle <- abs(varshape.out$s.wiggle) / varshape.out$s.total
-
-# Mean proportion and 95% CI of total variation of each measure
-mean(varshape.out$prop.mean)
-quantile(varshape.out$prop.mean, probs = c(0.05, 0.5, 0.95))
-
-mean(varshape.out$prop.slope)
-quantile(varshape.out$prop.slope, probs = c(0.05, 0.5, 0.95))
-
-mean(varshape.out$prop.curve)
-quantile(varshape.out$prop.curve, probs = c(0.05, 0.5, 0.95))
-
-mean(varshape.out$prop.wiggle)
-qwiggle <- quantile(varshape.out$prop.wiggle, probs = c(0.05, 0.5, 0.95))
-```
-
-## Temperature of gene activation
+### Compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes
 
@@
 
-Calculate T~on~ for *High* genes in each colony.
+Calculate T~on~ for *High* genes in each species.
 
@@
 }
-```
-
-Determine if T~on~ is greater in *A22* or *Ar* for *High* genes.
-
-```{r T_on_high_ttest}
+
+# determine if T~on~ is greater in *A22* or *Ar* for *High* genes.
+
 t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
@@
 
-Genes with increased expression at *High* temperatures are turned on at, on average, 1°C higher in *AcNC* than *ApVT*.
+Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.
 
@@
 
-Genes with increased expression at *High* temperatures are turned on at, on average, 1°C higher in *ApVT* than *AcNC*.
+Genes with increased expression at *Low* temperatures are on average turned on 1°C higher in *ApVT* than *AcNC*.
 
@@
 ```
+
+### Evaluate the extent to which differences in thermal reaction norms are due to mean or shape
+
+For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect?
+To test this, I rougly follow `r citet("10.1086/675302")` by defining differences among reation norms for individual genes due to changes in the overall mean, slope, curvature and all higher-order shape differences (i.e.
+wiggle).
+
+- *Mean, M*: overall difference in the mean expression value across all temperatures
+- *Slope, S*: difference in overall slope
+- *Curvature, C*: average difference in curvature of the reaction norm
+- *Wiggle, W*: variability in shape not captured by the previous three measures
+
+```{r varshape, echo=TRUE, eval=TRUE}
+varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)
+
+# delta mean
+boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
+boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))
+
+t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean), paired = TRUE)
+t.mean
+
+# delta slope
+boxplot(log(varshape.out$A22.slope), log(varshape.out$Ar.slope))
+t.slope <- t.test(log(varshape.out$A22.slope + 0.1), log(varshape.out$Ar.slope + 0.1), paired = TRUE)
+t.slope
+
+# delta curvature
+boxplot(log(varshape.out$A22.curve), log(varshape.out$Ar.curve))
+t.curvature <- t.test(log(varshape.out$A22.curve + 0.1), log(varshape.out$Ar.curve + 0.1), paired = TRUE)
+t.curvature
+
+# delta wiggle - small values so use un-transformed
+boxplot(varshape.out$A22.wiggle, varshape.out$Ar.wiggle)
+t.wiggle <- t.test(varshape.out$A22.wiggle, varshape.out$Ar.wiggle, paired = TRUE)
+t.wiggle
+```
+
+Reaction norms differ between species by mean and slope, but not by curvature or wiggle.
+
+Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means and slope, as curvature and wiggle did not differ between species.
+To do this, I first standardize each of these parameters so they are directly comparable.
+
+```{r partition_reaction_norm_variation}
+# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
+# mean = mean1 - mean2
+varshape.out$lmean <- log(varshape.out$A22.mean) - log(varshape.out$Ar.mean)
+
+# slope = slope1 - slope2
+varshape.out$lslope <- log(varshape.out$A22.slope + 0.1) - log(varshape.out$Ar.slope + 0.1)
+
+# take absolute value of each value and sum to get total differences between reaction norms
+varshape.out$ltotal <- abs(varshape.out$lmean) + abs(varshape.out$lslope)
+
+# variation in reaction norms due to differences in mean
+varshape.out$prop.lmean <- abs(varshape.out$lmean) / varshape.out$ltotal
+varshape.out$prop.lslope <- abs(varshape.out$lslope) / varshape.out$ltotal
+
+# Mean proportion and 95% CI of total variation of each measure
+mean(varshape.out$prop.lmean)
+(qlmean <- quantile(varshape.out$prop.lmean, probs = c(0.05, 0.5, 0.95)))
+
+mean(varshape.out$prop.lslope)
+(qlslope <- quantile(varshape.out$prop.lslope, probs = c(0.05, 0.5, 0.95)))
+```
+
+From this analysis, about 3/4 of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.
 
@@
 
+Perform GSEA for all thermally-responsive transcripts.
+
 ```{r gsea_all, results='hide'}
@@
 
-**2) High**
-
-Genes with *High* expression in both colonies
-
-Use `selectTranscript` function to select transcripts from 'Ap.response.type' for GSEA.
-
-```{r gsea_high, results = 'hide'}
-# Select transcripts
+As the molecular processes involved in cold and hot temperature tolerance are different, I perform GSEA separately for each response type and compile results in a single table.
+
+```{r gsea_by_type, results = 'hide'}
+## Genes with *High* expression in both colonies
 Ap.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High" & Ap.response.type$A22.type == "High"), "Transcript"]
@@
 table(Ap.geneList.high)
-
-# Run GSEA
+# run GSEA
 Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)
-#pandoc.table(Ap.high.gsea)
-# output for Revigo
-write.table(Ap.high.gsea[,"GO.ID"], file = "results/Ap_high_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
-```
-
-**3) Low**
-
-Genes with *Low* expression in both colonies
-
-```{r gsea_low, results = 'hide'}
-# Select transcripts
+
+## Genes with *Low* expression in both colonies
 Ap.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low" & Ap.response.type$A22.type == "Low"), "Transcript"]
@@
 table(Ap.geneList.low)
-
 # Run GSEA
 Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
-#pandoc.table(Ap.low.BP.ResTable)
-write.table(Ap.low.gsea[,"GO.ID"], file = "results/Ap_low_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
-```
-
-**4) Bimodal**
-
-Genes with *Bimodal* expression in both colonies
-
-```{r gsea_bimodal, results = 'hide'}
+
+## Genes with *Bimodal* expression in both colonies
 Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
@@
 table(Ap.geneList.bim)
-
 # Run GSEA
 Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
-#pandoc.table(Ap.bim.gsea)
-write.table(Ap.bim.gsea[,"GO.ID"], file = "results/Ap_bim_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
-```
-
-**5) Intermediate**
-
-Genes with *Intermediate* expression in both colonies
-
-```{r gsea_intermediate, results = 'hide'}
+
+## Genes with *Intermediate* expression in both colonies
 Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
@@
 table(Ap.geneList.int)
-
 # Run GSEA
 Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
-#pandoc.table(Ap.int.gsea)
-write.table(Ap.int.gsea[,"GO.ID"], file = "results/Ap_int_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
-```
-
-```{r gsea_table, echo=FALSE, eval=TRUE, results='asis'}
+
 ## merge results into single table
@@
 colnames(Ap.gsea1)[7] <- "P"
-# display
-Ap.gsea1[ , 1:3]
-```
-
-Next, I perform GSEA for genes in each functional type in one colony but not the other (e.g. the set difference) to gain insight on differences between the colonies.
-
-**A22 'High' genes not in Ar**
-
-```{r gsea_high_A22, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+# write to file
+write.csv(Ap.setdiff.gsea, file = paste(resultsdir, "Ap_setdiff_GSEA_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
+```
+
+```{r gsea_by_type_table, echo = FALSE, results = 'asis'}
+pandoc.table(Ap.gsea1[, 1:3], caption = "Enriched GO terms for each thermal-response category")
+#print(xtable(Ap.gsea1[ , 1:3], comment = getOption("xtable.comment", FALSE)))
+```
+
+Next, I perform GSEA for genes in each functional type in one species but not the other (e.g. the set difference) to gain insight on differences between the colonies.
+
+```{r gsea_setdiff, results = 'hide'}
+# A22 'High' genes not in Ar**
 A22.geneList.high <- rep(0, length = length(Ap.geneList))
@@
 A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
-# check correct number of values set to 1
-table(A22.geneList.high)
-
-# Run GSEA
 A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
-#pandoc.table(A22.high.gsea, split.table = Inf)
-```
-
-**Ar 'High' genes not in A22**
-
-```{r gsea_high_Ar, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+
+# *Ar 'High' genes not in A22**
 Ar.geneList.high <- rep(0, length = length(Ap.geneList))
@@
 Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high))] <- 1
-# check correct number of values set to 1
-table(Ar.geneList.high)
-
-# Run GSEA
 Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
-#pandoc.table(Ar.high.gsea, split.table = Inf)
-```
-
-**A22 'Low' genes not in Ar**
-
-```{r gsea_low_A22, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+
+# A22 'Low' genes not in Ar**
 A22.geneList.low <- rep(0, length = length(Ap.geneList))
@@
 A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low))] <- 1
-# check correct number of values set to 1
-table(A22.geneList.low)
-
-# Run GSEA
 A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
-#pandoc.table(A22.low.gsea, split.table = Inf)
-```
-
-**Ar 'Low' genes not in A22**
-
-```{r gsea_low_Ar, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+# Ar 'Low' genes not in A22**
 Ar.geneList.low <- rep(0, length = length(Ap.geneList))
@@
 Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low))] <- 1
-# check correct number of values set to 1
-table(Ar.geneList.low)
-
-# Run GSEA
 Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
-#pandoc.table(Ar.low.gsea, split.table = Inf)
-```
-
-**A22 'Bimodal' genes not in Ar**
-
-```{r gsea_bim_A22, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+
+# A22 'Bimodal' genes not in Ar**
 A22.geneList.bim <- rep(0, length = length(Ap.geneList))
@@
 A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim))] <- 1
-# check correct number of values set to 1
-table(A22.geneList.bim)
-
-# Run GSEA
 A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
-#pandoc.table(A22.bim.gsea, split.table = Inf)
-```
-
-**Ar 'Bimodal' genes not in A22**
-
-```{r gsea_bim_Ar, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+
+# Ar 'Bimodal' genes not in A22**
 Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
@@
 Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim))] <- 1
-# check correct number of values set to 1
-table(Ar.geneList.bim)
-
-# Run GSEA
 Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
-#pandoc.table(Ar.bim.gsea, split.table = Inf)
-```
-
-**A22 'Intermediate' genes not in Ar**
-
-```{r gsea_int_A22, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+
+# A22 'Intermediate' genes not in Ar**
 A22.geneList.int <- rep(0, length = length(Ap.geneList))
@@
 A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int))] <- 1
-# check correct number of values set to 1
 table(A22.geneList.int)
-
-# Run GSEA
 A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
-#pandoc.table(A22.int.gsea, split.table = Inf)
-```
-
-**Ar 'Intermediate' genes not in A22**
-
-```{r gsea_int_Ar, results = 'hide'}
-# create gene list, setting value to 1 for "bim" transcripts
+
+# Ar 'Intermediate' genes not in A22**
 Ar.geneList.int <- rep(0, length = length(Ap.geneList))
@@
 Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int))] <- 1
-# check correct number of values set to 1
-table(Ar.geneList.int)
-
-# Run GSEA
 Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
-#pandoc.table(Ar.int.gsea, split.table = Inf)
-```
-
-Table of overall GSEA results.
-
-```{r gsea_table_2, echo=FALSE, eval=TRUE, results='asis'}
+
+# combine into single table
 A22.high.gsea$Type <- "High"
@@
 A22.gsea <- rbind(A22.high.gsea, A22.low.gsea, A22.int.gsea, A22.bim.gsea)
-A22.gsea$Colony <- "ApVT"
+A22.gsea$Species <- "ApVT"
 
@@
 Ar.gsea <- rbind(Ar.high.gsea, Ar.low.gsea, Ar.int.gsea, Ar.bim.gsea)
-Ar.gsea$Colony <- "AcNC"
+Ar.gsea$Species <- "AcNC"
 
@@
 # reorder
-Ap.setdiff.gsea <- Ap.setdiff.gsea[,c("Colony", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
+Ap.setdiff.gsea <- Ap.setdiff.gsea[,c("Species", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
 colnames(Ap.setdiff.gsea)[8] <- "P"
 
-print(xtable(Ap.setdiff.gsea, comment = getOption("xtable.comment", FALSE)))
+write.csv(Ap.setdiff.gsea, file = paste(resultsdir, "Ap_setdiff_GSEA_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
+```
+
+```{r gsea_setdiff_table, echo = FALSE, results = 'asis'}
+pandoc.table(Ap.setdiff.gsea[ , 1:4], caption = "Enriched GO terms found in one species but not the other for each thermal-response category")
+#print(xtable(Ap.setdiff.gsea[ , 1:4], comment = getOption("xtable.comment", FALSE)))
 ```
@@
 
-# Line plot, expression against temp, faceted by colony
+# Line plot, expression against temp, faceted by species
 trp1 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
@@
 # save RData for manuscript file
-save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.offset, t.slope, t.curvature, t.wiggle, file = "ms.RData")
+save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.themsens, t.thermbreadth, file = "ms.RData")
 ```

### 2014-05-16 doc/ApTranscriptome_manuscript.Rmd

@@
 
-Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *AcNC* respectively.
+Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W).
 The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers.
@@
 ```{r expression_type_table, echo=FALSE, results='asis'}
-pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species. *AcNC* in rows and *ApVT* in columns.")
+pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species.
+*A.
+carolinensis* in rows and *A.
+picea* in columns.")
 ```
@@
 ```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
-mosaicplot(t(tt2), shade = FALSE, ylab = "AcNC", xlab = "ApVT", main = "", cex = 0.5)
+mosaicplot(t(tt2), shade = FALSE, ylab = "A. carolinensis", xlab = "A. picea", main = "", cex = 0.5)
 ```
@@
 We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$).
-Overall, we found mean expression was greater in *AcNC* (*t*~`r t.offset$parameter`~ = `r t.offset$statistic`, *P* < `r max(t.offset$p.value, 0.001)`),
-
-but changes in the shape due to slope , curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant.
-On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A.
-picea* than *A.
-carolinensis*, indicating a steeper response to temperature.
+Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate)` TPM greater in *A.
+carolinensis*.
+Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$slope)` $\Delta$TPM/degree greater in *A.
+picea*.
+Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r max(t.curvature$p.value, 0.001)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species.
+Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.
 
@@
 
-Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.
-
-Plasticity is a well-characterized mechanism by which organism tolerate environmental change.
-Given our sampling of a warmer (*A.
+By sampling a warmer (*A.
 carolinensis*) and colder (*A.
-picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater expression near optimal temperatures of 19°C in the warmer-climate species, and vice versa.
-Partially consistent with our prediction, we found that optimal-temperature expression was greater in the more southern *A.
+picea*) climate species, we were able to test if the same genes that are up-regulated in response to thermal stress have increased basal expression in the other species.
+Partially consistent with this prediction, basal expression was greater in the more southern *A.
 carolinensis* species for *High* genes in *A.
@@
 However, for genes with *Low* expression in *A.
-carolinensis* we also found that optimal-temperature expression was greater in *A.
+carolinensis* basal expression was also greater in *A.
 carolinensis* than *A.
 picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).
-Moreover, we also found that for genes with *Intermediate* expression in either colony, optimal-temperature expression was greater in *A.
-carolinensis*.
+Moreover, basal expression was also greater in *A.
+carolinensis* for genes with *Intermediate* expression in either colony.
 The overall conclusion from these analyses is that for thermally-responsive genes, the more southern species, *A.
@@
 
-We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* expressed genes.
+We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* and *Bimodal* expressed genes.
 Contrary to our expectations based on critical maximum and minimal temperatures, *A.
-carolinensis* had a transcriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
-In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
+carolinensis* had a broader thermal range of stable gene expression (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
+In contrast, there was no difference in thermal sensitivity among species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
 
 We also explored the extent to which the temperature at which thermally-responsive genes are activated differs between species.
-For genes that are upregulated at *High* temperatures, including both *High* and *Bimodal* genes, we found that ...
-In contrast, for genes that are upregulated at *Low* temperatures, including both *Low* and *Bimodal* genes, we found that...
+For genes that are upregulated at *High* temperatures, we found that *A.
+carolinensis* genes were activated on average 1°C hotter than *A.
+picea*.
+In contrast, for genes that are upregulated at *Low* temperaturese we found that...
 
@@
 carolinensis*.
+
+overall pattern *A. carolinensis*
+
+- more genes thermally-responsive
+- tolerates thermal stress by shutting down biological processes
+
+overall pattern *A. picea*
+
+- more *Low* genes
 

### 2014-05-16 ApTranscriptome_TR.Rmd

@@
 ```{r gsea_by_type_table, echo = FALSE, results = 'asis'}
-pandoc.table(Ap.gsea1[, 1:3], caption = "Enriched GO terms for each thermal-response category")
-#print(xtable(Ap.gsea1[ , 1:3], comment = getOption("xtable.comment", FALSE)))
+#pandoc.table(Ap.gsea1[, 1:3], caption = "Enriched GO terms for each thermal-response category")
+print(xtable(Ap.gsea1[ , 1:3], comment = getOption("xtable.comment", FALSE)))
 ```
@@
 ```{r gsea_setdiff_table, echo = FALSE, results = 'asis'}
-pandoc.table(Ap.setdiff.gsea[ , 1:4], caption = "Enriched GO terms found in one species but not the other for each thermal-response category")
-#print(xtable(Ap.setdiff.gsea[ , 1:4], comment = getOption("xtable.comment", FALSE)))
+#pandoc.table(Ap.setdiff.gsea[ , 1:4], caption = "Enriched GO terms found in one species but not the other for each thermal-response category")
+print(xtable(Ap.setdiff.gsea[ , 1:4], comment = getOption("xtable.comment", FALSE)))
 ```

### 2014-05-16 doc/ApTranscriptome_manuscript.Rmd
Updates to Results and starting on Discussion

@@
 
-### Thermally-responsive genes
-
 We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate.
@@
 
-**Table 1 and Figure 1 are redundant - which is preferable??**
-
-```{r expression_type_plot, echo=FALSE, fig.cap = "Mosaic plot that shows the proportion of thermally-responsive genes in each expression category for each species."}
-mosaicplot(t(tt2), shade = FALSE, ylab = "A. carolinensis", xlab = "A. picea", main = "", cex = 0.5)
-```
-
-As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately.
-Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "High"), ])` enriched (*P* < 0.01) for *High* transcripts, including categories such as `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][1]` and `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][6]` (Table S1).
-For *Low* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Low"), ])` GO terms were enriched including some implicated in stress response (e.g.
-`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][18]`) and cold tolerance (e.g.
-`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][6]`) (Table S1).
-For *Bimodal* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g.
-regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1).
-Finally, for *Intermediate* genes `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).
-
-In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Figure 2).
-Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
-
-```{r rxn_plot, echo = FALSE, warning = FALSE, fig.cap = "Gene expression in transcripts per million (TPM) for thermally-reactive genes in the GO category *response to stress* for each colony"}
-rxnplot <- ggplot(resp.TPM.dt.sub[GO0006950.responsive$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) +
-  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
-  facet_grid(. ~ colony2) +
-  scale_y_continuous(name = "Expression (scaled)") +
-  scale_x_continuous(name = expression(paste("Temperature", degree, "C")))
-print(rxnplot)
-```
-
-As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
-
-### Species-level comparisons
+We performed a number of comparisons to evaluate the extent to which the thermal reactionomes differed between *A.
+carolinensis* and *A.
+picea*.
+First, we found that the temperature of maximum expression for the thermally-responsive genes differed between species (Figure 3), with a similar number of genes with maximum expression at high temperatures, many more genes with maximum expression at intermediate temperatures in *A.
+carolinensis* and more genes with maximum expression at low temperatures in *A.
+picea*.
 
@@
 
+### Molecular comparisons
+
+As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately.
+Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "High"), ])` enriched (*P* < 0.01) for *High* transcripts, including categories such as `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][1]` and `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][6]` (Table S1).
+For *Low* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Low"), ])` GO terms were enriched including some implicated in stress response (e.g.
+`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][18]`) and cold tolerance (e.g.
+`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][6]`) (Table S1).
+For *Bimodal* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g.
+regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1).
+Finally, for *Intermediate* genes `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).
+
+In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Figure 2).
+Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
+
+```{r rxn_plot, echo = FALSE, warning = FALSE, fig.cap = "Gene expression in transcripts per million (TPM) for thermally-reactive genes in the GO category *response to stress* for each colony"}
+rxnplot <- ggplot(resp.TPM.dt.sub[GO0006950.responsive$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) +
+  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
+  facet_grid(. ~ colony2) +
+  scale_y_continuous(name = "Expression (scaled)") +
+  scale_x_continuous(name = expression(paste("Temperature", degree, "C")))
+print(rxnplot)
+```
+
+As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
+
 We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock.
@@
 ## Discussion
+
+- greater adaptability to cold than hot thermal stress
+    - *A. carolinensis* limited future potential to adapt to heat stress
+- different mechanisms of thermal tolernace
+    - *A. carolinensis* tolerates thermal stress by shutting down (*Intermediate* genes)
+    - *A. picea* tolerates by a combination of shutting down and activating 'protective' genes 
+       - some indication that *A.
+picea* maintains express in cold conditions that *A.
+carolinensis* shuts down based on overlap of *Low* genes in *A.
+picea* with *Intermediate* genes in *A.
+carolinensis*
+- metabolic compensation?
+
+-------------------------------------
 

### 2014-05-19 doc/ApTranscriptome_manuscript.Rmd
Edits

@@
 
-We performed a number of comparisons to evaluate the extent to which the thermal reactionomes differed between *A.
+We performed a number of additional comparisons to evaluate the extent to which the thermal reactionomes differed between *A.
 carolinensis* and *A.
 picea*.
-First, we found that the temperature of maximum expression for the thermally-responsive genes differed between species (Figure 3), with a similar number of genes with maximum expression at high temperatures, many more genes with maximum expression at intermediate temperatures in *A.
-carolinensis* and more genes with maximum expression at low temperatures in *A.
+First, we found that the profile of the temperature of maximum expression for the thermally-responsive genes differed between species (Figure 3) with a similar number of genes at high temperatures, more genes at intermediate temperatures in *A.
+carolinensis* and more genes at low temperatures in *A.
 picea*.
-
-We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$).
-Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate)` TPM greater in *A.
-carolinensis*.
-Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$slope)` $\Delta$TPM/degree greater in *A.
-picea*.
-Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r max(t.curvature$p.value, 0.001)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species.
-Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.
 
@@
 
-By sampling a warmer (*A.
-carolinensis*) and colder (*A.
-picea*) climate species, we were able to test if the same genes that are up-regulated in response to thermal stress have increased basal expression in the other species.
-Partially consistent with this prediction, basal expression was greater in the more southern *A.
-carolinensis* species for *High* genes in *A.
-picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`).
-However, for genes with *Low* expression in *A.
-carolinensis* basal expression was also greater in *A.
-carolinensis* than *A.
-picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).
-Moreover, basal expression was also greater in *A.
-carolinensis* for genes with *Intermediate* expression in either colony.
-The overall conclusion from these analyses is that for thermally-responsive genes, the more southern species, *A.
-carolinensis* has greater optimal-temperature expression.
+We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$).
+Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` TPM greater in *A.
+carolinensis*.
+Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A.
+picea*.
+Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species.
+Of the total differences ($\Delta T$) in reaction norms between species for each gene, about ...% (95% CI ...
+- ...%) was due to shifts in mean and ...% (95% CI ...
+- ...%) due to shifts in slope.
 
-We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* and *Bimodal* expressed genes.
-Contrary to our expectations based on critical maximum and minimal temperatures, *A.
+We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species.
+Instead, we found that for *High* genes in *A.
+picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`), *Low* genes in *A.
+carolinensis* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`) and *Intermediate* genes in both (Wilcoxon V = `r w3$statistic`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r w3$conf.int[1:2]`), basal expression near optimum temperatures of 19.5C was always higher in *A.
+carolinensis*.
+
+Sensitivity of thermally-responsive genes is indicated by the thermal range of *Bimodal* expressed genes.
+
+We tested for a transcriptome-wide pattern of thermal sensitivity and tolerance by examining the thermal range for the *Bimodal* and *Intermediate* expressed genes.
+We found no in thermal sensitivity between species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
+However, for *Intermediate* genes, *A.
 carolinensis* had a broader thermal range of stable gene expression (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
-In contrast, there was no difference in thermal sensitivity among species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
 

### 2014-05-19 ApTranscriptome_TR.Rmd
working on partitioning true 'Low' from 'Off high'

@@
 
-## Species comparison ##
+## Species comparison
 
@@
 wilcox.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+
+# list of transcripts that are 'Intermediate' expressed in either colony
+Ap.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" | Ap.response.type$A22.type == "Intermediate"), ]
+# T test
+t.test(log(Ap.int.transcripts$A22.opt+1), log(Ap.int.transcripts$Ar.opt+1))
+# Wilcoxon signed rank-test
+w3 <- wilcox.test(Ap.int.transcripts$A22.opt, Ap.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+w3
 ```
@@
 From this analysis, about 3/4 of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.
+
+### Overlap between *Ar* Intermediate and *A22* Low genes
+
+```{r gene_overlap}
+# transcripts that are low in A22 and intermediate in Ar
+A22.low.Ar.int <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+A22.low.Ar.int.lms <- responsive.lms[A22.low.Ar.int]
+
+# transcripts that are low in both A22 and Ar
+A22.low.Ar.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
+A22.low.Ar.low.lms <- responsive.lms[A22.low.Ar.low]
+
+head(int.low.overlap)
+
+# compare linear and quadratic model terms
+
+# transcripts where TPM increases at low temps & decreases at high temps
+A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low,Ar.int), "A22")]
+
+for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
+    s <- A22.low.Ar.int.TPM.dt.sub[i]
+    X11()
+    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
+    points(s$val, s$TPM, pch = 8, col = "red")
+    if(min(s[which(s$val < 30) , pTPM]) < s[which(s$val == 38.5), pTPM]) print("On low") else {
+        print("Off high")
+    }
+}
+
+# A22 low Ar int
+A22.low.Ar.int.low.type <- ddply(A22.low.Ar.int.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))
+
+table(A22.low.Ar.int.low.type$lt)
+
+# A22 low Ar low
+A22.low.Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low, Ar.low), "A22")]
+
+# transcripts where TPM increases at low temps & decreases at high temps
+
+A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low, Ar.int), "A22")]
+dim(A22.low.Ar.int.TPM.dt.sub)
+
+l5 <- length(unique(A22.low.Ar.int.TPM.dt.sub$Transcript))
+
+A22.low.Ar.int.T_on <- data.frame(Transcript = unique(A22.low.Ar.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l5), type = rep("High", length = l5), T_on = NA)
+
+# of these, how many have greatest change in expression as a decrease at high temperatures?
+for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)) {
+    subdf <- A22.low.Ar.int.TPM.dt.sub[i]
+    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
+    A22.low.Ar.int.T_on[which(A22.low.Ar.int.T_on$Transcript == i), "T_on"] <- T_on
+}
+
+```
 
@@
 # save RData for manuscript file
-save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.themsens, t.thermbreadth, file = "ms.RData")
+save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.themsens, t.thermbreadth, file = "ms.RData")
 ```

### 2014-05-20 ApTranscriptome_TR.Rmd
Updates to TR

@@
 
-**May 16, 2014**
+**May 20, 2014**
 
@@
 # from either AWS or GoogleDrive
-annotationURL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
+annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
 #a2 <- getURL("https://googledrive.com/host/0B75IymziRJ_9Tlg1U1Vxbjk1bzg") # GoogleDrive link
 
-annotationfile <- read.csv(textConnection(annotationURL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
-str(annotationfile)
+annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
+str(annotation.file)
 
 # Convert to data.table
-annotationtable <- data.table(annotationfile)
-head(annotationtable)
+annotation.table <- data.table(annotation.file)
+head(annotation.table)
 ```
@@
 
-```{r extractionTranscripts}
-A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
-Ar.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]
-
-A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
-Ar.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
-
-A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
-Ar.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
-
-A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
-Ar.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+```{r extract_Transcripts}
+A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
+Ar.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]
+
+A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
+Ar.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]
+
+A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
+Ar.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+
+A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
+Ar.int <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
 ```
@@
 
-t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
-```
-
-Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.
+(T_on.high.ttest <- t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on))
+```
+
+Genes with increased expression at *High* temperatures are on average turned on 2.5°C higher in *ApVT* than *AcNC*.
 
@@
 # transcripts expressed at *Low* temperatures in A22
-A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.low, "A22")]
+A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.low, A22.bim), "A22")]
+setkey(A22.low.TPM.dt.sub, Transcript)
 str(A22.low.TPM.dt.sub)
@@
 # loop across transcripts, calculating T_on
-
 for(i in unique(A22.low.TPM.dt.sub$Transcript)) {
@@
 # repeat for Ar
-Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.low, "Ar")]
+Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.low, Ar.bim), "A22")]
+setkey(Ar.low.TPM.dt.sub, Transcript)
 str(Ar.low.TPM.dt.sub)
@@
 
-t.test(Ar.low.T_on$T_on, A22.low.T_on$T_on)
+(T_on.low.ttest <- t.test(Ar.low.T_on$T_on, A22.low.T_on$T_on))
 ```
@@
 
-### Overlap between *Ar* Intermediate and *A22* Low genes
-
-```{r gene_overlap}
-# transcripts that are low in A22 and intermediate in Ar
-A22.low.Ar.int <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
-A22.low.Ar.int.lms <- responsive.lms[A22.low.Ar.int]
-
-# transcripts that are low in both A22 and Ar
-A22.low.Ar.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
-A22.low.Ar.low.lms <- responsive.lms[A22.low.Ar.low]
-
-head(int.low.overlap)
-
-# compare linear and quadratic model terms
-
-# transcripts where TPM increases at low temps & decreases at high temps
-A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low,Ar.int), "A22")]
-
-for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
-    s <- A22.low.Ar.int.TPM.dt.sub[i]
-    X11()
-    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
-    points(s$val, s$TPM, pch = 8, col = "red")
-    if(min(s[which(s$val < 30) , pTPM]) < s[which(s$val == 38.5), pTPM]) print("On low") else {
-        print("Off high")
-    }
-}
-
-# A22 low Ar int
-A22.low.Ar.int.low.type <- ddply(A22.low.Ar.int.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))
-
-table(A22.low.Ar.int.low.type$lt)
-
-# A22 low Ar low
-A22.low.Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low, Ar.low), "A22")]
-
-# transcripts where TPM increases at low temps & decreases at high temps
-
-A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low, Ar.int), "A22")]
-dim(A22.low.Ar.int.TPM.dt.sub)
-
-l5 <- length(unique(A22.low.Ar.int.TPM.dt.sub$Transcript))
-
-A22.low.Ar.int.T_on <- data.frame(Transcript = unique(A22.low.Ar.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l5), type = rep("High", length = l5), T_on = NA)
-
-# of these, how many have greatest change in expression as a decrease at high temperatures?
-for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)) {
-    subdf <- A22.low.Ar.int.TPM.dt.sub[i]
-    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
-    A22.low.Ar.int.T_on[which(A22.low.Ar.int.T_on$Transcript == i), "T_on"] <- T_on
-}
-
-```
-
 ## Gene set enrichment analysis for thermal-responsive genes
@@
 ```{r annotation}
-setkey(annotationtable, Sequence.Name)
+setkey(annotation.table, Sequence.Name)
 signif.transcripts <- data.table(signif.transcripts)
 setkey(signif.transcripts, Transcript)
-signif.transcripts <- annotationtable[signif.transcripts]
+signif.transcripts <- annotation.table[signif.transcripts]
 setnames(signif.transcripts, "Sequence.Name", "Transcript")
@@
 # GO 'response to stress' hits in all transcripts
-resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
-
-GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.lms.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)
+resp.stress.all.count <- nrow(annotation.table[grep("GO:0006950", annotation.table$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])
+
+GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.lms.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotation.table) - resp.stress.all.count), nrow = 2)
 
@@
 
-hsp_all <- annotationtable[grep("shock", annotationtable$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
+hsp_all <- annotation.table[grep("shock", annotation.table$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
 nrow(hsp_all)
@@
 ```{r geneid2go, echo=TRUE, eval=TRUE, cache=TRUE}
-# create geneid2go.map file from FastAnnotator AnnotationTable.txt
-geneid2GOmap(annotationfile)
+# create geneid2go.map file from FastAnnotator annotationtable.txt
+geneid2GOmap(annotation.file)
 ```
@@
 
+```{r extract_Transcripts}
+A22.only.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
+Ar.only.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]
+
+A22.only.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
+Ar.only.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
+
+A22.only.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
+Ar.only.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+
+A22.only.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
+Ar.only.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+```
+
 ```{r gsea_setdiff, results = 'hide'}
@@
 names(A22.geneList.high) <- names(Ap.geneList)
-A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
+A22.geneList.high[(which(names(A22.geneList.high) %in% A22.only.high))] <- 1
 A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
@@
 names(Ar.geneList.high) <- names(Ap.geneList)
-Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high))] <- 1
+Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.only.high))] <- 1
 Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
@@
 names(A22.geneList.low) <- names(Ap.geneList)
-A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low))] <- 1
+A22.geneList.low[(which(names(A22.geneList.low) %in% A22.only.low))] <- 1
 A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
@@
 names(Ar.geneList.low) <- names(Ap.geneList)
-Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low))] <- 1
+Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.only.low))] <- 1
 Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
@@
 names(A22.geneList.bim) <- names(Ap.geneList)
-A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim))] <- 1
+A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.only.bim))] <- 1
 A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
@@
 names(Ar.geneList.bim) <- names(Ap.geneList)
-Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim))] <- 1
+Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.only.bim))] <- 1
 Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
@@
 names(A22.geneList.int) <- names(Ap.geneList)
-A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int))] <- 1
+A22.geneList.int[(which(names(A22.geneList.int) %in% A22.only.int))] <- 1
 table(A22.geneList.int)
@@
 names(Ar.geneList.int) <- names(Ap.geneList)
-Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int))] <- 1
+Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.only.int))] <- 1
 Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
@@
 print(trp_A22_high)
-
+```
+
+### Overlap between *Ar* Intermediate and *A22* Low genes
+
+An observation from GO analysis is that there is an overlap between enriched terms in the *Low* group for *A22* and the *Intermediate* group for *Ar*.
+A potential explanation for this pattern is if the *Low* genes in *A22* shut down at high temperatures, but not low temperatures, rather than are up-regulated at low temperatures.
+
+```{r gene_overlap}
+# transcripts where TPM increases at low temps & decreases at high temps
+A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low,Ar.int), "A22")]
+
+for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
+    s <- A22.low.Ar.int.TPM.dt.sub[i]
+    X11()
+    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
+    points(s$val, s$TPM, pch = 8, col = "red")
+    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh") else {
+        print("OnLow")
+    }
+}
+
+#    if(min(s[which(s$val < 30) , pTPM]) < s[which(s$val == 38.5), pTPM] & s[which(s$val == 0), pTPM] < (s[which(s$val == 21), pTPM] + sd(s$pTPM))) print("On low") else {
+
+# tabulate the number of transcripts that have lowest expression at maximum temperature "OffHigh"
+A22.low.Ar.int.low.type <- ddply(A22.low.Ar.int.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))
+
+table(A22.low.Ar.int.low.type$lt)
+
+# A22 low Ar low
+A22.low.Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low, Ar.low), "A22")]
+
+A22.low.Ar.low.low.type <- ddply(A22.low.Ar.low.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))
+
+# table results and perform chi-square test
+low.table <- rbind(LowInt = table(A22.low.Ar.int.low.type$lt), LowLow = table(A22.low.Ar.low.low.type$lt))
+low.table
+chisq.test(low.table)
 ```

### 2014-05-20 doc/ApTranscriptome_manuscript.Rmd
Edits to ms

@@
   Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
+csl: ecology.csl
 output:
+  word_document: default
   pdf_document:
@@
 
-When ants are exposed to different temperatures, lots of genes change expression.
+When ants are exposed to different temperatures, lots of genes change expression. The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress.
+
+Comparisons between species indicate that the species from the warmer environment tolerates thermal stress by higher baseline expression of protective genes
 
@@
 While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991].
-Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance.
-To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*.
+We compared the thermal reactionomes between two parapatric species of *Aphaenogaster*, *A.
+carolinensis* from North Carolina, USA and *A.
+picea* from Vermont, USA, to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance.
+Specifically, we examined differences between the species in (i) the number of thermally-responsive genes, (ii) the distribution of responsive genes into functional categories, (iii) basal expression near optimal temperatures for responsive genes and (iv) the thermal sensitivity of the transcriptome.
 *A.
 picea*, the northern species, experiences less chronic and acute heat stress than *A.
-carolensis*.
-Thus, we predicted that *A.
-picea* would have more genes than *A.
-carolinensis* that are up-regulated to high temperatures, while *A.
-carolinensis* would have higher expression of these thermally-tolerant genes near optimal temperatures.
-Reciprocally, we expected *A.
-carolinensis* to have greater up-regulation of genes in response to cold temperatures, while *A.
-picea* would have greater basal expression of cold-tolerant genes.
-Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures.
-From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A.
-picea* would have a broader thermal breadth of gene expression than *A.
-carolinensis*.
+carolinensis* [@diamond2013] we predicted that it would show more plasticity of gene expression in response to heat shock, and lower basal levels of expression for heat-tolerant genes.
 
@@
 Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
-To reduce redundancy in the enriched GO terms, we used the web program [ReviGO](http://revigo.irb.hr/).
+In addition, we used the web-tool DAVID for ....
 
@@
 The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1).
-For both species, the majority of responsive transcripts were in the *Low* category (Fig.
-1).
-More transcripts had *Intermediate* expression in *A.
+The density function for the temperature of maximum expression (Fig.
+1) showed that the majority of responsive genes were in the *Low* category for both species, with *A.
+picea* having more *Low* genes than *A.
+carolinensis*.
+More genes had *Intermediate* expression in *A.
 carolinensis* than in *A.
-picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig.
+picea*, while the two species had a similar number of *High* genes (Fig.
 1, Table 1).
@@
 
-We performed a number of additional comparisons to evaluate the extent to which the thermal reactionomes differed between *A.
-carolinensis* and *A.
-picea*.
-First, we found that the profile of the temperature of maximum expression for the thermally-responsive genes differed between species (Figure 3) with a similar number of genes at high temperatures, more genes at intermediate temperatures in *A.
-carolinensis* and more genes at low temperatures in *A.
-picea*.
-
 ```{r exp_plot, echo=FALSE, warning = FALSE, fig.cap = "Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
@@
 
-We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$).
-Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` TPM greater in *A.
-carolinensis*.
-Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A.
+The reaction norms of thermally-responsive genes between species differed by mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A.
+carolinensis* than *A picea*, and mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A.
 picea*.
@@
 Of the total differences ($\Delta T$) in reaction norms between species for each gene, about ...% (95% CI ...
-- ...%) was due to shifts in mean and ...% (95% CI ...
-- ...%) due to shifts in slope.
+- ...%) and ...% (95% CI ...
+- ...%) were due to shifts in mean and slope, respectively.
 
-We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species.
-Instead, we found that for *High* genes in *A.
+We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for *High* genes 
+*A.
 picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`), *Low* genes in *A.
@@
 
-Sensitivity of thermally-responsive genes is indicated by the thermal range of *Bimodal* expressed genes.
-
-We tested for a transcriptome-wide pattern of thermal sensitivity and tolerance by examining the thermal range for the *Bimodal* and *Intermediate* expressed genes.
-We found no in thermal sensitivity between species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).
-However, for *Intermediate* genes, *A.
-carolinensis* had a broader thermal range of stable gene expression (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
-
-We also explored the extent to which the temperature at which thermally-responsive genes are activated differs between species.
-For genes that are upregulated at *High* temperatures, we found that *A.
+Overall greater thermal tolerance of one species would be indicated by finding that gene expression is stable across a greater portion of the thermal gradient.
+For *Bimodal* expressed genes, we found no difference (*t*~df=1406~ = 0.31, *P* = 0.76) between species in the variance of the reaction norms for thermally-responsive genes.
+In contrast, *Intermediate* genes from *A.
+carolinensis* had a broader thermal range of stable gene expression than those from *A.
+picea* (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
+In addition, the species differed in the temperatures at which thermally-responsive genes were activated.
+For genes with increased expression at *High* temperatures, we found that *A.
 carolinensis* genes were activated on average 1°C hotter than *A.
-picea*.
-In contrast, for genes that are upregulated at *Low* temperaturese we found that...
+picea* genes (Fig.
+2).
+Similarly, for genes with increased expression at *Low* temperatures, *A.
+carolinensis* genes were activated on average 1°C hotter than *A.
+picea* genes (Fig.
+2).
+Thus, while the activation temperatures differed, the size of the window between activation of *Low* and *High* genes remained the same between species, consistent with the results for *Bimodal* genes above.
 
@@
 
+## Discussion
+
 overall pattern *A. carolinensis*
@@
 - more *Low* genes
-
-## Discussion
 
@@
 heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness.
-In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*.
+In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis* found in colder and warmer environments, respectively.
 We found that nearly 10% of the genes in this species show significant thermal responsiveness.

### 2014-05-21 ApTranscriptome_TR.Rmd
Some updates

@@
 
-Genes with increased expression at *High* temperatures are on average turned on 2.5°C higher in *ApVT* than *AcNC*.
+Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.
 
@@
 
-Genes with increased expression at *Low* temperatures are on average turned on 1°C higher in *ApVT* than *AcNC*.
+Genes with increased expression at *Low* temperatures are on average turned on 0.2°C higher in *ApVT* than *AcNC*.
 
@@
 
-```{r plot_low_T_on, echo=FALSE, eval=TRUE}
+```{r plot_T_on, echo=FALSE, eval=TRUE}
 # merge data for plotting
@@
 Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means and slope, as curvature and wiggle did not differ between species.
-To do this, I first standardize each of these parameters so they are directly comparable.
 
@@
 
-```{r extract_Transcripts}
+```{r extract_Transcripts_only}
 A22.only.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
@@
 
-for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
-    s <- A22.low.Ar.int.TPM.dt.sub[i]
-    X11()
-    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
-    points(s$val, s$TPM, pch = 8, col = "red")
-    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh") else {
-        print("OnLow")
-    }
-}
-
-#    if(min(s[which(s$val < 30) , pTPM]) < s[which(s$val == 38.5), pTPM] & s[which(s$val == 0), pTPM] < (s[which(s$val == 21), pTPM] + sd(s$pTPM))) print("On low") else {
+#for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
+#    s <- A22.low.Ar.int.TPM.dt.sub[i]
+#    X11()
+#    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
+#    points(s$val, s$TPM, pch = 8, col = "red")
+#    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh"#) else {
+#        print("OnLow")
+#    }
+#}
 
@@
 # save RData for manuscript file
-save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.themsens, t.thermbreadth, file = "ms.RData")
+save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.themsens, t.thermbreadth, file = "ms.RData")
 ```

### 2014-05-29 ApTranscriptome_TR.Rmd
Added gene functional annotation using DAVID throught the RDAVIDWebService Bioconductor package

@@
 library(methods) # required by `knitcitations` when file is `knit` file from command-line
+library(reshape2)
 
@@
 #biocLite("Rgraphviz")
+#biocLite("RDAVIDWebService")
 library(topGO) 
 library(Rgraphviz)
+library(RDAVIDWebService)
 
@@
 
+## DAVID Gene Functional Annotation
+
+As an alternative to topGO, in this section I use the [DAVID Functional Annotation web resource](http://david.abcc.ncifcrf.gov/) throught the Bioconductor [RDAVIDWebService](http://www.bioconductor.org/packages/release/bioc/html/RDAVIDWebService.html) package.
+Installing this packages is troublesome at best and my notes on how to do so are here ....
+
+BLAST to *Apis mellifera* and *Drosophila* RefSeq Protein database
+
+~~~
+blastx
+...
+~~~
+
+First, I registered as a user on the [DAVID webservice](http://david.abcc.ncifcrf.gov/content.jsp?file=DAVID_WebService.html).
+
+I then connect and upload my complete transcriptome BLAST table as the gene background list.
+
+Documentation for RDAVIDWebService is available [here](http://www.bdmg.com.ar/?page_id=695)
+
+```{r DAVID_background}
+# set up connection
+david.connect <- DAVIDWebService$new(email = "johnsg@uvm.edu")
+
+# load gene list
+ApTranscriptome.Amel.Dros.refseq.prot <- read.table("results/trinity-full/blast_amel_dros_refseq_protein/ApTranscriptome_parallel_blastx_refseq_protein_amellifera_drosophila1.txt", sep = "\t", header = FALSE)
+ApTranscriptome.Amel.Dros.refseq.prot <- data.table(ApTranscriptome.Amel.Dros.refseq.prot)
+setnames(ApTranscriptome.Amel.Dros.refseq.prot, c("Transcript", "gi", "percID", "alnLength", "mismatchCount", "gapOpenCount", "queryStart", "queryEnd", "subjectStart", "subjectEnd", "eVal", "bitScore"))
+setkey(ApTranscriptome.Amel.Dros.refseq.prot, Transcript)
+str(ApTranscriptome.Amel.Dros.refseq.prot)
+
+# extract RefSeq protein ID list
+split.refseq.df <- colsplit(string = ApTranscriptome.Amel.Dros.refseq.prot$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
+# remove version information for each gene accession
+refseq.col <- strsplit(split.refseq.df$XP, split = "\\.")
+refseq.id <- unlist(refseq.col)[2*1:length(refseq.col)-1]
+str(refseq.id)
+
+# add background list to DAVID
+david.list <- addList(david.connect, refseq.id,
+                      idType = "REFSEQ_PROTEIN",
+                      listName = "ApTranscriptome_Amellifera_Drosophila_refseq_protein",
+                      listType = "Background")
+```
+
+Next, I need to add candidate gene lists for each response type. Extract these from the full BLAST table.
+
+```{r DAVID_geneList}
+A22.high.geneList <- ApTranscriptome.Amel.Dros.refseq.prot[A22.high]
+# most of these were were not BLASTed. extract those that did
+A22.high.geneList.blast <- A22.high.geneList[which(!is.na(A22.high.geneList$gi)), ]
+# extract RefSeq protein ID
+A22.high.geneList.blast.XP <- colsplit(string = A22.high.geneList.blast$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
+# remove version information for each gene accession
+A22.high.geneList.refseq.col <- strsplit(A22.high.geneList.blast.XP$XP, split = "\\.")
+A22.high.geneList.refseq.id <- unlist(A22.high.geneList.refseq.col)[2*1:length(A22.high.geneList.refseq.col)-1]
+tail(A22.high.geneList.refseq.id)
+
+# add genelist to DAVID
+david.list <- addList(david.connect, A22.high.geneList.refseq.id,
+                      idType = "REFSEQ_PROTEIN",
+                      listName = "ApTranscriptome_A22_high",
+                      listType = "Gene")
+
+# set annotation category and background gene list
+setAnnotationCategories(david.connect, "GOTERM_BP_ALL")
+getBackgroundListNames(david.connect)
+setCurrentBackgroundPosition(david.connect, 1)
+getCurrentBackgroundListPosition(david.connect)
+
+# gbtain Functional Annotation Cluster
+A22.high.termCluster <- getClusterReport(david.connect, type = "Term")
+head(summary(A22.high.termCluster))
+members(A22.high.termCluster)
+plot2D(A22.high.termCluster, 1)
+```
+
 ## Visualize responsive transcripts

### 2014-06-04 doc/ApTranscriptome_manuscript.Rmd
Updates to ms

@@
 
-When ants are exposed to different temperatures, lots of genes change expression. The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress.
-
-Comparisons between species indicate that the species from the warmer environment tolerates thermal stress by higher baseline expression of protective genes
+- When ants are exposed to different temperatures, lots of genes change expression. 
+- The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress. 
+- Comparisons between species indicate that *A.
+carolinensis*, which has greater basal gene expression and physiological thermal tolerance, tolerates thermal stress by shutting down gene expression.
+*A.
+picea* tolerates thermal stress by activating protective genes
 
@@
 
-We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate.
+We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$padj < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate.
 Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction.
@@
 Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species.
-Of the total differences ($\Delta T$) in reaction norms between species for each gene, about ...% (95% CI ...
-- ...%) and ...% (95% CI ...
-- ...%) were due to shifts in mean and slope, respectively.
+Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.
 
-We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for *High* genes 
-*A.
-picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`), *Low* genes in *A.
-carolinensis* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`) and *Intermediate* genes in both (Wilcoxon V = `r w3$statistic`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r w3$conf.int[1:2]`), basal expression near optimum temperatures of 19.5C was always higher in *A.
+We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species.
+Instead, we found that for genes expressed at *High* temperatures in *A.
+picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), genes expressed at *Low* temperatures in *A.
+carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) and *Intermediate* genes in both (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), basal expression near optimum temperatures of 19.5°C was always higher in *A.
 carolinensis*.
@@
 For genes with increased expression at *High* temperatures, we found that *A.
-carolinensis* genes were activated on average 1°C hotter than *A.
-picea* genes (Fig.
+carolinensis* genes were activated on average at 33.8°C compared to 32.9°C for *A.
+picea* (Fig.
 2).
-Similarly, for genes with increased expression at *Low* temperatures, *A.
-carolinensis* genes were activated on average 1°C hotter than *A.
-picea* genes (Fig.
+For genes with increased expression at *Low* temperatures, *A.
+carolinensis* genes were activated on average at 12.8°C compared to 13°C for a *A.
+picea* (Fig.
 2).
-Thus, while the activation temperatures differed, the size of the window between activation of *Low* and *High* genes remained the same between species, consistent with the results for *Bimodal* genes above.
 
@@
     scale_y_continuous(name="Density") +
-    scale_x_continuous(name=expression("Temperature"))
+    scale_x_continuous(name=expression("Temperature of gene activation"))
 print(T_on_plot)
@@
 
-overall pattern *A. carolinensis*
+- many genes thermally responsive
+- thermal reaction norms differ between species
+  - *A. carolinensis* limited future potential to adapt to heat stress
+- more genes activated in response to cold shock --> greater adaptability to cold
+- differences among species 
+  - *A. carolinensis* tolerates thermal stress by shutting down biological processes
+  - in *A. picea* the *Intermediate* genes from *A. carolinensis* are over-represented in *Low* and *Intermediate* categories
+  - potential explanation is that *A.
+carolinensis* tolerates thermal stress by shutting-down metabolism and relies on other mechanisms (behavior, protein stability) whereas *A.
+picea* activates protective genes which requires metabolism...
+- comparison of gene terms with other species (Sara?)
 
-- more genes thermally-responsive
-- tolerates thermal stress by shutting down biological processes
-
-overall pattern *A. picea*
-
-- more *Low* genes
-
-- greater adaptability to cold than hot thermal stress
-    - *A. carolinensis* limited future potential to adapt to heat stress
-- different mechanisms of thermal tolernace
-    - *A. carolinensis* tolerates thermal stress by shutting down (*Intermediate* genes)
-    - *A. picea* tolerates by a combination of shutting down and activating 'protective' genes 
-       - some indication that *A.
-picea* maintains express in cold conditions that *A.
-carolinensis* shuts down based on overlap of *Low* genes in *A.
-picea* with *Intermediate* genes in *A.
-carolinensis*
-- metabolic compensation?
+- regression approach allows us to detect patterns not evident from ANOVA design
 
@@
 While considerable research has focused on a few key genes (e.g.
-heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness.
-In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis* found in colder and warmer environments, respectively.
-We found that nearly 10% of the genes in this species show significant thermal responsiveness.
+heat shock proteins) involved in thermal stress response [@krebs1997; @gaitainespitia2013], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness.
+In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively.
+We found that nearly 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress.
+The patterns of gene expression between species differed substantially, with more down-regulation of gene expression in *A.
+carolinensis* and more up-regulation of gene expression in *A.
+picea* in response to thermal stress (Fig.
+2).
 
-### Molecular toolkit of thermal responsiveness
+Given the differences in the number of thermally-responsive genes between the two species (Table 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species, with greater variation between species at low temperatures (Fig.
+1).
+A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures.
+This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addobediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
+Moreover, in an experiment with 94 species of *Drosophila*, Kellerman et al [-@kellerman2012] showed that limited genetic variation among species (e.g.
+phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits.
+Thus, our results indicate that, when coupled with phenotypic data, transcriptomic studies can provide insight on adapatability.
+For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal].
 
-While the sheer number of thermally-responsive transcripts makes generalizations difficult, groupings such as Gene Ontology (GO) can be used to identify categories associated with thermal stress.
-For heat response, the semantic space of the enriched GO terms included *gene expression*, *regeneration* and *cellular component organization or biogenesis*.
-...previous work...
-For cold response, there were 65 enriched GO terms that grouped by semantic space into *protein modification process*, *sensory perception of mechanical stimulus*, *response to mechanical stimulus*, *folic acid metabolism*, *membrane lipid metabolism* and *protein complex biogenesis*.
-...previous work...
+Depending on the frequency with which organisms experience stressful conditions, the reaction norms of protective genes may differ considerably.
+Organisms that frequently experience stress may have high basal levels of expression with little change, while organisms that less frequently experiences stress would have lower basal levels with large changes in response to environmental stress [@cite...].
+For *Aphaenogaster*, we predicted that genes up-regulated to cold or hot temperatures in one species might be expressed at higher levels near optimal growth temperatures in the other species to confer greater basal stress tolerance.
+This prediction was not supported, as we found greater expression near optimal temperatures in *A.
+carolinensis* for *High* genes in *A.
+picea*, as predicted, but also for *Low* genes in *A.
+carolinensis* and *Intermediate* genes in both species.
+As *A.
+carolinensis* is from a warmer environment, a likely explanation for this is that this species simply has a higher metabolic rate and thus greater gene expression across it's thermal range.
+The observation that there are more *Intermediate* genes, which are shut-down under thermal stress, in *A.
+carolinensis* supports this explanation.
+...cite some coral literature....
 
-*Bimodal* genes are general stress response genes that are activated under both heat and cold stress.
-The semantic space of the enriched GO terms included *regulation of metabolism*, *biological regulation*, *macromolecule biosynthesis*, ...
+Of the `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes that were thermally-responsive, `r round(with(sigtable, sigtable[Coefficient=="Temperature:Colony","Number.significant"]) / with(sigtable, sum(sigtable[Coefficient %in% c("Temperature", "Temperature:Colony"),"Number.significant"])) * 100, 0)`% differed in their responsiveness by colony.
+The differences between species were due to changes in the mean ($\Delta M$) and slope ($\Delta S$) of the reaction norms for each thermally-responsive gene.
+This result contrasts with the findings of Murren et al.
+[-@murren2014], who found that most of the differences in phenotypic reaction norms were due to shape.
+As genes are not directly analogous to phenotypes, this contradiction is not surprising.
+Instead, it indicates that small shifts in the mean of many genes underlying a trait can result in non-linear phenotypic changes.
 
-*Intermediate* genes represent non-essential molecular functions that are shut-down as conditions become stressful.
-The semantic space for the enriched GO terms included *lipid metabolic process*, *glycoprotein biosynthetic process*, *membrane lipid metabolic process*, *cellular protein metabolic process*, *membrane lipid biosynthetic process*.
-...indicates shutting down of non essential molecule biosynthesis...
+To date, transcriptomic studies have performed differential expression analysis between a few different conditions or environments (but see [@hodginsdavis2012]).
+By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
+Thus, we were able to compare the temperature at which genes were activated near the upper and lower thermal limits between species (Fig.
+2).
+For genes activated at cold temperatures, there was little difference among species with *A.
+picea* genes turning on at 13°C compared to 12.8°C for *A.
+carolinensis*.
+For high temperature genes, *A.
+carolinensis* genes were turned on at 33.8°C compared to 32.9°C for *A.
+picea*.
 
-**Sara - comparisons of our gene list to other studies**
+### Molecular toolkit of thermal tolerance
 
-This result poses a paradox.
-Why is the heat shock functional response highly conserved at the biological process level, while the exact molecular pathways appear to be highly labile?
-We suggest two possible explanations.
-First, regulatory networks controlling gene expression are evolutionarily dynamic.
-Trade-offs between function in different organisms may constrain which genes are involved in thermal response, thus resulting in different pathways to the same phenotype.
-Second, variation in molecular pathways may be due to inconsistent selection on thermal stress response.
-That is, as the environment changes and species shift their distributions, selection may relax or strengthen repeatedly within and across lineages.
-Different genes may be recruited within each selective episode.
-Differentiating between these, and alternate, explanations will require greater geographic and population genetic sampling.
+REANALYZING...
 
-### Transcriptome-wide comparisions
+### Caveats/limitations
 
-Of the 8,492 genes that were thermally-responsive, 80% differed in their responsiveness by colony.
-The majority of this difference in the reactionomes was due to changes in the slope ($\Delta S$) and curvature ($\Delta C$) which together accounted for nearly 90% of the differences between reaction norms for all transcripts.
-In a meta-analysis of phenotypic reaction norms, Murren et al.
-[-@murren2014] similarly found that differences in the shape of the reaction norms accounted for the majority of the differences.
-Thus, our finding suggests that this phenotypic pattern may be explained by greater genetic lability in the shape than the mean of gene expression.
-The relative role of plasticity versus adaptation in response to rapid environmental change is widely debated [@vellend...?].
-Our data suggest that plasticity through changes in the responsiveness of gene expression are likely to occur more rapidly than changes in mean values.
-
-An important note is that we would have completely missed this pattern if we had performed a differential expression study using only a few temperatures.
-Only by measuring gene expression across a thermal gradient and fitting a function to each transcript were we able to characterize the shape of the reaction norm of expression...
-To our knowledge, this is the first time that a reaction norm approach has been applied to the study of global gene expression.
-
-Given the large differences in number (Table 1) and responsiveness of genes between the two species, a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures.
-Intriguingly, this result mirrors the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
-A plausible explanation is that organisms have fewer molecular mechanisms to tolerate heat than cold stress.
-Future studies ...
-implications for climate change as the evolutionary potential of ecototherms to climate change may be limited as suggested by ...
-[kingsolver?]
+As with all *de novo* transcriptomes, our work has some important caveats.
+First, given that all sequenced ant genomes to date have about 18,000 genes [@smith...], our transcriptome is highly fragmented.
+Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts.
+A genome assembly, in progress, would considerably resolve this problem.
+Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is poor, as we have to rely upon resources from distantly related *Drosophila* and *Apis*.
+Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005].
+Large differences in patterns of gene expression may only have small effects on organismal fitness.
+Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change.
+Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@cite...].
+In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@cite...].
 
@@
 
-A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*.
-Our comparative results are limited to two populations, and an important future step will be expanding this work to examine differences in gene expression among a larger number of populations.
-Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance.
-The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of species, and inability of ants to mate in the lab.
-However, future work could leverage genomic information to infer the relatedness among species in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.
-
-In summary, we have generated the first transcriptome of a temperate ant species, and have identified approximately 10% of genes that show thermal responsiveness.
-In this group, we find many known stress response genes, but also identify novel genetic pathways that respond to temperature.
+In this work, we have completed the first transcriptome for a temperate ant species, *Aphaenogaster picea* and *A.
+carolinensis*, to date.
+We found that a large proportion of the transcriptome is thermally-responsive.
+Transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures.
+...differences among species...*A.
+carolinensis* greater organismal tolerance and shuts down metabolism under thermal stress...*A.
+picea* less tolerant and activates protective proteins requiring metabolism...
 These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.

### 2014-06-04 ApTranscriptome_TR.Rmd

@@
 # add background list to DAVID
-david.list <- addList(david.connect, refseq.id,
+david.background <- addList(david.connect, refseq.id,
                       idType = "REFSEQ_PROTEIN",
@@
 # add genelist to DAVID
-david.list <- addList(david.connect, A22.high.geneList.refseq.id,
+david.geneList <- addList(david.connect, A22.high.geneList.refseq.id,
                       idType = "REFSEQ_PROTEIN",
@@
 
-# gbtain Functional Annotation Cluster
+# obtain Functional Annotation Cluster
 A22.high.termCluster <- getClusterReport(david.connect, type = "Term")
 head(summary(A22.high.termCluster))
-members(A22.high.termCluster)
-plot2D(A22.high.termCluster, 1)
+members(A22.high.termCluster)[[1]]$Term
+members(A22.high.termCluster)[[2]]$Term
+members(A22.high.termCluster)[[3]]$Term
+
+plot2D(A22.high.termCluster, 3)
 ```

### 2014-06-05 ApTranscriptome_TR.Rmd
Functioning RDAVIDWebService analysis. Potential bug in 'Enrichment' scores

--- 2014-05-21 ApTranscriptome_TR.Rmd
@@
 
-**May 20, 2014**
+**June 2, 2014**
 
@@
 library(xtable)
-library(methods) # required by `knitcitations` when file is `knit` file from command-line
+library(methods) # required by `knitcitations` when file is `knit` from command-line
+library(reshape2)
+library(psych)
 
@@
 #biocLite("Rgraphviz")
+#biocLite("RDAVIDWebService")
+#biocLite("GOSemSim")
 library(topGO) 
 library(Rgraphviz)
+library(RDAVIDWebService)
+library(GOSemSim)
 
@@
 
+```{r gsea_by_category}
+# A22 'High' genes
+A22.geneList.high <- rep(0, length = length(Ap.geneList))
+names(A22.geneList.high) <- names(Ap.geneList)
+A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
+A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
+```
+
 ```{r gsea_setdiff, results = 'hide'}
 # A22 'High' genes not in Ar**
-A22.geneList.high <- rep(0, length = length(Ap.geneList))
-names(A22.geneList.high) <- names(Ap.geneList)
-A22.geneList.high[(which(names(A22.geneList.high) %in% A22.only.high))] <- 1
-A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
+A22.geneList.high.only <- rep(0, length = length(Ap.geneList))
+names(A22.geneList.high.only) <- names(Ap.geneList)
+A22.geneList.high.only[(which(names(A22.geneList.high.only) %in% A22.only.high))] <- 1
+A22.high.only.gsea <- gsea(genelist = A22.geneList.high.only, geneID2GO = geneID2GO)
 
@@
 
+For many of the responsive categories, it appears that there is considerable redundancy in the enriched GO terms.
+I use the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then perform hierarhichal clustering based on this information distance.
+
+```{r GOSemSim}
+# similarity among enriched GO terms
+A22.GO.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = "fly")
+# distance matrix
+dist.A22.GO.term.sim <- dist(A22.GO.term.sim)
+plot(hclust(dist(A22.GO.term.sim)))
+
+```
+
+## DAVID Gene Functional Annotation
+
+As an alternative to topGO, in this section I use the [DAVID Functional Annotation web resource](http://david.abcc.ncifcrf.gov/) throught the Bioconductor [RDAVIDWebService](http://www.bioconductor.org/packages/release/bioc/html/RDAVIDWebService.html) package.
+Installing this packages is troublesome at best and my notes on how to do so are here ....
+
+BLAST to *Apis mellifera* and *Drosophila* RefSeq Protein database
+
+~~~
+blastx
+...
+~~~
+
+First, I registered as a user on the [DAVID webservice](http://david.abcc.ncifcrf.gov/content.jsp?file=DAVID_WebService.html).
+
+I then connect and upload my complete transcriptome BLAST table as the gene background list.
+
+Documentation for RDAVIDWebService is available [here](http://www.bdmg.com.ar/?page_id=695)
+
+```{r DAVID_background}
+# set up connection
+david.connect <- DAVIDWebService$new(email = "johnsg@uvm.edu")
+
+# load gene list
+ApTranscriptome.Amel.Dros.refseq.prot <- read.table("results/trinity-full/blast_amel_dros_refseq_protein/ApTranscriptome_parallel_blastx_refseq_protein_amellifera_drosophila1.txt", sep = "\t", header = FALSE)
+ApTranscriptome.Amel.Dros.refseq.prot <- data.table(ApTranscriptome.Amel.Dros.refseq.prot)
+setnames(ApTranscriptome.Amel.Dros.refseq.prot, c("Transcript", "gi", "percID", "alnLength", "mismatchCount", "gapOpenCount", "queryStart", "queryEnd", "subjectStart", "subjectEnd", "eVal", "bitScore"))
+setkey(ApTranscriptome.Amel.Dros.refseq.prot, Transcript)
+str(ApTranscriptome.Amel.Dros.refseq.prot)
+
+# extract RefSeq protein ID list
+split.refseq.df <- colsplit(string = ApTranscriptome.Amel.Dros.refseq.prot$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
+# remove version information for each gene accession
+refseq.col <- strsplit(split.refseq.df$XP, split = "\\.")
+refseq.id <- unlist(refseq.col)[2*1:length(refseq.col)-1]
+str(refseq.id)
+#write.table(refseq.id, "DAVID.background.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
+
+# add background list to DAVID
+david.background <- addList(david.connect, refseq.id,
+                      idType = "REFSEQ_PROTEIN",
+                      listName = "ApTranscriptome_Amellifera_Drosophila_refseq_protein",
+                      listType = "Background")
+```
+
+Next, I need to add candidate gene lists for each response type. Extract these from the full BLAST table.
+
+```{r DAVID_A22_high}
+A22.high.geneList <- ApTranscriptome.Amel.Dros.refseq.prot[A22.high]
+# most of these were were not BLASTed. extract those that did
+A22.high.geneList.blast <- A22.high.geneList[which(!is.na(A22.high.geneList$gi)), ]
+# extract RefSeq protein ID
+A22.high.geneList.blast.XP <- colsplit(string = A22.high.geneList.blast$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
+# remove version information for each gene accession
+A22.high.geneList.refseq.col <- strsplit(A22.high.geneList.blast.XP$XP, split = "\\.")
+A22.high.geneList.refseq.id <- unlist(A22.high.geneList.refseq.col)[2*1:length(A22.high.geneList.refseq.col)-1]
+
+# add genelist to DAVID
+david.A22.high.geneList <- addList(david.connect, A22.high.geneList.refseq.id,
+                      idType = "REFSEQ_PROTEIN",
+                      listName = "ApTranscriptome_A22_high",
+                      listType = "Gene")
+
+# set annotation category and background gene list
+#setAnnotationCategories(david.connect, "GOTERM_BP_ALL")
+getBackgroundListNames(david.connect)
+setCurrentBackgroundPosition(david.connect, 1)
+getCurrentBackgroundListPosition(david.connect)
+david.connect
+
+# Note that the current display for "species" shows only the first in list, but I confirmed in email with the programmer, Cristobal Fresno Rodriguez, that this is a bug and all species are being used, as shown in this code
+
+getCurrentSpeciesPosition2<-function(object){
+    ans <- NA
+    if(length(getSpecieNames(object)) != 0){
+        stub<-getStub(object)
+        ans<-as.integer(strsplit(stub$getCurrentSpecies(), split = ",")[[1]]) + 1
+    }
+    return(ans)
+}
+
+getCurrentSpeciesPosition2(david.connect)
+
+# obtain Functional Annotation Cluster
+A22.high.termCluster <- getClusterReport(david.connect, type = "Term")
+summary(A22.high.termCluster)
+members(A22.high.termCluster)[[1]]$Term
+members(A22.high.termCluster)[[2]]$Term
+members(A22.high.termCluster)[[3]]$Term
+
+# calculate Enrichment
+geometric_mean <- function(x) exp(mean(log(x)))
+-log10(geometric_mean(members(A22.high.termCluster)[[1]]$PValue))
+-log10(geometric_mean(members(A22.high.termCluster)[[2]]$PValue))
+-log10(geometric_mean(members(A22.high.termCluster)[[3]]$PValue))
+-log10(geometric_mean(members(A22.high.termCluster)[[4]]$PValue))
+-log10(geometric_mean(members(A22.high.termCluster)[[5]]$PValue))
+-log10(geometric_mean(members(A22.high.termCluster)[[6]]$PValue))
+-log10(geometric_mean(members(A22.high.termCluster)[[7]]$PValue))
+```
+
+Three clusters identified for *A. picea* genes turned on at high temperatures.
+
+1: protein synthesis
+2: regulation
+3: metabolism
+
+A22 low
+
+```{r DAVID_A22_low}
+A22.low.geneList <- ApTranscriptome.Amel.Dros.refseq.prot[A22.low]
+# most of these were were not BLASTed. extract those that did
+A22.low.geneList.blast <- A22.low.geneList[which(!is.na(A22.low.geneList$gi)), ]
+# extract RefSeq protein ID
+A22.low.geneList.blast.XP <- colsplit(string = A22.low.geneList.blast$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
+# remove version information for each gene accession
+A22.low.geneList.refseq.col <- strsplit(A22.low.geneList.blast.XP$XP, split = "\\.")
+A22.low.geneList.refseq.id <- unlist(A22.low.geneList.refseq.col)[2*1:length(A22.low.geneList.refseq.col)-1]
+str(A22.low.geneList.refseq.id)
+
+# add genelist to DAVID
+david.A22.low.geneList <- addList(david.connect, A22.low.geneList.refseq.id,
+                      idType = "REFSEQ_PROTEIN",
+                      listName = "ApTranscriptome_A22_low",
+                      listType = "Gene")
+david.A22.low.geneList
+
+# set annotation category and background gene list
+#setAnnotationCategories(david.connect, "GOTERM_BP_ALL")
+getBackgroundListNames(david.connect)
+setCurrentBackgroundPosition(david.connect, 1)
+getCurrentBackgroundListPosition(david.connect)
+david.connect
+getCurrentSpeciesPosition2(david.connect)
+
+# obtain Functional Annotation Cluster
+A22.low.termCluster <- getClusterReport(david.connect, type = "Term")
+summary(A22.low.termCluster)
+members(A22.low.termCluster)[[1]]$Term
+members(A22.low.termCluster)[[2]]$Term
+members(A22.low.termCluster)[[3]]$Term
+members(A22.low.termCluster)[[1]]$Term
+members(A22.low.termCluster)[[2]]$Term
+members(A22.low.termCluster)[[3]]$Term
+
+```
+
+7 clusters for *A. picea* low-temperature activated genes:
+
+1. 
+2.
+3. 
+4. membrane
+5. cytochrome P450
+6. zinc finger
+7. Ras/Rho protein signaling
+
 ## Visualize responsive transcripts
@@
 
-#for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
-#    s <- A22.low.Ar.int.TPM.dt.sub[i]
-#    X11()
-#    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
-#    points(s$val, s$TPM, pch = 8, col = "red")
-#    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh"#) else {
-#        print("OnLow")
-#    }
-#}
+for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
+    s <- A22.low.Ar.int.TPM.dt.sub[i]
+    X11()
+    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
+    points(s$val, s$TPM, pch = 8, col = "red")
+    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh") else {
+        print("OnLow")
+    }
+}
 

### 2014-06-05 ApTranscriptome_TR.Rmd
Clustered topGO GSEA results by GO term semantic similarity for A22 Low and High genes

@@
 library(reshape2)
-library(psych)
 
@@
 # similarity among enriched GO terms
-A22.GO.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = "fly")
+A22.low.GO.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = "fly")
 # distance matrix
-dist.A22.GO.term.sim <- dist(A22.GO.term.sim)
-plot(hclust(dist(A22.GO.term.sim)))
-
+dist.A22.low.GO.term.sim <- dist(A22.low.GO.term.sim)
+plot(hclust(dist(A22.low.GO.term.sim)))
+
+A22.high.GO.term.sim <- termSim(A22.high.gsea$GO.ID, A22.high.gsea$GO.ID, ont = "BP", organism = "fly")
+# distance matrix
+plot(hclust(dist(A22.high.GO.term.sim)))
 ```
@@
 getCurrentSpeciesPosition2<-function(object){
-    ans <- NA
-    if(length(getSpecieNames(object)) != 0){
-        stub<-getStub(object)
-        ans<-as.integer(strsplit(stub$getCurrentSpecies(), split = ",")[[1]]) + 1
-    }
-    return(ans)
-}
-
-getCurrentSpeciesPosition2(david.connect)
-
-# obtain Functional Annotation Cluster
-A22.high.termCluster <- getClusterReport(david.connect, type = "Term")
-summary(A22.high.termCluster)
-members(A22.high.termCluster)[[1]]$Term
-members(A22.high.termCluster)[[2]]$Term
-members(A22.high.termCluster)[[3]]$Term
-
-# calculate Enrichment
 geometric_mean <- function(x) exp(mean(log(x)))

### 2014-06-09 ApTranscriptome_TR.Rmd
Dropped DAVID analysis due to concerns about database quality

@@
 
-## DAVID Gene Functional Annotation
-
-As an alternative to topGO, in this section I use the [DAVID Functional Annotation web resource](http://david.abcc.ncifcrf.gov/) throught the Bioconductor [RDAVIDWebService](http://www.bioconductor.org/packages/release/bioc/html/RDAVIDWebService.html) package.
-Installing this packages is troublesome at best and my notes on how to do so are here ....
-
-BLAST to *Apis mellifera* and *Drosophila* RefSeq Protein database
-
-~~~
-blastx
-...
-~~~
-
-First, I registered as a user on the [DAVID webservice](http://david.abcc.ncifcrf.gov/content.jsp?file=DAVID_WebService.html).
-
-I then connect and upload my complete transcriptome BLAST table as the gene background list.
-
-Documentation for RDAVIDWebService is available [here](http://www.bdmg.com.ar/?page_id=695)
-
-```{r DAVID_background}
-# set up connection
-david.connect <- DAVIDWebService$new(email = "johnsg@uvm.edu")
-
-# load gene list
-ApTranscriptome.Amel.Dros.refseq.prot <- read.table("results/trinity-full/blast_amel_dros_refseq_protein/ApTranscriptome_parallel_blastx_refseq_protein_amellifera_drosophila1.txt", sep = "\t", header = FALSE)
-ApTranscriptome.Amel.Dros.refseq.prot <- data.table(ApTranscriptome.Amel.Dros.refseq.prot)
-setnames(ApTranscriptome.Amel.Dros.refseq.prot, c("Transcript", "gi", "percID", "alnLength", "mismatchCount", "gapOpenCount", "queryStart", "queryEnd", "subjectStart", "subjectEnd", "eVal", "bitScore"))
-setkey(ApTranscriptome.Amel.Dros.refseq.prot, Transcript)
-str(ApTranscriptome.Amel.Dros.refseq.prot)
-
-# extract RefSeq protein ID list
-split.refseq.df <- colsplit(string = ApTranscriptome.Amel.Dros.refseq.prot$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
-# remove version information for each gene accession
-refseq.col <- strsplit(split.refseq.df$XP, split = "\\.")
-refseq.id <- unlist(refseq.col)[2*1:length(refseq.col)-1]
-str(refseq.id)
-#write.table(refseq.id, "DAVID.background.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
-
-# add background list to DAVID
-david.background <- addList(david.connect, refseq.id,
-                      idType = "REFSEQ_PROTEIN",
-                      listName = "ApTranscriptome_Amellifera_Drosophila_refseq_protein",
-                      listType = "Background")
-```
-
-Next, I need to add candidate gene lists for each response type. Extract these from the full BLAST table.
-
-```{r DAVID_A22_high}
-A22.high.geneList <- ApTranscriptome.Amel.Dros.refseq.prot[A22.high]
-# most of these were were not BLASTed. extract those that did
-A22.high.geneList.blast <- A22.high.geneList[which(!is.na(A22.high.geneList$gi)), ]
-# extract RefSeq protein ID
-A22.high.geneList.blast.XP <- colsplit(string = A22.high.geneList.blast$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
-# remove version information for each gene accession
-A22.high.geneList.refseq.col <- strsplit(A22.high.geneList.blast.XP$XP, split = "\\.")
-A22.high.geneList.refseq.id <- unlist(A22.high.geneList.refseq.col)[2*1:length(A22.high.geneList.refseq.col)-1]
-
-# add genelist to DAVID
-david.A22.high.geneList <- addList(david.connect, A22.high.geneList.refseq.id,
-                      idType = "REFSEQ_PROTEIN",
-                      listName = "ApTranscriptome_A22_high",
-                      listType = "Gene")
-
-# set annotation category and background gene list
-#setAnnotationCategories(david.connect, "GOTERM_BP_ALL")
-getBackgroundListNames(david.connect)
-setCurrentBackgroundPosition(david.connect, 1)
-getCurrentBackgroundListPosition(david.connect)
-david.connect
-
-# Note that the current display for "species" shows only the first in list, but I confirmed in email with the programmer, Cristobal Fresno Rodriguez, that this is a bug and all species are being used, as shown in this code
-
-getCurrentSpeciesPosition2<-function(object){
-geometric_mean <- function(x) exp(mean(log(x)))
--log10(geometric_mean(members(A22.high.termCluster)[[1]]$PValue))
--log10(geometric_mean(members(A22.high.termCluster)[[2]]$PValue))
--log10(geometric_mean(members(A22.high.termCluster)[[3]]$PValue))
--log10(geometric_mean(members(A22.high.termCluster)[[4]]$PValue))
--log10(geometric_mean(members(A22.high.termCluster)[[5]]$PValue))
--log10(geometric_mean(members(A22.high.termCluster)[[6]]$PValue))
--log10(geometric_mean(members(A22.high.termCluster)[[7]]$PValue))
-```
-
-Three clusters identified for *A. picea* genes turned on at high temperatures.
-
-1: protein synthesis
-2: regulation
-3: metabolism
-
-A22 low
-
-```{r DAVID_A22_low}
-A22.low.geneList <- ApTranscriptome.Amel.Dros.refseq.prot[A22.low]
-# most of these were were not BLASTed. extract those that did
-A22.low.geneList.blast <- A22.low.geneList[which(!is.na(A22.low.geneList$gi)), ]
-# extract RefSeq protein ID
-A22.low.geneList.blast.XP <- colsplit(string = A22.low.geneList.blast$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
-# remove version information for each gene accession
-A22.low.geneList.refseq.col <- strsplit(A22.low.geneList.blast.XP$XP, split = "\\.")
-A22.low.geneList.refseq.id <- unlist(A22.low.geneList.refseq.col)[2*1:length(A22.low.geneList.refseq.col)-1]
-str(A22.low.geneList.refseq.id)
-
-# add genelist to DAVID
-david.A22.low.geneList <- addList(david.connect, A22.low.geneList.refseq.id,
-                      idType = "REFSEQ_PROTEIN",
-                      listName = "ApTranscriptome_A22_low",
-                      listType = "Gene")
-david.A22.low.geneList
-
-# set annotation category and background gene list
-#setAnnotationCategories(david.connect, "GOTERM_BP_ALL")
-getBackgroundListNames(david.connect)
-setCurrentBackgroundPosition(david.connect, 1)
-getCurrentBackgroundListPosition(david.connect)
-david.connect
-getCurrentSpeciesPosition2(david.connect)
-
-# obtain Functional Annotation Cluster
-A22.low.termCluster <- getClusterReport(david.connect, type = "Term")
-summary(A22.low.termCluster)
-members(A22.low.termCluster)[[1]]$Term
-members(A22.low.termCluster)[[2]]$Term
-members(A22.low.termCluster)[[3]]$Term
-members(A22.low.termCluster)[[1]]$Term
-members(A22.low.termCluster)[[2]]$Term
-members(A22.low.termCluster)[[3]]$Term
-
-```
-
-7 clusters for *A. picea* low-temperature activated genes:
-
-1. 
-2.
-3. 
-4. membrane
-5. cytochrome P450
-6. zinc finger
-7. Ras/Rho protein signaling
-
 ## Visualize responsive transcripts

### 2014-06-09 ApTranscriptome_TR.Rmd
Modified RxNtype to allow for more response types

@@
 ```{r responsive_transcripts}
-grepFunc <- function(lmitem, term = NA) {
-  coefs <- coefficients(lmitem)
-  coef.names <- names(coefs)
-  keep <- if(length(grep(term, coef.names)) > 0) TRUE else FALSE
-}
-
 interaction.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
@@
 ```{r expression_shape, eval=TRUE, echo=TRUE}
+
 # calculate response type for responsive transcripts
@@
 # reorder 
-tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]
+tt2 <- type.table[c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn"), c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn")]
 
@@
 # create geneid2go.map file from FastAnnotator annotationtable.txt
-geneid2GOmap(annotation.file)
+mapname = "geneid2go.map"
+if(file.exists(mapname)) print("Gene ID to GO term file exists") else{
+    geneid2GOmap(annotation.file, ontology = c("BP", "CC", "MF"), mapname = mapname)
+}
 ```
@@
 
-```{r gsea_by_type, results = 'hide'}
+```{r gsea_Ap_high, results = 'hide'}
 ## Genes with *High* expression in both colonies
@@
 Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)
-
+```
+
+```{r gsea_Ap_high_cluster}
+# similarity among enriched GO terms
+Ap.high.GO.term.sim <- termSim(Ap.high.gsea$GO.ID, Ap.high.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term
+colnames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term
+
+# distance matrix
+Ap.high.hclust <- hclust(dist(Ap.high.GO.term.sim))
+plot(Ap.high.hclust)
+
+# report items in each cluster at height 0.6 
+GSEAReportClusters(Ap.high.hclust, h = 0.6)
+```
+
+```{r gsea_Ap_low, results = "hide"}
 ## Genes with *Low* expression in both colonies
@@
 Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
-
+```
+
+```{r gsea_Ap_low_cluster}
+# similarity among enriched GO terms
+Ap.low.GO.term.sim <- termSim(Ap.low.gsea$GO.ID, Ap.low.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ap.low.GO.term.sim) <- Ap.low.gsea$Term
+colnames(Ap.low.GO.term.sim) <- Ap.low.gsea$Term
+
+# distance matrix
+Ap.low.hclust <- hclust(dist(Ap.low.GO.term.sim))
+plot(Ap.low.hclust)
+
+# report items in each cluster at height 1.5
+GSEAReportClusters(Ap.low.hclust, h = 1.5)
+```
+
+```{r gsea_Ap_bim, resuts = 'hide'}
 ## Genes with *Bimodal* expression in both colonies
@@
 Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
-
+```
+
+```{r gsea_Ap_bim_cluster}
+# similarity among enriched GO terms
+Ap.bim.GO.term.sim <- termSim(Ap.bim.gsea$GO.ID, Ap.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ap.bim.GO.term.sim) <- Ap.bim.gsea$Term
+colnames(Ap.bim.GO.term.sim) <- Ap.bim.gsea$Term
+
+# distance matrix
+Ap.bim.hclust <- hclust(dist(Ap.bim.GO.term.sim))
+plot(Ap.bim.hclust)
+
+# report items in each cluster at height 1.5
+GSEAReportClusters(Ap.bim.hclust, h = 1.5)
+```
+
+```{r gsea_Ap_int, results = 'hide'}
 ## Genes with *Intermediate* expression in both colonies
@@
 Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
+```
+
+```{r gsea_Ap_int_cluster}
+# similarity among enriched GO terms
+Ap.int.GO.term.sim <- termSim(Ap.int.gsea$GO.ID, Ap.int.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ap.int.GO.term.sim) <- Ap.int.gsea$Term
+colnames(Ap.int.GO.term.sim) <- Ap.int.gsea$Term
+
+# distance matrix
+Ap.int.hclust <- hclust(dist(Ap.int.GO.term.sim))
+plot(Ap.int.hclust)
+
+# report items in each cluster at height 1.5
+GSEAReportClusters(Ap.int.hclust, h = 1.5)
+```
 

### 2014-06-10 ApTranscriptome_TR.Rmd
Analysis with responsive genes separated into 'linear' and 'on' categories

@@
 library(Rgraphviz)
-library(RDAVIDWebService)
 library(GOSemSim)
@@
 
-* High - increase expression with temperature
-* Low - decrease expression with temperature
+* HighOn - expression activated at high temperatures
+* PositiveLinear - expression positive linear across thermal gradient (0 - 38°C)
+* LowOn - expression activated at low temperatures
+* NegativeLinear - expression negative linear across thermal gradient (0 - 38°C)
 * Intermediate - maximum expression at intermediate temperatures (14 - 28C)
-* Bimodal - expressed greater than two standard deviations of expression at both low and high temperatures
+* Bimodal - expressed greater than one standard deviation of expression at both low and high temperatures
 
@@
 ```{r expression_shape, eval=TRUE, echo=TRUE}
-
 # calculate response type for responsive transcripts
@@
 # table
-Ap.type.table <- rbind(A22.type.table, Ar.type.table)
+(Ap.type.table <- rbind(A22.type.table, Ar.type.table))
 
@@
 # reorder 
-tt2 <- type.table[c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn"), c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn")]
+tt2 <- type.table[c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn", "NotResp"), c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn", "NotResp")]
 
@@
 tt3 <- cbind(tt3, rowSums(tt3))
-colnames(tt3)[6] <- "Total"
-rownames(tt3)[6] <- "Total"
+colnames(tt3)[8] <- "Total"
+rownames(tt3)[8] <- "Total"
 # move names into table
@@
 # name col/row by species
-rownames(tt3) <- c("AcNC", "", "", "", "", "", "")
-colnames(tt3) <- c("ApVT", "", "", "", "", "", "")
+rownames(tt3) <- c("AcNC", "", "", "", "", "", "", "", "")
+colnames(tt3) <- c("ApVT", "", "", "", "", "", "", "", "")
 
@@
 # list of transcripts that are 'high' expressed in A22
-A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
+A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "HighOn" | Ap.response.type$A22.type == "PositiveLinear"), ]
 
@@
 # list of transcripts that are 'high' expressed in Ar
-Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
+Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "LowOn" | Ap.response.type$A22.type == "NegativeLinear"), ]
 
@@
 
-To confirm that there are not sample-level issues, I performed the same comparison using transcripts where I do *not* expect to see a difference in expression.
+To confirm that there are not sample-level issues, I performed the same comparison using *Intermediate* expressed-genes where I do not expect to see a difference in expression.
 
 ```{r intermediate_transcripts}
-# list of transcripts that are 'Intermediate' expressed in Ar
-Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
-# T test
-t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
-# Wilcoxon signed rank-test
-wilcox.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-
-# list of transcripts that are 'Intermediate' expressed in A22
-A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
-# T test
-t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
-# Wilcoxon signed rank-test
-wilcox.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-
 # list of transcripts that are 'Intermediate' expressed in either colony
@@
 ```{r extract_Transcripts}
-A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
-Ar.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]
-
-A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
-Ar.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]
-
-A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
-Ar.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
-
-A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
-Ar.int <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "HighOn"), "Transcript"]
+Ar.high.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "HighOn"), "Transcript"]
+
+A22.poslinear.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "PositiveLinear"), "Transcript"]
+Ar.poslinear.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "POsitiveLinear"), "Transcript"]
+
+A22.low.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "LowOn"), "Transcript"]
+Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "LowOn"), "Transcript"]
+
+A22.neglinear.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "NegativeLinear"), "Transcript"]
+Ar.neglinear.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "NegativeLinear"), "Transcript"]
+
+A22.bim.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
+Ar.bim.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+
+A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
+Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
 ```
@@
 names(A22.geneList.high) <- names(Ap.geneList)
-A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
+A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high.transcripts))] <- 1
 A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
@@
 ```{r ggplot_high}
-trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high], aes(x=val, y=TPM.scaled, group=Transcript)) + 
+trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high.transcripts][1:220,], aes(x=val, y=TPM.scaled, group=Transcript)) +
   geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
@@
 print(trp_A22_high)
+
+trp_A22_poslinear <- ggplot(resp.TPM.dt.sub[A22.poslinear.transcripts][1:220,], aes(x=val, y=TPM.scaled, group=Transcript)) +
+  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
+  facet_grid(. ~ colony2) + 
+  scale_y_continuous(name="Expression (scaled)") +
+  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
+print(trp_A22_poslinear)
+
+### predicted values
+resp.TPM.dt.sub.pred[,pTPM.scaled:=scale(pTPM), by = Transcript]
+
+pred_A22_high <- ggplot(resp.TPM.dt.sub.pred[A22.high.transcripts][1:220,], aes(x=val, y=pTPM.scaled, group=Transcript)) +
+  geom_line() + 
+  facet_grid(. ~ colony) + 
+  scale_y_continuous(name="Expression (scaled)") +
+  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
+print(pred_A22_high)
+
+pred_A22_poslinear <- ggplot(resp.TPM.dt.sub.pred[A22.poslinear.transcripts][1:220,], aes(x=val, y=pTPM.scaled, group=Transcript)) +
+  geom_line() + 
+  facet_grid(. ~ colony) + 
+  scale_y_continuous(name="Expression (scaled)") +
+  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
+print(pred_A22_poslinear)
+
+ggplot(resp.TPM.dt.sub.pred["100348|*|comp3450455_c0_seq1"], aes(x=val, y=pTPM.scaled, group=Transcript)) +
+  geom_line() + 
+  facet_grid(. ~ colony) + 
+  scale_y_continuous(name="Expression (scaled)") +
+  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
+
 ```

### 2014-06-10 ApTranscriptome_TR.Rmd
Rolling back to analysis with High and Low categories

--- 2014-06-09 ApTranscriptome_TR.Rmd
@@
 
-**June 2, 2014**
+**June 10, 2014**
 
@@
 
-```{r max_exp_PDF}
+```{r max_exp_PDF, warning = FALSE, message = FALSE}
 # reshape data
@@
   scale_x_continuous(name="Temperature of maximum expression")
-suppressWarnings(print(maxexplot))
+print(maxexplot)
 ```
@@
 # list of transcripts that are 'high' expressed in A22
-A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
+A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
 
 # Compare expression at optimum temp (A22.opt) between colonies using t-test
-t.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
-boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
+t.test(A22.high$A22.opt, A22.high$Ar.opt)
+boxplot(A22.high$A22.opt, A22.high$Ar.opt)
 
 # T test on log-transformed values to control for outliers
-t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
-boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
+t.test(log(A22.high$A22.opt+1), log(A22.high$Ar.opt+1))
+boxplot(log(A22.high$A22.opt+1), log(A22.high$Ar.opt+1))
 ```
@@
 ```{r A22_high_wilcoxon}
-w1 <- wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+w1 <- wilcox.test(A22.high$A22.opt, A22.high$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
 w1
@@
 # list of transcripts that are 'high' expressed in Ar
-Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
+Ar.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
 
 # t-test with log values
-t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
-boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
+t.test(log(Ar.low$A22.opt+1), log(Ar.low$Ar.opt+1))
+boxplot(log(Ar.low$A22.opt+1), log(Ar.low$Ar.opt+1))
 
 # Wilcoxon signed rank-test
-w2 <- wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+w2 <- wilcox.test(Ar.low$A22.opt, Ar.low$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
 w2
@@
 ```{r intermediate_transcripts}
-# list of transcripts that are 'Intermediate' expressed in Ar
-Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
-# T test
-t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
-# Wilcoxon signed rank-test
-wilcox.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-
-# list of transcripts that are 'Intermediate' expressed in A22
-A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
-# T test
-t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
-# Wilcoxon signed rank-test
-wilcox.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-
 # list of transcripts that are 'Intermediate' expressed in either colony
@@
 # extract 'Intermediate' expressed transcripts for A22
+A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
 A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
@@
 # repeat for Ar
+Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
 Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
@@
 ```{r extract_Transcripts}
-A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
-Ar.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]
-
-A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
-Ar.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]
-
-A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
-Ar.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
-
-A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
-Ar.int <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
+Ar.high.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]
+
+A22.low.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
+Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]
+
+A22.bim.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
+Ar.bim.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+
+A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
+Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
 ```
@@
 # transcripts expressed at *High* and *Bimodal* temperatures in A22
-A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.high, A22.bim), "A22")]
+A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.high.transcripts, A22.bim.transcripts), "A22")]
 setkey(A22.high.TPM.dt.sub, Transcript)
@@
 # repeat for Ar
-Ar.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.high, Ar.bim), "Ar")]
+Ar.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.high.transcripts, Ar.bim.transcripts), "Ar")]
 setkey(Ar.high.TPM.dt.sub, Transcript)
@@
 # transcripts expressed at *Low* temperatures in A22
-A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.low, A22.bim), "A22")]
+A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.low.transcripts, A22.bim.transcripts), "A22")]
 setkey(A22.low.TPM.dt.sub, Transcript)
@@
 # repeat for Ar
-Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.low, Ar.bim), "A22")]
+Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.low.transcripts, Ar.bim.transcripts), "A22")]
 setkey(Ar.low.TPM.dt.sub, Transcript)
@@
 ```{r ggplot_high}
-trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high], aes(x=val, y=TPM.scaled, group=Transcript)) + 
+trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high.transcripts], aes(x=val, y=TPM.scaled, group=Transcript)) + 
   geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 

### 2014-06-11 ApTranscriptome_TR.Rmd

--- 2014-06-10 ApTranscriptome_TR.Rmd
@@
 library(Rgraphviz)
+
 library(GOSemSim)
@@
 ```{r responsive_transcripts}
+grepFunc <- function(lmitem, term = NA) {
+  coefs <- coefficients(lmitem)
+  coef.names <- names(coefs)
+  keep <- if(length(grep(term, coef.names)) > 0) TRUE else FALSE
+}
+
 interaction.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
@@
 
-* HighOn - expression activated at high temperatures
-* PositiveLinear - expression positive linear across thermal gradient (0 - 38°C)
-* LowOn - expression activated at low temperatures
-* NegativeLinear - expression negative linear across thermal gradient (0 - 38°C)
+* High - increase expression with temperature
+* Low - decrease expression with temperature
 * Intermediate - maximum expression at intermediate temperatures (14 - 28C)
@@
 # reorder 
-tt2 <- type.table[c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn", "NotResp"), c("Bimodal", "PositiveLinear", "HighOn", "Intermediate", "NegativeLinear", "LowOn", "NotResp")]
+tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]
 
@@
 tt3 <- cbind(tt3, rowSums(tt3))
-colnames(tt3)[8] <- "Total"
-rownames(tt3)[8] <- "Total"
+colnames(tt3)[6] <- "Total"
+rownames(tt3)[6] <- "Total"
 # move names into table
@@
 # name col/row by species
-rownames(tt3) <- c("AcNC", "", "", "", "", "", "", "", "")
-colnames(tt3) <- c("ApVT", "", "", "", "", "", "", "", "")
+rownames(tt3) <- c("AcNC", "", "", "", "", "", "")
+colnames(tt3) <- c("ApVT", "", "", "", "", "", "")
 
@@
 # list of transcripts that are 'high' expressed in A22
-A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "HighOn" | Ap.response.type$A22.type == "PositiveLinear"), ]
+A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
 
@@
 # list of transcripts that are 'high' expressed in Ar
-Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "LowOn" | Ap.response.type$A22.type == "NegativeLinear"), ]
+Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
 
@@
 ```{r extract_Transcripts}
-A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "HighOn"), "Transcript"]
-Ar.high.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "HighOn"), "Transcript"]
-
-A22.poslinear.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "PositiveLinear"), "Transcript"]
-Ar.poslinear.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "POsitiveLinear"), "Transcript"]
-
-A22.low.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "LowOn"), "Transcript"]
-Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "LowOn"), "Transcript"]
-
-A22.neglinear.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "NegativeLinear"), "Transcript"]
-Ar.neglinear.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "NegativeLinear"), "Transcript"]
+A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
+Ar.high.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]
+
+A22.low.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
+Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]
 

### 2014-06-11 ApTranscriptome_TR.Rmd
Cleaning up file

@@
 library(Rgraphviz)	
-
 library(GOSemSim)
@@
 # set "trt" to true values - truncated in file names for convenience
-
 A22.TPM[val==3,val:=3.5]

### 2014-06-12 ApTranscriptome_TR.Rmd
Rolled back RxNtype function to only High, Low, Bimodal and Intermediate categories output

@@
 
-**June 2, 2014**
+**June 12, 2014**
 
@@
 library(reshape2)
+library(RColorBrewer)
 
@@
 # plot
-mosaicplot(t(tt2), ylab = "AcNC", xlab = "ApVT", main = "Mosaic plot of responsive transcripts")
+mosaic.colors <- brewer.pal(5, "Blues")
+
+par(mar = c(5.1, 4.1, 4.1, 5.1))
+mosaicplot(tt2, ylab = "A. picea", xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
+labels <- c("High", "Low", "Bimodal", "Intermediate", "NotResp")
+text(c(.1,.3,.55,.75,.95), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE)
+text(par("usr")[2] + 0.25, c(.95, .75, .55, .35, .1), srt = 0, adj = 1, labels = labels, xpd = TRUE)
+text(.5, par("usr")[4] + .1, "A. carolinensis", xpd = TRUE)
 ```

### 2014-06-12 scripts/RxNseq.R

@@ -147,23 +147,16 @@ RxNtype <- function(lmitem) {
 
         # determine expression type
         A22.exp.type <- NA
-        # "Bimodal" if max expression below 10C and above 30C is greater than one standard deviation above expression at the optimum temperature (19.5C)
+        # "Bimodal" if max expression below 10C and above 30C is greater than one standard deviation above expression at the optimum temperature (19.5C)  
         if(max(A22.pout[A22.pout$val < 10, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + sd(A22.pout$pTPM)) & max(A22.pout[A22.pout$val > 30, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + sd(A22.pout$pTPM))) A22.exp.type <- "Bimodal" else {
-              # "PositiveLinear" if max expression at temperature above 30C and minimum expression at 0C
-              if(A22.max.val > 30 & A22.min.val == 0) A22.exp.type <- "PositiveLinear" else {
-                  # "HighOn" if max expression above 30C and minimum not at 0C (due to quadratic curvature)
-                  if(A22.max.val > 30 & A22.min.val != 0) A22.exp.type <- "HighOn" else {
-                      # "NegativeLinear" if max expression below 10C and minimum expression at 38.5
-                      if(A22.max.val < 10 & A22.min.val == 38.5) A22.exp.type <- "NegativeLinear" else {
-                          # "LowOn" if max expression below 10C and minimum not at 38.5C (due to quadratic curvature)
-                          if(A22.max.val < 10 & A22.min.val != 38.5) A22.exp.type <- "LowOn" else {
-                              A22.exp.type <- "Intermediate"
-                          }
-                      }
-                  }
-              }
-          }
-    }
+          # "High" if max expression at temperature above 30C
+          if(A22.max.val > 30) A22.exp.type <- "High" else {
+            # "Low" if max expression at temperature below 10C
+            if(A22.max.val < 10) A22.exp.type <- "Low" else
+              A22.exp.type <- "Intermediate"
+            } # end else
+          } # end else
+        } # end if/else
     
 
         
@@ -186,21 +179,14 @@ RxNtype <- function(lmitem) {
             Ar.exp.type <- NA
[…]

### 2014-06-18 ApTranscriptome_TR.Rmd
Accidental but correct merge of rollback branch to master

@@
 
-**June 12, 2014**
+**June 16, 2014**
 
@@
 #biocLite("Rgraphviz")
-#biocLite("RDAVIDWebService")
 #biocLite("GOSemSim")
@@
 ```{r download, echo = TRUE}
-### Annotation file
-
-# from either AWS or GoogleDrive
+# URL for annotation file
 annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
-#a2 <- getURL("https://googledrive.com/host/0B75IymziRJ_9Tlg1U1Vxbjk1bzg") # GoogleDrive link
-
+# load 
 annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
@@
 annotation.table <- data.table(annotation.file)
-head(annotation.table)
+str(annotation.table)
 ```
@@
 length(unique(TPM.dt.sub$Transcript))
+
+# remove from annotation.table
+setkey(annotation.table, Sequence.Name)
+annotation.table <- annotation.table[!cont.list]
+str(annotation.table)
 ```
@@
 mosaic.colors <- brewer.pal(5, "Blues")
-
-par(mar = c(5.1, 4.1, 4.1, 5.1))
-mosaicplot(tt2, ylab = "A. picea", xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
+par(mar = c(5.1, 3.1, 2.1, 7.5))
+mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
 labels <- c("High", "Low", "Bimodal", "Intermediate", "NotResp")
-text(c(.1,.3,.55,.75,.95), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE)
-text(par("usr")[2] + 0.25, c(.95, .75, .55, .35, .1), srt = 0, adj = 1, labels = labels, xpd = TRUE)
-text(.5, par("usr")[4] + .1, "A. carolinensis", xpd = TRUE)
+text(c(.1,.3,.55,.75,.95), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
+text(.5, par("usr")[4], "A. carolinensis", xpd = TRUE, font = 3)
+legend(1.05, .75, fill = mosaic.colors, c("High", "Low", "Bimodal", "Intermediate", "NotResp"), xpd=TRUE)
 ```
@@
 
-I perform gene set enrichment analysis below, but a quick `grep` shows that there are 26 transcripts with GO term "response to stress", though this is not enriched compared to the frequency of this term in the full dataset.
+### Proportion of responsive transcripts annotated
+
+Of the responsive transcripts, `r prettyNum(length(which(responsive.lms.ann.type$best.hit.to.nr == "-"))/length(responsive.lms.ann.type$best.hit.to.nr)*100)`% are annotated, while `r prettyNum(length(which(annotation.table$best.hit.to.nr == "-"))/nrow(annotation.table) * 100)`% of all transcripts are annotated.
+
+### Candidate gene enrichment
+
+First, I explore of candidate genes with the GO term "response to stress" are enriched in the responsive data set.
 
@@
 
+While `r nrow(GO0006950.responsive)` "response to stress" genes are in the responsive set, this is not enriched compared to the whole transcriptome.
+
 There are also 7 heat shock related genes in the responsive transcripts, out of 130 total.
@@
 
+### Gene set enrichment analysis
+
 I use [topGO](http://www.bioconductor.org/packages/2.12/bioc/html/topGO.html) to perform gene set enrichment analysis (GSEA) seperately for each expression type (bimodal, intermediate, high, low).
@@
 # create geneid2go.map file from FastAnnotator annotationtable.txt
-mapname = "geneid2go.map"
+mapname = "results/geneid2go.map"
 if(file.exists(mapname)) print("Gene ID to GO term file exists") else{
-    geneid2GOmap(annotation.file, ontology = c("BP", "CC", "MF"), mapname = mapname)
+    geneid2GOmap(as.data.frame(annotation.table), ontology = c("BP", "CC", "MF"), mapname = mapname)
 }
@@
 # read mappings file
-geneID2GO <- readMappings(file = "geneid2go.map")
+geneID2GO <- readMappings(file = mapname)
 str(head(geneID2GO))
@@
 
-### GSEA for thermally-responsive transcripts ###
+### GSEA for thermally-responsive transcripts
 
@@
 Perform GSEA for all thermally-responsive transcripts.
+
+For many of the responsive categories, it appears that there is considerable redundancy in the enriched GO terms.
+I use the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then perform hierarhichal clustering based on this information distance.
 
@@
 
-# Function to select top genes (defined above)
+# Function to select genes with adjusted P-value < 0.05
 selectFDR <- function(padj) {
@@
                  description = "BP gene set analysis", ontology = "BP",
-                 allGenes = Ap.geneList, geneSel = selectFDR,
+                 allGenes = Ap.geneList, 
+                 geneSel = selectFDR,
                  nodeSize = 10,
@@
 
-# report items in each cluster at height 1.5
-GSEAReportClusters(Ap.low.hclust, h = 1.5)
+# report items in each cluster at height 2
+GSEAReportClusters(Ap.low.hclust, h = 2)
 ```
@@
 ```{r gsea_Ap_bim, resuts = 'hide'}
-## Genes with *Bimodal* expression in both colonies
+# Genes with *Bimodal* expression in both colonies
 Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
@@
 
-Next, I perform GSEA for genes in each functional type in one species but not the other (e.g. the set difference) to gain insight on differences between the colonies.
-
-```{r extract_Transcripts_only}
-A22.only.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
-Ar.only.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]
-
-A22.only.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
-Ar.only.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
-
-A22.only.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
-Ar.only.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
-
-A22.only.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
-Ar.only.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
-```
-
-```{r gsea_by_category}
+```{r gsea_Ar_int_A22_bim, results = 'hide'}
+# genes with *Intermediate* expression in Ar and *Bimodal* in A22
+Ar.int.A22.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Bimodal"), "Transcript"]
+Ar.int.A22.bim.geneList <- rep(0, length = length(Ap.geneList))
+names(Ar.int.A22.bim.geneList) <- names(Ap.geneList)
+Ar.int.A22.bim.geneList[(which(names(Ar.int.A22.bim.geneList) %in% Ar.int.A22.bim))] <- 1
+# check correct number of values set to 1
+table(Ar.int.A22.bim.geneList)
+# run GSEA
+Ar.int.A22.bim.gsea <- gsea(genelist = Ar.int.A22.bim.geneList, geneID2GO = geneID2GO, plotpath = NA)
+```
+
+```{r gsea_Ar_int_A22_bim_cluster}
+# similarity among enriched GO terms
+Ar.int.A22.bim.gsea.term.sim <- termSim(Ar.int.A22.bim.gsea$GO.ID, Ar.int.A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
+colnames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
+# distance matrix
+Ar.int.A22.bim.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.bim.gsea.term.sim))
+plot(Ar.int.A22.bim.gsea.term.sim.hclust)
+
+# report items in each cluster at height 1.5
+GSEAReportClusters(Ar.int.A22.bim.gsea.term.sim.hclust, h = 1)
+```
+
+```{r gsea_Ar_int_A22_low, results = 'hide'}
+# genes with *Intermediate* expression in Ar and *Low* in A22
+Ar.int.A22.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Low"), "Transcript"]
+Ar.int.A22.low.geneList <- rep(0, length = length(Ap.geneList))
+names(Ar.int.A22.low.geneList) <- names(Ap.geneList)
+Ar.int.A22.low.geneList[(which(names(Ar.int.A22.low.geneList) %in% Ar.int.A22.low))] <- 1
+# check correct number of values set to 1
+table(Ar.int.A22.low.geneList)
+# run GSEA
+Ar.int.A22.low.gsea <- gsea(genelist = Ar.int.A22.low.geneList, geneID2GO = geneID2GO, plotpath = NA)
+```
+
+```{r gsea_Ar_int_A22_low_cluster}
+# similarity among enriched GO terms
+Ar.int.A22.low.gsea.term.sim <- termSim(Ar.int.A22.low.gsea$GO.ID, Ar.int.A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
+colnames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
+# distance matrix
+Ar.int.A22.low.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.low.gsea.term.sim))
+plot(Ar.int.A22.low.gsea.term.sim.hclust)
+# report items in each cluster at height 1.8
+GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.6)
+```
+
+### GSEA for each functional type in each species
+
+GO enrichment for *A. picea* **High** genes.
+
+```{r gsea_A22_high, results = 'hide'}
 # A22 'High' genes
@@
 
-```{r gsea_setdiff, results = 'hide'}
-# A22 'High' genes not in Ar**
-A22.geneList.high.only <- rep(0, length = length(Ap.geneList))
-names(A22.geneList.high.only) <- names(Ap.geneList)
-A22.geneList.high.only[(which(names(A22.geneList.high.only) %in% A22.only.high))] <- 1
-A22.high.only.gsea <- gsea(genelist = A22.geneList.high.only, geneID2GO = geneID2GO)
-
-# *Ar 'High' genes not in A22**
-Ar.geneList.high <- rep(0, length = length(Ap.geneList))
-names(Ar.geneList.high) <- names(Ap.geneList)
-Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.only.high))] <- 1
-Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
-
-# A22 'Low' genes not in Ar**
+Clustering of enriched GO terms for *A. picea* **High** genes.
+
+```{r gsea_A22_high_cluster}
+# similarity among enriched GO terms
+A22.high.gsea.term.sim <- termSim(A22.high.gsea$GO.ID, A22.high.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
+colnames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
+# distance matrix
+A22.high.gsea.term.sim.hclust <- hclust(dist((A22.high.gsea.term.sim)))
+plot((A22.high.gsea.term.sim.hclust))
+# report items in each cluster at height 1.8
+GSEAReportClusters(A22.high.gsea.term.sim.hclust, h = 1.2)
+```
+GO enrichment for *A. picea* **Low** genes.
+
+```{r gsea_A22_low, results = 'hide'}
+# A22 'Low' genes
 A22.geneList.low <- rep(0, length = length(Ap.geneList))
 names(A22.geneList.low) <- names(Ap.geneList)
-A22.geneList.low[(which(names(A22.geneList.low) %in% A22.only.low))] <- 1
+A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low.transcripts))] <- 1
 A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
-# Ar 'Low' genes not in A22**
-Ar.geneList.low <- rep(0, length = length(Ap.geneList))
-names(Ar.geneList.low) <- names(Ap.geneList)
-Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.only.low))] <- 1
-Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
-
-# A22 'Bimodal' genes not in Ar**
+```
+
+Clustering of enriched GO terms for *A. picea* **Low** genes.
+
+```{r gsea_A22_low_cluster}
+# similarity among enriched GO terms
+A22.low.gsea.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
+colnames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
+# distance matrix
+A22.low.gsea.term.sim.hclust <- hclust(dist((A22.low.gsea.term.sim)))
+plot((A22.low.gsea.term.sim.hclust))
+# report items in each cluster at height 1.8
+GSEAReportClusters(A22.low.gsea.term.sim.hclust, h = 2)
+```
+
+GO enrichment for *A. picea* **Bimodal** genes.
+
+```{r gsea_A22_bim, results = 'hide'}
+# A22 'Bim' genes
 A22.geneList.bim <- rep(0, length = length(Ap.geneList))
 names(A22.geneList.bim) <- names(Ap.geneList)
-A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.only.bim))] <- 1
+A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim.transcripts))] <- 1
 A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
-
-# Ar 'Bimodal' genes not in A22**
-Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
-names(Ar.geneList.bim) <- names(Ap.geneList)
-Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.only.bim))] <- 1
-Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
-
-# A22 'Intermediate' genes not in Ar**
+```
+
+Clustering of enriched GO terms for *A. picea* **Low** genes.
+
+```{r gsea_A22_bim_cluster}
+# similarity among enriched GO terms
+A22.bim.gsea.term.sim <- termSim(A22.bim.gsea$GO.ID, A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
+colnames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
+# distance matrix
+A22.bim.gsea.term.sim.hclust <- hclust(dist((A22.bim.gsea.term.sim)))
+plot((A22.bim.gsea.term.sim.hclust))
+# report items in each cluster at height 1.8
+GSEAReportClusters(A22.bim.gsea.term.sim.hclust, h = 1)
+```
+
+GO enrichment for *A. picea* **Intermediate** genes.
+
+```{r gsea_A22_int, results = 'hide'}
+# A22 'Int' genes
 A22.geneList.int <- rep(0, length = length(Ap.geneList))
 names(A22.geneList.int) <- names(Ap.geneList)
-A22.geneList.int[(which(names(A22.geneList.int) %in% A22.only.int))] <- 1
-table(A22.geneList.int)
+A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int.transcripts))] <- 1
 A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
-
-# Ar 'Intermediate' genes not in A22**
-Ar.geneList.int <- rep(0, length = length(Ap.geneList))
-names(Ar.geneList.int) <- names(Ap.geneList)
-Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.only.int))] <- 1
-Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
-
+```
+
+Clustering of enriched GO terms for *A. picea* **Intermediate** genes.
+
+```{r gsea_A22_int_cluster}
+# similarity among enriched GO terms
+A22.int.gsea.term.sim <- termSim(A22.int.gsea$GO.ID, A22.int.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
+colnames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
+# distance matrix
+A22.int.gsea.term.sim.hclust <- hclust(dist((A22.int.gsea.term.sim)))
+plot((A22.int.gsea.term.sim.hclust))
+# report items in each cluster at height 1.8
+GSEAReportClusters(A22.int.gsea.term.sim.hclust, h = 1.3)
+```
+
+```{r gsea_combine, eval = FALSE}
 # combine into single table
@@
 
-```{r gsea_setdiff_table, echo = FALSE, results = 'asis'}
+```{r gsea_setdiff_table, echo = FALSE, results = 'asis', eval = FALSE}
 #pandoc.table(Ap.setdiff.gsea[ , 1:4], caption = "Enriched GO terms found in one species but not the other for each thermal-response category")
@@
 ```
-
-For many of the responsive categories, it appears that there is considerable redundancy in the enriched GO terms.
-I use the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then perform hierarhichal clustering based on this information distance.
 
@@
 
-```{r plot_GOstress, echo=FALSE, eval=TRUE}
+```{r plot_GOstress, echo=FALSE, eval=FALSE}
 high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]

### 2014-06-18 ApTranscriptome_TR.Rmd
Updates to GSEA clustering and mosaic plot

@@
 # reorder
-tt2 <- type.table[c(2,4,1,3,5), c(2,4,1,3,5)]
+tt2 <- type.table[c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", "Intermediate", "High", "Bimodal", "NotResp")]
 
 # plot
+png("mosaic_plot.png")
 mosaic.colors <- brewer.pal(5, "Blues")
@@
 mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
-labels <- c("High", "Low", "Bimodal", "Intermediate", "NotResp")
-text(c(.1,.3,.55,.75,.95), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
+labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
+text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
 text(.5, par("usr")[4], "A. carolinensis", xpd = TRUE, font = 3)
-legend(1.05, .75, fill = mosaic.colors, c("High", "Low", "Bimodal", "Intermediate", "NotResp"), xpd=TRUE)
+legend(1.05, .75, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
+dev.off()
 ```
@@
   geom_density(alpha=0.2, position="identity") + 
-  #scale_fill_manual(name = "Colony", values = c("white", "black")) + 
-  scale_y_continuous(name="Density") +
-  scale_x_continuous(name="Temperature of maximum expression")
+  labs(x = "Temperature of maximum expression", y = "Density") +
+  theme(axis.title = element_text(size = rel(2))) +
+  theme(axis.text = element_text(size = rel(1.2)))
 print(maxexplot)
+
+# Figure for presentation
+png("results/temp_max_expression.png")
+maxexplot + theme(legend.position = "none")
+dev.off()
 ```
@@
 
-# plot
+# low plot
+png("T_on_low.png")
+T_on_low_plot <- ggplot(rbind(A22.low.T_on, Ar.low.T_on), aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) +
+    labs(x = "Temperature of gene activation", y = "Density") +
+    theme(axis.title = element_text(size = rel(2))) +
+    theme(axis.text = element_text(size = rel(1.2))) +
+    theme(legend.position = "none")
+print(T_on_low_plot)
+dev.off()
+
+# high plot
+png("T_on_high.png")
+T_on_high_plot <- ggplot(rbind(A22.high.T_on, Ar.high.T_on), aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) +
+    labs(x = "Temperature of gene activation", y = "Density") +
+    theme(axis.title = element_text(size = rel(2))) +
+    theme(axis.text = element_text(size = rel(1.2))) +
+    theme(legend.position = "none")
+print(T_on_high_plot)
+dev.off()
+
+# combined plot
 T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
@@
     geom_density(adjust = 3, fill = NA, size = 2) + 
-    scale_y_continuous(name="Density") +
-    scale_x_continuous(name=expression("Temperature"))
+    labs(x = "Temperature of gene activation", y = "Density") +
+    theme(axis.title = element_text(size = rel(2))) +
+    theme(axis.text = element_text(size = rel(1.2))) +
+    theme(legend.position = "none")
 print(T_on_plot)
@@
 
-# report items in each cluster at height 0.6 
-GSEAReportClusters(Ap.high.hclust, h = 0.6)
+# report items in each cluster
+GSEAReportClusters(Ap.high.hclust, h = 1)
+write.table(unlist(GSEAReportClusters(Ap.high.hclust, h = 0.6)), file = "Ap_high_hclust.txt")
 ```
@@
 # report items in each cluster at height 2
-GSEAReportClusters(Ap.low.hclust, h = 2)
+(Ap.low.hclust.report <- GSEAReportClusters(Ap.low.hclust, h = 2))
 ```
@@
 # report items in each cluster at height 1.5
-GSEAReportClusters(Ap.bim.hclust, h = 1.5)
+(Ap.bim.hclust.report <- GSEAReportClusters(Ap.bim.hclust, h = 1.5))
 ```
@@
 ```{r gsea_Ap_int, results = 'hide'}
-## Genes with *Intermediate* expression in both colonies
+# genes with *Intermediate* expression in both colonies
 Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
@@
 table(Ap.geneList.int)
-# Run GSEA
+# run GSEA
 Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
@@
 # report items in each cluster at height 1.5
-GSEAReportClusters(Ap.int.hclust, h = 1.5)
-```
-
-## merge results into single table
-Ap.high.gsea$Type <- "High"
-Ap.low.gsea$Type <- "Low"
-Ap.bim.gsea$Type <- "Bimodal"
-Ap.int.gsea$Type <- "Intermediate"
-# combine
-Ap.gsea <- rbind(Ap.high.gsea, Ap.low.gsea, Ap.bim.gsea, Ap.int.gsea)
-# reorder
-Ap.gsea1 <- Ap.gsea[,c("Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
-colnames(Ap.gsea1)[7] <- "P"
-# write to file
-write.csv(Ap.setdiff.gsea, file = paste(resultsdir, "Ap_setdiff_GSEA_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
-```
-
-```{r gsea_by_type_table, echo = FALSE, results = 'asis'}
-#pandoc.table(Ap.gsea1[, 1:3], caption = "Enriched GO terms for each thermal-response category")
-print(xtable(Ap.gsea1[ , 1:3], comment = getOption("xtable.comment", FALSE)))
+(Ap.int.hclust.report <- GSEAReportClusters(Ap.int.hclust, h = 1.5))
 ```
@@
 # report items in each cluster at height 1.5
-GSEAReportClusters(Ar.int.A22.bim.gsea.term.sim.hclust, h = 1)
+(Ar.int.A22.bim.hclust.report <- GSEAReportClusters(Ar.int.A22.bim.gsea.term.sim.hclust, h = 1))
 ```
@@
 # report items in each cluster at height 1.8
-GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.6)
+(Ar.int.A22.low.hclust.report <- GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.6))
 ```
@@
 ```
+
 GO enrichment for *A. picea* **Low** genes.
@@
 
-Clustering of enriched GO terms for *A. picea* **Low** genes.
+Clustering of enriched GO terms for *A. picea* **Bimodal** genes.
 
@@
 
+GO enrichment for *A. carolinensis* **High** genes.
+
+```{r gsea_Ar_high, results = 'hide'}
+# Ar 'High' genes
+Ar.geneList.high <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.high) <- names(Ap.geneList)
+Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high.transcripts))] <- 1
+Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
+```
+
+Clustering of enriched GO terms for *A. carolinensis* **High** genes.
+
+```{r gsea_Ar_high_cluster}
+# similarity among enriched GO terms
+Ar.high.gsea.term.sim <- termSim(Ar.high.gsea$GO.ID, Ar.high.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ar.high.gsea.term.sim) <- Ar.high.gsea$Term
+colnames(Ar.high.gsea.term.sim) <- Ar.high.gsea$Term
+# distance matrix
+Ar.high.gsea.term.sim.hclust <- hclust(dist((Ar.high.gsea.term.sim)))
+plot((Ar.high.gsea.term.sim.hclust))
+# report items in each cluster
+GSEAReportClusters(Ar.high.gsea.term.sim.hclust, h = 1.2)
+```
+GO enrichment for *A. carolinensis* **Low** genes.
+
+```{r gsea_Ar_low, results = 'hide'}
+# Ar 'Low' genes
+Ar.geneList.low <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.low) <- names(Ap.geneList)
+Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low.transcripts))] <- 1
+Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
+```
+
+Clustering of enriched GO terms for *A. carolinensis* **Low** genes.
+
+```{r gsea_Ar_low_cluster}
+# similarity among enriched GO terms
+Ar.low.gsea.term.sim <- termSim(Ar.low.gsea$GO.ID, Ar.low.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ar.low.gsea.term.sim) <- Ar.low.gsea$Term
+colnames(Ar.low.gsea.term.sim) <- Ar.low.gsea$Term
+# distance matrix
+Ar.low.gsea.term.sim.hclust <- hclust(dist((Ar.low.gsea.term.sim)))
+plot((Ar.low.gsea.term.sim.hclust))
+# report items in each cluster at height 1.8
+GSEAReportClusters(Ar.low.gsea.term.sim.hclust, h = 2)
+```
+
+GO enrichment for *A. carolinensis* **Bimodal** genes.
+
+```{r gsea_Ar_bim, results = 'hide'}
+# Ar 'Bim' genes
+Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.bim) <- names(Ap.geneList)
+Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim.transcripts))] <- 1
+Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
+```
+
+Clustering of enriched GO terms for *A. carolinensis* **Bimodal** genes.
+
+```{r gsea_Ar_bim_cluster}
+# similarity among enriched GO terms
+Ar.bim.gsea.term.sim <- termSim(Ar.bim.gsea$GO.ID, Ar.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ar.bim.gsea.term.sim) <- Ar.bim.gsea$Term
+colnames(Ar.bim.gsea.term.sim) <- Ar.bim.gsea$Term
+# distance matrix
+Ar.bim.gsea.term.sim.hclust <- hclust(dist((Ar.bim.gsea.term.sim)))
+plot((Ar.bim.gsea.term.sim.hclust))
+# report items in each cluster at height 1.8
+GSEAReportClusters(Ar.bim.gsea.term.sim.hclust, h = 1)
+```
+
+GO enrichment for *A. carolinensis* **Intermediate** genes.
+
+```{r gsea_Ar_int, results = 'hide'}
+# Ar 'Int' genes
+Ar.geneList.int <- rep(0, length = length(Ap.geneList))
+names(Ar.geneList.int) <- names(Ap.geneList)
+Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int.transcripts))] <- 1
+Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
+```
+
+Clustering of enriched GO terms for *A. carolinensis* **Intermediate** genes.
+
+```{r gsea_Ar_int_cluster}
+# similarity among enriched GO terms
+Ar.int.gsea.term.sim <- termSim(Ar.int.gsea$GO.ID, Ar.int.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ar.int.gsea.term.sim) <- Ar.int.gsea$Term
+colnames(Ar.int.gsea.term.sim) <- Ar.int.gsea$Term
+# distance matrix
+Ar.int.gsea.term.sim.hclust <- hclust(dist((Ar.int.gsea.term.sim)))
+plot((Ar.int.gsea.term.sim.hclust))
+# report items in each cluster
+GSEAReportClusters(Ar.int.gsea.term.sim.hclust, h = 1.8)
+```
+
 ```{r gsea_combine, eval = FALSE}
@@
 
-ggplot(resp.TPM.dt.sub.pred["100348|*|comp3450455_c0_seq1"], aes(x=val, y=pTPM.scaled, group=Transcript)) +
+png("RxN_example.png")
+ggplot(resp.TPM.dt.sub.pred["100348|*|comp3450455_c0_seq1"], aes(x=val, y=pTPM.scaled, group=Transcript, color = Transcript)) +
   geom_line() + 
   facet_grid(. ~ colony) +
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
+  labs(y = "Expression (scaled)", x = expression("Temperature ", degree, "C")) +
+  theme(legend.position = "none")
+dev.off()
 

### 2014-07-15 doc/ApTranscriptome_manuscript.Rmd
Updates to ms

@@
 ---
-title: "Thermal reactionome of a temperate ant species"
+title: "Pattners of stress tolerance versus resistance in genome-wide expression data
+  of sympatric ant species"
 author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
@@
 output:
-  word_document: default
   pdf_document:
@@
   html_document: null
+  word_document: default
 bibliography: ApTranscriptome.bib
@@
 The study of whole organism gene expression (e.g.
-transcriptomics) has rapidly increased due to improved sequencing technology (RNAseq).
+transcriptomics) has rapidly increased due to improved RNA sequencing (RNAseq) technology.
 To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
@@
 An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014].
-The authors concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments.
+Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments.
 Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression.
-To overcome this limitation, we use a novel method to characterize the reaction norm of all genes in the transcriptome (e.g.
-the reactionome) in response to thermal stress.
+
+For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [cite..] and provide insight on evolutionary potential.
+However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [cite...] and miss genes with small expression changes that have large phenotypic effects [cite...], thus limiting their usefullness in this context.
+Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes is over-estimated while many causal genes of small effect are missed [@beavis; @larson2013].
+Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues.
+Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction.
+Analog
+
+In this study, we apply a novel method to characterize the reaction-norm of all genes in the transcriptome (e.g.
+the reactionome) in response to thermal stress to overcome the above limitations.
 Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach.
+Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially-divergent selection has changed gene expression patterns.
 
@@
 library(topGO)
+library(RColorBrewer)
 ```
@@
 
+```{r mosaic_plot, echo=FALSE, results='hide'}
+tiff("mosaic_plot.tiff")
+mosaic.colors <- brewer.pal(5, "Blues")
+par(mar = c(5.1, 3.1, 2.1, 7.5))
+mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
+labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
+text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
+text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
+text(.2, .6, "2,594")
+text(.85, .03, "223", cex = .75)
+legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
+dev.off()
+```
+
+![Mosaic plot of the number of thermally-responsive genes by expression type.
+The width of the columns is proportional to the number of genes for each expression type for *A.
+carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A.
+picea.* The rows are colored by expression type for *A.
+picea*.
+The number of genes are labeled in two boxes for scale.](mosaic_plot.tiff)
+
 We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$padj < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate.
 Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction.
-The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1).
+The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).
 The density function for the temperature of maximum expression (Fig.
@@
 1, Table 1).
-
-```{r expression_type_table, echo=FALSE, results='asis'}
-pandoc.table(tt3, style="rmarkdown", split.tables = Inf, caption = "Number of thermally-responsive genes in each expression category for each species.
-*A.
-carolinensis* in rows and *A.
-picea* in columns.")
-```
-
-```{r exp_plot, echo=FALSE, warning = FALSE, fig.cap = "Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
-ggplot(Ap.df, aes(x=max.val, fill=colony, colour = colony)) +
-  geom_density(alpha = 0.2, position = "identity") +
-  scale_y_continuous(name = "Density") +
-  scale_x_continuous(name = "Temperature of maximum expression")
-```
 

### 2014-07-16 ApTranscriptome_TR.Rmd
Cleaning up GSEA enrichment analysis and clustering for ms

@@
 # save results to file
-write.table(file = paste(resultsdir, "Ap-responsive-transcripts", Sys.Date(), ".txt", sep = ""), Ap.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
+write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), Ap.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
 ```
@@
 # T-test comparing standard deviation of expression between colonies
-(t.thermbreadth <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
+(t.varint <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
 ```
@@
 # t-test to compare standard deviation of 'Bimodal' transcripts between colonies
-(t.themsens <- t.test(Ar.bim.sd, A22.bim.sd))
+(t.varbim <- t.test(Ar.bim.sd, A22.bim.sd))
 ```
@@
 Ap.T_on <- rbind(A22.high.T_on, Ar.high.T_on, A22.low.T_on, Ar.low.T_on)
-
-# low plot
-png("T_on_low.png")
-T_on_low_plot <- ggplot(rbind(A22.low.T_on, Ar.low.T_on), aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+Ap.T_on$type <- factor(Ap.T_on$type, levels = c("Low", "High"))
+Ap.T_on$Species <- ifelse(Ap.T_on$colony == "ApVT", "A. picea", "A. carolinensis")
+
+# combined plot
+T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
+    facet_grid(. ~ type) +
     geom_histogram(position = "dodge", binwidth = 3) +
@@
     theme(axis.title = element_text(size = rel(2))) +
-    theme(axis.text = element_text(size = rel(1.2))) +
-    theme(legend.position = "none")
-print(T_on_low_plot)
-dev.off()
-
-# high plot
-png("T_on_high.png")
-T_on_high_plot <- ggplot(rbind(A22.high.T_on, Ar.high.T_on), aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
-    geom_histogram(position = "dodge", binwidth = 3) +
-    geom_density(adjust = 3, fill = NA, size = 2) +
-    labs(x = "Temperature of gene activation", y = "Density") +
-    theme(axis.title = element_text(size = rel(2))) +
-    theme(axis.text = element_text(size = rel(1.2))) +
-    theme(legend.position = "none")
-print(T_on_high_plot)
-dev.off()
-
-# combined plot
-T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
-    geom_histogram(position = "dodge", binwidth = 3) +
-    geom_density(adjust = 3, fill = NA, size = 2) + 
-    labs(x = "Temperature of gene activation", y = "Density") +
-    theme(axis.title = element_text(size = rel(2))) +
-    theme(axis.text = element_text(size = rel(1.2))) +
-    theme(legend.position = "none")
+    theme(axis.text = element_text(size = rel(1.2)))
+#    theme(legend.position = "none")
 print(T_on_plot)
@@
 # report items in each cluster
-GSEAReportClusters(Ap.high.hclust, h = 1)
-write.table(unlist(GSEAReportClusters(Ap.high.hclust, h = 0.6)), file = "Ap_high_hclust.txt")
+Ap.high.hclust.report <- GSEAReportClusters(Ap.high.hclust, h = 1)
+
+# add cluster information
+Ap.high.gsea$Cluster <- NA
+
+for(i in 1:length(Ap.high.hclust.report)) {
+    Ap.high.gsea[Ap.high.hclust.report[[i]], "Cluster"] <- i
+}
 ```
@@
 
-# report items in each cluster at height 2
-(Ap.low.hclust.report <- GSEAReportClusters(Ap.low.hclust, h = 2))
+# report items in each cluster
+Ap.low.gsea.cluster <- Ap.low.hclust.report <- GSEAReportClusters(Ap.low.hclust, h = 1.5)
+
+# add cluster information
+Ap.low.gsea$Cluster <- NA
+
+for(i in 1:length(Ap.low.hclust.report)) {
+    Ap.low.gsea[Ap.low.hclust.report[[i]], "Cluster"] <- i
+}
 ```
@@
 
-# report items in each cluster at height 1.5
-(Ap.bim.hclust.report <- GSEAReportClusters(Ap.bim.hclust, h = 1.5))
+# report items in each cluster
+Ap.bim.hclust.report <- GSEAReportClusters(Ap.bim.hclust, h = 1.5)
+
+# add cluster information
+Ap.bim.gsea$Cluster <- NA
+
+for(i in 1:length(Ap.bim.hclust.report)) {
+    Ap.bim.gsea[Ap.bim.hclust.report[[i]], "Cluster"] <- i
+}
 ```
@@
 
-# report items in each cluster at height 1.5
+# report items in each cluster
 (Ap.int.hclust.report <- GSEAReportClusters(Ap.int.hclust, h = 1.5))
-```
+
+# add cluster information
+Ap.int.gsea$Cluster <- NA
+
+for(i in 1:length(Ap.int.hclust.report)) {
+    Ap.int.gsea[Ap.int.hclust.report[[i]], "Cluster"] <- i
+}
+```
+
+```{r Ap_gsea_intersect_save, echo = FALSE}
+Ap.high.gsea$Type <- "High"
+Ap.low.gsea$Type <- "Low"
+Ap.bim.gsea$Type <- "Bimodal"
+Ap.int.gsea$Type <- "Intermediate"
+Ap.gsea.intersect <- rbind(Ap.low.gsea, Ap.int.gsea, Ap.high.gsea, Ap.bim.gsea)
+
+Ap.gsea.intersect <- Ap.gsea.intersect[,c("Type", "GO.ID", "Term", "Cluster", "Annotated", "Significant", "Expected", "parentchild")]
+colnames(Ap.gsea.intersect)[which(colnames(Ap.gsea.intersect) == "parentchild")] <- "P"
+
+write.csv(Ap.gsea.intersect, file = paste(resultsdir, "Ap_gsea_intersect_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
+```
+
+GSEA for genes that are *Intermediate* in A. carolinensis and *Bimodal* in A. picea
 
@@
 
+Cluster by semantic similarity
+
 ```{r gsea_Ar_int_A22_bim_cluster}
@@
 ```
+
+GSEA for genes that are *Intermediate* in A. carolinensis and *Low* in A. picea
 
@@
 ```
-
 Clustering of enriched GO terms for *A. picea* **High** genes.
@@
 
+Combine gene set enrichment analysis for each species and category into single table.
+
 ```{r gsea_combine, eval = FALSE}
@@
 # combine
-Ap.setdiff.gsea <- rbind(A22.gsea, Ar.gsea)
+Ap.gsea.union <- rbind(A22.gsea, Ar.gsea)
 # reorder
-Ap.setdiff.gsea <- Ap.setdiff.gsea[,c("Species", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
-colnames(Ap.setdiff.gsea)[8] <- "P"
-
-write.csv(Ap.setdiff.gsea, file = paste(resultsdir, "Ap_setdiff_GSEA_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
-```
-
-```{r gsea_setdiff_table, echo = FALSE, results = 'asis', eval = FALSE}
-#pandoc.table(Ap.setdiff.gsea[ , 1:4], caption = "Enriched GO terms found in one species but not the other for each thermal-response category")
-print(xtable(Ap.setdiff.gsea[ , 1:4], comment = getOption("xtable.comment", FALSE)))
-```
-
-```{r GOSemSim}
-# similarity among enriched GO terms
-A22.low.GO.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = "fly")
-# distance matrix
-dist.A22.low.GO.term.sim <- dist(A22.low.GO.term.sim)
-plot(hclust(dist(A22.low.GO.term.sim)))
-
-A22.high.GO.term.sim <- termSim(A22.high.gsea$GO.ID, A22.high.gsea$GO.ID, ont = "BP", organism = "fly")
-# distance matrix
-plot(hclust(dist(A22.high.GO.term.sim)))
+Ap.gsea.union <- Ap.gsea.union[,c("Species", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
+colnames(Ap.gsea.union)[8] <- "P"
+
+write.csv(Ap.gsea.union, file = paste(resultsdir, "Ap_gsea_union", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
 ```
@@
 ```{r shiny_file, eval=FALSE, echo=FALSE}
-shiny.dt <- TPM.dt.sub[Ap.response.type$Transcript]
+shiny.dt <- TPM.dt.sub[responsive.lms.ann.type[,list(Transcript, best.hit.to.nr.1, GO.Biological.Process.1)]]
 
@@
 # save RData for manuscript file
-save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, Ap.gsea1, Ap.setdiff.gsea, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.themsens, t.thermbreadth, file = "ms.RData")
+save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.intersect, Ap.gsea.union, file = "ms.RData")
 ```

### 2014-07-16 doc/ApTranscriptome_manuscript.Rmd

@@
 ---
-title: "Pattners of stress tolerance versus resistance in genome-wide expression data
+title: "Patterns of stress tolerance versus resistance in genome-wide expression data
   of sympatric ant species"
@@
 
-In this study, we apply a novel method to characterize the reaction-norm of all genes in the transcriptome (e.g.
-the reactionome) in response to thermal stress to overcome the above limitations.
+In this study, we apply a novel method to characterize the reaction norm of all genes in the transcriptome (e.g.
+the reactionome) in response to thermal extremes to overcome the above limitations.
 Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach.
-Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially-divergent selection has changed gene expression patterns.
+Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.
 
@@
 Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
-As ectothermic species, ants experience a wide range of temperature conditions.
-Throughout the year, ant species persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food.
-Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.
+Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food.
+*A.
+carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A.
+picea* [cite...].
+Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species.
 
-To explore the molecular toolkit used by *Aphaenogaster* to cope with thermal stress, we performed a whole organism RNA sequencing study.
-Our specific goals were to 1) generate the set of all genes (i.e.
-transcriptome) expressed by *Aphaenogaster* across physiologically-relevant temperatures, 2) identify genes that have responsive expression across this thermal gradient (i.e.
-thermal reactionome), 3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures.
+To explore differences in the thermal reactionomes between *A.
+picea* and *A.
+carolinensis*, we performed a whole organism RNA sequencing study spanning temperatures from 0° to 38.5°C.
+We identified genes that have responsive expression across this thermal gradient (i.e.
+thermal reactionome), compared the patterns
+
+3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures.
 We expected to observe candidate genes in the *response to stress* Gene Ontology (GO) category including *heat shock proteins* and ...
@@
 
-Finally, to test whether the temperature at which thremally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression.
+Finally, to test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression.
 We grouped the *High* genes along with the high temperature end of the *Bimodal* genes, and did the same for *Low* genes.
@@
 text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
-text(.2, .6, "2,594")
+text(.15, .7, "2,594")
 text(.85, .03, "223", cex = .75)

### 2014-07-16 doc/ApTranscriptome_manuscript.Rmd
Updates to ms

@@
 
+### Transcriptome assembly
+
 The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp.
@@
 Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage.
-DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts.
+We removed 5,675 genes identified as known contaminants, leaving 99,861 clean transcripts.
 Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
 
-```{r mosaic_plot, echo=FALSE, results='hide'}
-tiff("mosaic_plot.tiff")
+```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type.
+The width of the columns is proportional to the number of genes for each expression type for *A.
+carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A.
+picea.* The rows are colored by expression type for *A.
+picea*.
+The number of genes are labeled in two boxes for scale."}
 mosaic.colors <- brewer.pal(5, "Blues")
@@
 legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
-dev.off()
 ```
 
-![Mosaic plot of the number of thermally-responsive genes by expression type.
-The width of the columns is proportional to the number of genes for each expression type for *A.
-carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A.
-picea.* The rows are colored by expression type for *A.
-picea*.
-The number of genes are labeled in two boxes for scale.](mosaic_plot.tiff)
+### Thermally-responsive genes
 
-We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$padj < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate.
-Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction.
-The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).
-The density function for the temperature of maximum expression (Fig.
-1) showed that the majority of responsive genes were in the *Low* category for both species, with *A.
-picea* having more *Low* genes than *A.
-carolinensis*.
-More genes had *Intermediate* expression in *A.
-carolinensis* than in *A.
-picea*, while the two species had a similar number of *High* genes (Fig.
-1, Table 1).
-
-The reaction norms of thermally-responsive genes between species differed by mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A.
-carolinensis* than *A picea*, and mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A.
-picea*.
-Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species.
-Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.
-
-We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species.
-Instead, we found that for genes expressed at *High* temperatures in *A.
-picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), genes expressed at *Low* temperatures in *A.
-carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) and *Intermediate* genes in both (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), basal expression near optimum temperatures of 19.5°C was always higher in *A.
+We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR.
+Of these, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species.
+The number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
+1).
+In both species, the majority of genes were in the *Low* category (Fig.
+1).
+For *A.
+carolinensis*, this was followed by *Intermediate*, *High* and then *Bimodal*, whereas for *A.
+picea* the second largest category was *Bimodal* followed by *High* and finally *Intermediate* (Fig.
+1).
+There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A.
+carolinensis* but not *A.
+picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A.
+picea* but not *A.
 carolinensis*.
 
-Overall greater thermal tolerance of one species would be indicated by finding that gene expression is stable across a greater portion of the thermal gradient.
-For *Bimodal* expressed genes, we found no difference (*t*~df=1406~ = 0.31, *P* = 0.76) between species in the variance of the reaction norms for thermally-responsive genes.
+### Species comparisons
+
+To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for *High* and *Low* genes, and the variance in expression for the *Intermediate* and *Bimodal* genes.
+The *High* genes in *A.
+carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A.
+picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
+2).
+For genes with increased expression at *Low* temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+2).
+For *Bimodal* expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`).
 In contrast, *Intermediate* genes from *A.
 carolinensis* had a broader thermal range of stable gene expression than those from *A.
-picea* (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).
-In addition, the species differed in the temperatures at which thermally-responsive genes were activated.
-For genes with increased expression at *High* temperatures, we found that *A.
-carolinensis* genes were activated on average at 33.8°C compared to 32.9°C for *A.
-picea* (Fig.
-2).
-For genes with increased expression at *Low* temperatures, *A.
-carolinensis* genes were activated on average at 12.8°C compared to 13°C for a *A.
-picea* (Fig.
-2).
+picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`).
 
@@
 carolinensis* (red) species"}
-T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
+
+T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
+    facet_grid(. ~ type) +
     geom_histogram(position = "dodge", binwidth = 3) +
@@
 
-### Molecular comparisons
+We looked for evidence of genetic assimilation by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
+carolinensis* or high temperatures in *A.
+picea*.
+Contrary to expectations, for *Low* genes in *A.
+carolinensis* gene expression at intermediate temperatures was lower in *A.
+picea* than *A.
+carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) .
+Moreover, for *High* genes in *A.
+picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and *Intermediate* genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A.
+carolinensis*.
+
+In addition to shifts in position, we quantified the differences between species in the mean and shape of the reaction norms of thermally-responsive genes.
+Species differed by mean expression (*t*~df=`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A.
+carolinensis* than *A picea*, and mean slope (*t*~df=`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A.
+picea*.
+Curvature (*t*~df=`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~df=`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species.
+Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.
+
+### Gene set enrichment analysis
 
 As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately.
-Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "High"), ])` enriched (*P* < 0.01) for *High* transcripts, including categories such as `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][1]` and `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][6]` (Table S1).
-For *Low* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Low"), ])` GO terms were enriched including some implicated in stress response (e.g.
-`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][18]`) and cold tolerance (e.g.
-`r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][6]`) (Table S1).
-For *Bimodal* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g.
+Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* transcripts.
+Clustering of terms by semantic similarity yielded 3 clusters, with "regulation of biosynthetic/metabolic process" being the largest group.
+
+For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched.
+These terms grouped into 6 clusters, including "protein metabolic process", "lipid metabolic process", and "".
+
+including some implicated in stress response (e.g.
+`r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][18]`) and cold tolerance (e.g.
+`r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][6]`) (Table S1).
+For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g.
 regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1).
-Finally, for *Intermediate* genes `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).
+Finally, for *Intermediate* genes `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).
 
@@
 Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
-
-```{r rxn_plot, echo = FALSE, warning = FALSE, fig.cap = "Gene expression in transcripts per million (TPM) for thermally-reactive genes in the GO category *response to stress* for each colony"}
-rxnplot <- ggplot(resp.TPM.dt.sub[GO0006950.responsive$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) +
-  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
-  facet_grid(. ~ colony2) +
-  scale_y_continuous(name = "Expression (scaled)") +
-  scale_x_continuous(name = expression(paste("Temperature", degree, "C")))
-print(rxnplot)
-```
 

### 2014-07-17 doc/ApTranscriptome_manuscript.Rmd
completed draft results section

@@
 title: "Patterns of stress tolerance versus resistance in genome-wide expression data
-  of sympatric ant species"
+  of parapatric ant species"
 author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
@@
 
-### Transcriptome assembly
+### Thermally-responsive genes
 
@@
 
-### Thermally-responsive genes
-
 We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR.
 Of these, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species.
+Thus, at a minimum ...
+of genes are thermally-responsive.
+However, the closest related ant species with a sequenced genome has only ...
+genes.
+If we use this as
+
 The number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
@@
 As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately.
-Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* transcripts.
-Clustering of terms by semantic similarity yielded 3 clusters, with "regulation of biosynthetic/metabolic process" being the largest group.
+First, we looked at the genes shared in each expression category (e.g.
+the intersection) for each species (Table S1).
+Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* genes.
+Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".
+For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched.
+These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis".
+For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched.
+These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process".
+For *Intermediate* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched.
+These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".
+Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.
 
-For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched.
-These terms grouped into 6 clusters, including "protein metabolic process", "lipid metabolic process", and "".
+We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, *Intermediate*, in *A.
+carolinensis* to having increased expression in *Low* temperatures in *A.
+picea*.
+This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization".
+For *Intermediate* genes in *A.
+carolinensis* that shifted *Bimodal* in *A.
+picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification".
+These results indicate that in response to thermal extremes, the warmer-climate species, *A.
+carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A.
+picea*, which increases metabolism.
 
-including some implicated in stress response (e.g.
-`r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][18]`) and cold tolerance (e.g.
-`r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][6]`) (Table S1).
-For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g.
-regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1).
-Finally, for *Intermediate* genes `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).
+### Candidate genes
 
-In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Figure 2).
+We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig.
+3).
 Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
@@
 As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
-
-We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock.
-For *High* genes, there were 4 enriched GO terms in *A.
-picea* and 12 in *A.
-carolinensis*.
-For *Low* genes, there were 40 enriched GO terms in *A.
-picea* and 23 in *A.
-carolinensis*.
-For *Intermediate* genes, there were 3 enriched GO terms in *A.
-picea* and 54 in *A.
-carolinensis*.
-For *Bimodal* genes, there were 12 enriched GO terms in *A.
-picea* and 10 in *A.
-carolinensis*.
 
@@
 In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively.
-We found that nearly 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress.
+We found that at least 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress.
 The patterns of gene expression between species differed substantially, with more down-regulation of gene expression in *A.

### 2014-07-17 doc/ApTranscriptome_manuscript.Rmd
completed Discussion draft

@@
 For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (*High*), low (< 10°C) temperatures (*Low*), intermediate (10 - 30°C) temperatures (*Intermediate*), both high and low temperatures (*Bimodal*) or were not thermally-responsive in that species (*NotResp*).
-For the *Bimodal* group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
+We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
 
@@
 To gain greater insight on the ecological genetics of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction.
-For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Responsive*.
+For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category *Not Responsive*.
 To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.
@@
 carolinensis*.
-To test this with our data, for each *Intermediate* expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
+To test this with our data, for each **Intermediate** expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
 point of maximum expression had highest probability of being sampled).
 We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species.
-Conversely, we predicted that the thermal sensitivity of *Bimodal* expressed genes would be greater in *A.
+Conversely, we predicted that the thermal sensitivity of **Bimodal** expressed genes would be greater in *A.
 carolinensis* than *A.
 picea*.
-We performed the same analysis as with the *Intermediate* transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern.
-Thus, finding one species had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.
+We performed the same analysis as with the **Intermediate** transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern.
+Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity.
 
 Finally, to test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression.
-We grouped the *High* genes along with the high temperature end of the *Bimodal* genes, and did the same for *Low* genes.
-We then tested if the temperature at which the *High* or *Low* group differed among the two species.
+We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes.
+We then tested if the temperature at which the **High** or **Low** group differed among the two species.
 
@@
 library(data.table)
-library(pander)
 library(ggplot2)
+library(RColorBrewer)
 library(topGO)
-library(RColorBrewer)
 ```
@@
 We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR.
-Of these, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species.
-Thus, at a minimum ...
-of genes are thermally-responsive.
-However, the closest related ant species with a sequenced genome has only ...
-genes.
-If we use this as
+Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive.
+However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012], and assuming the number is similar for *Aphaenogaster*, this suggests that up to 50% of the genome could be thermally-responsive.
+The reaction norms for each responsive gene can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
 
-The number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
+Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species.
+These differences were not simply of magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
 1).
-In both species, the majority of genes were in the *Low* category (Fig.
+In both species, the majority of genes were in the **Low** category (Fig.
 1).
 For *A.
-carolinensis*, this was followed by *Intermediate*, *High* and then *Bimodal*, whereas for *A.
-picea* the second largest category was *Bimodal* followed by *High* and finally *Intermediate* (Fig.
+carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A.
+picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig.
 1).
@@
 
-To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for *High* and *Low* genes, and the variance in expression for the *Intermediate* and *Bimodal* genes.
-The *High* genes in *A.
+To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes, and the variance in expression for the **Intermediate** and **Bimodal** genes.
+The **High** genes in *A.
 carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A.
@@
 2).
-For genes with increased expression at *Low* temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+For genes with increased expression at **Low** temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
 2).
-For *Bimodal* expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`).
-In contrast, *Intermediate* genes from *A.
+For **Bimodal** expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`).
+In contrast, **Intermediate** genes from *A.
 carolinensis* had a broader thermal range of stable gene expression than those from *A.
@@
 
-```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures.
+```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at **Low** and **High** temperatures.
 Bars are Probability density function of peak expression for all thermally-responsive in *A.
@@
 picea*.
-Contrary to expectations, for *Low* genes in *A.
+Contrary to expectations, for **Low** genes in *A.
 carolinensis* gene expression at intermediate temperatures was lower in *A.
@@
 carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) .
-Moreover, for *High* genes in *A.
-picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and *Intermediate* genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A.
+Moreover, for **High** genes in *A.
+picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A.
 carolinensis*.
@@
 the intersection) for each species (Table S1).
-Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* genes.
+Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for **High** genes.
 Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".
-For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched.
+For **Low** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched.
 These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis".
-For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched.
+For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched.
 These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process".
-For *Intermediate* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched.
+For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched.
 These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".
@@
 
-We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, *Intermediate*, in *A.
-carolinensis* to having increased expression in *Low* temperatures in *A.
+We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A.
+carolinensis* to having increased expression in **Low** temperatures in *A.
 picea*.
 This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization".
-For *Intermediate* genes in *A.
-carolinensis* that shifted *Bimodal* in *A.
+For **Intermediate** genes in *A.
+carolinensis* that shifted **Bimodal** in *A.
 picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification".
@@
 
-As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
-
 ## Discussion
 
-- many genes thermally responsive
-- thermal reaction norms differ between species
-  - *A. carolinensis* limited future potential to adapt to heat stress
-- more genes activated in response to cold shock --> greater adaptability to cold
-- differences among species 
-  - *A. carolinensis* tolerates thermal stress by shutting down biological processes
-  - in *A. picea* the *Intermediate* genes from *A. carolinensis* are over-represented in *Low* and *Intermediate* categories
-  - potential explanation is that *A.
-carolinensis* tolerates thermal stress by shutting-down metabolism and relies on other mechanisms (behavior, protein stability) whereas *A.
-picea* activates protective genes which requires metabolism...
-- comparison of gene terms with other species (Sara?)
+Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species.
+In this study, we applied a novel regression design to characterize the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively, that are known to differ in their thermal tolerances [@warren2013].
+We found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.
 
-- regression approach allows us to detect patterns not evident from ANOVA design
+The patterns of gene expression between species differed substantially, with more down-regulation in *A.
+carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A.
+picea* in response to thermal extremes (**High** and **Low** genes; Fig.
+2).
+The results of the gene set enrichment analysis give an indication of what might drive this pattern.
+The genes that are up-regulated at cold temperatures (**Low**) in *A.
+picea* but shut-down **Intermediate** by *A.
+carolinensis* include metabolism, transport and regulation GO clusters.
+Thus, *A.
+carolinensis* shuts-down metabolism to a lesser extent than *A.
+picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress.
+In contrast, *A.
+picea* increases metabolism to a greater extent than *A.
+carolinensis*, potentially to mount a greater molecular response to thermal stress.
+Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A.
+picea*, but only 4 do in *A.
+carolinensis*.
+This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically-enduced resistance.
 
--------------------------------------
-
-Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species.
-While considerable research has focused on a few key genes (e.g.
-heat shock proteins) involved in thermal stress response [@krebs1997; @gaitainespitia2013], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness.
-In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively.
-We found that at least 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress.
-The patterns of gene expression between species differed substantially, with more down-regulation of gene expression in *A.
-carolinensis* and more up-regulation of gene expression in *A.
-picea* in response to thermal stress (Fig.
+Given the differences in the number of thermally-responsive genes between the two species (Fig.
+1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species (Fig.
+1, Fig.
 2).
-
-Given the differences in the number of thermally-responsive genes between the two species (Table 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species, with greater variation between species at low temperatures (Fig.
-1).
-A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures.
+A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [cite...kingsolver, chown].
 This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addobediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
@@
 phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits.
-Thus, our results indicate that, when coupled with phenotypic data, transcriptomic studies can provide insight on adapatability.
-For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal].
+Thus, when coupled with phenotypic data, our results demonstrate that transcriptomic studies can provide insight on evolutionary potential.
+For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal2012].
 
-Depending on the frequency with which organisms experience stressful conditions, the reaction norms of protective genes may differ considerably.
-Organisms that frequently experience stress may have high basal levels of expression with little change, while organisms that less frequently experiences stress would have lower basal levels with large changes in response to environmental stress [@cite...].
-For *Aphaenogaster*, we predicted that genes up-regulated to cold or hot temperatures in one species might be expressed at higher levels near optimal growth temperatures in the other species to confer greater basal stress tolerance.
-This prediction was not supported, as we found greater expression near optimal temperatures in *A.
-carolinensis* for *High* genes in *A.
-picea*, as predicted, but also for *Low* genes in *A.
-carolinensis* and *Intermediate* genes in both species.
-As *A.
-carolinensis* is from a warmer environment, a likely explanation for this is that this species simply has a higher metabolic rate and thus greater gene expression across it's thermal range.
-The observation that there are more *Intermediate* genes, which are shut-down under thermal stress, in *A.
-carolinensis* supports this explanation.
-...cite some coral literature....
+For species or populations that routinely, but unpredictably, experience stressful environments, genetic assimilation is the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed  [@barshis2013].
+However, in contrast to these expectations, we found lower basal gene expression in *A.
+picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A.
+carolinensis*.
+We also observed greater basal expression in *A.
+carolinensis* for **High** genes in *A.
+picea* as well as for genes that were **Intermediate** in both spcies.
+A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A.
+carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.
 
@@
 
-### Molecular toolkit of thermal tolerance
-
-REANALYZING...
-
 ### Caveats/limitations
@@
 As with all *de novo* transcriptomes, our work has some important caveats.
-First, given that all sequenced ant genomes to date have about 18,000 genes [@smith...], our transcriptome is highly fragmented.
+First, given that all sequenced ant genomes to date have about 18,000 genes [@gadau2012], our transcriptome is highly fragmented.
 Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts.
@@
 carolinensis*, to date.
-We found that a large proportion of the transcriptome is thermally-responsive.
-Transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures.
-...differences among species...*A.
-carolinensis* greater organismal tolerance and shuts down metabolism under thermal stress...*A.
-picea* less tolerant and activates protective proteins requiring metabolism...
+Using a novel regression approach, we found that a large proportion of the transcriptome is thermally-responsive.
+These data indicate that the two species have different responses to thermal extremes.
+*A.
+picea* responds by increasing metabolism and activating heat shock and other protective molecules, while *A.
+carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolernace.
+For both species, transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures.
 These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.

### 2014-07-18 doc/ApTranscriptome_manuscript.Rmd
Completed first draft manuscript sent to co-authors

@@
 output:
+  word_document: default
   pdf_document:
     fig_caption: yes
+    includes:
+      in_header: header.tex
+    keep_tex: yes
     latex_engine: xelatex
+    number_sections: yes
+    toc_depth: 2
   html_document: null
-  word_document: default
+includes: header.tex
 bibliography: ApTranscriptome.bib
@@
 
-## Abstract
+\linenumbers
+\tableofcontents
 
-- When ants are exposed to different temperatures, lots of genes change expression. 
-- The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress. 
-- Comparisons between species indicate that *A.
-carolinensis*, which has greater basal gene expression and physiological thermal tolerance, tolerates thermal stress by shutting down gene expression.
+\newpage
+
+# Abstract
+
+Thermal tolerance is a key trait governing the distributions of species, their abundances, and underlies their future responses to climate change.
+However, the genetic basis of thermal tolerance is poorly understood for most species.
+In this study, we present a novel regression approach to characterize the thermal reaction norm for all genes in an organism's transcriptome.
+We apply this approach to parapatric ant species in the eastern United States; *A.
+carolinensis* and *A.
+picea*, which have warmer and colder distributions, respectively.
+Using this approach, we find that at least 10% of genes are thermally responsive.
+In both species, more genes are up-regulated in response to cold than warm temperatures, suggesting that there is less adaptive potential to heat than cold stress.
+Between species, *A.
+carolinensis* shuts down more genes in response to thermal extremes, whereas *A.
+picea* activates more genes in response to cold and hot temperatures.
+The genes that are shut down in *A.
+carolinensis* but up-regulated in *A.
+picea* included many metabolism, cellular organization and repair gene ontology terms.
+We conclude that these species have different mechanisms for responding to thermal stress.
 *A.
-picea* tolerates thermal stress by activating protective genes
+picea* resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas *A.
+carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress.
+In summary, this study demonstrates how we can gain insight on the genetic basis of local adaption to climate using a reaction norm approach with transcriptomic data.
 
-## Introduction
+# Introduction
 
@@
 transcriptomics) has rapidly increased due to improved RNA sequencing (RNAseq) technology.
-To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
+To date, RNAseq studies have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
 However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature.
@@
 
-For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [cite..] and provide insight on evolutionary potential.
-However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [cite...] and miss genes with small expression changes that have large phenotypic effects [cite...], thus limiting their usefullness in this context.
-Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes is over-estimated while many causal genes of small effect are missed [@beavis; @larson2013].
+For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014].
+However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [] and may not be directly informative about organismal status [@feder2005].
+Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes are over-estimated while many causal genes of small effect are missed [@beavis1994; @larsson2013].
 Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues.
-Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction.
-Analog
+Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].
+Analogously, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient (e.g.
+the thermal reactionome).
+Instead of focusing on individual genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.
+We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
+picea* [@umphrey1996], which have more southern and northern distributions, respectively.
 
-In this study, we apply a novel method to characterize the reaction norm of all genes in the transcriptome (e.g.
-the reactionome) in response to thermal extremes to overcome the above limitations.
-Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach.
-Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.
-
-We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
-picea* [@umphrey1996], which have more southern and northern distributions, respectively.
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
 Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food.
@@
 carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A.
-picea* [cite...].
+picea* [@diamond2013].
 Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species.
+For example, thermally-responsive genes may shift the temperature at which they are up-regulated to higher temperatures in *A.
+carolinensis*, or colder temperatures in *A.
+picea*.
+Additionally, genes may change their sensitivity to thermal extremes.
+Further, through the process of genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A.
+carolinensis* may have increased basal expression in the colder-climate species *A.
+picea*.
 
-To explore differences in the thermal reactionomes between *A.
-picea* and *A.
-carolinensis*, we performed a whole organism RNA sequencing study spanning temperatures from 0° to 38.5°C.
-We identified genes that have responsive expression across this thermal gradient (i.e.
-thermal reactionome), compared the patterns
+Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance in these species.
+Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.
 
-3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures.
-We expected to observe candidate genes in the *response to stress* Gene Ontology (GO) category including *heat shock proteins* and ...
-in the thermally-responsive group.
+# Methods
 
-While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991].
-We compared the thermal reactionomes between two parapatric species of *Aphaenogaster*, *A.
-carolinensis* from North Carolina, USA and *A.
-picea* from Vermont, USA, to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance.
-Specifically, we examined differences between the species in (i) the number of thermally-responsive genes, (ii) the distribution of responsive genes into functional categories, (iii) basal expression near optimal temperatures for responsive genes and (iv) the thermal sensitivity of the transcriptome.
-*A.
-picea*, the northern species, experiences less chronic and acute heat stress than *A.
-carolinensis* [@diamond2013] we predicted that it would show more plasticity of gene expression in response to heat shock, and lower basal levels of expression for heat-tolerant genes.
-
-## Methods
-
-### Samples
+## Samples
 
@@
 
-### Sequencing, assembly and annotation
+## Sequencing, assembly and annotation
 
@@
 We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
-To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmeider2011] with the provided bacteria, virus, archae and human databases of contaminants.
+To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.
 
@@
 
-### Thermally-responsive genes
+## Thermally-responsive genes
 
-We quantified expression of each gene using the program `Sailfish` [@patro2013] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression.
+We quantified expression of each gene using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression.
 We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses.
@@
 
-To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
-Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
-In addition, we used the web-tool DAVID for ....
+## Species-level comparisons
 
-### Species-level comparisons
+We performed comparative analyses between the two species focusing on the genes that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of gene activation and sensitivity of the thermally-responsive genes.
+For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
+To test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression for each gene in each species.
+We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes.
+We then performed a *t*-test to determine if the mean temperature of gene activation differed between the two species for each group.
 
-To gain greater insight on the ecological genetics of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction.
-For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category *Not Responsive*.
-To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.
+To determine if the sensitivity of genes to thermal extremes differed between the species, we used the subset of **Intermediate** and **Bimodal** genes in each species.
+For each **Intermediate** gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
+point of maximum expression had highest probability of being sampled).
+We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species.
+Finding a greater standard deviation of expression would be indicative of less sensitivity to changes in temperature.
+For the **Bimodal** genes, we performed the same analysis as with the **Intermediate** genes, but with a greater standard deviation being indicative of a more sensitive gene (convex reaction norm) while a low standard deviation being a gene with greater inertia of expression in response to temperature changes (nearly flat).Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity for that species.
 
-With the same set of genes, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene.
+To test for genetic assimilation, we compared basal expression levels (expression at 19.5° C) for the subset of genes that were responsive to cold temperatures (**Low** genes) in the southern *A.
+carolinensis* species, or hot temperatures (**High** genes) in the northern *A.
+picea* species using non-parametric two-sided Wilcoxon tests.
+
+Finally, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene.
 Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix).
@@
 carolinensis*.
-
 Further, to quantify the relative contribution of changes in the mean, slope, curvature or wiggle on the overall differences in reaction norms, we defined the *Total* difference in the reaction norm for each gene as $\Delta T = \Delta M + \Delta S + \Delta C + \Delta W$.
@@
 
-To test if genes that were responsive to cold temperatures in the southern *A.
-carolinensis* species or hot temperatures in the northern *A.
-picea* species had higher basal expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.
+## Gene set enrichment analysis
 
-As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A.
-picea* species would have a broader thermal range of stable gene expression than *A.
-carolinensis*.
-To test this with our data, for each **Intermediate** expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
-point of maximum expression had highest probability of being sampled).
-We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species.
-Conversely, we predicted that the thermal sensitivity of **Bimodal** expressed genes would be greater in *A.
-carolinensis* than *A.
-picea*.
-We performed the same analysis as with the **Intermediate** transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern.
-Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity.
+To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the package `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
+Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
+We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.
 
-Finally, to test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression.
-We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes.
-We then tested if the temperature at which the **High** or **Low** group differed among the two species.
-
-### Data availability
+## Data availability
 
@@
 
-## Results
+# Results
 
@@
 
-### Thermally-responsive genes
+## Thermally-responsive genes
 
@@
 Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species.
-These differences were not simply of magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
+These differences were not only in magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
 1).
@@
 
-### Species comparisons
+## Species comparisons
 
@@
 
-### Gene set enrichment analysis
+## Gene set enrichment analysis
 
@@
 
-### Candidate genes
+**Table** Table of enriched GO terms for each expression category in each species.
+
+## Candidate genes
 
@@
 
-## Discussion
+# Discussion
 
@@
 picea* in response to thermal extremes (**High** and **Low** genes; Fig.
-2).
+1, 3).
 The results of the gene set enrichment analysis give an indication of what might drive this pattern.
@@
 
+```{r max_exp_PDF, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
+# reshape data
+Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")
+
+maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=Species)) + 
+  geom_density(alpha=0.2, position="identity") + 
+  labs(x = "Temperature of maximum expression", y = "Density")
+
+print(maxexplot)
+```
+
 Given the differences in the number of thermally-responsive genes between the two species (Fig.
 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species (Fig.
-1, Fig.
-2).
-A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [cite...kingsolver, chown].
-This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addobediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
-Moreover, in an experiment with 94 species of *Drosophila*, Kellerman et al [-@kellerman2012] showed that limited genetic variation among species (e.g.
+3).
+A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013].
+This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
+Moreover, in an experiment with 94 species of *Drosophila*, Kellermann et al [-@kellermann2012] showed that limited genetic variation among species (e.g.
 phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits.
 Thus, when coupled with phenotypic data, our results demonstrate that transcriptomic studies can provide insight on evolutionary potential.
-For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal2012].
+For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
 
-For species or populations that routinely, but unpredictably, experience stressful environments, genetic assimilation is the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed  [@barshis2013].
+For species or populations that routinely, but unpredictably, experience stressful environments, genetic assimilation is the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed [@pigliucci2006].
 However, in contrast to these expectations, we found lower basal gene expression in *A.
@@
 
-To date, transcriptomic studies have performed differential expression analysis between a few different conditions or environments (but see [@hodginsdavis2012]).
+To date, transcriptomic studies have performed differential expression analysis between a few different conditions or environments (but see [@hodgins-davis2012]).
 By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
@@
 
-### Caveats/limitations
-
 As with all *de novo* transcriptomes, our work has some important caveats.
@@
 Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change.
-Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@cite...].
-In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@cite...].
+Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@jensen2008; @kenkel2013].
+In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@oneil2014; @lemay2013].
 
-### Conclusions
+## Conclusions
 
-In this work, we have completed the first transcriptome for a temperate ant species, *Aphaenogaster picea* and *A.
-carolinensis*, to date.
-Using a novel regression approach, we found that a large proportion of the transcriptome is thermally-responsive.
-These data indicate that the two species have different responses to thermal extremes.
+In this work, we bring reaction norms to the genomic era by characterizing the thermal reactionomes of the temperate ant species, *Aphaenogaster picea* and *A.
+carolinensis*.
+We found that at least 10% of the transcriptome is thermally-responsive.
+For both species, more genes responded to cold than hot temperatures.
+In light of physiological data showing more variation in critical minimum than maximum temperatures, this suggests that these species and other ectotherms have less adaptive potential to hot than cold temperatures.
+Further, our results indicate that the two species have different responses to thermal extremes.
 *A.
@@
 carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolernace.
-For both species, transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures.
 These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.
 
-## References
+# References

### 2014-07-25 ApTranscriptome_TR.Rmd
updates

@@
 
-**June 16, 2014**
+**July 17, 2014**
 
@@
 plot(Ar.int.A22.low.gsea.term.sim.hclust)
-# report items in each cluster at height 1.8
-(Ar.int.A22.low.hclust.report <- GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.6))
+# report items in each cluster at height 1
+(Ar.int.A22.low.hclust.report <- GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.2))
 ```
@@
 # save RData for manuscript file
-save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.intersect, Ap.gsea.union, file = "ms.RData")
+save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.intersect, Ap.gsea.union, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
 ```

### 2014-07-25 ApTranscriptome_TR.Rmd
Updates. Change TR to download expression data rather than re-run sailfish as unclear if filtered fastq files will be archived.

@@
 
-In this technical report, which accompanies the manuscript **Thermal reactionome of a common ant species** (Stanton-Geddes et al., in press), we:
+In this technical report, which accompanies the manuscript **Patterns of stress tolerance versus resistance in genome-wide expression data of parapatric ant species** (Stanton-Geddes et al., in press), we:
 
@@
 2. identify thermally-responsive genes
-3. evaluate differences in the expression patterns between the two species
-3. perform gene set enrichment analysis of thermally-responsive genes for the two species
+3. evaluate differences in reaction norms of expression for each thermally-responsive gene between the two species
+4. perform gene set enrichment analysis of thermally-responsive genes for the two species
 
@@
 
+## Data ##
+
+The raw Illumina fastq files are available from the NCBI short read archive [link tbd].
 The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.
-
-## Data ##
-
-The Illumina fastq files are available from [https://minilims1.uvm.edu/BCProject-26-Cahan/downloads.html].
-The Trimmomatic filtered fastq files should be downloaded, uncompressed and moved to the appropriate directory using the following code.
-
-~~~
-# download filtered fastq files, uncompress and move
-wget --no-check-certificate https://minilims1.uvm.edu/BCProject-26-Cahan/_downloads/trimmomatic_output.tar.gz
-tar -zxvf trimmomatic_output.tar.gz
-mkdir -p data/
-mv ind_files data/.
-~~~
 
@@
 
-The Illumina reads were filtered using the program [Trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) `r citep("10.1093/nar/gks540")` to remove Ilumina adapter sequences and filter out bases with quality scores less than ??.
+The Illumina reads were filtered using the program [Trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) `r citep("10.1093/nar/gks540")` to remove Ilumina adapter sequences, trim bases with PHRED quality scores less than 15 in a 4 bp sliding window and remove final trimmed sequences with length less than 36 bp.
+The code used was
 
 ~~~
-TRIMMOMATIC CODE
+java -jar trimmomatic-0.30.jar PE -threads 40 -phred33 -trimlog trimmomatic.log sample.R1.fastq sample.R2.fastq sample.R1.trimmed.paired.fastq sample.R1.trimmed.unpaired.fastq sample.R2.trimmed.paired.fastq sample.R2.trimmed.unpaired.fastq ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:36
 ~~~
 
-This filtering yielded...
-
-These reads were combined and used in *de novo* transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`.
-Note that this required ???
-GB RAM and ???
-hours run-time and was run on the [Vermont Genetics Network](http://vgn.uvm.edu/) computing cluster.
+where "sample" was replaced by each sample. This filtering yielded 339,845,787 properly paired reads and 16,860,885 unpaired reads.
+
+Properly paired and unpaired reads passing the Trimmomatic filter were combined and used in de novo transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`.
+Unpaired reads resulting from Trimmomatic were concatenated to the set of left (R1) reads per Trinity usage documentation.
+Assembly on a single 40 CPU machine with 1TB of memory required 20 hours of compute time.
 
 ~~~
-TRINITY CODE
+Trinity.pl --seqType=fq --JM=90G —left=all.R1.trimmed.fastq —right=all.R2.trimmed.fastq --output=trinity --CPU=40 --inchworm_cpu=40 --bflyCPU=5
 ~~~
@@
 
-Then build the index of the assembly:
-
-```{r sailfish_index}
-system("sailfish index -t results/trinity-full/Trinity_cap3_uclust.fasta -o results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4")
-```
-
-Once this is done, quantify expression for the Trimmomatic filtered reads from each species-treatment sample separately. Note that for each sample, there are three four filtered read files:
+An index of the assembly is built with the command:
+
+~~~
+sailfish index -t results/trinity-full/Trinity_cap3_uclust.fasta -o results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4
+~~~
+
+Once this is done, expression is quantified for the Trimmomatic filtered reads from each species-treatment sample separately. Note that for each sample, there are four filtered read files:
 
@@
 
-Make a directory for the expression values
-
-```{r mk_sailfish_dir}
-system("mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust")
-```
-
-Then, for each sample, run the following command in the `results/trinity-full/` directory:
-
+~~~
+# make a directory for the expression values
+mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust
+# change to "trinity-full" directory
+cd results/trinity-full
+# for each sample, run the following command
 sailfish quant -i sailfish-index-Trinity-cap3-uclust -o sailfish-expression-Trinity-cap3-uclust/A22-0 -l "T=SE:S=U" -r A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq A22-0_ATCACG.paired.left.fastq A22-0_ATCACG.paired.right.fastq -p 4
+~~~
 
@@
 
-Or, using a loop in R:
-
-```{r sailfish_quant, results = 'hide'}
-# directory containing trimmed reads
-readdir <- "data/ind_files/" 
-# sailfish index directory
-sfindex <- "results/trinity-full/sailfish-index-Trinity-cap3-uclust"
-# sailfish expression directory
-sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
-
-# list of reads in each of four trimmed classes
-readlist <- list.files(readdir)
-paired.left <- readlist[grep(".\\.paired.left.fastq$", readlist)]
-paired.right <- readlist[grep("\\.paired.right.fastq$", readlist)]
-unpaired.left <- readlist[grep("unpaired.left.fastq$", readlist)]
-unpaired.right <- readlist[grep("unpaired.right.fastq$", readlist)]
-
-# Loop across each sample and quantify expression
-
-# NOTE - samples listed in same order as given by the above lists
-samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")
-
-for (j in 1:length(samples)) {
-    message("Start expression quantification for sample ", samples[j], ": ", Sys.time())
-    quantdir <- paste(sfexpressionroot, samples[j], "_quant", sep="")
-    samp.pos <- grep(paste(paste(samples[j], "_", sep="")), paired.left)
-    samp.paired.l <- paste(readdir, paired.left[samp.pos], sep="")
-    samp.paired.r <- paste(readdir, paired.right[samp.pos], sep="")
-    samp.unpaired.l <- paste(readdir, unpaired.left[samp.pos], sep="")
-    samp.unpaired.r <- paste(readdir, unpaired.right[samp.pos], sep="")
-    sailfishcmd <- paste("sailfish quant -i ", sfindex, " -o ", quantdir, " -l 'T=SE:S=U' -r ", samp.paired.l, " ", samp.paired.r, " ", samp.unpaired.l, " ", samp.unpaired.r, " -p 4", sep="")
-    #print(sailfishcmd)
-    system(sailfishcmd)
-    message("Done with expression quantification for sample ", samples[j], ": ", Sys.time(), "\n")
-}
-```
-
-This generated a directory for each sample
-
-`r list.files(sfexpressionroot)`
-
-and within each directory there are the following r:
-
-`r list.files(quantdir)`
-
-The file *quant_bias_corrected.sf* contains the following columns, following a number of header lines:
+These files are downloaded for convenience:
+
+~~~
+wget ...
+mkdir -p /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
+mv .. /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
+tar -zxvf ...
+~~~
+
+For each sample, there is a directory containting a file *quant_bias_corrected.sf*. This file has the following columns, following a number of header lines:
 
@@
 ```{r load_expression_data, eval=TRUE, echo=FALSE, results='hide'}
+# sailfish expression directory
+sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
+
 # read in each file using loop

### 2014-07-25 doc/ApTranscriptome_manuscript.Rmd
Updates. Change TR to download expression data rather than re-run sailfish as unclear if filtered fastq files will be archived.

@@
 carolinensis*.
-This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically-enduced resistance.
+This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically induced resistance.
 

### 2014-07-30 doc/ApTranscriptome_manuscript.Rmd
Incorporated comments from AE/RD/SCH/AN to abstract

@@
   of parapatric ant species"
-author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
-  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
+author: "John Stanton-Geddes, Andrew D. Nguyen, Mahesh Vangala, Jim Vincent, Robert R. Dunn,
+  Aaron Ellison, Nate Sanders, Nicholas J. Gotelli, Sara Helms-Cahan"
 csl: ecology.csl
@@
 
-Thermal tolerance is a key trait governing the distributions of species, their abundances, and underlies their future responses to climate change.
-However, the genetic basis of thermal tolerance is poorly understood for most species.
-In this study, we present a novel regression approach to characterize the thermal reaction norm for all genes in an organism's transcriptome.
-We apply this approach to parapatric ant species in the eastern United States; *A.
+Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change.
+However, the full set of genes and pathways underlying thermal tolerance are poorly understood for most species.
+Here, we demonstrate how insights about the genetic basis of local adaption to climate can be gained by applying a reaction norm approach to transcriptomic data.
+We introduce the thermal *reactionome*, the characterization of the thermal reaction norm of all genes in an organism's transcriptome.
+We characterized transcriptome-wide gene expression over a 38-degree thermal range for two closely-related ant species, *A.
 carolinensis* and *A.
-picea*, which have warmer and colder distributions, respectively.
-Using this approach, we find that at least 10% of genes are thermally responsive.
-In both species, more genes are up-regulated in response to cold than warm temperatures, suggesting that there is less adaptive potential to heat than cold stress.
-Between species, *A.
-carolinensis* shuts down more genes in response to thermal extremes, whereas *A.
-picea* activates more genes in response to cold and hot temperatures.
-The genes that are shut down in *A.
-carolinensis* but up-regulated in *A.
-picea* included many metabolism, cellular organization and repair gene ontology terms.
-We conclude that these species have different mechanisms for responding to thermal stress.
-*A.
-picea* resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas *A.
+picea*, which are parapatrically distributed along a temperature gradient in eastern deciduous forests of North America.
+At least 10% of genes are thermally responsive.
+When the two species were compared, the warm-adapted *A.
+carolinensis* displayed higher basal levels of gene expression, \todo{a lower magnitude of upregulation at thermal extremes}, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes.
+This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability.
+Moreover, a surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress.
+*Aphaenogaster carolinensis* shut down more genes than *A.
+picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A.
+picea*.
+These included genes functionally involved in \todo{metabolism, cellular organization and repair}.
+Altogether, these results suggest that *A.
+picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A.
 carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress.
-In summary, this study demonstrates how we can gain insight on the genetic basis of local adaption to climate using a reaction norm approach with transcriptomic data.
+More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention, despite enzyme efficiency tradeoffs that such mechanisms entail.
 

### 2014-07-30 doc/ApTranscriptome_manuscript.Rmd
Incorporating comments from RD/AE/SCH/AN into Introduction

@@
 
-Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009].
-At the population level, thermal tolerance has been assessed in dozens of studies often demonstrating adaptive differentiation [@hoffmann2013].
-In combination with environmental data, this thermal tolerance data has provided the backbone to dozens of studies examining limits to species geographic ranges [@angert2011; @diamond2013; @gouveia2014], and the impacts of climate change on population viability [@deutsch2008; @kingsolver2013].
-At the organismal level, studies of thermal tolerance have largely focused on candidate genes such as heat shock proteins [@dahlgaard1998].
+Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@cowles1939; @kingsolver2009].
+The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@deutsch2008; @kingsolver2013; @hoffmann2013] and these differences are often genetic [@krebs1997a; @kellermann2012].
+To understand how this diversity of thermal tolerance evolved, we need more information on the genetic basis of thermal tolerance.
+Studies of the genetic basis of thermal tolerance have largely focused on candidate genes such as heat shock proteins [@dahlgaard1998].
 However, estimates of the standing genetic variation [@krebs1997; @krebs1998; @williams2012], studies of differential gene expression [@leemans2000; @sorensen2005; @teets2012], quantitative trait loci (QTL) [@morgan2006] and knock-outs [@takahashi2011] in the model system *Drosophila melanogaster* all implicate dozens of additional genes underlying thermal tolerance.
-Little is known about the diversity of these genes or their expression patterns, especially in non-model species.
+Little is known about the diversity of these genes or their expression patterns, especially in non-model terrestrial species.
 
 The study of whole organism gene expression (e.g.
-transcriptomics) has rapidly increased due to improved RNA sequencing (RNAseq) technology.
-To date, RNAseq studies have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
+transcriptomics) has increased in tandem with improved RNA sequencing (RNAseq) technology.
+In evolutionary ecology, RNAseq studies to date have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
 However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature.
@@
 phenotypic plasticity) is referred to as a reaction norm.
-A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014].
-An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014].
+Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014] and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].
 Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments.
-Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression.
+Similarly, gene expression studies that look at just a few environments (e.g.
+hot vs.
+cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.
 
-For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014].
-However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [] and may not be directly informative about organismal status [@feder2005].
-Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes are over-estimated while many causal genes of small effect are missed [@beavis1994; @larsson2013].
+Moreover, the results provided by RNAseq data do not always directly address the aims of studies.
+For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014].
+However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [@padovan2013] and may not be directly informative about organismal status [@feder2005].
+Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of the few statistically-significant genes are over-estimated while many causal genes of small average effect are missed [@beavis1994; @larsson2013].
 Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues.
 Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].
-Analogously, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient (e.g.
-the thermal reactionome).
-Instead of focusing on individual genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.
+
+Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*.
+Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.
 We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
-picea* [@umphrey1996], which have more southern and northern distributions, respectively.
+picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately ....
+years ago [bernice??].
+
+Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.
+Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.
+
+## Why ants?
 
@@
 picea*.
-
-Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance in these species.
-Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.
 

### 2014-07-31 doc/ApTranscriptome_manuscript.Rmd
Aborted attempt at re-ogranization of Intro

@@
   of parapatric ant species"
-author: "John Stanton-Geddes, Andrew D. Nguyen, Mahesh Vangala, Jim Vincent, Robert R. Dunn,
-  Aaron Ellison, Nate Sanders, Nicholas J. Gotelli, Sara Helms-Cahan"
+author: \centering{John Stanton-Geddes\textsuperscript{1}, Andrew D.
+Nguyen\textsuperscript{1}, Mahesh Vangala\textsuperscript{2}, James Vincent\textsuperscript{2}, Robert R.
+Dunn\textsuperscript{3}, Nathan Sanders\textsuperscript{4}, Nicholas J.
+Gotelli\textsuperscript{1}, Sara Helms Cahan\textsuperscript{1} \\ \textsuperscript{1}: Department of Biology, University of Vermont \\ \textsuperscript{2}: Vermont Genetics Network, University of Vermont \\ \textsuperscript{3}: North Carolina State University \\ \textsuperscript{4}: University of Copenhagen}
 csl: ecology.csl
 output:
-  word_document: default
   pdf_document:
@@
   html_document: null
+  word_document: default
 includes: header.tex
@@
 
-Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@cowles1939; @kingsolver2009].
-The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@deutsch2008; @kingsolver2013; @hoffmann2013] and these differences are often genetic [@krebs1997a; @kellermann2012].
-To understand how this diversity of thermal tolerance evolved, we need more information on the genetic basis of thermal tolerance.
-Studies of the genetic basis of thermal tolerance have largely focused on candidate genes such as heat shock proteins [@dahlgaard1998].
-However, estimates of the standing genetic variation [@krebs1997; @krebs1998; @williams2012], studies of differential gene expression [@leemans2000; @sorensen2005; @teets2012], quantitative trait loci (QTL) [@morgan2006] and knock-outs [@takahashi2011] in the model system *Drosophila melanogaster* all implicate dozens of additional genes underlying thermal tolerance.
-Little is known about the diversity of these genes or their expression patterns, especially in non-model terrestrial species.
+Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009].
+The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
+These differences in thermal tolerance are often genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
+
+. To understand how this variation of thermal tolerance evolved, we need information on the diversity of genes, and their expression patterns, involved in thermal tolerance.
+
+In this study, we present a novel method for gene expression analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient, and report the set of genes that are thermally-responsive; the *thermal reactionome*.
+We then use a comparative approach to compare the thermal reactionomes between two closely-related ant species to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.
 
@@
 years ago [bernice??].
-
-Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.
-Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.
-
-## Why ants?
-
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
 Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food.
@@
 
+Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.
+Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.
+
 # Methods
+
+## Why ants?
+
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
+Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].
+...previous results from chambers...???
 
@@
 
-Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W).
+We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina ( 36.037° N -78.874° W).
 The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers.

### 2014-08-01 doc/ApTranscriptome_manuscript.Rmd
revised Intro

--- 2014-07-31 doc/ApTranscriptome_manuscript.Rmd
@@
 The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
-These differences in thermal tolerance are often genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-
-. To understand how this variation of thermal tolerance evolved, we need information on the diversity of genes, and their expression patterns, involved in thermal tolerance.
-
-In this study, we present a novel method for gene expression analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient, and report the set of genes that are thermally-responsive; the *thermal reactionome*.
-We then use a comparative approach to compare the thermal reactionomes between two closely-related ant species to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.
+These differences in thermal tolerance are largely genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
+How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffman2008], is not well understood.
+To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their expression patterns, involved in thermal tolerance.
 
@@
 years ago [bernice??].
+
 Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food.
@@
 picea* [@diamond2013].
-Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species.
-For example, thermally-responsive genes may shift the temperature at which they are up-regulated to higher temperatures in *A.
-carolinensis*, or colder temperatures in *A.
-picea*.
-Additionally, genes may change their sensitivity to thermal extremes.
-Further, through the process of genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A.
+Thus, patterns of gene expression underlying tolerance of thermal extremes is likely to differ between these species in a number of ways.
+
+First, the number of genes showing plasticity of expression to thermal extremes may differ between the species.
+Finding more genes are up-regulated to high temperatures in the colder-climate species *A.
+picea* would indicate that this species relies on activation of the cellular stress response [@...] and other protective genes to a greater extent than *A.
+carolinensis* , while the opposite would be true for cold temperatures.
+Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies upon proteome stability [@...] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.
+
+Second, responsive genes may change their sensitivity to thermal extremes.
+Finding that heat-responsive genes are up-regulated at higher temperatures in *A.
+carolinensis* than *A.
+picea* would be consistent with local adaptation for greater heat tolerance.
+That is, more extreme temperatures are required for activation of the heat shock response [@somero...].
+Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A.
+picea* than *A.
+carolinensis* would indicate local adaptation for cold tolerance.
+Similar patterns may be observed for genes that are down-regulated in response to thermal extremes.
+
+Moreover, insight on the evolution of thermal tolerance can be gained by examining shifts in the patterns of responsive genes between species.
+For example, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A.
 carolinensis* may have increased basal expression in the colder-climate species *A.
@@
 # Methods
-
-## Why ants?
 

### 2014-08-01 doc/ApTranscriptome_manuscript.Rmd
Revised Intro incorporating comments from AE/NG

@@
 ---
-title: "Patterns of stress tolerance versus resistance in genome-wide expression data
-  of parapatric ant species"
-author: \centering{John Stanton-Geddes\textsuperscript{1}, Andrew D.
-Nguyen\textsuperscript{1}, Mahesh Vangala\textsuperscript{2}, James Vincent\textsuperscript{2}, Robert R.
-Dunn\textsuperscript{3}, Nathan Sanders\textsuperscript{4}, Nicholas J.
-Gotelli\textsuperscript{1}, Sara Helms Cahan\textsuperscript{1} \\ \textsuperscript{1}: Department of Biology, University of Vermont \\ \textsuperscript{2}: Vermont Genetics Network, University of Vermont \\ \textsuperscript{3}: North Carolina State University \\ \textsuperscript{4}: University of Copenhagen}
+title: "Patterns of stress tolerance versus resistance in the thermal reactionomes from populations of cold and warm-climate ant species"
+author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
+Dunn \textsuperscript{3}, Nathan J.
+Sanders \textsuperscript{4}, Nicholas J.
+Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1} Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences, North Carolina State University \\ \textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen}
 csl: ecology.csl
@@
 
+**250 words allowed - need to cut 56**
+
 Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change.
-However, the full set of genes and pathways underlying thermal tolerance are poorly understood for most species.
-Here, we demonstrate how insights about the genetic basis of local adaption to climate can be gained by applying a reaction norm approach to transcriptomic data.
-We introduce the thermal *reactionome*, the characterization of the thermal reaction norm of all genes in an organism's transcriptome.
-We characterized transcriptome-wide gene expression over a 38-degree thermal range for two closely-related ant species, *A.
-carolinensis* and *A.
-picea*, which are parapatrically distributed along a temperature gradient in eastern deciduous forests of North America.
-At least 10% of genes are thermally responsive.
+In this study, we demonstrate how insight on the evolution of thermal tolerance can be gained through characterizing an organism’s thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient.
+We characterized the reactionomes of two populations of parapatric ant species of eastern U.S.
+deciduous forests: the northern cold-climate Aphaenogaster rudis and the southern warm-climate Aphaenogaster carolinensis .
+After exposing laboratory-acclimated individuals to a range of realistic temperatures (0° - 38.5°C), we found that 10% of genes were thermally responsive.
 When the two species were compared, the warm-adapted *A.
-carolinensis* displayed higher basal levels of gene expression, \todo{a lower magnitude of upregulation at thermal extremes}, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes.
+carolinensis* displayed higher basal levels of gene expression, \todo{check this statement} a lower magnitude of upregulation at thermal extremes, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes.
 This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability.
@@
 picea*.
-These included genes functionally involved in \todo{metabolism, cellular organization and repair}.
+These included genes functionally involved in \todo{check} metabolism, cellular organization and repair.
 Altogether, these results suggest that *A.
@@
 
+# Significance Statement
+
+**120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty**
+
 # Introduction
@@
 The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
-These differences in thermal tolerance are largely genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffman2008], is not well understood.
-To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their expression patterns, involved in thermal tolerance.
+While organisms differ in fixed traits such as proteome stability [@ghosh2010] and behavior [@sunday2014] that allow them to function in the typical range of temperatures to which they are exposed, tolerance to thermal extremes often involves rapid changes in gene expression [@kultz2005; @somero2011].
+These differences within and among species in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
+How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood.
+To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.
 
-The study of whole organism gene expression (e.g.
-transcriptomics) has increased in tandem with improved RNA sequencing (RNAseq) technology.
-In evolutionary ecology, RNAseq studies to date have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach.
+In recent years, whole organism gene expression approaches (e.g.
+transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], drought tolerance [@villar2011] and many other traits.
+These RNAseq studies have used an ANOVA-type experimental or sampleing design with only a few levels of temperature or other environmental stressors tested.
 However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature.
-Variation in the phenotype of an organism in response to such a continuous predictor (i.e.
-phenotypic plasticity) is referred to as a reaction norm.
-Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014] and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].
-Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments.
+The reaction norm describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992].
+Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].
+Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable.
 Similarly, gene expression studies that look at just a few environments (e.g.
@@
 Moreover, the results provided by RNAseq data do not always directly address the aims of studies.
-For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014].
-However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [@padovan2013] and may not be directly informative about organismal status [@feder2005].
-Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of the few statistically-significant genes are over-estimated while many causal genes of small average effect are missed [@beavis1994; @larsson2013].
-Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues.
+For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014].
+However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005].
+Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013].
+Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues.
 Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].
 
-Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*.
-Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.
+Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across a thermal gradient.
+Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes of two closely-related temperate-zone species to gain insight into how spatially divergent selection has changed patterns of thermally-responsive gene expression.
 We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
 picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately ....
-years ago [bernice??].
+years ago (bernice??).
+Temperature is a potentially strong selective force on ground-nesting *Aphaenogaster* colonies, which must tolerate seasonally freezing winters and hot summers.
+On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@lubertazzi2012].
 
-Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food.
 *A.
 carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A.
-picea* [@diamond2013].
+picea* (Supplemental Information).
 Thus, patterns of gene expression underlying tolerance of thermal extremes is likely to differ between these species in a number of ways.
-
 First, the number of genes showing plasticity of expression to thermal extremes may differ between the species.
 Finding more genes are up-regulated to high temperatures in the colder-climate species *A.
-picea* would indicate that this species relies on activation of the cellular stress response [@...] and other protective genes to a greater extent than *A.
-carolinensis* , while the opposite would be true for cold temperatures.
-Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies upon proteome stability [@...] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.
+picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A.
+carolinensis*, while the opposite would be true for cold temperatures.
+Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.
 
-Second, responsive genes may change their sensitivity to thermal extremes.
+In addition, responsive genes may change in their sensitivity to thermal extremes.
 Finding that heat-responsive genes are up-regulated at higher temperatures in *A.
@@
 picea* would be consistent with local adaptation for greater heat tolerance.
-That is, more extreme temperatures are required for activation of the heat shock response [@somero...].
+That is, more extreme temperatures are required for activation of the heat shock response [@somero2011].
 Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A.
@@
 Similar patterns may be observed for genes that are down-regulated in response to thermal extremes.
+Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance.
+Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A.
+carolinensis* may have increased basal expression in the colder-climate species *A.
+picea*, and vice versa.
 
-Moreover, insight on the evolution of thermal tolerance can be gained by examining shifts in the patterns of responsive genes between species.
-For example, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A.
-carolinensis* may have increased basal expression in the colder-climate species *A.
-picea*.
-
-Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.
-Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.
-
-# Methods
-
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
-Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].
-...previous results from chambers...???
-
-## Samples
-
-We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina ( 36.037° N -78.874° W).
-The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers.
-communication).
-Colonies of these species were maintained in the lab for 6 months prior to experimentation.
-
-From each species, we haphazardly collected 12 ants at the same time on 12 days.
-Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour.
-The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
-At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
-We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
-
-## Sequencing, assembly and annotation
-
-For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the *A.
-picea* and *A.
-carolinensis* samples respectively.
-Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
-
-We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
-*De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g.
-regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013].
-We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
-Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters).
-We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
-To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.
-
-To determine the putative function of the transcripts, we performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
-
-We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
-
-## Thermally-responsive genes
-
-We quantified expression of each gene using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression.
-We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses.
-As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis.
-The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).
-
-To identify genes that have significant changes in expression across the thermal gradient, we fit the regression model
-
-$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$
-
-to each gene.
-We used $log(TPM + 1)$ as the response to control for skew in the expression data.
-For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005].
-This method is robust to over-dispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
-To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.
-
-We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
-All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
-
-For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
-
-## Species-level comparisons
-
-We performed comparative analyses between the two species focusing on the genes that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of gene activation and sensitivity of the thermally-responsive genes.
-For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
-To test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression for each gene in each species.
-We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes.
-We then performed a *t*-test to determine if the mean temperature of gene activation differed between the two species for each group.
-
-To determine if the sensitivity of genes to thermal extremes differed between the species, we used the subset of **Intermediate** and **Bimodal** genes in each species.
-For each **Intermediate** gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
-point of maximum expression had highest probability of being sampled).
-We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species.
-Finding a greater standard deviation of expression would be indicative of less sensitivity to changes in temperature.
-For the **Bimodal** genes, we performed the same analysis as with the **Intermediate** genes, but with a greater standard deviation being indicative of a more sensitive gene (convex reaction norm) while a low standard deviation being a gene with greater inertia of expression in response to temperature changes (nearly flat).Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity for that species.
-
-To test for genetic assimilation, we compared basal expression levels (expression at 19.5° C) for the subset of genes that were responsive to cold temperatures (**Low** genes) in the southern *A.
-carolinensis* species, or hot temperatures (**High** genes) in the northern *A.
-picea* species using non-parametric two-sided Wilcoxon tests.
-
-Finally, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene.
-Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix).
-As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A.
-picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A.
-picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively.
-Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A.
-picea* than *A.
-carolinensis*.
-Further, to quantify the relative contribution of changes in the mean, slope, curvature or wiggle on the overall differences in reaction norms, we defined the *Total* difference in the reaction norm for each gene as $\Delta T = \Delta M + \Delta S + \Delta C + \Delta W$.
-We then partitioned the contribution of each measure by dividing it by $\Delta T$.
-
-## Gene set enrichment analysis
-
-To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the package `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
-Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
-We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.
-
-## Data availability
-
-The raw Illumina reads are available at [...], the script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).
+Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012].
+However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes.
+For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].
 
@@
 
+# Material and Methods
+
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
+Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].
+...previous results from chambers...???
+
+## Samples
+
+We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina ( 36.037° N -78.874° W).
+The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers.
+communication).
+Colonies of these species were maintained in the lab for 6 months prior to experimentation.
+
+From each species, we haphazardly collected 12 ants at the same time on 12 days.
+Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour.
+The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
+At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
+We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
+
+## Sequencing, assembly and annotation
+
+For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the *A.
+picea* and *A.
+carolinensis* samples respectively.
+Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
+
+We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
+*De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g.
+regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013].
+We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
+Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters).
+We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
+To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.
+
+To determine the putative function of the transcripts, we performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
+
+We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
+
+## Thermally-responsive genes
+
+We quantified expression of each gene using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression.
+We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses.
+As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis.
+The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).
+
+To identify genes that have significant changes in expression across the thermal gradient, we fit the regression model
+
+$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$
+
+to each gene.
+We used $log(TPM + 1)$ as the response to control for skew in the expression data.
+For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005].
+This method is robust to over-dispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
+To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.
+
+We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
+
+For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
+
+## Species-level comparisons
+
+We performed comparative analyses between the two species focusing on the genes that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of gene activation and sensitivity of the thermally-responsive genes.
+For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
+To test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression for each gene in each species.
+We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes.
+We then performed a *t*-test to determine if the mean temperature of gene activation differed between the two species for each group.
+
+To determine if the sensitivity of genes to thermal extremes differed between the species, we used the subset of **Intermediate** and **Bimodal** genes in each species.
+For each **Intermediate** gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
+point of maximum expression had highest probability of being sampled).
+We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species.
+Finding a greater standard deviation of expression would be indicative of less sensitivity to changes in temperature.
+For the **Bimodal** genes, we performed the same analysis as with the **Intermediate** genes, but with a greater standard deviation being indicative of a more sensitive gene (convex reaction norm) while a low standard deviation being a gene with greater inertia of expression in response to temperature changes (nearly flat).Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity for that species.
+
+To test for genetic assimilation, we compared basal expression levels (expression at 19.5° C) for the subset of genes that were responsive to cold temperatures (**Low** genes) in the southern *A.
+carolinensis* species, or hot temperatures (**High** genes) in the northern *A.
+picea* species using non-parametric two-sided Wilcoxon tests.
+
+Finally, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene.
+Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix).
+As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A.
+picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A.
+picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively.
+Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A.
+picea* than *A.
+carolinensis*.
+Further, to quantify the relative contribution of changes in the mean, slope, curvature or wiggle on the overall differences in reaction norms, we defined the *Total* difference in the reaction norm for each gene as $\Delta T = \Delta M + \Delta S + \Delta C + \Delta W$.
+We then partitioned the contribution of each measure by dividing it by $\Delta T$.
+
+## Gene set enrichment analysis
+
+To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the package `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
+Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
+We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.
+
+## Data availability
+
+The raw Illumina reads are available at [...], the script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).
+
+# Figure Legends
+
 # References

### 2014-08-01 doc/ApTranscriptome_manuscript.Rmd
manuscript v-1.2 sent to co-authors

@@
 ---
-title: "Patterns of stress tolerance versus resistance in the thermal reactionomes from populations of cold and warm-climate ant species"
-author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
-Dunn \textsuperscript{3}, Nathan J.
-Sanders \textsuperscript{4}, Nicholas J.
-Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1} Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences, North Carolina State University \\ \textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen}
+title: "Patterns of stress tolerance versus resistance in the thermal reactionomes of cold and warm-climate ant species"
+author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
+  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
+  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
+  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
+  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
+  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
+  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
+  Evolution and Climate, University of Copenhagen}
 csl: ecology.csl
 output:
+  word_document: default
   pdf_document:
@@
   html_document: null
-  word_document: default
 includes: header.tex
@@
 
-**250 words allowed - need to cut 56**
+**250 words allowed - need to cut 26**
 
@@
 In this study, we demonstrate how insight on the evolution of thermal tolerance can be gained through characterizing an organism’s thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient.
-We characterized the reactionomes of two populations of parapatric ant species of eastern U.S.
-deciduous forests: the northern cold-climate Aphaenogaster rudis and the southern warm-climate Aphaenogaster carolinensis .
-After exposing laboratory-acclimated individuals to a range of realistic temperatures (0° - 38.5°C), we found that 10% of genes were thermally responsive.
+We characterized the reactionomes of two populations of ant species of eastern U.S.
+forests: the northern cold-climate Aphaenogaster picea and the southern warm-climate Aphaenogaster carolinensis .
+We found that  at least 10% of genes were thermally responsive.
 When the two species were compared, the warm-adapted *A.
-carolinensis* displayed higher basal levels of gene expression, \todo{check this statement} a lower magnitude of upregulation at thermal extremes, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes.
+carolinensis* displayed higher basal levels of gene expression,, upregulated of fewer heat shock proteins, and had greater inertia of changes in gene expression in response to thermal extremes.
 This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability.
-Moreover, a surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress.
+A surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress.
 *Aphaenogaster carolinensis* shut down more genes than *A.
@@
 picea*.
-These included genes functionally involved in \todo{check} metabolism, cellular organization and repair.
+These included genes functionally involved in metabolism, cellular organization and repair.
 Altogether, these results suggest that *A.
@@
 carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress.
-More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention, despite enzyme efficiency tradeoffs that such mechanisms entail.
+More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
 
@@
 
-Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across a thermal gradient.
+Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across an ecologically-relevant thermal gradient.
 Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes of two closely-related temperate-zone species to gain insight into how spatially divergent selection has changed patterns of thermally-responsive gene expression.
@@
 On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@lubertazzi2012].
+We collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina (cite) paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Information).
 
-*A.
+As *A.
 carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A.
-picea* (Supplemental Information).
-Thus, patterns of gene expression underlying tolerance of thermal extremes is likely to differ between these species in a number of ways.
+picea* (Supplemental Information), we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways.
 First, the number of genes showing plasticity of expression to thermal extremes may differ between the species.
@@
 Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012].
-However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes.
+However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes using gene set enrichment analysis.
 For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].
@@
 
-## Thermally-responsive genes
+## Species comparisons in the number of thermally-responsive genes
 
-The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp.
-Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes.
-The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp.
-Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database.
-Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage.
-We removed 5,675 genes identified as known contaminants, leaving 99,861 clean transcripts.
-Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
+The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts, hereafter referred to as genes, of which 47% had a best hit to an organism in the class Insecta.
+Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that were thermally-responsive at 5% False Discovery Rate.
+Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive.
+However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012].
+Assuming that the number of genes is similar for *Aphaenogaster*, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.
 
@@
 picea*.
-The number of genes are labeled in two boxes for scale."}
+The total number of genes for two categories (*A.
+picea* Low - *A.
+carolinensis* Low; *A.
+picea* NotResp - *A.
+carolinensis* Bimodal) is labeled in two boxes for perspective."}
 mosaic.colors <- brewer.pal(5, "Blues")
@@
 
-We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR.
-Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive.
-However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012], and assuming the number is similar for *Aphaenogaster*, this suggests that up to 50% of the genome could be thermally-responsive.
-The reaction norms for each responsive gene can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].
-
-Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species.
-These differences were not only in magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
+For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
+Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` had an expression pattern that differs between the two species.
+While some of these were simply responsive in only one species, `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A.
+carolinensis* but not *A.
+picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A.
+picea* but not *A.
+carolinensis*, most of these were genes that changed from one expression category to another between the species (e.g.
+the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
 1).
@@
 1).
-There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A.
-carolinensis* but not *A.
-picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A.
-picea* but not *A.
-carolinensis*.
 
-## Species comparisons
+## Sensitivity of thermally-responsive genes to changes in temperature
 
-To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes, and the variance in expression for the **Intermediate** and **Bimodal** genes.
+To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes.
 The **High** genes in *A.
@@
 2).
-For **Bimodal** expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`).
-In contrast, **Intermediate** genes from *A.
-carolinensis* had a broader thermal range of stable gene expression than those from *A.
-picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`).
+
+For **Intermediate** and **Bimodal** genes, genes had two inflection points so it was not possible to calculate a single temperature of gene activation.
+Instead, we calculated the variance of 1,000 random temperatures, weighted by expression at that temperature.
+For **Intermediate** genes, a greater variance would be indicative of less sensitivity to changes in temperature.
+For **Bimodal** genes, a greater variance would be indicative of a more sensitivity to changes in temperature (convex reaction norm).
+We found that **Intermediate** genes from *A.
+carolinensis* had a greater variance of expression than those from *A.
+picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`), indicating greater inertia of gene expression to changes in temperature.
+However, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`) for the **Bimodal** genes.
 
@@
 
-We looked for evidence of genetic assimilation by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
+We looked for evidence of genetic assimilation (the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed )by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
 carolinensis* or high temperatures in *A.
@@
 
-In addition to shifts in position, we quantified the differences between species in the mean and shape of the reaction norms of thermally-responsive genes.
-Species differed by mean expression (*t*~df=`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A.
-carolinensis* than *A picea*, and mean slope (*t*~df=`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A.
-picea*.
-Curvature (*t*~df=`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~df=`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species.
-Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.
-
 ## Gene set enrichment analysis
 
-As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately.
+As the molecular pathways involved in thermal response to cold and hot temperatures differ, we used gene set enrichment analysis on each expression category separately.
 First, we looked at the genes shared in each expression category (e.g.
@@
 
-We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A.
+We further used GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A.
 carolinensis* to having increased expression in **Low** temperatures in *A.
@@
 In this study, we applied a novel regression design to characterize the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively, that are known to differ in their thermal tolerances [@warren2013].
-We found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.
+To date, transcriptomic studies have used differential expression analysis between a few different conditions or environments (but see [@hodgins-davis2012]).
+By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
+Specifically, we found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.
 
-The patterns of gene expression between species differed substantially, with more down-regulation in *A.
-carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A.
-picea* in response to thermal extremes (**High** and **Low** genes; Fig.
+The patterns of gene expression between species differed substantially, with more genes down-regulated in *A.
+carolinensis* and more up-regulated in *A.
+picea* in response to thermal extremes (Fig.
 1, 3).
 The results of the gene set enrichment analysis give an indication of what might drive this pattern.
-The genes that are up-regulated at cold temperatures (**Low**) in *A.
-picea* but shut-down **Intermediate** by *A.
-carolinensis* include metabolism, transport and regulation GO clusters.
+The genes that are shut-down by *A.
+carolinensis* in response to cold temperatures, but up-regulated by *A.
+picea* include gene ontology clusters involved in metabolism, transport and regulation.
 Thus, *A.
-carolinensis* shuts-down metabolism to a lesser extent than *A.
-picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress.
+carolinensis*, the more southerly of our two species, shuts-down metabolism to a lesser extent than *A.
+picea* and so must rely on other mechanisms such as behavior (e.g.
+retreat into its nest) or proteome stability to tolerate thermal stress.
 In contrast, *A.
 picea* increases metabolism to a greater extent than *A.
-carolinensis*, potentially to mount a greater molecular response to thermal stress.
+carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress.
 Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A.
@@
 carolinensis*.
-This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically induced resistance.
+This result supports our expectation that *A.
+picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A.
+carolinensis*.
+In general this contributes to evidence from other systems [@barshis2013]  that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
+...Sara - something about enzyme efficiency tradeoffs???
 
@@
 Given the differences in the number of thermally-responsive genes between the two species (Fig.
-1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species (Fig.
+1), it is worth noting the similarity in expression patterns between species.
+There are 3-4\times more genes upregulated at cold than high temperatures in both species (Fig.
 3).
 A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013].
-This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures.
-Moreover, in an experiment with 94 species of *Drosophila*, Kellermann et al [-@kellermann2012] showed that limited genetic variation among species (e.g.
-phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits.
-Thus, when coupled with phenotypic data, our results demonstrate that transcriptomic studies can provide insight on evolutionary potential.
-For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
+This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
+At least among *Drosophila* taxa, these limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012].
+In this light, for *A.
+picea* and *A.
+carolinensis*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
 
-For species or populations that routinely, but unpredictably, experience stressful environments, genetic assimilation is the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed [@pigliucci2006].
-However, in contrast to these expectations, we found lower basal gene expression in *A.
+In contrast to expectations that adaptation occurs via changes from plastically-induced to constitutive expression (e.g.
+genetic assimilation), we found lower basal gene expression in *A.
 picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A.
@@
 carolinensis* for **High** genes in *A.
-picea* as well as for genes that were **Intermediate** in both spcies.
+picea* as well as for genes that were **Intermediate** in both species.
 A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A.
@@
 
-Of the `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes that were thermally-responsive, `r round(with(sigtable, sigtable[Coefficient=="Temperature:Colony","Number.significant"]) / with(sigtable, sum(sigtable[Coefficient %in% c("Temperature", "Temperature:Colony"),"Number.significant"])) * 100, 0)`% differed in their responsiveness by colony.
-The differences between species were due to changes in the mean ($\Delta M$) and slope ($\Delta S$) of the reaction norms for each thermally-responsive gene.
-This result contrasts with the findings of Murren et al.
-[-@murren2014], who found that most of the differences in phenotypic reaction norms were due to shape.
-As genes are not directly analogous to phenotypes, this contradiction is not surprising.
-Instead, it indicates that small shifts in the mean of many genes underlying a trait can result in non-linear phenotypic changes.
-
-To date, transcriptomic studies have performed differential expression analysis between a few different conditions or environments (but see [@hodgins-davis2012]).
-By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
-Thus, we were able to compare the temperature at which genes were activated near the upper and lower thermal limits between species (Fig.
-2).
-For genes activated at cold temperatures, there was little difference among species with *A.
-picea* genes turning on at 13°C compared to 12.8°C for *A.
-carolinensis*.
-For high temperature genes, *A.
-carolinensis* genes were turned on at 33.8°C compared to 32.9°C for *A.
-picea*.
-
 As with all *de novo* transcriptomes, our work has some important caveats.
 First, given that all sequenced ant genomes to date have about 18,000 genes [@gadau2012], our transcriptome is highly fragmented.
-Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts.
-A genome assembly, in progress, would considerably resolve this problem.
-Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is poor, as we have to rely upon resources from distantly related *Drosophila* and *Apis*.
+Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts.
+A genome assembly, in progress, will help to reduce fragmentation.
+Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for distantly-related model organisms *Drosophila* and *Apis*.
 Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005].
@@
 
-In this work, we bring reaction norms to the genomic era by characterizing the thermal reactionomes of the temperate ant species, *Aphaenogaster picea* and *A.
+In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A.
 carolinensis*.
-We found that at least 10% of the transcriptome is thermally-responsive.
+At least 10% of their transcriptomes are thermally-responsive.
 For both species, more genes responded to cold than hot temperatures.
-In light of physiological data showing more variation in critical minimum than maximum temperatures, this suggests that these species and other ectotherms have less adaptive potential to hot than cold temperatures.
-Further, our results indicate that the two species have different responses to thermal extremes.
+Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential to hot than cold temperatures.
+Our results also indicate that these two ant species have different responses to thermal extremes.
 *A.
-picea* responds by increasing metabolism and activating heat shock and other protective molecules, while *A.
-carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolernace.
-These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.
+picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A.
+carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance.
+The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a tremendous resource for the future study of ecological adaptation in ant species.
+Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.
 
@@
 
-To determine the putative function of the transcripts, we performed functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
+The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 genes with a total length of 100 million bp.
+Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes.
+The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp.
+Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database.
+Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage.
+We removed 5,675 genes identified as known contaminants, leaving 99,861 clean transcripts.
+Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
 
-We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
+To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
+
+We used subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
 
@@
 
-Finally, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene.
-Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix).
-As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A.
-picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A.
-picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively.
-Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A.
-picea* than *A.
-carolinensis*.
-Further, to quantify the relative contribution of changes in the mean, slope, curvature or wiggle on the overall differences in reaction norms, we defined the *Total* difference in the reaction norm for each gene as $\Delta T = \Delta M + \Delta S + \Delta C + \Delta W$.
-We then partitioned the contribution of each measure by dividing it by $\Delta T$.
-
 ## Gene set enrichment analysis

### 2014-08-08 doc/ApTranscriptome_manuscript.Rmd
Updated with comments from NS

@@
 ---
-title: "Patterns of stress tolerance versus resistance in the thermal reactionomes of cold and warm-climate ant species"
+title: Investigations of thermal reactionomes reveals stress tolerance vs. resistance
+  warm and cold-climate ant species
 author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
@@
 
-**250 words allowed - need to cut 26**
-
 Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change.
-In this study, we demonstrate how insight on the evolution of thermal tolerance can be gained through characterizing an organism’s thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient.
-We characterized the reactionomes of two populations of ant species of eastern U.S.
-forests: the northern cold-climate Aphaenogaster picea and the southern warm-climate Aphaenogaster carolinensis .
+Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the evolution of thermal tolerance.
+We characterized the reactionomes of two common ant species in eastern U.S.
+forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
 We found that at least 10% of genes were thermally responsive.
-When the two species were compared, the warm-adapted *A.
-carolinensis* displayed higher basal levels of gene expression,, upregulated of fewer heat shock proteins, and had greater inertia of changes in gene expression in response to thermal extremes.
+Interestingly, the warm-adapted *A.
+carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change to thermal extremes than did *A.
+picea*.
 This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability.
-A surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress.
-*Aphaenogaster carolinensis* shut down more genes than *A.
+*A.
+carolinensis* shut down more genes than *A.
 picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A.
 picea*.
-These included genes functionally involved in metabolism, cellular organization and repair.
-Altogether, these results suggest that *A.
+These included genes functionally involved in metabolism, cellular organization and repair \todo{Sara - where did the 'repair' term come from?}.
+These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress, namely *A.
 picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A.
@@
 
-Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009].
-The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
-While organisms differ in fixed traits such as proteome stability [@ghosh2010] and behavior [@sunday2014] that allow them to function in the typical range of temperatures to which they are exposed, tolerance to thermal extremes often involves rapid changes in gene expression [@kultz2005; @somero2011].
+Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009].
+Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
 These differences within and among species in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood.
+How this variation in thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood.
 To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.
@@
 In recent years, whole organism gene expression approaches (e.g.
-transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], drought tolerance [@villar2011] and many other traits.
-These RNAseq studies have used an ANOVA-type experimental or sampleing design with only a few levels of temperature or other environmental stressors tested.
-However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature.
-The reaction norm describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992].
-Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].
-Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable.
+transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], among other traits.
+These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few levels of temperature.
+However, temperature varies continuously in nature.
+As a results, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014].
 Similarly, gene expression studies that look at just a few environments (e.g.
@@
 cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.
+What is needed is a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992].
+Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].
 
-Moreover, the results provided by RNAseq data do not always directly address the aims of studies.
-For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014].
-However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005].
-Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013].
+Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, an important suite of issues often hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014].
+Namely, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005].
+Genome-wide association studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013].
 Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues.
-Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].
 
-Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across an ecologically-relevant thermal gradient.
-Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes of two closely-related temperate-zone species to gain insight into how spatially divergent selection has changed patterns of thermally-responsive gene expression.
-We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
-picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately ....
-years ago (bernice??).
-Temperature is a potentially strong selective force on ground-nesting *Aphaenogaster* colonies, which must tolerate seasonally freezing winters and hot summers.
-On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@lubertazzi2012].
-We collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina (cite) paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Information).
+Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences.
+Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes to gain insight into how spatially divergent selection has led to patterns thermally-responsive gene expression in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
+picea* [@umphrey1996].
+Temperature is a potentially strong selective force on ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
+On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
 
-As *A.
-carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A.
-picea* (Supplemental Information), we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways.
+As the more southern *A.
+carolinensis* occurs in climates that are hotter with longer growing seasons than that of the more northerly *A.
+picea*, we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways.
 First, the number of genes showing plasticity of expression to thermal extremes may differ between the species.
-Finding more genes are up-regulated to high temperatures in the colder-climate species *A.
+Finding that more genes are up-regulated to high temperatures in the colder-climate species *A.
 picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A.
@@
 
-In addition, responsive genes may change in their sensitivity to thermal extremes.
+Second, responsive genes may change in their sensitivity to thermal extremes.
 Finding that heat-responsive genes are up-regulated at higher temperatures in *A.
-carolinensis* than *A.
+carolinensis* than in *A.
 picea* would be consistent with local adaptation for greater heat tolerance.
@@
 For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].
+To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
 
@@
 
-The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts, hereafter referred to as genes, of which 47% had a best hit to an organism in the class Insecta.
+The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts, hereafter referred to as genes.
 Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that were thermally-responsive at 5% False Discovery Rate.
 Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive.
-However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012].
-Assuming that the number of genes is similar for *Aphaenogaster*, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.
+However, the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012].
+Assuming that the number of genes is similar for these two *Aphaenogaster* species, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.
 
 ```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type.
+**Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.
 The width of the columns is proportional to the number of genes for each expression type for *A.
@@
 picea*.
-The total number of genes for two categories (*A.
-picea* Low - *A.
-carolinensis* Low; *A.
-picea* NotResp - *A.
-carolinensis* Bimodal) is labeled in two boxes for perspective."}
+**Low** The total number of genes for two categories (*A.
+picea* **Low** - *A.
+carolinensis* **Low**; *A.
+picea* **NotResp** - *A.
+carolinensis* **Bimodal**) is labeled in two boxes for perspective."}
 mosaic.colors <- brewer.pal(5, "Blues")
@@
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
-Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` had an expression pattern that differs between the two species.
-While some of these were simply responsive in only one species, `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A.
+Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` had an expression pattern that differed between the two species.
+While some of these were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes were responsive in *A.
 carolinensis* but not *A.
-picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A.
+picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` were responsive in *A.
 picea* but not *A.
-carolinensis*, most of these were genes that changed from one expression category to another between the species (e.g.
+carolinensis*), most of these were genes that changed from one expression category to another between the species (e.g.
 the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
@@
 The **High** genes in *A.
-carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A.
+carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A.
 picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
 2).
-For genes with increased expression at **Low** temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+For genes with increased expression at **Low** temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
+carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
+picea*; *t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
 2).
@@
 ```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at **Low** and **High** temperatures.
-Bars are Probability density function of peak expression for all thermally-responsive in *A.
-carolinensis* (blue) and *A.
-carolinensis* (red) species"}
+Red bars and lines are for *A.
+carolinensis* while blue bars and lines are for *A.
+picea*."}
 
@@
 
-As the molecular pathways involved in thermal response to cold and hot temperatures differ, we used gene set enrichment analysis on each expression category separately.
+To describe the molecular toolkit underlying *Aphaenogaster* response to thermal extremes, we used gene set enrichment analysis to identify gene ontology (GO) categories enriched for each expression category separately.
 First, we looked at the genes shared in each expression category (e.g.
@@
 
-We further used GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A.
+We further used gene set enrichment analysis for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A.
 carolinensis* to having increased expression in **Low** temperatures in *A.
@@
 
-Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species.
-In this study, we applied a novel regression design to characterize the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively, that are known to differ in their thermal tolerances [@warren2013].
-To date, transcriptomic studies have used differential expression analysis between a few different conditions or environments (but see [@hodgins-davis2012]).
 By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
 Specifically, we found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.
+This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have variable expression in response to copper exposure [@hodgins-davis2012].
+In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014].
+Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.
 
@@
 1, 3).
-The results of the gene set enrichment analysis give an indication of what might drive this pattern.
+The results of the gene set enrichment analysis give an indication of what might drive this difference.
 The genes that are shut-down by *A.
@@
 Thus, *A.
-carolinensis*, the more southerly of our two species, shuts-down metabolism to a lesser extent than *A.
+carolinensis*, the more southerly species, shuts-down metabolism to a lesser extent than *A.
 picea* and so must rely on other mechanisms such as behavior (e.g.
@@
 carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress.
-Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A.
+Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A.
 picea*, but only 4 do in *A.
@@
 carolinensis*.
-In general this contributes to evidence from other systems [@barshis2013]  that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
+Our results are at least congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
 ...Sara - something about enzyme efficiency tradeoffs???
@@
 1), it is worth noting the similarity in expression patterns between species.
-There are 3-4\times more genes upregulated at cold than high temperatures in both species (Fig.
+There are 3-4 $\times$ more genes upregulated at cold than high temperatures in both species (Fig.
 3).
@@
 Large differences in patterns of gene expression may only have small effects on organismal fitness.
-Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change.
-Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@jensen2008; @kenkel2013].
-In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@oneil2014; @lemay2013].
+Given these caveats, transcriptomic studies promise to offer many new insights into how organisms respond and will respond to environmental change, especially for long-lived species that are intractable to study in the lab.
 
@@
 For both species, more genes responded to cold than hot temperatures.
-Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential to hot than cold temperatures.
+Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to hot than cold temperatures.
 Our results also indicate that these two ant species have different responses to thermal extremes.
@@
 carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance.
-The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a tremendous resource for the future study of ecological adaptation in ant species.
+The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species.
 Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.
@@
 
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006].
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2011].
 Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].
-...previous results from chambers...???
 

### 2014-08-19 doc/ApTranscriptome_manuscript.Rmd
some changes

@@
 ---
-title: Investigations of thermal reactionomes reveals stress tolerance vs. resistance
-  warm and cold-climate ant species
+title: Investigations of thermal reactionomes reveal stress tolerance vs. resistance in
+  warm and cold-climate ant species \\
 author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
@@
 \linenumbers
-\tableofcontents
 
@@
 picea*.
-These included genes functionally involved in metabolism, cellular organization and repair \todo{Sara - where did the 'repair' term come from?}.
-These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress, namely *A.
+These included genes functionally involved in metabolism, cellular organization, glycoprotein synthesis and DNA repair.
+These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress.
+Namely *A.
 picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A.
@@
 More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
-
-# Significance Statement
-
-**120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty**
 
@@
 picea* in response to thermal extremes (Fig.
-1, 3).
+3).
 The results of the gene set enrichment analysis give an indication of what might drive this difference.
@@
 
-```{r max_exp_PDF, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
+```{r max_exp_PDF, eval = FALSE,aniopts= echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
 # reshape data
@@
 print(maxexplot)
+
+####################################
+### CHANGE TO REACTION NORM FOR HSPs 
+####################################
+
 ```
@@
 carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.
+
+Thermal safety margins...
 
@@
 
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2011].
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].
 Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].

### 2014-08-19 doc/ApTranscriptome_manuscript.Rmd
edits to Intro from SHC

@@
 ---
-title: Investigations of thermal reactionomes reveal stress tolerance vs. resistance in
+title: Investigation of thermal reactionomes reveals stress tolerance vs. resistance in
   warm and cold-climate ant species \\
@@
 Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change.
-Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the evolution of thermal tolerance.
+Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the genetic basis of thermal tolerance.
 We characterized the reactionomes of two common ant species in eastern U.S.
@@
 We found that at least 10% of genes were thermally responsive.
-Interestingly, the warm-adapted *A.
-carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change to thermal extremes than did *A.
+The warm-adapted *A.
+carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression chang than did *A.
 picea*.
@@
 *A.
-carolinensis* shut down more genes than *A.
-picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A.
+carolinensis* also shut down more genes in response to thermal extremes than *A.
 picea*.
-These included genes functionally involved in metabolism, cellular organization, glycoprotein synthesis and DNA repair.
+A subset of these genes were upregulated in *A.
+picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair.
 These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress.
@@
 
+# Significance Statement
+
+Thermal tolerance is a key parameter underlying species distributions, and varies considerably within species.
+However, little is known about the genetic basis of thermal tolerance or how it has evolved.
+In recent years, whole organism gene expression (e.g.
+transcriptomic) studies have been used to gain greater insight on these questions, but these studies are limited to finding genes with large effects in a few conditions.
+Here, we present a method to characterize the reaction norms of gene expression across an ecologically-relevant thermal gradient (e.g.
+the thermal reactionome) for all genes in the transcriptomes of two closely related ant species.
+We find that a large fraction of the transcriptome is thermally-responsive.
+The warmer-climate species down-regulates more genes in response to thermal extremes than the colder-climate species.
+These differences in expression patterns are indicative of divergent strategies in these two species; stress tolerance in the warmer-climate species and stress-resistance in the colder-climate species.
+
 # Introduction
@@
 Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
-These differences within and among species in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-How this variation in thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood.
+These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
+Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved.
 To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.
+This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short term acclimation and adaptation to increased temperatures [@somero2010].
 
 In recent years, whole organism gene expression approaches (e.g.
-transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], among other traits.
-These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few levels of temperature.
-However, temperature varies continuously in nature.
-As a results, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014].
+transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits.
+These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few environmental levels.
+However, temperature and other environmental factors vary continuously in nature.
+As a result, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014].
 Similarly, gene expression studies that look at just a few environments (e.g.
@@
 Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences.
-Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes to gain insight into how spatially divergent selection has led to patterns thermally-responsive gene expression in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
-picea* [@umphrey1996].
-Temperature is a potentially strong selective force on ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
+We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
+picea* [@umphrey1996], and then compare the thermal reactionomes of these species to gain insight into how spatially divergent selection has led to changes in thermally-responsive gene expression.
+
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].
+Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
 On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
+Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
 
@@
 
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].
-Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].
-
 ## Samples

### 2014-08-22 doc/ApTranscriptome_manuscript.Rmd
Added Significance Statement

@@
 
-Thermal tolerance is a key parameter underlying species distributions, and varies considerably within species.
-However, little is known about the genetic basis of thermal tolerance or how it has evolved.
-In recent years, whole organism gene expression (e.g.
-transcriptomic) studies have been used to gain greater insight on these questions, but these studies are limited to finding genes with large effects in a few conditions.
-Here, we present a method to characterize the reaction norms of gene expression across an ecologically-relevant thermal gradient (e.g.
-the thermal reactionome) for all genes in the transcriptomes of two closely related ant species.
-We find that a large fraction of the transcriptome is thermally-responsive.
-The warmer-climate species down-regulates more genes in response to thermal extremes than the colder-climate species.
-These differences in expression patterns are indicative of divergent strategies in these two species; stress tolerance in the warmer-climate species and stress-resistance in the colder-climate species.
+Successfully coping with progressive changes in climate will require both short-term physiological acclimation and longer-term adaptive changes.
+To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
+carolinensis*.
+We found surprisingly divergent transcriptomic responses in the two species.
+*A.
+picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a more robust heat shock response.
+In contrast, *A.
+carolinensis* reacted to thermal stress to a greater extent through downregulation, reducing metabolic gene expression and requiring more extreme temperatures to trigger a protective response.
+These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatic zones.
 

### 2014-08-22 doc/ApTranscriptome_manuscript.Rmd
edits to Intro

@@
 To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.
-This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short term acclimation and adaptation to increased temperatures [@somero2010].
+This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2010].
 
@@
 We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
-picea* [@umphrey1996], and then compare the thermal reactionomes of these species to gain insight into how spatially divergent selection has led to changes in thermally-responsive gene expression.
+picea* [@umphrey1996].
+Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].
 
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].
-Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
-On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
-Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
+We compared the thermal reactionomes of these species to gain insight on the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature.
+First, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses.
+Finding that the same genes up-regulated in response to cold temperatures in *A.
+carolinensis* have increased basal expression in the colder-climate species *A.
+picea*, and vice versa, would be consistent with this pattern.
 
-As the more southern *A.
-carolinensis* occurs in climates that are hotter with longer growing seasons than that of the more northerly *A.
-picea*, we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways.
-First, the number of genes showing plasticity of expression to thermal extremes may differ between the species.
-Finding that more genes are up-regulated to high temperatures in the colder-climate species *A.
+Second, spatially-divergent adaptive responses to temperature may involve changes
+
+the molecular toolkit involved in short-term acclimation to temperature is the same as
+
+finding that more genes are up-regulated to high temperatures in the colder-climate species *A.
 picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A.
-carolinensis*, while the opposite would be true for cold temperatures.
-Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.
+carolinensis*, which might rely to a greater extent on proteome stability [@ghosh2010] and behavioral avoidance mechanisms [@sunday2014], while the opposite would be true for cold temperatures.
 
@@
 Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance.
-Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A.
-carolinensis* may have increased basal expression in the colder-climate species *A.
-picea*, and vice versa.
+Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes
 
@@
 
+...ants...Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
+On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
+Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
+
 We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina ( 36.037° N -78.874° W).

### 2014-08-26 doc/ApTranscriptome_manuscript.Rmd
more edits to Intro

@@
 
-Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences.
+Here, we present a novel method for RNAseq analysis: the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
 We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
@@
 First, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses.
-Finding that the same genes up-regulated in response to cold temperatures in *A.
+Finding that the genes up-regulated in response to cold temperatures in *A.
 carolinensis* have increased basal expression in the colder-climate species *A.
 picea*, and vice versa, would be consistent with this pattern.
+Second, species from more stressful environments may rely more upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
+Finding that fewer genes are up-regulated to high temperatures in the warmer-climate species *A.
+carolinensis* would support this hypothesis.
+Third, spatially-divergent selection often results in local adaptation (cite...).
+For gene expression, this local adaptation may be manifested through changes in the sensitivity of responsive genes to changes in temperature.
+Evidence for local adaptation in gene expression would be observed in heat-responsive genes are up-regulated at higher temperatures in *A.
+carolinensis* than in *A.
+picea*, and conversely, that cold-responsive genes are activated at colder temperatures in *A.
+picea*.
 
-Second, spatially-divergent adaptive responses to temperature may involve changes
+Importantly, none of the above predictions depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012].
+However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis.
+For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].
 
-the molecular toolkit involved in short-term acclimation to temperature is the same as
-
-finding that more genes are up-regulated to high temperatures in the colder-climate species *A.
-picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A.
-carolinensis*, which might rely to a greater extent on proteome stability [@ghosh2010] and behavioral avoidance mechanisms [@sunday2014], while the opposite would be true for cold temperatures.
-
-Second, responsive genes may change in their sensitivity to thermal extremes.
-Finding that heat-responsive genes are up-regulated at higher temperatures in *A.
-carolinensis* than in *A.
-picea* would be consistent with local adaptation for greater heat tolerance.
-That is, more extreme temperatures are required for activation of the heat shock response [@somero2011].
-Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A.
-picea* than *A.
-carolinensis* would indicate local adaptation for cold tolerance.
-Similar patterns may be observed for genes that are down-regulated in response to thermal extremes.
-Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance.
-Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes
-
-Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012].
-However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes using gene set enrichment analysis.
-For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].
 To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

### 2014-08-27 doc/ApTranscriptome_manuscript.Rmd
fixed Rmd to compile

@@
 title: Investigation of thermal reactionomes reveals stress tolerance vs. resistance in
-  warm and cold-climate ant species \\
+  warm and cold-climate ant species
 author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
@@
 
-The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts, hereafter referred to as genes.
-Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that were thermally-responsive at 5% False Discovery Rate.
+The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts.
+Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts genes with expression patterns that were thermally-responsive at 5% False Discovery Rate.
 Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive.
@@
 
-```{r max_exp_PDF, eval = FALSE,aniopts= echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
+```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
 # reshape data

### 2014-08-29 doc/ApTranscriptome_manuscript.Rmd
Added plot for changes in expression type between species using modified contingency table

@@
 
-We compared the thermal reactionomes of these species to gain insight on the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature.
-First, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses.
-Finding that the genes up-regulated in response to cold temperatures in *A.
-carolinensis* have increased basal expression in the colder-climate species *A.
-picea*, and vice versa, would be consistent with this pattern.
-Second, species from more stressful environments may rely more upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
+To gain insight on the genetic basis of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions.
+First, species from more thermally-stressful environments may rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
 Finding that fewer genes are up-regulated to high temperatures in the warmer-climate species *A.
-carolinensis* would support this hypothesis.
-Third, spatially-divergent selection often results in local adaptation (cite...).
+carolinensis* would support this prediction.
+Second, spatially-divergent selection often results in local adaptation [@turesson1922].
 For gene expression, this local adaptation may be manifested through changes in the sensitivity of responsive genes to changes in temperature.
-Evidence for local adaptation in gene expression would be observed in heat-responsive genes are up-regulated at higher temperatures in *A.
+Evidence for local adaptation in gene expression would be observed if heat-responsive genes are up-regulated at higher temperatures in *A.
 carolinensis* than in *A.
@@
 
-Importantly, none of the above predictions depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012].
+In addition, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature.
+Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses.
+Finding that the genes up-regulated in response to cold temperatures in *A.
+carolinensis* have increased basal expression in the colder-climate species *A.
+picea*, and vice versa, would be consistent with this prediction.
+
+Importantly, none of the above predictions depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012; @pavlidis2012].
 However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis.
-For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].
+For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].
 
@@
 library(ggplot2)
+library(reshape2)
+library(pander)
 library(RColorBrewer)
@@
 The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts.
-Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts genes with expression patterns that were thermally-responsive at 5% False Discovery Rate.
-Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive.
-However, the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012].
-Assuming that the number of genes is similar for these two *Aphaenogaster* species, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.
+Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate.
+Of these, xxx had a non-linear relationship with temperature that would not have been detected with a standard differential expression approach.
+At a minimum `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive.
+~~However, the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012].
+Assuming that the number of genes is similar for these two *Aphaenogaster* species, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.~~
 
@@
 
-For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
-Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` had an expression pattern that differed between the two species.
-While some of these were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes were responsive in *A.
-carolinensis* but not *A.
-picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` were responsive in *A.
-picea* but not *A.
-carolinensis*), most of these were genes that changed from one expression category to another between the species (e.g.
-the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
-1).
+For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
 In both species, the majority of genes were in the **Low** category (Fig.
@@
 1).
+Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
+While some of these were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes were responsive in *A.
+carolinensis* but not *A.
+picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` were responsive in *A.
+picea* but not *A.
+carolinensis*), most of these were genes that changed from one expression category to another between the species (i.e.
+the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
+1).
+
+Specifically, all the diagonals (e.g.
+**Low** -- **Low**) the *A.
+carolinensis* **Bimodal** -- *A.
+picea* **Not responsive** and *A.
+carolinensis* **Not responsive** -- *A.
+picea* **High** categories were over-represented compared to a hypothetical homogenous distribution for the contingency table.
+
+Specifically, 
+...categories that were over-expressed relative to null...
+
+```{r expression_change, echo=FALSE}
+cont.tab <- structure(unlist(as.vector(tt2)), .Dim = c(5L, 5L), .Dimnames = structure(list(
+    c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", 
+    "Intermediate", "High", "Bimodal", "NotResp")), .Names = c("Ac", 
+"Ap")))
+
+# marginal sums and total
+rs <- rowSums(cont.tab)
+cs <- colSums(cont.tab)
+n <- sum(cont.tab)
+
+# grand mean for each class
+gm <- (rs + cs) / sum(cont.tab * 2)
+
+# calculate observed values for each cell using overall mean for each expression type
+Ec <- outer(gm, gm, "*") *sum(cont.tab)
+
+# get deviations of observed from expected
+Ec.dev <- cont.tab - Ec
+
+# overall table P-value
+Ec.chistat <- sum((cont.tab-Ec)^2 / Ec)
+# P-value
+pchisq(Ec.chistat, df=(nrow(cont.tab)-1) * (ncol(cont.tab)-1), lower.tail = FALSE)
+
+# overall minimum chi-squared value `X2.df` for a significantly non-homogenous distribution of counts
+df = (ncol(cont.tab) - 1) * (nrow(cont.tab) - 1)
+X2.df <- qchisq(p = 0.05, df = df, lower.tail = FALSE)
+
+# posthoc result for each cell
+Ec.cells <- sign(cont.tab - Ec) * (cont.tab - Ec)^2 / Ec
+Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
+                                                            ifelse(x > X2.df, "+", "-")))
+
+# plot
+qplot(x=Ap, y=Ac, data=melt(Ec.cells), fill=value, geom="tile") + 
+  scale_fill_gradient2(limits=c(-600, 600)) 
+```
 
@@
 
-We looked for evidence of genetic assimilation (the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed )by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
+## Adaptive changes in basal expression are not linked to genes involved in short-term physiological acclimation
+
+We looked for evidence of genetic assimilation (the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed) by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
 carolinensis* or high temperatures in *A.

### 2014-09-16 scripts/sailfish_bootstrap.sh
Rolled back RxNseq function to previous version. Added sailfish_bootstrap.sh script to bootstrap gene expression quantification with Sailfish and modified sailfish_quant.R script to work with this.

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+export PATH=/home/ubuntu/software/Sailfish-0.6.3-Linux_x86-64/bin:$PATH
+export LD_LIBRARY_PATH=/home/ubuntu/software/Sailfish-0.6.3-Linux_x86-64/lib:$LD_LIBRARY_PATH
+
+# script to bootstrap gene expression quantification using sailfish
+parallel Rscript sailfish_quant.R {1} ::: {3..20}
+

### 2014-09-16 scripts/sailfish_quant.R

@@ -1,14 +1,19 @@
 #!/bin/Rscript
 
+args <- commandArgs(trailingOnly = TRUE)
+iout <- args[1]
+
 # list of reads in each of four trimmed classes
 readdir <- "../data/ind_files/"
 
 # sailfish index directory
-system('mkdir -p ../results/trinity-full/sailfish-index-Trinity-cap3-uclust')
 sfindex <- "../results/trinity-full/sailfish-index-Trinity-cap3-uclust"
 
-# make sailfish index
-system("sailfish index -t ../results/trinity-full/Trinity_cap3_uclust.fasta -o ../results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4")
+# make sailfish index if it does not exist
+if(!file.exists(paste(sfindex, "/transcriptome.klut", sep=""))) {
+  print("Making sailfish index")
+  system("sailfish index -t ../results/trinity-full/Trinity_cap3_uclust.fasta -o ../results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4")
+} else print("Sailfish index already exists!")
 
 # prep for sailfish quantification
 readlist <- list.files(readdir)
@@ -17,12 +22,16 @@ paired.right <- readlist[grep("\\.paired.right.fastq$", readlist)]
 unpaired.left <- readlist[grep("unpaired.left.fastq$", readlist)]
 unpaired.right <- readlist[grep("unpaired.right.fastq$", readlist)]
 
-# output directiry
-system('mkdir -p ../results/trinity-full/sailfish-expression-Trinity-cap3-uclust')
-sfexpressionroot <- "../results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
+iout = 1
 
-samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")
- 
+# output directiry
+system(paste('mkdir -p ../results/trinity-full/sailfish-expression-Trinity-cap3-uclust/rep', iout, sep=""))
+
[…]

### 2014-09-18 ApTranscriptome_TR.Rmd
working pipeline to bootstrap gene expression quantification

@@
 # Load Bioconductor libraries
-#source("http://bioconductor.org/biocLite.R")
+source("http://bioconductor.org/biocLite.R")
 #biocLite("topGO")
@@
 sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
+
+samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")
 

### 2014-09-18 scripts/RxNseq.R

@@ -77,26 +77,6 @@ lmFunc <- function(bar) {
 }
 
 
-############################################################################################
-## grepFunc
-############################################################################################
-
-grepFunc <- function(lmitem, term = NA) {
-    # Function to identify models (after stepwise AIC selection) that contain a specific term
-    #
-    # Arguments:
-    #  lmitem: object of class `lm`
-    #  term: term to grep if in model
-    #
-    # Returns
-    #   TRUE/FALSE list of linear models that do or do not include term
-    
-    coefs <- coefficients(lmitem)
-    coef.names <- names(coefs)
-    keep <- if(length(grep(term, coef.names)) > 0) TRUE else FALSE
-}
-
-
 
 
 ############################################################################################
@@ -144,29 +124,19 @@ RxNtype <- function(lmitem) {
         # for min.val, take median point as it may be string of points at zero
         A22.min.val <- median(A22.pout[which(A22.pout$pTPM == min(A22.pout$pTPM)), "val"])
         A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
-
-        # determine expression type
-        A22.exp.type <- NA
-        # "Bimodal" if max expression below 10C and above 30C is greater than one standard deviation above expression at the optimum temperature (19.5C)
+        # determine expression shape
+         A22.exp.type <- NA
+        # expression type defined as "Bimodal" if max expression below 10C and above 30C is greater than one standard deviation above expression at the optimum temperature (19.5C)
         if(max(A22.pout[A22.pout$val < 10, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + sd(A22.pout$pTPM)) & max(A22.pout[A22.pout$val > 30, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + sd(A22.pout$pTPM))) A22.exp.type <- "Bimodal" else {
-              # "PositiveLinear" if max expression at temperature above 30C and minimum expression at 0C
-              if(A22.max.val > 30 & A22.min.val == 0) A22.exp.type <- "PositiveLinear" else {
-                  # "HighOn" if max expression above 30C and minimum not at 0C (due to quadratic curvature)
-                  if(A22.max.val > 30 & A22.min.val != 0) A22.exp.type <- "HighOn" else {
[…]

### 2014-09-18 scripts/sailfish_bootstrap.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+export PATH=/home/ubuntu/software/Sailfish-0.6.3-Linux_x86-64/bin:$PATH
+export LD_LIBRARY_PATH=/home/ubuntu/software/Sailfish-0.6.3-Linux_x86-64/lib:$LD_LIBRARY_PATH
+
+# script to bootstrap gene expression quantification using sailfish
+parallel Rscript sailfish_quant.R {1} ::: {1..20}
+

### 2014-09-18 scripts/sailfish_quant.R

@@ -1,14 +1,19 @@
 #!/bin/Rscript
 
+args <- commandArgs(trailingOnly = TRUE)
+iout <- args[1]
+
 # list of reads in each of four trimmed classes
 readdir <- "../data/ind_files/"
 
 # sailfish index directory
-system('mkdir -p ../results/trinity-full/sailfish-index-Trinity-cap3-uclust')
 sfindex <- "../results/trinity-full/sailfish-index-Trinity-cap3-uclust"
 
-# make sailfish index
-system("sailfish index -t ../results/trinity-full/Trinity_cap3_uclust.fasta -o ../results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4")
+# make sailfish index if it does not exist
+if(!file.exists(paste(sfindex, "/transcriptome.klut", sep=""))) {
+  print("Making sailfish index")
+  system("sailfish index -t ../results/trinity-full/Trinity_cap3_uclust.fasta -o ../results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4")
+} else print("Sailfish index already exists!")
 
 # prep for sailfish quantification
 readlist <- list.files(readdir)
@@ -18,11 +23,13 @@ unpaired.left <- readlist[grep("unpaired.left.fastq$", readlist)]
 unpaired.right <- readlist[grep("unpaired.right.fastq$", readlist)]
 
 # output directiry
-system('mkdir -p ../results/trinity-full/sailfish-expression-Trinity-cap3-uclust')
-sfexpressionroot <- "../results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
+system(paste('mkdir -p ../results/trinity-full/sailfish-expression-Trinity-cap3-uclust/rep', iout, sep=""))
 
-samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")
- 
+sfexpressionroot <- paste('../results/trinity-full/sailfish-expression-Trinity-cap3-uclust/rep', iout, "/", sep="")
+
[…]

### 2014-09-18 ApTranscriptome_TR.Rmd
working on analysis of Sailfish replicates

--- 2014-07-25 ApTranscriptome_TR.Rmd
@@
 
+Preliminary runs of Sailfish revealed that while quantification is highly consistent among runs, small stochastic differences impact which transcripts are found to be thermally-responsive.
+To address this, I repeated that expression quantification 10 times and retain only the transcripts that are responsive in at least 8 of to 10 replicates.
+
+The gene expression replicates are generated by running the `scripts/sailfish_rep.sh` script, which uses GNU parallel to repeat the `scripts/sailfish_quant.R` script which runs Sailfish for each of the 24 samples.
+
 These files are downloaded for convenience:
@@
 
+**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.**
+
 ```{r load_expression_data, eval=TRUE, echo=FALSE, results='hide'}
@@
 sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
-
-# read in each file using loop
-for (j in 1:length(samples)) {
-    samp <- samples[j]
-    trtval <- as.numeric(str_split_fixed(samp, "-", 2)[2])
-    outpre <- gsub("-", "_", samp)
-    outname <- paste(outpre, "_quant", sep="")
-    read.sailfish.quant(filein=paste(sfexpressionroot, samp, "_quant/quant_bias_corrected.sf", sep=""), outname=outname, samp = samp, trtval = trtval)
+# samples
+samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")
+
+# loop across replicates
+reps <- c("rep1", "rep2", "rep3", "rep4", "rep5", "rep6", "rep7", "rep8", "rep9", "rep10")
+
+for(sr in reps) {
+  read.sailfish.reps(sr, fd=sfexpressionroot, repout=paste("TPM.dt.", sr, sep=""))
   }
-
-# combine into long format
-A22.TPM <- rbind(A22_0_quant, A22_3_quant, A22_7_quant, A22_10_quant, A22_14_quant, A22_17_quant, A22_21_quant, A22_24_quant, A22_28_quant, A22_31_quant, A22_35_quant, A22_38_quant)
-str(A22.TPM)
-
-# convert to data.table
-A22.TPM <- data.table(A22.TPM)
-setkey(A22.TPM, Transcript, val)
-
-# set "trt" to true values - truncated in file names for convenience
-A22.TPM[val==3,val:=3.5]
-A22.TPM[val==10,val:=10.5]
-A22.TPM[val==17,val:=17.5]
-A22.TPM[val==24,val:=24.5]
-A22.TPM[val==31,val:=31.5]
-A22.TPM[val==38,val:=38.5]
-str(A22.TPM)
-
-# Repeat for Ar
-Ar.TPM <- rbind(Ar_0_quant, Ar_3_quant, Ar_7_quant, Ar_10_quant, Ar_14_quant, Ar_17_quant, Ar_21_quant, Ar_24_quant, Ar_28_quant, Ar_31_quant, Ar_35_quant, Ar_38_quant)
-Ar.TPM <- data.table(Ar.TPM)
-setkey(Ar.TPM, Transcript, val)
-Ar.TPM[val==3,val:=3.5]
-Ar.TPM[val==10,val:=10.5]
-Ar.TPM[val==17,val:=17.5]
-Ar.TPM[val==24,val:=24.5]
-Ar.TPM[val==31,val:=31.5]
-Ar.TPM[val==38,val:=38.5]
-str(Ar.TPM)
-```
-
-Note that expression levels at each temperature treatment are highly correlated between the two colonies.
+```
+
+For any single replicate, the expression levels at each temperature treatment are highly correlated between the two colonies.
 
 ```{r exp_correlations, echo=FALSE, results='hide'}
-temp <- c(0, 3.5, 7, 10.5, 14, 17.5, 21, 24.5, 28, 31.5, 35, 38.5)
-cors <- c(round(cor(Ar_0_quant$TPM, A22_0_quant$TPM), 2), round(cor(Ar_3_quant$TPM, A22_3_quant$TPM), 2), round(cor(Ar_7_quant$TPM, A22_7_quant$TPM), 2), round(cor(Ar_10_quant$TPM, A22_10_quant$TPM), 2), round(cor(Ar_14_quant$TPM, A22_14_quant$TPM), 2), round(cor(Ar_17_quant$TPM, A22_17_quant$TPM), 2), round(cor(Ar_21_quant$TPM, A22_21_quant$TPM), 2), round(cor(Ar_24_quant$TPM, A22_24_quant$TPM), 2), round(cor(Ar_28_quant$TPM, A22_28_quant$TPM), 2), round(cor(Ar_31_quant$TPM, A22_31_quant$TPM), 2), round(cor(Ar_35_quant$TPM, A22_35_quant$TPM), 2), round(cor(Ar_38_quant$TPM, A22_38_quant$TPM), 2))
-
-cortable <- cbind(Temperature = temp, Correlation = cors)
+cortable <- ddply(TPM.dt.rep1, .(val), summarize, cor(TPM[which(colony=="A22")], TPM[which(colony=="Ar")]))
 ```
@@
 ```{r exp_correlations_table, echo=FALSE, results='asis'}
-pandoc.table(cortable, style="rmarkdown", caption = "Correlations between species for gene expression at temperature treatment")
-```
-
-**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.**
-
-```{r combined_TPM, results = 'hide'}
-A22.TPM[,colony:="A22"]
-Ar.TPM[,colony:="Ar"]
-TPM.dt <- rbind(A22.TPM, Ar.TPM)
-TPM.dt$colony <- as.factor(TPM.dt$colony)
-str(TPM.dt)
-
-setkey(TPM.dt, val)
-TPM.dt.sub <- TPM.dt[val != 7] 
-unique(TPM.dt.sub$val)
-length(unique(TPM.dt.sub$Transcript))
+colnames(cortable) <- c("Temperature", "r")
+pandoc.table(cortable, style="rmarkdown", caption = "Correlations between species for gene expression at temperature treatment", round=2)
 ```
@@
 
-# remove from TPM.dt.sub
-setkey(TPM.dt.sub, Transcript)
-TPM.dt.sub <- TPM.dt.sub[!cont.list]
-str(TPM.dt.sub)
-length(unique(TPM.dt.sub$Transcript))
+# setkey of each TPM replicate to 'Transcript'
+ls1 <- ls(pattern="TPM")
+sapply(mget(ls1), function(x) setkey(x, Transcript))
+# remove Transcripts found in contaminant list
+sapply(mget(ls1), function(x) x[!cont.list])
 

### 2014-09-18 scripts/RxNseq.R

@@ -25,6 +25,53 @@ read.sailfish.quant <- function(filein, outname, samp, trtval) {
 }
 
 
+############################################################################################
+## read.sailfish.reps
+############################################################################################
+
+
+read.sailfish.reps <- function(r, fd=".", repout){
+  repdir <- paste(fd, r, "/", sep="")
+  # read in each file using loop
+  for (j in 1:length(samples)) {
+    samp <- samples[j]
+    trtval <- as.numeric(str_split_fixed(samp, "-", 2)[2])
+    outpre <- gsub("-", "_", samp)
+    outname <- paste(outpre, "_quant", sep="")
+    read.sailfish.quant(filein=paste(repdir, samp, "_quant/quant_bias_corrected.sf", sep=""), 
+                        outname=outname, samp = samp, trtval = trtval)
+  } 
+  
+  # combine results for each treatment into long format as data.table
+  A22.TPM <- data.table(rbind(A22_0_quant, A22_3_quant, A22_7_quant, A22_10_quant, A22_14_quant, A22_17_quant, 
+                              A22_21_quant, A22_24_quant, A22_28_quant, A22_31_quant, A22_35_quant, A22_38_quant))
+  A22.TPM[,colony:="A22"]
+  Ar.TPM <- data.table(rbind(Ar_0_quant, Ar_3_quant, Ar_7_quant, Ar_10_quant, Ar_14_quant, Ar_17_quant, 
+                             Ar_21_quant, Ar_24_quant, Ar_28_quant, Ar_31_quant, Ar_35_quant, Ar_38_quant))
+  Ar.TPM[,colony:="Ar"]
+  
+  # combine each colony into single data.table
+  TPM.dt <- rbind(A22.TPM, Ar.TPM)
+  TPM.dt$colony <- as.factor(TPM.dt$colony)
+  setkey(TPM.dt, val)
+  
+  # set "trt" to true values - truncated in file names for convenience
+  TPM.dt[val==3,val:=3.5]
+  TPM.dt[val==10,val:=10.5]
+  TPM.dt[val==17,val:=17.5]
+  TPM.dt[val==24,val:=24.5]
+  TPM.dt[val==31,val:=31.5]
+  TPM.dt[val==38,val:=38.5]
+  
+  # drop treatment value of 7
+  TPM.dt.sub <- TPM.dt[val != 7] 
+  unique(TPM.dt.sub$val)
+  length(unique(TPM.dt.sub$Transcript))
+  
+  # return data.table
[…]

### 2014-09-18 ApTranscriptome_TR.Rmd
Completed analysis of replicated Sailfish gene expression quantification. Turns out that quantification is *very* consistent among repeated runs. The differences between my previous and current results are (probably) due to upgraded version of the Sailfish software.

@@
 
-# loop across replicates
+# replicates
 reps <- c("rep1", "rep2", "rep3", "rep4", "rep5", "rep6", "rep7", "rep8", "rep9", "rep10")
 
+# loop across replicates and load expression data
 for(sr in reps) {
@@
 # setkey of each TPM replicate to 'Transcript'
-ls1 <- ls(pattern="TPM")
+ls1 <- ls(pattern="TPM.dt.rep")
 sapply(mget(ls1), function(x) setkey(x, Transcript))
+sapply(mget(ls1), function(x) key(x))
 # remove Transcripts found in contaminant list
-sapply(mget(ls1), function(x) x[!cont.list])
-
-# remove from annotation.table
+TPM.dt.rep1 <- TPM.dt.rep1[!cont.list]
+TPM.dt.rep2 <- TPM.dt.rep2[!cont.list]
+TPM.dt.rep3 <- TPM.dt.rep3[!cont.list]
+TPM.dt.rep4 <- TPM.dt.rep4[!cont.list]
+TPM.dt.rep5 <- TPM.dt.rep5[!cont.list]
+TPM.dt.rep6 <- TPM.dt.rep6[!cont.list]
+TPM.dt.rep7 <- TPM.dt.rep7[!cont.list]
+TPM.dt.rep8 <- TPM.dt.rep8[!cont.list]
+TPM.dt.rep9 <- TPM.dt.rep9[!cont.list]
+TPM.dt.rep10 <- TPM.dt.rep10[!cont.list]
+
+# remove contaminants from annotation.table
 setkey(annotation.table, Sequence.Name)
@@
 # calculate overall P value and R^2 for each transcript
-RxNpval <- ddply(TPM.dt.sub, .(Transcript), .inform="TRUE", modpFunc)
+RxNpval.r1 <- ddply(TPM.dt.rep1, .(Transcript), .inform="TRUE", modpFunc)
+RxNpval.r2 <- ddply(TPM.dt.rep2, .(Transcript), .inform="TRUE", modpFunc)
+RxNpval.r3 <- ddply(TPM.dt.rep3, .(Transcript), .inform="TRUE", modpFunc)
+RxNpval.r4 <- ddply(TPM.dt.rep4, .(Transcript), .inform="TRUE", modpFunc)
+
 ```
@@
 ```{r padjust}
-RxNpval$padj <- p.adjust(RxNpval$pval, method = "fdr")
+RxNpval.r1$padj <- p.adjust(RxNpval.r1$pval, method = "fdr")
+RxNpval.r2$padj <- p.adjust(RxNpval.r2$pval, method = "fdr")
+RxNpval.r3$padj <- p.adjust(RxNpval.r3$pval, method = "fdr")
+RxNpval.r4$padj <- p.adjust(RxNpval.r4$pval, method = "fdr")
+
 # Plot FDR values against initial pvalues
@@
 # subset to significant transcripts
-signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]
+signif.transcripts.r1 <- RxNpval.r1[which(RxNpval.r1$padj < 0.05), ]
+signif.transcripts.r2 <- RxNpval.r2[which(RxNpval.r2$padj < 0.05), ]
+signif.transcripts.r3 <- RxNpval.r3[which(RxNpval.r3$padj < 0.05), ]
+signif.transcripts.r4 <- RxNpval.r4[which(RxNpval.r4$padj < 0.05), ]
 

### 2014-09-19 scripts/RxNseq.R
merged in update to read.sailfish.quant function for new data format

@@ -16,7 +16,7 @@ read.sailfish.quant <- function(filein, outname, samp, trtval) {
     #  outname: name for data.frame in R
     
     file.df <- read.table(filein, header=FALSE, sep="\t", stringsAsFactors = FALSE)
-    colnames(file.df) <- c("Transcript", "Length", "TPM", "RPKM", "KPKM", "EstimatedNumReads")
+    colnames(file.df) <- c("Transcript", "Length", "TPM", "RPKM", "KPKM", "EstimatedNumKmers", "EstimatedNumReads")
     #head(file.df)
     # add columns with sample ID and treatment
     file.df$sample <- samp

### 2014-09-19 scripts/RxNseq.R
fixed some broken scripts due to a bad previous rollback

@@ -16,7 +16,8 @@ read.sailfish.quant <- function(filein, outname, samp, trtval) {
     #  outname: name for data.frame in R
     
     file.df <- read.table(filein, header=FALSE, sep="\t", stringsAsFactors = FALSE)
-    colnames(file.df) <- c("Transcript", "Length", "TPM", "RPKM", "KPKM", "EstimatedNumKmers", "EstimatedNumReads")
+    colnames(file.df) <- c("Transcript", "Length", "TPM", "RPKM", "KPKM", "EstimatedNumKmers", 
+                           "EstimatedNumReads")
     #head(file.df)
     # add columns with sample ID and treatment
     file.df$sample <- samp
@@ -24,6 +25,25 @@ read.sailfish.quant <- function(filein, outname, samp, trtval) {
     assign(outname, file.df, envir = .GlobalEnv)
 }
 
+
+############################################################################################
+## grepFunc
+############################################################################################
+
+grepFunc <- function(lmitem, term = NA) {
+  # Function to identify models (after stepwise AIC selection) that contain a specific term
+  #
+  # Arguments:
+  #  lmitem: object of class `lm`
+  #  term: term to grep if in model
+  #
+  # Returns
+  #   TRUE/FALSE list of linear models that do or do not include term
+  
+  coefs <- coefficients(lmitem)
+  coef.names <- names(coefs)
+  keep <- if(length(grep(term, coef.names)) > 0) TRUE else FALSE
+}
 
 
 ############################################################################################
@@ -294,7 +314,7 @@ RxNvarshape <- function(lmitem) {
 ## geneid2GOmap
 ############################################################################################
 
-geneid2GOmap <- function(annotmat) {
+geneid2GOmap <- function(annotmat, ontology = c("BP", "CC", "MF"), mapname = "geneid2go.map") {
   # Create geneid2go.map file from AnnotationTable.txt produced by FastAnnotator 
   #
   # Args:
@@ -306,24 +326,35 @@ geneid2GOmap <- function(annotmat) {
   # requires
   require(stringr)
   
+  # check that mapname doesn't already exist
+  if(file.exists(mapname)) stop(paste("File ", mapname, " already exists!", sep =""))
+  
   # Extract all GO terms and combine, write to file
   for(r in 1:nrow(annotmat)) {
[…]

### 2014-09-19 ApTranscriptome_TR.Rmd
complete functioning script

--- 2014-09-18 ApTranscriptome_TR.Rmd
@@
 
-To remove contigs that are likely contaminants from bacterial, archael, virus or human sources, we used the program [DeconSeq](http://deconseq.sourceforge.net/) `r citep("10.1371/journal.pone.0017288")`.
+To remove contigs that are likely contaminants from bacterial, archael, virus or human sources, we used the program [DeconSeq](http://deconseq.sourceforge.net/) r citep("10.1371/journal.pone.0017288").
 We downloaded the bacteria, virus, archae and human [databases of contaminants](ftp://edwards.sdsu.edu:7009/deconseq/db), modified the `DeconSeqConfig.pm` file as described [here](http://www.vcru.wisc.edu/simonlab/bioinformatics/programs/install/deconseq.htm) to point to the databases, and ran DeconSeq specifiying 95% identity over 50% the length of contig
@@
 
-To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) `r citep("doi:10.1186/1471-2105-12-385")`. I ...
+To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) r citep("doi:10.1186/1471-2105-12-385"). I ...
 
@@
 
-Annotation was performed by uploading the reduced assembly "Trinity_cap3_uclust.fasta" to the web-based annotation program [FastAnnotator](http://fastannotator.cgu.edu.tw/index.php) `r citep("10.1186/1471-2164-13-S7-S9")`.
+Annotation was performed by uploading the reduced assembly "Trinity_cap3_uclust.fasta" to the web-based annotation program [FastAnnotator](http://fastannotator.cgu.edu.tw/index.php).
 
@@
 
-Preliminary runs of Sailfish revealed that while quantification is highly consistent among runs, small stochastic differences impact which transcripts are found to be thermally-responsive.
-To address this, I repeated that expression quantification 10 times and retain only the transcripts that are responsive in at least 8 of to 10 replicates.
-
-The gene expression replicates are generated by running the `scripts/sailfish_rep.sh` script, which uses GNU parallel to repeat the `scripts/sailfish_quant.R` script which runs Sailfish for each of the 24 samples.
-
 These files are downloaded for convenience:
@@
 sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"
-# samples
+
 samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")
 
-# loop across replicates
-reps <- c("rep1", "rep2", "rep3", "rep4", "rep5", "rep6", "rep7", "rep8", "rep9", "rep10")
-
-for(sr in reps) {
-  read.sailfish.reps(sr, fd=sfexpressionroot, repout=paste("TPM.dt.", sr, sep=""))
+# read in each file using loop
+for (j in 1:length(samples)) {
+    samp <- samples[j]
+    trtval <- as.numeric(str_split_fixed(samp, "-", 2)[2])
+    outpre <- gsub("-", "_", samp)
+    outname <- paste(outpre, "_quant", sep="")
+    read.sailfish.quant(filein=paste(sfexpressionroot, samp, "_quant/quant_bias_corrected.sf", sep=""), outname=outname, samp = samp, trtval = trtval)
 }
-```
-
-For any single replicate, the expression levels at each temperature treatment are highly correlated between the two colonies.
+
+# combine into long format data.table
+A22.TPM <- data.table(rbind(A22_0_quant, A22_3_quant, A22_7_quant, A22_10_quant, A22_14_quant, A22_17_quant, A22_21_quant, A22_24_quant, A22_28_quant, A22_31_quant, A22_35_quant, A22_38_quant))
+A22.TPM[,colony:="A22"]
+
+Ar.TPM <- data.table(rbind(Ar_0_quant, Ar_3_quant, Ar_7_quant, Ar_10_quant, Ar_14_quant, Ar_17_quant, Ar_21_quant, Ar_24_quant, Ar_28_quant, Ar_31_quant, Ar_35_quant, Ar_38_quant))
+Ar.TPM[,colony:="Ar"]
+
+# combine colonies into single data.frame
+TPM.dt <- rbind(A22.TPM, Ar.TPM)
+TPM.dt$colony <- as.factor(TPM.dt$colony)
+str(TPM.dt)
+
+# drop temperature 7 
+setkey(TPM.dt, val)
+TPM.dt.sub <- TPM.dt[val != 7]
+
+length(unique(TPM.dt.sub$Transcript))
+
+# set "trt" to true values - truncated in file names for convenience
+TPM.dt.sub[val==3,val:=3.5]
+TPM.dt.sub[val==10,val:=10.5]
+TPM.dt.sub[val==17,val:=17.5]
+TPM.dt.sub[val==24,val:=24.5]
+TPM.dt.sub[val==31,val:=31.5]
+TPM.dt.sub[val==38,val:=38.5]
+str(TPM.dt.sub)
+unique(TPM.dt.sub$val)
+```
+
+Note that expression levels at each temperature treatment are highly correlated between the two colonies.
 
 ```{r exp_correlations, echo=FALSE, results='hide'}
-cortable <- ddply(TPM.dt.rep1, .(val), summarize, cor(TPM[which(colony=="A22")], TPM[which(colony=="Ar")]))
+cortable <- ddply(TPM.dt.sub, .(val), summarize, cor(TPM[which(colony=="A22")], TPM[which(colony=="Ar")]))
 ```
@@
 
-# setkey of each TPM replicate to 'Transcript'
-ls1 <- ls(pattern="TPM")
-sapply(mget(ls1), function(x) setkey(x, Transcript))
-# remove Transcripts found in contaminant list
-sapply(mget(ls1), function(x) x[!cont.list])
+# remove from TPM.dt.sub
+setkey(TPM.dt.sub, Transcript)
+TPM.dt.sub <- TPM.dt.sub[!cont.list]
+str(TPM.dt.sub)
+length(unique(TPM.dt.sub$Transcript))
 
@@
 
-pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts of 99,861 total with expression that depends on species, temperature or their interaction at 5% FDR.")
+pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts with expression that depends on species, temperature or their interaction at 5% FDR  out of 98,186 total transcripts.")
 ```
@@
 
-From this analysis, about 3/4 of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.
+From this analysis, about 80% of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.
 
@@
 
-There are also 7 heat shock related genes in the responsive transcripts, out of 130 total.
-
 ```{r grep_shock}
@@
 ```
+
+There are also `r nrow(hsp_responsive)` heat shock related genes in the responsive transcripts, out of `r nrow(hsp_all)` total heat shock related genes in the transcriptome annotation.
 
@@
 
-```{r gsea_combine, eval = FALSE}
+```{r gsea_combine, eval = TRUE}
 # combine into single table
@@
 
-Make plots for all genes expressed at *High* temps in GO category "GO:0006950: response to stress"
-
-```{r plot_GOstress, echo=FALSE, eval=FALSE}
+Plots for all genes expressed at *High* temps in GO category "GO:0006950: response to stress"
+
+```{r plot_GOstress, echo=FALSE, eval=TRUE}
 high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
@@
 
-trp_A22_poslinear <- ggplot(resp.TPM.dt.sub[A22.poslinear.transcripts][1:220,], aes(x=val, y=TPM.scaled, group=Transcript)) +
-  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
-  facet_grid(. ~ colony2) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(trp_A22_poslinear)
-
 ### predicted values
@@
 print(pred_A22_high)
-
-pred_A22_poslinear <- ggplot(resp.TPM.dt.sub.pred[A22.poslinear.transcripts][1:220,], aes(x=val, y=pTPM.scaled, group=Transcript)) +
-  geom_line() + 
-  facet_grid(. ~ colony) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(pred_A22_poslinear)
 
@@
 dev.off()
-
-```
-
-### Overlap between *Ar* Intermediate and *A22* Low genes
-
-An observation from GO analysis is that there is an overlap between enriched terms in the *Low* group for *A22* and the *Intermediate* group for *Ar*.
-A potential explanation for this pattern is if the *Low* genes in *A22* shut down at high temperatures, but not low temperatures, rather than are up-regulated at low temperatures.
-
-```{r gene_overlap}
-# transcripts where TPM increases at low temps & decreases at high temps
-A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low,Ar.int), "A22")]
-
-for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
-    s <- A22.low.Ar.int.TPM.dt.sub[i]
-    X11()
-    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
-    points(s$val, s$TPM, pch = 8, col = "red")
-    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh") else {
-        print("OnLow")
-    }
-}
-
-# tabulate the number of transcripts that have lowest expression at maximum temperature "OffHigh"
-A22.low.Ar.int.low.type <- ddply(A22.low.Ar.int.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))
-
-table(A22.low.Ar.int.low.type$lt)
-
-# A22 low Ar low
-A22.low.Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low, Ar.low), "A22")]
-
-A22.low.Ar.low.low.type <- ddply(A22.low.Ar.low.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))
-
-# table results and perform chi-square test
-low.table <- rbind(LowInt = table(A22.low.Ar.int.low.type$lt), LowLow = table(A22.low.Ar.low.low.type$lt))
-low.table
-chisq.test(low.table)
 ```
@@
 # save RData for manuscript file
-save(RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.intersect, Ap.gsea.union, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
-```
-
-## Session information ##
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.union, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
+```
+
+## Session information
 

### 2014-09-22 doc/ApTranscriptome_manuscript.Rmd
writing

@@
 carolinensis* would support this prediction.
-Second, spatially-divergent selection often results in local adaptation [@turesson1922].
+Second, spatially-divergent local adaptation leaves a signal at the molecular level [@savolainen2013].
 For gene expression, this local adaptation may be manifested through changes in the sensitivity of responsive genes to changes in temperature.
@@
 picea*, and vice versa, would be consistent with this prediction.
-
-Importantly, none of the above predictions depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012; @pavlidis2012].
-However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis.
+While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis.
 For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].
-
-To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
+To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
 
@@
 
-## Species comparisons in the number of thermally-responsive genes
+## Species differences in thermally-responsive genes
 
@@
 
-To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes.
-The **High** genes in *A.
+To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures.
+Genes turned on at high temperatures in *A.
 carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A.
-picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
+picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
 2).
-For genes with increased expression at **Low** temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
+For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
-picea*; *t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
 2).
@@
 
-We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina ( 36.037° N -78.874° W).
+We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W).
 The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers.

### 2014-09-22 ApTranscriptome_TR.Rmd

@@
 
-pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts with expression that depends on species, temperature or their interaction at 5% FDR  out of 98,186 total transcripts.")
+pandoc.table(sigtable, style="rmarkdown", big.mark = ",", caption = "Number of transcripts with expression that depends on species, temperature or their interaction at 5% FDR  out of 98,186 total transcripts.")
 ```
@@
 # Reorganize for plotting to show overlap among categories between colonies
-type.table <- table(Ap.response.type[ , 'Ar.type'], Ap.response.type[ , 'A22.type'])
+type.table <- table(Acar = Ap.response.type[ , 'Ar.type'], Apic = Ap.response.type[ , 'A22.type'])
 # reorder
@@
 dev.off()
+```
+
+The question of biological interest is whether the marginal frequencies differ between the two species.
+Statistically, this can be addressed using the generalized [McNemar's test](http://en.wikipedia.org/wiki/McNemar's_test) of marginal homogeneity.
+
+```{r mh_test}
+mh.test <- mh_test(tt2)
+
+# calculate contribution of each off-diagonal to Z~0~
+d1 <- sum(tt2[1,2:5]) - sum(tt2[2:5,1])
+d2 <- sum(tt2[2,c(1,3:5)]) - sum(tt2[c(1,3:5),2])
+d3 <- sum(tt2[3,c(1:2,4:5)]) - sum(tt2[c(1:2,4:5),3])
+d4 <- sum(tt2[4,c(1:3,5)]) - sum(tt2[c(1:3,5),4])
+d5 <- sum(tt2[5,1:4]) - sum(tt2[1:4,5])
+
+# proportion of test statistic due to off-diagonals
+dsum <- sum(abs(d1), abs(d2), abs(d3), abs(d4), abs(d5))
+d1/dsum
+d2/dsum
+d3/dsum
+d4/dsum
+d5/dsum
+
+# e2 (Acar Intermediate) contributes ~44% to Z~0~
+```
+
+```{r mh_plot}
+# overall mean for each class
+rs <- rowSums(tt2)
+cs <- colSums(tt2)
+(gm <- (rs + cs) / sum(tt2 * 2))
+
+# calculate observed values for each cell using overall mean for each expression type
+Ec <- outer(gm, gm, "*") * sum(tt2)
+
+# get deviations of observed from expected
+Ec.dev <- tt2 - Ec
+
+# calculate chi-squared deviation 
+Ec.cells <- sign(tt2 - Ec) * (tt2 - Ec)^2 / Ec
+
+md <- melt(Ec.cells)
+
+qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
+  scale_fill_gradient2(limits=c(-600, 600)) 
 ```
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.union, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, md, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.union, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
 ```

### 2014-09-23 doc/ApTranscriptome_manuscript.Rmd
updates to ms

@@
 
-## Species differences in thermally-responsive genes
-
 The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts.
@@
 
+## Thermal responses involve greater down-regulation in the warmer-climate species
+
 For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
@@
 
-Specifically, 
-...categories that were over-expressed relative to null...
+In contrast to expectations, more genes did not respond to high temperatures in *A.
+carolinensis*.
+In fact, the only differences were for **Intermediate** genes in *A.
+carolinensis*, which disproportionatly switched to **Low** or **Bimodal** expression in *A.
+picea*.
+Thus, *A.
+carolinensis*, the warmer-climate species, appears to rely upon shutting down gene expression to a greater extent than *A.
+picea*, which activates gene expression...
 
@@
 
-## Sensitivity of thermally-responsive genes to changes in temperature
+## Greater inertia of gene expression response to heat stress in the warmer climate species
 
-To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures.
+To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures, including *Bimodal* genes.
 Genes turned on at high temperatures in *A.
@@
 
-For **Intermediate** and **Bimodal** genes, genes had two inflection points so it was not possible to calculate a single temperature of gene activation.
-Instead, we calculated the variance of 1,000 random temperatures, weighted by expression at that temperature.
-For **Intermediate** genes, a greater variance would be indicative of less sensitivity to changes in temperature.
-For **Bimodal** genes, a greater variance would be indicative of a more sensitivity to changes in temperature (convex reaction norm).
-We found that **Intermediate** genes from *A.
-carolinensis* had a greater variance of expression than those from *A.
-picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`), indicating greater inertia of gene expression to changes in temperature.
-However, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`) for the **Bimodal** genes.
+Similarly, for the **Intermediate** genes we compared the mean temperatures of downregulation of gene expression among species at both high (> 21°) and low (< 21°C) temperatures.
+Consistent with the results for up-regulated genes, we found that with increasing temperatures, the **Intermediate** genes were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
+carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
+picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A.
+carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A.
+picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
+Overall, these results indicate greater intertia of gene expression to changes in temperature for *A.
+carolinensis* than *A.
+picea*.
 
@@
 picea*.
+While finding that genes upregulated to thermal extremes
+
 Contrary to expectations, for **Low** genes in *A.

### 2014-09-23 ApTranscriptome_TR.Rmd

@@
 # download filtered Trinity assembly, uncompress and move
-wget http://johnstantongeddes.org/assets/files/Aphaenogaster_transcriptome.tar
+wget http://johnstantongeddes.org/assets/files/ApTranscriptome/Aphaenogaster_transcriptome.tar
 # check md5sum
@@
 
-# uncompress and move
-tar -xvf Aphaenogaster_transcriptome.tar
+# move and extract
 mkdir -p results/
 mkdir -p results/trinity-full/
-mv Trinity_cap3_uclust.fasta results/trinity-full/.
-mv Trinity_cap3_uclust_clean.fasta results/trinity-full/.
+mv Aphaenogaster_transcriptome.tar results/trinity-full/.
+tar -xvf Aphaenogaster_transcriptome.tar
 ~~~
@@
 # URL for annotation file
-annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
+annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome/ApTranscriptome_AnnotationTable_20140113.txt")
 # load 
@@
 ~~~
-wget ...
+# download gene expression quantification
+wget http://johnstantongeddes.org/assets/files/ApTranscriptome/sailfish_quant_20140916.tar
+# check md5sum
+md5sum sailfish_quant_20140916.tar
+# 26102c7ef86cf30b5f8e923640378185
+
+# move and extract
 mkdir -p /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
-mv .. /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
-tar -zxvf ...
+mv sailfish_quant_20140916.tar results/trinity-full/.
+tar -xvf sailfish_quant_20140916.tar
 ~~~
@@
 Calculate T~on~ for *High* genes in each species.
+I use the observed (rather than predicted) values for each species as by nature of fitting a quadratic function, the maximum change tends to be at the extremes (38.5 or 0) for the predicted values.
 
@@
 ```
+
+Repeat analysis for when **Intermediate** genes are turned *off*.
+
+```{r T_int_off}
+# transcripts expressed at *Intermediate* temperatures in A22
+A22.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.int.transcripts, "A22")]
+setkey(A22.int.TPM.dt.sub, Transcript)
+str(A22.int.TPM.dt.sub)
+
+# make data.frame for results
+l5 <- length(unique(A22.int.TPM.dt.sub$Transcript))
+A22.int.T_off <- data.frame(Transcript = unique(A22.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l5), type = rep("Low", length = l5), T_off_high = NA, T_off_low = NA)
+
+# loop across transcripts, calculating T_off at both high and low temperatures
+for(i in unique(A22.int.TPM.dt.sub$Transcript)) {
+    subdf <- A22.int.TPM.dt.sub[i]
+
+    subdf.high <- subdf[which(subdf$val >= 21), ]
+    T_off_high <- subdf.high[median(which(diff(subdf.high$TPM) == min(diff(subdf.high$TPM)))), val]
+    A22.int.T_off[which(A22.int.T_off$Transcript == i), "T_off_high"] <- T_off_high
+
+    subdf.low <- subdf[which(subdf$val <= 21), ]
+    T_off_low <- subdf.low[median(which(diff(subdf.low$TPM) == min(diff(subdf.low$TPM)))), val]
+    A22.int.T_off[which(A22.int.T_off$Transcript == i), "T_off_low"] <- T_off_low
+}
+
+# repeat for Ar
+Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.int.transcripts, "Ar")]
+setkey(Ar.int.TPM.dt.sub, Transcript)
+str(Ar.int.TPM.dt.sub)
+
+# make data.frame for results
+l6 <- length(unique(Ar.int.TPM.dt.sub$Transcript))
+Ar.int.T_off <- data.frame(Transcript = unique(Ar.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l6), type = rep("Low", length = l6), T_off_high = NA, T_off_low = NA)
+
+# loop across transcripts, calculating T_off at both high and low temperatures
+for(i in unique(Ar.int.TPM.dt.sub$Transcript)) {
+    subdf <- Ar.int.TPM.dt.sub[i]
+
+    subdf.high <- subdf[which(subdf$val >= 21), ]
+    T_off_high <- subdf.high[median(which(diff(subdf.high$TPM) == min(diff(subdf.high$TPM)))), val]
+    Ar.int.T_off[which(Ar.int.T_off$Transcript == i), "T_off_high"] <- T_off_high
+
+    subdf.low <- subdf[which(subdf$val <= 21), ]
+    T_off_low <- subdf.low[median(which(diff(subdf.low$TPM) == min(diff(subdf.low$TPM)))), val]
+    Ar.int.T_off[which(Ar.int.T_off$Transcript == i), "T_off_low"] <- T_off_low
+}
+
+# compare T_off among species
+(T_off.int.high.ttest <- t.test(Ar.int.T_off$T_off_high, A22.int.T_off$T_off_high))
+
+(T_off.int.low.ttest <- t.test(Ar.int.T_off$T_off_low, A22.int.T_off$T_off_low))
+
+```
+
+As with genes turned on at high temperatures, *A. carolinensis* **Intermediate** genes are down-regulated on average at 1.2C higher temperatures than *A. picea* **Intermediate** genes.
 
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, md, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.union, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, md, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.union, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
 ```

### 2014-09-24 ApTranscriptome_TR.Rmd
updates to script

@@
 rownames(tt3) <- c("AcNC", "", "", "", "", "", "")
-colnames(tt3) <- c("ApVT", "", "", "", "", "", "")
+colnames(tt3) <- c("**ApVT**", "", "", "", "", "", "")
 
@@
 (T_off.int.low.ttest <- t.test(Ar.int.T_off$T_off_low, A22.int.T_off$T_off_low))
-
 ```
@@
 # table results
-Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 10)
+Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 132)
 Ap.BP.ResTable
@@
 
+GSEA for genes that are *Intermediate* in A. carolinensis and *Low* or *Bimodal* in A. picea
+
+```{r gsea_Ar_int_A22_lowbim, results = 'hide'}
+# genes with *Intermediate* expression in Ar and *Low* in A22
+Ar.int.A22.lowbim <- c(Ar.int.A22.low, Ar.int.A22.bim)
+Ar.int.A22.lowbim.geneList <- rep(0, length = length(Ap.geneList))
+names(Ar.int.A22.lowbim.geneList) <- names(Ap.geneList)
+Ar.int.A22.lowbim.geneList[(which(names(Ar.int.A22.lowbim.geneList) %in% Ar.int.A22.lowbim))] <- 1
+# check correct number of values set to 1
+table(Ar.int.A22.lowbim.geneList)
+# run GSEA
+Ar.int.A22.lowbim.gsea <- gsea(genelist = Ar.int.A22.lowbim.geneList, geneID2GO = geneID2GO, plotpath = NA)
+```
+
+```{r gsea_Ar_int_A22_lowbim_cluster}
+# similarity among enriched GO terms
+Ar.int.A22.lowbim.gsea.term.sim <- termSim(Ar.int.A22.lowbim.gsea$GO.ID, Ar.int.A22.lowbim.gsea$GO.ID, ont = "BP", organism = 'fly')
+# replace GO IDs with terms
+rownames(Ar.int.A22.lowbim.gsea.term.sim) <- Ar.int.A22.lowbim.gsea$Term
+colnames(Ar.int.A22.lowbim.gsea.term.sim) <- Ar.int.A22.lowbim.gsea$Term
+# distance matrix
+Ar.int.A22.lowbim.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.lowbim.gsea.term.sim))
+plot(Ar.int.A22.lowbim.gsea.term.sim.hclust)
+# report items in each cluster at height 1
+(Ar.int.A22.lowbim.hclust.report <- GSEAReportClusters(Ar.int.A22.lowbim.gsea.term.sim.hclust, h = 1.2))
+```
+
 ### GSEA for each functional type in each species
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, md, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.union, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, file = "doc/ms.RData")
 ```

### 2014-09-24 doc/ApTranscriptome_manuscript.Rmd
incorporating SCH revisions to Results and Discussion

@@
 
-To gain insight on the genetic basis of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions.
-First, species from more thermally-stressful environments may rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
-Finding that fewer genes are up-regulated to high temperatures in the warmer-climate species *A.
-carolinensis* would support this prediction.
-Second, spatially-divergent local adaptation leaves a signal at the molecular level [@savolainen2013].
-For gene expression, this local adaptation may be manifested through changes in the sensitivity of responsive genes to changes in temperature.
-Evidence for local adaptation in gene expression would be observed if heat-responsive genes are up-regulated at higher temperatures in *A.
-carolinensis* than in *A.
-picea*, and conversely, that cold-responsive genes are activated at colder temperatures in *A.
-picea*.
-
-In addition, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature.
+To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions.
+First, species from more thermally-stressful environments will rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
+This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature.
+Finding that fewer genes are thermally responsive in the warmer-climate species *A.
+carolinensis* would support this prediction would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A.
+carolinensis* would support the latter.
+Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature.
 Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses.
@@
 picea*, and vice versa, would be consistent with this prediction.
-While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis.
+While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes.
 For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].
-To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
+To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
 
@@
 options(scipen = 1, digits = 3)
-load("../ms.RData")
+load("ms_20140924.RData")
 library(data.table)
@@
 ```
+
+## Species differences in the number of thermally-responsive genes
 
@@
 
-## Thermal responses involve greater down-regulation in the warmer-climate species
-
-For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
+We found that the molecular diversity of thermal response to cold temperatures is substantially greater than of that to hot temperatures.
+In particular, we grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
-In both species, the majority of genes were in the **Low** category (Fig.
-1).
+In both species,  the majority of genes were in the **Low** category (Table 1).
 For *A.
 carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A.
-picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig.
-1).
+picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Table 1).
+
+Gene set enrichment analysis (GSEA) of the responsive genes indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.
+Specifically, out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** genes.
+These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis".
+For *High* genes, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".
+For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process".
+For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".
+
+# Shifts in the expression patterns of thermally-responsive genes between species
+
 Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
@@
 carolinensis*), most of these were genes that changed from one expression category to another between the species (i.e.
-the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig.
+the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).
+
+In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A.
+carolinensis*, the total number of thermally-responsive genes did not differ between the species (...`r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])`...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ...
+).
+Specifically, compared to a hypothetical homogenous distribution, there were an excess of **Intermediate** genes in *A.
+carolinensis* which disproportionatly switched to **Low** or **Bimodal** expression categories in *A.
+picea* (Fig.
 1).
 
-Specifically, all the diagonals (e.g.
-**Low** -- **Low**) the *A.
-carolinensis* **Bimodal** -- *A.
-picea* **Not responsive** and *A.
-carolinensis* **Not responsive** -- *A.
-picea* **High** categories were over-represented compared to a hypothetical homogenous distribution for the contingency table.
-
-In contrast to expectations, more genes did not respond to high temperatures in *A.
-carolinensis*.
-In fact, the only differences were for **Intermediate** genes in *A.
-carolinensis*, which disproportionatly switched to **Low** or **Bimodal** expression in *A.
-picea*.
-Thus, *A.
-carolinensis*, the warmer-climate species, appears to rely upon shutting down gene expression to a greater extent than *A.
-picea*, which activates gene expression...
+The molecular processes in this set of genes that switched from down-regulation in *A.
+carolinensis* to up-regulation in *A.
+picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ...
+were related to metabolism.
+In addition, there were GO terms related to repair (e.g.
+"histone modification"), gene regulation (e.g.
+"positive regulation of cellular process") and cellular organization (e.g.
+"transport", "organelle organization").
+These results indicate that in response to thermal extremes, the warmer-climate species, *A.
+carolinensis*, shuts down metabolism while the colder-climate species, *A.
+picea*, increases metabolism.
 
@@
 
-## Greater inertia of gene expression response to heat stress in the warmer climate species
+## Species differences in the sensitivity of thermally-responsive genes to changes in temperature
 
@@
 
-We looked for evidence of genetic assimilation (the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed) by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
+We looked for evidence of genetic assimilation, the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed, by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
 carolinensis* or high temperatures in *A.
 picea*.
-While finding that genes upregulated to thermal extremes
+Contrary to expectations, we found that gene expression at optimal temperatures (about 19.5°C) was always greater in *A.
+carolinensis* than in *A.
+picea*.
+This pattern was true for genes that had *Low** expression in *A.
+carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) **High** expression in *A.
+picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).
 
-Contrary to expectations, for **Low** genes in *A.
-carolinensis* gene expression at intermediate temperatures was lower in *A.
-picea* than *A.
-carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) .
-Moreover, for **High** genes in *A.
-picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A.
-carolinensis*.
+In addition, we examined the magnitude of upregulation...
+...overall mean level of expression...
 
-## Gene set enrichment analysis
-
-To describe the molecular toolkit underlying *Aphaenogaster* response to thermal extremes, we used gene set enrichment analysis to identify gene ontology (GO) categories enriched for each expression category separately.
-First, we looked at the genes shared in each expression category (e.g.
-the intersection) for each species (Table S1).
-Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for **High** genes.
-Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".
-For **Low** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched.
-These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis".
-For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched.
-These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process".
-For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched.
-These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".
-Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.
-
-We further used gene set enrichment analysis for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A.
-carolinensis* to having increased expression in **Low** temperatures in *A.
-picea*.
-This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization".
-For **Intermediate** genes in *A.
-carolinensis* that shifted **Bimodal** in *A.
-picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification".
-These results indicate that in response to thermal extremes, the warmer-climate species, *A.
-carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A.
-picea*, which increases metabolism.
-
-**Table** Table of enriched GO terms for each expression category in each species.
+Further, we compared the mean expression of genes that were not thermally-responsive in one species to the expression level near optimum temperatures in the other species.
+Finding that mean expression of the non-responsive genes than the optimum genes would be indicative of adaptative basal expression....
 
@@
 Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
+In addition, of the ...
+heat shock proteins in the transcriptome, ...
+were upregulated to cold or hot temperatures in *A.
+picea* while only ...
+were upregulated in *A.
+carolinensis*.
 
@@
 By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
-Specifically, we found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.
-This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have variable expression in response to copper exposure [@hodgins-davis2012].
-In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014].
+We found that ...% of genes have non-linear responses to temperature.
+
+This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012].
+In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014] which only used a few temperature conditions.
 Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.
@@
 Thus, *A.
-carolinensis*, the more southerly species, shuts-down metabolism to a lesser extent than *A.
+carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A.
 picea* and so must rely on other mechanisms such as behavior (e.g.
-retreat into its nest) or proteome stability to tolerate thermal stress.
+retreat into its nest) or increased proteome stability to tolerate thermal stress.
 In contrast, *A.
@@
 carolinensis*.
-Our results are at least congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
-...Sara - something about enzyme efficiency tradeoffs???
+
+Our results are at congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
+Although the heat-shock response is one of the most conserved across living organisms (ref), it is energetically expensive, particularly under chronic stress conditions (ref).
+Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher basal level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al.
+2001).
+Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.
+Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.
 
@@
 picea* and *A.
-carolinensis*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
+carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
 

### 2014-09-25 ApTranscriptome_TR.Rmd
updated TR

@@
 library(RColorBrewer)
+library(coin)
 
 # Load Bioconductor libraries
-#source("http://bioconductor.org/biocLite.R")
+source("http://bioconductor.org/biocLite.R")
 #biocLite("topGO")
@@
 rm(other.lms)
+
+quadratic.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "2") == TRUE)]
 ```
@@
 ```{r mh_test}
-mh.test <- mh_test(tt2)
+mh.test <- mh_test(as.table(tt2))
 
@@
 dsum <- sum(abs(d1), abs(d2), abs(d3), abs(d4), abs(d5))
-d1/dsum
-d2/dsum
-d3/dsum
-d4/dsum
-d5/dsum
+abs(d1)/dsum
+abs(d2)/dsum
+abs(d3)/dsum
+abs(d4)/dsum
+abs(d5)/dsum
 
@@
 
-qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
+mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
   scale_fill_gradient2(limits=c(-600, 600)) 
+mh_plot
 ```
@@
 
-## Species comparison
+An interesting observation from the matched observations plot (Fig.
+...) is that for **NotResp** genes in *A.
+carolinensis* the **High** and **Bimodal** categories in *A.
+picea* are over-represented.
+Finding that the mean expression level of these genes in *A.
+carolinensis* is greater than the expression level at the optimum temperature (19.5°C) in *A.
+picea* would be consistent with genetic assimilation.
+
+```{r gen_assimilation_high}
+Ap.high.Ac.nr <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type == "NotResp"), ]
+
+# get mean expression level for Ar transcripts
+Ap.high.Ac.nr.TPM <- TPM.dt.sub[Ap.high.Ac.nr$Transcript]
+Ap.high.Ac.nr.mean.exp.NR <- ddply(Ap.high.Ac.nr.TPM, .(Transcript), summarize, meanTPM = mean(TPM))
+
+# get expression at optimum for A22 transcripts
+
+Ap.high.Ac.nr.opt.exp <- Ap.high.Ac.nr$A22.opt
+
+t.test(Ap.high.Ac.nr.mean.exp.NR$meanTPM, Ap.high.Ac.nr.opt.exp)
+```
+
+Contrary to expectations, mean expression of the non-responsive *A. carolinensis* genes is less than expression of the **High** genes in *A. picea*.
+
+Next, I check the same for the **Bimodal** genes in *A. picea*.
+
+```{r gen_assimilation_bim}
+Ap.bim.Ac.nr <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "NotResp"), ]
+
+# get mean expression level for Ar transcripts
+Ap.bim.Ac.nr.TPM <- TPM.dt.sub[Ap.bim.Ac.nr$Transcript]
+Ap.bim.Ac.nr.mean.exp.NR <- ddply(Ap.bim.Ac.nr.TPM, .(Transcript), summarize, meanTPM = mean(TPM))
+
+# get expression at optimum for A22 transcripts
+Ap.bim.Ac.nr.opt.exp <- Ap.bim.Ac.nr$A22.opt
+
+t.test(Ap.bim.Ac.nr.mean.exp.NR$meanTPM, Ap.bim.Ac.nr.opt.exp)
+```
+
+As with the **High** transcripts, expression much higher for the responsive *A. picea* transcripts than for the non-responsive *A. carolinensis* transcripts.
+
+## Species comparisons
 
@@
 
-Of the responsive transcripts, `r prettyNum(length(which(responsive.lms.ann.type$best.hit.to.nr == "-"))/length(responsive.lms.ann.type$best.hit.to.nr)*100)`% are annotated, while `r prettyNum(length(which(annotation.table$best.hit.to.nr == "-"))/nrow(annotation.table) * 100)`% of all transcripts are annotated.
+Of the responsive transcripts, `r round(length(which(responsive.lms.ann.type$best.hit.to.nr != "-"))/length(responsive.lms.ann.type$best.hit.to.nr)*100, 0)`% are annotated, while `r round(length(which(annotation.table$best.hit.to.nr != "-"))/nrow(annotation.table) * 100, 0)`% of all transcripts are annotated.
 
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, file = "doc/ms.RData")
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
 ```

### 2014-09-25 doc/ApTranscriptome_manuscript.Rmd
updates to ms

@@
 ---
-title: Investigation of thermal reactionomes reveals stress tolerance vs. resistance in
-  warm and cold-climate ant species
-author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
-  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
-  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
-  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
-  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
-  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
-  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
-  Evolution and Climate, University of Copenhagen}
+title: Investigation of thermal reactionomes reveals differing responses to thermal
+  extremes in warm and cold-climate ant species
+author: "null"
 csl: ecology.csl
 output:
+  word_document: default
   pdf_document:
@@
     latex_engine: xelatex
-    number_sections: yes
+    number_sections: no
     toc_depth: 2
   html_document: null
-  word_document: default
 includes: header.tex
@@
 \linenumbers
+
+\newpage
+
+**Authors**
+
+John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2},
+Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4},
+Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}
+
+\vspace{10 mm}
+
+**Affilitiations**
+
+\textsuperscript{1} Department of Biology, University of Vermont \newline
+\textsuperscript{2} Vermont Genetics Network, University of Vermont \newline
+\textsuperscript{3} Department of Biological Sciences, North Carolina State University \newline
+\textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen
 
@@
 carolinensis* would support the latter.
+
 Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature.
@@
 For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].
+
 To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
@@
 ```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
-options(scipen = 1, digits = 3)
-load("ms_20140924.RData")
+#options(scipen = 1, digits = 3)
+load("ms_2014-09-25.RData")
 library(data.table)
@@
 Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate.
-Of these, xxx had a non-linear relationship with temperature that would not have been detected with a standard differential expression approach.
-At a minimum `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive.
-~~However, the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012].
-Assuming that the number of genes is similar for these two *Aphaenogaster* species, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.~~
+Of these, `prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
+In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive.
+This result is the same if we focus on only the transcripts that have a BLAST hit.
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts (i.e.
+the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]) indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.
 
-```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type.
-**Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.
-The width of the columns is proportional to the number of genes for each expression type for *A.
-carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A.
-picea.* The rows are colored by expression type for *A.
+```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive genes by expression type for *A.
+carolinensis* and *A.
 picea*.
-**Low** The total number of genes for two categories (*A.
-picea* **Low** - *A.
-carolinensis* **Low**; *A.
-picea* **NotResp** - *A.
-carolinensis* **Bimodal**) is labeled in two boxes for perspective."}
-mosaic.colors <- brewer.pal(5, "Blues")
-par(mar = c(5.1, 3.1, 2.1, 7.5))
-mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
-labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
-text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
-text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
-text(.15, .7, "2,594")
-text(.85, .03, "223", cex = .75)
-legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
+**Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other."}
+
+rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")
+
+pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
+             big.mark = ",", split.tables = Inf, 
+             caption = "Table of the number of thermally-responsive genes by expression type for *A.
+carolinensis* and *A.
+picea*.
+**Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.")
 ```
@@
 In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A.
-carolinensis*, the total number of thermally-responsive genes did not differ between the species (...`r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])`...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ...
+carolinensis*, the total number of thermally-responsive genes did not differ between the species (...r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ...
 ).
@@
 
-```{r expression_change, echo=FALSE}
-cont.tab <- structure(unlist(as.vector(tt2)), .Dim = c(5L, 5L), .Dimnames = structure(list(
-    c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", 
-    "Intermediate", "High", "Bimodal", "NotResp")), .Names = c("Ac", 
-"Ap")))
+```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A.
+carolinensis* (rows) and *A.
+picea* (columns).
+The color of each cell represent the deviation of the observed from the expected number of genes based on hypothetical equivalence of the marginal frequencies.
+Blue cells are those with significant excesses, while red cells are those with significant absences."}
+#tt2
 
 # marginal sums and total
-rs <- rowSums(cont.tab)
-cs <- colSums(cont.tab)
-n <- sum(cont.tab)
+rs <- rowSums(tt2)
+cs <- colSums(tt2)
+n <- sum(tt2)
 
 # grand mean for each class
-gm <- (rs + cs) / sum(cont.tab * 2)
+gm <- (rs + cs) / sum(tt2 * 2)
 
 # calculate observed values for each cell using overall mean for each expression type
-Ec <- outer(gm, gm, "*") *sum(cont.tab)
+Ec <- outer(gm, gm, "*") *sum(tt2)
 
 # get deviations of observed from expected
-Ec.dev <- cont.tab - Ec
+Ec.dev <- tt2 - Ec
 
 # overall table P-value
-Ec.chistat <- sum((cont.tab-Ec)^2 / Ec)
+Ec.chistat <- sum((tt2-Ec)^2 / Ec)
 # P-value
-pchisq(Ec.chistat, df=(nrow(cont.tab)-1) * (ncol(cont.tab)-1), lower.tail = FALSE)
+pchisq(Ec.chistat, df=(nrow(tt2)-1) * (ncol(tt2)-1), lower.tail = FALSE)
 
 # overall minimum chi-squared value `X2.df` for a significantly non-homogenous distribution of counts
-df = (ncol(cont.tab) - 1) * (nrow(cont.tab) - 1)
+df = (ncol(tt2) - 1) * (nrow(tt2) - 1)
 X2.df <- qchisq(p = 0.05, df = df, lower.tail = FALSE)
@@
 # posthoc result for each cell
-Ec.cells <- sign(cont.tab - Ec) * (cont.tab - Ec)^2 / Ec
+Ec.cells <- sign(tt2 - Ec) * (tt2 - Ec)^2 / Ec
 Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
@@
 
+md <- melt(Ec.cells)
+
 # plot
-qplot(x=Ap, y=Ac, data=melt(Ec.cells), fill=value, geom="tile") + 
+mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
   scale_fill_gradient2(limits=c(-600, 600)) 
+mh_plot
 ```
@@
 
-We looked for evidence of genetic assimilation, the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed, by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
+We looked for evidence of genetic assimilation, the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed, in multiple comparisons.
+First, we predicted that gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
 carolinensis* or high temperatures in *A.
-picea*.
-Contrary to expectations, we found that gene expression at optimal temperatures (about 19.5°C) was always greater in *A.
+picea* should be higher in the opposite species (e.g.
+greater basal expression but lower inducibility).
+Instead, we found that gene expression at optimal temperatures (about 19.5°C) was always greater in *A.
 carolinensis* than in *A.
 picea*.
-This pattern was true for genes that had *Low** expression in *A.
+This pattern was true for genes that had **Low** expression in *A.
 carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) **High** expression in *A.
@@
 
-In addition, we examined the magnitude of upregulation...
-...overall mean level of expression...
-
-Further, we compared the mean expression of genes that were not thermally-responsive in one species to the expression level near optimum temperatures in the other species.
-Finding that mean expression of the non-responsive genes than the optimum genes would be indicative of adaptative basal expression....
+Second, for the genes that had **High** or **Bimodal** expression in *A.
+picea* but were not thermally-responsive in *A.
+carolinensis* (blue cells in bottom row of Fig.
+1), we predicted that mean expression level of the non-responsive *A.
+carolinensis* genes would be greater than the expression level near the optimum temperature (19.5°C) of the responsive *A.
+picea* genes.
+Again, we found no support for genetic assimilation as the non-responsive genes had lower mean expression levels.
 
@@
 Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
-In addition, of the ...
-heat shock proteins in the transcriptome, ...
-were upregulated to cold or hot temperatures in *A.
-picea* while only ...
-were upregulated in *A.
+In addition, of the `r nrow(hsp_all)` heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
+picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
 carolinensis*.
@@
 By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
-We found that ...% of genes have non-linear responses to temperature.
+We found that about 70% of the thermally-responsive genes (7,202) have non-linear responses to temperature.
 
@@
 carolinensis* and more up-regulated in *A.
-picea* in response to thermal extremes (Fig.
-3).
+picea* in response to thermal extremes (Table 1, Fig.
+1).
 The results of the gene set enrichment analysis give an indication of what might drive this difference.

### 2014-10-01 doc/ApTranscriptome_manuscript.Rmd
updates to Intro with comments from SHC, NS, RD

@@
 
-Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change.
-Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the genetic basis of thermal tolerance.
-We characterized the reactionomes of two common ant species in eastern U.S.
-forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
+Thermal tolerance is a key trait governing the distributions of species and is likely useful in predicting their future responses to climate change.
+Here, we describe the thermal reactionome, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient, to identify genes underlying thermal tolerance.
+We characterize the reactionomes of two common ant species in the eastern U.S: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
 We found that at least 10% of genes were thermally responsive.
 The warm-adapted *A.
-carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression chang than did *A.
-picea*.
+carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change than did *A.
+picea* in response to warming.
 This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability.
-*A.
-carolinensis* also shut down more genes in response to thermal extremes than *A.
-picea*.
-A subset of these genes were upregulated in *A.
+*Aphaenogaster carolinensis* also shut down more genes in response to thermal extremes than did *A.
+picea*, while a subset of these same genes were upregulated in *A.
 picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair.
-These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress.
-Namely *A.
-picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A.
+These differences indicate that the two closely related species employ different strategies for coping with temperature stress.
+The northern *A.
+picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas the southern *A.
 carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress.
-More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
+These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
 
@@
 
-Successfully coping with progressive changes in climate will require both short-term physiological acclimation and longer-term adaptive changes.
+Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptive changes.
 To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
 carolinensis*.
-We found surprisingly divergent transcriptomic responses in the two species.
-*A.
-picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a more robust heat shock response.
+Surprisingly, we found divergent transcriptomic responses in these species.
+*Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response.
 In contrast, *A.
-carolinensis* reacted to thermal stress to a greater extent through downregulation, reducing metabolic gene expression and requiring more extreme temperatures to trigger a protective response.
-These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatic zones.
+carolinensis* required more extreme temperatures to trigger a protective response and downregulated genes related to metabolism.
+These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatices.
 
@@
 Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009].
-Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
+Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature (sunday, kingsolve, deutch, hoffman).
+Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage (...refs...).
+Thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
 These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved.
-To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.
-This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2010].
+Outside of candidate genes [@krebs1998], little is known about the molecular basis of such variation or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes.
+To address this gap of knowlege, we need information on the diversity of genes, and their plasticity of expression, involved in thermal tolerance [@hoffman2008; @somero2011].
 
 In recent years, whole organism gene expression approaches (e.g.
-transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits.
-These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few environmental levels.
+transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits.
+Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes.
 However, temperature and other environmental factors vary continuously in nature.
-As a result, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014].
-Similarly, gene expression studies that look at just a few environments (e.g.
+As a result, many patterns of biologically-relevant plasticity are likely to be missed in studies that compare only a few levels of an environmental variable [@murren2014].
+Similarly, species comparisons look at just a few environments (e.g.
 hot vs.
 cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.
-What is needed is a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992].
+Such differences in degree are better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992].
 Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].
 
-Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, an important suite of issues often hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014].
-Namely, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005].
-Genome-wide association studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013].
-Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues.
-
-Here, we present a novel method for RNAseq analysis: the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
+Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
 We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
 picea* [@umphrey1996].
-Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].
+*Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
+*Aphaenogaster picea* are found ...
+while *A.
+carolinensis* occupy habitats with ...
+.
+Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).
 
-To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions.
-First, species from more thermally-stressful environments will rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
+To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species to test two specific predictions.
+First, we predicted that species from more thermally-stressful environments will rely to a greater extent than species from cooler environments upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
 This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature.
@@
 carolinensis* would support the latter.
-
-Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature.
-Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses.
+Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g.
+genetic assimilation [@pigliucci2006]).
 Finding that the genes up-regulated in response to cold temperatures in *A.
@@
 picea*, and vice versa, would be consistent with this prediction.
-While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes.
+
+While neither of these predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes.
 For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].
-
-To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
 
@@
 #options(scipen = 1, digits = 3)
-load("ms_2014-09-25.RData")
+load("ms_2014-09-26.RData")
 library(data.table)

### 2014-10-01 doc/ApTranscriptome_manuscript.Rmd
updates to ms

--- 2014-09-25 doc/ApTranscriptome_manuscript.Rmd
@@
   extremes in warm and cold-climate ant species
-author: "null"
+author: 'null'
 csl: ecology.csl
@@
 John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2},
-Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4},
-Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}
+Mahesh Vangala \textsuperscript{2}, Robert R.
+Dunn \textsuperscript{3}, Aaron M.
+Ellison \textsuperscript{4}, Nathan J.
+Sanders \textsuperscript{5}, Nicholas J.
+Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}
 
@@
 
-\textsuperscript{1} Department of Biology, University of Vermont \newline
-\textsuperscript{2} Vermont Genetics Network, University of Vermont \newline
-\textsuperscript{3} Department of Biological Sciences, North Carolina State University \newline
-\textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen
+\textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
+\textsuperscript{2} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
+\textsuperscript{3} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
+\textsuperscript{4} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
+\textsuperscript{5} Center for Macroecology, Evolution and Climate, University of Copenhagen
 
@@
 Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change.
-Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the genetic basis of thermal tolerance.
+However, little is known about the genetic basis of thermal tolerance in most species.
+Here, we describe the thermal “reactionome”, the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient, to understand the genetic basis of thermal tolerance.
 We characterized the reactionomes of two common ant species in eastern U.S.
@@
 The warm-adapted *A.
-carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression chang than did *A.
+carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change than did *A.
 picea*.
@@
 carolinensis* reacted to thermal stress to a greater extent through downregulation, reducing metabolic gene expression and requiring more extreme temperatures to trigger a protective response.
-These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatic zones.
+These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of the northern species in the face of rapidly shifting climatic zones.
 
@@
 To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.
-This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2010].
+This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2011].
 
@@
 
-Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, an important suite of issues often hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014].
+Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, a suite of issues hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014].
 Namely, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005].
@@
 picea*, and vice versa, would be consistent with this prediction.
+C
+
 While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes.
 For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].
-
 To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
@@
 The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts.
-Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate.
-Of these, `prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
-In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive.
-This result is the same if we focus on only the transcripts that have a BLAST hit.
+Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes), `r round(9100/99861, 2)*100`% of the transcriptome, with expression patterns that were thermally-responsive at 5% False Discovery Rate.
+Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
 However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts (i.e.
@@
              big.mark = ",", split.tables = Inf, 
-             caption = "Table of the number of thermally-responsive genes by expression type for *A.
+             caption = "Number of thermally-responsive genes by expression type for *A.
 carolinensis* and *A.
@@
 
-We found that the molecular diversity of thermal response to cold temperatures is substantially greater than of that to hot temperatures.
+We found that the number of genes that are responsive to cold temperatures is substantially greater than of that to hot temperatures (Table 1).
 In particular, we grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
@@
 
-Gene set enrichment analysis (GSEA) of the responsive genes indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.
+Gene set enrichment analysis (GSEA) of the responsive genes indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes.
 Specifically, out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** genes.
@@
 
-# Shifts in the expression patterns of thermally-responsive genes between species
+## Shifts in the expression patterns of thermally-responsive genes between species
 
@@
 
+```{r, echo=FALSE}
+Ap.t.test <- t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])
+```
+
 In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A.
-carolinensis*, the total number of thermally-responsive genes did not differ between the species (...r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ...
+carolinensis*, the total number of thermally-responsive genes did not differ between the species (*t*~df~=`r round(Ap.t.test$parameter, 0)` = `r round(Ap.t.test$statistic, 2)`, *P* = `r round(Ap.t.test$p.value, 2)`), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ...
 ).
@@
 "transport", "organelle organization").
-These results indicate that in response to thermal extremes, the warmer-climate species, *A.
-carolinensis*, shuts down metabolism while the colder-climate species, *A.
-picea*, increases metabolism.
+These results suggest that in response to thermal extremes, the warmer-climate species *A.
+carolinensis* shuts down metabolism while the colder-climate species *A.
+picea* increases metabolism, potentially to support cellular resistance to thermal stress.
 
@@
                                                             ifelse(x > X2.df, "+", "-")))
-
 md <- melt(Ec.cells)
@@
 carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
-picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+picea*; *t*~df~=`r round(T_on.low.ttest$parameter, 0)` = `r round(T_on.low.ttest$statistic, 1)`, *P* = `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
 2).
@@
 
-By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
-We found that about 70% of the thermally-responsive genes (7,202) have non-linear responses to temperature.
+Species persistence in face of climate change will require short-term physiological acclimation as well as longer-term adaptive changes.
+In this study, we explored the relationship between these processes using whole organism gene expression data from closely-related ant species.
+By applying a regression-approach to gene expression data, we were able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
+In the combined transcriptome of these species, we found that a minimum of 9% of genes are thermally-responsive, and that about 70% of these genes (7,202) have non-linear responses to temperature that would not have been detected by standard differential expression approaches.
 
-This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012].
+The result that a large fraction of the genome is responsive is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012].
 In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014] which only used a few temperature conditions.
@@
 
-```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
-# reshape data
-Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")
-
-maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=Species)) + 
-  geom_density(alpha=0.2, position="identity") + 
-  labs(x = "Temperature of maximum expression", y = "Density")
-
-print(maxexplot)
-
-####################################
-### CHANGE TO REACTION NORM FOR HSPs 
-####################################
-
-```
-
 Given the differences in the number of thermally-responsive genes between the two species (Fig.
@@
 carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.
-
-Thermal safety margins...
 

### 2014-10-01 ApTranscriptome_TR.Rmd

@@
 # table results
-Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 132)
+Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 131)
 Ap.BP.ResTable
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
 ```

### 2014-10-01 doc/ApTranscriptome_manuscript.Rmd
tweaks

@@
   extremes in warm and cold-climate ant species
-author: 'null'
+author: ' '
 csl: ecology.csl
 output:
-  word_document: default
   pdf_document:

### 2014-10-03 doc/PNAS_cover_letter.tex
added draft PNAS cover letter

\documentclass[letterpaper]{article}

\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{color} % color!
\usepackage[hidelinks]{hyperref} % allow html links, hide defaul link box

\usepackage[textheight=8in]{geometry} % reduce textbox height to fit header (next section) and change page margins to 1 in

\usepackage{parskip} % change format to blank line between paragraphs instead of tab

\usepackage{fancyhdr}

\fancypagestyle{first}{%
  \fancyhf{}
  \setlength{\headheight}{40pt}
  \fancyhead[L]{\includegraphics[scale=0.1]{UVM_Official_Logo.jpg}} 
  \fancyhead[R]{\small Department of Biology \\102  Marsh Life Sciences \\ Burlington, VT 05405 }
  \renewcommand{\headrulewidth}{0.4pt}
  \renewcommand{\footrulewidth}{0pt}}
\pagestyle{first}

\fancypagestyle{plain}{%
  \fancyhf{}
  \setlength{\headheight}{0pt}
  \fancyhead[L]{ } 
  \fancyhead[R]{\small John Stanton-Geddes} 
  \fancyfoot[C]{\thepage}
  \renewcommand{\headrulewidth}{0.4pt}
  \renewcommand{\footrulewidth}{0pt}}
\pagestyle{plain}

\begin{document}

\thispagestyle{first}

\noindent{Dear \emph{PNAS} Editorial Board,}

We are excited to submit our manuscript \emph{Investigation of thermal
reactionomes reveals differing responses to thermal extremes in warm and
cold-climate ant species} for consideration as a Research Report in
\emph{The Proceedings of the National Academy of Sciences}. There are
several aspects of the work that we think make it worthy of publication
in \emph{PNAS}.

First, we make a methodological contribution by extending the reaction
norm approach to the study of gene expression. We define the
\emph{reactionome} as a characterization of the reaction norm for all
genes in an organism's transcriptome across an environmental gradient. 
We believe that this approach is an important contribution as it allows us
to make quantitative comparisons of plasticity in gene expression between
species. Importantly, this analysis does not depend on the functional 
annotation of the genes, and thus is robust to criticisms of "storytelling" 
in RNAseq studies based on annotation of differentially expressed 
genes (Pavlidis et al. 2012 MBE).

Second, by applying the reactionome approach to two related ant species,
\emph{Aphaenogaster picea} and \emph{A. carolinensis}, from colder and
warmer climates, respectively, we gain insight on the relationship
between short-term physiological acclimation and longer-term adaptive
change in response to thermal extremes. In particular, we find that
adaptation to chronic high temperatures involves different mechanisms
than those used to cope with rarely encountered stressors, which may
limit adaptive potential of northern species in the face of rapidly
shifting climatices.

We believe that the data, including the first transcriptome of a
temperate ant species, analyses and results will be of interest to
multiple audiences including ecologists working on species
distributions, physiologists interested in thermal tolerance and
evolutionary geneticists interested in the evolution of gene expression.

This manuscript is most appropriate for the major classification
BIOLOGICAL SCIENCES, minor classification Ecology.

\textbf{Suggested Editors:} Douglas J. Futuyma, David L. Denlinger, David M. Hillis

\textbf{Suggested NAS Members:} James H. Brown, Bert Hoelldobler, Trudy F. Mackay, Nevo Eviata, Johanna Schmitt

\textbf{Suggested reviewers:}

\begin{itemize}
  \item{Dr. Nicholas Teets, Postdoc, University of Florida}
  \item{Dr. Vanessa Kellerman, Postdoc, Monash University}
  \item{Dr. Pavlos Pavlidis, Postdoc, Heidelberg Institute of Theoretical Studies}
  \item{Dr. Stephen Palumbi, Professor, Hopkins Marine Station, Stanford University}
  \item{Dr. Joel Kingsolver, Professor, University of North Carolina}
\end{itemize}

\noindent{Thank you for consideration of our work,} \\
\includegraphics[scale=1]{JSG_signature.png}  \\
Dr. John Stanton-Geddes

\end{document}

### 2014-10-03 doc/ApTranscriptome_manuscript.Rmd
updates to Intro from SHC

@@
 ---
-title: Investigation of thermal reactionomes reveals differing responses to thermal
+title: Thermal reactionomes reveal adaptive responses to thermal
   extremes in warm and cold-climate ant species
@@
 Thermal tolerance is a key trait governing the distributions of species and is likely useful in predicting their future responses to climate change.
-Here, we describe the thermal reactionome, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient, to identify genes underlying thermal tolerance.
+To understand how plasticity of gene expression underlies thermal tolerance, we describe the *thermal reactionome*, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient.
 We characterize the reactionomes of two common ant species in the eastern U.S: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
@@
 The warm-adapted *A.
-carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change than did *A.
-picea* in response to warming.
-This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability.
-*Aphaenogaster carolinensis* also shut down more genes in response to thermal extremes than did *A.
-picea*, while a subset of these same genes were upregulated in *A.
-picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair.
+carolinensis* displayed higher mean levels of expression for responsive genes, exhibited greater inertia of gene expression change and upregulated fewer heat shock proteins than did *A.
+picea* in response to thermal extremes.
+This pattern is consistent with hypotheses of constitutive thermal defense and enhanced proteome stability.
+*Aphaenogaster carolinensis* also down-regulated expression of more genes in response to thermal extremes than did *A.
+picea*.
+A subset of these same genes were upregulated in *A.
+picea*, including genes mapped to ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair.
 These differences indicate that the two closely related species employ different strategies for coping with temperature stress.
@@
 Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009].
-Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature (sunday, kingsolve, deutch, hoffman).
+Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014].
 Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage (...refs...).
-Thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
+Thermal defense strategies are shaped by the environmental regime organisms experience (...ref...), and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
 These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-Outside of candidate genes [@krebs1998], little is known about the molecular basis of such variation or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes.
-To address this gap of knowlege, we need information on the diversity of genes, and their plasticity of expression, involved in thermal tolerance [@hoffman2008; @somero2011].
+Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes.
+This information is critical for understanding the adaptive potential of species to future climates [@hoffman2008].
 
+To address this gap of knowlege, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011].
 In recent years, whole organism gene expression approaches (e.g.
@@
 However, temperature and other environmental factors vary continuously in nature.
-As a result, many patterns of biologically-relevant plasticity are likely to be missed in studies that compare only a few levels of an environmental variable [@murren2014].
-Similarly, species comparisons look at just a few environments (e.g.
+As a result, such comparisons (e.g.
 hot vs.
-cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.
-Such differences in degree are better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992].
-Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].
+cold) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is.
+Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992].
+Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].
 
@@
 *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-*Aphaenogaster picea* are found ...
-while *A.
-carolinensis* occupy habitats with ...
-.
+*Aphaenogaster carolinensis* experiences significantly warmer and less seasonally variable temperatures than *A.
+picea* (...data...), with correspondingly higher minimal and maximal critical thermal limits (...CTmin and CTmax from HF/DF...).
 Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).
 
-To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species to test two specific predictions.
-First, we predicted that species from more thermally-stressful environments will rely to a greater extent than species from cooler environments upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013].
+To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material).
+We used these data to test two specific predictions.
+First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms relative to species from cooler environments.
+[@barshis2013].
 This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature.
 Finding that fewer genes are thermally responsive in the warmer-climate species *A.
-carolinensis* would support this prediction would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A.
+carolinensis* would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A.
 carolinensis* would support the latter.
@@
 picea*, and vice versa, would be consistent with this prediction.
-
-While neither of these predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes.
-For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].
 
@@
 library(topGO)
+
+options(digits = 2)
+panderOptions('round', 2)
+panderOptions('keep.trailing.zeros', TRUE)
 ```
@@
 The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts.
-Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate.
-Of these, `prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
+Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995].
+Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
 In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive.
 This result is the same if we focus on only the transcripts that have a BLAST hit.
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts (i.e.
-the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]) indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.
 
@@
 
-We found that the molecular diversity of thermal response to cold temperatures is substantially greater than of that to hot temperatures.
-In particular, we grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
+The number of genes that were responsive to cold temperatures was substantially greater than of the number responsive to hot temperatures.
+We grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
@@
 Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
-While some of these were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes were responsive in *A.
+While some of these genes were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` in *A.
 carolinensis* but not *A.
-picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` were responsive in *A.
+picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` in *A.
 picea* but not *A.
@@
 3).
-Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).
+Although `r resp.stress.responsive.count` genes were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
 In addition, of the `r nrow(hsp_all)` heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
@@
 This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012].
-In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014] which only used a few temperature conditions.
+Finding that thousands of genes responded to temperature highlights the limitations of previous studies [@meyer2011; @barshis2013; @palumbi2014; @oneil2014] that used only a few temperature levels and found that only a few dozen to hundred genes are involved in thermal responses.
 Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.
@@
 carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A.
-picea* and so must rely on other mechanisms such as behavior (e.g.
-retreat into its nest) or increased proteome stability to tolerate thermal stress.
+picea* and so must rely on other mechanisms such as behavior or increased proteome stability to tolerate thermal stress.
 In contrast, *A.
@@
 
-Our results are at congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
+Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
 Although the heat-shock response is one of the most conserved across living organisms (ref), it is energetically expensive, particularly under chronic stress conditions (ref).
@@
 1), it is worth noting the similarity in expression patterns between species.
-There are 3-4 $\times$ more genes upregulated at cold than high temperatures in both species (Fig.
+In both species, there are 3-4 $\times$ more genes upregulated at cold than high temperatures  (Fig.
 3).
@@
 This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
-At least among *Drosophila* taxa, these limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012].
+At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012].
 In this light, for *A.
@@
 
-In contrast to expectations that adaptation occurs via changes from plastically-induced to constitutive expression (e.g.
+In contrast to the predictions of the genetic assimilation hypothesis that adaptation occurs via changes from plastically-induced to constitutive expression (e.g.
 genetic assimilation), we found lower basal gene expression in *A.
@@
 
-Thermal safety margins...
-
-As with all *de novo* transcriptomes, our work has some important caveats.
-First, given that all sequenced ant genomes to date have about 18,000 genes [@gadau2012], our transcriptome is highly fragmented.
+As with all *de novo* transcriptomes, our work has some caveats.
+First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented.
 Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts.
@@
 Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005].
-Large differences in patterns of gene expression may only have small effects on organismal fitness.
+Large differences in patterns of gene expression may have only small effects on organismal fitness.
 Given these caveats, transcriptomic studies promise to offer many new insights into how organisms respond and will respond to environmental change, especially for long-lived species that are intractable to study in the lab.
+
+An additional concern is differences in gene expression among species due to drift.
+While we do in fact expect some genes to differ in expession due to neutral processes, if this was the primary driver of differences among species we would expect to see equal distribution of counts in the off-diagonals of the matched observations (Fig.
+1).
+Instead we find that certain categories are over-represented, and that the differences in the reaction norm shapes are consistent with *a priori* expectations.
 
@@
 
-...ants...Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
+Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
+Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
 On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
-Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
 
@@
 
-From each species, we haphazardly collected 12 ants at the same time on 12 days.
-Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour.
+From each species, we haphazardly collected 12 ants from a single colony at the same time on 12 days.
+Each day, the 12 ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
 The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
@@
 We used $log(TPM + 1)$ as the response to control for skew in the expression data.
-For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005].
-This method is robust to over-dispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
+For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005].
+This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
 To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.
 
-We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+We identified thermally-responsive genes as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
 All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).

### 2014-10-03 ApTranscriptome_TR.Rmd
added some analyses for ms

@@
 
+Table with 'Total Transcripts' category
+
+```{r total_table}
+
+```
+
 The question of biological interest is whether the marginal frequencies differ between the two species.
@@
 
+Does the mean expression level for all responsive genes differ among colonies?
+
+```{r mean_exp}
+t.test(log(Ap.response.type$A22.opt), log(Ap.response.type$Ar.opt))
+# expression level at optimum temp greater in Ar than A22
+
+# non-parametric test
+w4 = wilcox.test(Ap.response.type$A22.opt, 
+                 Ap.response.type$Ar.opt,
+                 alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+
+mean.exp <- ddply(resp.TPM.dt.sub.pred, .(colony, Transcript), summarise,
+                  mean.TPM = mean(pTPM))
+
+t.test(log(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"]), 
+       log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]))
+
+# non-parametric test
+w5 = wilcox.test(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"],
+                 mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"],
+                 alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+
+plot(log(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"]), 
+     log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]))
+```
+
 For next analyses, extract list of gene names by response type.
@@
 Of the responsive transcripts, `r round(length(which(responsive.lms.ann.type$best.hit.to.nr != "-"))/length(responsive.lms.ann.type$best.hit.to.nr)*100, 0)`% are annotated, while `r round(length(which(annotation.table$best.hit.to.nr != "-"))/nrow(annotation.table) * 100, 0)`% of all transcripts are annotated.
+
+length(which(responsive.lms.ann.type$Transcript %in% annotation.table[which(annotation.table$best.hit.to.nr != "-"), "Sequence.Name"]))
+
+best.hit.to.nr != "-")
 
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
 ```

### 2014-10-03 doc/ApTranscriptome_manuscript.Rmd

@@
 
-## Species differences in the number of thermally-responsive genes
+## Transcripts differ in thermal responsiveness between species
 
-The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts.
-Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995].
+The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts.
+About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
+Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995].
 Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
-In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive.
-This result is the same if we focus on only the transcripts that have a BLAST hit.
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.
+The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246).
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.
 
-```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive genes by expression type for *A.
+```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A.
 carolinensis* and *A.
 picea*.
-**Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other."}
+**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other."}
 
@@
              big.mark = ",", split.tables = Inf, 
-             caption = "Table of the number of thermally-responsive genes by expression type for *A.
+             caption = "Table of the number of thermally-responsive transcripts by expression type for *A.
 carolinensis* and *A.
 picea*.
-**Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.")
+**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other.")
 ```
 
-The number of genes that were responsive to cold temperatures was substantially greater than of the number responsive to hot temperatures.
-We grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
+Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient.
+We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
-In both species,  the majority of genes were in the **Low** category (Table 1).
+
+We found that the number of transcripts that responded to low temperatures is substantially greater than of that to high temperatures (Table 1).
+In both species, the majority of transcripts were in the **Low** category (Table 1).
 For *A.
 carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A.
-picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Table 1).
+picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate**.
 
-Gene set enrichment analysis (GSEA) of the responsive genes indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.
-Specifically, out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** genes.
+Gene set enrichment analysis (GSEA) of the responsive transcripts indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes.
+Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** transcripts.
 These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis".
-For *High* genes, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".
-For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process".
-For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".
+For *High* transcripts, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 8 of the 9 terms being involved in regulation of biosynthetic or metabolic process".
+For **Bimodal** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "transcript expression", "signaling", and "protein modification process".
+For **Intermediate** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".
 
-# Shifts in the expression patterns of thermally-responsive genes between species
+# Shifts in the expression patterns of thermally-responsive transcripts between species
 
-Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
-While some of these genes were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` in *A.
-carolinensis* but not *A.
-picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` in *A.
-picea* but not *A.
-carolinensis*), most of these were genes that changed from one expression category to another between the species (i.e.
-the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).
+Of the thermally-responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
+This was not due to a species difference in the overall number of responsive genes (...stat...).
+Instead, most of these were transcripts that changed from one expression category to another (i.e.
+the number of transcripts in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).
+These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...)
 
-In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A.
-carolinensis*, the total number of thermally-responsive genes did not differ between the species (...r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ...
-).
-Specifically, compared to a hypothetical homogenous distribution, there were an excess of **Intermediate** genes in *A.
-carolinensis* which disproportionatly switched to **Low** or **Bimodal** expression categories in *A.
-picea* (Fig.
+**Intermediate** transcripts in *A.
+carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A.
+picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A.
+picea* **Intermediate** transcripts to any upregulated category in *A.
+carolinensis* (Fig.
+1).
+Similarly, both **High** and **Bimodal** transcripts in *A.
+picea* were more likely to be **Not Responsive** in *A.
+carolinensis* than expected, while the reverse was not significant (Fig.
 1).
 
-The molecular processes in this set of genes that switched from down-regulation in *A.
+The molecular processes in the set of transcripts that switched from down-regulation in *A.
 carolinensis* to up-regulation in *A.
@@
 In addition, there were GO terms related to repair (e.g.
-"histone modification"), gene regulation (e.g.
+"histone modification"), transcript regulation (e.g.
 "positive regulation of cellular process") and cellular organization (e.g.
 "transport", "organelle organization").
-These results indicate that in response to thermal extremes, the warmer-climate species, *A.
-carolinensis*, shuts down metabolism while the colder-climate species, *A.
-picea*, increases metabolism.
+Altogether, these results suggest that in response to thermal extremes, the warmer-climate species, *A.
+carolinensis*, primarily shuts down metabolism while the colder-climate species, *A.
+picea*, increases metabolism and initiates damage repair responses.
+
+...GO terms **High** and **Bimodal** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* ...
 
@@
 picea* (columns).
-The color of each cell represent the deviation of the observed from the expected number of genes based on hypothetical equivalence of the marginal frequencies.
+The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
 Blue cells are those with significant excesses, while red cells are those with significant absences."}
@@
 
-## Species differences in the sensitivity of thermally-responsive genes to changes in temperature
+## Species differences in the sensitivity of thermally-responsive transcripts to changes in temperature
 
-To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures, including *Bimodal* genes.
-Genes turned on at high temperatures in *A.
+To examine how the sensitivity of transcript expression to changes in temperature differed between the two species, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression.
+**High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes wre included for analysis at both high and low temperature activation.
+Transcripts turned on at high temperatures in *A.
 carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A.
-picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
+picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
 2).
-For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
+For low temperatures, there was no difference in the mean temperature of transcript activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
-picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+picea*; *t*~`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
 2).
 
-Similarly, for the **Intermediate** genes we compared the mean temperatures of downregulation of gene expression among species at both high (> 21°) and low (< 21°C) temperatures.
-Consistent with the results for up-regulated genes, we found that with increasing temperatures, the **Intermediate** genes were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
+Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures.
+Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ...
+at the higher temperature extremes and ...
+at low temperature extremes.
+Consistent with the results for up-regulated transcripts, we found greater inertia of gene expression to changes in temperature for *A.
+carolinensis* than *A.
+picea* drive by a differential response to heat stress.
+With increasing temperatures, **Intermediate** transcripts were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
@@
 picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
-Overall, these results indicate greater intertia of gene expression to changes in temperature for *A.
-carolinensis* than *A.
-picea*.
 
-```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at **Low** and **High** temperatures.
+**ADD PANEL TO FIG2 WITH DOWNREGULATION OF INTERMEDIATE TRANSCRIPTS**
+
+```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
 Red bars and lines are for *A.
@@
     scale_y_continuous(name="Density") +
-    scale_x_continuous(name=expression("Temperature of gene activation"))
+    scale_x_continuous(name=expression("Temperature of transcript activation"))
 print(T_on_plot)
@@
 
-## Adaptive changes in basal expression are not linked to genes involved in short-term physiological acclimation
+## Adaptive changes in basal expression are not linked to transcripts involved in short-term physiological acclimation
 
-We looked for evidence of genetic assimilation, the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed, in multiple comparisons.
-First, we predicted that gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A.
-carolinensis* or high temperatures in *A.
-picea* should be higher in the opposite species (e.g.
-greater basal expression but lower inducibility).
-Instead, we found that gene expression at optimal temperatures (about 19.5°C) was always greater in *A.
+We tested the hypothesis of adaptation by genetic assimilation, the process by which transcripts change from being environmentally induced to constitutively expressed, in multiple comparisons.
+First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression (e.g.
+"frontloading", [@barshis2013]).
+Thus, the warm-adaptated *A.
+carolinensis* should display higher basal expression of **High** transcripts than *A.
+picea*, while the reverse should be true for **Low** transcripts.
+Instead, we found that transcript expression at optimal temperatures (about 19.5°C) was always greater in *A.
 carolinensis* than in *A.
 picea*.
-This pattern was true for genes that had **Low** expression in *A.
-carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) **High** expression in *A.
-picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).
+This pattern was true for transcripts that had **High** expression in *A.
+picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), **Low** expression in *A.
+carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).
 
-Second, for the genes that had **High** or **Bimodal** expression in *A.
+In general, thermally-responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.
+
+The only exception was the set of genes that had **High** or **Bimodal** expression in *A.
 picea* but were not thermally-responsive in *A.
 carolinensis* (blue cells in bottom row of Fig.
-1), we predicted that mean expression level of the non-responsive *A.
-carolinensis* genes would be greater than the expression level near the optimum temperature (19.5°C) of the responsive *A.
-picea* genes.
-Again, we found no support for genetic assimilation as the non-responsive genes had lower mean expression levels.
+1).
+Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A.
+carolinensis*.
+Overall, we found no support for genetic assimilation as the primary mechanism for adaptive shifts in thermal tolerance.
 
@@
 
-We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig.
+We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig.
 3).
-Although `r resp.stress.responsive.count` genes were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
-In addition, of the `r nrow(hsp_all)` heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
+Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
+In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
 picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
@@
 
-By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes.
-We found that about 70% of the thermally-responsive genes (7,202) have non-linear responses to temperature.
+By using a regression-approach, we are able to quantitatively describe changes in transcript expression, rather than simply report the numbers of differentially-expressed transcripts.
+We found that about 70% of the thermally-responsive transcripts (7,202) have non-linear responses to temperature.
 
@@
 
-The patterns of gene expression between species differed substantially, with more genes down-regulated in *A.
+The patterns of transcript expression between species differed substantially, with more transcripts down-regulated in *A.
 carolinensis* and more up-regulated in *A.
@@
 1).
-The results of the gene set enrichment analysis give an indication of what might drive this difference.
-The genes that are shut-down by *A.
+The results of gene set enrichment analysis give an indication of what might drive this difference.
+The transcripts that are shut-down by *A.
 carolinensis* in response to cold temperatures, but up-regulated by *A.
@@
 
-```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
+```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive transcripts in each species."}
 # reshape data
@@
 
-Given the differences in the number of thermally-responsive genes between the two species (Fig.
+Given the differences in the number of thermally-responsive transcripts between the two species (Fig.
 1), it is worth noting the similarity in expression patterns between species.
-In both species, there are 3-4 $\times$ more genes upregulated at cold than high temperatures  (Fig.
+In both species, there are 3-4 $\times$ more transcripts upregulated at cold than high temperatures  (Fig.
 3).
@@
 In contrast to the predictions of the genetic assimilation hypothesis that adaptation occurs via changes from plastically-induced to constitutive expression (e.g.
-genetic assimilation), we found lower basal gene expression in *A.
-picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A.
+genetic assimilation), we found lower basal transcript expression in *A.
+picea* for transcripts that were up-regulated in response to cold temperatures (**Low** transcripts) in *A.
 carolinensis*.
 We also observed greater basal expression in *A.
-carolinensis* for **High** genes in *A.
-picea* as well as for genes that were **Intermediate** in both species.
-A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A.
-carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.
+carolinensis* for **High** transcripts in *A.
+picea* as well as for transcripts that were **Intermediate** in both species.
+A potential explanation for the consistently higher basal transcript expression in the warmer-environment species, *A.
+carolinensis*, is that this species has a higher metabolic rate and thus requires greater transcript expression.
 
@@
 
-The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 genes with a total length of 100 million bp.
-Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes.
-The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp.
-Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database.
-Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage.
-We removed 5,675 genes identified as known contaminants, leaving 99,861 clean transcripts.
-Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
+The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 transcripts with a total length of 100 million bp.
+Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 transcripts.
+The total length was 63 million bp with an N~50~ length of 895 bp and a mean transcript size of 593 bp.
+Of the 105,536 filtered transcripts, 55,432 had hits to the NCBI-nr database.
+Of these, 38,711 transcripts mapped to GO terms, 1,659 transcripts were identified to an enzyme and 18,935 transcripts mapped to a domain with >50% coverage.
+We removed 5,675 transcripts identified as known contaminants, leaving 99,861 clean transcripts.
+Of these, 47% have best hits to transcripts from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
 
-To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
+To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies transcripts by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
 
-We used subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.
+## Thermally-responsive transcripts
 
-## Thermally-responsive genes
-
-We quantified expression of each gene using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression.
-We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses.
+We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression.
+We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
 As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis.
@@
 
-To identify genes that have significant changes in expression across the thermal gradient, we fit the regression model
+To identify transcripts that have significant changes in expression across the thermal gradient, we fit the regression model
 
@@
 
-to each gene.
+to each transcript.
 We used $log(TPM + 1)$ as the response to control for skew in the expression data.
@@
 This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
-To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.
+To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
 
-We identified thermally-responsive genes as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
 All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
 
-For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
+For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
@@
 
-We performed comparative analyses between the two species focusing on the genes that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of gene activation and sensitivity of the thermally-responsive genes.
-For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
-To test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression for each gene in each species.
-We grouped the **High** genes along with the high temperature end of the **Bimodal** genes, and did the same for **Low** genes.
-We then performed a *t*-test to determine if the mean temperature of gene activation differed between the two species for each group.
+We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of transcript activation and sensitivity of the thermally-responsive transcripts.
+For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
+To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
+We grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
+We then performed a *t*-test to determine if the mean temperature of transcript activation differed between the two species for each group.
 
-To determine if the sensitivity of genes to thermal extremes differed between the species, we used the subset of **Intermediate** and **Bimodal** genes in each species.
-For each **Intermediate** gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
+To determine if the sensitivity of transcripts to thermal extremes differed between the species, we used the subset of **Intermediate** and **Bimodal** transcripts in each species.
+For each **Intermediate** transcript in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
 point of maximum expression had highest probability of being sampled).
@@
 Finding a greater standard deviation of expression would be indicative of less sensitivity to changes in temperature.
-For the **Bimodal** genes, we performed the same analysis as with the **Intermediate** genes, but with a greater standard deviation being indicative of a more sensitive gene (convex reaction norm) while a low standard deviation being a gene with greater inertia of expression in response to temperature changes (nearly flat).Thus, finding one species had overall greater standard deviation of expression for **Bimodal** genes would be indicative of greater thermal sensitivity for that species.
+For the **Bimodal** transcripts, we performed the same analysis as with the **Intermediate** transcripts, but with a greater standard deviation being indicative of a more sensitive transcript (convex reaction norm) while a low standard deviation being a transcript with greater inertia of expression in response to temperature changes (nearly flat).Thus, finding one species had overall greater standard deviation of expression for **Bimodal** transcripts would be indicative of greater thermal sensitivity for that species.
 
-To test for genetic assimilation, we compared basal expression levels (expression at 19.5° C) for the subset of genes that were responsive to cold temperatures (**Low** genes) in the southern *A.
-carolinensis* species, or hot temperatures (**High** genes) in the northern *A.
+To test for genetic assimilation, we compared basal expression levels (expression at 19.5° C) for the subset of transcripts that were responsive to cold temperatures (**Low** transcripts) in the southern *A.
+carolinensis* species, or hot temperatures (**High** transcripts) in the northern *A.
 picea* species using non-parametric two-sided Wilcoxon tests.
@@
 To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the package `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
-Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
+Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
 We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.

### 2014-10-06 doc/ApTranscriptome_manuscript.Rmd
revisions to ms, mostly Discussion

@@
 Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014].
-Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage (...refs...).
-Thermal defense strategies are shaped by the environmental regime organisms experience (...ref...), and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
+Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010].
+Thermal defense strategies are shaped by the environmental regime organisms experience [@gilchrist1995; @angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
 These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
@@
 Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
-We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
+We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
 picea* [@umphrey1996].
@@
 We used these data to test two specific predictions.
-First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms relative to species from cooler environments.
-[@barshis2013].
-This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature.
+First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance (e.g.
+"frontloading", [@barshis2013]) than plastically-induced resistance mechanisms relative to species from cooler environments.
+This pattern could be manifested through the number of thermally responsive genes, or their sensitivity to changes in temperature.
 Finding that fewer genes are thermally responsive in the warmer-climate species *A.
@@
 genetic assimilation [@pigliucci2006]).
-Finding that the genes up-regulated in response to cold temperatures in *A.
-carolinensis* have increased basal expression in the colder-climate species *A.
+Specifically, finding that the genes up-regulated in response to cold temperatures in *A.
+carolinensis* have increased constitutive expression in the colder-climate species *A.
 picea*, and vice versa, would be consistent with this prediction.
@@
 About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995].
+Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
 Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
 The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246).
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally responsive.
 
-```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A.
+```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally responsive transcripts by expression type for *A.
 carolinensis* and *A.
 picea*.
-**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other."}
+**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other."}
 
@@
              big.mark = ",", split.tables = Inf, 
-             caption = "Table of the number of thermally-responsive transcripts by expression type for *A.
+             caption = "Table of the number of thermally responsive transcripts by expression type for *A.
 carolinensis* and *A.
 picea*.
-**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other.")
+**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other.")
 ```
@@
 Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient.
-We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**).
+We grouped the thermally responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
@@
 
-# Shifts in the expression patterns of thermally-responsive transcripts between species
+# Shifts in the expression patterns of thermally responsive transcripts between species
 
-Of the thermally-responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
+Of the thermally responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
 This was not due to a species difference in the overall number of responsive genes (...stat...).
@@
 the number of transcripts in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).
-These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...)
+These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...) indicating that these differences are not simply due to stochastic processes.
 
@@
 
-## Species differences in the sensitivity of thermally-responsive transcripts to changes in temperature
+## Species differences in the sensitivity of thermally responsive transcripts to changes in temperature
 
@@
 
-## Adaptive changes in basal expression are not linked to transcripts involved in short-term physiological acclimation
+## Adaptive changes in constitutive expression are not linked to transcripts involved in short-term physiological acclimation
 
 We tested the hypothesis of adaptation by genetic assimilation, the process by which transcripts change from being environmentally induced to constitutively expressed, in multiple comparisons.
-First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression (e.g.
-"frontloading", [@barshis2013]).
+First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression.
 Thus, the warm-adaptated *A.
-carolinensis* should display higher basal expression of **High** transcripts than *A.
+carolinensis* should display higher constitutive expression of **High** transcripts than *A.
 picea*, while the reverse should be true for **Low** transcripts.
@@
 
-In general, thermally-responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.
+In general, thermally responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.
 
 The only exception was the set of genes that had **High** or **Bimodal** expression in *A.
-picea* but were not thermally-responsive in *A.
+picea* but were not thermally responsive in *A.
 carolinensis* (blue cells in bottom row of Fig.
@@
 
-## Candidate genes
+## Greater up-regulation of heat shock proteins in the cool-climate species *A. picea*
 
@@
 3).
-Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
+Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
 In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
@@
 
-By using a regression-approach, we are able to quantitatively describe changes in transcript expression, rather than simply report the numbers of differentially-expressed transcripts.
-We found that about 70% of the thermally-responsive transcripts (7,202) have non-linear responses to temperature.
+Species persistence in face of climate change requires both short-term physiological acclimation and longer-term adapative changes.
+In this study, we explored the relationship between these processes using whole organism gene expression data from closely-related ant species.
 
-This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012].
-Finding that thousands of genes responded to temperature highlights the limitations of previous studies [@meyer2011; @barshis2013; @palumbi2014; @oneil2014] that used only a few temperature levels and found that only a few dozen to hundred genes are involved in thermal responses.
-Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.
+Recent RNA-seq studies of thermal tolerance, using only a few temperature conditions, have identified a few dozen to hundred genes that are differentially expressed between conditions [@meyer2011; @barshis2013; @palumbi2014; @oneil2014].
+By applying a regression approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
+In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
+carolinensis* we found differences indicative of adaptation to climate, even while the total number of differentially-expressed genes was similar between the two species (Table 1).
+In total, we found that about 8% of transcripts (~9,000) are thermally responsive in each species, and that about 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.
+Generally, this indicates that the molecular response to thermal extremes is much more complex than indicated by previous studies, as would be expected given the ecological importance of thermal tolerance [@kingsolver2009].
 
-The patterns of transcript expression between species differed substantially, with more transcripts down-regulated in *A.
-carolinensis* and more up-regulated in *A.
-picea* in response to thermal extremes (Table 1, Fig.
-1).
-The results of gene set enrichment analysis give an indication of what might drive this difference.
-The transcripts that are shut-down by *A.
-carolinensis* in response to cold temperatures, but up-regulated by *A.
-picea* include gene ontology clusters involved in metabolism, transport and regulation.
-Thus, *A.
-carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A.
-picea* and so must rely on other mechanisms such as behavior or increased proteome stability to tolerate thermal stress.
-In contrast, *A.
-picea* increases metabolism to a greater extent than *A.
-carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress.
-Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A.
-picea*, but only 4 do in *A.
+Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript.
+In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A.
+carolinensis* than *A.
+picea* in nearly all comparisons.
+The exception to this was for genes that had **High** or **Bimodal** expression in *A.
+picea* and were not responsive in *A.
 carolinensis*.
-This result supports our expectation that *A.
-picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A.
+Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A.
 carolinensis*.
+...gene ontology terms...genes recruited to thermal tolerance...
+Consistent with our results but at a much shorter time-scale, no evidence for genetic assimilation at the expression level was found in an experimental evolution study with *Caenorhabditis remanei* selected for 10 generations to increased heat tolerance, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations [@sikkink2014].
+In this study, the genetic assimilation at the phenotypic level was caused by a shift in the threshold for temperature response, similar to what we observed here (Fig.
+2).
 
-Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
-Although the heat-shock response is one of the most conserved across living organisms (ref), it is energetically expensive, particularly under chronic stress conditions (ref).
-Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher basal level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al.
-2001).
-Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.
-Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.
+While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally responsive transcripts is consistent with the prediction that the warm-climate species *A.
+carolinensis* would rely more upon "frontloading" than the cool-climate *A.
+picea*.
+Moreover, *A.
+carolinensis* shut-down expression of more transcripts (Fig.
+2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig.
+2) than did *A.
+picea*.
+In constrast, the cool-climate species *A.
+picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation.
+One possible explanation for this difference in responses is that *A.
+picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A.
+carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress.
+Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A.
+picea*, but only 3 do in *A.
+carolinensis*.
+In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress.
+These differences are in line with the expectations if *A.
+picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A.
+carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions.
+In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
 
-```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive transcripts in each species."}
-# reshape data
-Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")
-
-maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=Species)) + 
-  geom_density(alpha=0.2, position="identity") + 
-  labs(x = "Temperature of maximum expression", y = "Density")
-
-print(maxexplot)
-
-####################################
-### CHANGE TO REACTION NORM FOR HSPs 
-####################################
-
-```
-
-Given the differences in the number of thermally-responsive transcripts between the two species (Fig.
-1), it is worth noting the similarity in expression patterns between species.
+Given the differences in the patterns thermal-responsiveness between species (Fig.
+2), it is worth noting the similarity in expression patterns between species.
 In both species, there are 3-4 $\times$ more transcripts upregulated at cold than high temperatures  (Fig.
@@
 
-In contrast to the predictions of the genetic assimilation hypothesis that adaptation occurs via changes from plastically-induced to constitutive expression (e.g.
-genetic assimilation), we found lower basal transcript expression in *A.
-picea* for transcripts that were up-regulated in response to cold temperatures (**Low** transcripts) in *A.
-carolinensis*.
-We also observed greater basal expression in *A.
-carolinensis* for **High** transcripts in *A.
-picea* as well as for transcripts that were **Intermediate** in both species.
-A potential explanation for the consistently higher basal transcript expression in the warmer-environment species, *A.
-carolinensis*, is that this species has a higher metabolic rate and thus requires greater transcript expression.
+A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels).
+A number of issues have hampered RNA-seq studies to date.
+Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2013].
+Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive genes in each species.
+Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive genes and shifts in temperatures of induction do not depend on functional annotation.
+Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].
 
@@
 Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for distantly-related model organisms *Drosophila* and *Apis*.
-Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005].
+This is a problem that plagues  Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005].
 Large differences in patterns of gene expression may have only small effects on organismal fitness.
-Given these caveats, transcriptomic studies promise to offer many new insights into how organisms respond and will respond to environmental change, especially for long-lived species that are intractable to study in the lab.
-
-An additional concern is differences in gene expression among species due to drift.
-While we do in fact expect some genes to differ in expession due to neutral processes, if this was the primary driver of differences among species we would expect to see equal distribution of counts in the off-diagonals of the matched observations (Fig.
-1).
-Instead we find that certain categories are over-represented, and that the differences in the reaction norm shapes are consistent with *a priori* expectations.
 
@@
 
+Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
+Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions (.....) .
+Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al.
+2001).
+Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.
+Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.
+
 In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A.
 carolinensis*.
-At least 10% of their transcriptomes are thermally-responsive.
+At least 10% of their transcriptomes are thermally responsive.
 For both species, more genes responded to cold than hot temperatures.
@@
 We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W).
-The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers.
-communication).
+The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice DeMarco, pers.
+communication), and there is about 2% pairwise divergence between these clades using five nuclear genes (Bernice DeMarco, unpublished data).
 Colonies of these species were maintained in the lab for 6 months prior to experimentation.
@@
 
-## Thermally-responsive transcripts
+## Thermally responsive transcripts
 
@@
 
-We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+We identified thermally responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
 All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
 
-For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*).
+For each thermally responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+We used the predicted transcript expression values to partition the thermally responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
@@
 
-We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of transcript activation and sensitivity of the thermally-responsive transcripts.
+We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of transcript activation and sensitivity of the thermally responsive transcripts.
 For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
-To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
+To test whether the temperature at which thermally responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
 We grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
@@
 
-To test for genetic assimilation, we compared basal expression levels (expression at 19.5° C) for the subset of transcripts that were responsive to cold temperatures (**Low** transcripts) in the southern *A.
+To test for genetic assimilation, we compared constitutive expression levels (expression at 19.5° C) for the subset of transcripts that were responsive to cold temperatures (**Low** transcripts) in the southern *A.
 carolinensis* species, or hot temperatures (**High** transcripts) in the northern *A.

### 2014-10-08 ApTranscriptome_TR.Rmd
updates to analysis

@@
 resp.TPM.dt.sub.pred[,pTPM.scaled:=scale(pTPM), by = Transcript]
+setkey(resp.TPM.dt.sub.pred, "Transcript")
 
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
 ```

### 2014-10-08 doc/ApTranscriptome_manuscript.Rmd

@@
 
-John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2},
-Mahesh Vangala \textsuperscript{2}, Robert R.
+John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
 Dunn \textsuperscript{3}, Aaron M.
@@
 Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010].
-Thermal defense strategies are shaped by the environmental regime organisms experience [@gilchrist1995; @angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
+Thermal defense strategies are shaped by the environmental regime organisms experience [@angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
 These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
 Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes.
-This information is critical for understanding the adaptive potential of species to future climates [@hoffman2008].
+This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].
 
@@
 To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material).
-We used these data to test two specific predictions.
-First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance (e.g.
-"frontloading", [@barshis2013]) than plastically-induced resistance mechanisms relative to species from cooler environments.
-This pattern could be manifested through the number of thermally responsive genes, or their sensitivity to changes in temperature.
-Finding that fewer genes are thermally responsive in the warmer-climate species *A.
-carolinensis* would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A.
-carolinensis* would support the latter.
-Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g.
-genetic assimilation [@pigliucci2006]).
-Specifically, finding that the genes up-regulated in response to cold temperatures in *A.
-carolinensis* have increased constitutive expression in the colder-climate species *A.
-picea*, and vice versa, would be consistent with this prediction.
+We used these data to quantitatively test a number of alternative mechanistic hypotheses to explain how populations evolutionarily increase their thermal breadth in response to stressful environments.
+First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species recruits additional molecular pathways to provide greater protection from damage.
+This hypothesis predicts that *A.
+picea* induces more genes in response to cold temperatures, whereas *A.
+carolinensis* induces more genes in response to high temperatures.
+An additional prediction of this hypothesis is that upregulation of the shared molecular pathways should be strong or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A.
+picea*, high temperatures for *A.
+carolinensis*.
+Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes.
+This hypothesis predicts that transcripts responsive to high temperatures in *A.
+picea* will have higher constitutive expression in *A.
+carolinensis*, while transcripts responsive to low temperatures in *A.
+carolinensis* will have higher constitutive expression in *A.
+picea*.
+Finally, the *tolerance* hypothesis (...ref...fields 2001?...) proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@fields2001] or behavioral [@gordon2013; @sunday2014] adapations.
+This hypothesis predicts that the two species should differ in the identity and timing of gene induction in response to temperature changes.
+Of course, these hypotheses are not mutually exclusive.
+By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.
 
@@
 #options(scipen = 1, digits = 3)
-load("ms_2014-09-26.RData")
+load("ms_2014-10-08.RData")
 library(data.table)
@@
 
-## Transcripts differ in thermal responsiveness between species
+## Thermally responsive gene responsed non-linearly to changes in temperature
 
@@
 
-```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally responsive transcripts by expression type for *A.
+```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A.
 carolinensis* and *A.
@@
              big.mark = ",", split.tables = Inf, 
-             caption = "Table of the number of thermally responsive transcripts by expression type for *A.
+             caption = "Table of the number of thermally-responsive transcripts by expression type for *A.
 carolinensis* and *A.
@@
 
+## *A. picea* recruits additional molecular pathways in response to heat stress
+
 Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient.
-We grouped the thermally responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
+We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
+The expression pattern differed between species for `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) of the thermally-responsive transcripts.
 
-We found that the number of transcripts that responded to low temperatures is substantially greater than of that to high temperatures (Table 1).
-In both species, the majority of transcripts were in the **Low** category (Table 1).
-For *A.
-carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A.
-picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate**.
+Consistent with the *enhanced response* hypothesis, *A.
+picea* induced expression of about 1,000 (30%) more transcripts than *A.
+carolinensis* in response to cold temperatures (Table 1, ...stats...).
+However, *A.
+picea* also induced expression of more transcripts, though to a smaller extent (~14%), than did *A.
+carolinensis* in response to high temperatures and induced more transcripts that are responsive to both high low temperatures (*Bimodal* transcripts; Table 1).
 
-Gene set enrichment analysis (GSEA) of the responsive transcripts indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes.
-Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** transcripts.
-These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis".
-For *High* transcripts, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 8 of the 9 terms being involved in regulation of biosynthetic or metabolic process".
-For **Bimodal** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "transcript expression", "signaling", and "protein modification process".
-For **Intermediate** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".
+The additional genes recruited included molecular processes such as...GO terms...
 
-# Shifts in the expression patterns of thermally responsive transcripts between species
+## *A. carolinensis* shuts-down gene expression in response to thermal extremes
 
-Of the thermally responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species.
-This was not due to a species difference in the overall number of responsive genes (...stat...).
-Instead, most of these were transcripts that changed from one expression category to another (i.e.
-the number of transcripts in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).
-These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...) indicating that these differences are not simply due to stochastic processes.
-
+While *A.
+picea* induced more genes, the total number of thermally-responsive transcripts did not differ between species (...stat...).
+Instead, most of these transcripts shifted from one responsive category to another.
+These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes.
 **Intermediate** transcripts in *A.
@@
 Blue cells are those with significant excesses, while red cells are those with significant absences."}
-#tt2
-
-# marginal sums and total
-rs <- rowSums(tt2)
-cs <- colSums(tt2)
-n <- sum(tt2)
-
-# grand mean for each class
-gm <- (rs + cs) / sum(tt2 * 2)
-
-# calculate observed values for each cell using overall mean for each expression type
-Ec <- outer(gm, gm, "*") *sum(tt2)
-
-# get deviations of observed from expected
-Ec.dev <- tt2 - Ec
-
-# overall table P-value
-Ec.chistat <- sum((tt2-Ec)^2 / Ec)
-# P-value
-pchisq(Ec.chistat, df=(nrow(tt2)-1) * (ncol(tt2)-1), lower.tail = FALSE)
-
-# overall minimum chi-squared value `X2.df` for a significantly non-homogenous distribution of counts
-df = (ncol(tt2) - 1) * (nrow(tt2) - 1)
-X2.df <- qchisq(p = 0.05, df = df, lower.tail = FALSE)
-
-# posthoc result for each cell
-Ec.cells <- sign(tt2 - Ec) * (tt2 - Ec)^2 / Ec
-Ec.cells.sign.table <- apply(Ec.cells, c(1,2), function(x) ifelse(abs(x) <= X2.df, "0", 
-                                                            ifelse(x > X2.df, "+", "-")))
-
-md <- melt(Ec.cells)
 
@@
 
-## Species differences in the sensitivity of thermally responsive transcripts to changes in temperature
+## Greater constitutive expression of thermally-responsive genes in *A. carolinensis*
 
-To examine how the sensitivity of transcript expression to changes in temperature differed between the two species, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression.
+We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by comparing constitutive expression, here defined as expression at 19.5°C, between species for high temperature responsive transcripts in *A.
+picea* and *A.
+carolinensis*.
+Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A.
+picea* for high temperature responsive transcripts in *A.
+picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`).
+However, this pattern was also true for low temperature responsive transcripts in *A.
+carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).
+Thus, the overall pattern is that thermally-responsive transcripts in *A.
+carolinensis* had higher constitutive expression than those in *A.
+picea*.
+
+An exception to this overall pattern is the set of genes that had **High** or **Bimodal** expression in *A.
+picea* but were not thermally responsive in *A.
+carolinensis* (blue cells in bottom row of Fig.
+1).
+Rather than having constitutively elevated expression as predicted, these genes were not expressed (...mean TPM...) in *A.
+carolinensis*.
+
+## *A. carolinensis* tolerates heat stress to a greater extent than *A. picea*
+
+To test the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes, and the identity of these genes.
+Specifically, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species.
 **High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes wre included for analysis at both high and low temperature activation.
-Transcripts turned on at high temperatures in *A.
+In support of the *tolerance* but not the *enhanced defense* hypothesis, transcripts turned on at high temperatures in *A.
 carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A.
@@
 at low temperature extremes.
-Consistent with the results for up-regulated transcripts, we found greater inertia of gene expression to changes in temperature for *A.
+Again, consistent with the *tolerance* hypothesis, we found greater inertia of gene expression to changes in temperature for *A.
 carolinensis* than *A.
@@
 ```
-
-## Adaptive changes in constitutive expression are not linked to transcripts involved in short-term physiological acclimation
-
-We tested the hypothesis of adaptation by genetic assimilation, the process by which transcripts change from being environmentally induced to constitutively expressed, in multiple comparisons.
-First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression.
-Thus, the warm-adaptated *A.
-carolinensis* should display higher constitutive expression of **High** transcripts than *A.
-picea*, while the reverse should be true for **Low** transcripts.
-Instead, we found that transcript expression at optimal temperatures (about 19.5°C) was always greater in *A.
-carolinensis* than in *A.
-picea*.
-This pattern was true for transcripts that had **High** expression in *A.
-picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), **Low** expression in *A.
-carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).
-
-In general, thermally responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.
-
-The only exception was the set of genes that had **High** or **Bimodal** expression in *A.
-picea* but were not thermally responsive in *A.
-carolinensis* (blue cells in bottom row of Fig.
-1).
-Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A.
-carolinensis*.
-Overall, we found no support for genetic assimilation as the primary mechanism for adaptive shifts in thermal tolerance.
-
-## Greater up-regulation of heat shock proteins in the cool-climate species *A. picea*
-
-We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig.
-3).
-Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
-In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
-picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
-carolinensis*.
 
@@
 
-While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally responsive transcripts is consistent with the prediction that the warm-climate species *A.
+While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is consistent with the prediction that the warm-climate species *A.
 carolinensis* would rely more upon "frontloading" than the cool-climate *A.
@@
 
+**put heat shock protein in Discussion**
+
+> We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig.
+3).
+Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
+In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
+picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
+carolinensis*.
+
 Given the differences in the patterns thermal-responsiveness between species (Fig.
@@
 A number of issues have hampered RNA-seq studies to date.
-Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2013].
+Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012].
 Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive genes in each species.
@@
 Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
-Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions (.....) .
+Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].
 Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al.
@@
 
-## Thermally responsive transcripts
+## Thermally-responsive transcripts
 
@@
 
-We identified thermally responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
 All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
 
-For each thermally responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted transcript expression values to partition the thermally responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
+For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
@@
 
-We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of transcript activation and sensitivity of the thermally responsive transcripts.
+We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of transcript activation and sensitivity of the thermally-responsive transcripts.
 For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
-To test whether the temperature at which thermally responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
+To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
 We grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.

### 2014-10-09 doc/ApTranscriptome_manuscript.Rmd
started edits to Discussion

@@
 ---
-title: Thermal reactionomes reveal adaptive responses to thermal
-  extremes in warm and cold-climate ant species
+title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
+  and cold-climate ant species
 author: ' '
@@
 output:
+  word_document: default
   pdf_document:
@@
 
-Recent RNA-seq studies of thermal tolerance, using only a few temperature conditions, have identified a few dozen to hundred genes that are differentially expressed between conditions [@meyer2011; @barshis2013; @palumbi2014; @oneil2014].
-By applying a regression approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
-In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
-carolinensis* we found differences indicative of adaptation to climate, even while the total number of differentially-expressed genes was similar between the two species (Table 1).
-In total, we found that about 8% of transcripts (~9,000) are thermally responsive in each species, and that about 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.
-Generally, this indicates that the molecular response to thermal extremes is much more complex than indicated by previous studies, as would be expected given the ecological importance of thermal tolerance [@kingsolver2009].
+Under the *enhanced defense* hypothesis, we predicted that the cool-climate species *A.
+picea* would induce more genes in response to low temperatures than *A.
+carolinensis*, while *A.
+carolinensis* would induce more genes in response to high temperatures.
+This prediction was only true for *A.
+picea*, which induced the same genes and then recruited additional genes in response to both low and high temperatures (Fig.
+1).
+In addition, we predicted that upregulation should occur in response to less extreme cold and high temperatures for *A.
+picea* and *A.
+carolinensis*, respectively.
+In contrast, we found no difference between species in the mean temperature of expression response for cold temperatures (Fig.
+2A,C), and that higher temperatures were required to induce the expression response for *A.
+carolinensis* than *A.
+picea* for cold temperatures.
+While conflicting with the *enhanced defense* hypothesis, this is consistent with the *tolerance* hypothesis for *A.
+carolinensis*.
 
-Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript.
+We also explored the *genetic assimilation* hypothesis that exposure to more frequent stressors would select for a shift from inducible to constitutive expression of stress-response genes.
 In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A.
@@
 
-While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is consistent with the prediction that the warm-climate species *A.
+While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is again consistent with the *tolerance* hypothesis for *A.
+carolinensis*.
+
+Thus, we find evidence that these two species have responded to temperature using distinct mechanisms. *A. picea* has enhanced its thermal response, while *A. carolinensis* tolerates thermal stress...
+
+----------------------------------------------------
+
+with the prediction that the warm-climate species *A.
 carolinensis* would rely more upon "frontloading" than the cool-climate *A.
@@
 
+Recent RNA-seq studies of thermal tolerance, using only a few temperature conditions, have identified a few dozen to hundred genes that are differentially expressed between conditions [@meyer2011; @barshis2013; @palumbi2014; @oneil2014].
+By applying a regression approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
+In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
+carolinensis* we found differences indicative of adaptation to climate, even while the total number of differentially-expressed genes was similar between the two species (Table 1).
+In total, we found that about 8% of transcripts (~9,000) are thermally responsive in each species, and that about 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.
+Generally, this indicates that the molecular response to thermal extremes is much more complex than indicated by previous studies, as would be expected given the ecological importance of thermal tolerance [@kingsolver2009].
+
+Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript.
+
 **put heat shock protein in Discussion**
@@
 carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
+
+CTmax and CTmin hypothesized to be genetically correlated, but not evident in terms of gene expression.
+Only 15-20% of genes upregulated in response to temp are bimodal (ie generalized stress response), and although upper activation thresholds differed significantly btwn species, lower thresholds were virtually identical.
+Implications .
+.
+.
 

### 2014-10-09 doc/ApTranscriptome_manuscript.Rmd
edits through end of Results

@@
 
-Thermal tolerance is a key trait governing the distributions of species and is likely useful in predicting their future responses to climate change.
-To understand how plasticity of gene expression underlies thermal tolerance, we describe the *thermal reactionome*, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient.
-We characterize the reactionomes of two common ant species in the eastern U.S: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
-We found that at least 10% of genes were thermally responsive.
-The warm-adapted *A.
-carolinensis* displayed higher mean levels of expression for responsive genes, exhibited greater inertia of gene expression change and upregulated fewer heat shock proteins than did *A.
-picea* in response to thermal extremes.
-This pattern is consistent with hypotheses of constitutive thermal defense and enhanced proteome stability.
-*Aphaenogaster carolinensis* also down-regulated expression of more genes in response to thermal extremes than did *A.
+**Limit 250 words, at 250! If you add words...find others to delete!**
+
+Thermal tolerance is a key trait governing the distributions of species and is useful in predicting their responses to climate change.
+To quantitatively evaluate among three hypotheses for the evolution of thermal tolerance, we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
+We characterized the thermal reactionomes of two common ant species in the eastern U.S: the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
+We found that at least 8% of genes were thermally responsive.
+The cool-adapted *A.
+picea* induced expression of more genes in response to extreme temperatures than *A.
+carolinensis*, including heat shock proteins, consistent with the *enhanced defense* hypothesis.
+In contrast, *A.
+carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than did *A.
 picea*.
-A subset of these same genes were upregulated in *A.
-picea*, including genes mapped to ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair.
-These differences indicate that the two closely related species employ different strategies for coping with temperature stress.
-The northern *A.
-picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas the southern *A.
-carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress.
+This pattern was consistent with *tolerance* hypothesis at high, but not low, temperatures.
+We found no evidence for the *genetic assimilation* hypothesis as there was no trade-off between constitutive and inducible levels of gene expression.
+Overall, our results indicate that the northern *A.
+picea* actively resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas the southern *A.
+carolinensis* relies upon other mechanisms to tolerate thermal stress.
 These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
@@
 # Significance Statement
+
+**Limit 120 words, at 120! If you add words...find others to delete!**
 
@@
 carolinensis* required more extreme temperatures to trigger a protective response and downregulated genes related to metabolism.
-These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatices.
+These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.
 
@@
 As a result, such comparisons (e.g.
-hot vs.
-cold) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is.
+high vs.
+low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is.
 Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992].
@@
 *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-*Aphaenogaster carolinensis* experiences significantly warmer and less seasonally variable temperatures than *A.
-picea* (...data...), with correspondingly higher minimal and maximal critical thermal limits (...CTmin and CTmax from HF/DF...).
-Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).
+*Aphaenogaster carolinensis* (mean annual temperature at collection site = ...) experiences significantly warmer and less seasonally variable temperatures than *A.
+picea* (mean annual temperature from collection site = ; [@hijmans2005]), with correspondingly higher minimal and maximal critical thermal limits that allow us to investigate adaptation to both lower and upper thermal extremes (...CTmin and CTmax from HF/DF...) [@diamond2012].
 
 To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material).
-We used these data to quantitatively test a number of alternative mechanistic hypotheses to explain how populations evolutionarily increase their thermal breadth in response to stressful environments.
-First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species recruits additional molecular pathways to provide greater protection from damage.
-This hypothesis predicts that *A.
-picea* induces more genes in response to cold temperatures, whereas *A.
+We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation.
+First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage.
+This hypothesis would predict that *A.
+picea* induces more genes in response to low temperatures, whereas *A.
 carolinensis* induces more genes in response to high temperatures.
-An additional prediction of this hypothesis is that upregulation of the shared molecular pathways should be strong or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A.
+Additionally, upregulation of molecular pathways shared by both species should be stronger or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A.
 picea*, high temperatures for *A.
@@
 Finally, the *tolerance* hypothesis (...ref...fields 2001?...) proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@fields2001] or behavioral [@gordon2013; @sunday2014] adapations.
-This hypothesis predicts that the two species should differ in the identity and timing of gene induction in response to temperature changes.
+This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways.
 Of course, these hypotheses are not mutually exclusive.
@@
 
-## Thermally responsive gene responsed non-linearly to changes in temperature
+## Reaction norms of thermally-responsive transcripts differ between species
 
@@
 The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246).
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally responsive.
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is likely a lower-bound for the true proportion of genes that are thermally responsive.
 
@@
 
-## *A. picea* recruits additional molecular pathways in response to heat stress
-
-Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient.
+The total number of thermally-responsive transcripts did not differ between species (...stat...).
+Within this set, there was significant variation in response both among transcripts and between the two species.
+Transcripts varied both qualitatively and quantitatively in their their thermal reaction norms.
 We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
-The expression pattern differed between species for `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) of the thermally-responsive transcripts.
+The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A.
+carolinensis* and *A.
+picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A.
+carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A.
+picea*.
 
-Consistent with the *enhanced response* hypothesis, *A.
-picea* induced expression of about 1,000 (30%) more transcripts than *A.
-carolinensis* in response to cold temperatures (Table 1, ...stats...).
-However, *A.
-picea* also induced expression of more transcripts, though to a smaller extent (~14%), than did *A.
-carolinensis* in response to high temperatures and induced more transcripts that are responsive to both high low temperatures (*Bimodal* transcripts; Table 1).
+## Expression response to thermal extremes differs between species
 
-The additional genes recruited included molecular processes such as...GO terms...
+Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Table 1).
+*Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), while *A.
+carolinensis* downregulated more transcripts under these conditions (*Intermediate*, ...stats...).
+Consistent with the *enhanced defense* hypothesis, *A.
+picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A.
+carolinensis* (Table 1, ...
+stats...).
+At high temperatures, *A.
+carolinensis* showed the opposite pattern, matching the predictions of the *tolerance* hypothesis, inducing expression of ~14% fewer transcripts than *A.
+picea* (Table 1).
 
-## *A. carolinensis* shuts-down gene expression in response to thermal extremes
-
-While *A.
-picea* induced more genes, the total number of thermally-responsive transcripts did not differ between species (...stat...).
-Instead, most of these transcripts shifted from one responsive category to another.
+In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species.
+Expression patterns differed for 6,792 (65%) of the thermally-responsive transcripts.
+Most of these shifted from one responsive category to another.
 These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes.
@@
 "transport", "organelle organization").
-Altogether, these results suggest that in response to thermal extremes, the warmer-climate species, *A.
-carolinensis*, primarily shuts down metabolism while the colder-climate species, *A.
+Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A.
+carolinensis*, primarily shuts down metabolism while the cooler-climate species, *A.
 picea*, increases metabolism and initiates damage repair responses.
-
-...GO terms **High** and **Bimodal** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* ...
 
@@
 
-## Greater constitutive expression of thermally-responsive genes in *A. carolinensis*
+## No evidence for *genetic assimilation*
 
-We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by comparing constitutive expression, here defined as expression at 19.5°C, between species for high temperature responsive transcripts in *A.
-picea* and *A.
-carolinensis*.
+We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by testing for each temperature extreme whether the suite of transcripts induced in response to that extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions.
 Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A.
 picea* for high temperature responsive transcripts in *A.
-picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`).
-However, this pattern was also true for low temperature responsive transcripts in *A.
-carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).
-Thus, the overall pattern is that thermally-responsive transcripts in *A.
-carolinensis* had higher constitutive expression than those in *A.
-picea*.
+picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`).
+However, the reverse pattern of higher constitutive expression in *A.
+picea* was not observed for low temperature responsive transcripts in *A.
+carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`).
 
-An exception to this overall pattern is the set of genes that had **High** or **Bimodal** expression in *A.
+Instead, the overall pattern is that thermally-responsive transcripts in *A.
+carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A.
+picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`).
+An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A.
 picea* but were not thermally responsive in *A.
@@
 1).
-Rather than having constitutively elevated expression as predicted, these genes were not expressed (...mean TPM...) in *A.
+Rather than having constitutively elevated expression as predicted, these genes were uniformally not expressed (...mean TPM...) in *A.
 carolinensis*.
+
+...GO terms **High** and **Bimodal** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* ...
 
@@
 
-To test the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes, and the identity of these genes.
-Specifically, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species.
-**High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes wre included for analysis at both high and low temperature activation.
+As an additional test of the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes.
+We compared the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species.
+**High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes were included for analysis at both high and low temperatures.
 In support of the *tolerance* but not the *enhanced defense* hypothesis, transcripts turned on at high temperatures in *A.
-carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A.
+carolinensis* were upregulated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A.
 picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
 2).
-For low temperatures, there was no difference in the mean temperature of transcript activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
+For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
-picea*; *t*~`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
 2).
@@
 at low temperature extremes.
-Again, consistent with the *tolerance* hypothesis, we found greater inertia of gene expression to changes in temperature for *A.
-carolinensis* than *A.
-picea* drive by a differential response to heat stress.
-With increasing temperatures, **Intermediate** transcripts were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
+Again, we found greater inertia of gene expression only in response to high temperature extremes.
+With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
-picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A.
+picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of downregulation (*A.
 carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A.
-picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
+picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
 
@@
 1).
-In addition, we predicted that upregulation should occur in response to less extreme cold and high temperatures for *A.
+In addition, we predicted that upregulation should occur in response to less extreme low and high temperatures for *A.
 picea* and *A.
 carolinensis*, respectively.
-In contrast, we found no difference between species in the mean temperature of expression response for cold temperatures (Fig.
+In contrast, we found no difference between species in the mean temperature of expression response for low temperatures (Fig.
 2A,C), and that higher temperatures were required to induce the expression response for *A.
 carolinensis* than *A.
-picea* for cold temperatures.
+picea* for low temperatures.
 While conflicting with the *enhanced defense* hypothesis, this is consistent with the *tolerance* hypothesis for *A.
@@
 Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
-In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A.
+In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to low or high temperatures in *A.
 picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
@@
 2), it is worth noting the similarity in expression patterns between species.
-In both species, there are 3-4 $\times$ more transcripts upregulated at cold than high temperatures  (Fig.
+In both species, there are 3-4 $\times$ more transcripts upregulated at low than high temperatures  (Fig.
 3).
@@
 picea* and *A.
-carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
+carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to low temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
 
@@
 At least 10% of their transcriptomes are thermally responsive.
-For both species, more genes responded to cold than hot temperatures.
-Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to hot than cold temperatures.
+For both species, more genes responded to low than high temperatures.
+Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to high than low temperatures.
 Our results also indicate that these two ant species have different responses to thermal extremes.
@@
 
-To test for genetic assimilation, we compared constitutive expression levels (expression at 19.5° C) for the subset of transcripts that were responsive to cold temperatures (**Low** transcripts) in the southern *A.
-carolinensis* species, or hot temperatures (**High** transcripts) in the northern *A.
+To test for genetic assimilation, we compared constitutive expression levels (expression at 19.5° C) for the subset of transcripts that were responsive to low temperatures (**Low** transcripts) in the southern *A.
+carolinensis* species, or high temperatures (**High** transcripts) in the northern *A.
 picea* species using non-parametric two-sided Wilcoxon tests.

### 2014-10-10 doc/ApTranscriptome_manuscript.Rmd
added site climate info and species CTmax to Intro

@@
 Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
-We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
+We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
+rudis*-species complex, B.
+DeMarco personal communication) and *A.
 picea* [@umphrey1996].
 *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-*Aphaenogaster carolinensis* (mean annual temperature at collection site = ...) experiences significantly warmer and less seasonally variable temperatures than *A.
-picea* (mean annual temperature from collection site = ; [@hijmans2005]), with correspondingly higher minimal and maximal critical thermal limits that allow us to investigate adaptation to both lower and upper thermal extremes (...CTmin and CTmax from HF/DF...) [@diamond2012].
+*Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A.
+picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (40.75°C for *A.
+carolinensis* in Raleigh, North Carolina compared to 38°C for *A.
+picea* at Harvard Forest, Massachusetts [@diamond2012]).
+Critical minimum temperatures have been shown to co-vary with critical maximum temperatures between and within *A.
+picea* and *A.
+rudis* [@warren2013], allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
+
+```{r climate, echo=FALSE, message=FALSE, warning=FALSE}
+library(raster)
+w <- getData('worldclim', var='bio', res=10)
+#names(w)
+
+# Plot first bioclim dimension (MAT) to check download worked
+#plot(w, 1, col=colorRampPalette(c("blue","green","red"))(255))
+
+# sites
+sites <- data.frame(site = c("Ac", "Ap"), lon=c(-78.874, -72.702), lat= c(36.037, 44.508))
+
+# extract bioclimatic variables from worldclim data. Note bioclim temperature data is multiplied by 10 (no decimal), so transform by dividing by 10.
+
+dbio <- extract(w, sites[,c("lon","lat")])
+# bio1 = MAT
+# bio4 = Temperature Seasonality (standard deviation *100)
+
+dbio.vars <- cbind(sites, dbio[,1], dbio[,4])
+colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 1]")] <- "MATx10"
+dbio.vars$MAT <- dbio.vars$MATx10/10
+colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 4]")] <- "Seasonality"
+
+dbio.vars
+```
 

### 2014-10-10 doc/ApTranscriptome_manuscript.Rmd
edits to text, mostly Discussion

--- 2014-10-09 doc/ApTranscriptome_manuscript.Rmd
@@
 title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
-  and cold-climate ant species
+  and cool-climate ant species
 author: ' '
@@
 
-John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
-Dunn \textsuperscript{3}, Aaron M.
-Ellison \textsuperscript{4}, Nathan J.
-Sanders \textsuperscript{5}, Nicholas J.
+John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, Lacy Chick \textsuperscript{2}, James Vincent \textsuperscript{3}, Mahesh Vangala \textsuperscript{3}, Robert R.
+Dunn \textsuperscript{4}, Aaron M.
+Ellison \textsuperscript{5}, Nathan J.
+Sanders \textsuperscript{2,6}, Nicholas J.
 Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}
@@
 \textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
-\textsuperscript{2} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
-\textsuperscript{3} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
-\textsuperscript{4} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
-\textsuperscript{5} Center for Macroecology, Evolution and Climate, University of Copenhagen
+\textsuperscript{2} Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, Tenessee 37996 \newline
+\textsuperscript{3} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
+\textsuperscript{4} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
+\textsuperscript{5} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
+\textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen
 
@@
 Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010].
-Thermal defense strategies are shaped by the environmental regime organisms experience [@angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
+Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta_jr2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
 These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes.
+Outside of candidate genes [@krebs1998], little is known about the evolution of thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes.
 This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].
@@
 Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
-We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
+We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
+rudis*-species complex, B.
+DeMarco personal communication) and *A.
 picea* [@umphrey1996].
 *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-*Aphaenogaster carolinensis* (mean annual temperature at collection site = ...) experiences significantly warmer and less seasonally variable temperatures than *A.
-picea* (mean annual temperature from collection site = ; [@hijmans2005]), with correspondingly higher minimal and maximal critical thermal limits that allow us to investigate adaptation to both lower and upper thermal extremes (...CTmin and CTmax from HF/DF...) [@diamond2012].
+*Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A.
+picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A.
+carolinensis* versus 41.4°C for *A.
+picea*; Supplemental Material) and minimum temperatures (6.1°C for *A.
+carolinensis* versus -0.1°C for *A.
+picea*; Supplemental Material) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
+
+```{r climate, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
+library(raster)
+w <- getData('worldclim', var='bio', res=10)
+#names(w)
+# Plot first bioclim dimension (MAT) to check download worked
+#plot(w, 1, col=colorRampPalette(c("blue","green","red"))(255))
+# sites
+sites <- data.frame(site = c("Ac", "Ap"), lon=c(-78.874, -72.702), lat= c(36.037, 44.508))
+# extract bioclimatic variables from worldclim data. Note bioclim temperature data is multiplied by 10 (no decimal), so transform by dividing by 10.
+dbio <- extract(w, sites[,c("lon","lat")])
+# bio1 = MAT
+# bio4 = Temperature Seasonality (standard deviation *100)
+dbio.vars <- cbind(sites, dbio[,1], dbio[,4])
+colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 1]")] <- "MATx10"
+dbio.vars$MAT <- dbio.vars$MATx10/10
+colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 4]")] <- "Seasonality"
+dbio.vars
+```
 
@@
 We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation.
-First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage.
+First, the *enhanced defense* hypothesis [@hofmann1996; @feder1999; @kultz2005] proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage.
 This hypothesis would predict that *A.
@@
 picea*.
-Finally, the *tolerance* hypothesis (...ref...fields 2001?...) proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@fields2001] or behavioral [@gordon2013; @sunday2014] adapations.
+Finally, the *tolerance* hypothesis [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@lockwood2012] or behavioral [@gordon2013; @sunday2014] adapations.
 This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways.
@@
 
-Species persistence in face of climate change requires both short-term physiological acclimation and longer-term adapative changes.
-In this study, we explored the relationship between these processes using whole organism gene expression data from closely-related ant species.
+The potential for many species to persist in face of contemporary climate change will depend in part upon their thermal tolerance.
+However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerance.
+By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
+In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
+carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species.
+Overall, we found that about 8% of transcripts (~9,000) are thermally responsive in each species.
+About 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.
 
-Under the *enhanced defense* hypothesis, we predicted that the cool-climate species *A.
-picea* would induce more genes in response to low temperatures than *A.
-carolinensis*, while *A.
-carolinensis* would induce more genes in response to high temperatures.
-This prediction was only true for *A.
-picea*, which induced the same genes and then recruited additional genes in response to both low and high temperatures (Fig.
+Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
+We found evidence for this hypothesis from the cool-climate species *A.
+picea* which induced more transcripts in response to low temperatures than did *A.
+carolinensis* (Table 1, Fig.
 1).
-In addition, we predicted that upregulation should occur in response to less extreme low and high temperatures for *A.
-picea* and *A.
-carolinensis*, respectively.
-In contrast, we found no difference between species in the mean temperature of expression response for low temperatures (Fig.
-2A,C), and that higher temperatures were required to induce the expression response for *A.
+However, *A.
+picea* also induced more transcripts in response to high temperatures than did *A.
+carolinensis*.
+We also found that higher temperatures were required to induce the expression response for *A.
 carolinensis* than *A.
-picea* for low temperatures.
-While conflicting with the *enhanced defense* hypothesis, this is consistent with the *tolerance* hypothesis for *A.
+picea* towards the high thermal limit (Fig.
+2B, D), consistent with the *tolerance* but not the *enhanced defense* hypothesis for *A.
 carolinensis*.
+There was no difference between species in temperature of response to low temperatures (Fig.
+2A,C).
+...compare to Drosophila...
 
-We also explored the *genetic assimilation* hypothesis that exposure to more frequent stressors would select for a shift from inducible to constitutive expression of stress-response genes.
+Exposure to more frequent stressors is hypothesized to select for a shift from inducible to constitutive expression of stress-response transcripts under the *genetic assimilation* hypothesis.
 In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A.
@@
 picea* in nearly all comparisons.
-The exception to this was for genes that had **High** or **Bimodal** expression in *A.
-picea* and were not responsive in *A.
-carolinensis*.
-Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A.
-carolinensis*.
-...gene ontology terms...genes recruited to thermal tolerance...
-Consistent with our results but at a much shorter time-scale, no evidence for genetic assimilation at the expression level was found in an experimental evolution study with *Caenorhabditis remanei* selected for 10 generations to increased heat tolerance, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations [@sikkink2014].
-In this study, the genetic assimilation at the phenotypic level was caused by a shift in the threshold for temperature response, similar to what we observed here (Fig.
-2).
-
+Consistent with our results but at a much shorter time-scale, Sikkink et al.
+[-@sikkink2014] found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
 While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is again consistent with the *tolerance* hypothesis for *A.
@@
 
-Thus, we find evidence that these two species have responded to temperature using distinct mechanisms. *A. picea* has enhanced its thermal response, while *A. carolinensis* tolerates thermal stress...
+The exception the higher constitutive expression of *A.
+carolinensis* relative to *A.
+picea* transcripts was for transcripts that had **High** or **Bimodal** expression in *A.
+picea* and were not responsive in *A.
+carolinensis*.
+Rather than being constitutively elevated as predicted, these transcripts were not expressed (...mean TPM...) in *A.
+carolinensis*.
+This further supports the *enhanced defense* hypothesis for adaptation of *A.
+picea* to cold temperatures as these are novel genes recruited to the low temperature response in this species.
+...gene ontology terms...
 
-----------------------------------------------------
-
-with the prediction that the warm-climate species *A.
-carolinensis* would rely more upon "frontloading" than the cool-climate *A.
+Taken together, these results indicate that these two species have distinct mechanisms of response to thermal extremes.
+*Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] to resist the impacts of thermal stress, requiring increased metabolism, while *A.
+carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress.
+If this hypothesis is correct, we would expect that candidate genes such as heat shock proteins would be over-represented in the thermally-responsive category for *A.
 picea*.
-Moreover, *A.
-carolinensis* shut-down expression of more transcripts (Fig.
-2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig.
-2) than did *A.
-picea*.
-In constrast, the cool-climate species *A.
-picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation.
-One possible explanation for this difference in responses is that *A.
-picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A.
-carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress.
-Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A.
-picea*, but only 3 do in *A.
+Consistent with this, of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
+picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
 carolinensis*.
@@
 
-Recent RNA-seq studies of thermal tolerance, using only a few temperature conditions, have identified a few dozen to hundred genes that are differentially expressed between conditions [@meyer2011; @barshis2013; @palumbi2014; @oneil2014].
-By applying a regression approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
-In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
-carolinensis* we found differences indicative of adaptation to climate, even while the total number of differentially-expressed genes was similar between the two species (Table 1).
-In total, we found that about 8% of transcripts (~9,000) are thermally responsive in each species, and that about 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.
-Generally, this indicates that the molecular response to thermal extremes is much more complex than indicated by previous studies, as would be expected given the ecological importance of thermal tolerance [@kingsolver2009].
-
-Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript.
-
-**put heat shock protein in Discussion**
-
-> We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig.
-3).
-Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`).
-In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to low or high temperatures in *A.
-picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
-carolinensis*.
+Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
+Only 15-20% of transcripts upregulated in response to temp are bimodal (i.e.
+generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical.
+Implications ...
 
@@
 
-CTmax and CTmin hypothesized to be genetically correlated, but not evident in terms of gene expression.
-Only 15-20% of genes upregulated in response to temp are bimodal (ie generalized stress response), and although upper activation thresholds differed significantly btwn species, lower thresholds were virtually identical.
-Implications .
-.
-.
-
 A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels).
 A number of issues have hampered RNA-seq studies to date.
-Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012].
-Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive genes in each species.
-Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive genes and shifts in temperatures of induction do not depend on functional annotation.
+Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify transcripts that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012].
+Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species.
+Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive transcripts and shifts in temperatures of induction do not depend on functional annotation.
 Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].
@@
 At least 10% of their transcriptomes are thermally responsive.
-For both species, more genes responded to low than high temperatures.
+For both species, more transcripts responded to low than high temperatures.
 Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to high than low temperatures.
@@
 
-To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the package `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013].
+To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013].
 Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
@@
 
-The raw Illumina reads are available at [...], the script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).
+The raw Illumina reads are available from the NCBI Short Read Archive .... The script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).
 

### 2014-10-10 doc/PNAS_cover_letter.tex
edits to text, mostly Discussion

@@
 
-First, we make a methodological contribution by extending the reaction
+In face of contemporary climate change, many species are predicted to be at 
+risk of extinction. However, there is also evidence that species can physiologically
+acclimate and rapidly adapt to changing environmental conditions, potentially
+rescuing them from extinction. This paradox poses a dilemma for evolutionary
+ecologists. In this study, we evaluate plasticity and adaptive differentiation
+in gene expression of closely related cool and warm-climate ant species.
+
+We present a novel analysis by extending the reaction
 norm approach to the study of gene expression. We define the
@@
 
-Second, by applying the reactionome approach to two related ant species,
+By applying the reactionome approach to two related ant species,
 \emph{Aphaenogaster picea} and \emph{A. carolinensis}, from colder and
@@
 
-\textbf{Suggested NAS Members:} James H. Brown, Bert Hoelldobler, Trudy F. Mackay, Nevo Eviata, Johanna Schmitt
+\textbf{Suggested NAS Members:} Trudy F. Mackay, Nevo Eviata, George N. Somero
 
-\textbf{Suggested reviewers:}
-
-\begin{itemize}
-  \item{Dr. Nicholas Teets, Postdoc, University of Florida}
-  \item{Dr. Vanessa Kellerman, Postdoc, Monash University}
-  \item{Dr. Pavlos Pavlidis, Postdoc, Heidelberg Institute of Theoretical Studies}
-  \item{Dr. Stephen Palumbi, Professor, Hopkins Marine Station, Stanford University}
-  \item{Dr. Joel Kingsolver, Professor, University of North Carolina}
-\end{itemize}
+\textbf{Suggested reviewers:}  Dr.
+Nicholas Teets, Postdoc, University of Florida; Dr.
+Vanessa Kellerman, Postdoc, Monash University; Dr.
+Pavlos Pavlidis, Postdoc, Heidelberg Institute of Theoretical Studies; Dr.
+Stephen Palumbi, Professor, Hopkins Marine Station, Stanford University; Dr.
+Joel Kingsolver, Professor, University of North Carolina
 
@@
 \includegraphics[scale=1]{JSG_signature.png}  \\
+
 Dr. John Stanton-Geddes

### 2014-10-13 doc/ApTranscriptome_manuscript.Rmd
edits from NG

@@
 
-**Limit 250 words, at 250! If you add words...find others to delete!**
-
 Thermal tolerance is a key trait governing the distributions of species and is useful in predicting their responses to climate change.
@@
 We characterized the thermal reactionomes of two common ant species in the eastern U.S: the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
-We found that at least 8% of genes were thermally responsive.
-The cool-adapted *A.
+We found that at least 8% of all genes were thermally responsive.
+Consistent with the *enhanced defense hypothesis*, the cool-adapted *A.
 picea* induced expression of more genes in response to extreme temperatures than *A.
-carolinensis*, including heat shock proteins, consistent with the *enhanced defense* hypothesis.
+carolinensis*, including heat shock proteins.
 In contrast, *A.
-carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than did *A.
+carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than *A.
 picea*.
-This pattern was consistent with *tolerance* hypothesis at high, but not low, temperatures.
-We found no evidence for the *genetic assimilation* hypothesis as there was no trade-off between constitutive and inducible levels of gene expression.
-Overall, our results indicate that the northern *A.
+This pattern was consistent with the *tolerance hypothesis* at high, but not low, temperatures.
+We found no evidence for the *genetic assimilation hypothesis* as there was no trade-off between constitutive and inducible levels of gene expression.
+Collectively, our results indicate that the northern *A.
 picea* actively resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas the southern *A.
 carolinensis* relies upon other mechanisms to tolerate thermal stress.
-These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
+These results are consistent with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
 
@@
 
-**Limit 120 words, at 120! If you add words...find others to delete!**
-
-Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptive changes.
-To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
+Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation.
+To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
 carolinensis*.
@@
 In contrast, *A.
-carolinensis* required more extreme temperatures to trigger a protective response and downregulated genes related to metabolism.
-These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.
+carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes.
+These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with low temperatures and other stressors.
 
@@
 Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009].
-Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014].
-Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010].
+Many species, especially small-bodied arthropods, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014].
+Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010].
 Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta_jr2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
@@
 
-To address this gap of knowlege, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011].
+To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011].
 In recent years, whole organism gene expression approaches (e.g.
@@
 Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
-We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
+We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
 rudis*-species complex, B.
@@
 To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material).
+
 We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation.
-First, the *enhanced defense* hypothesis [@hofmann1996; @feder1999; @kultz2005] proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage.
-This hypothesis would predict that *A.
-picea* induces more genes in response to low temperatures, whereas *A.
-carolinensis* induces more genes in response to high temperatures.
-Additionally, upregulation of molecular pathways shared by both species should be stronger or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A.
+First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced responses to provide greater protection from more frequently encountered stressors.
+This hypothesis would predict that the cool-adapted *A.
+picea* would induce more genes in response to low temperatures than would the warm-adapted *A.
+carolinensis*, which would induce more genes in response to high temperatures.
+Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A.
 picea*, high temperatures for *A.
 carolinensis*.
-Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes.
+Second, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes.
 This hypothesis predicts that transcripts responsive to high temperatures in *A.
 picea* will have higher constitutive expression in *A.
-carolinensis*, while transcripts responsive to low temperatures in *A.
+carolinensis*, whereas transcripts responsive to low temperatures in *A.
 carolinensis* will have higher constitutive expression in *A.
 picea*.
-Finally, the *tolerance* hypothesis [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@lockwood2012] or behavioral [@gordon2013; @sunday2014] adapations.
+Finally, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014].
 This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways.
@@
 About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
-Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach.
-The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246).
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is likely a lower-bound for the true proportion of genes that are thermally responsive.
+Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
+Of these 9,052 transcripts, 80% (`r prettyNum(length(quadratic.lms), big.mark = ",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
+high vs.
+low temperature).
+The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246).
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.
 
@@
 picea*.
-**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other."}
+**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}
 
@@
 The total number of thermally-responsive transcripts did not differ between species (...stat...).
-Within this set, there was significant variation in response both among transcripts and between the two species.
-Transcripts varied both qualitatively and quantitatively in their their thermal reaction norms.
+Within this set, there was significant qualitative and quantitative variation in thermal responses; both among transcripts and between the two species.
 We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
@@
 Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Table 1).
-*Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), while *A.
+*Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), whereas *A.
 carolinensis* downregulated more transcripts under these conditions (*Intermediate*, ...stats...).
@@
 stats...).
-At high temperatures, *A.
-carolinensis* showed the opposite pattern, matching the predictions of the *tolerance* hypothesis, inducing expression of ~14% fewer transcripts than *A.
+In contrast, at high temperatures, *A.
+carolinensis* matched the predictions of the *tolerance hypothesis*, inducing expression of ~14% fewer transcripts than *A.
 picea* (Table 1).
@@
 In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species.
-Expression patterns differed for 6,792 (65%) of the thermally-responsive transcripts.
-Most of these shifted from one responsive category to another.
+Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another.
 These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes.
@@
 carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A.
-picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A.
+picea* than expected by chance (...stat...), whereas there was a significant deficit of the reverse shift of *A.
 picea* **Intermediate** transcripts to any upregulated category in *A.
@@
 Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A.
-carolinensis*, primarily shuts down metabolism while the cooler-climate species, *A.
+carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A.
 picea*, increases metabolism and initiates damage repair responses.
@@
 
-We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by testing for each temperature extreme whether the suite of transcripts induced in response to that extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions.
+We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions.
 Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A.
-picea* for high temperature responsive transcripts in *A.
+picea* for transcripts that were responsive to high temperature in *A.
 picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`).
@@
 
-Instead, the overall pattern is that thermally-responsive transcripts in *A.
+Instead, the overall pattern is that thermally responsive transcripts in *A.
 carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A.
@@
 1).
-Rather than having constitutively elevated expression as predicted, these genes were uniformally not expressed (...mean TPM...) in *A.
+Rather than having constitutively elevated expression as predicted, these genes were uniformly not expressed (...mean TPM...) in *A.
 carolinensis*.
@@
 
-## *A. carolinensis* tolerates heat stress to a greater extent than *A. picea*
+## *A. carolinensis* has greater inertia of expression change to increases in temperature than does *A. picea*
 
-As an additional test of the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes.
-We compared the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species.
-**High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes were included for analysis at both high and low temperatures.
-In support of the *tolerance* but not the *enhanced defense* hypothesis, transcripts turned on at high temperatures in *A.
-carolinensis* were upregulated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A.
+As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures.
+We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression.
+In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A.
+carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A.
 picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
@@
 
-Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures.
-Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ...
-at the higher temperature extremes and ...
+Similarly, for the **Intermediate** transcripts we compared the mean temperatures of transcript downregulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures.
+Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately ...
+at the higher temperature extremes and a shift of ...
 at low temperature extremes.
-Again, we found greater inertia of gene expression only in response to high temperature extremes.
+Again, there was greater inertia of gene expression only in response to high temperature extremes.
 With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
-picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of downregulation (*A.
+picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures the species did not differ in mean temperature of downregulation (*A.
 carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A.
-picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
+picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
 
@@
 
-The potential for many species to persist in face of contemporary climate change will depend in part upon their thermal tolerance.
-However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerance.
-By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
-In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
-carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species.
-Overall, we found that about 8% of transcripts (~9,000) are thermally responsive in each species.
-About 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.
-
-Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
-We found evidence for this hypothesis from the cool-climate species *A.
-picea* which induced more transcripts in response to low temperatures than did *A.
-carolinensis* (Table 1, Fig.
-1).
-However, *A.
-picea* also induced more transcripts in response to high temperatures than did *A.
+Collectively, our results indicate that these two species have distinct mechanisms of response to thermal extremes.
+The cool-adapated species *Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] with elevated gene expression at both low and high temperatures, which has been shown to impose a metabolic cost [@hoekstra2013].
+In contrast, the warm-adapted *A.
+carolinensis* relies upon constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress.
+If the *enhanced defense* is hypothesis is correct, candidate genes such as heat shock proteins should be over-represented in the thermally-responsive category for *A.
+picea*.
+Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
+picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
 carolinensis*.
-We also found that higher temperatures were required to induce the expression response for *A.
-carolinensis* than *A.
-picea* towards the high thermal limit (Fig.
-2B, D), consistent with the *tolerance* but not the *enhanced defense* hypothesis for *A.
-carolinensis*.
-There was no difference between species in temperature of response to low temperatures (Fig.
-2A,C).
-...compare to Drosophila...
-
-Exposure to more frequent stressors is hypothesized to select for a shift from inducible to constitutive expression of stress-response transcripts under the *genetic assimilation* hypothesis.
-In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A.
-carolinensis* than *A.
-picea* in nearly all comparisons.
-Consistent with our results but at a much shorter time-scale, Sikkink et al.
-[-@sikkink2014] found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
-While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is again consistent with the *tolerance* hypothesis for *A.
-carolinensis*.
-
-The exception the higher constitutive expression of *A.
-carolinensis* relative to *A.
-picea* transcripts was for transcripts that had **High** or **Bimodal** expression in *A.
-picea* and were not responsive in *A.
-carolinensis*.
-Rather than being constitutively elevated as predicted, these transcripts were not expressed (...mean TPM...) in *A.
-carolinensis*.
-This further supports the *enhanced defense* hypothesis for adaptation of *A.
-picea* to cold temperatures as these are novel genes recruited to the low temperature response in this species.
-...gene ontology terms...
-
-Taken together, these results indicate that these two species have distinct mechanisms of response to thermal extremes.
-*Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] to resist the impacts of thermal stress, requiring increased metabolism, while *A.
-carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress.
-If this hypothesis is correct, we would expect that candidate genes such as heat shock proteins would be over-represented in the thermally-responsive category for *A.
-picea*.
-Consistent with this, of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
-picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
-carolinensis*.
-In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress.
 These differences are in line with the expectations if *A.
@@
 
-Given the differences in the patterns thermal-responsiveness between species (Fig.
+Given the differences in the patterns of thermal responsiveness between species (Fig.
 2), it is worth noting the similarity in expression patterns between species.
-In both species, there are 3-4 $\times$ more transcripts upregulated at low than high temperatures  (Fig.
+In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Fig.
 3).
@@
 This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
-At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012].
-In this light, for *A.
-picea* and *A.
-carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to low temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].
+Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012].
+To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012].
 
-A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels).
+A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels).
 A number of issues have hampered RNA-seq studies to date.
@@
 Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species.
-Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive transcripts and shifts in temperatures of induction do not depend on functional annotation.
-Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].
+While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
+Thus, our method provides an example of how RNA-seq data can be used to understand differences among species  ...
 
@@
 First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented.
-Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts.
+Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts.
 A genome assembly, in progress, will help to reduce fragmentation.
-Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for distantly-related model organisms *Drosophila* and *Apis*.
-This is a problem that plagues  Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005].
-Large differences in patterns of gene expression may have only small effects on organismal fitness.
-
-## Conclusions
+Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*.
+Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.
 
@@
 Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.
+
+## Conclusion
 

### 2014-10-14 ApTranscriptome_TR.Rmd
added plot with critical temperature of up and downregulation

@@
 library(coin)
+library(gridExtra)
 
@@
 
-Thermally-responsive genes could also differ in the temperatures at which they have increased or decreased expression in response to temperature changes.
-To examine this, I determine the temperature at which each responsive gene has the greated increase or decrease in expression.
+Thermally-responsive genes could also differ in the critical temperature of gene induction.
+To examine this, I determine the temperature at which each responsive gene has the greatest increase or decrease in expression.
 
@@
 
-Does the mean expression level for all responsive genes differ among colonies?
-
-```{r mean_exp}
-t.test(log(Ap.response.type$A22.opt), log(Ap.response.type$Ar.opt))
-# expression level at optimum temp greater in Ar than A22
-
-# non-parametric test
-w4 = wilcox.test(Ap.response.type$A22.opt, 
-                 Ap.response.type$Ar.opt,
-                 alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-
-mean.exp <- ddply(resp.TPM.dt.sub.pred, .(colony, Transcript), summarise,
-                  mean.TPM = mean(pTPM))
-
-t.test(log(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"]), 
-       log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]))
-
-# non-parametric test
-w5 = wilcox.test(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"],
-                 mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"],
-                 alternative = "two.sided", paired = TRUE, conf.int = TRUE)
-
-plot(log(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"]), 
-     log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]))
-```
-
 For next analyses, extract list of gene names by response type.
@@
 l5 <- length(unique(A22.int.TPM.dt.sub$Transcript))
-A22.int.T_off <- data.frame(Transcript = unique(A22.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l5), type = rep("Low", length = l5), T_off_high = NA, T_off_low = NA)
+A22.int.T_off <- data.frame(Transcript = unique(A22.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l5), type = rep("Intermediate", length = l5), High = NA, Low = NA)
 
@@
     subdf.high <- subdf[which(subdf$val >= 21), ]
-    T_off_high <- subdf.high[median(which(diff(subdf.high$TPM) == min(diff(subdf.high$TPM)))), val]
-    A22.int.T_off[which(A22.int.T_off$Transcript == i), "T_off_high"] <- T_off_high
+    High <- subdf.high[median(which(diff(subdf.high$TPM) == min(diff(subdf.high$TPM)))), val]
+    A22.int.T_off[which(A22.int.T_off$Transcript == i), "High"] <- High
 
     subdf.low <- subdf[which(subdf$val <= 21), ]
-    T_off_low <- subdf.low[median(which(diff(subdf.low$TPM) == min(diff(subdf.low$TPM)))), val]
-    A22.int.T_off[which(A22.int.T_off$Transcript == i), "T_off_low"] <- T_off_low
+    Low <- subdf.low[median(which(diff(subdf.low$TPM) == min(diff(subdf.low$TPM)))), val]
+    A22.int.T_off[which(A22.int.T_off$Transcript == i), "Low"] <- Low
 }
@@
 l6 <- length(unique(Ar.int.TPM.dt.sub$Transcript))
-Ar.int.T_off <- data.frame(Transcript = unique(Ar.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l6), type = rep("Low", length = l6), T_off_high = NA, T_off_low = NA)
+Ar.int.T_off <- data.frame(Transcript = unique(Ar.int.TPM.dt.sub$Transcript), colony = rep("AcNC", length = l6), type = rep("Intermediate", length = l6), High = NA, Low = NA)
 
@@
     subdf.high <- subdf[which(subdf$val >= 21), ]
-    T_off_high <- subdf.high[median(which(diff(subdf.high$TPM) == min(diff(subdf.high$TPM)))), val]
-    Ar.int.T_off[which(Ar.int.T_off$Transcript == i), "T_off_high"] <- T_off_high
+    High <- subdf.high[median(which(diff(subdf.high$TPM) == min(diff(subdf.high$TPM)))), val]
+    Ar.int.T_off[which(Ar.int.T_off$Transcript == i), "High"] <- High
 
     subdf.low <- subdf[which(subdf$val <= 21), ]
-    T_off_low <- subdf.low[median(which(diff(subdf.low$TPM) == min(diff(subdf.low$TPM)))), val]
-    Ar.int.T_off[which(Ar.int.T_off$Transcript == i), "T_off_low"] <- T_off_low
+    Low <- subdf.low[median(which(diff(subdf.low$TPM) == min(diff(subdf.low$TPM)))), val]
+    Ar.int.T_off[which(Ar.int.T_off$Transcript == i), "Low"] <- Low
 }
@@
 # compare T_off among species
-(T_off.int.high.ttest <- t.test(Ar.int.T_off$T_off_high, A22.int.T_off$T_off_high))
-
-(T_off.int.low.ttest <- t.test(Ar.int.T_off$T_off_low, A22.int.T_off$T_off_low))
-```
-
-As with genes turned on at high temperatures, *A. carolinensis* **Intermediate** genes are down-regulated on average at 1.2C higher temperatures than *A. picea* **Intermediate** genes.
+(T_off.int.high.ttest <- t.test(Ar.int.T_off$High, A22.int.T_off$High))
+
+(T_off.int.low.ttest <- t.test(Ar.int.T_off$Low, A22.int.T_off$Low))
+```
+
+As with genes turned on at high temperatures, *A.
+carolinensis* **Intermediate** genes are down-regulated on average at 2.7°C higher temperatures than *A.
+picea* **Intermediate** genes.
+In contrast, towards low temperatures, the temperature of mean temperature of downregulation was higher for *A.
+picea* (10.6°) than *A.
+carolinensis* (9.75°).
+
+Visualize T~off~ for both *Low* and *High* genes on the same plot.
+
+```{r activation_plot, echo=FALSE, eval=TRUE}
+# reshape data
+Ar.T_off.df <- melt(Ar.int.T_off, id.vars = c("Transcript", "colony", "type"), 
+                 measure.vars = c("High", "Low"),
+                 variable.name = "class",
+                 value.name = "T_off")
+
+A22.T_off.df <- melt(A22.int.T_off, id.vars = c("Transcript", "colony", "type"), 
+                 measure.vars = c("High", "Low"),
+                 variable.name = "class",
+                 value.name = "T_off")
+
+# merge data for plotting
+Ap.T_off <- rbind(Ar.T_off.df, A22.T_off.df)
+Ap.T_off$Species <- ifelse(Ap.T_off$colony == "ApVT", "A. picea", "A. carolinensis")
+# change factor levels to "Low" first
+Ap.T_off$class = factor(Ap.T_off$class, levels = c("Low", "High"))
+
+# combined plot
+T_off_plot <- ggplot(Ap.T_off, aes(x = T_off, y = ..density.., colour = Species, group = Species, fill = Species)) +
+    facet_grid(. ~ class) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    labs(x = "Temperature of gene activation", y = "Density") +
+    theme(axis.title = element_text(size = rel(2))) +
+    theme(axis.text = element_text(size = rel(1.2)))
+#    theme(legend.position = "none")
+print(T_off_plot)
+```
+
+Make a plot with all 4 gene activation comparisons.
+
+```{r combined_plot}
+gg2 <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
+    facet_grid(. ~ type) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    labs(x = "Critical temperature of upregulation", y = "Density") +
+    theme(legend.position = "none")
+
+gg3 <- ggplot(Ap.T_off, aes(x = T_off, y = ..density.., colour = Species, group = Species, fill = Species)) +
+    facet_grid(. ~ class) +
+    geom_histogram(position = "dodge", binwidth = 3) +
+    geom_density(adjust = 3, fill = NA, size = 2) + 
+    labs(x = "Critical temperature of downregulation", y = "Density") +
+    theme(legend.position = "bottom", legend.text = element_text(size = 12, face = 3))
+
+# grid.arrange
+grid.arrange(gg2, gg3, nrow = 2)
+```
+
+Does the mean expression level for all responsive genes differ among colonies?
+
+```{r mean_exp}
+t.test(log(Ap.response.type$A22.opt), log(Ap.response.type$Ar.opt))
+# expression level at optimum temp greater in Ar than A22
+
+# non-parametric test
+w4 = wilcox.test(Ap.response.type$A22.opt, 
+                 Ap.response.type$Ar.opt,
+                 alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+
+mean.exp <- ddply(resp.TPM.dt.sub.pred, .(colony, Transcript), summarise,
+                  mean.TPM = mean(pTPM))
+
+t.test(log(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"]), 
+       log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]))
+
+# non-parametric test
+w5 = wilcox.test(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"],
+                 mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"],
+                 alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+
+plot(log(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"]), 
+     log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]))
+```
 

### 2014-10-14 ApTranscriptome_TR.Rmd
cleaning up files by removing analyses no longer referred to in ms

@@
 - compare basal expression at optimal temperature for thermally-responsive genes
-- compare thermal stability of *Intermediate* genes
-- compare thermal sensitivity of *Bimodal* genes
-- compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes
-- evaluate the extent to which thermal reaction norms differ between species by mean and shape
+- compare degree of uregulation to constitutive expression
+- compare the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures 
+- compare the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures
 
@@
 
-### Compare thermal stability of *Intermediate* genes
-
-*Intermediate* genes are those that have expression that is shut-down as conditions become stressful, likely non-essential molecular processes.
-We hypothesized that if the more southern *Ar* species was more thermally-tolerant than *A22*, transcripts with *Intermediate* expression (10-30C) would be active across a wider range of temperatures.
-To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each species.
-
-```{r thermal_breadth}
-# extract 'Intermediate' expressed transcripts for A22
-A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
-A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
-length(A22.int.lm)
-
-# apply `transcriptSD` function to all transcripts
-A22.int.sd <- unlist(Map(transcriptSD, A22.int.lm, colony = "A22"))
-A22.int.sd <- data.frame(colony = "A22", exp_sd = A22.int.sd)
-
-# repeat for Ar
-Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
-Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
-Ar.int.sd <- unlist(Map(transcriptSD, Ar.int.lm, colony = "Ar"))
-Ar.int.sd <- data.frame(colony = "Ar", exp_sd = Ar.int.sd)
-
-# T-test comparing standard deviation of expression between colonies
-(t.varint <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
-```
-
-Consistent with our hypothesis, *Intermediate* transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.
-
-```{r plot_thermal_breadth, echo=FALSE, eval=TRUE}
-# prep data for plotting
-Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)
-
-# plot thermal breadth
-g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
-g1 + scale_y_continuous(name="Density") +
-  scale_x_continuous(name=expression("Standard deviation of expression"))
-```
-
-### Compare thermal sensitivity of *Bimodal* genes
-
-As the converse of the above hypothesis, a species that is especially thermally-sensitive is likely to activate expression of molecular processes for thermal tolerance more quickly.
-We tested this using the same approach as for the *Intermediate* transcripts, but using the inverse of the *Bimodal* expressed transcripts.
-
-```{r calc_Bimodal_variance}
-# extract 'Bimodal' expressed transcripts for A22 species
-A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
-length(A22.bim.lm)
-
-# apply `transcriptSD` function to all transcripts
-A22.bim.sd <- unlist(Map(transcriptSD, A22.bim.lm, colony = "A22"))
-
-# repeat for Ar 
-Ar.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]]
-length(Ar.bim.lm)
-Ar.bim.sd <- unlist(Map(transcriptSD, Ar.bim.lm, colony = "Ar"))
-
-# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
-(t.varbim <- t.test(Ar.bim.sd, A22.bim.sd))
-```
-
-No difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.
-
-### Compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes
+### Comparison of the degree of upregulation against constitutive expression
+
+To test the genetic assimilation hypothesis, I compare the degree of upregulation of a transcript against constitutive expression at optimal temperatures.
+
+```{r induction_vs_constitutive}
+Ap.constitutive <- resp.TPM.dt.sub.pred[resp.TPM.dt.sub.pred$val == ]
+
+```
+
+### Compare the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures
 
@@
 
-Repeat analysis for when **Intermediate** genes are turned *off*.
+### Compare the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures
+
+This analysis is comparable to the one above, but for the *Intermediate* genes that are downregulated, split into the low and high temperature extremes.
 
@@
 # grid.arrange
+#tiff("results/crit_temp_regulation_Fig2.tiff", compression = "lzw", bg = "white")
 grid.arrange(gg2, gg3, nrow = 2)
+#dev.off()
 ```
@@
 ```
-
-### Evaluate the extent to which differences in thermal reaction norms are due to mean or shape
-
-For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect?
-To test this, I rougly follow `r citet("10.1086/675302")` by defining differences among reation norms for individual genes due to changes in the overall mean, slope, curvature and all higher-order shape differences (i.e.
-wiggle).
-
-- *Mean, M*: overall difference in the mean expression value across all temperatures
-- *Slope, S*: difference in overall slope
-- *Curvature, C*: average difference in curvature of the reaction norm
-- *Wiggle, W*: variability in shape not captured by the previous three measures
-
-```{r varshape, echo=TRUE, eval=TRUE}
-varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)
-
-# delta mean
-boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
-boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))
-
-t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean), paired = TRUE)
-t.mean
-
-# delta slope
-boxplot(log(varshape.out$A22.slope), log(varshape.out$Ar.slope))
-t.slope <- t.test(log(varshape.out$A22.slope + 0.1), log(varshape.out$Ar.slope + 0.1), paired = TRUE)
-t.slope
-
-# delta curvature
-boxplot(log(varshape.out$A22.curve), log(varshape.out$Ar.curve))
-t.curvature <- t.test(log(varshape.out$A22.curve + 0.1), log(varshape.out$Ar.curve + 0.1), paired = TRUE)
-t.curvature
-
-# delta wiggle - small values so use un-transformed
-boxplot(varshape.out$A22.wiggle, varshape.out$Ar.wiggle)
-t.wiggle <- t.test(varshape.out$A22.wiggle, varshape.out$Ar.wiggle, paired = TRUE)
-t.wiggle
-```
-
-Reaction norms differ between species by mean and slope, but not by curvature or wiggle.
-
-Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means and slope, as curvature and wiggle did not differ between species.
-
-```{r partition_reaction_norm_variation}
-# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
-# mean = mean1 - mean2
-varshape.out$lmean <- log(varshape.out$A22.mean) - log(varshape.out$Ar.mean)
-
-# slope = slope1 - slope2
-varshape.out$lslope <- log(varshape.out$A22.slope + 0.1) - log(varshape.out$Ar.slope + 0.1)
-
-# take absolute value of each value and sum to get total differences between reaction norms
-varshape.out$ltotal <- abs(varshape.out$lmean) + abs(varshape.out$lslope)
-
-# variation in reaction norms due to differences in mean
-varshape.out$prop.lmean <- abs(varshape.out$lmean) / varshape.out$ltotal
-varshape.out$prop.lslope <- abs(varshape.out$lslope) / varshape.out$ltotal
-
-# Mean proportion and 95% CI of total variation of each measure
-mean(varshape.out$prop.lmean)
-(qlmean <- quantile(varshape.out$prop.lmean, probs = c(0.05, 0.5, 0.95)))
-
-mean(varshape.out$prop.lslope)
-(qlslope <- quantile(varshape.out$prop.lslope, probs = c(0.05, 0.5, 0.95)))
-```
-
-From this analysis, about 80% of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.
 
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
+save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.T_off, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
 ```

### 2014-10-14 scripts/RxNseq.R

@@ -196,88 +196,7 @@ RxNtype <- function(lmitem) {
 } # end RxNtype
 
 
-############################################################################################
-## transcriptSD
-############################################################################################
 
-transcriptSD <- function(lmitem, colony) {
-
-    # function passed to Map() that calculates the standard deviation of expression for each 'Intermediate' transcript
-    #
-    # Arguments:
-    #  lmitem: List item of class `lm`
-    #  colony: Colony that standard deviation of expression is being calculated for
-    #
-    # Returns:
-    #   list of standard deviation of expression for each transcript
-
-    # check class lm
-    if(!"lm" %in% class(lmitem)) stop("Object not of class lm")
-    
-    # predict only for A22
-    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
-    newdf <- data.frame(val = pred.vals, colony = rep(colony, length = length(pred.vals)))
-    pout <- predict(lmitem, newdata=newdf)
-    pout <- cbind(newdf, pTPM = pout) # note this is on log scale
-
-    # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
-    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))
-
-    # draw 1000 'temps' weighted by predicted expression
-    random.draw <- sample(size = 1000, pred.vals, prob = pout$pTPM, replace = TRUE)
-
-    # calculate standard deviation of weighted random.draw
-    data.frame(exp_sd = sd(random.draw))
-    } # end function
-
-
-############################################################################################
-## RxNvarshape
-############################################################################################
-
-RxNvarshape <- function(lmitem) {
-    # check class lm
-    if(!"lm" %in% class(lmitem)) stop("Object not of class lm")
-
-    # calculate mean expression across all temps
-    # predict 
-    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
-    colony.levels <- as.factor(rep(unlist(lmitem$xlevels)))
-    # for transcripts with no 'colony' effect in model, no levels, so create dummy
[…]

### 2014-10-14 ApTranscriptome_TR.Rmd
visualization of induction vs constitutive expression

@@
 
-**July 17, 2014**
+**October 14, 2014**
 
@@
 colnames(Ap.response.type)[which(colnames(Ap.response.type) == ".id")] <- "Transcript"
+Ap.response.type <- data.table(Ap.response.type)
+setkey(Ap.response.type, Transcript)
 str(Ap.response.type)
@@
 mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
-  scale_fill_gradient2(limits=c(-600, 600)) 
+  scale_fill_gradient2(limits=c(-600, 600)) +
+  theme(axis.title = element_text(face="italic")) +
+  labs(x = "A. picea", y = "A. carolinensis")
+
+tiff("results/matched_observations_Fig1.tiff", compression = "lzw")
 mh_plot
+dev.off()
 ```
@@
 ```{r induction_vs_constitutive}
-Ap.constitutive <- resp.TPM.dt.sub.pred[resp.TPM.dt.sub.pred$val == ]
-
+# calculate degree of induction
+Ap.induction <- ddply(resp.TPM.dt.sub.pred, .(colony, Transcript), summarise, 
+                      upreg = (max(pTPM) - min(pTPM))/min(pTPM) * 100 )
+# merge optimum expression, in Ap.response.type data.frame, with predicted TPM data
+Ap.induction <- data.table(Ap.induction)
+setkey(Ap.induction, Transcript)
+Ap.response.type2 <- Ap.induction[Ap.response.type]
+
+# plot constitutive expression vs induction for High, Low and Bimodal genes only
+
+setkey(Ap.response.type2, A22.type)
+Ap.response.type2.up <- Ap.response.type2[A22.type!="Intermediate", ]
+setkey(Ap.response.type2.up, Ar.type)
+Ap.response.type2.up <- Ap.response.type2.up[Ar.type!="Intermediate", ]
+
+gg5 <- ggplot(Ap.response.type2.up, aes(x = log(A22.opt), y = log(upreg), colour = A22.type)) + 
+  geom_point() + 
+  facet_grid(. ~ colony)
+
+print(gg5)
 ```

### 2014-10-17 ApTranscriptome_TR.Rmd
working on figure for genetic assimilation

@@
 # list of transcripts that are 'high' expressed in A22
-A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
+A22.high.transcripts.df <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
 
 # Compare expression at optimum temp (A22.opt) between colonies using t-test
-t.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
-boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
+t.test(A22.high.transcripts.df$A22.opt, A22.high.transcripts.df$Ar.opt)
+boxplot(A22.high.transcripts.df$A22.opt, A22.high.transcripts.df$Ar.opt)
 
 # T test on log-transformed values to control for outliers
-t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
-boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
+t.test(log(A22.high.transcripts.df$A22.opt+1), log(A22.high.transcripts.df$Ar.opt+1))
+boxplot(log(A22.high.transcripts.df$A22.opt+1), log(A22.high.transcripts.df$Ar.opt+1))
 ```
@@
 ```{r A22_high_wilcoxon}
-w1 <- wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+w1 <- wilcox.test(A22.high.transcripts.df$A22.opt, A22.high.transcripts.df$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
 w1
@@
 # list of transcripts that are 'high' expressed in Ar
-Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
+Ar.low.transcripts.df <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
 
 # t-test with log values
-t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
-boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
+t.test(log(Ar.low.transcripts.df$A22.opt+1), log(Ar.low.transcripts.df$Ar.opt+1))
+boxplot(log(Ar.low.transcripts.df$A22.opt+1), log(Ar.low.transcripts.df$Ar.opt+1))
 
 # Wilcoxon signed rank-test
-w2 <- wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
+w2 <- wilcox.test(Ar.low.transcripts.df$A22.opt, Ar.low.transcripts.df$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
 w2
@@
                       upreg = (max(pTPM) - min(pTPM))/min(pTPM) * 100 )
+# cast to wide
+Ap.induction.wide <- dcast(Ap.induction, Transcript ~ colony, value.var = "upreg")
+names(Ap.induction.wide)[names(Ap.induction.wide)=="A22"] <- "A22.upreg"
+names(Ap.induction.wide)[names(Ap.induction.wide)=="Ar"] <- "Ar.upreg"
+
 # merge optimum expression, in Ap.response.type data.frame, with predicted TPM data
-Ap.induction <- data.table(Ap.induction)
-setkey(Ap.induction, Transcript)
-Ap.response.type2 <- Ap.induction[Ap.response.type]
+Ap.induction.wide <- data.table(Ap.induction.wide)
+setkey(Ap.induction.wide, Transcript)
+Ap.response.type2 <- Ap.induction.wide[Ap.response.type]
 
 # plot constitutive expression vs induction for High, Low and Bimodal genes only
-
 setkey(Ap.response.type2, A22.type)
-Ap.response.type2.up <- Ap.response.type2[A22.type!="Intermediate", ]
+up.terms <- c("Bimodal", "Low", "High")
+Ap.response.type2.up <- Ap.response.type2[A22.type %in% up.terms, ]
+
 setkey(Ap.response.type2.up, Ar.type)
-Ap.response.type2.up <- Ap.response.type2.up[Ar.type!="Intermediate", ]
-
-gg5 <- ggplot(Ap.response.type2.up, aes(x = log(A22.opt), y = log(upreg), colour = A22.type)) + 
+Ap.response.type2.up <- Ap.response.type2.up[Ar.type %in% up.terms, ]
+Ap.response.type2.up <- droplevels(Ap.response.type2.up)
+
+gg5 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.opt), y = log(A22.upreg), colour = A22.type)) + 
   geom_point() + 
-  facet_grid(. ~ colony)
-
-print(gg5)
+  geom_smooth() +
+  labs(x="log(consitutive expression of A. carolinensis)", y = "log(Degree of inducibility in A. picea)") +
+  theme(legend.position = "none")
+
+gg6 <- ggplot(Ap.response.type2.up, aes(x = log(A22.opt), y = log(Ar.upreg), colour = Ar.type)) + 
+  geom_point() + 
+  geom_smooth() +
+  labs(x="log(consitutive expression of A. picea)", y = "log(Degree of inducibility in A. carolinensis)") +
+  theme(legend.justification=c(1,0), legend.position=c(1,0), legend.text = element_text(size = 12))
+
+# grid.arrange
+tiff("results/constitutive_vs_inducibility.tiff", compression = "lzw", bg = "white")
+grid.arrange(gg5, gg6, nrow = 1)
+dev.off()
+
+### plot constitutive expression in one species vs the other
+gg7 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.opt), y = log(A22.opt), colour = A22.type)) + 
+  geom_point() + 
+  geom_smooth() +
+  labs(x="log(consitutive expression of A. carolinensis)", y = "log(constitutive expression in A. picea)") +
+  theme(legend.position = "none")  
+gg7
+
+### plot inducibility in one species vs the other
+gg8 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.upreg), y = log(A22.upreg), colour = A22.type)) + 
+  geom_point() + 
+  geom_smooth() +
+  labs(x="log(upregulation in A. carolinensis)", y = "log(upregulation in A. picea)") +
+  theme(legend.position = "none")  
+gg8
+
+### constitutive vs upregulated for *High* genes only
+Ap.response.type2.A22.high <- Ap.response.type2[Ap.response.type2$Transcript %in% A22.high.transcripts.df$Transcript, ]
+str(Ap.response.type2.A22.high)
+
+gg9 <- ggplot(Ap.response.type2.A22.high, aes(x = log(Ar.opt), y = log(A22.opt), colour = Ar.type)) + 
+  geom_point() + 
+  geom_smooth() +
+  labs(x="log(consitutive expression of A. carolinensis)", y = "log(Degree of inducibility in A. picea)") +
+  theme(legend.position = "none")  
+gg9
+
 ```

### 2014-10-17 doc/ApTranscriptome_manuscript.Rmd

@@
 
-Thermal tolerance is a key trait governing the distributions of species and is useful in predicting their responses to climate change.
-To quantitatively evaluate among three hypotheses for the evolution of thermal tolerance, we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
-We characterized the thermal reactionomes of two common ant species in the eastern U.S: the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
+The distributions of species and their responses to climate change are in part determined by their thermal tolerances.
+However, little is known about how thermal tolerance evolves.
+To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
+We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
 We found that at least 8% of all genes were thermally responsive.
-Consistent with the *enhanced defense hypothesis*, the cool-adapted *A.
+The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A.
 picea* induced expression of more genes in response to extreme temperatures than *A.
 carolinensis*, including heat shock proteins.
-In contrast, *A.
-carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than *A.
-picea*.
-This pattern was consistent with the *tolerance hypothesis* at high, but not low, temperatures.
-We found no evidence for the *genetic assimilation hypothesis* as there was no trade-off between constitutive and inducible levels of gene expression.
-Collectively, our results indicate that the northern *A.
-picea* actively resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas the southern *A.
-carolinensis* relies upon other mechanisms to tolerate thermal stress.
-These results are consistent with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
+In contrast, the warm-adapted *A.
+carolinensis* downregulated many of the genes upregulated in *A.
+picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis.
+We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.
+These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
 
@@
 carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes.
-These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with low temperatures and other stressors.
+These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.
 
@@
 
-To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011].
+To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity and plasticity of genes involved in thermal tolerance [@hoffmann2008; @somero2011].
 In recent years, whole organism gene expression approaches (e.g.
@@
 
-```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A.
-carolinensis* (rows) and *A.
-picea* (columns).
-The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
-Blue cells are those with significant excesses, while red cells are those with significant absences."}
-
-# plot
-mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
-  scale_fill_gradient2(limits=c(-600, 600)) 
-mh_plot
-```
-
 ## No evidence for *genetic assimilation*
@@
 
-**ADD PANEL TO FIG2 WITH DOWNREGULATION OF INTERMEDIATE TRANSCRIPTS**
+# Discussion
 
-```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
-Red bars and lines are for *A.
-carolinensis* while blue bars and lines are for *A.
-picea*."}
+The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances.
+However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerances.
+By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
+In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
+carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species.
 
-T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
-    facet_grid(. ~ type) +
-    geom_histogram(position = "dodge", binwidth = 3) +
-    geom_density(alpha = 0.2, adjust = 3) + 
-    scale_y_continuous(name="Density") +
-    scale_x_continuous(name=expression("Temperature of transcript activation"))
-print(T_on_plot)
-```
+Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
+We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A.
+picea* is substantially lower than *A.
+carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig.
+2A).
+Moreover, the transcripts upregulated in *A.
+picea* included nearly the entire set upregulated in *A.
+carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Supplemental...).
+Two non-mutually exclusive hypotheses may explain this pattern.
+First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants and a suite of preparatory physiological modifications [@denlinger2002].
+The northern species *A.
+picea* might need to induce a greater response to survive the longer winter period.
+Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012].
+This requirement may be under stronger selection in *A.
+picea* given the short growing season that would necessitate foraging under broader ranges of temperature (...stuble data...).
 
-# Discussion
+In contrast to cold tolerance, the enhanced upper thermal limit in *A.
+picea* is best explained by the *tolerance* hypothesis.
+High temperatures were associated with significantly fewer upregulated transcripts in *A.
+carolinensis*, and a large proportion (361 of 1,238) of those transcripts upregulated in *A.
+picea* were either downregulated or expressed at negligible levels overall in *A.
+carolinensis*.
+Similarly, signficantly higher temperatures were required to induce the expression response for *A.
+carolinensis* than *A.
+picea* towards the high thermal limit (Fig.
+2B, D), suggesting that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.
+
+The one hypothesis that did not receive unequivocal support was the *genetic assimilation hypothesis*, which predicted that exposure to more frequent stressors is hypothesized to select for a shift from inducible to constitutive expression of stress-response transcripts.
+In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A.
+carolinensis* than *A.
+picea* in nearly all comparisons, with the specific notable exception of **High** transcripts in *A.
+picea* that were **Not Responsive** in *A.
+carolinensis* not because their expression was enhanced, but because they were not expressed at all.
+This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al.
+[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
+Both the genetic assimilation and enhanced defense routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
+Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.
 
@@
 In contrast, the warm-adapted *A.
-carolinensis* relies upon constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress.
-If the *enhanced defense* is hypothesis is correct, candidate genes such as heat shock proteins should be over-represented in the thermally-responsive category for *A.
-picea*.
+carolinensis* relies upon other constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress.
+Although the reactionome approach is designed to identify large-scale response patterns, this difference between species is also evident among candidate genes known to be involved in stress response.
 Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
@@
 In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
-
-Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
-Only 15-20% of transcripts upregulated in response to temp are bimodal (i.e.
-generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical.
-Implications ...
 
@@
 To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012].
+At the same time, critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
+Only 15-20% of transcripts upregulated in response to temperature are bimodal (i.e.
+generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
+2).
+This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.
 
@@
 While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
-Thus, our method provides an example of how RNA-seq data can be used to understand differences among species  ...
+Thus, our method provides an example of how RNA-seq data can be used to understand gain meaningful insight on the process of evolution.
 
@@
 
+Figure 1.
+Matched expression type data for *A.
+carolinensis* (rows) and *A.
+picea* (columns).
+The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
+Blue cells are those with significant excesses, while red cells are those with significant absences.
+
+Figure 2.
+Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
+Red bars and lines are for *A.
+carolinensis* while blue bars and lines are for *A.
+picea*.
+
+Figure 3.
+Scatterplots of (A) constitutive expression in *A.
+picea* against degree of induction for upregulated transcripts in *A.
+carolinensis* and (B) constitutive expression in *A.
+carolinensis* against degree of induction for upregulated transcripts in *A.
+picea.*.
+For each species, the transcripts are split into expression category based on the ...
+The top row is for *High* transcripts, middle is for *Intermediate* transcripts and bottom row is for *Low* transcripts.
+
+# Figures
+
+![Figure 1](matched_observations_Fig1.tiff)
+
+![Figure 2](crit_temp_regulation_Fig2.tiff)
+
+![Figure 3](constitutive_vs_upreg_Fig3.tiff)
+
 # References

### 2014-10-20 doc/ApTranscriptome_manuscript.Rmd
cleaning up Rmd file so it compiles

@@
 
-![Figure 1](matched_observations_Fig1.tiff)
+![Matched expression type data for *A.
+carolinensis* (rows) and *A.
+picea* (columns).
+The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
+Blue cells are those with significant excesses, while red cells are those with significant absences.](../results/matched_observations_Fig1.png)
 
-![Figure 2](crit_temp_regulation_Fig2.tiff)
+![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
+Red bars and lines are for *A.
+carolinensis* while blue bars and lines are for *A.
+picea*.](../results/crit_temp_regulation_Fig2.png)
 
-![Figure 3](constitutive_vs_upreg_Fig3.tiff)
+![Scatterplots of (A) constitutive expression in *A.
+picea* against degree of induction for upregulated transcripts in *A.
+carolinensis* and (B) constitutive expression in *A.
+carolinensis* against degree of induction for upregulated transcripts in *A.
+picea.*.
+For each species, the transcripts are split into expression category based on the ...
+The top row is for *High* transcripts, middle is for *Intermediate* transcripts and bottom row is for *Low* transcripts.](../results/constitutive_vs_upreg_Fig3.png)
 

### 2014-10-20 doc/ApTranscriptome_manuscript.Rmd
more final tweaks

@@
 author: ' '
-csl: ecology.csl
+csl: pnas1.csl
 output:
@@
 To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
-We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*.
+We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* in a heat-shock experiment.
 We found that at least 8% of all genes were thermally responsive.
@@
 We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.
-These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.
+These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.
 
@@
 carolinensis*.
-Surprisingly, we found divergent transcriptomic responses in these species.
+We found substantial differences in their patterns of gene expression.
 *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response.
@@
 Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010].
-Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta_jr2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
+Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
 These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
@@
 *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-*Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A.
+*Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A.
 picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A.
@@
 carolinensis* versus -0.1°C for *A.
-picea*; Supplemental Material) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
+picea*; Table S1) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
 
@@
 
-To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material).
+To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
 
@@
 
-options(digits = 2)
-panderOptions('round', 2)
+options(digits = 3)
+panderOptions('round', 3)
 panderOptions('keep.trailing.zeros', TRUE)
@@
 picea*.
-**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other.")
+**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
 ```
 
-The total number of thermally-responsive transcripts did not differ between species (...stat...).
-Within this set, there was significant qualitative and quantitative variation in thermal responses; both among transcripts and between the two species.
-We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
+Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
+To examine this, we grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
@@
 
-Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Table 1).
-*Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), whereas *A.
-carolinensis* downregulated more transcripts under these conditions (*Intermediate*, ...stats...).
+Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure 1).
+*Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, post-hoc $\chi^2$ = ...stats...), whereas *A.
+carolinensis* downregulated more transcripts under these conditions (*Intermediate*, post-hoc $\chi^2$ = ...stats...).
 Consistent with the *enhanced defense* hypothesis, *A.
 picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A.
-carolinensis* (Table 1, ...
-stats...).
+carolinensis* (Table 1, $\chi^2$ = ...).
 In contrast, at high temperatures, *A.
@@
 Two non-mutually exclusive hypotheses may explain this pattern.
-First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants and a suite of preparatory physiological modifications [@denlinger2002].
+First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002].
 The northern species *A.
@@
 For both species, more transcripts responded to low than high temperatures.
-Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to high than low temperatures.
+Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have more limited genetic accomodation to high than low temperatures.
 Our results also indicate that these two ant species have different responses to thermal extremes.
@@
 communication), and there is about 2% pairwise divergence between these clades using five nuclear genes (Bernice DeMarco, unpublished data).
-Colonies of these species were maintained in the lab for 6 months prior to experimentation.
+Colonies of these species were maintained in common conditions at ??
+for 6 months prior to experimentation.
 

### 2014-10-20 ApTranscriptome_TR.Rmd

@@
 
-# Pearson Chi-square test
-chi1 <- chisq.test(Ap.type.table)
-chi1
-
+# test if the total number of responsive transcripts differs between species 
+resp.table <- matrix(nrow=2, ncol=2, dimnames = list(c("R", "NR"), c("Ap", "Ac")))
+resp.table[1,1] <- sum(A22.type.table[which(names(A22.type.table) != "NotResp")])
+resp.table[2,1] <- nrow(annotation.table) - sum(A22.type.table[which(names(A22.type.table) != "NotResp")])
+resp.table[1,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "NotResp")])
+resp.table[2,2] <- nrow(annotation.table) - sum(Ar.type.table[which(names(Ar.type.table) != "NotResp")])
+resp.table
+
+chi1 <- chisq.test(resp.table)
+
+# test if the number of transcripts in each category differs from null expectations using Pearson Chi-square test
+chi2 <- chisq.test(Ap.type.table)
+chi2
+```
+
+The total number of responsive transcripts does not differ between species, though their groupings into expression categories does differ.
+
+```{r mosaic_plot}
 # Reorganize for plotting to show overlap among categories between colonies
@@
 
-Table with 'Total Transcripts' category
-
-```{r total_table}
-
-```
-
 The question of biological interest is whether the marginal frequencies differ between the two species.
@@
 
-```{r mh_plot}
+To identify specific cells that deviate from null expectation, calculate expected observations assuming that marginal frequencies equal the overall frequency for each expression category, then determine which cells have deviations between observed and expected that are greater than the overall X^2^ statistic.
+
+```{r chi_posthoc}
 # overall mean for each class
@@
 
+# which of these are significantly great than expected chi-squared?
+```
+
+To visualize this, make a heatmap that shows deviation from expected count.
+
+```{r mh_plot}
 md <- melt(Ec.cells)
@@
 # grid.arrange
-tiff("results/constitutive_vs_inducibility.tiff", compression = "lzw", bg = "white")
+png("results/constitutive_vs_inducibility.png")
 grid.arrange(gg5, gg6, nrow = 1)
@@
   theme(legend.position = "none")  
+gg7 <- gg7 + geom_abline(intercept=0, slope=1)
+
+png("results/constitutive_comparison.png")
 gg7
+dev.off()
 
@@
 
+png("results/upregulation_comparison.png")
+gg8
+dev.off()
+
 ### constitutive vs upregulated for *High* genes only
@@
   theme(legend.position = "none")  
+gg9 + geom_abline(intercept=0, slope=1)
+
+png("results/constitutive_vs_inducibility_A22high.png")
 gg9
-
+dev.off()
 ```
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.T_off, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
+save(TPM.dt.sub, RxNpval, sigtable, chi1, chi2, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
 ```

### 2014-10-21 ApTranscriptome_TR.Rmd
added figure to compare log ratio of inducibility to log ratio of baseline expression

@@
 dev.off()
+
+Ar.high.transcripts.df <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), ]
+Ap.response.type2.Ar.high <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.high.transcripts.df$Transcript, ]
+str(Ap.response.type2.Ar.high)
+
+gg10 <- ggplot(Ap.response.type2.Ar.high, aes(x = log(A22.opt), y = log(Ar.opt), colour = A22.type)) + 
+  geom_point() + 
+  geom_smooth() +
+  labs(x="log(consitutive expression of A. picea)", y = "log(Degree of inducibility in A. carolinensis)")
+
+#+
+  #theme(legend.position = "none")
+
+gg10 + geom_abline(intercept=0, slope=1)
+```
+
+```{r log_inducibility}
+str(Ar.high.transcripts.df)
+Ap.response.type2.A22.high <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.high.transcripts.df$Transcript, ]
+str(Ap.response.type2.A22.high)
+
+# exclude transcripts with zero upregulation for denominator
+Ap.response.type2.A22.high <- Ap.response.type2.A22.high[which(Ap.response.type2.A22.high$Ar.upreg != 0), ]
+
+Ap.response.type2.A22.high$A22.Ar.upreg <- Ap.response.type2.A22.high$A22.upreg / Ap.response.type2.A22.high$Ar.upreg
+
+# exclude transcripts with zero ratio of upregulation
+Ap.response.type2.A22.high <- Ap.response.type2.A22.high[which(Ap.response.type2.A22.high$A22.Ar.upreg != 0), ]
+
+# ratio of baseline expression
+Ap.response.type2.A22.high$Ar.A22.baseline <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt
+
+head(Ap.response.type2.A22.high)
+
+# plot!
+gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg+0.01), x = log(Ar.A22.baseline))) + 
+  geom_point() + 
+  geom_smooth(method = "lm") +
+  labs(y="log(inducibility A. picea / inducibility A. car)", x = "log(baseline A. carolinensis / baseline A. picea)")
+
+summary(lm(log(A22.Ar.upreg) ~ log(Ar.A22.baseline), data = Ap.response.type2.A22.high))
+
+cor.test(log(Ap.response.type2.A22.high$A22.Ar.upreg), log(Ap.response.type2.A22.high$Ar.A22.baseline))
+
+### Repeat for A. carolinensis *Low* transcripts
+
+str(Ar.low.transcripts.df)
+Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
+str(Ap.response.type2.Ar.low)
+
+# ratio of upregulation
+Ap.response.type2.Ar.low$Ar.A22.upreg <- Ap.response.type2.Ar.low$Ar.upreg / Ap.response.type2.Ar.low$A22.upreg
+
+# exclude transcripts with zero ratio of upregulation
+Ap.response.type2.Ar.low <- Ap.response.type2.Ar.low[which(Ap.response.type2.Ar.low$Ar.A22.upreg != 0), ]
+
+Ap.response.type2.Ar.low$A22.Ar.baseline <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt
+
+head(Ap.response.type2.Ar.low)
+
+gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.baseline))) + 
+  geom_point() + 
+  geom_smooth(method = "lm") +
+  labs(y="log(inducibility A. car / inducibility A. picea)", x = "log(baseline A. picea / baseline A. car)")
+
+gg12
 ```

### 2014-10-21 ApTranscriptome_TR.Rmd
modifing figures for manuscript

--- 2014-10-20 ApTranscriptome_TR.Rmd
@@
 
+# make subplot of rows that match hypotheses
+
+# A. carolinensis **Low** and **Bimodal**
+md_sub1 <- subset(md, Acar %in% c("Low", "Bimodal"), )
+md_sub1 <- droplevels(md_sub1)
+
+mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
+  scale_fill_gradient2(limits=c(-600, 600)) +
+  theme(axis.title = element_text(face="italic")) +
+  labs(x = "A. picea", y = "A. carolinensis") +
+  theme(legend.position = "none")
+mh_plot_sub1
+
+md_sub2 <- subset(md, Apic %in% c("High", "Bimodal"), )
+md_sub2 <- droplevels(md_sub2)
+
+mh_plot_sub2 <- qplot(x=Apic, y=Acar, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Acar))) + 
+  scale_fill_gradient2(limits=c(-600, 600)) +
+  theme(axis.title = element_text(face="italic")) +
+  labs(x = "A. picea", y = "A. carolinensis") + 
+  theme(legend.position = "none")
+mh_plot_sub2
+
+# get legend from http://stackoverflow.com/questions/11883844/inserting-a-table-under-the-legend-in-a-ggplot2-histogram
+#Extract Legend
+g_legend <- function(a.gplot){
+  tmp <- ggplot_gtable(ggplot_build(a.gplot))
+  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+  legend <- tmp$grobs[[leg]]
+  return(legend)}
+
+legend <- g_legend(mh_plot)
+
+# remove legend from mh_plot
+mh_plot <- mh_plot + theme(legend.position = "none")
+
+# add bars
+geom_rect(data = dark.subset, aes(xmin = phase_start, xmax = phase_end), 
+              ymin = -Inf,
+              xmax = Inf,
+              fill = alpha("#000000", 0.2))
+
+# arrange plots
+
+grid.arrange(mh_plot, mh_plot_sub2, mh_plot_sub1, legend,
+             ncol=2, widths=c(1,0.5), heights=c(1, 0.5))
+
+arrangeGrob(mh_plot, mh_plot_sub1, mh_plot_sub2, ncol=1), 
+                    ncol=2, widths=c(1,1.2))
+
 tiff("results/matched_observations_Fig1.tiff", compression = "lzw")
@@
 
-### Comparison of the degree of upregulation against constitutive expression
-
-To test the genetic assimilation hypothesis, I compare the degree of upregulation of a transcript against constitutive expression at optimal temperatures.
+To visualize this data, I plot the log ratio of inducibility between species against the log ratio of baseline expression between species.
+I do this separately for **High** transcripts in *A.
+picea* and then **Low** transcripts in *A.
+carolinensis*.
+
+First, I calculate the degree of expression induction as the proportion difference between the maximum and minimum expression levels for each transcript.
 
@@
 Ap.response.type2 <- Ap.induction.wide[Ap.response.type]
-
-# plot constitutive expression vs induction for High, Low and Bimodal genes only
-setkey(Ap.response.type2, A22.type)
-up.terms <- c("Bimodal", "Low", "High")
-Ap.response.type2.up <- Ap.response.type2[A22.type %in% up.terms, ]
-
-setkey(Ap.response.type2.up, Ar.type)
-Ap.response.type2.up <- Ap.response.type2.up[Ar.type %in% up.terms, ]
-Ap.response.type2.up <- droplevels(Ap.response.type2.up)
-
-gg5 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.opt), y = log(A22.upreg), colour = A22.type)) + 
-  geom_point() + 
-  geom_smooth() +
-  labs(x="log(consitutive expression of A. carolinensis)", y = "log(Degree of inducibility in A. picea)") +
-  theme(legend.position = "none")
-
-gg6 <- ggplot(Ap.response.type2.up, aes(x = log(A22.opt), y = log(Ar.upreg), colour = Ar.type)) + 
-  geom_point() + 
-  geom_smooth() +
-  labs(x="log(consitutive expression of A. picea)", y = "log(Degree of inducibility in A. carolinensis)") +
-  theme(legend.justification=c(1,0), legend.position=c(1,0), legend.text = element_text(size = 12))
-
-# grid.arrange
-png("results/constitutive_vs_inducibility.png")
-grid.arrange(gg5, gg6, nrow = 1)
-dev.off()
-
-### plot constitutive expression in one species vs the other
-gg7 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.opt), y = log(A22.opt), colour = A22.type)) + 
-  geom_point() + 
-  geom_smooth() +
-  labs(x="log(consitutive expression of A. carolinensis)", y = "log(constitutive expression in A. picea)") +
-  theme(legend.position = "none")  
-gg7 <- gg7 + geom_abline(intercept=0, slope=1)
-
-png("results/constitutive_comparison.png")
-gg7
-dev.off()
-
-### plot inducibility in one species vs the other
-gg8 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.upreg), y = log(A22.upreg), colour = A22.type)) + 
-  geom_point() + 
-  geom_smooth() +
-  labs(x="log(upregulation in A. carolinensis)", y = "log(upregulation in A. picea)") +
-  theme(legend.position = "none")  
-gg8
-
-png("results/upregulation_comparison.png")
-gg8
-dev.off()
-
-### constitutive vs upregulated for *High* genes only
+```
+
+I extract the **High** transcripts for *A.
+picea* and plot the log ratio of inducibility in *A.
+picea* over inducibility in *A.
+carolinensis*, against the log ratio of baseline expression in *A.
+carolinensis* against expression in *A.
+picea*.
+According to the *genetic assimilation hypothesis* there should be a positive relationship between these ratios.
+That is, greater baseline expression in *A.
+carolinensis* should have greater inducibility towards **High** temperatures in *A.
+picea*.
+
+```{r log_inducibility}
 Ap.response.type2.A22.high <- Ap.response.type2[Ap.response.type2$Transcript %in% A22.high.transcripts.df$Transcript, ]
@@
 
-gg9 <- ggplot(Ap.response.type2.A22.high, aes(x = log(Ar.opt), y = log(A22.opt), colour = Ar.type)) + 
+# filter out NotResp transcripts in Ar
+Ap.response.type2.A22.high <- Ap.response.type2.A22.high[which(Ap.response.type2.A22.high$Ar.type != "NotResp"), ]
+
+# ratio of upregulation
+Ap.response.type2.A22.high$A22.Ar.upreg <- Ap.response.type2.A22.high$A22.upreg / Ap.response.type2.A22.high$Ar.upreg
+
+# ratio of baseline expression
+Ap.response.type2.A22.high$Ar.A22.baseline <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt
+
+# plot!
+gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg+0.01), x = log(Ar.A22.baseline))) + 
   geom_point() + 
-  geom_smooth() +
-  labs(x="log(consitutive expression of A. carolinensis)", y = "log(Degree of inducibility in A. picea)") +
-  theme(legend.position = "none")  
-gg9 + geom_abline(intercept=0, slope=1)
-
-png("results/constitutive_vs_inducibility_A22high.png")
-gg9
+  geom_smooth(method = "lm") +
+  labs(y = expression(paste("log(inducibility ", italic("A. picea / A. carolinensis"), ")")),
+       x = expression(paste("log(baseline ", italic("A. carolinensis / A. picea"), ")")))    
+gg11
+
+A22.high.lm <- lm(log(A22.Ar.upreg) ~ log(Ar.A22.baseline), data = Ap.response.type2.A22.high)
+
+summary(A22.high.lm)
+```
+
+In contrast to the prediction there is a negative relationship between these ratios.
+That is, genes with greater upregulation of expression also have greater baseline expression.
+This is also consistent with the results simply comparing baseline expression between species for the *A.
+picea* **High** transcripts.
+
+I repeated the above but for **Low** transcripts in *A. carolinensis*.
+
+```{r assimilation_Ar_low}
+### Repeat for A. carolinensis *Low* transcripts
+
+Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
+str(Ap.response.type2.Ar.low)
+
+# filter out NotResp transcripts in A22
+Ap.response.type2.Ar.low <- Ap.response.type2.Ar.low[which(Ap.response.type2.Ar.low$A22.type != "NotResp"), ]
+
+# ratio of upregulation
+Ap.response.type2.Ar.low$Ar.A22.upreg <- Ap.response.type2.Ar.low$Ar.upreg / Ap.response.type2.Ar.low$A22.upreg
+
+# ratio of baseline expression
+Ap.response.type2.Ar.low$A22.Ar.baseline <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt
+
+gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.baseline))) + 
+  geom_point() + 
+  geom_smooth(method = "lm") +
+  labs(y = expression(paste("log(inducibility ", italic("A. carolinensis / A. picea"), ")")),
+       x = expression(paste("log(baseline ", italic("A. picea / A. carolinensis"), ")")))
+
+gg12
+
+Ar.low.lm <- lm(log(Ar.A22.upreg) ~ log(A22.Ar.baseline), data = Ap.response.type2.Ar.low)
+
+summary(Ar.low.lm)
+```
+
+I combine the two above figures into a single figure.
+
+```{r fig3}
+# grid.arrange
+png(file = "results/genetic_assimilation.png")
+grid.arrange(gg11, gg12, nrow = 1)
 dev.off()
 ```
+
+As with the previous comparison, I also find no evidence for the genetic assimilation hypothesis.
 

### 2014-10-22 ApTranscriptome_TR.Rmd
updated genetic assimilation analysis and figure

@@
 These colonies were maintained in the lab for 6 months prior to sample collection.
-Bernice Bacon DeMarco (Michigan State University) identified colony *A22* as *A.
+Bernice DeMarco (Michigan State University) identified colony *A22* as *A.
 picea* and *Ar* as *A.
 carolinensis*.
-For historical reasons, I refer to these species as colonies at times throughout this technical report.
+For historical reasons, I sometimes refer to these species as colonies at times throughout this technical report.
 
@@
 
+# test if the number of transcripts in each category differs from null expectations using Pearson Chi-square test
+(chi1 <- chisq.test(Ap.type.table))
+
 # test if the total number of responsive transcripts differs between species 
@@
 
-chi1 <- chisq.test(resp.table)
-
-# test if the number of transcripts in each category differs from null expectations using Pearson Chi-square test
-chi2 <- chisq.test(Ap.type.table)
-chi2
+(chi2 <- chisq.test(resp.table))
+
+# test if the number of Low responsive transcripts differs between species 
+low.table <- matrix(nrow=2, ncol=2, dimnames = list(c("Low", "NotLow"), c("Ap", "Ac")))
+low.table[1,1] <- A22.type.table[which(names(A22.type.table) == "Low")]
+low.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "Low")])
+low.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "Low")]
+low.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "Low")])
+low.table
+
+(chi3 <- chisq.test(low.table))
+
+# test if the number of Intermediate responsive transcripts differs between species 
+intermediate.table <- matrix(nrow=2, ncol=2, dimnames = list(c("Intermediate", "Others"), c("Ap", "Ac")))
+intermediate.table[1,1] <- A22.type.table[which(names(A22.type.table) == "Intermediate")]
+intermediate.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "Intermediate")])
+intermediate.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "Intermediate")]
+intermediate.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "Intermediate")])
+intermediate.table
+
+(chi4 <- chisq.test(intermediate.table))
+
+# test if the number of High responsive transcripts differs between species 
+high.table <- matrix(nrow=2, ncol=2, dimnames = list(c("High", "Others"), c("Ap", "Ac")))
+high.table[1,1] <- A22.type.table[which(names(A22.type.table) == "High")]
+high.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "High")])
+high.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "High")]
+high.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "High")])
+high.table
+
+(chi5 <- chisq.test(high.table))
+
+# test if the number of Bimodal responsive transcripts differs between species 
+bimodal.table <- matrix(nrow=2, ncol=2, dimnames = list(c("Bimodal", "Others"), c("Ap", "Ac")))
+bimodal.table[1,1] <- A22.type.table[which(names(A22.type.table) == "Bimodal")]
+bimodal.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "Bimodal")])
+bimodal.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "Bimodal")]
+bimodal.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "Bimodal")])
+bimodal.table
+
+(chi6 <- chisq.test(bimodal.table))
 ```
@@
 # A. carolinensis **Low** and **Bimodal**
-md_sub1 <- subset(md, Acar %in% c("Low", "Bimodal"), )
+md_sub1 <- subset(md, Acar %in% c("Low", "Bimodal", "High"), )
 md_sub1 <- droplevels(md_sub1)
@@
   labs(x = "A. picea", y = "A. carolinensis") +
-  theme(legend.position = "none")
+  theme(legend.position = "bottom") + 
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 3.5, xmax = 4.5, ymin = .5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+
 mh_plot_sub1
 
-md_sub2 <- subset(md, Apic %in% c("High", "Bimodal"), )
+md_sub2 <- subset(md, Apic %in% c("Low", "High", "Bimodal"), )
 md_sub2 <- droplevels(md_sub2)
 
-mh_plot_sub2 <- qplot(x=Apic, y=Acar, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Acar))) + 
+mh_plot_sub2 <- qplot(x=Acar, y=Apic, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Apic))) + 
   scale_fill_gradient2(limits=c(-600, 600)) +
   theme(axis.title = element_text(face="italic")) +
-  labs(x = "A. picea", y = "A. carolinensis") + 
-  theme(legend.position = "none")
+  labs(x = "A. carolinensis", y = "A. picea") + 
+  theme(legend.position = "none") + 
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+
 mh_plot_sub2
 
-# get legend from http://stackoverflow.com/questions/11883844/inserting-a-table-under-the-legend-in-a-ggplot2-histogram
-#Extract Legend
-g_legend <- function(a.gplot){
-  tmp <- ggplot_gtable(ggplot_build(a.gplot))
-  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
-  legend <- tmp$grobs[[leg]]
-  return(legend)}
-
-legend <- g_legend(mh_plot)
-
-# remove legend from mh_plot
-mh_plot <- mh_plot + theme(legend.position = "none")
-
-# add bars
-geom_rect(data = dark.subset, aes(xmin = phase_start, xmax = phase_end), 
-              ymin = -Inf,
-              xmax = Inf,
-              fill = alpha("#000000", 0.2))
-
-# arrange plots
-
-grid.arrange(mh_plot, mh_plot_sub2, mh_plot_sub1, legend,
-             ncol=2, widths=c(1,0.5), heights=c(1, 0.5))
-
-arrangeGrob(mh_plot, mh_plot_sub1, mh_plot_sub2, ncol=1), 
-                    ncol=2, widths=c(1,1.2))
-
-tiff("results/matched_observations_Fig1.tiff", compression = "lzw")
+# arrange plots to make figure for manuscript
+mh_plot_sub2 <- mh_plot_sub2 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
+mh_plot_sub1 <- mh_plot_sub1 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
+
+png("results/matched_observations_Fig1.png")
+grid.arrange(mh_plot_sub2, mh_plot_sub1, ncol=1, heights = c(0.45, 0.6))
+dev.off()
+
+# full figure for supplemental
+png("results/matched_observations_FigS1.png")
 mh_plot
@@
 An interesting observation from the matched observations plot (Fig.
-...) is that for **NotResp** genes in *A.
+1) is that for **NotResp** genes in *A.
 carolinensis* the **High** and **Bimodal** categories in *A.
@@
 
-To visualize this data, I plot the log ratio of inducibility between species against the log ratio of baseline expression between species.
+To visualize this data, I plot the log ratio of inducibility between species against the log ratio of constitutive expression between species.
 I do this separately for **High** transcripts in *A.
@@
 picea* over inducibility in *A.
-carolinensis*, against the log ratio of baseline expression in *A.
+carolinensis*, against the log ratio of constitutive expression in *A.
 carolinensis* against expression in *A.
@@
 According to the *genetic assimilation hypothesis* there should be a positive relationship between these ratios.
-That is, greater baseline expression in *A.
+That is, greater constitutive expression in *A.
 carolinensis* should have greater inducibility towards **High** temperatures in *A.
@@
 
-# ratio of baseline expression
-Ap.response.type2.A22.high$Ar.A22.baseline <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt
+# ratio of constitutive expression
+Ap.response.type2.A22.high$Ar.A22.constitutive <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt
 
 # plot!
-gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg+0.01), x = log(Ar.A22.baseline))) + 
+gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg), x = log(Ar.A22.constitutive))) + 
   geom_point() + 
@@
   labs(y = expression(paste("log(inducibility ", italic("A. picea / A. carolinensis"), ")")),
-       x = expression(paste("log(baseline ", italic("A. carolinensis / A. picea"), ")")))    
+       x = expression(paste("log(constitutive ", italic("A. carolinensis / A. picea"), ")")))
+
 gg11
 
-A22.high.lm <- lm(log(A22.Ar.upreg) ~ log(Ar.A22.baseline), data = Ap.response.type2.A22.high)
+A22.high.lm <- lm(log(A22.Ar.upreg) ~ log(Ar.A22.constitutive), data = Ap.response.type2.A22.high)
 
@@
 In contrast to the prediction there is a negative relationship between these ratios.
-That is, genes with greater upregulation of expression also have greater baseline expression.
-This is also consistent with the results simply comparing baseline expression between species for the *A.
+That is, genes with greater upregulation of expression also have greater constitutive expression.
+This is also consistent with the results simply comparing constitutive expression between species for the *A.
 picea* **High** transcripts.
@@
 ### Repeat for A. carolinensis *Low* transcripts
-
 Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
@@
 
-# ratio of baseline expression
-Ap.response.type2.Ar.low$A22.Ar.baseline <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt
-
-gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.baseline))) + 
+# ratio of constitutive expression
+Ap.response.type2.Ar.low$A22.Ar.constitutive <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt
+
+gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.constitutive))) + 
   geom_point() + 
@@
   labs(y = expression(paste("log(inducibility ", italic("A. carolinensis / A. picea"), ")")),
-       x = expression(paste("log(baseline ", italic("A. picea / A. carolinensis"), ")")))
+       x = expression(paste("log(constitutive ", italic("A. picea / A. carolinensis"), ")")))
 
@@
 
-Ar.low.lm <- lm(log(Ar.A22.upreg) ~ log(A22.Ar.baseline), data = Ap.response.type2.Ar.low)
+Ar.low.lm <- lm(log(Ar.A22.upreg) ~ log(A22.Ar.constitutive), data = Ap.response.type2.Ar.low)
 
@@
 
-```{r fig3}
+```{r fig2}
+# add labels
+gg11 <- gg11 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
+gg12 <- gg12 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
 # grid.arrange
-png(file = "results/genetic_assimilation.png")
+png(file = "results/genetic_assimilation_Fig2.png")
 grid.arrange(gg11, gg12, nrow = 1)
@@
 As with genes turned on at high temperatures, *A.
-carolinensis* **Intermediate** genes are down-regulated on average at 2.7°C higher temperatures than *A.
+carolinensis* **Intermediate** genes are down-regulated on at higher temperatures than *A.
 picea* **Intermediate** genes.
 In contrast, towards low temperatures, the temperature of mean temperature of downregulation was higher for *A.
-picea* (10.6°) than *A.
-carolinensis* (9.75°).
+picea* than *A.
+carolinensis*.
 
@@
 # grid.arrange
-#tiff("results/crit_temp_regulation_Fig2.tiff", compression = "lzw", bg = "white")
-grid.arrange(gg2, gg3, nrow = 2)
-#dev.off()
+gg2 <- gg2 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
+gg3 <- gg3 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
+
+png("results/crit_temp_regulation_Fig3.png")
+grid.arrange(gg2, gg3, nrow = 2, heights = c(.5,.6))
+dev.off()
 ```

### 2014-10-22 doc/ApTranscriptome_manuscript.Rmd
updates to ms

@@
 carolinensis* versus 41.4°C for *A.
-picea*; Supplemental Material) and minimum temperatures (6.1°C for *A.
+picea*; see Methods) and minimum temperatures (6.1°C for *A.
 carolinensis* versus -0.1°C for *A.
-picea*; Table S1) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
+picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
 
@@
 #options(scipen = 1, digits = 3)
-load("ms_2014-10-08.RData")
+load("ms_slim_2014-10-22.RData")
 library(data.table)
@@
 Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
-Of these 9,052 transcripts, 80% (`r prettyNum(length(quadratic.lms), big.mark = ",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
+Of these 9,052 transcripts, 80% (...) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
 high vs.
@@
 
-Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure 1).
-*Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, post-hoc $\chi^2$ = ...stats...), whereas *A.
-carolinensis* downregulated more transcripts under these conditions (*Intermediate*, post-hoc $\chi^2$ = ...stats...).
+The two species differed in the number of transcripts in each response category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
+*Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A.
+carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`).
 Consistent with the *enhanced defense* hypothesis, *A.
-picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A.
-carolinensis* (Table 1, $\chi^2$ = ...).
-In contrast, at high temperatures, *A.
-carolinensis* matched the predictions of the *tolerance hypothesis*, inducing expression of ~14% fewer transcripts than *A.
-picea* (Table 1).
+picea* induced 1,146 (30%) more transcripts in response to low temperatures than *A.
+carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`).
+However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, at high temperatures *A.
+carolinensis* induced expression of ~14% fewer transcripts than *A.
+picea* (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* < `r max(chi5$p.value, 0.001)`).
+
+```{r echo=FALSE, results='hide'}
+# results of Stuart-Maxwell test
+mh.test
+```
 
@@
 Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another.
-These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes.
+These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure S1) indicating that these differences are not simply due to neutral processes.
 **Intermediate** transcripts in *A.
 carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A.
-picea* than expected by chance (...stat...), whereas there was a significant deficit of the reverse shift of *A.
+picea* than expected by chance (Fig.
+1A), whereas there was a significant deficit of the reverse shift of *A.
 picea* **Intermediate** transcripts to any upregulated category in *A.
 carolinensis* (Fig.
-1).
+1B).
 Similarly, both **High** and **Bimodal** transcripts in *A.
@@
 
+![Deviations from expected numbers of transcripts in matched observation data for transcript expression type data.
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
+Shown are **Low**, **High** and **Bimodal** expression types for (A) *A.
+picea** and (B) *A.
+carolinensis* against all expression types in the opposite species.
+The dashed boxes indicate the diagonals (e.g.
+High - High) which are positive](../results/matched_observations_Fig1.png)
+
 The molecular processes in the set of transcripts that switched from down-regulation in *A.
 carolinensis* to up-regulation in *A.
-picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ...
-were related to metabolism.
+picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
 In addition, there were GO terms related to repair (e.g.
-"histone modification"), transcript regulation (e.g.
-"positive regulation of cellular process") and cellular organization (e.g.
-"transport", "organelle organization").
+histone modification), transcript regulation (e.g.
+positive regulation of cellular process) and cellular organization (e.g.
+transport, organelle organization).
 Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A.
@@
 
-We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions.
-Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A.
-picea* for transcripts that were responsive to high temperature in *A.
-picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`).
-However, the reverse pattern of higher constitutive expression in *A.
-picea* was not observed for low temperature responsive transcripts in *A.
-carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`).
+We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed higher constitutive expression, here defined as expression at 19.5°C, in the species better adapted to these conditions.
+This would be evident as a positive relationship between the ratio of inducibility between species against the opposite ratio of constitutive expression between species.
+However, for both high temperature activated transcripts in *A.
+picea* (Fig.
+2A) and low temperature activated transcripts in *A.
+carolinensis* (Fig.
+2B) there was no evidence for greater constitutive expression for transcripts with lower inducibility.
+In fact, for both comparisons there was the opposite relationship as predicted by the *genetic assimilation hypothesis*, with greater inducibility for transcripts with lower constitutive expression in the opposite species.
 
-Instead, the overall pattern is that thermally responsive transcripts in *A.
+![Scatterplots of (A) log ratio of inducibility in *A.
+picea* over inducibility in *A.
+carolinensis* against log ration of constitutive expression in *A.
+carolinensis* over constitutive expression in *A.
+picae* for transcripts with **High** expression in *A.
+picea*, and (B) log ratio of inducibility in *A.
+carolinensis* over inducibility in *A.
+picea* against log ration of constitutive expression in *A.
+picea* over constitutive expression in *A.
+carolinensis*.
+In both cases, a positive relationship would be Lines are ordinary least squares regressions on the data.](../results/genetic_assimilation_Fig2.png)
+
+In addition, the thermally responsive transcripts in *A.
 carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A.
@@
 1).
-Rather than having constitutively elevated expression as predicted, these genes were uniformly not expressed (...mean TPM...) in *A.
+Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A.
 carolinensis*.
-
-...GO terms **High** and **Bimodal** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* ...
 
@@
 2).
+
+![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
+Red bars and lines are for *A.
+carolinensis* while blue bars and lines are for *A.
+picea*.](../results/crit_temp_regulation_Fig3.png)
 
@@
 picea* included nearly the entire set upregulated in *A.
-carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Supplemental...).
+carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1).
 Two non-mutually exclusive hypotheses may explain this pattern.
@@
 This requirement may be under stronger selection in *A.
-picea* given the short growing season that would necessitate foraging under broader ranges of temperature (...stuble data...).
+picea* given the short growing season that would necessitate foraging under a broader range of temperatures.
 
@@
 Thus, our method provides an example of how RNA-seq data can be used to understand gain meaningful insight on the process of evolution.
+
+One possible explanation for the differences between species in their thermal reactionomes is neutral drift of gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
+While we cannot rule this possibility out, two lines of evidence make us think that it is unlikely.
+First, if gene expression differed by drift between species, we would not expect to see any striking differences among expression type classes as we do here (Fig.
+1, Fig.
+S1).
+Second, even if these differences were due to drift, it would be unlikely that there would be transcriptome-wide differences in the critical temperatures of gene activation such as we see here (Fig.
+2).
 
@@
 
-![Matched expression type data for *A.
-carolinensis* (rows) and *A.
-picea* (columns).
-The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
-Blue cells are those with significant excesses, while red cells are those with significant absences.](../results/matched_observations_Fig1.png)
-
-![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
-Red bars and lines are for *A.
-carolinensis* while blue bars and lines are for *A.
-picea*.](../results/crit_temp_regulation_Fig2.png)
-
-![Scatterplots of (A) constitutive expression in *A.
-picea* against degree of induction for upregulated transcripts in *A.
-carolinensis* and (B) constitutive expression in *A.
-carolinensis* against degree of induction for upregulated transcripts in *A.
-picea.*.
-For each species, the transcripts are split into expression category based on the ...
-The top row is for *High* transcripts, middle is for *Intermediate* transcripts and bottom row is for *Low* transcripts.](../results/constitutive_vs_upreg_Fig3.png)
-
 # References

### 2014-10-23 doc/ApTranscriptome_manuscript.Rmd

@@
 However, little is known about how thermal tolerance evolves.
-To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
-We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* in a heat-shock experiment.
+To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
+We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* across 12 temperature intervals that spanned their entire thermal breadth.
 We found that at least 8% of all genes were thermally responsive.
@@
 carolinensis*, including heat shock proteins.
-In contrast, the warm-adapted *A.
+In contrast, under high temperatures the warm-adapted *A.
 carolinensis* downregulated many of the genes upregulated in *A.
@@
 We found substantial differences in their patterns of gene expression.
-*Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response.
+*Aphaenogaster picea* responded to extreme temperatures primarily through upregulation, with a robust heat shock response.
 In contrast, *A.
 carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes.
-These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.
+These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a qualitative shift in mechanisms from those used to cope with rarely encountered stressors.
 
@@
 carolinensis*.
-Second, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes.
+Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014].
+This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways.
+Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.
 This hypothesis predicts that transcripts responsive to high temperatures in *A.
@@
 picea*.
-Finally, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014].
-This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways.
 Of course, these hypotheses are not mutually exclusive.
-By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.
+By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation.
 
@@
 picea*.
-**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
+**Low** are transcripts with increased expression at low temperatures (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
 ```
@@
 Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another.
-These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure S1) indicating that these differences are not simply due to neutral processes.
-**Intermediate** transcripts in *A.
-carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A.
-picea* than expected by chance (Fig.
-1A), whereas there was a significant deficit of the reverse shift of *A.
-picea* **Intermediate** transcripts to any upregulated category in *A.
+These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Figure S1) indicating that these differences are not simply due to neutral processes.
+The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species.
+The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.
+For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A.
+carolinensis* for low temperatures, *A.
+picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species.
+As with the overall results, the patterns of association differed for the two temperature extremes.
+Genes upregulated at low temperatures in *A.
+carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A.
+picea*, suggesting shared response pathways (Fig.
+1A).
+In contrast, genes upregulated in response to high temperatures in *A.
+picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A.
 carolinensis* (Fig.
 1B).
-Similarly, both **High** and **Bimodal** transcripts in *A.
-picea* were more likely to be **Not Responsive** in *A.
-carolinensis* than expected, while the reverse was not significant (Fig.
-1).
+Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig.
+1B).
 
@@
 Shown are **Low**, **High** and **Bimodal** expression types for (A) *A.
-picea** and (B) *A.
+picea* and (B) *A.
 carolinensis* against all expression types in the opposite species.
 The dashed boxes indicate the diagonals (e.g.
-High - High) which are positive](../results/matched_observations_Fig1.png)
+**High** - **High**).](../results/matched_observations_Fig1.png)
 
-The molecular processes in the set of transcripts that switched from down-regulation in *A.
-carolinensis* to up-regulation in *A.
-picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
+The molecular processes in the set of transcripts that switched from upregulation in *A.
+picea* to downregulation in *A.
+carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
 In addition, there were GO terms related to repair (e.g.
@@
 picea*, increases metabolism and initiates damage repair responses.
-
-## No evidence for *genetic assimilation*
-
-We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed higher constitutive expression, here defined as expression at 19.5°C, in the species better adapted to these conditions.
-This would be evident as a positive relationship between the ratio of inducibility between species against the opposite ratio of constitutive expression between species.
-However, for both high temperature activated transcripts in *A.
-picea* (Fig.
-2A) and low temperature activated transcripts in *A.
-carolinensis* (Fig.
-2B) there was no evidence for greater constitutive expression for transcripts with lower inducibility.
-In fact, for both comparisons there was the opposite relationship as predicted by the *genetic assimilation hypothesis*, with greater inducibility for transcripts with lower constitutive expression in the opposite species.
-
-![Scatterplots of (A) log ratio of inducibility in *A.
-picea* over inducibility in *A.
-carolinensis* against log ration of constitutive expression in *A.
-carolinensis* over constitutive expression in *A.
-picae* for transcripts with **High** expression in *A.
-picea*, and (B) log ratio of inducibility in *A.
-carolinensis* over inducibility in *A.
-picea* against log ration of constitutive expression in *A.
-picea* over constitutive expression in *A.
-carolinensis*.
-In both cases, a positive relationship would be Lines are ordinary least squares regressions on the data.](../results/genetic_assimilation_Fig2.png)
-
-In addition, the thermally responsive transcripts in *A.
-carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A.
-picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`).
-An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A.
-picea* but were not thermally responsive in *A.
-carolinensis* (blue cells in bottom row of Fig.
-1).
-Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A.
-carolinensis*.
 
@@
 carolinensis* while blue bars and lines are for *A.
-picea*.](../results/crit_temp_regulation_Fig3.png)
+picea*.](../results/crit_temp_regulation_Fig2.png)
 
@@
 
+## No evidence for *genetic assimilation*
+
+We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by comparing the log ratios of relative inducibility to relative constitutive expression.
+Specifically, this hypothesis predicted a negative relationship between the two.
+We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A.
+picea* (Fig.
+3A) or transcripts upregulated at low temperatures in *A.
+carolinensis* (Fig.
+3B).
+In fact, for both comparisons we found a weakly positive relationship.
+
+![Scatterplots of (A) log ratio of inducibility in *A.
+picea* over inducibility in *A.
+carolinensis* against log ration of constitutive expression in *A.
+carolinensis* over constitutive expression in *A.
+picea* for transcripts with **High** expression in *A.
+picea*, and (B) log ratio of inducibility in *A.
+carolinensis* over inducibility in *A.
+picea* against log ration of constitutive expression in *A.
+picea* over constitutive expression in *A.
+carolinensis*.
+Blue lines and confidence intervals are from are ordinary least squares regressions.](../results/genetic_assimilation_Fig3.png)
+
+In addition, the thermally responsive transcripts in *A.
+carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A.
+picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`).
+An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A.
+picea* but were not thermally responsive in *A.
+carolinensis* (blue cells in bottom row of Fig.
+1).
+Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A.
+carolinensis*.
+
 # Discussion
@@
 The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances.
-However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerances.
-By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient.
-In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A.
-carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species.
+However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance.
+By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify evolutionary changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A.
+carolinensis*.
+We found large-scale, non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.
 
 Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
-We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A.
+We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A.
 picea* is substantially lower than *A.
@@
 The northern species *A.
-picea* might need to induce a greater response to survive the longer winter period.
+picea* may induce a greater response to survive the longer winter period.
 Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012].
 This requirement may be under stronger selection in *A.
-picea* given the short growing season that would necessitate foraging under a broader range of temperatures.
+picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.
 
@@
 High temperatures were associated with significantly fewer upregulated transcripts in *A.
-carolinensis*, and a large proportion (361 of 1,238) of those transcripts upregulated in *A.
+carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A.
 picea* were either downregulated or expressed at negligible levels overall in *A.
 carolinensis*.
-Similarly, signficantly higher temperatures were required to induce the expression response for *A.
+Similarly, significantly higher temperatures were required to induce the expression response for *A.
 carolinensis* than *A.
 picea* towards the high thermal limit (Fig.
-2B, D), suggesting that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.
-
-The one hypothesis that did not receive unequivocal support was the *genetic assimilation hypothesis*, which predicted that exposure to more frequent stressors is hypothesized to select for a shift from inducible to constitutive expression of stress-response transcripts.
-In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A.
-carolinensis* than *A.
-picea* in nearly all comparisons, with the specific notable exception of **High** transcripts in *A.
-picea* that were **Not Responsive** in *A.
-carolinensis* not because their expression was enhanced, but because they were not expressed at all.
-This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al.
-[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
-Both the genetic assimilation and enhanced defense routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
-Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.
-
-Collectively, our results indicate that these two species have distinct mechanisms of response to thermal extremes.
-The cool-adapated species *Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] with elevated gene expression at both low and high temperatures, which has been shown to impose a metabolic cost [@hoekstra2013].
-In contrast, the warm-adapted *A.
-carolinensis* relies upon other constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress.
-Although the reactionome approach is designed to identify large-scale response patterns, this difference between species is also evident among candidate genes known to be involved in stress response.
+2B-D).
+This difference between species is also evident among candidate genes known to be involved in stress response.
 Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
@@
 carolinensis*.
-These differences are in line with the expectations if *A.
-picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A.
-carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions.
+Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include constitutive defenses [@barshis2013],  enhanced proteome stability [@ghosh2010] or behavioral changes [@sunday2014] to tolerate thermal stress.
+These differences are in line with expectations that *A.
+carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions.
 In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
 
+The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts.
+This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al.
+[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
+Both the genetic assimilation and tolerance routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
+Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.
+
 Given the differences in the patterns of thermal responsiveness between species (Fig.
-2), it is worth noting the similarity in expression patterns between species.
-In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Fig.
-3).
+1, 2), it is worth noting a number of similarities.
+In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1).
 A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013].
@@
 To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012].
-At the same time, critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
-Only 15-20% of transcripts upregulated in response to temperature are bimodal (i.e.
-generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
+The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role.
+Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
+Only 15-20% of transcripts upregulated in response to temperature are bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
 2).
@@
 
-One possible explanation for the differences between species in their thermal reactionomes is neutral drift of gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
-While we cannot rule this possibility out, two lines of evidence make us think that it is unlikely.
-First, if gene expression differed by drift between species, we would not expect to see any striking differences among expression type classes as we do here (Fig.
+Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
+While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig.
 1, Fig.
 S1).
-Second, even if these differences were due to drift, it would be unlikely that there would be transcriptome-wide differences in the critical temperatures of gene activation such as we see here (Fig.
-2).
+Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.
 
@@
 
-The raw Illumina reads are available from the NCBI Short Read Archive .... The script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).
+Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626.
+The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at [TBD: Harvard Forest Archives].
+Reproducible R script for analysis is available from GitHub [@stanton-geddes2014].
+
+## Acknowledgements
+
+Support made possible by NSF DEB Award #1136644 and the Vermont Genetics Network through Grant Number 8P20GM103449 from the INBRE Program of the National Institute of General Medical Sciences (NIGMS) of the National Institutes of Health (NIH).
+Its contents are solely the responsibility of the authors and do not necessarily represent the official views of NIGMS or NIH.
 
@@
 
-Figure 1.
-Matched expression type data for *A.
-carolinensis* (rows) and *A.
-picea* (columns).
-The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
-Blue cells are those with significant excesses, while red cells are those with significant absences.
-
-Figure 2.
-Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
-Red bars and lines are for *A.
-carolinensis* while blue bars and lines are for *A.
-picea*.
-
-Figure 3.
-Scatterplots of (A) constitutive expression in *A.
-picea* against degree of induction for upregulated transcripts in *A.
-carolinensis* and (B) constitutive expression in *A.
-carolinensis* against degree of induction for upregulated transcripts in *A.
-picea.*.
-For each species, the transcripts are split into expression category based on the ...
-The top row is for *High* transcripts, middle is for *Intermediate* transcripts and bottom row is for *Low* transcripts.
+Figure legends go here.
 
@@
 
+Figures here.
+
 # References

### 2014-10-24 doc/PNAS_cover_letter.tex
final version PNAS cover letter

@@
   \fancyhf{}
-  \setlength{\headheight}{40pt}
+  \setlength{\headheight}{30pt}
   \fancyhead[L]{\includegraphics[scale=0.1]{UVM_Official_Logo.jpg}} 
   \fancyhead[R]{\small Department of Biology \\102  Marsh Life Sciences \\ Burlington, VT 05405 }
-  \renewcommand{\headrulewidth}{0.4pt}
+  \renewcommand{\headrulewidth}{0.3pt}
   \renewcommand{\footrulewidth}{0pt}}
@@
 
-We are excited to submit our manuscript \emph{Investigation of thermal
-reactionomes reveals differing responses to thermal extremes in warm and
+We are excited to submit our manuscript \emph{Thermal
+reactionomes reveal adaptive responses to thermal extremes in warm and
 cold-climate ant species} for consideration as a Research Report in
@@
 acclimate and rapidly adapt to changing environmental conditions, potentially
-rescuing them from extinction. This paradox poses a dilemma for evolutionary
-ecologists. In this study, we evaluate plasticity and adaptive differentiation
-in gene expression of closely related cool and warm-climate ant species.
+rescuing them from extinction. How these two processes are related at the molecular
+level is poorly understood, hampered by the limitations inherent in extrapolating 
+complex responses and functional interactions from the relatively simple experimental
+design of most gene expression experiments to date.
 
-We present a novel analysis by extending the reaction
+In this study, we resolve this problem by extending the reaction
 norm approach to the study of gene expression. We define the
@@
 between short-term physiological acclimation and longer-term adaptive
-change in response to thermal extremes. In particular, we find that
-adaptation to chronic high temperatures involves different mechanisms
-than those used to cope with rarely encountered stressors, which may
-limit adaptive potential of northern species in the face of rapidly
-shifting climatices.
+change in response to thermal extremes. In particular, we find that while
+extensions of lower critical thermal limits involve enhancing existing plasticity,
+adaptation to chronic high temperatures involves a shift from induced resistance
+toward novel tolerance mechanisms, which may limit adaptive potential 
+in the face of rapidly rising temperatures associated with global climate change.
 
@@
 \noindent{Thank you for consideration of our work,} \\
-\includegraphics[scale=1]{JSG_signature.png}  \\
-
-Dr. John Stanton-Geddes
+\includegraphics[scale=0.8]{JSG_signature.png}  \\
+Dr. John Stanton-Geddes on behalf of all the authors
 

### 2014-10-24 ApTranscriptome_TR.Rmd

@@
 # A. carolinensis **Low** and **Bimodal**
-md_sub1 <- subset(md, Acar %in% c("Low", "Bimodal", "High"), )
+md_sub1 <- subset(md, Acar %in% c("Bimodal", "Low"), )
 md_sub1 <- droplevels(md_sub1)
@@
   labs(x = "A. picea", y = "A. carolinensis") +
-  theme(legend.position = "bottom") + 
-  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black") +
-  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
-  geom_rect(xmin = 3.5, xmax = 4.5, ymin = .5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+  theme(legend.position = "none") + 
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
 
@@
 
-md_sub2 <- subset(md, Apic %in% c("Low", "High", "Bimodal"), )
+# A. picea **High** and **Bimodal**
+md_sub2 <- subset(md, Apic %in% c("High", "Bimodal"), )
 md_sub2 <- droplevels(md_sub2)
@@
   labs(x = "A. carolinensis", y = "A. picea") + 
-  theme(legend.position = "none") + 
-  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black") +
+  theme(legend.position = "bottom") + 
   geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
@@
 png("results/matched_observations_Fig1.png")
-grid.arrange(mh_plot_sub2, mh_plot_sub1, ncol=1, heights = c(0.45, 0.6))
+grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=1, heights = c(0.45, 0.6))
 dev.off()
@@
 # ratio of constitutive expression
-Ap.response.type2.A22.high$Ar.A22.constitutive <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt
+Ap.response.type2.A22.high$A22.Ar.constitutive <- Ap.response.type2.A22.high$A22.opt / Ap.response.type2.A22.high$Ar.opt
 
 # plot!
-gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg), x = log(Ar.A22.constitutive))) + 
+gg11 <- ggplot(Ap.response.type2.A22.high, aes(x = log(A22.Ar.upreg), y = log(A22.Ar.constitutive))) + 
   geom_point() + 
   geom_smooth(method = "lm") +
-  labs(y = expression(paste("log(inducibility ", italic("A. picea / A. carolinensis"), ")")),
-       x = expression(paste("log(constitutive ", italic("A. carolinensis / A. picea"), ")")))
+  labs(x = expression(paste("log(inducibility ", italic("A. picea / A. carolinensis"), ")")),
+       y = expression(paste("log(constitutive expression ", italic("A. picea / A. carolinensis"), ")")))
 
@@
 
-A22.high.lm <- lm(log(A22.Ar.upreg) ~ log(Ar.A22.constitutive), data = Ap.response.type2.A22.high)
+A22.high.lm <- lm(log(Ar.A22.constitutive) ~ log(A22.Ar.upreg), data = Ap.response.type2.A22.high)
 
@@
 # ratio of constitutive expression
-Ap.response.type2.Ar.low$A22.Ar.constitutive <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt
-
-gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.constitutive))) + 
+Ap.response.type2.Ar.low$Ar.A22.constitutive <- Ap.response.type2.Ar.low$Ar.opt / Ap.response.type2.Ar.low$A22.opt
+
+gg12 <- ggplot(Ap.response.type2.Ar.low, aes(x = log(Ar.A22.upreg), y = log(Ar.A22.constitutive))) + 
   geom_point() + 
   geom_smooth(method = "lm") +
-  labs(y = expression(paste("log(inducibility ", italic("A. carolinensis / A. picea"), ")")),
-       x = expression(paste("log(constitutive ", italic("A. picea / A. carolinensis"), ")")))
+  labs(x = expression(paste("log(inducibility ", italic("A. carolinensis / A. picea"), ")")),
+       y = expression(paste("log(constitutive expression ", italic("A. carolinensis / A. picea"), ")")))
 
@@
 
-Ar.low.lm <- lm(log(Ar.A22.upreg) ~ log(A22.Ar.constitutive), data = Ap.response.type2.Ar.low)
+Ar.low.lm <- lm(log(Ar.A22.constitutive) ~ log(Ar.A22.upreg), data = Ap.response.type2.Ar.low)
 
@@
 
-```{r fig2}
+```{r fig3}
 # add labels
@@
 # grid.arrange
-png(file = "results/genetic_assimilation_Fig2.png")
+png(file = "results/genetic_assimilation_Fig3.png")
 grid.arrange(gg11, gg12, nrow = 1)
@@
 
-png("results/crit_temp_regulation_Fig3.png")
+png("results/crit_temp_regulation_Fig2.png")
 grid.arrange(gg2, gg3, nrow = 2, heights = c(.5,.6))
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, chi2, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
+save(TPM.dt.sub, RxNpval, sigtable, chi1, chi2, chi3, chi4, chi5, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
+
+save(sigtable, chi1, chi2, chi3, chi4, chi5, chi6, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, md, file = paste("doc/ms_slim_", Sys.Date(), ".RData", sep = ""))
 ```

### 2014-10-24 doc/ApTranscriptome_manuscript.Rmd
FINAL edits to ms

@@
 
+**Short title**: Thermal reactionomes of cool and warm-climate ant species
+
 **Authors**
@@
 
-**Affilitiations**
+**Affiliations**
 
@@
 \textsuperscript{5} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
-\textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen
+\textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen, Universitetsparken 15, DK-2100 Copenhagen
+
+**Corresponding author**
+
+John Stanton-Geddes
+
+Department of Biology, University of Vermont, 109 Carrigan Dr, Burlington, VT 05405.
+
+Phone: 802-656-2922
+
+Email: john.stantongeddes.research@uvm.edu
+
+**Keywords**
+
+*Aphaenogaster*, gene expression, plasticity, reactionome, transcriptome
 
@@
 To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
-We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* across 12 temperature intervals that spanned their entire thermal breadth.
+We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperature intervals that spanned their entire thermal breadth.
 We found that at least 8% of all genes were thermally responsive.
@@
 
-Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation.
-To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
+Coping with progressive changes in climate requires both acclimation and long-term adaptation.
+To better understand the relationship between these mechanisms, we compared gene expression as a function of temperature in two ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
 carolinensis*.
 We found substantial differences in their patterns of gene expression.
-*Aphaenogaster picea* responded to extreme temperatures primarily through upregulation, with a robust heat shock response.
+Aphaenogaster picea responded to extreme temperatures primarily through upregulation, with a robust heat shock response.
 In contrast, *A.
 carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes.
-These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a qualitative shift in mechanisms from those used to cope with rarely encountered stressors.
+These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a shift in mechanisms from those used to cope with rarely encountered stressors.
 
@@
 picea* [@umphrey1996].
-*Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
+*Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
 *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A.
 picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A.
-carolinensis* versus 41.4°C for *A.
+carolinensis* versus 41.3°C for *A.
 picea*; see Methods) and minimum temperatures (6.1°C for *A.
@@
 
-To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
+To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated a *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
 
 We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation.
-First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced responses to provide greater protection from more frequently encountered stressors.
+First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
 This hypothesis would predict that the cool-adapted *A.
@@
 library(topGO)
+library(plyr)
 
@@
 
-## Reaction norms of thermally-responsive transcripts differ between species
+## Reaction norms of thermally-responsive transcripts
 
@@
 Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
-Of these 9,052 transcripts, 80% (...) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
+Of these 9,052 transcripts, ~80% (7,201) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
 high vs.
@@
 However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.
+
+Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
+To examine this, we grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
+The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A.
+carolinensis* and *A.
+picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A.
+carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A.
+picea*.
 
@@
 ```
-
-Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
-To examine this, we grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
-The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A.
-carolinensis* and *A.
-picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A.
-carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A.
-picea*.
 
@@
 
-![Deviations from expected numbers of transcripts in matched observation data for transcript expression type data.
+![Deviations from expected numbers of transcripts in matched observations of transcript expression type.
 The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
-Shown are **Low**, **High** and **Bimodal** expression types for (A) *A.
-picea* and (B) *A.
-carolinensis* against all expression types in the opposite species.
 The dashed boxes indicate the diagonals (e.g.
-**High** - **High**).](../results/matched_observations_Fig1.png)
+**High** - **High**).
+(A) **Low** and **Bimodal** expression types for *A.
+carolinensis* showing bias away from all other categories in *A.
+picea*.
+(B) **High** and **Bimodal** expression types for *A.
+picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**).
+The complete set of matched observations is shown in Fig.
+S1.](../results/matched_observations_Fig1.png)
 
@@
 picea* to downregulation in *A.
-carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
+carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` significantly enriched GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
 In addition, there were GO terms related to repair (e.g.
@@
 
-Similarly, for the **Intermediate** transcripts we compared the mean temperatures of transcript downregulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures.
-Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately ...
-at the higher temperature extremes and a shift of ...
-at low temperature extremes.
-Again, there was greater inertia of gene expression only in response to high temperature extremes.
+Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
+Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes.
+There was greater inertia of gene expression in response to high and low temperature extremes.
 With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
-picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures the species did not differ in mean temperature of downregulation (*A.
-carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A.
-picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
+picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A.
+picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A.
+carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
 
@@
 
-We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by comparing the log ratios of relative inducibility to relative constitutive expression.
-Specifically, this hypothesis predicted a negative relationship between the two.
+We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the midpoint temperature (19.5°C).
+If stress-response genes have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two.
 We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A.
@@
 
-![Scatterplots of (A) log ratio of inducibility in *A.
-picea* over inducibility in *A.
-carolinensis* against log ration of constitutive expression in *A.
-carolinensis* over constitutive expression in *A.
-picea* for transcripts with **High** expression in *A.
-picea*, and (B) log ratio of inducibility in *A.
-carolinensis* over inducibility in *A.
-picea* against log ration of constitutive expression in *A.
-picea* over constitutive expression in *A.
-carolinensis*.
-Blue lines and confidence intervals are from are ordinary least squares regressions.](../results/genetic_assimilation_Fig3.png)
+![Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
+picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A.
+carolinensis* (*P* < 0.001, r^2^ = 0.1).
+Blue lines and confidence intervals are from ordinary least squares regressions.
+](../results/genetic_assimilation_Fig3.png)
 
@@
 picea* but were not thermally responsive in *A.
-carolinensis* (blue cells in bottom row of Fig.
-1).
+carolinensis* (blue cells in far-right column of Fig.
+1A).
 Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A.
@@
 carolinensis*.
-Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include constitutive defenses [@barshis2013],  enhanced proteome stability [@ghosh2010] or behavioral changes [@sunday2014] to tolerate thermal stress.
+Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress.
 These differences are in line with expectations that *A.
 carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions.
-In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
+In fact, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
 
@@
 [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
-Both the genetic assimilation and tolerance routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
+Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
 Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.
@@
 Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
-Only 15-20% of transcripts upregulated in response to temperature are bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
+Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
 2).
@@
 A number of issues have hampered RNA-seq studies to date.
-Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify transcripts that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012].
+Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012].
 Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species.
 While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
-Thus, our method provides an example of how RNA-seq data can be used to understand gain meaningful insight on the process of evolution.
 
@@
 Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].
-Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al.
-2001).
+Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins [@krebs1997] or by increasing the thermal stability of proteins at the expense of catalytic activity [@arnold2001].
 Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.
@@
 carolinensis*.
-At least 10% of their transcriptomes are thermally responsive.
+At least 8% of their transcriptomes are thermally responsive.
 For both species, more transcripts responded to low than high temperatures.
@@
 
-We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W).
-The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice DeMarco, pers.
-communication), and there is about 2% pairwise divergence between these clades using five nuclear genes (Bernice DeMarco, unpublished data).
-Colonies of these species were maintained in common conditions at ??
-for 6 months prior to experimentation.
+We collected a single colony of each of the ant species *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and *Aphaenogaster carolinensis*, part of the *A.
+rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012.
+Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University).
+Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation.
+We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W).
+Following the protocol of Warren and Chick [-@warren2013], we tested the upper and lower critical thermal limits for 5 ants from each of these colonies.
 
-From each species, we haphazardly collected 12 ants from a single colony at the same time on 12 days.
-Each day, the 12 ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
+```{r therm_limits, echo=FALSE, results='hide'}
+therm.limits <- read.csv("../data/thermal_limits.csv")
+avg.therm.limits <- ddply(therm.limits, .(Species), summarise,
+                          CTmax = mean(CTmax),
+                          CTmin = mean(CTmin))
+
+therm.limits.table <- data.frame(Species = avg.therm.limits[,"Species"],
+                    Lat = therm.limits[c(1,6), "Lat"],
+                    Lon = therm.limits[c(1,6), "Lon"],
+                    MAT = therm.limits[c(1,6), "MAT"],
+                    CTmax = avg.therm.limits["CTmax"],
+                    CTmin = avg.therm.limits["CTmin"])
+
+therm.limits.table
+
+#pandoc.table(therm.limits.table, style="rmarkdown", digits = 3, split.table = Inf,
+#             caption = "Upper (CTmax) and lower (CTmin) critical thermal limits for 5 ants from colonies of *Aphaenogaster picea* from Molly Bog, Vermont and *Aphaenogaster rudis* from Duke Forest, North Carolina.
+*Aphaenogaster rudis* and *A.
+carolinensis* are in the same species-complex.
+Latitude (Lat), longitude (Long) and mean annual temperature (MAT) of each collection site are also given.")
+```
+
+To characterize the thermal reactionome, we haphazardly collected 10 ants from each colony at the same time on 12 days.
+Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
 The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
@@
 We removed 5,675 transcripts identified as known contaminants, leaving 99,861 clean transcripts.
-Of these, 47% have best hits to transcripts from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
 
@@
 
-## Thermally-responsive transcripts
+## Identification of thermally-responsive transcripts
 
@@
 We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
-As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis.
-The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).
+As preliminary examination of the data (Supplemental Methods 1) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
+The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods 1).
 
@@
 For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
+We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°C) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
 
-## Species-level comparisons
+## Statistical analyses
 
-We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction to determine if the two species differed in the tempertaure of transcript activation and sensitivity of the thermally-responsive transcripts.
+We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction.
 For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
+We used $\chi^2$ tests to determine if the total number of responsive transcripts, and the number of transcripts in each expression category differed between species.
+To evaluate if shifts from one expression category to another were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equality.
+
 To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
-We grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
+For upregulated transcripts, we grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
 We then performed a *t*-test to determine if the mean temperature of transcript activation differed between the two species for each group.
+For downregulated transcripts (i.e.
+**Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test as before.
 
-To determine if the sensitivity of transcripts to thermal extremes differed between the species, we used the subset of **Intermediate** and **Bimodal** transcripts in each species.
-For each **Intermediate** transcript in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g.
-point of maximum expression had highest probability of being sampled).
-We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species.
-Finding a greater standard deviation of expression would be indicative of less sensitivity to changes in temperature.
-For the **Bimodal** transcripts, we performed the same analysis as with the **Intermediate** transcripts, but with a greater standard deviation being indicative of a more sensitive transcript (convex reaction norm) while a low standard deviation being a transcript with greater inertia of expression in response to temperature changes (nearly flat).Thus, finding one species had overall greater standard deviation of expression for **Bimodal** transcripts would be indicative of greater thermal sensitivity for that species.
-
-To test for genetic assimilation, we compared constitutive expression levels (expression at 19.5° C) for the subset of transcripts that were responsive to low temperatures (**Low** transcripts) in the southern *A.
-carolinensis* species, or high temperatures (**High** transcripts) in the northern *A.
-picea* species using non-parametric two-sided Wilcoxon tests.
+To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A.
+picea* and for **Low** transcripts in *A.
+carolinensis*.
+Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*.
 
@@
 
-To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013].
+To describe the molecular processes involved in thermal adaptation, we performed gene set enrichment analysis (GSEA) using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013].
 Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
-We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.
 
@@
 
-# Figures
-
-Figures here.
-
 # References

### 2014-10-24 doc/ApTranscriptome_manuscript.Rmd

@@
 However, little is known about how thermal tolerance evolves.
-To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
-We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperature intervals that spanned their entire thermal breadth.
+To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced defense*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
+We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth.
 We found that at least 8% of all genes were thermally responsive.
-The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A.
+The majority of upregulation was specific to low temperatures, and consistent with *enhanced defense* the cool-adapted *A.
 picea* induced expression of more genes in response to extreme temperatures than *A.
@@
 
-```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A.
-carolinensis* and *A.
-picea*.
-**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}
-
-rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")
-
-pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
-             big.mark = ",", split.tables = Inf, 
-             caption = "Table of the number of thermally-responsive transcripts by expression type for *A.
-carolinensis* and *A.
-picea*.
-**Low** are transcripts with increased expression at low temperatures (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
-```
-
 ## Expression response to thermal extremes differs between species
@@
 
-![Deviations from expected numbers of transcripts in matched observations of transcript expression type.
-The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
-The dashed boxes indicate the diagonals (e.g.
-**High** - **High**).
-(A) **Low** and **Bimodal** expression types for *A.
-carolinensis* showing bias away from all other categories in *A.
-picea*.
-(B) **High** and **Bimodal** expression types for *A.
-picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**).
-The complete set of matched observations is shown in Fig.
-S1.](../results/matched_observations_Fig1.png)
-
 The molecular processes in the set of transcripts that switched from upregulation in *A.
@@
 picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
-2).
+2A).
 For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
@@
 picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
-2).
-
-![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
-Red bars and lines are for *A.
-carolinensis* while blue bars and lines are for *A.
-picea*.](../results/crit_temp_regulation_Fig2.png)
+2A).
 
 Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
-Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes.
+Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig.
+2).
 There was greater inertia of gene expression in response to high and low temperature extremes.
@@
 picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A.
-carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
+carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
+2B).
 
@@
 In fact, for both comparisons we found a weakly positive relationship.
-
-![Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
-picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A.
-carolinensis* (*P* < 0.001, r^2^ = 0.1).
-Blue lines and confidence intervals are from ordinary least squares regressions.
-](../results/genetic_assimilation_Fig3.png)
 
@@
 Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*.
+In addition, we used a Mann-Whitney test to compare the baseline constitutive expression between species for all responsive transcripts.
 
@@
 
-Figure legends go here.
+**Figure 1**.
+Deviations from expected numbers of transcripts in matched observations of transcript expression type.
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
+The dashed boxes indicate the diagonals (e.g.
+**High** - **High**).
+(A) **Low** and **Bimodal** expression types for *A.
+carolinensis* showing bias away from all other categories in *A.
+picea*.
+(B) **High** and **Bimodal** expression types for *A.
+picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**).
+The complete set of matched observations is shown in Fig.
+S1.
+
+**Figure 2**.
+Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
+Red bars and lines are for *A.
+carolinensis* while blue bars and lines are for *A.
+picea*.
+
+**Figure 3**.
+Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
+picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A.
+carolinensis* (*P* < 0.001, r^2^ = 0.1).
+Blue lines and confidence intervals are from ordinary least squares regressions.
+
+# Tables
+
+```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A.
+carolinensis* and *A.
+picea*.
+**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}
+
+rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")
+
+pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
+             big.mark = ",", split.tables = Inf, 
+             caption = "Table of the number of thermally-responsive transcripts by expression type for *A.
+carolinensis* and *A.
+picea*.
+**Low** are transcripts with increased expression at low temperatures (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
+```
+
+# Figures
+
+![Deviations from expected numbers of transcripts in matched observations of transcript expression type.
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
+The dashed boxes indicate the diagonals (e.g.
+**High** - **High**).
+(A) **Low** and **Bimodal** expression types for *A.
+carolinensis* showing bias away from all other categories in *A.
+picea*.
+(B) **High** and **Bimodal** expression types for *A.
+picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**).
+The complete set of matched observations is shown in Fig.
+S1.](../results/matched_observations_Fig1.png)
+
+![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
+Red bars and lines are for *A.
+carolinensis* while blue bars and lines are for *A.
+picea*.](../results/crit_temp_regulation_Fig2.png)
+
+![Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
+picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A.
+carolinensis* (*P* < 0.001, r^2^ = 0.1).
+Blue lines and confidence intervals are from ordinary least squares regressions.
+](../results/genetic_assimilation_Fig3.png)
+
+\newpage
 

### 2014-10-24 ApTranscriptome_TR.Rmd
changes to TR for final submission to PNAS

@@
 # arrange plots to make figure for manuscript
-mh_plot_sub2 <- mh_plot_sub2 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
-mh_plot_sub1 <- mh_plot_sub1 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
+mh_plot_sub1 <- mh_plot_sub1 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
+mh_plot_sub2 <- mh_plot_sub2 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
 

### 2014-11-04 doc/ApTranscriptome_manuscript.Rmd
added legend for Fig S1 to manuscript and separate file for Fig S1

@@
 ---
-title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
-  and cool-climate ant species
+title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm and cool-climate ant species
 author: ' '
@@
 
-**Short title**: Thermal reactionomes of cool and warm-climate ant species
+**Short title**: Thermal reactionomes of warm and cool-climate ants
 
@@
 Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another.
-These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Figure S1) indicating that these differences are not simply due to neutral processes.
+These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Fig.
+S1) indicating that these differences are not simply due to neutral processes.
 The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species.
@@
 Deviations from expected numbers of transcripts in matched observations of transcript expression type.
-The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
-The dashed boxes indicate the diagonals (e.g.
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (blue = excess, red = deficit).
+The dashed boxes indicate the matched (e.g.
 **High** - **High**).
@@
 
-# Tables
+**Figure S1.** Deviations from expected numbers of transcripts in matched observations of transcript expression type between species (*A.
+carolinensis* on rows, *A.
+picea* on columns).
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (blue = excess, red = deficit).
+The expression types are **Low** transcripts that had greatest expression temperatures < 10°C, **Intermediate** transcripts with greatest expression between 10 and 30°C, **High** transcripts that had greatest expression at temperatures > 31°, **Bimodal** transcripts with increased expression at both high and low temperatures, and **Not Responsive** transcripts that were not thermally responsive in that species.
+
+\newpage
+
+# Tables and Figures
 
@@
 ```
-
-# Figures
 

### 2014-11-04 doc/FigS1.Rmd
added legend for Fig S1 to manuscript and separate file for Fig S1

--- 2014-10-24 doc/ApTranscriptome_manuscript.Rmd
@@
 ---
-title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
-  and cool-climate ant species
-author: ' '
-csl: pnas1.csl
-output:
-  word_document: default
-  pdf_document:
-    fig_caption: yes
-    includes:
-      in_header: header.tex
-    keep_tex: yes
-    latex_engine: xelatex
-    number_sections: no
-    toc_depth: 2
-  html_document: null
-includes: header.tex
-bibliography: ApTranscriptome.bib
+output: pdf_document
 ---
-
-\linenumbers
-
-\newpage
-
-**Short title**: Thermal reactionomes of cool and warm-climate ant species
-
-**Authors**
-
-John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, Lacy Chick \textsuperscript{2}, James Vincent \textsuperscript{3}, Mahesh Vangala \textsuperscript{3}, Robert R.
-Dunn \textsuperscript{4}, Aaron M.
-Ellison \textsuperscript{5}, Nathan J.
-Sanders \textsuperscript{2,6}, Nicholas J.
-Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}
-
-\vspace{10 mm}
-
-**Affiliations**
-
-\textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
-\textsuperscript{2} Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, Tenessee 37996 \newline
-\textsuperscript{3} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
-\textsuperscript{4} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
-\textsuperscript{5} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
-\textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen, Universitetsparken 15, DK-2100 Copenhagen
-
-**Corresponding author**
-
-John Stanton-Geddes
-
-Department of Biology, University of Vermont, 109 Carrigan Dr, Burlington, VT 05405.
-
-Phone: 802-656-2922
-
-Email: john.stantongeddes.research@uvm.edu
-
-**Keywords**
-
-*Aphaenogaster*, gene expression, plasticity, reactionome, transcriptome
-
-\newpage
-
-# Abstract
-
-The distributions of species and their responses to climate change are in part determined by their thermal tolerances.
-However, little is known about how thermal tolerance evolves.
-To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
-We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperature intervals that spanned their entire thermal breadth.
-We found that at least 8% of all genes were thermally responsive.
-The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A.
-picea* induced expression of more genes in response to extreme temperatures than *A.
-carolinensis*, including heat shock proteins.
-In contrast, under high temperatures the warm-adapted *A.
-carolinensis* downregulated many of the genes upregulated in *A.
-picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis.
-We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.
-These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.
-
-# Significance Statement
-
-Coping with progressive changes in climate requires both acclimation and long-term adaptation.
-To better understand the relationship between these mechanisms, we compared gene expression as a function of temperature in two ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
-carolinensis*.
-We found substantial differences in their patterns of gene expression.
-Aphaenogaster picea responded to extreme temperatures primarily through upregulation, with a robust heat shock response.
-In contrast, *A.
-carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes.
-These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a shift in mechanisms from those used to cope with rarely encountered stressors.
-
-# Introduction
-
-Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009].
-Many species, especially small-bodied arthropods, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014].
-Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010].
-Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013].
-These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].
-Outside of candidate genes [@krebs1998], little is known about the evolution of thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes.
-This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].
-
-To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity and plasticity of genes involved in thermal tolerance [@hoffmann2008; @somero2011].
-In recent years, whole organism gene expression approaches (e.g.
-transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits.
-Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes.
-However, temperature and other environmental factors vary continuously in nature.
-As a result, such comparisons (e.g.
-high vs.
-low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is.
-Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992].
-Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].
-
-Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
-We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
-rudis*-species complex, B.
-DeMarco personal communication) and *A.
-picea* [@umphrey1996].
-*Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-*Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A.
-picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A.
-carolinensis* versus 41.3°C for *A.
-picea*; see Methods) and minimum temperatures (6.1°C for *A.
-carolinensis* versus -0.1°C for *A.
-picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
-
-```{r climate, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
-library(raster)
-w <- getData('worldclim', var='bio', res=10)
-#names(w)
-# Plot first bioclim dimension (MAT) to check download worked
-#plot(w, 1, col=colorRampPalette(c("blue","green","red"))(255))
-# sites
-sites <- data.frame(site = c("Ac", "Ap"), lon=c(-78.874, -72.702), lat= c(36.037, 44.508))
-# extract bioclimatic variables from worldclim data. Note bioclim temperature data is multiplied by 10 (no decimal), so transform by dividing by 10.
-dbio <- extract(w, sites[,c("lon","lat")])
-# bio1 = MAT
-# bio4 = Temperature Seasonality (standard deviation *100)
-dbio.vars <- cbind(sites, dbio[,1], dbio[,4])
-colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 1]")] <- "MATx10"
-dbio.vars$MAT <- dbio.vars$MATx10/10
-colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 4]")] <- "Seasonality"
-dbio.vars
+```{r type_table, echo=FALSE, results='asis', fig.cap = "Deviations from expected numbers of transcripts in matched observations of transcript expression type.
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
+The diagonal is the matched observations for each class (e.g.
+**High** - **High**)"}
+library(ggplot2)
+load("ms_slim_2014-10-22.RData")
+mh_plot
 ```
 
-To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated a *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
-
-We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation.
-First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
-This hypothesis would predict that the cool-adapted *A.
-picea* would induce more genes in response to low temperatures than would the warm-adapted *A.
-carolinensis*, which would induce more genes in response to high temperatures.
-Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A.
-picea*, high temperatures for *A.
-carolinensis*.
-Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014].
-This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways.
-Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.
-This hypothesis predicts that transcripts responsive to high temperatures in *A.
-picea* will have higher constitutive expression in *A.
-carolinensis*, whereas transcripts responsive to low temperatures in *A.
-carolinensis* will have higher constitutive expression in *A.
-picea*.
-Of course, these hypotheses are not mutually exclusive.
-By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation.
-
-# Results
-
-```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
-#options(scipen = 1, digits = 3)
-load("ms_slim_2014-10-22.RData")
-library(data.table)
-library(ggplot2)
-library(reshape2)
-library(pander)
-library(RColorBrewer)
-library(topGO)
-library(plyr)
-
-options(digits = 3)
-panderOptions('round', 3)
-panderOptions('keep.trailing.zeros', TRUE)
-```
-
-## Reaction norms of thermally-responsive transcripts
-
-The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts.
-About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
-Of these 9,052 transcripts, ~80% (7,201) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
-high vs.
-low temperature).
-The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246).
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.
-
-Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
-To examine this, we grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
-The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A.
-carolinensis* and *A.
-picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A.
-carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A.
-picea*.
-
-```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A.
-carolinensis* and *A.
-picea*.
-**Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}
-
-rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")
-
-pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
-             big.mark = ",", split.tables = Inf, 
-             caption = "Table of the number of thermally-responsive transcripts by expression type for *A.
-carolinensis* and *A.
-picea*.
-**Low** are transcripts with increased expression at low temperatures (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
-```
-
-## Expression response to thermal extremes differs between species
-
-The two species differed in the number of transcripts in each response category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
-*Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A.
-carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`).
-Consistent with the *enhanced defense* hypothesis, *A.
-picea* induced 1,146 (30%) more transcripts in response to low temperatures than *A.
-carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`).
-However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, at high temperatures *A.
-carolinensis* induced expression of ~14% fewer transcripts than *A.
-picea* (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* < `r max(chi5$p.value, 0.001)`).
-
-```{r echo=FALSE, results='hide'}
-# results of Stuart-Maxwell test
-mh.test
-```
-
-In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species.
-Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another.
-These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Figure S1) indicating that these differences are not simply due to neutral processes.
-The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species.
-The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.
-For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A.
-carolinensis* for low temperatures, *A.
-picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species.
-As with the overall results, the patterns of association differed for the two temperature extremes.
-Genes upregulated at low temperatures in *A.
-carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A.
-picea*, suggesting shared response pathways (Fig.
-1A).
-In contrast, genes upregulated in response to high temperatures in *A.
-picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A.
-carolinensis* (Fig.
-1B).
-Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig.
-1B).
-
-![Deviations from expected numbers of transcripts in matched observations of transcript expression type.
-The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
-The dashed boxes indicate the diagonals (e.g.
-**High** - **High**).
-(A) **Low** and **Bimodal** expression types for *A.
-carolinensis* showing bias away from all other categories in *A.
-picea*.
-(B) **High** and **Bimodal** expression types for *A.
-picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**).
-The complete set of matched observations is shown in Fig.
-S1.](../results/matched_observations_Fig1.png)
-
-The molecular processes in the set of transcripts that switched from upregulation in *A.
-picea* to downregulation in *A.
-carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` significantly enriched GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
-In addition, there were GO terms related to repair (e.g.
-histone modification), transcript regulation (e.g.
-positive regulation of cellular process) and cellular organization (e.g.
-transport, organelle organization).
-Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A.
-carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A.
-picea*, increases metabolism and initiates damage repair responses.
-
-## *A. carolinensis* has greater inertia of expression change to increases in temperature than does *A. picea*
-
-As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures.
-We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression.
-In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A.
-carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A.
-picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
-2).
-For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
-carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
-picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
-2).
-
-![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
-Red bars and lines are for *A.
-carolinensis* while blue bars and lines are for *A.
-picea*.](../results/crit_temp_regulation_Fig2.png)
-
-Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
-Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes.
-There was greater inertia of gene expression in response to high and low temperature extremes.
-With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
-carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
-picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A.
-picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A.
-carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).
-
-## No evidence for *genetic assimilation*
-
-We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the midpoint temperature (19.5°C).
-If stress-response genes have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two.
-We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A.
-picea* (Fig.
-3A) or transcripts upregulated at low temperatures in *A.
-carolinensis* (Fig.
-3B).
-In fact, for both comparisons we found a weakly positive relationship.
-
-![Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
-picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A.
-carolinensis* (*P* < 0.001, r^2^ = 0.1).
-Blue lines and confidence intervals are from ordinary least squares regressions.
-](../results/genetic_assimilation_Fig3.png)
-
-In addition, the thermally responsive transcripts in *A.
-carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A.
-picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`).
-An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A.
-picea* but were not thermally responsive in *A.
-carolinensis* (blue cells in far-right column of Fig.
-1A).
-Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A.
-carolinensis*.
-
-# Discussion
-
-The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances.
-However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance.
-By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify evolutionary changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A.
-carolinensis*.
-We found large-scale, non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.
-
-Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
-We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A.
-picea* is substantially lower than *A.
-carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig.
-2A).
-Moreover, the transcripts upregulated in *A.
-picea* included nearly the entire set upregulated in *A.
-carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1).
-Two non-mutually exclusive hypotheses may explain this pattern.
-First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002].
-The northern species *A.
-picea* may induce a greater response to survive the longer winter period.
-Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012].
-This requirement may be under stronger selection in *A.
-picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.
-
-In contrast to cold tolerance, the enhanced upper thermal limit in *A.
-picea* is best explained by the *tolerance* hypothesis.
-High temperatures were associated with significantly fewer upregulated transcripts in *A.
-carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A.
-picea* were either downregulated or expressed at negligible levels overall in *A.
-carolinensis*.
-Similarly, significantly higher temperatures were required to induce the expression response for *A.
-carolinensis* than *A.
-picea* towards the high thermal limit (Fig.
-2B-D).
-This difference between species is also evident among candidate genes known to be involved in stress response.
-Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
-picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
-carolinensis*.
-Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress.
-These differences are in line with expectations that *A.
-carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions.
-In fact, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
-
-The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts.
-This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al.
-[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
-Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
-Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.
-
-Given the differences in the patterns of thermal responsiveness between species (Fig.
-1, 2), it is worth noting a number of similarities.
-In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1).
-A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013].
-This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
-Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012].
-To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012].
-The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role.
-Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
-Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
-2).
-This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.
-
-A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels).
-A number of issues have hampered RNA-seq studies to date.
-Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012].
-Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species.
-While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
-
-Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
-While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig.
-1, Fig.
-S1).
-Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.
-
-As with all *de novo* transcriptomes, our work has some caveats.
-First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented.
-Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts.
-A genome assembly, in progress, will help to reduce fragmentation.
-Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*.
-Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.
-
-Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
-Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].
-Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins [@krebs1997] or by increasing the thermal stability of proteins at the expense of catalytic activity [@arnold2001].
-Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.
-Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.
-
-## Conclusion
-
-In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A.
-carolinensis*.
-At least 8% of their transcriptomes are thermally responsive.
-For both species, more transcripts responded to low than high temperatures.
-Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have more limited genetic accomodation to high than low temperatures.
-Our results also indicate that these two ant species have different responses to thermal extremes.
-*A.
-picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A.
-carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance.
-The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species.
-Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.
-
-# Material and Methods
-
-## Samples
-
-Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
-Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
-On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
-
-We collected a single colony of each of the ant species *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and *Aphaenogaster carolinensis*, part of the *A.
-rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012.
-Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University).
-Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation.
-We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W).
-Following the protocol of Warren and Chick [-@warren2013], we tested the upper and lower critical thermal limits for 5 ants from each of these colonies.
-
-```{r therm_limits, echo=FALSE, results='hide'}
-therm.limits <- read.csv("../data/thermal_limits.csv")
-avg.therm.limits <- ddply(therm.limits, .(Species), summarise,
-                          CTmax = mean(CTmax),
-                          CTmin = mean(CTmin))
-
-therm.limits.table <- data.frame(Species = avg.therm.limits[,"Species"],
-                    Lat = therm.limits[c(1,6), "Lat"],
-                    Lon = therm.limits[c(1,6), "Lon"],
-                    MAT = therm.limits[c(1,6), "MAT"],
-                    CTmax = avg.therm.limits["CTmax"],
-                    CTmin = avg.therm.limits["CTmin"])
-
-therm.limits.table
-
-#pandoc.table(therm.limits.table, style="rmarkdown", digits = 3, split.table = Inf,
-#             caption = "Upper (CTmax) and lower (CTmin) critical thermal limits for 5 ants from colonies of *Aphaenogaster picea* from Molly Bog, Vermont and *Aphaenogaster rudis* from Duke Forest, North Carolina.
-*Aphaenogaster rudis* and *A.
-carolinensis* are in the same species-complex.
-Latitude (Lat), longitude (Long) and mean annual temperature (MAT) of each collection site are also given.")
-```
-
-To characterize the thermal reactionome, we haphazardly collected 10 ants from each colony at the same time on 12 days.
-Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
-The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
-At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
-We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
-
-## Sequencing, assembly and annotation
-
-For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the *A.
-picea* and *A.
-carolinensis* samples respectively.
-Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
-
-We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
-*De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g.
-regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013].
-We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
-Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters).
-We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
-To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.
-
-The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 transcripts with a total length of 100 million bp.
-Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 transcripts.
-The total length was 63 million bp with an N~50~ length of 895 bp and a mean transcript size of 593 bp.
-Of the 105,536 filtered transcripts, 55,432 had hits to the NCBI-nr database.
-Of these, 38,711 transcripts mapped to GO terms, 1,659 transcripts were identified to an enzyme and 18,935 transcripts mapped to a domain with >50% coverage.
-We removed 5,675 transcripts identified as known contaminants, leaving 99,861 clean transcripts.
-
-To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies transcripts by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
-
-## Identification of thermally-responsive transcripts
-
-We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression.
-We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
-As preliminary examination of the data (Supplemental Methods 1) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
-The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods 1).
-
-To identify transcripts that have significant changes in expression across the thermal gradient, we fit the regression model
-
-$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$
-
-to each transcript.
-We used $log(TPM + 1)$ as the response to control for skew in the expression data.
-For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005].
-This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
-To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
-
-We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
-All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
-
-For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°C) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
-
-## Statistical analyses
-
-We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction.
-For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
-We used $\chi^2$ tests to determine if the total number of responsive transcripts, and the number of transcripts in each expression category differed between species.
-To evaluate if shifts from one expression category to another were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equality.
-
-To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
-For upregulated transcripts, we grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
-We then performed a *t*-test to determine if the mean temperature of transcript activation differed between the two species for each group.
-For downregulated transcripts (i.e.
-**Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test as before.
-
-To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A.
-picea* and for **Low** transcripts in *A.
-carolinensis*.
-Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*.
-
-## Gene set enrichment analysis
-
-To describe the molecular processes involved in thermal adaptation, we performed gene set enrichment analysis (GSEA) using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013].
-Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
-
-## Data availability
-
-Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626.
-The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at [TBD: Harvard Forest Archives].
-Reproducible R script for analysis is available from GitHub [@stanton-geddes2014].
-
-## Acknowledgements
-
-Support made possible by NSF DEB Award #1136644 and the Vermont Genetics Network through Grant Number 8P20GM103449 from the INBRE Program of the National Institute of General Medical Sciences (NIGMS) of the National Institutes of Health (NIH).
-Its contents are solely the responsibility of the authors and do not necessarily represent the official views of NIGMS or NIH.
-
-# Figure Legends
-
-Figure legends go here.
-
-# References
+Figure S1.
+Deviations from expected numbers of transcripts in matched observations of transcript expression type between species (*A.
+carolinensis* on rows, *A.
+picea* on columns).
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (blue = excess, red = deficit).
+The expression types are **Low** transcripts that had greatest expression temperatures < 10°C, **Intermediate** transcripts with greatest expression between 10 and 30°C, **High** transcripts that had greatest expression at temperatures > 31°, **Bimodal** transcripts with increased expression at both high and low temperatures, and **Not Responsive** transcripts that were not thermally responsive in that species.

### 2014-11-26 doc/ApTranscriptome_manuscript.Rmd
updates to files for submission to Molecular Ecology

@@
 ---
-title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm and cool-climate ant species
+title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
+  and cool-climate ant species
 author: ' '
-csl: pnas1.csl
+csl: molecol.csl
 output:
+  word_document: default
   pdf_document:
@@
 These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.
-
-# Significance Statement
-
-Coping with progressive changes in climate requires both acclimation and long-term adaptation.
-To better understand the relationship between these mechanisms, we compared gene expression as a function of temperature in two ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A.
-carolinensis*.
-We found substantial differences in their patterns of gene expression.
-Aphaenogaster picea responded to extreme temperatures primarily through upregulation, with a robust heat shock response.
-In contrast, *A.
-carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes.
-These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a shift in mechanisms from those used to cope with rarely encountered stressors.
 
@@
 
-# Results
+# Materials and Methods
 
@@
 ```
+
+## Samples
+
+Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
+Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
+On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
+
+We collected a single colony of each of the ant species *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and *Aphaenogaster carolinensis*, part of the *A.
+rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012.
+Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University).
+Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation.
+We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W).
+Following the protocol of Warren and Chick [-@warren2013], we tested the upper and lower critical thermal limits for 5 ants from each of these colonies.
+
+```{r therm_limits, echo=FALSE, results='hide'}
+therm.limits <- read.csv("../data/thermal_limits.csv")
+avg.therm.limits <- ddply(therm.limits, .(Species), summarise,
+                          CTmax = mean(CTmax),
+                          CTmin = mean(CTmin))
+
+therm.limits.table <- data.frame(Species = avg.therm.limits[,"Species"],
+                    Lat = therm.limits[c(1,6), "Lat"],
+                    Lon = therm.limits[c(1,6), "Lon"],
+                    MAT = therm.limits[c(1,6), "MAT"],
+                    CTmax = avg.therm.limits["CTmax"],
+                    CTmin = avg.therm.limits["CTmin"])
+
+therm.limits.table
+
+#pandoc.table(therm.limits.table, style="rmarkdown", digits = 3, split.table = Inf,
+#             caption = "Upper (CTmax) and lower (CTmin) critical thermal limits for 5 ants from colonies of *Aphaenogaster picea* from Molly Bog, Vermont and *Aphaenogaster rudis* from Duke Forest, North Carolina.
+*Aphaenogaster rudis* and *A.
+carolinensis* are in the same species-complex.
+Latitude (Lat), longitude (Long) and mean annual temperature (MAT) of each collection site are also given.")
+```
+
+To characterize the thermal reactionome, we haphazardly collected 10 ants from each colony at the same time on 12 days.
+Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
+The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
+At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
+We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
+
+## Sequencing, assembly and annotation
+
+For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the *A.
+picea* and *A.
+carolinensis* samples respectively.
+Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
+
+We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
+*De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g.
+regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013].
+We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
+Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Methods).
+We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
+To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.
+
+The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 transcripts with a total length of 100 million bp.
+Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 transcripts.
+The total length was 63 million bp with an N~50~ length of 895 bp and a mean transcript size of 593 bp.
+Of the 105,536 filtered transcripts, 55,432 had hits to the NCBI-nr database.
+Of these, 38,711 transcripts mapped to GO terms, 1,659 transcripts were identified to an enzyme and 18,935 transcripts mapped to a domain with >50% coverage.
+We removed 5,675 transcripts identified as known contaminants, leaving 99,861 clean transcripts.
+
+To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies transcripts by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
+
+## Identification of thermally-responsive transcripts
+
+We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression.
+We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
+As preliminary examination of the data (Supplemental Methods) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
+The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods).
+
+To identify transcripts that have significant changes in expression across the thermal gradient, we fit the regression model
+
+$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$
+
+to each transcript.
+We used $log(TPM + 1)$ as the response to control for skew in the expression data.
+For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005].
+This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
+To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
+
+We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental Methods).
+
+For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°C) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
+
+## Statistical analyses
+
+We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction.
+For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
+We used $\chi^2$ tests to determine if the total number of responsive transcripts, and the number of transcripts in each expression category differed between species.
+To evaluate if shifts from one expression category to another were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equality.
+
+To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
+For upregulated transcripts, we grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
+We then performed a *t*-test to determine if the mean temperature of transcript activation differed between the two species for each group.
+For downregulated transcripts (i.e.
+**Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test as before.
+
+To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A.
+picea* and for **Low** transcripts in *A.
+carolinensis*.
+Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*.
+In addition, we used a Mann-Whitney test to compare the baseline constitutive expression between species for all responsive transcripts.
+
+## Gene set enrichment analysis
+
+To describe the molecular processes involved in thermal adaptation, we performed gene set enrichment analysis (GSEA) using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013].
+Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
+
+# Results
 
@@
 
-A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels).
+A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach).
 A number of issues have hampered RNA-seq studies to date.
@@
 
-# Material and Methods
-
-## Samples
-
-Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
-Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
-On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].
-
-We collected a single colony of each of the ant species *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and *Aphaenogaster carolinensis*, part of the *A.
-rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012.
-Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University).
-Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation.
-We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W).
-Following the protocol of Warren and Chick [-@warren2013], we tested the upper and lower critical thermal limits for 5 ants from each of these colonies.
-
-```{r therm_limits, echo=FALSE, results='hide'}
-therm.limits <- read.csv("../data/thermal_limits.csv")
-avg.therm.limits <- ddply(therm.limits, .(Species), summarise,
-                          CTmax = mean(CTmax),
-                          CTmin = mean(CTmin))
-
-therm.limits.table <- data.frame(Species = avg.therm.limits[,"Species"],
-                    Lat = therm.limits[c(1,6), "Lat"],
-                    Lon = therm.limits[c(1,6), "Lon"],
-                    MAT = therm.limits[c(1,6), "MAT"],
-                    CTmax = avg.therm.limits["CTmax"],
-                    CTmin = avg.therm.limits["CTmin"])
-
-therm.limits.table
-
-#pandoc.table(therm.limits.table, style="rmarkdown", digits = 3, split.table = Inf,
-#             caption = "Upper (CTmax) and lower (CTmin) critical thermal limits for 5 ants from colonies of *Aphaenogaster picea* from Molly Bog, Vermont and *Aphaenogaster rudis* from Duke Forest, North Carolina.
-*Aphaenogaster rudis* and *A.
-carolinensis* are in the same species-complex.
-Latitude (Lat), longitude (Long) and mean annual temperature (MAT) of each collection site are also given.")
-```
-
-To characterize the thermal reactionome, we haphazardly collected 10 ants from each colony at the same time on 12 days.
-Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
-The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
-At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
-We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
-
-## Sequencing, assembly and annotation
-
-For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the *A.
-picea* and *A.
-carolinensis* samples respectively.
-Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].
-
-We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011].
-*De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g.
-regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013].
-We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
-Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters).
-We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
-To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.
-
-The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 transcripts with a total length of 100 million bp.
-Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 transcripts.
-The total length was 63 million bp with an N~50~ length of 895 bp and a mean transcript size of 593 bp.
-Of the 105,536 filtered transcripts, 55,432 had hits to the NCBI-nr database.
-Of these, 38,711 transcripts mapped to GO terms, 1,659 transcripts were identified to an enzyme and 18,935 transcripts mapped to a domain with >50% coverage.
-We removed 5,675 transcripts identified as known contaminants, leaving 99,861 clean transcripts.
-
-To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies transcripts by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
-
-## Identification of thermally-responsive transcripts
-
-We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression.
-We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
-As preliminary examination of the data (Supplemental Methods 1) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
-The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods 1).
-
-To identify transcripts that have significant changes in expression across the thermal gradient, we fit the regression model
-
-$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$
-
-to each transcript.
-We used $log(TPM + 1)$ as the response to control for skew in the expression data.
-For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005].
-This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
-To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
-
-We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
-All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
-
-For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°C) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
-
-## Statistical analyses
-
-We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction.
-For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
-We used $\chi^2$ tests to determine if the total number of responsive transcripts, and the number of transcripts in each expression category differed between species.
-To evaluate if shifts from one expression category to another were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equality.
-
-To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
-For upregulated transcripts, we grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
-We then performed a *t*-test to determine if the mean temperature of transcript activation differed between the two species for each group.
-For downregulated transcripts (i.e.
-**Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test as before.
-
-To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A.
-picea* and for **Low** transcripts in *A.
-carolinensis*.
-Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*.
-In addition, we used a Mann-Whitney test to compare the baseline constitutive expression between species for all responsive transcripts.
-
-## Gene set enrichment analysis
-
-To describe the molecular processes involved in thermal adaptation, we performed gene set enrichment analysis (GSEA) using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013].
-Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
-
-## Data availability
-
-Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626.
-The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at [TBD: Harvard Forest Archives].
-Reproducible R script for analysis is available from GitHub [@stanton-geddes2014].
-
-## Acknowledgements
+# Acknowledgements
 
@@
 Its contents are solely the responsibility of the authors and do not necessarily represent the official views of NIGMS or NIH.
+
+## Data Accessibility
+
+Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626.
+The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive).
+The reproducible and version-controlled script for data analysis is available from GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
+
+## Author Contributions
+
+JSG, NG and SHC designed research.
+JSG, ADN and LC performed research.
+JSG, JV, MV and SCH analyzed data and wrote the paper.
+ADN and LC performed research.
+JV and MV analyzed data.
+JSG, RD, AE, NS, NG and SHC wrote the paper.
 

### 2014-11-26 doc/MolEco_cover_letter.tex
updates to files for submission to Molecular Ecology

--- 2014-10-24 doc/PNAS_cover_letter.tex
@@
   \fancyhf{}
-  \setlength{\headheight}{30pt}
+  \setlength{\headheight}{40pt}
   \fancyhead[L]{\includegraphics[scale=0.1]{UVM_Official_Logo.jpg}} 
@@
 
-\noindent{Dear \emph{PNAS} Editorial Board,}
+\noindent{Dear Dr. Riesberg,}
 
@@
 reactionomes reveal adaptive responses to thermal extremes in warm and
-cold-climate ant species} for consideration as a Research Report in
-\emph{The Proceedings of the National Academy of Sciences}. There are
+cold-climate ant species} for consideration as an Original Article in
+\emph{Molecular Ecology}. There are
 several aspects of the work that we think make it worthy of publication
-in \emph{PNAS}.
+in \emph{Molecular Ecology}.
 
@@
 
-This manuscript is most appropriate for the major classification
-BIOLOGICAL SCIENCES, minor classification Ecology.
+\textbf{Suggested reviewers:}
 
-\textbf{Suggested Editors:} Douglas J. Futuyma, David L. Denlinger, David M. Hillis
+\begin{itemize}
+  \item Dr. Nicholas Teets - expertise with insect genomics and transcriptomics
+  \item Dr. Vanessa Kellerman - expertise with insect population and quantitative genetics
+  \item Dr. Pavlos Pavlidis - critical insight on function annotation and gene ontologies
+  \item Dr. Joel Kingsolver - expertise in insect thermal limits
+  \item Dr. Daniel Barshis - expertise in ecological genomics
+\end{itemize}
 
-\textbf{Suggested NAS Members:} Trudy F. Mackay, Nevo Eviata, George N. Somero
+\textbf{Suggested editors:}
 
-\textbf{Suggested reviewers:}  Dr.
-Nicholas Teets, Postdoc, University of Florida; Dr.
-Vanessa Kellerman, Postdoc, Monash University; Dr.
-Pavlos Pavlidis, Postdoc, Heidelberg Institute of Theoretical Studies; Dr.
-Stephen Palumbi, Professor, Hopkins Marine Station, Stanford University; Dr.
-Joel Kingsolver, Professor, University of North Carolina
+\begin{itemize}
+  \item Dr. Ary Hoffman - expertise with insect population and quantitative genetics
+  \item Dr. Stephen Palumbi - expertise in ecological genomics
+  \item Dr. Aurelie Bonin - expertise in population genomics
+\end{itemize}
 

### 2015-04-28 doc/ApTranscriptome_manuscript.Rmd
Edits in response to reviewer comments.

@@
 ---
-title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
+title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
   and cool-climate ant species
@@
 
-To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated a *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
-
+To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) and characterized the reaction norm for each gene using a custom analysis.
 We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation.
@@
 Of course, these hypotheses are not mutually exclusive.
-By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation.
+By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adaptation.
 
@@
 
-We collected a single colony of each of the ant species *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and *Aphaenogaster carolinensis*, part of the *A.
+We collected a single colony of *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of *Aphaenogaster carolinensis*, part of the *A.
 rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012.
+These locations are central to each species distributions.
 Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University).
@@
 We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W).
-Following the protocol of Warren and Chick [-@warren2013], we tested the upper and lower critical thermal limits for 5 ants from each of these colonies.
+We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves [@warren2013].
 
@@
 
-To characterize the thermal reactionome, we haphazardly collected 10 ants from each colony at the same time on 12 days.
+To characterize the thermal reactionome, we haphazardly collected three ants from each 2012 colony at the same time on 12 days.
 Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
-The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013].
+The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality.
 At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
-We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
+We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
 
@@
 
-$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$
+$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temperature) + \beta_3(temperature^2) + \beta_4(species * temperature) + \beta_5(species * temperature^2) + \epsilon $$
 
-to each transcript.
+with to each transcript.
+Temperature and species were both fixed effects, with a quadratic term included for temperature.
 We used $log(TPM + 1)$ as the response to control for skew in the expression data.
@@
 picea*, suggesting shared response pathways (Fig.
-1A).
+2A).
 In contrast, genes upregulated in response to high temperatures in *A.
@@
 carolinensis* (Fig.
-1B).
+2B).
 Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig.
-1B).
+2B).
 
@@
 picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
-2A).
+3A).
 For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
@@
 picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
-2A).
+3A).
 
@@
 Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig.
-2).
+3).
 There was greater inertia of gene expression in response to high and low temperature extremes.
@@
 carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
-2B).
+3B).
 
@@
 picea* (Fig.
-3A) or transcripts upregulated at low temperatures in *A.
+4A) or transcripts upregulated at low temperatures in *A.
 carolinensis* (Fig.
-3B).
+4B).
 In fact, for both comparisons we found a weakly positive relationship.
@@
 picea* but were not thermally responsive in *A.
-carolinensis* (blue cells in far-right column of Fig.
-1A).
+carolinensis* (far-right column of Fig.
+2B).
 Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A.
@@
 carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig.
-2A).
+3A).
 Moreover, the transcripts upregulated in *A.
@@
 picea* towards the high thermal limit (Fig.
-2B-D).
+3B-D).
 This difference between species is also evident among candidate genes known to be involved in stress response.
@@
 Given the differences in the patterns of thermal responsiveness between species (Fig.
-1, 2), it is worth noting a number of similarities.
+2, 2), it is worth noting a number of similarities.
 In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1).
-A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013].
+A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013].
 This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
-Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012].
-To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012].
 The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role.
@@
 Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
-2).
+3).
 This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.
@@
 While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig.
-1, Fig.
+2, Fig.
 S1).
@@
 Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.
+In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability (...citations…).
 
@@
 *A.
-picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A.
-carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance.
+picea* responds by increasing expression of transcripts related to metabolism, heat shock proteins and other protective molecules, whereas *A.
+carolinensis* decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance.
 The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species.
@@
 Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626.
-The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive).
-The reproducible and version-controlled script for data analysis is available from GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
+The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40.
+The reproducible and version-controlled script for data analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
 
@@
 **Figure 1**.
-Deviations from expected numbers of transcripts in matched observations of transcript expression type.
-The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (blue = excess, red = deficit).
+Cartoon figure showing the predicted results for the deviations from the expected number of transcripts in each expression type.
+The color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
+The (A) *enhanced defense* hypothesis assumes that temperature adapation uses the same mechanisms, and thus that there is a strong overlap of response (e.g.
+excess in shared response types between species).
+In contrast, the (B) *tolerance* hypothesis predicts that responsive transcripts in one species will become non-responsive or shift expression type in the other species (e.g.
+excess in non-shared response).
+Expression types are defined in the text and Table 1.
+We show only the results for *A.
+carolinensis* in response to low temperatures as the pattern would be similar *A.
+picea* in response to high temperatures.
+
+**Figure 2**.
+Deviations from expected number of transcripts in expression type.
+The color of each cell  (purple = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
 The dashed boxes indicate the matched (e.g.
 **High** - **High**).
-(A) **Low** and **Bimodal** expression types for *A.
-carolinensis* showing bias away from all other categories in *A.
-picea*.
-(B) **High** and **Bimodal** expression types for *A.
-picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**).
+(A) For *A.
+carolinensis* there is an excess of shared **Low** and **Bimodal** expression types and a bias away from all other categories in *A.
+picea*, consistent with the *enhanced defense* hypothesis (Fig.
+1).
+(B) For *A.
+picea*, an excess of **High** and **Bimodal** transcripts shift to expression patterns that are not upregulated (**Intermediate** and **Not Responsive**), partially consistent with the *tolerance* hypothesis.
 The complete set of matched observations is shown in Fig.
 S1.
+Expression types are defined in Table 1.
 
-**Figure 2**.
+**Figure 3**.
 Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
@@
 
-**Figure 3**.
+**Figure 4**.
 Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
@@
 picea* on columns).
-The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (blue = excess, red = deficit).
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (purple = excess, orange = deficit).
 The expression types are **Low** transcripts that had greatest expression temperatures < 10°C, **Intermediate** transcripts with greatest expression between 10 and 30°C, **High** transcripts that had greatest expression at temperatures > 31°, **Bimodal** transcripts with increased expression at both high and low temperatures, and **Not Responsive** transcripts that were not thermally responsive in that species.
@@
 
-![Deviations from expected numbers of transcripts in matched observations of transcript expression type.
-The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit).
-The dashed boxes indicate the diagonals (e.g.
-**High** - **High**).
-(A) **Low** and **Bimodal** expression types for *A.
-carolinensis* showing bias away from all other categories in *A.
-picea*.
-(B) **High** and **Bimodal** expression types for *A.
-picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**).
-The complete set of matched observations is shown in Fig.
-S1.](../results/matched_observations_Fig1.png)
+![Fig. 2](../results/matched_observations_Fig2.png)
 
-![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
-Red bars and lines are for *A.
-carolinensis* while blue bars and lines are for *A.
-picea*.](../results/crit_temp_regulation_Fig2.png)
+![Fig. 3](../results/crit_temp_regulation_Fig3.png)
 
-![Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
-picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A.
-carolinensis* (*P* < 0.001, r^2^ = 0.1).
-Blue lines and confidence intervals are from ordinary least squares regressions.
-](../results/genetic_assimilation_Fig3.png)
+![Fig. 4](../results/genetic_assimilation_Fig4.png)
 

### 2015-04-28 doc/ApTranscriptome_manuscript.Rmd
Added paragraph from SHC about common garden approach.

@@
 ```
+
+## Common Garden Design
+
+Ideally, genetically-based variation in gene expression profiles would be identified by comparing individuals completely reared under common-garden conditions to eliminate environmental variation experienced either as adults or during development.
+However, *Aphaenogaster* colonies are long-lived, cannot be bred under laboratory conditions, and do not achieve complete turnover of the workforce for at least a year or longer.
+Thus, as is commonly done with other long-lived organisms [@barshis2013, ], we exposed both colonies to common-garden rearing conditions for three months to fully acclimate adult workers to common temperatures.
+Over this time, roughly 1-2 cohorts of new  workers are expected to join the workforce (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.
 

### 2015-04-28 ApTranscriptome_TR.Rmd
Made cartoon figure showing predictions for shifts in transcript expression.

@@
+---
+output: word_document
+---
 Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
@@
 
-To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) r citep("doi:10.1186/1471-2105-12-385"). I ...
-
-~~~
-KRONA code
-~~~
-
-The interactive visualization is available [here]().
+To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) r citep("doi:10.1186/1471-2105-12-385") which provides a taxonomic profile of all transcripts by BLAST annotation to species.
+
+The interactive visualization is available [here](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).
 
@@
 
-**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.**
+**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.**  The 'Ar_7' sample clusters with all the other A22 samples, while the 'A22_7' sample clusters with all the other Ar samples.
+The parsimonious explanation for this is a labeling mistake.
+Removing these samples, rather than relabeling, is the conservative approach.
+
+![clustering](results/aphaenogaster_sample_clustering.png)
 
@@
 # extract transcript names from fasta file rather than loading whole file
-system("grep '^>' results/trinity-full/Trinity_cap3_uclust_cont.fa | cut -f 1 -d ' ' | sed -r 's/^.{1}//' > results/trinity-full/cont.list")
+system("grep '^>' results/trinity-full/Trinity_cap3_uclust_cont.fa | cut -f 1 -d ' ' | sed 's/^.{1}//' > results/trinity-full/cont.list")
 
@@
 ```{r temperature_response}
-A22.type.table <- table(Ap.response.type[ , 'A22.type'])
-Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])
+A22.type.table <- table(Ap.response.type$A22.type)
+Ar.type.table <- table(Ap.response.type$Ar.type)
 
@@
 # Reorganize for plotting to show overlap among categories between colonies
-type.table <- table(Acar = Ap.response.type[ , 'Ar.type'], Apic = Ap.response.type[ , 'A22.type'])
+type.table <- table(Acar = Ap.response.type$Ar.type, Apic = Ap.response.type$A22.type)
 # reorder
@@
 ```{r mh_plot}
-md <- melt(Ec.cells)
+md <- reshape2::melt(Ec.dev)
 
 mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) +
-  scale_fill_gradient2(limits=c(-600, 600)) +
+  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 
+mh_plot
 # make subplot of rows that match hypotheses
@@
 mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
-  scale_fill_gradient2(limits=c(-600, 600)) +
+  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 mh_plot_sub2 <- qplot(x=Acar, y=Apic, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Apic))) + 
-  scale_fill_gradient2(limits=c(-600, 600)) +
+  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 
-png("results/matched_observations_Fig1.png")
+png("results/matched_observations_Fig2.png")
 grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=1, heights = c(0.45, 0.6))
@@
 mh_plot
+dev.off()
+```
+
+Per suggestion of a reviewer, make a cartoon figure that shows the contrasting results of the *enhance defense* and the *tolerance* hypotheses.
+
+The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response.
+This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g.
+more shared "Low" transcripts) and a deficit of transcripts that shift to other response patterns or become not responsive.
+
+```{r ed_cartoon}
+enhanced_defense_dat <- md_sub1
+enhanced_defense_dat$value <- 0
+# set non-shared response to random negative value 
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min = -300, max = 0)
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- runif(5, min = -300, max = 0)
+# change shared response to positive
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 500
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 300
+
+enhanced_defense_dat
+
+# plot
+enhanced_defense_plot <- qplot(x=Apic, y=Acar, data=enhanced_defense_dat, fill=value, geom="tile", ylim = rev(levels(enhanced_defense_dat$Acar))) + 
+  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
+  theme(axis.title = element_text(face="italic")) +
+  labs(x = "A. picea", y = "A. carolinensis") +
+  theme(legend.position = "none") + 
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+
+# title
+enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced defense hypothesis') + theme(plot.title = element_text(hjust=0))
+enhanced_defense_plot
+```
+
+In contrast, the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.
+
+```{r tol_cartoon}
+tolerance_dat <- md_sub1
+tolerance_dat$value <- 0
+# set non-shared response to random positive value 
+tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min = 100, max = 300)
+tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min = 100, max = 200)
+# change shared response to near zero
+tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- 30
+tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- 20
+
+tolerance_dat
+
+# plot
+tolerance_plot <- qplot(x=Apic, y=Acar, data=tolerance_dat, fill=value, geom="tile", ylim = rev(levels(tolerance_dat$Acar))) + 
+  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
+  theme(axis.title = element_text(face="italic")) +
+  labs(x = "A. picea", y = "A. carolinensis") +
+  theme(legend.position = "bottom") + 
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+
+# title
+tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0))
+tolerance_plot
+
+# plot together
+png("results/predictions_Fig1.png")
+grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=1, heights = c(0.45, 0.6))
 dev.off()
@@
 # grid.arrange
-png(file = "results/genetic_assimilation_Fig3.png")
+png(file = "results/genetic_assimilation_Fig4.png")
 grid.arrange(gg11, gg12, nrow = 1)
@@
 
-png("results/crit_temp_regulation_Fig2.png")
+png("results/crit_temp_regulation_Fig3.png")
 grid.arrange(gg2, gg3, nrow = 2, heights = c(.5,.6))

### 2015-04-28 ApTranscriptome_TR.Rmd
Changed baseline temperature to 25C (the rearing temp) from 19.5C. No qualitative changes to results.

@@
 # get expression at optimum for A22 transcripts
-
 Ap.high.Ac.nr.opt.exp <- Ap.high.Ac.nr$A22.opt
@@
 - compare basal expression at optimal temperature for thermally-responsive genes
-- compare degree of uregulation to constitutive expression
+- compare degree of upregulation to constitutive expression
 - compare the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures 
 - compare the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures
+
+Predict expression for responsive transcripts.
+
+```{r responsive_pred}
+# apply predFunc to all responsive transcripts
+resp.TPM.dt.sub.pred <- ddply(resp.TPM.dt.sub, .(Transcript), .inform="TRUE", predFunc)
+
+# setkey to Transcript and colony 
+resp.TPM.dt.sub.pred <- data.table(resp.TPM.dt.sub.pred)
+setkey(resp.TPM.dt.sub.pred, Transcript, colony)
+```
 
@@
 Genes upregulated in response to thermal stress in one species may have greater basal levels of expression in the other species that experiences those stressful conditions more often.
-To test this hypothesis, we compared expression levels near the optimal temperature (19.5C) between the two species for genes in that are either in the 'High' or 'Low' expression group in the other species.
+To test this hypothesis, we compared expression levels at the rearing temperature (25C) between the two species for genes in that are either in the 'High' or 'Low' expression group in the other species.
 Specifically, do genes upregulated at high temperatures in *A22* have greater basal expression at optimal temperatures in *Ar*?
@@
 
-Consistent with expectation, there is greater expression at 19.5C for *Ar* than *A22* transcripts for the set of transcripts that are transcripts that are up-regulated at high temperatures in *A22*.
+Consistent with expectations, there is greater expression at 25C for *Ar* than *A22* transcripts for the set of transcripts that are transcripts that are up-regulated at high temperatures in *A22*.
 Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.
@@
 I extract the **High** transcripts for *A.
-picea* and plot the log ratio of inducibility in *A.
-picea* over inducibility in *A.
-carolinensis*, against the log ratio of constitutive expression in *A.
-carolinensis* against expression in *A.
-picea*.
-According to the *genetic assimilation hypothesis* there should be a positive relationship between these ratios.
-That is, greater constitutive expression in *A.
-carolinensis* should have greater inducibility towards **High** temperatures in *A.
+picea* and examine the log ratio of inducibility against the log ratio of constitutive expression in *A.
+picea* over *A.
+carolinensis**.
+According to the *genetic assimilation hypothesis* there should be a negative relationship between these ratios.
+That is, transcripts with greater constitutive expression in *A.
+carolinensis* should have greater inducibility in response to **High** temperatures in *A.
 picea*.
@@
 
-# ratio of upregulation
+# ratio of upregulation in A22 vs Ar
 Ap.response.type2.A22.high$A22.Ar.upreg <- Ap.response.type2.A22.high$A22.upreg / Ap.response.type2.A22.high$Ar.upreg
 
-# ratio of constitutive expression
+# ratio of constitutive expression in Ar vs A22
 Ap.response.type2.A22.high$A22.Ar.constitutive <- Ap.response.type2.A22.high$A22.opt / Ap.response.type2.A22.high$Ar.opt
@@
 
-A22.high.lm <- lm(log(Ar.A22.constitutive) ~ log(A22.Ar.upreg), data = Ap.response.type2.A22.high)
+A22.high.lm <- lm(log(A22.Ar.constitutive) ~ log(A22.Ar.upreg), data = Ap.response.type2.A22.high)
 
@@
 
-In contrast to the prediction there is a negative relationship between these ratios.
-That is, genes with greater upregulation of expression also have greater constitutive expression.
-This is also consistent with the results simply comparing constitutive expression between species for the *A.
-picea* **High** transcripts.
+In contrast to the prediction there is a positive relationship between these ratios. That is, genes with greater upregulation of expression also have greater constitutive expression.
 
@@
 ```{r assimilation_Ar_low}
-### Repeat for A. carolinensis *Low* transcripts
+### filter for A. carolinensis *Low* transcripts
 Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
@@
 
-Predict expression for responsive transcripts.
-
-```{r responsive_pred}
-# apply predFunc to all responsive transcripts
-resp.TPM.dt.sub.pred <- ddply(resp.TPM.dt.sub, .(Transcript), .inform="TRUE", predFunc)
-
-# setkey to Transcript and colony 
-resp.TPM.dt.sub.pred <- data.table(resp.TPM.dt.sub.pred)
-setkey(resp.TPM.dt.sub.pred, Transcript, colony)
-```
-
 For next analyses, extract list of gene names by response type.
@@
 ```{r extract_Transcripts}
+Ap.response.type <- as.data.frame(Ap.response.type)
 A22.high.transcripts <-  Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
@@
 # transcripts expressed at *High* and *Bimodal* temperatures in A22
-A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.high.transcripts, A22.bim.transcripts), "A22")]
+A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.high.transcripts.df$Transcript, A22.bim.transcripts), "A22")]
 setkey(A22.high.TPM.dt.sub, Transcript)
@@
 picea* **Intermediate** genes.
-In contrast, towards low temperatures, the temperature of mean temperature of downregulation was higher for *A.
-picea* than *A.
-carolinensis*.
+Towards low temperatures, the mean temperature of downregulation was marginally higher for *A.
+carolinensis* than *A.
+picea*.
 

### 2015-04-28 scripts/RxNseq.R

@@ -136,18 +136,18 @@ RxNtype <- function(lmitem) {
     if(sd(A22.pout$pTPM) < 0.00001) {
         A22.max.val <- NA
         A22.min.val <- NA
-        A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
+        A22.opt <- A22.pout[A22.pout$val == 25, "pTPM"]
         A22.exp.type <- "NotResp"
     } else {
         # else, set values from data
         A22.max.val <- median(A22.pout[which(A22.pout$pTPM == max(A22.pout$pTPM)), "val"])
         # for min.val, take median point as it may be string of points at zero
         A22.min.val <- median(A22.pout[which(A22.pout$pTPM == min(A22.pout$pTPM)), "val"])
-        A22.opt <- A22.pout[A22.pout$val == 19.5, "pTPM"]
+        A22.opt <- A22.pout[A22.pout$val == 25, "pTPM"]
         # determine expression shape
          A22.exp.type <- NA
-        # expression type defined as "Bimodal" if max expression below 10C and above 30C is greater than one standard deviation above expression at the optimum temperature (19.5C)
-        if(max(A22.pout[A22.pout$val < 10, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + sd(A22.pout$pTPM)) & max(A22.pout[A22.pout$val > 30, "pTPM"])[1] > (A22.pout[A22.pout$val == 19.5, "pTPM"] + sd(A22.pout$pTPM))) A22.exp.type <- "Bimodal" else {
+        # expression type defined as "Bimodal" if max expression below 10C and above 30C is greater than one standard deviation above expression at the optimum temperature (25C)
+        if(max(A22.pout[A22.pout$val < 10, "pTPM"])[1] > (A22.pout[A22.pout$val == 25, "pTPM"] + sd(A22.pout$pTPM)) & max(A22.pout[A22.pout$val > 30, "pTPM"])[1] > (A22.pout[A22.pout$val == 25, "pTPM"] + sd(A22.pout$pTPM))) A22.exp.type <- "Bimodal" else {
               # expression type "High" if max expression at temperature above 30C
               if(A22.max.val > 30) A22.exp.type <- "High" else {
               # expression type "Low" if max expression at temperature below 10C
@@ -163,18 +163,18 @@ RxNtype <- function(lmitem) {
[…]

### 2015-04-28 ApTranscriptome_TR.Rmd
Changed figures of deviation from expectations to vertical orientation.

@@
 
-mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
+mh_plot_sub1 <- qplot(y=Apic, x=Acar, data=md_sub1, fill=value, geom="tile", xlim = rev(levels(md_sub1$Acar))) + 
   scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
-  labs(x = "A. picea", y = "A. carolinensis") +
+  labs(y = "A. picea", x = "A. carolinensis") +
   theme(legend.position = "none") + 
-  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
-  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
 
@@
 
-mh_plot_sub2 <- qplot(x=Acar, y=Apic, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Apic))) + 
+mh_plot_sub2 <- qplot(y=Acar, x=Apic, data=md_sub2, fill=value, geom="tile", xlim = rev(levels(md_sub2$Apic))) + 
   scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
-  labs(x = "A. carolinensis", y = "A. picea") + 
-  theme(legend.position = "bottom") + 
-  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
-  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+  labs(y = "A. carolinensis", x = "A. picea") + 
+  theme(legend.position = "right") + 
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black")
 
@@
 png("results/matched_observations_Fig2.png")
-grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=1, heights = c(0.45, 0.6))
+grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=2)
 dev.off()
@@
 
-The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response.
+The *enhanced response hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response.
 This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g.
@@
 # plot
-enhanced_defense_plot <- qplot(x=Apic, y=Acar, data=enhanced_defense_dat, fill=value, geom="tile", ylim = rev(levels(enhanced_defense_dat$Acar))) + 
+enhanced_defense_plot <- qplot(y=Apic, x=Acar, data=enhanced_defense_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_defense_dat$Acar))) + 
   scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
-  labs(x = "A. picea", y = "A. carolinensis") +
+  labs(y = "A. picea", x = "A. carolinensis") +
   theme(legend.position = "none") + 
-  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
-  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
 
 # title
-enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced defense hypothesis') + theme(plot.title = element_text(hjust=0))
+enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0))
 enhanced_defense_plot
@@
 # plot
-tolerance_plot <- qplot(x=Apic, y=Acar, data=tolerance_dat, fill=value, geom="tile", ylim = rev(levels(tolerance_dat$Acar))) + 
+tolerance_plot <- qplot(y=Apic, x=Acar, data=tolerance_dat, fill=value, geom="tile", xlim = rev(levels(tolerance_dat$Acar))) + 
   scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
-  labs(x = "A. picea", y = "A. carolinensis") +
+  labs(y = "A. picea", x = "A. carolinensis") +
   theme(legend.position = "bottom") + 
-  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
-  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
 
@@
 png("results/predictions_Fig1.png")
-grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=1, heights = c(0.45, 0.6))
+grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=2)
 dev.off()
@@
 plot(log(mean.exp[which(mean.exp$colony == "A22"), "mean.TPM"]), 
-     log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]))
+     log(mean.exp[which(mean.exp$colony == "Ar"), "mean.TPM"]),
+     xlab = "log(expression) A. picea",
+     ylab = "log(expression) A. carolinensis")
+abline(a=0, b=1)
 ```
@@
 Of the responsive transcripts, `r round(length(which(responsive.lms.ann.type$best.hit.to.nr != "-"))/length(responsive.lms.ann.type$best.hit.to.nr)*100, 0)`% are annotated, while `r round(length(which(annotation.table$best.hit.to.nr != "-"))/nrow(annotation.table) * 100, 0)`% of all transcripts are annotated.
-
-length(which(responsive.lms.ann.type$Transcript %in% annotation.table[which(annotation.table$best.hit.to.nr != "-"), "Sequence.Name"]))
-
-best.hit.to.nr != "-")
 

### 2015-06-03 doc/ApTranscriptome_manuscript.Rmd
revisions for MolEco

@@
 
-John Stanton-Geddes
+John Stanton-Geddes, Data Scientist
 
-Department of Biology, University of Vermont, 109 Carrigan Dr, Burlington, VT 05405.
+Dealer.com, 1 Howard St, Burlington, VT 05401
 
@@
 
-Email: john.stantongeddes.research@uvm.edu
+Email: john.stantongeddes.research@google.com
 
@@
 However, little is known about how thermal tolerance evolves.
-To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced defense*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
+To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
 We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth.
 We found that at least 8% of all genes were thermally responsive.
-The majority of upregulation was specific to low temperatures, and consistent with *enhanced defense* the cool-adapted *A.
+The majority of upregulation was specific to low temperatures, and consistent with *enhanced response* the cool-adapted *A.
 picea* induced expression of more genes in response to extreme temperatures than *A.
@@
 Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
-We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
+While temporal patterns of transcriptional activity (e.g.
+fast vs.
+slow responding genes) are also important components of an organism's transcriptional response to environmental conditions [@sorensen2005], we focus of the response of transcripts across conditions at the same time point.
+We use the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
 rudis*-species complex, B.
@@
 
-To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) and characterized the reaction norm for each gene using a custom analysis.
-We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation.
-First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
+To characterize the thermal reactionome, we measured the reaction norm for each gene using a regression-based statistical approach to identify temperature-dependent patterns of change in gene expression.
+We used these response patterns to quantitatively test three mechanistic hypotheses of thermal adaptation.
+First, the *enhanced response hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
 This hypothesis would predict that the cool-adapted *A.
@@
 Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014].
-This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways.
+This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011].
 Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.
@@
 However, *Aphaenogaster* colonies are long-lived, cannot be bred under laboratory conditions, and do not achieve complete turnover of the workforce for at least a year or longer.
-Thus, as is commonly done with other long-lived organisms [@barshis2013, ], we exposed both colonies to common-garden rearing conditions for three months to fully acclimate adult workers to common temperatures.
+Thus, as is commonly done with other long-lived organisms [@whitehead2006; @barshis2013], we exposed both colonies to common-garden rearing conditions for six months to fully acclimate adult workers to common temperatures.
 Over this time, roughly 1-2 cohorts of new workers are expected to join the workforce (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.
 
-To characterize the thermal reactionome, we haphazardly collected three ants from each 2012 colony at the same time on 12 days.
+Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable [@cottingham2005].
+Thus, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature.
+To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time on 12 days.
+We measured response to temperature with a one-hour static temperature application, which is ecologically-relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012].
 Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
-The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality.
+The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum (~0°C) and maximum (~43°C) temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality.
 At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C.
+Thus, our reactionome characterized early, but not late, responding genes.
 We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.
@@
 We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
-Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Methods).
+Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supporting Information).
 We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
@@
 We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
-As preliminary examination of the data (Supplemental Methods) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
-The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods).
+As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
+The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).
 
@@
 To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
-
-We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
-All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental Methods).
-
-For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted transcript expression values to partition the thermally-responsive transcripts into five expression categories for each species; transcripts that had greatest expression at high (> 31°C) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (*NotResp*).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
 
@@
 carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`).
-Consistent with the *enhanced defense* hypothesis, *A.
+Consistent with the *enhanced response* hypothesis, *A.
 picea* induced 1,146 (30%) more transcripts in response to low temperatures than *A.
 carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`).
-However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, at high temperatures *A.
+However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced response hypothesis*, at high temperatures *A.
 carolinensis* induced expression of ~14% fewer transcripts than *A.
@@
 S1) indicating that these differences are not simply due to neutral processes.
-The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species.
-The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.
+The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species.
+The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.
 For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A.
@@
 We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression.
-In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A.
+In support of the *tolerance hypothesis* but not the *enhanced response hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A.
 carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A.
@@
 
-Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
+Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
 We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A.
@@
 Given the differences in the patterns of thermal responsiveness between species (Fig.
-2, 2), it is worth noting a number of similarities.
+2), it is worth noting a number of similarities.
 In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1).
-A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013].
+The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures.
+However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures.
+At these extreme low temperatures we also found few upregulated transcripts (Fig.
+3A), while the peak of low-temperature transcriptional activity occurred near 10°C (Fig.
+3).
+A potential explanation for this pattern is that increased gene expression is required to support elevated metabolism at cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002].
+The observation that more transcripts were upregulated at low than high temperatures could be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013].
 This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
@@
 
-A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach).
-A number of issues have hampered RNA-seq studies to date.
+A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach).
+For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as we discussed above.
+Further, a number of issues have hampered RNA-seq studies to date.
 Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012].
@@
 Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
-While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig.
+While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig.
 2, Fig.
@@
 
-As with all *de novo* transcriptomes, our work has some caveats.
-First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented.
+A number of caveats do apply to our work.
+First, species may differ in gene expression along axes which we have not measure here, primarily temporal patterns of gene expression [@sorensen2005], and could be studied in futher work.
+Second, the *de novo* transcriptome assembly is highly fragmented, given that all sequenced ant genomes to date have only about 18,000 genes [@gadau2012].
 Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts.
 A genome assembly, in progress, will help to reduce fragmentation.
-Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*.
+Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*.
 Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.
-In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability (...citations…).
+In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].
 
@@
 **Figure 1**.
-Cartoon figure showing the predicted results for the deviations from the expected number of transcripts in each expression type.
-The color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
-The (A) *enhanced defense* hypothesis assumes that temperature adapation uses the same mechanisms, and thus that there is a strong overlap of response (e.g.
-excess in shared response types between species).
-In contrast, the (B) *tolerance* hypothesis predicts that responsive transcripts in one species will become non-responsive or shift expression type in the other species (e.g.
-excess in non-shared response).
-Expression types are defined in the text and Table 1.
-We show only the results for *A.
-carolinensis* in response to low temperatures as the pattern would be similar *A.
-picea* in response to high temperatures.
+Prediction illustrations of expected thermal response patterns in the two species under alternative mechanistic hypotheses of temperature adaptation.
+Although both temperature extremes were investigated in a similar way, for simplicity only the response to low temperatures is illustrated here.
+Each column indicates the distribution across all response categories in *A.
+picea*, which has a lower CT_{min}_ and is therefore better adapated to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CT_{min}, *A.
+carolinensis*, relative to the null hypothesis of equal marginal frequencies.
+The dashed boxes highlight cells that would indicated matched responses in the two species, and the color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts.
+The (A) *enhanced response* hypothesis proposes that the increase in cold tolerance in *A.
+picea* is achieved by amplifying existing molecular mechanisms, and thus there should be an excess of shared resonse types between species.
+In contrast, the (B) *tolerance* predicts that *A.
+picea* is less reliant on induced responses to confer cold-tolerance than *A.
+carolinensis*, leading to an excess of shifts from induction in *A.
+carolinensis* to the **Not Responsive** or down-regulation categories in *A.
+picea*.
 
 **Figure 2**.
-Deviations from expected number of transcripts in expression type.
+Results of the marginal frequencies analysis of thermal response categories.
 The color of each cell  (purple = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
-The dashed boxes indicate the matched (e.g.
+The units are number of transcripts.
+For each temperature extreme, the species expected to be less well adapated to that extreme is displayed on the x-axis for the two response categories corresponding to upregulation (**Bimodal** and **Low** for the low temperatures, or **Bimodal** and **High** for high temperatures).
+The distribution of response categories for those transcripts in the better-adapted species is arrayed along the y-axis.
+The dashed boxes indicate the matched responses (e.g.
 **High** - **High**).
-(A) For *A.
-carolinensis* there is an excess of shared **Low** and **Bimodal** expression types and a bias away from all other categories in *A.
-picea*, consistent with the *enhanced defense* hypothesis (Fig.
+(A) Low temperature extreme: there is an excess of shared **Low** and **Bimodal** expression types and a bias away from all other categories in *A.
+picea*, consistent with the *enhanced response* hypothesis (Fig.
 1).
-(B) For *A.
-picea*, an excess of **High** and **Bimodal** transcripts shift to expression patterns that are not upregulated (**Intermediate** and **Not Responsive**), partially consistent with the *tolerance* hypothesis.
+(B) High temperature extreme: in addition to an excess of matched categories, there is an excess of **High** and **Bimodal** transcripts in *A.
+picea* that are not upregulated in *A.
+carolinensis* (**Intermediate** and **Not Responsive**), partially consistent with the *tolerance* hypothesis.
 The complete set of matched observations is shown in Fig.
@@
 
+![Fig. 1](../results/predictions_Fig1.png)
+
 ![Fig. 2](../results/matched_observations_Fig2.png)

### 2015-06-03 ApTranscriptome_TR.Rmd
revisions for MolEco

@@
   labs(y = "A. picea", x = "A. carolinensis") +
-  theme(legend.position = "none") + 
+  theme(legend.position = "right") + 
   geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
@@
   labs(y = "A. carolinensis", x = "A. picea") + 
-  theme(legend.position = "right") + 
+  theme(legend.position = "none") + 
   geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
@@
 png("results/matched_observations_Fig2.png")
-grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=2)
+grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=2, widths = c(3.75, 3))
 dev.off()
@@
 # title
-enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0))
+enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
 enhanced_defense_plot
@@
   labs(y = "A. picea", x = "A. carolinensis") +
-  theme(legend.position = "bottom") + 
+  theme(legend.position = "none") + 
   geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
@@
 # title
-tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0))
+tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
 tolerance_plot
@@
 png("results/predictions_Fig1.png")
-grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=2)
+grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=2, widths = c(3.75,3))
 dev.off()
@@
 ```{r shiny_file, eval=FALSE, echo=FALSE}
-shiny.dt <- TPM.dt.sub[responsive.lms.ann.type[,list(Transcript, best.hit.to.nr.1, GO.Biological.Process.1)]]
+shiny.dt <- TPM.dt.sub[responsive.lms.ann.type[,list(Transcript, best.hit.to.nr, GO.Biological.Process)]]
 

### 2015-06-03 doc/ApTranscriptome_manuscript.Rmd
revisions for MolEco

@@
 
-We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression.
-We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
-As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
+We quantified expression of each transcript using the program `Sailfish` [@patro2014] and used the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression.
+We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
+As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, so we omitted these data from the analysis.
 The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).
 
-To identify transcripts that have significant changes in expression across the thermal gradient, we fit the regression model
+To identify transcripts that had significant changes in expression across the thermal gradient, we fit the regression model
 
@@
 To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
+
+We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+All analyses were performed with R 3.1 [@r_core_team2013] and are fully reproducible (Supporting Information).
+
+For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+We used the predicted transcript expression levels to partition transcripts into five expression categories which were defined *a priori*, and could therefore be used to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in one species but did respond in the other.
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
+The two categories which unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).
 
@@
 Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
-To examine this, we grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**).
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).
 The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A.
@@
 Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.
-In particular, functional protein levels cannot be expected to be directly linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].
+In particular, functional protein levels cannot be expected to be fully linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].
 

### 2015-06-11 ApTranscriptome_TR.Rmd
in response to reviewer 2, changed comparisons between species to focus on only the transcripts with a significant species x temperature interaction. previously, this also included transcripts with a main effect of temperature only.

@@
 ---
-output: word_document
+output: pdf_document
 ---
-Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
+
+Technical Report for **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**
 ========================================================================================
@@
 
-**October 14, 2014**
+**`r format(Sys.time(), '%d %B, %Y')`**
 
@@
 
-```{r setup, echo=FALSE, results='hide', message = FALSE}
+```{r setup, echo=FALSE, results='hide', message = FALSE, warning=FALSE}
 # Global settings
@@
 # Load Bioconductor libraries
-source("http://bioconductor.org/biocLite.R")
+#source("http://bioconductor.org/biocLite.R")
 #biocLite("topGO")
@@
 
-In this technical report, which accompanies the manuscript **Patterns of stress tolerance versus resistance in genome-wide expression data of parapatric ant species** (Stanton-Geddes et al., in press), we:
-
-1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
-2. identify thermally-responsive genes
-3. evaluate differences in reaction norms of expression for each thermally-responsive gene between the two species
-4. perform gene set enrichment analysis of thermally-responsive genes for the two species
-
-This script is completely reproducible assuming that R, `knitr` and the other required libraries (listed within the source document) are installed on a standard linux system using the following:
-
-    Rscript -e "library(knitr); knit('ApTranscriptome_TR.Rmd')"
+In this technical report, which accompanies the manuscript **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species** (Stanton-Geddes et al., in press), we:
+
+1. Describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`.
+2. Identify thermally-responsive genes.
+3. Evaluate differences in transcriptome-wide patterns of thermal responsiveness (i.e. the thermal reactionome) between the two species
+4. Perform gene set enrichment analysis of thermally-responsive genes for the two species.
 
@@
 
-# merge results
-Ap.response.type <- rbind(interaction.response.type, temperature.response.type)
-colnames(Ap.response.type)[which(colnames(Ap.response.type) == ".id")] <- "Transcript"
-Ap.response.type <- data.table(Ap.response.type)
-setkey(Ap.response.type, Transcript)
-str(Ap.response.type)
+# for subsequent analyses, focus on the transcripts with significant temperature x species interaction only
+colnames(interaction.response.type)[which(colnames(interaction.response.type) == ".id")] <- "Transcript"
+interaction.response.type <- data.table(interaction.response.type)
+setkey(interaction.response.type, Transcript)
+str(interaction.response.type)
 
 # save results to file
-write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), Ap.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
+write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), interaction.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
 ```
@@
 ```{r temperature_response}
-A22.type.table <- table(Ap.response.type$A22.type)
-Ar.type.table <- table(Ap.response.type$Ar.type)
+A22.type.table <- table(interaction.response.type$A22.type)
+Ar.type.table <- table(interaction.response.type$Ar.type)
 
@@
 # Reorganize for plotting to show overlap among categories between colonies
-type.table <- table(Acar = Ap.response.type$Ar.type, Apic = Ap.response.type$A22.type)
+type.table <- table(Acar = interaction.response.type$Ar.type, Apic = interaction.response.type$A22.type)
 # reorder
@@
 # set non-shared response to random negative value 
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min = -300, max = 0)
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- runif(5, min = -300, max = 0)
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=150)
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- rep(-200, 5)
 # change shared response to positive
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 500
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 300
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 50
+enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 50
 
@@
 # set non-shared response to random positive value 
-tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min = 100, max = 300)
-tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min = 100, max = 200)
+tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=160)
+tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min=140, max=150)
 # change shared response to near zero
-tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- 30
+tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- 20
 tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- 20
@@
 ```{r gen_assimilation_high}
-Ap.high.Ac.nr <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type == "NotResp"), ]
+Ap.high.Ac.nr <- interaction.response.type[which(interaction.response.type$A22.type == "High" & interaction.response.type$Ar.type == "NotResp"), ]
 
@@
 ```{r gen_assimilation_bim}
-Ap.bim.Ac.nr <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "NotResp"), ]
+Ap.bim.Ac.nr <- interaction.response.type[which(interaction.response.type$A22.type == "Bimodal" & interaction.response.type$Ar.type == "NotResp"), ]
 
@@
 # reshape data
-Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))
+Ap.df <- data.frame(Transcript = rep(interaction.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(interaction.response.type$Transcript)), max.val = c(interaction.response.type$A22.max, interaction.response.type$Ar.max))
 
@@
 # list of transcripts that are 'high' expressed in A22
-A22.high.transcripts.df <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]
+A22.high.transcripts.df <- interaction.response.type[which(interaction.response.type$A22.type == "High"), ]
 
@@
 # list of transcripts that are 'high' expressed in Ar
-Ar.low.transcripts.df <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]
+Ar.low.transcripts.df <- interaction.response.type[which(interaction.response.type$Ar.type == "Low"), ]
 
@@
 # list of transcripts that are 'Intermediate' expressed in either colony
-Ap.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" | Ap.response.type$A22.type == "Intermediate"), ]
+Ap.int.transcripts <- interaction.response.type[which(interaction.response.type$Ar.type == "Intermediate" | interaction.response.type$A22.type == "Intermediate"), ]
 # T test
@@
 
-# merge optimum expression, in Ap.response.type data.frame, with predicted TPM data
+# merge optimum expression, in interaction.response.type data.frame, with predicted TPM data
 Ap.induction.wide <- data.table(Ap.induction.wide)
 setkey(Ap.induction.wide, Transcript)
-Ap.response.type2 <- Ap.induction.wide[Ap.response.type]
+interaction.response.type2 <- Ap.induction.wide[interaction.response.type]
 ```
@@
 ```{r log_inducibility}
-Ap.response.type2.A22.high <- Ap.response.type2[Ap.response.type2$Transcript %in% A22.high.transcripts.df$Transcript, ]
-str(Ap.response.type2.A22.high)
+interaction.response.type2.A22.high <- interaction.response.type2[interaction.response.type2$Transcript %in% A22.high.transcripts.df$Transcript, ]
+str(interaction.response.type2.A22.high)
 
 # filter out NotResp transcripts in Ar
-Ap.response.type2.A22.high <- Ap.response.type2.A22.high[which(Ap.response.type2.A22.high$Ar.type != "NotResp"), ]
+interaction.response.type2.A22.high <- interaction.response.type2.A22.high[which(interaction.response.type2.A22.high$Ar.type != "NotResp"), ]
 
 # ratio of upregulation in A22 vs Ar
-Ap.response.type2.A22.high$A22.Ar.upreg <- Ap.response.type2.A22.high$A22.upreg / Ap.response.type2.A22.high$Ar.upreg
+interaction.response.type2.A22.high$A22.Ar.upreg <- interaction.response.type2.A22.high$A22.upreg / interaction.response.type2.A22.high$Ar.upreg
 
 # ratio of constitutive expression in Ar vs A22
-Ap.response.type2.A22.high$A22.Ar.constitutive <- Ap.response.type2.A22.high$A22.opt / Ap.response.type2.A22.high$Ar.opt
+interaction.response.type2.A22.high$A22.Ar.constitutive <- interaction.response.type2.A22.high$A22.opt / interaction.response.type2.A22.high$Ar.opt
 
 # plot!
-gg11 <- ggplot(Ap.response.type2.A22.high, aes(x = log(A22.Ar.upreg), y = log(A22.Ar.constitutive))) + 
+gg11 <- ggplot(interaction.response.type2.A22.high, aes(x = log(A22.Ar.upreg), y = log(A22.Ar.constitutive))) + 
   geom_point() + 
@@
 
-A22.high.lm <- lm(log(A22.Ar.constitutive) ~ log(A22.Ar.upreg), data = Ap.response.type2.A22.high)
+A22.high.lm <- lm(log(A22.Ar.constitutive) ~ log(A22.Ar.upreg), data = interaction.response.type2.A22.high)
 
@@
 ### filter for A. carolinensis *Low* transcripts
-Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
-str(Ap.response.type2.Ar.low)
+interaction.response.type2.Ar.low <- interaction.response.type2[interaction.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
+str(interaction.response.type2.Ar.low)
 
 # filter out NotResp transcripts in A22
-Ap.response.type2.Ar.low <- Ap.response.type2.Ar.low[which(Ap.response.type2.Ar.low$A22.type != "NotResp"), ]
+interaction.response.type2.Ar.low <- interaction.response.type2.Ar.low[which(interaction.response.type2.Ar.low$A22.type != "NotResp"), ]
 
 # ratio of upregulation
-Ap.response.type2.Ar.low$Ar.A22.upreg <- Ap.response.type2.Ar.low$Ar.upreg / Ap.response.type2.Ar.low$A22.upreg
+interaction.response.type2.Ar.low$Ar.A22.upreg <- interaction.response.type2.Ar.low$Ar.upreg / interaction.response.type2.Ar.low$A22.upreg
 
 # ratio of constitutive expression
-Ap.response.type2.Ar.low$Ar.A22.constitutive <- Ap.response.type2.Ar.low$Ar.opt / Ap.response.type2.Ar.low$A22.opt
-
-gg12 <- ggplot(Ap.response.type2.Ar.low, aes(x = log(Ar.A22.upreg), y = log(Ar.A22.constitutive))) + 
+interaction.response.type2.Ar.low$Ar.A22.constitutive <- interaction.response.type2.Ar.low$Ar.opt / interaction.response.type2.Ar.low$A22.opt
+
+gg12 <- ggplot(interaction.response.type2.Ar.low, aes(x = log(Ar.A22.upreg), y = log(Ar.A22.constitutive))) + 
   geom_point() + 
@@
 
-Ar.low.lm <- lm(log(Ar.A22.constitutive) ~ log(Ar.A22.upreg), data = Ap.response.type2.Ar.low)
+Ar.low.lm <- lm(log(Ar.A22.constitutive) ~ log(Ar.A22.upreg), data = interaction.response.type2.Ar.low)
 
@@
 ```{r extract_Transcripts}
-Ap.response.type <- as.data.frame(Ap.response.type)
-A22.high.transcripts <-  Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
-Ar.high.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]
-
-A22.low.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
-Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]
-
-A22.bim.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
-Ar.bim.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
-
-A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
-Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+interaction.response.type <- as.data.frame(interaction.response.type)
+A22.high.transcripts <-  interaction.response.type[which(interaction.response.type$A22.type == "High"), "Transcript"]
+Ar.high.transcripts <- interaction.response.type[which(interaction.response.type$Ar.type == "High"), "Transcript"]
+
+A22.low.transcripts <- interaction.response.type[which(interaction.response.type$A22.type == "Low"), "Transcript"]
+Ar.low.transcripts <- interaction.response.type[which(interaction.response.type$Ar.type == "Low"), "Transcript"]
+
+A22.bim.transcripts <- interaction.response.type[which(interaction.response.type$A22.type == "Bimodal"), "Transcript"]
+Ar.bim.transcripts <- interaction.response.type[which(interaction.response.type$Ar.type == "Bimodal"), "Transcript"]
+
+A22.int.transcripts <- interaction.response.type[which(interaction.response.type$A22.type == "Intermediate"), "Transcript"]
+Ar.int.transcripts <- interaction.response.type[which(interaction.response.type$Ar.type == "Intermediate"), "Transcript"]
 ```
@@
 ```{r mean_exp}
-t.test(log(Ap.response.type$A22.opt), log(Ap.response.type$Ar.opt))
+t.test(log(interaction.response.type$A22.opt), log(interaction.response.type$Ar.opt))
 # expression level at optimum temp greater in Ar than A22
@@
 # non-parametric test
-w4 = wilcox.test(Ap.response.type$A22.opt, 
-                 Ap.response.type$Ar.opt,
+w4 = wilcox.test(interaction.response.type$A22.opt, 
+                 interaction.response.type$Ar.opt,
                  alternative = "two.sided", paired = TRUE, conf.int = TRUE)
@@
 # combine responsive.lms.ann with "type" information
-responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
+responsive.lms.ann.type <- merge(responsive.lms.ann.type, interaction.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
 str(responsive.lms.ann.type)
@@
 ## Genes with *High* expression in both colonies
-Ap.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High" & Ap.response.type$A22.type == "High"), "Transcript"]
+Ap.high <- interaction.response.type[which(interaction.response.type$Ar.type == "High" & interaction.response.type$A22.type == "High"), "Transcript"]
 Ap.geneList.high <- rep(0, length = length(Ap.geneList))
@@
 ## Genes with *Low* expression in both colonies
-Ap.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low" & Ap.response.type$A22.type == "Low"), "Transcript"]
+Ap.low <- interaction.response.type[which(interaction.response.type$Ar.type == "Low" & interaction.response.type$A22.type == "Low"), "Transcript"]
 Ap.geneList.low <- rep(0, length = length(Ap.geneList))
@@
 # Genes with *Bimodal* expression in both colonies
-Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
+Ap.bim <- interaction.response.type[which(interaction.response.type$A22.type == "Bimodal" & interaction.response.type$Ar.type == "Bimodal"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 # genes with *Intermediate* expression in both colonies
-Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
+Ap.int <- interaction.response.type[which(interaction.response.type$A22.type == "Intermediate" & interaction.response.type$Ar.type == "Intermediate"), "Transcript"]
 # create gene list, setting value to 1 for "int" transcripts
@@
 # genes with *Intermediate* expression in Ar and *Bimodal* in A22
-Ar.int.A22.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Bimodal"), "Transcript"]
+Ar.int.A22.bim <- interaction.response.type[which(interaction.response.type$Ar.type == "Intermediate" & interaction.response.type$A22.type == "Bimodal"), "Transcript"]
 Ar.int.A22.bim.geneList <- rep(0, length = length(Ap.geneList))
@@
 # genes with *Intermediate* expression in Ar and *Low* in A22
-Ar.int.A22.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Low"), "Transcript"]
+Ar.int.A22.low <- interaction.response.type[which(interaction.response.type$Ar.type == "Intermediate" & interaction.response.type$A22.type == "Low"), "Transcript"]
 Ar.int.A22.low.geneList <- rep(0, length = length(Ap.geneList))

### 2015-07-02 doc/mec-14-1243-response-letter.md
Final edits for response. Added response letter

Dear Dr. Hoffmann,

We appreciate the comments from you and the reviewers. Both reviewers raise important points and concerns, to which we have responded below with changes to the manuscript and/or rebuttal if we disagreed with the reviewer’s suggestion. We feel that this process has helped us to build a stronger case both for the methodology and main results of our study. We especially thank the reviewers for highlighting ambiguities in some of our original language.  While we respond to the reviewers comments in details below, we address the three major concerns here.

1) Acclimation vs adaptation

Reviewer 1 is concerned that we “cannot tease apart genetic or plastic differences in expression profiles” due to using field-collected ant colonies in this experiment. While a complete-generation common garden experiment is preferable for distinguishing genetic from plastic differences, this is neither possible nor practical for the vast majority of species, including Aphaenogaster. Moreover, rearing a single generation in common conditions does not resolve all these issues (e.g. maternal genetic effects). A key outcome of our work is demonstrating a method to gain insight on the relative role of genetic or plastic differences even when rearing organisms in common gardens is not feasible. We limited differences in developmental acclimation by maintaining colonies in the lab for 6 months prior to sampling. Given that a new cohort is reared in 2-3 months, 6 months allowed for a substantial number of workers to be fully reared under common-garden conditions, such that our samples would have contained a mix of truly common-garden and adult-acclimated workers. This is actually better than has been ascribed to adaptation and published for other long-lived taxa including corals (Barshis et al. 2013 PNAS 110:1387) and fish (Smith et al. BMC Genomics 2013, 14:375). With hindsight, the reviewer’s suggestion of sampling the species from sites where they overlap is great, but unfortunately was not done. However, it is worth noting that in previous work where the species were collected from overlapping sites along an altitudinal gradient in the Smokies (Warren and Chick 2013), both CTmax and CTmin differed significantly, suggesting different underlying physiological adaptations such as we have described here.

We have addressed this concern, noting the points above, in the section “Common Garden Design” in the Methods (lns 165 - 177).

2) Replication of RNA-seq

The standard practice of replication in RNA-seq experiments was developed for ANOVA to identify differences in expression among tissues and discrete conditions (e.g. disease states). For these studies, replication is absolutely essential. In contrast, ecologists are often interested in the response across a gradient, which is more directly modeled by a linear regression analysis (Cottingham et al. 2005). Variance in the expression profiles is estimated from the linear regression model. Replication at any particular temperature is not necessary, as power to detect a relationship is most improved by adding additional temperatures, not by replicating those that have already been assessed.  In fact, this approach results in a more precise variance estimate than if we had replicated only a few temperatures (Cottingham et al 2005). Only transcripts with a significant linear or quadratic relationship with temperature, after correcting for false discovery rate, were retained in the analysis.  Such analyses are just beginning to be used in a transcriptomic context (the only other example we are aware of is Hodgins-Davis et al 2012 GBE). To better clarify our approach for readers, we have added an expanded justification for the regression design in the Introduction  (lns 113-116):

To characterize the thermal reactionome, we characterized the reaction norm for each gene using a regression-based statistical approach to identify temperature-dependent patterns of change in gene expression. We used these response patterns to quantitatively test three mechanistic hypotheses of thermal adaptation.

and explanation in the methods  (lns 178-183):

	Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable (Cottingham et al 2005). Thus, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature.

2) Validation with qPCR

This is an important issue to which we gave careful consideration. For two reasons, we did not feel that the results of qPCR would improve this study.

First, RNA-seq is no longer a new methodology. Many studies have been published in ecological journals using RNAseq data without qPCR validation, such as Barshis et al. (2012 PNAS) and Smith et al. (2013 BMC Genomics).

Second, it is only feasible to perform qPCR on a small subset of transcripts. In our approach, we avoid focusing on the response of individual transcripts (e.g we *do not* provide a list of differentially-expressed genes). Instead, we focus on transcriptome-wide responses to temperature, which groups large numbers of genes into distinct response patterns. While our results are undoubtedly wrong for some individual transcripts, they are likely to be distributed at random in regards to these classifications. Thus, while qPCR may prove that the expression patterns of some transcripts are incorrect, it is unlikely to change the results for whole functional groups.

> Editor Comments to Author:
> 
> This interesting paper has received three careful reviews. The first reviewer recommended rejection, the second was enthusiastic about the paper, and the third sat somewhere in the middle. The first reviewer points out that the work does not tightly control for genetic versus plastic effects and therefore the paper does not precisely test the issues of interest to the authors. I suspect that this can be fixed with reworking of the paper in terms of interpretations, taking care to separate out the different components where appropriate. The reviewer also makes some important points about replication and the measurement of CTmax as well as the nature of control thermal treatments. The second reviewer was complimentary about the overall approach and had only minor comments, which cover some confusion stemming from the text and one of the Figures. Some apparent errors in the wording are also noted. The third reviewer also raises the issue of replication at the different temperatures and has an issue with the terminology in the paper and division into a predefined set of categories. The comments are worth taking on board. The reviewer also points out that some aspects of expression changes are likely to be missed, and calls for new qPCR data as is often carried out. The reviewer points out where results are inconsistent with the literature and novel patterns that appear not to be consistent with expectations, which all need discussion. The authors should take these comments on board when making interpretations of their data, and when interpreting the patterns established from their work. The issue of replication needs consideration along with measurement error and validation of expression patterns established from RNAseq.
> 
> Dr. Ary Hoffmann
> Subject Editor, Molecular Ecology

> Reviewer Comments to Author:
> Reviewer: 1
> 
> Recommendation: Reject
> 
> Comments:
> 
> The current paper by Stanton-Geddes et al generates “reactionomes” by generating RNA seq data  for two different species of ant at 12 different temperatures (0-38.5C).  The authors estimate CTmax in both species and link changes in their expression profiles to these estimates of CTmax and the environments where the species were collected. I think the authors were ambitious, the questions they present relevant and highly topical however, I believe there are significant design issues which limit the interpretation of the results.
> 
> First and fore mostly, if I understand the methods correctly, ants were collected from the field acclimated at 25C for 6 months and then experiments performed. While I understand that many species can be difficult to rear in the laboratory. Without performing a true common garden experiment (i.e both species of ants developed and reared under standard conditions) you simply cannot tease apart genetic or plastic differences in expression profiles. Are the observed differences between the species due to evolved differences related to thermal evolution or simply a consequence of developmental acclimation, by collecting cold adapted species from a cold environment and vice versa only confounds the results even more. One possible way that the authors may have alleviated this issue, to some degree, is to choose 2 species that vary in their distribution but where their distribution overlaps. This may be the case for these two species but the authors have not made that clear in their ms. If their species do overlap in distribution the authors should at the very least have collected these species from the same site. With such a design you could have then assumed that developmental acclimation may have been similar across the two species. Albeit you would still have GxE effects and these are likely to differ between species.

We address this concern as (1) above.

> My other major concern, which also depends critically on whether I have understood the methods correctly (see below comments) is that for the RNA seq the authors have pooled 3 ants per treatment (not a particular high number) and actually have no replicates per treatment. How do you estimate the variance in your expression profiles?

We address this concern as (2) above.

> Minor comments
> 
> -Although authors have characterized CTmax in the two species, how do these estimates of CTmax relate to the heat treatments that the authors use? Have the authors really examined a phenotype here? The authors CTmax estimates will also be confounded by developmental acclimation.
>

The maximum temperature we heated ants to was 38.5℃, below the CTmax for both A. carolinensis (~43℃) and A. picea (~42℃) (Fig 3. Warren & Chick 2013), so we have confidence that gene expression changes reflected active responses to thermal stress rather than dysregulation caused by loss of function. We have added this information to the “Common Garden Design” sub-section of the Methods  (lns 191 - 194).

Both colonies were maintained in the lab for 6 months prior to experimentation, so differences in developmental acclimation are as minimized as the biology of the species will allow, but it is of course true that we cannot rule out some residual developmental effect. To reflect this remaining potential source of phenotypic plasticity, we have attempted throughout the ms to make our word choices more precise in regard to whether we have demonstrated evidence for evolutionary differences.

> -The authors need to clearer about where in the distribution of the ant species they are sampling. Are they sampling them from their extremes? Middle?

We have clarified this in the Methods  (lns 154 - 155).

    These locations are centrally located within each species’ geographic range.

> 
> -Line 40-55 The end of this paragraph reads like methods
> 
> -Line 56-60 also reads like methods
>

We appreciate the reviewer’s concern, and removed some of the detailed information (lines 56-57 from draft 1), but we do feel feel that some methodological details are necessary here to explain how we tested our hypotheses.  We have attempted to reframe these such that the reason for their inclusion is more clear.

> -Line 57 it would be helpful to have a figure of the temperatures you used here, or at least have it written out in the methods
> 
> -I found the methods section confusing and not detailed enough
> 
> -Line 97 makes it sound like you collected each ant species form Molly Bog Vermont, but I don’t think that’s  what you mean I think you mean you collected a single colony of A. picea from Molly bog and a single population of A. carolinensis from Durham North Carolina

True - we have clarified this in the text  (lns 150 - 153):

	In fall 2012, we collected a single colony of Aphaenogaster picea from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of Aphaenogaster carolinensis, part of the A. rudis species complex (Umphrey 1996), from Durham, North Carolina (36.037° N -78.874° W).

> 
> -Why were you not able to determine the CTmax from the original,

It would have been ideal to have paired measures such as these, but both transcriptome and CTmax measurements require a large number of workers, and the colonies were simply not large enough to do both. We’ve added this information to ln 157 of the manuscript.

could you please describe your CTmax methods, there are many different ways this is calculated. It shouldn’t be necessary for me to have to look up another paper to understand your methods.

We have added this information as requested  (lns 160 - 163):

	We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves (Warren and Chick 2013).

> 
> -Line 109 To be clearer could the authors state that this part of the experiment was performed on ants collected in 2012

Done  (ln 150).

> 
> -The sampling and replicates are not clear Line 109-120. The authors collected 10 ants from each colony and exposed them to a single random temperature on a given day for a period of 12 days. Ants were then pooled into 3 which would make me think you had 10 ants per species per treatment you pooled them into 3 to make almost 4 replicates but on line 122 you say for each species the 12 samples were barcoded?

We apologize for the confusion. We collected 10 ants so that we would have extra samples if necessary, though only 3 were used in pooling (per temperature) for sequencing. For clarity, we now simply refer to 3 ants.

> 
> -Did the authors ensure these treatments did not cause mortality?

Yes. We have added this to the text (lns 191 - 193).

> 
> -Should the authors have a control here i.e 24.5 and look at species specific expression changes?
>

It’s unclear what the reviewer means here. A major advantage of the regression approach is that we have quantitative information on transcript expression in response to temperature and can predict expression at 24.5, or any other temperature, from the model of each transcripts’ reaction norm.  Species-specific expression changes are identified by a significant species x temperature interaction.  We have adopted the reviewer’s suggestion below to use 24.5 as our baseline value (see response to comment for line 222).

> -Line 162 I am unsure what the authors are suggesting here, that the data was odd so the got rid of it?

As you can see in this figure, which we now include in the Supporting Information, the “Ar_7” (Ar = A. carolinensis) sample clusters with all the other A22 (= A. picea) samples, while the “A22_7” sample clusters with all the other Ar samples. The misplaced samples are marked with red arrows. The most parsimonious explanation for this is a labeling mistake. Removing these samples, rather than relabeling, is the conservative approach.

> 
> -Line 169 Rather than writing out formula could you at least include a description of your model. Where species considered random or fixed effects? You can run mixed model glm’s in R and maybe this is what you did but your description is unclear. It is also unclear to me what temp and temp2 are in the model, this is not explained anywhere .
>

The formatting of this equation was lost in the file conversion, temp2 should have been temp2 indicating that we were fitting a second-degree polynomial regression. All effects were fixed as temperature is continuous and we only had two species. We have clarified in the text (lns 250 - 255):

	Temperature and species were both fixed effects, with a quadratic term included for temperature.

> -Line 180-184 I am unsure what exactly you are performing the step wise regression on?

The statistical significance of the species and linear and quadratic temperature terms in the model were tested via step-wise model selection.

> -Line 188-192 What about down regulated transcripts?

In a reaction norm context, downregulation is the inverse of up-regulation. That is, transcripts downregulated at low temperatures fall into the *High* category, transcripts downregulated at high temperatures fall into the *Low* category.  Splitting the temperature treatments into two at the midpoint to consider only low or only high temperatures would potentially have allowed us to make distinctions between upregulation and downregulation in these cases, but with only 12 temperatures in total, we would unfortunately be left with too few temperatures in each case to estimate the model.  We have clarified this in the sub-section “Identification of thermally-responsive transcripts”  (lns 279 - 284)

Because expression category was defined by the temperature of maximal expression, both Low and High categories were biased toward transcripts up-regulated at that temperature extreme, but also likely included some transcripts down-regulated at the opposing extreme. The two categories which could unambiguously distinguish up- from down-regulation are Bimodal (up at both extremes) and Intermediate (down at both extremes).

> 
> -The authors grouped their transcripts by expression level and temperature, the temperature categories need better justification

The temperature categories were decided a priori to allow us to apply the results to the biological hypotheses we were testing. This is elaborated in the sub-section “Identification of thermally-responsive transcripts”  (lns 271 - 284):

	We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined a priori to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species...

> 
> -Line 222 It seems odd that the authors have used 19.5C as their baseline for constitutive expression when ants were reared for 6 months at 25C. It seem 24.5 C would be a more suitable control.

This is a good point that we had not considered.  As suggested by the reviewer, we changed the baseline for constitutive expression to the rearing temperature of 25℃. See commit afdd69d on github (https://github.com/johnstantongeddes/ApTranscriptome). We use this version of the results in the revised manuscript, though it has no qualitative effect on the major results.

> 
> -Line 322 the authors start talking about down regulation but as I understood their categories high was high expression at high temperatures, low was high expression at low temperatures etc. Maybe I have misunderstood something here?

See our comment above.

> 
> -Line 424 It is unclear what the authors are trying to say here are they suggesting you can link the number of responsive genes to the level of additive genetic variation?
>

We were presenting a post-hoc explanation for the observation that many more genes were responsive to cold than warm temperature extremes. As we emphasized in the text (e.g. To the extent that …) this was a speculative explanation for this observation, though consistent with other work (Kellerman et al 2012, Warren and Chick 2013). However, given the uncertainty about the explanation we have chose to remove this section from the manuscript.

> 
> 
> Reviewer: 2
> 
> The authors of this study characterize the thermal reaction norms of all genes in a de novo transcriptome assembly across 12 temperatures spanning the entire thermal breadth of two species of ants: the cool-climate Aphaenogaster picea and the warm-climate Aphaenogaster carolinensis. In the past, studies exploring transcriptomic responses to temperature have mostly used ANOVA-type experimental designs, with only a few temperature levels. But continuous variation is much better characterized with the reaction norm approach used in this study, describing variation among genes and among organisms in the shape of expression responses across temperature. I really enjoyed this paper -- the approach taken here is exciting, and I found the manuscript to be quite well written, with clearly defined hypotheses an expectations. The comparative reaction norm approach taken here is, to my knowledge, unprecedented, and will make an important contribution to the field of thermal biology. I strongly recommend this manuscript for publication (and look forward to its acceptance so I can cite it!). I have only a few minor comments and revisions, outlined below.
>

We really appreciate these comments. While this study isn't perfect, we also believe that this study is pushing back the boundaries of our knowledge of thermal biology.

> Lines 61-86. As the authors note, studies exploring transcriptomic responses to stress have sometimes been prone to storytelling, so I appreciate the specific predictions generated by each of the three potential mechanisms of stress adaptation.
>

This problem of storytelling is rife in the ecological genetic literature, and one of our motivations for writing this paper was to demonstrate how similar methods can be analyzed in an alternate approach to gain novel insight.

> Figure 1
> 
> -Figure 1 is key to the whole paper, but connecting the results to the predictions generated by each of the three hypotheses made my head hurt a little. I wonder if there is a way to make it easier for the reader to draw inferences from this figure? One possibility would be an inset cartoon showing the results that could be expected in support of each of the competing hypotheses? Also maybe include little line cartoons along the axes depicting each of the norms of reaction.
>

This is a useful suggestion and we have made a figure (Fig. 1) with illustrations of the expected results under the *enhanced response* and the *tolerance* hypotheses that can be compared to the true results.

> The categories “High” “Low” “Intermediate” and “Bimodal” are defined well in the manuscript, but it’s always nice when key figures can stand alone without reference to the text.
>

We define these in the caption to Table 1, and refer to this in Figure 1 rather than repeating. We could also switch this if preferred.

> - What are the units in the color scale?

The units are number of transcripts. We have updated the legend to clarify this:

	...the color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts.

> 
> 196-197. I’m missing something about the methods - lines 196-197 state that the analyses focused on transcripts with a significant temperature x species interaction but in Figure S1, 4/5 of the boxes on the diagonal are blue (= excess of transcripts in the same reaction norm category for both species). Shouldn’t these have been excluded by focusing only on transcripts where temperature effects differed between species?
>

We apologize for being unclear in the text. For the analyses of the how the number of responsive transcripts differs between species (first paragraph under “Statistical Analyses”), we used all responsive transcripts, both those with a significant temperature x species interaction and those with a main effect of temperature only. The justification for including all thermally-responsive transcripts is that the enhanced response hypothesis explicitly predicts that the genes upregulated by the species that sees the stressor infrequently should retain that function in the other, while under the tolerance hypothesis they should not.

In contrast, for the comparisons of the *temperature of gene activation* (second paragraph under “Statistical Analyses”), only transcripts with a significant temperature x species interaction are used as, by definition, there is no difference in expression levels between species 
We have changed the text (lns 292-295) to clarify this.

	To test whether the temperature at which thermally-responsive transcripts were activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species, using only the transcripts with a significant species x temperature interaction.

> -207-216 –is t test appropriate here, given that data are categorical rather than continuous? (Expression is observed at 12 distinct temperatures).
>

In this case, we aren't testing expression across the 12 temperatures. For each transcript, we used the predicted expression values from the statistical model to identify the temperature where expression had the greatest change (the temperature of transcript activation). We then compared the vectors of temperature of activation (a continuous variable) between species using the t-test. We now explain this in the text (lns 297 - 299):

	We then performed a t-test to determine if the mean temperature of transcript activation differed between the two species for each group.

> -235-245 Something’s not adding up about the number of transcripts in each category 237-239 says, “Of the Formicidae hits, we identified 9,052 transcripts, or 8% of the total, with expression patterns that were thermally responsive” (implying that the 9052 thermally responsive transcripts all had hits to Formicidae) But then later in the paragraph, “The proportion of responsive transcripts is the same if we focus only on those transcript (4,618 transcripts out of 51,246).” (Implying that only half of the thermally responsive transcripts had blast hits). 
>

This was our mistake as the phrase “Of the Formicidae hits..." was left in unintentionally from a previous draft. The percentage of responsive transcripts is out of the total.

> 414-418 “In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures (Kingsolver et al. 2013; Kelly et al. 2013).” I would argue that the data show reduced plasticity of gene expression at higher temperatures rather than reduced genetic variation (= variation among individuals).
>

As this section raised concern from multiple reviewers and is not essential to our main conclusions, we have removed it from the manuscript.

> Reviewer: 3
> 
> Recommendation: Accept, minor revisions
> 
> Comments:
> 
> MEC-14-1243- MS TITLE: Thermal reactionomes reveal adaptive responses to thermal extremes in warm and cool-climate ant species
> 
> This paper brings valuable contribution to our understanding of thermal adaptation of two Ant species living in contrasted thermal regions. The authors found that a significant part of the transcriptome was thermally responsive. For both species, more transcripts responded to low than high temperatures, a result that I find counterintuitive and unexpected. The authors also found differential thermal-responsive transcriptional responses between the two species suggesting different thermal adaptation (and putative underlying mechanisms).
> 
> The authors present a set of experiments with insects that were assessed with a range of thermal conditions (12), ranging from lower to upper thermal limits, as well as transcriptional changes in reaction to temperature exposure (RNA seq). The topic is timely with the increasing interest in the molecular and physiological basis of response to temperature/climate change. The nature of cold response/tolerance is highly complex which makes the effort to unravel it very difficult.
> 
> Experiments were carefully conducted under well controlled conditions. The paper is logically arranged and well written, in good English. The methods used in this study were appropriate and the experiments were correctly conducted which makes the data set reliable. I also particularly appreciated that this is an hypothesis-driven study.
> 
> In spite of this, I have a number comments and queries that needs to be answered or at least discussed (see below).
> 
> -Abstract and elsewhere in MS : “enhanced defense hypothesis” . I find the term “defense” a bit confusing as it often used for immune process (defense genes …). “Stronger induced response hypothesis” could be an alternative.
>

We agree with this comment and have changed the ‘enhanced defense’ to the ‘enhanced response’ hypothesis throughout the paper.

> -L77: please provide factual examples for “shift in alternate suite of tolerance genes”

We have added citations to Neelakanta et al (2012), who showed that inducing expression of an antifreeze glycoprotein can increase the ability of fruit flies to tolerate cold, and Franssen et al. (2011), who showed that a southern population of Zosteria marina is more resilient to heat shock than a northern population (ln 131).

> 
> -L61-86: I am not totally convinced that the whole set of thermal-responsive genes will coordinately change in such way that it will fit to one of the three proposed mechanistic hypotheses of thermal adaptation (1 defense, 2 tolerance or 3 assimilation). I assume that transcriptional response will be more complex and gene-specific. Some genes will fit the first hypothesis, while other will fit the others. Please comment.
>

The reviewer is almost certainly correct, and it is our goal to determine the relative proportion of the transcriptome that falls into each category. We state this in the final sentence of the paragraph (lns 138-139:

Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify if, and under what conditions, these mechanisms contribute to thermal adaptation.

> -L107: why such a small number of individuals (5) were used to monitor thermal limits?
>

A previous study (Warren and Chick 2013) used larger sample sizes (10 ants per colony) to demonstrate differences in the thermal limits of these species. Our measurement of thermal limits were simply a confirmation of their results.

> -L109-120: Ten ants were exposed to 12 different temperatures for each species. I may have misunderstood but it is not clear to me if there were replicates for each temperature?

Please see responses to reviewer 1 on this topic.

> 
> -L109-115: Twelve temperatures were used to establish a reaction norm of transcript expression patterns. Expression was checked at a single time point : after one hour of temperature exposure. It is well known that many thermal-responsive genes show temporal patterns of expression peaking at various time after exposure (see Sorensen 2005 Cell Stress & Chaperones  10, 312–328). Here, interpretations are based on early responsive genes (1h of exposure), which might only explain a small fraction of molecular response to temperature. Instead of taking 12 temperatures, it could have been of interest to check temporal patterns, perhaps with less thermal conditions. In fact, were 12 temperatures truly necessary to establish a reaction norm? My point is that this study should acknowledge that a large suite of thermally-responsive transcripts are overlooked and cannot be captured by the experimental design.

The temporal dynamics of expression response to thermal response are clearly crucial. We acknowledge this complexity by specifying that our reactionome is focused on capturing a wide temperature gradient, at the expense of having to focus only on a single time-point in the temporal response. Specifically, after defining the reactionome in the Intro we state (lns 94-97:

	Although temporal patterns of transcriptional activity (e.g. fast- vs. slow- responding genes) are also important components of an organism's transcriptional response to environmental conditions (Sørensen et al. 2005), we focus here on the response of transcripts across conditions at the same time point.

And in the Methods (ln 197):

	Thus, our reactionome characterized early, but not late, responding genes.

Finally, in the section on caveats in the Discussion (ln 554-556):

	… species may differ in gene expression along axes which we have not measured here, primarily temporal patterns of gene expression (Sorensen et al. 2005).

> 
> -Microarray and NGS data are regularly validated by qPCR. I am surprised that authors did not include validation step (on 5-10 genes) in their experiments. Previous studies have shown extremely close correlations between qPCR and RNAseq data [1-4]. Ideally, the authors should re-validate their findings (potentially by qPCR) in a separate cohort of samples.

Please see our general comments above.

> 
> 1.Griffith M, Griffith OL, Mwenifumbo J, Goya R, Morrissy a S, et al. (2010) Alternative expression analysis by RNA sequencing. Nat Methods 7: 843–847. doi:10.1038/nmeth.1503.
> 
> 2.Asmann YW, Klee EW, Thompson EA, Perez E a, Middha S, et al. (2009) 3’ tag digital gene expression profiling of human brain and universal reference RNA using Illumina Genome Analyzer. BMC Genomics 10: 531. doi:10.1186/1471-2164-10-531.
> 
> 3.Wu AR, Neff NF, Kalisky T, Dalerba P, Treutlein B, et al. (2014) Quantitative assessment of single-cell RNA-sequencing methods. Nat Methods 11: 41–46. doi:10.1038/nmeth.2694.
> 
> 4.Shi Y, He M (2014) Differential gene expression identified by RNA-Seq and qPCR in two sizes of pearl oyster (Pinctada fucata). Gene 538: 313–322. doi:10.1016/j.gene.2014.01.031.
> 
> -L197: comparative analyses between species were performed by focusing on transcripts that had species x temperature interaction. What about genes that had unique profile, that is, genes that were species-specific? These might contribute to thermal adaptation. How these genes were treated?

The species-specific genes were included in the analysis. In Table 1, these are the genes that were expressed in one species, but were “Not Responsive” or missing in the other species. We discuss some of these genes in the paragraph in the Discussion that begins “In contrast to cold tolerance, the enhanced thermal limit in A. picea is best explained by the tolerance hypothesis…” (ln 469 - 489).

> 
> -L196-201 & 207-216. It wasn’t clear to me why authors arbitrary defined 5 expression patterns (categories) ? Authors could have use k-means algorithm that can be used to partition the input data set into k partitions (clusters). In such way, main groups of temperature-responsive genes could be identified by K-mean clustering without a priori, and this would also provide statistical measures of over-representation (EASE scores).
>

The expression categories were not arbitrary, but represent a priori predictions that are biologically meaningful and can be related to the predictions of the three hypotheses (lns 272 - 275). While statistical clustering may be valuable, it would make hypothesis testing difficult. What would one do with a cluster that increased at 26 degrees, decreased at 29 and then leveled off, for example? This could be explored in another paper, but is not a good analytical fit with the goal here. Further, we note that Reviewer 2 found our focus on specific predictions explored by the specific expression patterns to be a strength of the paper.

> -L210. it seems that authors only focused on genes that were upregulated to fall into 5 categories. What about genes that showed downregulation in expression patterns ?
>

Please see our response to reviewer 1 on this topic.

> -L258: it is rather surprising that majority of transcripts were upregulated at the low temperature, especially since there was no recovery to allow transcriptional machinery to operate. It has repeatedly observed that at low temperature, close to lower thermal limit, transcriptional activity is very weak. For instance, Teets et al. (Physiol. Genomics 44, 764–777, 2012) did not observe any differentially expressed transcript during the cold treatment and Vesala et al. (Insect Mol. Biol. 21, 107–118.2012) only found a few transcripts (3 or 0) modulated by cold in D. montana and D. virilis, respectively. Please comment.
>

This is an interesting observation that actually strengthens our reactionome approach, as we can quantify not only what is upregulated but the temperature at which that response occurs. We now comment on this in the Discussion (lns 505 - 515):

	In both species, there were 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1). The degree of upregulation at low temperatures is surprising given previous studies (Vesala et al. 2012; Teets et al. 2012b) that found little transcriptional activity at low temperatures. However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures. At these extreme low temperatures we also found few upregulated transcripts (Fig. 3A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig. 3). A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis (Addo-Bediako et al. 2002).

> -L362-379: A.picea is more cold-adapted than A. carolinensis. A.picea also shows more transcripts being expressed at cold than A. carolinensis. Hence, this correlated pattern is assumed to support enhanced defense hypothesis, that is, stress-adapted species show stronger response. This discussion is based on global expression patterns of hundreds of genes that show expression variation according to temperature. The metrics used here is the number of responsive transcripts. These global patterns are assumed to determine differential thermal adaptation between species. It is certainly interesting to look a general patterns and numbers of transcripts, but this bulk of transcripts obviously include a number of genes that have nothing to do with thermal response or are indirect consequences of exposure. My point is that general trends cannot explain all. Some genes are functionally essential for thermal response, while other are not, and I wonder whether looking a GOs and pathways would not be more functionally meaningful. I am not fully convinced that higher number of cold-responsive transcripts directly translates higher cold hardiness.
>

We completely agree that general trends cannot completely explain anything, but studies that simply report lists of differentially-expressed genes without strong functional work, as are rife in the field, also have substantial explanatory weaknesses. This project focused on broad patterns, testing alternative mechanistic hypotheses (enhanced response, tolerance, genetic assimilation) that did not depend on the response of any specific GO terms or pathways. However, as suggested, we do look at GO terms that are enriched for the global responses - see the section “Gene set enrichment analysis” in the methods, and subsequent discussion of GO terms (metabolism, transcript regulation, etc) in the Results and Discussion. The results of these analyses, which again are prone to storytelling, are used to support the tests of our specific hypotheses. As as example, we find the result that in response to thermal extremes, the warmer-climate species shuts down metabolism-related genes while the cool-climate species increases metabolism and damage repair responses, including  7 heat-shock proteins, to be quite compelling (third paragraph of Discussion, lns 476 - 482).

> -L388-390: More Hsp transcripts are expressed in A. carolinensis compared to A.picea when exposed to high thermal limit. This is assumed to support tolerance hypothesis. Interpretations depend on whether Hsp expression is viewed as stress reactive or protective effect.
>

Our main line of evidence for the tolerance hypothesis in A. carolinensis comes from requiring higher temperatures to get a transcriptional response than A. picea. That is, A. carolinensis tolerates greater heat before mounting a transcriptional response. This indicates that there must be other mechanisms that support this tolerance. We note that A. carolinensis activates fewer heat shock proteins than A. picea, and suggest other mechanisms (proteome stability, behavioral quiescence) that may underlie A. carolinensis’ increased thermal tolerance.

> -L413-418 : More transcripts are expressed at lower temperature. This pattern is explained by greater genetic and phenotypic variation at the lower temperature extreme.  Does more genetic and phenotypic variation necessary translates into more genes being responsive?
>

See our response above - we have removed this section.

> -L492-496:  It is mentioned that A. Picea responds by increasing metabolism and activating heat shock proteins and other protective molecules. I am not sure that transcriptomics can reach such a level of interpretation. RNAseq shows molecular correlates but can certainly not show direct evidence that proteins abundance and activity or more generally pathways and active molecules are actually altered.

We have changed this statement to (lns 593 - 596):

	A. picea responds by increasing expression of transcripts related to metabolism, heat shock proteins and other protective molecules, whereas A. carolinensis decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance.

> Transcriptomics targets genes expression which is situated upstream of the functional molecules. Moreover, it does not account for downstream regulations, and genes and proteins expression might operate with considerable time lag. The molecular machinery provide ample possibilities for post-transcriptional modifications and regulation, so that a direct relationship between mRNA and protein levels cannot be assumed or expected. In addition, mRNA fluctuations (birth/death) are considered as an important source of stochasticity in gene expression, especially when mRNA abundance is low.

This point is absolutely correct, and we briefly acknowledged it in the line

	Finally, the mapping of changes in gene expression to organismal fitness is far from direct (Feder & Walser 2005), and large differences in patterns of gene expression may have only small effects on organismal fitness.

We have added the reviewer’s caveats (lns 566 - 569):

	In particular, functional protein levels cannot be expected to be fully linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability (Feder & Walser 2005).

### 2015-07-02 ApTranscriptome_TR.Rmd
Final edits for response. Added response letter

--- 2015-06-03 ApTranscriptome_TR.Rmd
@@
 ---
-output: word_document
+output: html_document
 ---
-Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
+
+Technical Report for *Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species*
 ========================================================================================
@@
 
-**October 14, 2014**
-
-**Technical Report No. 3**
+**June 24, 2015**
 
@@
 
-```{r setup, echo=FALSE, results='hide', message = FALSE}
+```{r setup, echo=FALSE, results='hide', message = FALSE, warning = FALSE}
 # Global settings
 options(stringsAsFactors=FALSE, digits = 3, scipen = 1)
+options(width = 80)
 
@@
 
-In this technical report, which accompanies the manuscript **Patterns of stress tolerance versus resistance in genome-wide expression data of parapatric ant species** (Stanton-Geddes et al., in press), we:
-
-1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
-2. identify thermally-responsive genes
-3. evaluate differences in reaction norms of expression for each thermally-responsive gene between the two species
-4. perform gene set enrichment analysis of thermally-responsive genes for the two species
-
-This script is completely reproducible assuming that R, `knitr` and the other required libraries (listed within the source document) are installed on a standard linux system using the following:
-
-    Rscript -e "library(knitr); knit('ApTranscriptome_TR.Rmd')"
+In this technical report, which accompanies the manuscript **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**, we:
+
+1. Describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`.
+2. Measure the reaction norm across a thermal gradient (0°C to 38.5°C) for all transcripts and characterize the *thermal reactionome* as the set of transcripts that are thermally-responsive.
+3. Quantitatively test three mechanistic hypotheses of thermal adapation:.
+  - The *Enhanced response* hypothesis proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
+  - The *Tolerance hypothesis* proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency.
+  - The *Genetic assimilation hypothesis* proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.
+
+We quantify the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species.
+We use gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.
 
@@
 The raw Illumina fastq files are available from the NCBI short read archive [link tbd].
-The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.
+The assembled transcriptome, annotation and expression values are available from the Harvard Forest Archives, HF-113 [http://harvardforest.fas.harvard.edu/data-archive] datasets [hf113-36](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-36-ap-ann-table.csv), [hf113-40](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-40-ap-trans.tar) and [hf113-41](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-gene-exp-quant.tgz).
+These processed files are used for this analysis due to the computational demands of re-generating the files, but the exact commands for each of these steps are documented below.
 
@@
 carolinensis*.
-For historical reasons, I sometimes refer to these species as colonies at times throughout this technical report.
-
-For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5C ranging from 0C to 38.5C, for one hour in glass tubes in a water bath.
+
+For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5°C ranging from 0°C to 38.5°C, for one hour in glass tubes in a water bath.
 The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen).
@@
 ~~~
-java -jar trimmomatic-0.30.jar PE -threads 40 -phred33 -trimlog trimmomatic.log sample.R1.fastq sample.R2.fastq sample.R1.trimmed.paired.fastq sample.R1.trimmed.unpaired.fastq sample.R2.trimmed.paired.fastq sample.R2.trimmed.unpaired.fastq ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:36
+java -jar trimmomatic-0.30.jar PE -threads 40 -phred33 -trimlog trimmomatic.log \\
+sample.R1.fastq sample.R2.fastq sample.R1.trimmed.paired.fastq sample.R1.trimmed \\
+.unpaired.fastq sample.R2.trimmed.paired.fastq sample.R2.trimmed.unpaired.fastq \\
+ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:36
 ~~~
@@
 Unpaired reads resulting from Trimmomatic were concatenated to the set of left (R1) reads per Trinity usage documentation.
-Assembly on a single 40 CPU machine with 1TB of memory required 20 hours of compute time.
+Assembly on a single 40 CPU machine with 1TB of memory required about 20 hours of compute time.
 
 ~~~
-Trinity.pl --seqType=fq --JM=90G —left=all.R1.trimmed.fastq —right=all.R2.trimmed.fastq --output=trinity --CPU=40 --inchworm_cpu=40 --bflyCPU=5
+Trinity.pl --seqType=fq --JM=90G —left=all.R1.trimmed.fastq —right=all.R2.trimmed.fastq \\
+--output=trinity --CPU=40 --inchworm_cpu=40 --bflyCPU=5
 ~~~
 
-This assembly contained 100,381 unique components (roughly genes) in 126,172 total transcripts (Table 1).
+This assembly contained 100,381 unique components (~genes) in 126,172 total transcripts (Table 1).
 
@@
 # make table
-trinity <- c("126,172", "100,389,539", "358", "795", "16,201", "1,631")
-reduced <- c("105,536", "62,648,997", "320", "593", "15,491", "895")
-
-assemstats <- rbind(trinity, reduced)
+Trinity <- c("126,172", "100,389,539", "358", "795", "16,201", "1,631")
+Reduced <- c("105,536", "62,648,997", "320", "593", "15,491", "895")
+
+assemstats <- rbind(Trinity, Reduced)
 colnames(assemstats) <- c("Total contigs", "Total length", "Median contig size", "Mean contig size", "N50 contig", "N50 Length")
@@
 
-    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust -dbs hsref,bast,vir,arch
+    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust \\
+    -dbs hsref,bast,vir,arch
 
@@
 
-Running Trinity and subsequent programs is time and memory-intensive so the final assembly can be downloaded and used for all further analyses.
-In addition, the compressed archive contains the "clean" and "contaminant" sequences after filtering with DeconSeq.
+Running Trinity and subsequent programs is time and memory-intensive so the final assembly is downloaded and used for all further analyses.
+The downloaded archive contains the "clean" and "contaminant" sequences after filtering with DeconSeq.
 
@@
 # download filtered Trinity assembly, uncompress and move
-wget http://johnstantongeddes.org/assets/files/ApTranscriptome/Aphaenogaster_transcriptome.tar
-# check md5sum
-md5sum Aphaenogaster_transcriptome.tar
-# fd6dbb0b3e88e1c797c9e74611b245b2
+wget http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-40-ap-trans.tar
 
@@
 mkdir -p results/trinity-full/
-mv Aphaenogaster_transcriptome.tar results/trinity-full/.
-tar -xvf Aphaenogaster_transcriptome.tar
+mv hf113-40-ap-trans.tar results/trinity-full/.
+tar -xvf hf113-40-ap-trans.tar
 ~~~
@@
 
-Results are available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).
-
-This annotation file can be read directly to R:
+Results were available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).
+
+This annotation file can be downloaded from the Harvard Forest [archive](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-36-ap-ann-table.csv), or read directly into R.
 
@@
 annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome/ApTranscriptome_AnnotationTable_20140113.txt")
+
 # load 
@@
 
-### Quantify gene expression ###
+### Gene expression ###
 
@@
 # download gene expression quantification
-wget http://johnstantongeddes.org/assets/files/ApTranscriptome/sailfish_quant_20140916.tar
+wget http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-gene-exp-quant.tgz
+# http://johnstantongeddes.org/assets/files/ApTranscriptome/sailfish_quant_20140916.tar
 # check md5sum
@@
 
-Note that expression levels at each temperature treatment are highly correlated between the two colonies.
-
 ```{r exp_correlations, echo=FALSE, results='hide'}
@@
 ```
+
+\newpage
+
+Note that expression levels at each temperature treatment are highly correlated between the two colonies.
 
@@
 
-### Grouping of thermally-responsive transcripts
-
 The set of transcripts with significant expression patterns include those with expression that differs by species, temperature and the interaction of species and temperature.
-In this section, I am specifically interested in the thermally-responsive transcripts (temperature and species x temperature) so I subset the significant transcripts to examine these.
+I am specifically interested in the thermally-responsive transcripts (temperature and species x temperature) so I subset the significant transcripts to examine these.
 
@@
 colony.lms <- other.lms[setdiff(names(other.lms), names(temperature.lms))]
+
+# combine all responsive transcripts
 responsive.lms <- c(temperature.lms, interaction.lms)
@@
 
-### Thermal-response functional types ###
-
-The previous section identified the transcripts with thermally-responsive expression. In this section, I determine the shape of the expression response to temperature for each transcript. Catego
-ries of expression response are:
-
-* High - increase expression with temperature
-* Low - decrease expression with temperature
-* Intermediate - maximum expression at intermediate temperatures (14 - 28C)
-* Bimodal - expressed greater than one standard deviation of expression at both low and high temperatures
-
-I do this first for the thermally-responsive transcripts where there is no interaction with species.
-For the transcripts where thermal-responsive expression depends on species, I determine the functional type of the expression response separately for each species.
+I checked the proportion of responsive transcripts against those that have a BLAST hit.
+
+```{r responsive_with_BLAST}
+annotation.table.responsive <- annotation.table[which(annotation.table$best.hit.to.nr != "-"), ]
+
+responsive_transcripts_w_BLAST <- length(which(unique(resp.TPM.dt.sub$Transcript) %in% annotation.table.responsive$Sequence.Name))
+
+responsive_transcripts_w_BLAST / round(nrow(annotation.table.responsive))
+```
+
+### Thermal-response expression categories ###
+
+The previous section identified the transcripts with thermally-responsive expression.
+In this section, I determined the shape of the expression response to temperature for each transcript.
+Categories of expression response are:
+
+- **High** - increase expression with temperature
+- **Low** - decrease expression with temperature
+- **Intermediate** - maximum expression at intermediate temperatures (14 - 28C)
+- **Bimodal** - expressed greater than one standard deviation of expression at both low and high temperatures
+- **Not Responsive** - not thermally-responsive in one species but does respond in the other
+
+For the transcripts where thermal-responsive expression depends on species, I determined the functional type of the expression response separately for each species.
 
@@
 
-Next, I compare the number of thermally-responsive in each response category between the two colonies.
+Next, I compared the number of thermally-responsive in each response category between the two colonies.
 
@@
 (Ap.type.table <- rbind(A22.type.table, Ar.type.table))
+
+# proportion
+Ap.type.table.prop <- round(Ap.type.table / (sum(Ap.type.table)/2) * 100)
 
@@
 
-To identify specific cells that deviate from null expectation, calculate expected observations assuming that marginal frequencies equal the overall frequency for each expression category, then determine which cells have deviations between observed and expected that are greater than the overall X^2^ statistic.
+To identify specific cells that deviate from null expectation, I calculated the expected observations assuming that marginal frequencies equal the overall frequency for each expression category, then determined which cells have deviations between observed and expected that are greater than the overall X^2^ statistic.
 
@@
 
-To visualize this, make a heatmap that shows deviation from expected count.
+To visualize this, I made a heatmap that showed the deviation of the actual number of transcripts in each cell from the expected count.
 
@@
 
-Per suggestion of a reviewer, make a cartoon figure that shows the contrasting results of the *enhance defense* and the *tolerance* hypotheses.
-
-The *enhanced response hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response.
+The *enhanced response hypothesis* posits that temperature adaptation uses existing response mechanisms, which should result in significant overlap in response.
 This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g.
@@
 
-```{r ed_cartoon}
-enhanced_defense_dat <- md_sub1
-enhanced_defense_dat$value <- 0
+```{r er_cartoon}
+enhanced_response_dat <- md_sub1
+enhanced_response_dat$value <- 0
 # set non-shared response to random negative value 
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min = -300, max = 0)
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- runif(5, min = -300, max = 0)
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Low"), "value"] <- runif(5, min = -100, max = -150)
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Bimodal"), "value"] <- runif(5, min = -100, max = -150)
 # change shared response to positive
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 500
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 300
-
-enhanced_defense_dat
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Low" & enhanced_response_dat$Apic == "Low"), "value"] <- 10
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Bimodal" & enhanced_response_dat$Apic == "Bimodal"), "value"] <- 10
+
+enhanced_response_dat
 
 # plot
-enhanced_defense_plot <- qplot(y=Apic, x=Acar, data=enhanced_defense_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_defense_dat$Acar))) + 
+enhanced_response_plot <- qplot(y=Apic, x=Acar, data=enhanced_response_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_response_dat$Acar))) + 
   scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
@@
 
+enhanced_response_plot <- ggplot(enhanced_response_dat, aes(y = Apic, x = Acar, fill=value)) + 
+  geom_tile(colour = "#ffffff", size =1) +
+  xlim(rev(levels(enhanced_response_dat$Acar))) + 
+  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
+  theme(axis.title = element_text(face="italic")) +
+  labs(y = "A. picea", x = "A. carolinensis") +
+  theme(legend.position = "right") + 
+  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
+  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
+
 # title
-enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
-enhanced_defense_plot
-```
-
-In contrast, the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.
+enhanced_response_plot <- enhanced_response_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0, size = 10))
+enhanced_response_plot
+```
+
+In contrast, the *tolerance hypothesis* predicts that genes involved in active response will become non-responsive or shift to other response categories in the better-adapted species.
 
@@
 # set non-shared response to random positive value 
-tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min = 100, max = 300)
-tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min = 100, max = 200)
+tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min = 250, max = 300)
+tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min = 250, max = 300)
 # change shared response to near zero
-tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- 30
-tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- 20
+tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- -100
+tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- -100
 
@@
 # plot
-tolerance_plot <- qplot(y=Apic, x=Acar, data=tolerance_dat, fill=value, geom="tile", xlim = rev(levels(tolerance_dat$Acar))) + 
+tolerance_plot <- ggplot(tolerance_dat, aes(y=Apic, x=Acar, fill=value)) + 
+  geom_tile(colour = "#ffffff", size =1) +
+  xlim(rev(levels(enhanced_response_dat$Acar))) + 
   scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
@@
 # title
-tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
+tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0, size = 10))
 tolerance_plot
@@
 png("results/predictions_Fig1.png")
-grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=2, widths = c(3.75,3))
+grid.arrange(enhanced_response_plot, tolerance_plot, ncol=2, widths = c(3.7, 3))
 dev.off()
@@
 
-The number of thermally-responsive in each response category differs between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies.
+The number of thermally-responsive in each response category differed between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies.
 For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts.
@@
 Finding that the mean expression level of these genes in *A.
-carolinensis* is greater than the expression level at the optimum temperature (19.5°C) in *A.
+carolinensis* is greater than the expression level at the rearing temperature (25°C) in *A.
 picea* would be consistent with genetic assimilation.
@@
 
-Next, I check the same for the **Bimodal** genes in *A. picea*.
+Next, I checked the same for the **Bimodal** genes in *A. picea*.
 
@@
 
-As with the **High** transcripts, expression much higher for the responsive *A. picea* transcripts than for the non-responsive *A. carolinensis* transcripts.
+As with the **High** transcripts, expression was much higher for the responsive *A. picea* transcripts than for the non-responsive *A. carolinensis* transcripts.
 
@@
 ```{r responsive_pred}
+# subset to transcripts with significant temperature x species interaction
+resp.TPM.dt.sub <- TPM.dt.sub[names(interaction.lms)]
+
 # apply predFunc to all responsive transcripts
@@
 The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.
+The web app is available at https://johnsg.shinyapps.io/ApRxN-shinyapp/
 
@@
 
-save(sigtable, chi1, chi2, chi3, chi4, chi5, chi6, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, md, file = paste("doc/ms_slim_", Sys.Date(), ".RData", sep = ""))
+save(sigtable, chi1, chi2, chi3, chi4, chi5, chi6, resp.TPM.dt.sub, Ap.type.table, Ap.type.table.prop, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, md, responsive_transcripts_w_BLAST, annotation.table.responsive, file = paste("doc/ms_slim.RData", sep = ""))
 ```

### 2015-07-02 doc/ApTranscriptome_manuscript.Rmd
Final edits for response. Added response letter

@@
 output:
-  word_document: default
   pdf_document:
-    fig_caption: yes
+    fig_caption: true
     includes:
@@
     toc_depth: 2
-  html_document: null
+  html_document:
+        keep_md: true
+  word_document: default
 includes: header.tex
@@
 However, little is known about how thermal tolerance evolves.
-To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
+To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*: the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient.
 We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth.
-We found that at least 8% of all genes were thermally responsive.
-The majority of upregulation was specific to low temperatures, and consistent with *enhanced response* the cool-adapted *A.
-picea* induced expression of more genes in response to extreme temperatures than *A.
-carolinensis*, including heat shock proteins.
+We found that at least 9% of all genes were thermally responsive.
+The majority of upregulation was specific to exposure to low temperatures.
+The cool-adapted *A.
+picea* induced expression of more genes in response to extreme temperatures, including genes coding for heat shock proteins, than did *A.
+carolinensis*, consistent with the *enhanced response* hypothesis.
 In contrast, under high temperatures the warm-adapted *A.
@@
 We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.
-These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.
+These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.
 
@@
 To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity and plasticity of genes involved in thermal tolerance [@hoffmann2008; @somero2011].
-In recent years, whole organism gene expression approaches (e.g.
+In recent years, whole-organism gene expression approaches (e.g.
 transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits.
@@
 However, temperature and other environmental factors vary continuously in nature.
-As a result, such comparisons (e.g.
+As a result, such categorical comparisons (e.g.
 high vs.
@@
 Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992].
-Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].
+Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often larger than differences in mean values at both the species and the population level [@murren2014].
 
-Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
-While temporal patterns of transcriptional activity (e.g.
-fast vs.
-slow responding genes) are also important components of an organism's transcriptional response to environmental conditions [@sorensen2005], we focus of the response of transcripts across conditions at the same time point.
+In this study, we extend the reaction norm approach to RNAseq analysis and introduce the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient.
+Although temporal patterns of transcriptional activity (e.g.
+fast- vs.
+slow- responding genes) are also important components of an organism's transcriptional response to environmental conditions [@sorensen2005], we focus here on the response of transcripts across conditions at the same time point.
+
 We use the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
@@
 *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-*Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A.
-picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A.
+*Aphaenogaster carolinensis* experiences a higher mean annual temperature (MAT) (14.6°C) and less seasonal temperature variation (temperature seasonality = 7,678°) than does *A.
+picea* (MAT = 4.6°C, seasonality = 10,008°; [@hijmans2005]) at their respective collection sites.
+In controlled laboratory experiments, these warm- and cold-climate species exhibit corresponding differences in their critical maximum (44.7°C for *A.
 carolinensis* versus 41.3°C for *A.
@@
 carolinensis* versus -0.1°C for *A.
-picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.
+picea*).
+These differences between species in their thermal environments and physiological tolerances allowed us to investigate adaptation to both lower and upper thermal extremes in this system.
 
@@
 This hypothesis would predict that the cool-adapted *A.
-picea* would induce more genes in response to low temperatures than would the warm-adapted *A.
-carolinensis*, which would induce more genes in response to high temperatures.
-Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A.
-picea*, high temperatures for *A.
-carolinensis*.
-Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014].
+picea* would activate more genes, and induce them more strongly, in response to low temperatures than would the warm-adapted *A.
+carolinensis*, which would show greater induction in response to high temperatures.
+
+Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adaptations [@gordon2013; @sunday2014].
 This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011].
-Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.
+
+Finally, the *genetic assimilation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.
 This hypothesis predicts that transcripts responsive to high temperatures in *A.
@@
 Of course, these hypotheses are not mutually exclusive.
-By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adaptation.
+By taking a reactionome approach, we can quantify if, and under what conditions, these mechanisms contribute to thermal adaptation.
 
@@
 #options(scipen = 1, digits = 3)
-load("ms_slim_2014-10-22.RData")
+load("ms_slim.RData")
 library(data.table)
@@
 
-Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
+Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in critical maximum and minimum temperatures [@warren2013].
 Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers.
@@
 
-We collected a single colony of *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of *Aphaenogaster carolinensis*, part of the *A.
-rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W) in fall 2012.
-These locations are central to each species distributions.
-Species identity was confirmed by morphology (Bernice DeMarco, Michigan State University).
-Colonies of these species were maintained in common conditions at 25°C for 6 months prior to experimentation.
-We were unable to determine the critical thermal limits of these exact colonies, but in summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W).
-We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves [@warren2013].
+In fall 2012, we collected a single colony of *Aphaenogaster picea* from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of *Aphaenogaster carolinensis*, part of the *A.
+rudis* species complex [@umphrey1996], from Durham, North Carolina (36.037° N -78.874° W).
+These sites are centrally located within each species' geographic range.
+Species identity was confirmed with morphological characters (Bernice DeMarco, Michigan State University).
+Colonies of both species were maintained in common conditions at 25°C for 6 months prior to experimentation.
+Due to colony size limitations, we were unable to determine the critical thermal limits of these particular colonies.
+In summer 2013 we collected additional colonies of *Aphaenogaster* from Molly Bog, VT and North Carolina (Duke Forest, 36.036° N, 79.077° W).
+We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves, following the protocol of Warren & Chick [-@warren2013].
 
@@
 Thus, as is commonly done with other long-lived organisms [@whitehead2006; @barshis2013], we exposed both colonies to common-garden rearing conditions for six months to fully acclimate adult workers to common temperatures.
-Over this time, roughly 1-2 cohorts of new workers are expected to join the workforce (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.
+Over this time, roughly 1-2 cohorts of new workers are expected to join each colony (~1/3 of the total), such that the workers sampled for thermal traits and gene expression are likely to have included a mix of adult-acclimated and fully lab-reared individuals.
 
@@
 To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time on 12 days.
-We measured response to temperature with a one-hour static temperature application, which is ecologically-relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012].
-Each day, the ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
+We measured response to temperature with a one-hour static temperature application, which is ecologically relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012].
+Each day, the ants were placed in glass tubes immersed in a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour.
 The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum (~0°C) and maximum (~43°C) temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality.
@@
 We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
-To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archae and human databases of contaminants.
+To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archaen and human databases of contaminants.
 
@@
 We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
-As preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, so we omitted these data from the analysis.
+Because preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
 The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).
@@
 
-with to each transcript.
+to each transcript.
 Temperature and species were both fixed effects, with a quadratic term included for temperature.
@@
 To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
+The reaction norms of individual transcripts can be searched by gene ontology and viewed at https://johnsg.shinyapps.io/ApRxN-shinyapp.
 
@@
 All analyses were performed with R 3.1 [@r_core_team2013] and are fully reproducible (Supporting Information).
-
 For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
-We used the predicted transcript expression levels to partition transcripts into five expression categories which were defined *a priori*, and could therefore be used to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in one species but did respond in the other.
-For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.
-The two categories which unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).
+We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other.
+For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the rearing temperature of 25°C.
+Because expression category was defined by the temperature of maximal expression, both **Low** and **High** categories were biased toward transcripts up-regulated at that temperature extreme, but also likely included some transcripts down-regulated at the opposing extreme.
+The two categories which could unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).
 
@@
 
-We performed comparative analyses between the two species focusing on the transcripts that had a significant species x temperature interaction.
-For each transcript, we predicted expression levels across temperatures for each species using the full linear model, and then grouped transcripts into the four responsive categories, **High**, **Low**, **Intermediate**, and **Bimodal**, as well as a fifth category **Not Responsive**.
 We used $\chi^2$ tests to determine if the total number of responsive transcripts, and the number of transcripts in each expression category differed between species.
-To evaluate if shifts from one expression category to another were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equality.
+To evaluate if shifts from one expression category to another between the two species were randomly distributed, we used the Stuart-Maxwell test of marginal homogeneity from the `coin` package [@hothorn2008] in R [@r_core_team2013] which tests if the row and column marginal proportions are in equity.
 
-To test whether the temperature at which thermally-responsive transcripts are activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species.
+To test whether the temperature at which thermally-responsive transcripts were activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species, using only the transcripts with a significant species x temperature interaction.
 For upregulated transcripts, we grouped the **High** transcripts along with the high temperature end of the **Bimodal** transcripts, and did the same for **Low** transcripts.
@@
 For downregulated transcripts (i.e.
-**Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test as before.
+**Intermediate**), we identified the greatest change in expression for each transcript in response to both increasing (> 20°C) and decreasing (< 20°C) temperatures, and used a *t*-test to compare the mean temperature of down-regulation between species.
 
-To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A.
+To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response variable and the log ratio of relative inducibility as the predictor variable for **High** transcripts in *A.
 picea* and for **Low** transcripts in *A.
 carolinensis*.
-Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*.
+Constitutive expression was defined as predicted expression at 25°C, whereas inducibility of each transcript was defined as *((maximum TPM - minimum TPM) / minimum TPM) x 100*.
 In addition, we used a Mann-Whitney test to compare the baseline constitutive expression between species for all responsive transcripts.
@@
 To describe the molecular processes involved in thermal adaptation, we performed gene set enrichment analysis (GSEA) using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013].
-Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.
+Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, after accounting for dependencies among the GO terms.
 
@@
 About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
-Of these 9,052 transcripts, ~80% (7,201) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
+We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 9% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
+Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
 high vs.
 low temperature).
-The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246).
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.
+The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`).
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.
 
-Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
-The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A.
+Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
+The majority of the responsive transcripts were in the **Low** category for both species (`r Ap.type.table.prop[2,"Low"]`% and `r Ap.type.table.prop[1,"Low"]`% for *A.
 carolinensis* and *A.
-picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A.
-carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A.
+picea*, respectively), followed by **Intermediate** (`r Ap.type.table.prop[2,"Intermediate"]`%), **High** (`r Ap.type.table.prop[2,"High"]`%),**Bimodal** (`r Ap.type.table.prop[2,"Bimodal"]`%) and **Not Responsive** (`r Ap.type.table.prop[2,"NotResp"]`%) for *A.
+carolinensis*, and **Bimodal** (`r Ap.type.table.prop[1,"Bimodal"]`%), **High** (`r Ap.type.table.prop[1, "High"]`%), **Intermediate** (`r Ap.type.table.prop[1,"Intermediate"]`%) and **Not Responsive** (`r Ap.type.table.prop[1,"NotResp"]`%) for *A.
 picea*.
@@
 
-The two species differed in the number of transcripts in each response category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
+The two species differed in the number of transcripts in each expression category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
 *Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A.
 carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`).
-Consistent with the *enhanced response* hypothesis, *A.
-picea* induced 1,146 (30%) more transcripts in response to low temperatures than *A.
+Consistent with the *enhanced response* hypothesis, the cool-climate *A.
+picea* induced 1,146 (30%) more transcripts in response to low temperatures than the warm-climate *A.
 carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`).
@@
 
-In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species.
-Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another.
-These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Fig.
-S1) indicating that these differences are not simply due to neutral processes.
-The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species.
-The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.
-For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A.
-carolinensis* for low temperatures, *A.
-picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species.
+In addition, we also examined the specific patterns of shifts from one expression category to another between species.
+Expression patterns differed for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts, most of which shifted from one expression category to another.
+These shifts in expression category between species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, P < 0.001, Fig.
+S1), indicating that these differences were unlikely to be due to neutral processes.
+Specifically,
+
+The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species (Fig.
+1).
+The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig.
+1).
+In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig.
+1).
+For each temperature extreme, we tested these predictions by focusing on the transcripts upregulated by the species that expriences less extreme temperatures (the warm-climate *A.
+carolinensis* for cool temperatures, the cool-climate *A.
+picea* for warm temperatures) and testing whether those transcripts displayed the same response profile in species experiencing more extreme temperatures.
 As with the overall results, the patterns of association differed for the two temperature extremes.
-Genes upregulated at low temperatures in *A.
-carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A.
-picea*, suggesting shared response pathways (Fig.
-2A).
-In contrast, genes upregulated in response to high temperatures in *A.
-picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A.
+Transcripts upregulated at low temperatures in *A.
+carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from other expression categories in *A.
+picea* (Fig.
+2A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis.
+In contrast, transcripts upregulated in response to high temperatures in *A.
+picea* (**High** and **Bimodal**) shifted expression categories in *A.
 carolinensis* (Fig.
+2B), primarily to the **Not Responsive** and **Intermediate** expression categories (Fig.
 2B).
-Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig.
-2B).
+These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.
 
@@
 carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` significantly enriched GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
-In addition, there were GO terms related to repair (e.g.
+In addition, there were GO terms related to organelle repair (e.g.
 histone modification), transcript regulation (e.g.
@@
 transport, organelle organization).
-Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A.
+Collectively, these results suggest that in response to both thermal extremes, the warmer-climate species, *A.
 carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A.
@@
 3A).
-For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
+For low temperatures, *A.
+carolinensis* also had the more extreme induction temperature (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
 picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
-3A).
+3A), though the magnitude of difference was smaller than at high temperatures.
 
-Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
-Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig.
+Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
+Thresholds for downregulation were shifted more toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig.
 3).
-There was greater inertia of gene expression in response to high and low temperature extremes.
+As for the upregulated transcripts, *A.
+carolinensis* had greater inertia of gene expression in response to high and low temperature extremes.
 With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
-picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A.
-picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A.
-carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
+picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`).
+The difference between temperatures was not significant with decreasing temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)` for *A.
+picea* compared to `r round(T_off.int.low.ttest$estimate[1], 1)` for *A.
+carolinensis*; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
 3B).
@@
 
-We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the midpoint temperature (19.5°C).
-If stress-response genes have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two.
-We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A.
+We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C).
+If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two.
+We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A.
+picea* were not expressed at higher baseline levels in the warm-climate *A.
+carolinensis* (Fig.
+4A).
+Similarly, transcripts more upregulated at low temperatures in *A.
+carolinensis* did not show higher baseline levels in *A.
 picea* (Fig.
-4A) or transcripts upregulated at low temperatures in *A.
-carolinensis* (Fig.
 4B).
-In fact, for both comparisons we found a weakly positive relationship.
+In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species.
 
 In addition, the thermally responsive transcripts in *A.
-carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A.
+carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A.
 picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`).
@@
 2B).
-Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A.
+Rather than having constitutively elevated expression as predicted, these transcripts were uniformly low or not expressed (mean TPM = 0.08) in *A.
 carolinensis*.
@@
 However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance.
-By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify evolutionary changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A.
+By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify putative changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A.
 carolinensis*.
@@
 picea* is substantially lower than *A.
-carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig.
+carolinensis*, the two species upregulated responsive transcripts near the same threshold temperature (Fig.
 3A).
@@
 In contrast to cold tolerance, the enhanced upper thermal limit in *A.
-picea* is best explained by the *tolerance* hypothesis.
+carolinensis* is best explained by the *tolerance* hypothesis.
 High temperatures were associated with significantly fewer upregulated transcripts in *A.
@@
 3B-D).
-This difference between species is also evident among candidate genes known to be involved in stress response.
+This difference between species is also evident among candidate transcripts known to be involved in stress response.
 Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
@@
 carolinensis*.
-Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress.
+Collectively, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.
+Such mechanisms may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress.
 These differences are in line with expectations that *A.
 carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions.
-In fact, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
+Indeed, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
 
 The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts.
-This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al.
-[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations.
+This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013].
+However, in a short-term selection experiment for heat tolerance, Sikkink et al.
+[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance.
 Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
@@
 2), it is worth noting a number of similarities.
-In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1).
+The two species showed similar numbers of transcripts with a **Bimodal** generalized stress-response (Table 1).
+In both species, there were 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1).
 The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures.
@@
 At these extreme low temperatures we also found few upregulated transcripts (Fig.
-3A), while the peak of low-temperature transcriptional activity occurred near 10°C (Fig.
+3A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig.
 3).
-A potential explanation for this pattern is that increased gene expression is required to support elevated metabolism at cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002].
-The observation that more transcripts were upregulated at low than high temperatures could be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013].
+A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002].
+The observation that more transcripts were upregulated at low than high temperatures could also be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013].
 This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
-The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role.
+
 Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
-Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig.
+Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were much more similar (Fig.
 3).
@@
 
-A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach).
+A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach), which revealed quantitative patterns of temperature response not captured in categorical comparisons.
 For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as we discussed above.
 Further, a number of issues have hampered RNA-seq studies to date.
-Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012].
+Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" interpretations [@pavlidis2012].
 Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species.
-While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
+Although we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
 
 Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
-While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig.
+Although we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig.
 2, Fig.
 S1).
-Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.
+Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression -- as opposed to identifying lists of differentially-expressed transcripts -- can provide meaningful insight on the process of evolution.
 
 A number of caveats do apply to our work.
-First, species may differ in gene expression along axes which we have not measure here, primarily temporal patterns of gene expression [@sorensen2005], and could be studied in futher work.
+First, species may differ in gene expression along axes which we have not measured here, especially temporal patterns of gene expression [@sorensen2005], which could be studied in further work.
 Second, the *de novo* transcriptome assembly is highly fragmented, given that all sequenced ant genomes to date have only about 18,000 genes [@gadau2012].
@@
 A genome assembly, in progress, will help to reduce fragmentation.
-Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for the model organisms *Drosophila* and *Apis*.
+Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for model arthropods such as *Drosophila* and *Apis*.
 Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.
@@
 
-Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance.
+Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, whereas organisms from less stressful environments rely on plastically-induced resistance.
 Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].
@@
 carolinensis*.
-At least 8% of their transcriptomes are thermally responsive.
-For both species, more transcripts responded to low than high temperatures.
-Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have more limited genetic accomodation to high than low temperatures.
-Our results also indicate that these two ant species have different responses to thermal extremes.
+At least 9% of their transcriptomes are thermally responsive.
+Our results indicate that these two ant species have different responses to thermal extremes.
 *A.
@@
 Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626.
-The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40.
+The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).
 The reproducible and version-controlled script for data analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
@@
 Each column indicates the distribution across all response categories in *A.
-picea*, which has a lower CT_{min}_ and is therefore better adapated to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CT_{min}, *A.
+picea*, which has a lower CTmin and is therefore better adapated to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CTmin, *A.
 carolinensis*, relative to the null hypothesis of equal marginal frequencies.
-The dashed boxes highlight cells that would indicated matched responses in the two species, and the color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts.
+The dashed boxes highlight cells that would indicate matched responses in the two species, and the color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts.
 The (A) *enhanced response* hypothesis proposes that the increase in cold tolerance in *A.
 picea* is achieved by amplifying existing molecular mechanisms, and thus there should be an excess of shared resonse types between species.
-In contrast, the (B) *tolerance* predicts that *A.
+In contrast, the (B) *tolerance hypothesis* predicts that *A.
 picea* is less reliant on induced responses to confer cold-tolerance than *A.
@@
 **Figure 2**.
-Results of the marginal frequencies analysis of thermal response categories.
+Results of marginal frequencies analysis of thermal response categories.
 The color of each cell  (purple = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
@@
 **Figure 3**.
-Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures.
+Histogram with smooth density estimate of temperature of maximum rate of change in expression for transcripts that have (A) increased expression at **Low** and **High** temperatures and (B) decreased expression at **Low** and **High** temperatures.
 Red bars and lines are for *A.
@@
 **Figure 4**.
-Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A.
+Scatterplots of log ratios of relative inducibility to relative constitutive expression, defined as expression level at the common rearing temperature (25°C) for (A) **High** transcripts in *A.
 picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A.

### 2015-07-02 ApTranscriptome_TR.Rmd

--- 2015-06-11 ApTranscriptome_TR.Rmd
@@
 options(stringsAsFactors=FALSE, digits = 3, scipen = 1)
+options(width = 80)
 
@@
 
-In this technical report, which accompanies the manuscript **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species** (Stanton-Geddes et al., in press), we:
+In this technical report, which accompanies the manuscript **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**, we:
 
 1. Describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`.
-2. Identify thermally-responsive genes.
-3. Evaluate differences in transcriptome-wide patterns of thermal responsiveness (i.e. the thermal reactionome) between the two species
-4. Perform gene set enrichment analysis of thermally-responsive genes for the two species.
+2. Measure the reaction norm across a thermal gradient (0°C to 38.5°C) for all transcripts and characterize the *thermal reactionome* as the set of transcripts that are thermally-responsive.
+3. Quantitatively test three mechanistic hypotheses of thermal adapation:.
+  - The *Enhanced response* hypothesis proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
+  - The *Tolerance hypothesis* proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency.
+  - The *Genetic assimilation hypothesis* proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.
+
+We quantify the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species.
+We use gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.
 
@@
 
-```{r ed_cartoon}
-enhanced_defense_dat <- md_sub1
-enhanced_defense_dat$value <- 0
+```{r er_cartoon}
+enhanced_response_dat <- md_sub1
+enhanced_response_dat$value <- 0
 # set non-shared response to random negative value 
@@
 
-enhanced_defense_dat
+enhanced_response_dat
 
 # plot
-enhanced_defense_plot <- qplot(y=Apic, x=Acar, data=enhanced_defense_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_defense_dat$Acar))) + 
+enhanced_response_plot <- qplot(y=Apic, x=Acar, data=enhanced_response_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_response_dat$Acar))) + 
   scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +

### 2015-09-30 ApTranscriptome_TR.Rmd
added resampling analysis: greatly reduces number of significant transcripts

@@
 ---
-output: pdf_document
+title: Technical Report for **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**
+author: [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
+output:
+  word_document: default
+  pdf_document:
+    fig_caption: yes
+    includes:
+      in_header: header.tex
+    keep_tex: yes
+    latex_engine: xelatex
+    number_sections: no
+    toc_depth: 2
+  html_document:
+    keep_md: yes
 ---
-
-Technical Report for **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**
-========================================================================================
-
-**Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
-
-**`r format(Sys.time(), '%d %B, %Y')`**
-
-**Technical Report No. 3**
-
-**Department of Biology**
-
-**University of Vermont**
 
@@
 #biocLite("topGO")
-#biocLite("Rgraphviz")
-#biocLite("GOSemSim")
 library(topGO)
-library(Rgraphviz)	
-library(GOSemSim)
 
@@
 
-We quantify the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species.
-We use gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.
+We quantified the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species.
+We used gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.
 
@@
 ```
-
-Note that I used the reduced assembly, prior to cleaning out contaminants.
-As I use the "clean" assembly for read mapping and identification of responsive genes, the contaminants are dis-regarded in downstream analyses.
 
@@
 
-Note that expression levels at each temperature treatment are highly correlated between the two colonies.
+Note that expression levels at each temperature treatment are highly correlated between the two colonies, indicating common patterns of expression response to temperature.
 
@@
 
-With gene expression quantified using all reads, I now remove the transcripts identified as contaminants by DeconSeq.
+With gene expression quantified using all reads, I removed the transcripts identified as contaminants by DeconSeq.
 
@@
 cont.list <- read.table("results/trinity-full/cont.list")
+# remote ">" leader
+cont.list$V1 <- gsub(">", "", cont.list$V1)
 str(cont.list)
@@
 
-Many of these are likely false positives, so I adjust P-values using false discovery rate (FDR). Only those transcripts with less than 5% FDR are retained as significant.
+Many of these are likely false positives, so I adjusted P-values using false discovery rate (FDR).
 
@@
 # subset to significant transcripts
-signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]
+pre.signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]
+```
+
+At the 5% FDR significance threshold, there are `r nrow(pre.signif.transcripts)` transcripts with an overall significant model.
+
+Feedback from reviewers questioned whether this was sufficient to remove all false positives.
+To examine this, we performed a resampling simulation to determine how many transcripts would be identified as significant using the above approach.
+Specifically, we randomly re-assigned the expression levels to the temperature treatments for each species, fit the same linear model as for the real data, calculated the number of significant transcripts under the 5% FDR threshold, and repeated this 100 times.
+We then took the 95th percentile of the distribution of significant transcripts from the resampled datasets as the true null level of false positives.
+
+As this resampling is slow, I ran it in a separate file (ApTranscriptome_resample.Rmd) and load the results here.
+
+```{r resample}
+# load resampling results
+load("data/resample.Rda")
+```
+
+Using this approach, the empirical false positive, *after* adjusting *P*-values for FDR, is `r quantile (D$num_signif, probs = 0.95)`, indicating that only `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` are true positives.
+
+```{r}
+# order by padj
+pre.signif.transcripts <- pre.signif.transcripts[order(pre.signif.transcripts$padj), ]
+# number to keep
+keep <- nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)
+signif.transcripts <- pre.signif.transcripts[1:keep, ]
 
@@
 
-At the 5% FDR significance threshold, there are `r nrow(signif.transcripts)` transcripts with an overall significant model.
-
-(2) Fit linear model to overall significant transcripts; perform stepAIC to retain only significant terms, and save `lm` output to list
+(2) Fit linear model to significantly responsive transcripts, perform stepAIC to retain only significant terms, and save `lm` output to list
 
@@
 
-I checked the proportion of responsive transcripts against those that have a BLAST hit.
-
-```{r responsive_with_BLAST}
-annotation.table.responsive <- annotation.table[which(annotation.table$best.hit.to.nr != "-"), ]
-
-responsive_transcripts_w_BLAST <- length(which(unique(resp.TPM.dt.sub$Transcript) %in% annotation.table.responsive$Sequence.Name))
-
-responsive_transcripts_w_BLAST / round(nrow(annotation.table.responsive))
-```
-
 ### Thermal-response expression categories ###
@@
 ```{r expression_shape, eval=TRUE, echo=TRUE}
-# calculate response type for responsive transcripts
+# calculate response type for all temperature responsive transcripts
+sig.response.type <- ldply(responsive.lms, .progress = "none", .inform = TRUE, RxNtype)
+colnames(sig.response.type)[which(colnames(sig.response.type) == ".id")] <- "Transcript"
+sig.response.type <- data.table(sig.response.type)
+setkey(sig.response.type, Transcript)
+str(sig.response.type)
+
+# calculate response type for transcripts with species x temperature interaction only
 interaction.response.type <- ldply(interaction.lms, .progress = "none", .inform = TRUE, RxNtype)
-stopifnot(nrow(interaction.response.type) == length(interaction.lms))
-
-# calculate response types for transcripts without interactions
-temperature.response.type <- ldply(temperature.lms, .progress = "none", .inform = TRUE, RxNtype)
-stopifnot(nrow(temperature.response.type) == length(temperature.lms))
-
-# for subsequent analyses, focus on the transcripts with significant temperature x species interaction only
 colnames(interaction.response.type)[which(colnames(interaction.response.type) == ".id")] <- "Transcript"
@@
 
+# calculate response types for transcripts without interactions
+temperature.response.type <- ldply(temperature.lms, .progress = "none", .inform = TRUE, RxNtype)
+
 # save results to file
-write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), interaction.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
+write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), sig.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
 ```
@@
 ```{r temperature_response}
-A22.type.table <- table(interaction.response.type$A22.type)
-Ar.type.table <- table(interaction.response.type$Ar.type)
+A22.type.table <- table(sig.response.type$A22.type)
+Ar.type.table <- table(sig.response.type$Ar.type)
 
@@
 
-```{r mosaic_plot}
-# Reorganize for plotting to show overlap among categories between colonies
-type.table <- table(Acar = interaction.response.type$Ar.type, Apic = interaction.response.type$A22.type)
+## Test of marginal frequencies between transcripts in each expression category between species
+
+The question of biological interest is whether the marginal frequencies differ between the two species.
+Statistically, this can be addressed using the generalized [McNemar's test](http://en.wikipedia.org/wiki/McNemar's_test) of marginal homogeneity.
+
+```{r mh_test}
+type.table <- table(Acar = sig.response.type$Ar.type, Apic = sig.response.type$A22.type)
 # reorder
@@
 
-# plot
-png("mosaic_plot.png")
-mosaic.colors <- brewer.pal(5, "Blues")
-par(mar = c(5.1, 3.1, 2.1, 7.5))
-mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
-labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
-text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
-text(.5, par("usr")[4], "A. carolinensis", xpd = TRUE, font = 3)
-legend(1.05, .75, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
-dev.off()
-```
-
-The question of biological interest is whether the marginal frequencies differ between the two species.
-Statistically, this can be addressed using the generalized [McNemar's test](http://en.wikipedia.org/wiki/McNemar's_test) of marginal homogeneity.
-
-```{r mh_test}
+# McNemar's test
 mh.test <- mh_test(as.table(tt2))
@@
 
-# e2 (Acar Intermediate) contributes ~44% to Z~0~
+# e2 (Acar Intermediate) contributes ~48% to Z~0~
 ```
@@
 mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) +
-  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
+  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 mh_plot_sub1 <- qplot(y=Apic, x=Acar, data=md_sub1, fill=value, geom="tile", xlim = rev(levels(md_sub1$Acar))) + 
-  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
+  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 mh_plot_sub2 <- qplot(y=Acar, x=Apic, data=md_sub2, fill=value, geom="tile", xlim = rev(levels(md_sub2$Apic))) + 
-  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
+  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 # set non-shared response to random negative value 
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=150)
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- rep(-200, 5)
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Low"), "value"] <- runif(5, min=-20, max=-10)
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Bimodal"), "value"] <- runif(5, min=-20, max=-10)
 # change shared response to positive
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 50
-enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 50
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Low" & enhanced_response_dat$Apic == "Low"), "value"] <- 30
+enhanced_response_dat[which(enhanced_response_dat$Acar == "Bimodal" & enhanced_response_dat$Apic == "Bimodal"), "value"] <- 30
 
@@
 # plot
-enhanced_response_plot <- qplot(y=Apic, x=Acar, data=enhanced_response_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_response_dat$Acar))) + 
-  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
+enhanced_response_plot <- ggplot(enhanced_response_dat, aes(y = Apic, x = Acar, fill=value)) + 
+  geom_tile(colour = "#ffffff", size =1) +
+  xlim(rev(levels(enhanced_response_dat$Acar))) + 
+  scale_fill_gradient2(limits=c(-100, 100), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 
-enhanced_response_plot <- ggplot(enhanced_response_dat, aes(y = Apic, x = Acar, fill=value)) + 
-  geom_tile(colour = "#ffffff", size =1) +
-  xlim(rev(levels(enhanced_response_dat$Acar))) + 
-  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
-  theme(axis.title = element_text(face="italic")) +
-  labs(y = "A. picea", x = "A. carolinensis") +
-  theme(legend.position = "right") + 
-  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
-  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")
-
 # title
@@
 # set non-shared response to random positive value 
-tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=160)
-tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min=140, max=150)
+tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min=40, max=60)
+tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min=40, max=60)
 # change shared response to near zero
-tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- 20
-tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- 20
+tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- -20
+tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- -20
 
@@
   xlim(rev(levels(enhanced_response_dat$Acar))) + 
-  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
+  scale_fill_gradient2(limits=c(-100, 100), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 
-The number of thermally-responsive in each response category differed between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies.
-For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts.
-For *AcNC*, transcripts expressed at *Intermediate* temperatures are next most common, followed by *High* and *Bimodal*.
-
-Interestingly, nearly half of the *High* genes in *AcNC* are *Low* in *ApVT*, and vice versa. In contrast, most of the *Low* genes in one species are also *Low* in the other species.
+The number of thermally-responsive in each response category differed between the colonies.
+For *A.
+picea*, the majority of the responsive transcripts were in the *Low* category, followed by *Bimodal*, *High*, *Intermediate* and *Not Responsive*.
+For *A.
+carolinensis*, the majority of responsive transcripts were in the *Bimodal* and *High* categories, followed by *Intermediate*, *Low* and *Not Responsive*.
+
+Interestingly, over half of the *High* genes in *AcNC* are *Low* in *ApVT*, and vice versa. In contrast, most of the *Low* genes in one species are also *Low* in the other species.
 
@@
 Table 4 shows the number of transcripts that fall into each expression type for each each species.
-The totals for each species include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
+The totals for each species *do not* include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.
 
@@
 1) is that for **NotResp** genes in *A.
-carolinensis* the **High** and **Bimodal** categories in *A.
-picea* are over-represented.
+carolinensis* the **Bimodal** category in *A.
+picea* is over-represented.
 Finding that the mean expression level of these genes in *A.
@@
 picea* would be consistent with genetic assimilation.
-
-```{r gen_assimilation_high}
-Ap.high.Ac.nr <- interaction.response.type[which(interaction.response.type$A22.type == "High" & interaction.response.type$Ar.type == "NotResp"), ]
-
-# get mean expression level for Ar transcripts
-Ap.high.Ac.nr.TPM <- TPM.dt.sub[Ap.high.Ac.nr$Transcript]
-Ap.high.Ac.nr.mean.exp.NR <- ddply(Ap.high.Ac.nr.TPM, .(Transcript), summarize, meanTPM = mean(TPM))
-
-# get expression at optimum for A22 transcripts
-Ap.high.Ac.nr.opt.exp <- Ap.high.Ac.nr$A22.opt
-
-t.test(Ap.high.Ac.nr.mean.exp.NR$meanTPM, Ap.high.Ac.nr.opt.exp)
-```
-
-Contrary to expectations, mean expression of the non-responsive *A. carolinensis* genes is less than expression of the **High** genes in *A. picea*.
-
-Next, I checked the same for the **Bimodal** genes in *A. picea*.
+That is, they have evolved to be constitutively activated in *A.
+carolinensis* and thus are no longer thermally-responsive as they are in *A.
+picea*.
 
 ```{r gen_assimilation_bim}
-Ap.bim.Ac.nr <- interaction.response.type[which(interaction.response.type$A22.type == "Bimodal" & interaction.response.type$Ar.type == "NotResp"), ]
+Ap.bim.Ac.nr <- sig.response.type[which(sig.response.type$A22.type == "Bimodal" & sig.response.type$Ar.type == "NotResp"), ]
 
@@
 
-As with the **High** transcripts, expression was much higher for the responsive *A. picea* transcripts than for the non-responsive *A. carolinensis* transcripts.
+This hypothesis was not supported as expression was higher for the **Bimodal** transcripts in *A. picea* than the non-responsive *A. carolinensis* transcripts.
 
@@
 
-In this section, I perform a number of comparisons of the thermal reactionomes between the *Ar* and *A22* species. Specifically, I:
-
-- compare profiles of the temperature of maximum expression for thermally-responsive transcripts
-- compare basal expression at optimal temperature for thermally-responsive genes
-- compare degree of upregulation to constitutive expression
-- compare the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures 
-- compare the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures
-
-Predict expression for responsive transcripts.
+In this section, I performed a number of comparisons of the thermal reactionomes between the *Ar* and *A22* species. Specifically, I:
+
+- compared profiles of the temperature of maximum expression for thermally-responsive transcripts
+- compared basal expression at optimal temperature for thermally-responsive genes
+- compared degree of upregulation to constitutive expression
+- compared the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures 
+- compared the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures
+
+For these tests, I predicted expression for responsive transcripts.
 
@@
 
+I checked the proportion of responsive transcripts against those that have a BLAST hit.
+
+```{r responsive_with_BLAST}
+annotation.table.responsive <- annotation.table[which(annotation.table$best.hit.to.nr != "-"), ]
+
+responsive_transcripts_w_BLAST <- length(which(unique(resp.TPM.dt.sub$Transcript) %in% annotation.table.responsive$Sequence.Name))
+
+responsive_transcripts_w_BLAST / round(nrow(annotation.table.responsive))
+```
+
 ### Compare temperature of maximum expression between species
 
-Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*
+Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*.
+For this analysis, I used only the transcripts with a significant temperature by species interaction.
 
@@
 # reshape data
-Ap.df <- data.frame(Transcript = rep(interaction.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(interaction.response.type$Transcript)), max.val = c(interaction.response.type$A22.max, interaction.response.type$Ar.max))
+Ap.df <- data.frame(Transcript = rep(interaction.response.type$Transcript, times = 2), 
+                    colony = rep(c("ApVT", "AcNC"), 
+                                 each = length(interaction.response.type$Transcript)), 
+                    max.val = c(interaction.response.type$A22.max, interaction.response.type$Ar.max))
 
@@
 This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`.
-As my goal is simply to determine if expression is typically greater at optimal temperatures (19.5 C) in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test
+As my goal is simply to determine if expression is typically greater at optimal temperatures (19.5 C) in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test.
 
@@
 Consistent with expectations, there is greater expression at 25C for *Ar* than *A22* transcripts for the set of transcripts that are transcripts that are up-regulated at high temperatures in *A22*.
-Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.
+Note that *A22* had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.
 
@@
 
-I extract the **High** transcripts for *A.
-picea* and examine the log ratio of inducibility against the log ratio of constitutive expression in *A.
+I extracted the **High** transcripts for *A.
+picea* and examined the log ratio of inducibility against the log ratio of constitutive expression in *A.
 picea* over *A.
@@
 
-As with the previous comparison, I also find no evidence for the genetic assimilation hypothesis.
+As with the previous comparison, I also found no evidence for the genetic assimilation hypothesis.
 
@@
 Thermally-responsive genes could also differ in the critical temperature of gene induction.
-To examine this, I determine the temperature at which each responsive gene has the greatest increase or decrease in expression.
+To examine this, I determined the temperature at which each responsive gene had the greatest increase or decrease in expression.
 
@@
 
-For next analyses, extract list of gene names by response type.
+For next analyses, I extracted the list of gene names by response type.
 
@@
 
-Calculate T~on~ for *High* genes in each species.
-I use the observed (rather than predicted) values for each species as by nature of fitting a quadratic function, the maximum change tends to be at the extremes (38.5 or 0) for the predicted values.
+I calculated T~on~ for *High* genes in each species using the observed (rather than predicted) values for each species as by nature of fitting a quadratic function, the maximum change tends to be at the extremes (38.5 or 0) for the predicted values.
 
@@
 
-Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.
-
-Repeat analysis for *Low* genes.
+This test revealed no differences among species in T~on~ for *High* genes.
+
+I performed the same test comparing T~on~ for *Low* genes.
 
@@
 picea* **Intermediate** genes.
-Towards low temperatures, the mean temperature of downregulation was marginally higher for *A.
-carolinensis* than *A.
-picea*.
+Towards low temperatures, the mean temperature of downregulation was not different between species.
 
@@
 
-Make a plot with all 4 gene activation comparisons.
+I combined these gene activation comparisons into a single plot.
 
@@
 
-Does the mean expression level for all responsive genes differ among colonies?
+Next, I tested if the mean expression level for all responsive genes differed among colonies.
 
@@
 ```
+
+On average, *A. carolinensis* has higher expression than *A. picea*.
 
@@
 # combine responsive.lms.ann with "type" information
-responsive.lms.ann.type <- merge(responsive.lms.ann.type, interaction.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
+cols <- c("Transcript", "A22.type", "Ar.type")
+subtab <- sig.response.type[, cols, with = FALSE]
+setkey(subtab, "Transcript")
+responsive.lms.ann.type <- base::merge(responsive.lms.ann.type, subtab, by = "Transcript", all = TRUE)
 str(responsive.lms.ann.type)
@@
 
-There are also `r nrow(hsp_responsive)` heat shock related genes in the responsive transcripts, out of `r nrow(hsp_all)` total heat shock related genes in the transcriptome annotation.
+There are no heat shock protein-related genes in the responsive transcripts.
 
@@
 
-I use [topGO](http://www.bioconductor.org/packages/2.12/bioc/html/topGO.html) to perform gene set enrichment analysis (GSEA) seperately for each expression type (bimodal, intermediate, high, low).
-
-First need to create gene ID to GO term map file
+I used [topGO](http://www.bioconductor.org/packages/2.12/bioc/html/topGO.html) to perform gene set enrichment analysis (GSEA) seperately for each expression type (bimodal, intermediate, high, low).
+With the GSEA results, I applied the `selectFDR` function to select transcripts with adjusted P < 0.05.
+For many of the responsive categories, it appeared that there is considerable redundancy in the enriched GO terms.
+I used the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then performed hierarhichal clustering based on this information distance.
+
+For the GSEA, I first created a gene ID to GO term map file.
 
@@
 
-then read map file.
+Then I read the map file into memory.
 
@@
 
-Using this gene2GO map file, perform GSEA for:
+Using this gene2GO map file, I performed GSEA for:
 
 **1) all responsive transcripts**
-
-Use `selectFDR` function to select transcripts with adjusted P < 0.05.
-
-Perform GSEA for all thermally-responsive transcripts.
-
-For many of the responsive categories, it appears that there is considerable redundancy in the enriched GO terms.
-I use the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then perform hierarhichal clustering based on this information distance.
 
@@
 
-As the molecular processes involved in cold and hot temperature tolerance are different, I perform GSEA separately for each response type and compile results in a single table.
+As the molecular processes involved in cold and hot temperature tolerance are different, I performed GSEA separately for each response type and compiled results in a single table.
 
 ```{r gsea_Ap_high, results = 'hide'}
+sig.response.type <- as.data.frame(sig.response.type)
 ## Genes with *High* expression in both colonies
-Ap.high <- interaction.response.type[which(interaction.response.type$Ar.type == "High" & interaction.response.type$A22.type == "High"), "Transcript"]
+Ap.high <- sig.response.type[which(sig.response.type$Ar.type == "High" & sig.response.type$A22.type == "High"), "Transcript"]
 Ap.geneList.high <- rep(0, length = length(Ap.geneList))
@@
 
-```{r gsea_Ap_high_cluster}
-# similarity among enriched GO terms
-Ap.high.GO.term.sim <- termSim(Ap.high.gsea$GO.ID, Ap.high.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term
-colnames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term
-
-# distance matrix
-Ap.high.hclust <- hclust(dist(Ap.high.GO.term.sim))
-plot(Ap.high.hclust)
-
-# report items in each cluster
-Ap.high.hclust.report <- GSEAReportClusters(Ap.high.hclust, h = 1)
-
-# add cluster information
-Ap.high.gsea$Cluster <- NA
-
-for(i in 1:length(Ap.high.hclust.report)) {
-    Ap.high.gsea[Ap.high.hclust.report[[i]], "Cluster"] <- i
-}
-```
+A single GO term, *`r paste(Ap.high.gsea[, 1:2], collapse = " ")`*, is enriched for **High** genes in both species.
 
@@
 ## Genes with *Low* expression in both colonies
-Ap.low <- interaction.response.type[which(interaction.response.type$Ar.type == "Low" & interaction.response.type$A22.type == "Low"), "Transcript"]
+Ap.low <- sig.response.type[which(sig.response.type$Ar.type == "Low" & sig.response.type$A22.type == "Low"), "Transcript"]
 Ap.geneList.low <- rep(0, length = length(Ap.geneList))
@@
 
-```{r gsea_Ap_low_cluster}
-# similarity among enriched GO terms
-Ap.low.GO.term.sim <- termSim(Ap.low.gsea$GO.ID, Ap.low.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ap.low.GO.term.sim) <- Ap.low.gsea$Term
-colnames(Ap.low.GO.term.sim) <- Ap.low.gsea$Term
-
-# distance matrix
-Ap.low.hclust <- hclust(dist(Ap.low.GO.term.sim))
-plot(Ap.low.hclust)
-
-# report items in each cluster
-Ap.low.gsea.cluster <- Ap.low.hclust.report <- GSEAReportClusters(Ap.low.hclust, h = 1.5)
-
-# add cluster information
-Ap.low.gsea$Cluster <- NA
-
-for(i in 1:length(Ap.low.hclust.report)) {
-    Ap.low.gsea[Ap.low.hclust.report[[i]], "Cluster"] <- i
-}
-```
+The GO terms, *`r paste(Ap.low.gsea[, 2], collapse = ", ")`*, are enriched for **Low** genes in both species.
 
@@
 # Genes with *Bimodal* expression in both colonies
-Ap.bim <- interaction.response.type[which(interaction.response.type$A22.type == "Bimodal" & interaction.response.type$Ar.type == "Bimodal"), "Transcript"]
+Ap.bim <- sig.response.type[which(sig.response.type$A22.type == "Bimodal" & sig.response.type$Ar.type == "Bimodal"), "Transcript"]
 # create gene list, setting value to 1 for "bim" transcripts
@@
 
-```{r gsea_Ap_bim_cluster}
-# similarity among enriched GO terms
-Ap.bim.GO.term.sim <- termSim(Ap.bim.gsea$GO.ID, Ap.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ap.bim.GO.term.sim) <- Ap.bim.gsea$Term
-colnames(Ap.bim.GO.term.sim) <- Ap.bim.gsea$Term
-
-# distance matrix
-Ap.bim.hclust <- hclust(dist(Ap.bim.GO.term.sim))
-plot(Ap.bim.hclust)
-
-# report items in each cluster
-Ap.bim.hclust.report <- GSEAReportClusters(Ap.bim.hclust, h = 1.5)
-
-# add cluster information
-Ap.bim.gsea$Cluster <- NA
-
-for(i in 1:length(Ap.bim.hclust.report)) {
-    Ap.bim.gsea[Ap.bim.hclust.report[[i]], "Cluster"] <- i
-}
-```
+The GO terms, *`r paste(Ap.bim.gsea[, 2], collapse = ", ")`*, are enriched for **Bimodal** genes in both species.
 
@@
 # genes with *Intermediate* expression in both colonies
-Ap.int <- interaction.response.type[which(interaction.response.type$A22.type == "Intermediate" & interaction.response.type$Ar.type == "Intermediate"), "Transcript"]
+Ap.int <- sig.response.type[which(sig.response.type$A22.type == "Intermediate" & sig.response.type$Ar.type == "Intermediate"), "Transcript"]
 # create gene list, setting value to 1 for "int" transcripts
@@
 
-```{r gsea_Ap_int_cluster}
-# similarity among enriched GO terms
-Ap.int.GO.term.sim <- termSim(Ap.int.gsea$GO.ID, Ap.int.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ap.int.GO.term.sim) <- Ap.int.gsea$Term
-colnames(Ap.int.GO.term.sim) <- Ap.int.gsea$Term
-
-# distance matrix
-Ap.int.hclust <- hclust(dist(Ap.int.GO.term.sim))
-plot(Ap.int.hclust)
-
-# report items in each cluster
-(Ap.int.hclust.report <- GSEAReportClusters(Ap.int.hclust, h = 1.5))
-
-# add cluster information
-Ap.int.gsea$Cluster <- NA
-
-for(i in 1:length(Ap.int.hclust.report)) {
-    Ap.int.gsea[Ap.int.hclust.report[[i]], "Cluster"] <- i
-}
-```
-
-```{r Ap_gsea_intersect_save, echo = FALSE}
+The GO terms, *`r paste(Ap.int.gsea[, 2], collapse = ", ")`*, are enriched for **Intermediate** genes in both species.
+
+```{r Ap_gsea_intersect_save, echo = FALSE, eval = FALSE}
 Ap.high.gsea$Type <- "High"
@@
 Ap.int.gsea$Type <- "Intermediate"
-Ap.gsea.intersect <- rbind(Ap.low.gsea, Ap.int.gsea, Ap.high.gsea, Ap.bim.gsea)
-
-Ap.gsea.intersect <- Ap.gsea.intersect[,c("Type", "GO.ID", "Term", "Cluster", "Annotated", "Significant", "Expected", "parentchild")]
+Ap.gsea.intersect <- rbind(Ap.high.gsea, Ap.low.gsea, Ap.int.gsea, Ap.bim.gsea)
+
+Ap.gsea.intersect <- Ap.gsea.intersect[,c("Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
 colnames(Ap.gsea.intersect)[which(colnames(Ap.gsea.intersect) == "parentchild")] <- "P"
@@
 
-GSEA for genes that are *Intermediate* in A. carolinensis and *Bimodal* in A. picea
-
-```{r gsea_Ar_int_A22_bim, results = 'hide'}
-# genes with *Intermediate* expression in Ar and *Bimodal* in A22
-Ar.int.A22.bim <- interaction.response.type[which(interaction.response.type$Ar.type == "Intermediate" & interaction.response.type$A22.type == "Bimodal"), "Transcript"]
-Ar.int.A22.bim.geneList <- rep(0, length = length(Ap.geneList))
-names(Ar.int.A22.bim.geneList) <- names(Ap.geneList)
-Ar.int.A22.bim.geneList[(which(names(Ar.int.A22.bim.geneList) %in% Ar.int.A22.bim))] <- 1
+GSEA for genes that are **Bimodal** in *A. picea* and downregulated (**Intermediate**) in *A. carolinensis*.
+
+```{r gsea_A22_bim_Ar_int, results = 'hide'}
+A22.bim.Ar.int <- sig.response.type[which(sig.response.type$Ar.type == "Intermediate" & sig.response.type$A22.type == "Bimodal"), "Transcript"]
+A22.bim.Ar.int.geneList <- rep(0, length = length(Ap.geneList))
+names(A22.bim.Ar.int.geneList) <- names(Ap.geneList)
+A22.bim.Ar.int.geneList[(which(names(A22.bim.Ar.int.geneList) %in% A22.bim.Ar.int))] <- 1
 # check correct number of values set to 1
-table(Ar.int.A22.bim.geneList)
+table(A22.bim.Ar.int.geneList)
 # run GSEA
-Ar.int.A22.bim.gsea <- gsea(genelist = Ar.int.A22.bim.geneList, geneID2GO = geneID2GO, plotpath = NA)
-```
-
-Cluster by semantic similarity
-
-```{r gsea_Ar_int_A22_bim_cluster}
-# similarity among enriched GO terms
-Ar.int.A22.bim.gsea.term.sim <- termSim(Ar.int.A22.bim.gsea$GO.ID, Ar.int.A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
-colnames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
-# distance matrix
-Ar.int.A22.bim.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.bim.gsea.term.sim))
-plot(Ar.int.A22.bim.gsea.term.sim.hclust)
-
-# report items in each cluster at height 1.5
-(Ar.int.A22.bim.hclust.report <- GSEAReportClusters(Ar.int.A22.bim.gsea.term.sim.hclust, h = 1))
-```
-
-GSEA for genes that are *Intermediate* in A. carolinensis and *Low* in A. picea
-
-```{r gsea_Ar_int_A22_low, results = 'hide'}
-# genes with *Intermediate* expression in Ar and *Low* in A22
-Ar.int.A22.low <- interaction.response.type[which(interaction.response.type$Ar.type == "Intermediate" & interaction.response.type$A22.type == "Low"), "Transcript"]
-Ar.int.A22.low.geneList <- rep(0, length = length(Ap.geneList))
-names(Ar.int.A22.low.geneList) <- names(Ap.geneList)
-Ar.int.A22.low.geneList[(which(names(Ar.int.A22.low.geneList) %in% Ar.int.A22.low))] <- 1
-# check correct number of values set to 1
-table(Ar.int.A22.low.geneList)
-# run GSEA
-Ar.int.A22.low.gsea <- gsea(genelist = Ar.int.A22.low.geneList, geneID2GO = geneID2GO, plotpath = NA)
-```
-
-```{r gsea_Ar_int_A22_low_cluster}
-# similarity among enriched GO terms
-Ar.int.A22.low.gsea.term.sim <- termSim(Ar.int.A22.low.gsea$GO.ID, Ar.int.A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
-colnames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
-# distance matrix
-Ar.int.A22.low.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.low.gsea.term.sim))
-plot(Ar.int.A22.low.gsea.term.sim.hclust)
-# report items in each cluster at height 1
-(Ar.int.A22.low.hclust.report <- GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.2))
-```
-
-GSEA for genes that are *Intermediate* in A. carolinensis and *Low* or *Bimodal* in A. picea
-
-```{r gsea_Ar_int_A22_lowbim, results = 'hide'}
-# genes with *Intermediate* expression in Ar and *Low* in A22
-Ar.int.A22.lowbim <- c(Ar.int.A22.low, Ar.int.A22.bim)
-Ar.int.A22.lowbim.geneList <- rep(0, length = length(Ap.geneList))
-names(Ar.int.A22.lowbim.geneList) <- names(Ap.geneList)
-Ar.int.A22.lowbim.geneList[(which(names(Ar.int.A22.lowbim.geneList) %in% Ar.int.A22.lowbim))] <- 1
-# check correct number of values set to 1
-table(Ar.int.A22.lowbim.geneList)
-# run GSEA
-Ar.int.A22.lowbim.gsea <- gsea(genelist = Ar.int.A22.lowbim.geneList, geneID2GO = geneID2GO, plotpath = NA)
-```
-
-```{r gsea_Ar_int_A22_lowbim_cluster}
-# similarity among enriched GO terms
-Ar.int.A22.lowbim.gsea.term.sim <- termSim(Ar.int.A22.lowbim.gsea$GO.ID, Ar.int.A22.lowbim.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ar.int.A22.lowbim.gsea.term.sim) <- Ar.int.A22.lowbim.gsea$Term
-colnames(Ar.int.A22.lowbim.gsea.term.sim) <- Ar.int.A22.lowbim.gsea$Term
-# distance matrix
-Ar.int.A22.lowbim.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.lowbim.gsea.term.sim))
-plot(Ar.int.A22.lowbim.gsea.term.sim.hclust)
-# report items in each cluster at height 1
-(Ar.int.A22.lowbim.hclust.report <- GSEAReportClusters(Ar.int.A22.lowbim.gsea.term.sim.hclust, h = 1.2))
-```
+A22.bim.Ar.int.gsea <- gsea(genelist = A22.bim.Ar.int.geneList, geneID2GO = geneID2GO, plotpath = NA)
+```
+
+The GO terms, *`r paste(A22.bim.Ar.int.gsea[, 2], collapse = ", ")`*, are enriched for genes that are **Bimodal** in *A. picea* and downregulated (**Intermediate**) in *A. carolinensis*.
 
@@
 
-GO enrichment for *A. picea* **High** genes.
+*A. picea*
 
@@
 A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
-```
-Clustering of enriched GO terms for *A. picea* **High** genes.
-
-```{r gsea_A22_high_cluster}
-# similarity among enriched GO terms
-A22.high.gsea.term.sim <- termSim(A22.high.gsea$GO.ID, A22.high.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
-colnames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
-# distance matrix
-A22.high.gsea.term.sim.hclust <- hclust(dist((A22.high.gsea.term.sim)))
-plot((A22.high.gsea.term.sim.hclust))
-# report items in each cluster at height 1.8
-GSEAReportClusters(A22.high.gsea.term.sim.hclust, h = 1.2)
-```
-
-GO enrichment for *A. picea* **Low** genes.
-
-```{r gsea_A22_low, results = 'hide'}
+
 # A22 'Low' genes
@@
 A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
-```
-
-Clustering of enriched GO terms for *A. picea* **Low** genes.
-
-```{r gsea_A22_low_cluster}
-# similarity among enriched GO terms
-A22.low.gsea.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
-colnames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
-# distance matrix
-A22.low.gsea.term.sim.hclust <- hclust(dist((A22.low.gsea.term.sim)))
-plot((A22.low.gsea.term.sim.hclust))
-# report items in each cluster at height 1.8
-GSEAReportClusters(A22.low.gsea.term.sim.hclust, h = 2)
-```
-
-GO enrichment for *A. picea* **Bimodal** genes.
-
-```{r gsea_A22_bim, results = 'hide'}
+
 # A22 'Bim' genes
@@
 A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
-```
-
-Clustering of enriched GO terms for *A. picea* **Bimodal** genes.
-
-```{r gsea_A22_bim_cluster}
-# similarity among enriched GO terms
-A22.bim.gsea.term.sim <- termSim(A22.bim.gsea$GO.ID, A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
-colnames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
-# distance matrix
-A22.bim.gsea.term.sim.hclust <- hclust(dist((A22.bim.gsea.term.sim)))
-plot((A22.bim.gsea.term.sim.hclust))
-# report items in each cluster at height 1.8
-GSEAReportClusters(A22.bim.gsea.term.sim.hclust, h = 1)
-```
-
-GO enrichment for *A. picea* **Intermediate** genes.
-
-```{r gsea_A22_int, results = 'hide'}
+
 # A22 'Int' genes
@@
 
-Clustering of enriched GO terms for *A. picea* **Intermediate** genes.
-
-```{r gsea_A22_int_cluster}
-# similarity among enriched GO terms
-A22.int.gsea.term.sim <- termSim(A22.int.gsea$GO.ID, A22.int.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
-colnames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
-# distance matrix
-A22.int.gsea.term.sim.hclust <- hclust(dist((A22.int.gsea.term.sim)))
-plot((A22.int.gsea.term.sim.hclust))
-# report items in each cluster at height 1.8
-GSEAReportClusters(A22.int.gsea.term.sim.hclust, h = 1.3)
-```
-
-GO enrichment for *A. carolinensis* **High** genes.
+*A. carolinensis*
 
@@
 Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
-```
-
-Clustering of enriched GO terms for *A. carolinensis* **High** genes.
-
-```{r gsea_Ar_high_cluster}
-# similarity among enriched GO terms
-Ar.high.gsea.term.sim <- termSim(Ar.high.gsea$GO.ID, Ar.high.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ar.high.gsea.term.sim) <- Ar.high.gsea$Term
-colnames(Ar.high.gsea.term.sim) <- Ar.high.gsea$Term
-# distance matrix
-Ar.high.gsea.term.sim.hclust <- hclust(dist((Ar.high.gsea.term.sim)))
-plot((Ar.high.gsea.term.sim.hclust))
-# report items in each cluster
-GSEAReportClusters(Ar.high.gsea.term.sim.hclust, h = 1.2)
-```
-GO enrichment for *A. carolinensis* **Low** genes.
-
-```{r gsea_Ar_low, results = 'hide'}
+
 # Ar 'Low' genes
@@
 Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
-```
-
-Clustering of enriched GO terms for *A. carolinensis* **Low** genes.
-
-```{r gsea_Ar_low_cluster}
-# similarity among enriched GO terms
-Ar.low.gsea.term.sim <- termSim(Ar.low.gsea$GO.ID, Ar.low.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ar.low.gsea.term.sim) <- Ar.low.gsea$Term
-colnames(Ar.low.gsea.term.sim) <- Ar.low.gsea$Term
-# distance matrix
-Ar.low.gsea.term.sim.hclust <- hclust(dist((Ar.low.gsea.term.sim)))
-plot((Ar.low.gsea.term.sim.hclust))
-# report items in each cluster at height 1.8
-GSEAReportClusters(Ar.low.gsea.term.sim.hclust, h = 2)
-```
-
-GO enrichment for *A. carolinensis* **Bimodal** genes.
-
-```{r gsea_Ar_bim, results = 'hide'}
+
 # Ar 'Bim' genes
@@
 Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
-```
-
-Clustering of enriched GO terms for *A. carolinensis* **Bimodal** genes.
-
-```{r gsea_Ar_bim_cluster}
-# similarity among enriched GO terms
-Ar.bim.gsea.term.sim <- termSim(Ar.bim.gsea$GO.ID, Ar.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ar.bim.gsea.term.sim) <- Ar.bim.gsea$Term
-colnames(Ar.bim.gsea.term.sim) <- Ar.bim.gsea$Term
-# distance matrix
-Ar.bim.gsea.term.sim.hclust <- hclust(dist((Ar.bim.gsea.term.sim)))
-plot((Ar.bim.gsea.term.sim.hclust))
-# report items in each cluster at height 1.8
-GSEAReportClusters(Ar.bim.gsea.term.sim.hclust, h = 1)
-```
-
-GO enrichment for *A. carolinensis* **Intermediate** genes.
-
-```{r gsea_Ar_int, results = 'hide'}
+
 # Ar 'Int' genes
@@
 
-Clustering of enriched GO terms for *A. carolinensis* **Intermediate** genes.
-
-```{r gsea_Ar_int_cluster}
-# similarity among enriched GO terms
-Ar.int.gsea.term.sim <- termSim(Ar.int.gsea$GO.ID, Ar.int.gsea$GO.ID, ont = "BP", organism = 'fly')
-# replace GO IDs with terms
-rownames(Ar.int.gsea.term.sim) <- Ar.int.gsea$Term
-colnames(Ar.int.gsea.term.sim) <- Ar.int.gsea$Term
-# distance matrix
-Ar.int.gsea.term.sim.hclust <- hclust(dist((Ar.int.gsea.term.sim)))
-plot((Ar.int.gsea.term.sim.hclust))
-# report items in each cluster
-GSEAReportClusters(Ar.int.gsea.term.sim.hclust, h = 1.8)
-```
-
-Combine gene set enrichment analysis for each species and category into single table.
+Combined gene set enrichment analysis for each species and category into single table.
 
@@
 # combine
-Ap.gsea.union <- rbind(A22.gsea, Ar.gsea)
+Ap.gsea.by.species <- rbind(A22.gsea, Ar.gsea)
 # reorder
-Ap.gsea.union <- Ap.gsea.union[,c("Species", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
-colnames(Ap.gsea.union)[8] <- "P"
-
-write.csv(Ap.gsea.union, file = paste(resultsdir, "Ap_gsea_union", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
-```
-
-## Visualize responsive transcripts
-
-Plots for all genes expressed at *High* temps in GO category "GO:0006950: response to stress"
-
-```{r plot_GOstress, echo=FALSE, eval=TRUE}
-high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
-
-# in low category
-low.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
-# in bimodal category
-bim.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Bimodal", Ar.type)), with(GO0006950.responsive, grep("Bimodal", A22.type))), ]
-# in intermediate category
-int.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Intermediate", Ar.type)), with(GO0006950.responsive, grep("Intermediate", A22.type))), ]
-
-# Line plot, expression against temp, faceted by species
-trp1 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
-  geom_line() + 
-  facet_grid(. ~ colony2) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(trp1)
-
-# Smooth plot, expression against temp, faceted by colony
-trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
-  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
-  facet_grid(. ~ colony2) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(trp2)
-```
-
-Make plots for all genes expressed at *Low* temps in GO category "GO:0006950: response to stress"
-
-```{r plot_GOstress_low, echo=FALSE, eval=TRUE, cache=TRUE}
-low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
-
-# Line plot, expression against temp, faceted by colony
-trp3 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
-  geom_line() + 
-  facet_grid(. ~ colony2) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(trp3)
-
-# Smooth plot, expression against temp, faceted by colony
-trp4 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
-  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
-  facet_grid(. ~ colony2) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(trp4)
-```
-
-```{r ggplot_high}
-trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high.transcripts][1:220,], aes(x=val, y=TPM.scaled, group=Transcript)) +
-  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
-  facet_grid(. ~ colony2) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(trp_A22_high)
-
-### predicted values
-resp.TPM.dt.sub.pred[,pTPM.scaled:=scale(pTPM), by = Transcript]
-setkey(resp.TPM.dt.sub.pred, "Transcript")
-
-pred_A22_high <- ggplot(resp.TPM.dt.sub.pred[A22.high.transcripts][1:220,], aes(x=val, y=pTPM.scaled, group=Transcript)) +
-  geom_line() + 
-  facet_grid(. ~ colony) + 
-  scale_y_continuous(name="Expression (scaled)") +
-  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
-print(pred_A22_high)
-
-png("RxN_example.png")
-ggplot(resp.TPM.dt.sub.pred["100348|*|comp3450455_c0_seq1"], aes(x=val, y=pTPM.scaled, group=Transcript, color = Transcript)) +
-  geom_line() + 
-  facet_grid(. ~ colony) +
-  labs(y = "Expression (scaled)", x = expression("Temperature ", degree, "C")) +
-  theme(legend.position = "none")
-dev.off()
+Ap.gsea.by.species <- Ap.gsea.by.species[,c("Species", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
+colnames(Ap.gsea.by.species)[8] <- "P"
+
+write.csv(Ap.gsea.by.species, file = paste(resultsdir, "Ap_gsea_by_species", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
 ```
@@
 # save RData for manuscript file
-save(TPM.dt.sub, RxNpval, sigtable, chi1, chi2, chi3, chi4, chi5, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, Ap.BP.resultParentChild, resp.stress.responsive.count, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, quadratic.lms, resp.TPM.dt.sub.pred, md, file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
-
-save(sigtable, chi1, chi2, chi3, chi4, chi5, chi6, resp.TPM.dt.sub, Ap.type.table, Ap.type.table.prop, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, md, responsive_transcripts_w_BLAST, annotation.table.responsive, file = paste("doc/ms_slim.RData", sep = ""))
+save(Ap.type.table, annotation.table, annotation.table.responsive, Ap.type.table.prop, 
+     chi1, chi2, chi3, chi4, chi5, chi6, D, mh.test, pre.signif.transcripts,  
+     responsive_transcripts_w_BLAST, sigtable, signif.transcripts, T_on.low.ttest, 
+     T_on.high.ttest, T_off.int.high.ttest, T_off.int.low.ttest, w3, 
+     file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
 ```

### 2015-09-30 doc/ApTranscriptome_manuscript.Rmd
added resampling analysis: greatly reduces number of significant transcripts

@@
 output:
+  word_document: default
   pdf_document:
-    fig_caption: true
+    fig_caption: yes
     includes:
@@
   html_document:
-        keep_md: true
-  word_document: default
+    keep_md: yes
 includes: header.tex
@@
 We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth.
-We found that at least 9% of all genes were thermally responsive.
+We found that at least 2% of all genes were thermally responsive.
 The majority of upregulation was specific to exposure to low temperatures.
 The cool-adapted *A.
-picea* induced expression of more genes in response to extreme temperatures, including genes coding for heat shock proteins, than did *A.
+picea* induced expression of more genes in response to extreme temperatures than did *A.
 carolinensis*, consistent with the *enhanced response* hypothesis.
@@
 carolinensis* downregulated many of the genes upregulated in *A.
-picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis.
+picea*, and counterintuitively, required more extreme temperatures to induce changes in gene expression towards low temperature extremes, consistent with the *tolerance* hypothesis.
 We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.
@@
 #options(scipen = 1, digits = 3)
-load("ms_slim.RData")
+load("ms_2015-09-29.RData")
 library(data.table)
@@
 To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
-The reaction norms of individual transcripts can be searched by gene ontology and viewed at https://johnsg.shinyapps.io/ApRxN-shinyapp.
+We evaluated the robustness of this approach by re-assigning each transcript within a species randomly to a temperature, fitting the polynomial model as above, calculating FDR and then determining how many transcripts were under the 5% FDR threshold for the resampled data, when we would expect there to be no significant transcripts.
+We repeated this resampling approach 100 times, and used the 95th quantile of false significant transcripts as the null expectation, only retaining the transcripts that were more significant from the real data.
+We identified thermally-responsive transcripts as the subset that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
+All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supporting Information).
 
-We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
-All analyses were performed with R 3.1 [@r_core_team2013] and are fully reproducible (Supporting Information).
-For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species.
+For each thermally-responsive transcript, we predicted expression levels using the final linear model for each species across the tested thermal range.
 We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other.
@@
 
-The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts.
-About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 9% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995].
+The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts.
+About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
+The resampling analysis revealed that the polynomial model was prone to overfitting the data.
+The 95th percentile of the resampled data returned a null expectation of `r quantile(D$num_signif, probs = 0.95)` putatively significant transcripts, when we would expect no significant relationship.
+With the true data, there were `r nrow(pre.signif.transcripts)` transcripts below the 5% FDR threshold, of which we retained only the top `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` as true positives given the null expectation from the resampling.
 Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
@@
 
-Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species.
-The majority of the responsive transcripts were in the **Low** category for both species (`r Ap.type.table.prop[2,"Low"]`% and `r Ap.type.table.prop[1,"Low"]`% for *A.
-carolinensis* and *A.
-picea*, respectively), followed by **Intermediate** (`r Ap.type.table.prop[2,"Intermediate"]`%), **High** (`r Ap.type.table.prop[2,"High"]`%),**Bimodal** (`r Ap.type.table.prop[2,"Bimodal"]`%) and **Not Responsive** (`r Ap.type.table.prop[2,"NotResp"]`%) for *A.
-carolinensis*, and **Bimodal** (`r Ap.type.table.prop[1,"Bimodal"]`%), **High** (`r Ap.type.table.prop[1, "High"]`%), **Intermediate** (`r Ap.type.table.prop[1,"Intermediate"]`%) and **Not Responsive** (`r Ap.type.table.prop[1,"NotResp"]`%) for *A.
-picea*.
-
 ## Expression response to thermal extremes differs between species
 
-The two species differed in the number of transcripts in each expression category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
+Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`), the two species differed in the number of transcripts in each expression category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
 *Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A.
@@
 Consistent with the *enhanced response* hypothesis, the cool-climate *A.
-picea* induced 1,146 (30%) more transcripts in response to low temperatures than the warm-climate *A.
+picea* induced `r Ap.type.table["A.
+picea","Low"] - Ap.type.table["A.
+carolinensis","Low"]` (~50%) more transcripts in response to low temperatures than the warm-climate *A.
 carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`).
-However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced response hypothesis*, at high temperatures *A.
-carolinensis* induced expression of ~14% fewer transcripts than *A.
-picea* (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* < `r max(chi5$p.value, 0.001)`).
+However, there was no difference among species in the number of transcripts upregulated at high temperatures (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* = `r round(chi5$p.value, 2)`).
 
@@
 S1), indicating that these differences were unlikely to be due to neutral processes.
-Specifically,
 
-The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species (Fig.
+The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig.
 1).
 The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig.
-1).
+1, left).
 In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig.
-1).
-For each temperature extreme, we tested these predictions by focusing on the transcripts upregulated by the species that expriences less extreme temperatures (the warm-climate *A.
-carolinensis* for cool temperatures, the cool-climate *A.
-picea* for warm temperatures) and testing whether those transcripts displayed the same response profile in species experiencing more extreme temperatures.
-As with the overall results, the patterns of association differed for the two temperature extremes.
+1, right).
+We tested these predictions by testing if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A.
+carolinensis*, and warm temperatures for the cool-climate *A.
+picea*) displayed the same response profile in the species that more frequently experiences those conditions.
+
 Transcripts upregulated at low temperatures in *A.
-carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from other expression categories in *A.
+carolinensis* (**Low** and **Bimodal** transcripts) were less likely to be upregulated in any expression category in *A.
 picea* (Fig.
-2A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis.
+2A), with the exception of an excess of transcripts that shifted to the **High** category.
+This pattern is partially consistent with the *enhanced response* hypothesis, as the same transcripts are used to respond the the temperature extreme less frequently encountered in each species.
 In contrast, transcripts upregulated in response to high temperatures in *A.
@@
 carolinensis* (Fig.
-2B), primarily to the **Not Responsive** and **Intermediate** expression categories (Fig.
+2B), primarily to the **Intermediate** expression category (Fig.
 2B).
@@
 
-The molecular processes in the set of transcripts that switched from upregulation in *A.
-picea* to downregulation in *A.
-carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` significantly enriched GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism.
-In addition, there were GO terms related to organelle repair (e.g.
-histone modification), transcript regulation (e.g.
-positive regulation of cellular process) and cellular organization (e.g.
-transport, organelle organization).
-Collectively, these results suggest that in response to both thermal extremes, the warmer-climate species, *A.
-carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A.
-picea*, increases metabolism and initiates damage repair responses.
+## Molecular processes suggest a generalized stress response mechanism
+
+The results of the gene set enrichment analysis revealed a number of gene groups enriched in each expression category.
+Across both species, there were 9 terms enriched in the **Bimodal** category, including terms such involved in stress response (regulation of cellular response to stress, signal transduction by p53 class mediator), cell death (apoptotic signaling pathway) and cellular organization (e.g.
+protein complex localization).
+The 6 terms enriched in the **Low** category suggest that proteins undergo structural (e.g protein acylation) and organizational (single-organism organelle organization) changes to tolerate colder temperatures, possibly to maintain membrane fluidity [@ohtsu1998].
+In the **High** category had only a single enriched term, "nicotinamide metabolic process", while the **Intermediate** category had 5 terms including DNA packaging and metabolic process terms.
 
@@
 We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression.
-In support of the *tolerance hypothesis* but not the *enhanced response hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A.
-carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A.
-picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig.
-3A).
-For low temperatures, *A.
-carolinensis* also had the more extreme induction temperature (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A.
-carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A.
-picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
-3A), though the magnitude of difference was smaller than at high temperatures.
+In support of the *tolerance hypothesis* but not the *enhanced response hypothesis*, the temperature of induction at low temperatures was more extreme for *A.
+carolinensis* (`r round(T_on.low.ttest$estimate[1], 1)`°C) than *A.
+picea* (`r round(T_on.low.ttest$estimate[2], 1)`°C; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+3A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).
 
 Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
-Thresholds for downregulation were shifted more toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig.
-3).
 As for the upregulated transcripts, *A.
-carolinensis* had greater inertia of gene expression in response to high and low temperature extremes.
-With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
+carolinensis* had greater inertia of gene expression in response to increasing temperatures.
+The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
 carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
 picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`).
-The difference between temperatures was not significant with decreasing temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)` for *A.
-picea* compared to `r round(T_off.int.low.ttest$estimate[1], 1)` for *A.
-carolinensis*; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
+The difference between temperatures was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
 3B).
@@
 4B).
-In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species.
-
+In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001).
 In addition, the thermally responsive transcripts in *A.
@@
 2B).
-Rather than having constitutively elevated expression as predicted, these transcripts were uniformly low or not expressed (mean TPM = 0.08) in *A.
-carolinensis*.
 
@@
 carolinensis*.
-We found large-scale, non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.
+We found non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.
 
@@
 picea* is substantially lower than *A.
-carolinensis*, the two species upregulated responsive transcripts near the same threshold temperature (Fig.
+carolinensis*, *A.
+picea* also upregulated responsive transcripts at slightly less exterme temperatures (Fig.
 3A).
 Moreover, the transcripts upregulated in *A.
-picea* included nearly the entire set upregulated in *A.
-carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1).
+picea* included about half the set upregulated in *A.
+carolinensis* as well as an additional set of ~250 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1).
 Two non-mutually exclusive hypotheses may explain this pattern.
@@
 High temperatures were associated with significantly fewer upregulated transcripts in *A.
-carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A.
+carolinensis* (Table 1), and a large proportion (28%) of those transcripts upregulated at high temperatures in *A.
 picea* were either downregulated or expressed at negligible levels overall in *A.
 carolinensis*.
-Similarly, significantly higher temperatures were required to induce the expression response for *A.
-carolinensis* than *A.
-picea* towards the high thermal limit (Fig.
-3B-D).
-This difference between species is also evident among candidate transcripts known to be involved in stress response.
-Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A.
-picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A.
-carolinensis*.
-Collectively, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.
+These results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.
 Such mechanisms may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress.
@@
 2), it is worth noting a number of similarities.
-The two species showed similar numbers of transcripts with a **Bimodal** generalized stress-response (Table 1).
-In both species, there were 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1).
+In both species, there were 2 -- 3 times more transcripts upregulated at low than high temperatures (Table 1).
 The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures.
@@
 Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
-Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were much more similar (Fig.
+Only ~10% of transcripts upregulated in response to temperature were bimodal, and while activation thresholds differed at low temperatures they did not at high temperatures (Fig.
 3).
@@
 The (A) *enhanced response* hypothesis proposes that the increase in cold tolerance in *A.
-picea* is achieved by amplifying existing molecular mechanisms, and thus there should be an excess of shared resonse types between species.
+picea* is achieved by amplifying existing molecular mechanisms, and thus there should be an excess of shared response types between species.
 In contrast, the (B) *tolerance hypothesis* predicts that *A.

### 2015-10-01 ApRxN-shinyapp/server.R
clean up - removed shiny app

@@ -1,46 +1,0 @@
-library(shiny)
-library(ggplot2)
-library(stringr)
-library(plyr)
-
-# load data
-gxdata <- read.csv("responsive.transcripts.TPM.csv")
-
-# Define server logic required to plot variables 
-shinyServer(function(input, output) {
-  
-  # Caption text
-  captionText <- reactive({
-    paste("Reaction norm for ", input$searchtext, "related genes")
-  })
-  
-  # Return the formula text for printing as a caption
-  output$caption <- renderText({
-    captionText()
-  })
-  
-  # Subset data to transcripts that have a match to 'searchtext' in best.hit.to.nr or GO.Biological.Process
-  
-  ## gsub out any punctuation to avoid misspellings
-  
-  datasub <- reactive({
-    gxdata[union(grep(input$searchtext, gxdata$best.hit.to.nr), grep(input$searchtext, gxdata$GO.Biological.Process)), ]
-  })
-  
-  
-  # Extract GO annotation to display
-  output$GOannotation <- renderPrint({
-    foo <- datasub()
-    unlist(str_split(foo[1, "GO.Biological.Process"], " ", n=2))[2]
-  })
-  
-  # Generate a plot of the requested variable against temperature
-  output$RxNplot <- renderPlot({
-    #cat("gdata", dim(gdata), "\n")
-    g <- ggplot(datasub(), aes(x=val, y=exp.scaled, group=Transcript)) + 
-      geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(colour = exp_type)) + 
-      facet_grid(. ~ colony)
-    print(g)
-  })
-  
-})

### 2015-10-01 ApRxN-shinyapp/ui.R

@@ -1,25 +1,0 @@
-library(shiny)
-
-# Define UI for selecting GO term
-shinyUI(pageWithSidebar(
-    
-  # Application title
-  headerPanel("Aphaenogaster thermally-responsive genes"),
-  
-  # Sidebar to select which GO term to plot transcripts for
-  sidebarPanel(
-    textInput("searchtext", "Search term:", value = "response to stress"),
-    submitButton(text = "Apply Changes"),
-    br(),
-    p("Search terms to try: heat shock protein, behavior, phagocytosis, immune system, GO:0006950")
-  ),
-  
-  # Display caption and plot of requested GO term
-  mainPanel(
-    h3(textOutput("caption")),
-    
-    plotOutput("RxNplot"),
-    
-    verbatimTextOutput("GOannotation")
-    )
-))

### 2015-10-01 ApTranscriptome_TR.Rmd
tech report to md

@@
 ---
-title: Technical Report for **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**
-author: [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
+title: Technical Report for **Thermal reactionomes reveal divergent responses to thermal
+  extremes in warm and cool-climate ant species**
+author: '[John Stanton-Geddes](john.stantongeddes.research@gmail.com)'
 output:
-  word_document: default
+  html_document:
+    keep_md: yes
   pdf_document:
@@
     toc_depth: 2
-  html_document:
-    keep_md: yes
+  word_document: default
 ---
@@
 mh_plot_sub1 <- qplot(y=Apic, x=Acar, data=md_sub1, fill=value, geom="tile", xlim = rev(levels(md_sub1$Acar))) + 
-  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
+  scale_fill_gradient2(limits=c(-205, 205), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 mh_plot_sub2 <- qplot(y=Acar, x=Apic, data=md_sub2, fill=value, geom="tile", xlim = rev(levels(md_sub2$Apic))) + 
-  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
+  scale_fill_gradient2(limits=c(-205, 205), low="#f1a340", high="#998ec3") +
   theme(axis.title = element_text(face="italic")) +
@@
 
-## Shiny interactive web-app
-
-To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package.
-The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.
-The web app is available at https://johnsg.shinyapps.io/ApRxN-shinyapp/
-
-Export data for interactive shiny app.
-
-```{r shiny_file, eval=FALSE, echo=FALSE}
-shiny.dt <- TPM.dt.sub[responsive.lms.ann.type[,list(Transcript, best.hit.to.nr, GO.Biological.Process)]]
-
-# scale expression values 
-shiny.dt[,TPM.scaled:=scale(TPM), by = Transcript]
-str(shiny.dt)
-write.csv(shiny.dt, file = "ApRxN-shinyapp/responsive.lms.TPM.csv", quote = TRUE, row.names = FALSE)
-```
-
 ```{r rData, echo=FALSE}

### 2015-10-01 ApTranscriptome_resample.Rmd
script for resampling to calculate false-positive rate

# script to calculate empirical false-positive rate

library(MASS)
library(plyr)
library(dplyr)

# load functions
source("scripts/RxNseq.R")

# load real data
load("ApTranscriptome_exp_data.Rda")
str(TPM.dt.sub)

# dataframe to collect results
D <- data.frame(
  resamp = 1:10,
  num_signif = rep(NA, 10))

for(i in 1:100) {
  RxNresample <- TPM.dt.sub %>%
    group_by(Transcript, colony) %>%
    select(TPM) %>%
    sample_frac() %>%
    mutate(val = c(0, 3.5, 10.5, 14, 17.5, 21, 24.5, 28, 31.5, 35, 38.5))

  # define model for RxN function
  model <-  "log(TPM+1) ~ colony + val + I(val^2) + colony:val + colony:I(val^2)"

  # calculate overall P value and R^2 for each transcript
  RxNpval_resamp <- ddply(RxNresample, .(Transcript), .inform= TRUE, modpFunc, .parallel = TRUE)

  # adjust FDR
  RxNpval_resamp$padj <- p.adjust(RxNpval_resamp$pval, method = "fdr")

  # number of transcripts with FDR < 0.05
  numsignif <- length(which(RxNpval_resamp$padj < 0.05))

  # add to dataframe
  D[i, "num_signif"] <- numsignif
  print(D)
  }

save(D, file = "resample.out")

### 2015-10-26 doc/ApTranscriptome_manuscript.Rmd
revisions for BMC Genomics

@@
 author: ' '
-csl: molecol.csl
+csl: bmcgenomics.csl
 output:
@@
 
-**Short title**: Thermal reactionomes of warm and cool-climate ants
-
 **Authors**
@@
 \textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
+
 \textsuperscript{2} Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, Tenessee 37996 \newline
+
 \textsuperscript{3} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
+
 \textsuperscript{4} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
+
 \textsuperscript{5} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
+
 \textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen, Universitetsparken 15, DK-2100 Copenhagen
@@
 
-**Keywords**
+```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
+#options(scipen = 1, digits = 3)
+load("ms_2015-10-22.RData")
+library(data.table)
+library(ggplot2)
+library(reshape2)
+library(pander)
+library(RColorBrewer)
+library(topGO)
+library(plyr)
 
-*Aphaenogaster*, gene expression, plasticity, reactionome, transcriptome
+options(digits = 3)
+panderOptions('round', 3)
+panderOptions('keep.trailing.zeros', TRUE)
+```
 
@@
 carolinensis* downregulated many of the genes upregulated in *A.
-picea*, and counterintuitively, required more extreme temperatures to induce changes in gene expression towards low temperature extremes, consistent with the *tolerance* hypothesis.
+picea*, and required more extreme temperatures to induce down-regulation in gene expression, consistent with the *tolerance* hypothesis.
 We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.
@@
 
-# Introduction
+**Keywords**
+
+*Aphaenogaster*, gene expression, plasticity, reactionome, transcriptome
+
+# Background
 
@@
 
-# Materials and Methods
+# Results
 
-```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
-#options(scipen = 1, digits = 3)
-load("ms_2015-09-29.RData")
-library(data.table)
-library(ggplot2)
-library(reshape2)
-library(pander)
-library(RColorBrewer)
-library(topGO)
-library(plyr)
+## Reaction norms of thermally-responsive transcripts
 
-options(digits = 3)
-panderOptions('round', 3)
-panderOptions('keep.trailing.zeros', TRUE)
+The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts.
+About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
+The resampling analysis revealed that the polynomial model was prone to overfitting the data.
+The 95th percentile of the resampled data returned a null expectation of `r quantile(D$num_signif, probs = 0.95)` putatively significant transcripts, when we would expect no significant relationship.
+With the true data, there were `r nrow(pre.signif.transcripts)` transcripts below the 5% FDR threshold, of which we retained only the top `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`)% as true positives given the null expectation from the resampling.
+Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
+high vs.
+low temperature).
+The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`).
+However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.
+
+## Expression response to thermal extremes differs between species
+
+Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`), the two species differed in the number of transcripts in each expression category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
+*Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A.
+carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`).
+Consistent with the *enhanced response* hypothesis, the cool-climate *A.
+picea* induced `r Ap.type.table["A.
+picea","Low"] - Ap.type.table["A.
+carolinensis","Low"]` (~50%) more transcripts in response to low temperatures than the warm-climate *A.
+carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`).
+However, there was no difference among species in the number of transcripts upregulated at high temperatures (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* = `r round(chi5$p.value, 2)`).
+
+```{r echo=FALSE, results='hide'}
+# results of Stuart-Maxwell test
+mh.test
 ```
+
+In addition, we also examined the specific patterns of shifts from one expression category to another between species.
+When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, P < 0.001, Fig.
+S1).
+Specifically, expression patterns, which in slope as well as category, differed significantly between species for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts, most of which shifted from one expression category to another.
+
+The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig.
+1).
+The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig.
+1, left).
+In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig.
+1, right).
+We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A.
+carolinensis*, and warm temperatures for the cool-climate *A.
+picea*) displayed the same response profile in the species that more frequently experiences those conditions.
+
+Transcripts upregulated at low temperatures in *A.
+carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from other expression categories in *A.
+picea* (Fig.
+2A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis.
+In contrast, transcripts upregulated in response to high temperatures in *A.
+picea* (**High** and **Bimodal**) shifted expression categories in *A.
+carolinensis* (Fig.
+2B), primarily to the **Intermediate** category (Fig.
+2B).
+These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.
+
+## Molecular processes suggest a generalized stress response mechanism
+
+The gene set enrichment analysis revealed a number of gene groups enriched in each expression category.
+Across both species, there were 9 terms enriched in the **Bimodal** category, including terms involved in stress response (regulation of cellular response to stress, signal transduction by p53 class mediator), cell death (apoptotic signaling pathway) and cellular organization (e.g.
+protein complex localization).
+The 6 terms enriched in the **Low** category suggest that proteins undergo structural (e.g protein acylation) and organizational (single-organism organelle organization) changes to tolerate colder temperatures, possibly to maintain membrane fluidity [@ohtsu1998].
+The **High** category included only a single enriched GO term, "nicotinamide metabolic process", while the **Intermediate** category had 5 terms including DNA packaging and metabolic process terms.
+
+## *A. carolinensis* has greater inertia of expression change to increases in temperature than does *A. picea*
+
+As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures.
+We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression.
+In support of the *enhanced response* but not the *tolerance hypothesis*, the temperature of induction at low temperatures was significantly higher for the cool-climate *A.
+picea* than for *A.
+carolinensis* (`r round(T_on.low.ttest$estimate[1], 1)`°C) than *A.
+picea* (`r round(T_on.low.ttest$estimate[2], 1)`°C; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
+3A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).
+
+Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
+Consistent with the *tolerance hypothesis*, *A.
+carolinensis* had greater inertia of gene expression in response to increasing temperatures.
+The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
+carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
+picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`).
+The difference between temperatures was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
+3B).
+
+## No evidence for *genetic assimilation*
+
+We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C).
+If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two.
+We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A.
+picea* were not expressed at higher baseline levels in the warm-climate *A.
+carolinensis* (Fig.
+4A).
+Similarly, transcripts more upregulated at low temperatures in *A.
+carolinensis* did not show higher baseline levels in *A.
+picea* (Fig.
+4B).
+In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001).
+In addition, the thermally responsive transcripts in *A.
+carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A.
+picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`).
+An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A.
+picea* but were not thermally responsive in *A.
+carolinensis* (top-row of Fig.
+2B).
+These transcripts are less likely to be upregulated in any context, consistent with the tolerance hypothesis.
+
+# Discussion
+
+The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances.
+However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance.
+By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify putative changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A.
+carolinensis*.
+We found non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.
+
+Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
+We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A.
+picea* is substantially lower than *A.
+carolinensis*, *A.
+picea* upregulated responsive transcripts at slightly less extreme temperatures (Fig.
+3A).
+Moreover, the transcripts upregulated in *A.
+picea* included about about half (`r round(tt2["Low", "Low"] / colSums(tt2)["Low"], 2)*100`%) those upregulated in *A.
+carolinensis* as well as an additional set of `r round(rowSums(tt2)["Low"] - tt2["Low", "Low"], 2)` transcripts (Table 1), enriched for metabolism, organization and translation processes (Table S1).
+Two non-mutually exclusive hypotheses may explain this pattern.
+First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002].
+The northern species *A.
+picea* may induce a greater response to survive the longer winter period.
+Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012].
+This requirement may be under stronger selection in *A.
+picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.
+
+In contrast to cold tolerance, the enhanced upper thermal limit in *A.
+carolinensis* is best explained by the *tolerance* hypothesis.
+High temperatures were associated with significantly fewer upregulated transcripts in *A.
+carolinensis* (Table 1), and a large proportion (`r round(sum(tt2[c("Intermediate", "NotResp"), "High"]) / Ap.type.table["A.
+picea", "High"], 2) * 100`%) of the transcripts upregulated at high temperatures in *A.
+picea* were either downregulated or expressed at negligible levels overall in *A.
+carolinensis*.
+These results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.
+Such mechanisms may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress.
+These differences are in line with expectations that *A.
+carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions.
+Indeed, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
+
+The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts.
+This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013].
+However, in a short-term selection experiment for heat tolerance, Sikkink et al.
+[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance.
+Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
+Whether a particular taxon evolves one strategy over another may be related to availability of alternative mechanisms as well as the the intensity, frequency and duration of temperature stress in a given environment.
+
+Given the differences in the patterns of thermal responsiveness between species (Fig.
+2), it is worth noting a number of similarities.
+In both species, there were 2 -- 3 times more transcripts upregulated at low than high temperatures (Table 1).
+The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures.
+However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures.
+At these extremes, we also found few upregulated transcripts (Fig.
+3A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig.
+3).
+A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002].
+The observation that more transcripts were upregulated at low than high temperatures could also be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013].
+This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
+
+Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression in this study.
+Only ~10% of transcripts upregulated in response to temperature were bimodal, and for both activation and down-regulation, thresholds differed between species at only one temperature extreme (Fig.
+3).
+This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.
+
+A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach), which revealed quantitative patterns of temperature response not captured in categorical comparisons.
+For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as discussed above.
+Further, a number of issues have hampered RNA-seq studies to date.
+Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" interpretations [@pavlidis2012].
+Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species.
+Although we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
+
+Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
+Although we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig.
+2, Fig.
+S1).
+Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression -- as opposed to identifying lists of differentially-expressed transcripts -- can provide meaningful insight on the process of evolution.
+It should be noted, however, that although including non-linear relationships between expression and temperature captured a significantly larger range of biologically-relevant responses, it also led to a substantial increase in false positives.
+Empirical estimation of these rates via randomization tests, combined with robust sampling designs, can help to minimize this bias and focus results on biologically-meaningful gene sets.
+
+A number of caveats do apply to our work.
+First, species may differ in gene expression along axes which we have not measured here, especially temporal patterns of gene expression [@sorensen2005], which could be studied in further work.
+Second, the *de novo* transcriptome assembly is highly fragmented, given that all sequenced ant genomes to date have only about 18,000 genes [@gadau2012].
+Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts.
+A genome assembly, in progress, will help to reduce fragmentation.
+Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for model arthropods such as *Drosophila* and *Apis*.
+Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on fitness.
+In particular, functional protein levels cannot be expected to be fully linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].
+
+Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, whereas organisms from less stressful environments rely on plastically-induced resistance.
+Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].
+Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins [@krebs1997] or by increasing the thermal stability of proteins at the expense of catalytic activity [@arnold2001].
+Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.
+Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.
+
+## Conclusions
+
+In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A.
+carolinensis*.
+At least 2.5% of their transcriptomes are thermally responsive.
+Our results indicate that these two ant species have different responses to thermal extremes.
+*A.
+picea* responds by increasing expression of transcripts related to metabolism, stress response and other protective molecules, whereas *A.
+carolinensis* decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance.
+The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species.
+Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.
+
+# Methods
 
@@
 
-Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable [@cottingham2005].
-Thus, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature.
-To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time on 12 days.
+Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable [@gotelli2012].
+Ideally, the treatments should be replicated at each level of the predictor variable [@cottingham2005].
+However, even with no replication, the regression design is still more powerful than an ANOVA design with comparable replication, and provides an unbiased estimator of the slope [@gotelli2012].
+For these reasons, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature.
+
+To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time of day on 12 different days.
 We measured response to temperature with a one-hour static temperature application, which is ecologically relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012].
@@
 
-To identify transcripts that had significant changes in expression across the thermal gradient, we fit the regression model
+To identify transcripts that had significant changes in expression across the thermal gradient, we fit to each transcript an ordinary least-squares polynomial regression model
 
@@
 
-to each transcript.
 Temperature and species were both fixed effects, with a quadratic term included for temperature.
 We used $log(TPM + 1)$ as the response to control for skew in the expression data.
-For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005].
+For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@gotelli2012].
 This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature.
-To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold.
-We evaluated the robustness of this approach by re-assigning each transcript within a species randomly to a temperature, fitting the polynomial model as above, calculating FDR and then determining how many transcripts were under the 5% FDR threshold for the resampled data, when we would expect there to be no significant transcripts.
-We repeated this resampling approach 100 times, and used the 95th quantile of false significant transcripts as the null expectation, only retaining the transcripts that were more significant from the real data.
-We identified thermally-responsive transcripts as the subset that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
-All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supporting Information).
 
+To evaluate the statistical significance of the patterns, we computed parametric *P*-values for each model and adjusted these *P*-values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995].
+As a more stringent filter for false positives, we then randomly re-assigned each transcript within a species to a different temperature, fit the polynomial models as above, and again calculated *P*-values and FDR.
+Ideally, these randomized data sets should not yield any significant associations.
+We repeated this resampling approach 100 times, and used the 95th quantile of false significant transcripts as the null expectation for retaining transcripts from the true data.
+
+Of these overall significant transcripts, we identified thermally-responsive transcripts as the subset that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC.
 For each thermally-responsive transcript, we predicted expression levels using the final linear model for each species across the tested thermal range.
@@
 
-# Results
+All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supporting Information).
 
-## Reaction norms of thermally-responsive transcripts
+## Availability of supporting data
 
-The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts.
-About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-The resampling analysis revealed that the polynomial model was prone to overfitting the data.
-The 95th percentile of the resampled data returned a null expectation of `r quantile(D$num_signif, probs = 0.95)` putatively significant transcripts, when we would expect no significant relationship.
-With the true data, there were `r nrow(pre.signif.transcripts)` transcripts below the 5% FDR threshold, of which we retained only the top `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` as true positives given the null expectation from the resampling.
-Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
-high vs.
-low temperature).
-The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`).
-However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.
+The Illumina short-read sequence data supporting the results of this article are available in the NCBI Short Read Archive BioProject repository, PRJNA260626 http://www.ncbi.nlm.nih.gov/bioproject/PRJNA260626/.
 
-## Expression response to thermal extremes differs between species
+The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available at the Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).
 
-Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`), the two species differed in the number of transcripts in each expression category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`).
-*Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A.
-carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`).
-Consistent with the *enhanced response* hypothesis, the cool-climate *A.
-picea* induced `r Ap.type.table["A.
-picea","Low"] - Ap.type.table["A.
-carolinensis","Low"]` (~50%) more transcripts in response to low temperatures than the warm-climate *A.
-carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`).
-However, there was no difference among species in the number of transcripts upregulated at high temperatures (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* = `r round(chi5$p.value, 2)`).
-
-```{r echo=FALSE, results='hide'}
-# results of Stuart-Maxwell test
-mh.test
-```
-
-In addition, we also examined the specific patterns of shifts from one expression category to another between species.
-Expression patterns differed for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts, most of which shifted from one expression category to another.
-These shifts in expression category between species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, P < 0.001, Fig.
-S1), indicating that these differences were unlikely to be due to neutral processes.
-
-The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig.
-1).
-The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig.
-1, left).
-In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig.
-1, right).
-We tested these predictions by testing if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A.
-carolinensis*, and warm temperatures for the cool-climate *A.
-picea*) displayed the same response profile in the species that more frequently experiences those conditions.
-
-Transcripts upregulated at low temperatures in *A.
-carolinensis* (**Low** and **Bimodal** transcripts) were less likely to be upregulated in any expression category in *A.
-picea* (Fig.
-2A), with the exception of an excess of transcripts that shifted to the **High** category.
-This pattern is partially consistent with the *enhanced response* hypothesis, as the same transcripts are used to respond the the temperature extreme less frequently encountered in each species.
-In contrast, transcripts upregulated in response to high temperatures in *A.
-picea* (**High** and **Bimodal**) shifted expression categories in *A.
-carolinensis* (Fig.
-2B), primarily to the **Intermediate** expression category (Fig.
-2B).
-These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.
-
-## Molecular processes suggest a generalized stress response mechanism
-
-The results of the gene set enrichment analysis revealed a number of gene groups enriched in each expression category.
-Across both species, there were 9 terms enriched in the **Bimodal** category, including terms such involved in stress response (regulation of cellular response to stress, signal transduction by p53 class mediator), cell death (apoptotic signaling pathway) and cellular organization (e.g.
-protein complex localization).
-The 6 terms enriched in the **Low** category suggest that proteins undergo structural (e.g protein acylation) and organizational (single-organism organelle organization) changes to tolerate colder temperatures, possibly to maintain membrane fluidity [@ohtsu1998].
-In the **High** category had only a single enriched term, "nicotinamide metabolic process", while the **Intermediate** category had 5 terms including DNA packaging and metabolic process terms.
-
-## *A. carolinensis* has greater inertia of expression change to increases in temperature than does *A. picea*
-
-As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures.
-We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression.
-In support of the *tolerance hypothesis* but not the *enhanced response hypothesis*, the temperature of induction at low temperatures was more extreme for *A.
-carolinensis* (`r round(T_on.low.ttest$estimate[1], 1)`°C) than *A.
-picea* (`r round(T_on.low.ttest$estimate[2], 1)`°C; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
-3A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).
-
-Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures.
-As for the upregulated transcripts, *A.
-carolinensis* had greater inertia of gene expression in response to increasing temperatures.
-The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A.
-carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A.
-picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`).
-The difference between temperatures was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
-3B).
-
-## No evidence for *genetic assimilation*
-
-We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C).
-If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two.
-We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A.
-picea* were not expressed at higher baseline levels in the warm-climate *A.
-carolinensis* (Fig.
-4A).
-Similarly, transcripts more upregulated at low temperatures in *A.
-carolinensis* did not show higher baseline levels in *A.
-picea* (Fig.
-4B).
-In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001).
-In addition, the thermally responsive transcripts in *A.
-carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A.
-picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`).
-An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A.
-picea* but were not thermally responsive in *A.
-carolinensis* (far-right column of Fig.
-2B).
-
-# Discussion
-
-The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances.
-However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance.
-By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify putative changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A.
-carolinensis*.
-We found non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.
-
-Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions.
-We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A.
-picea* is substantially lower than *A.
-carolinensis*, *A.
-picea* also upregulated responsive transcripts at slightly less exterme temperatures (Fig.
-3A).
-Moreover, the transcripts upregulated in *A.
-picea* included about half the set upregulated in *A.
-carolinensis* as well as an additional set of ~250 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1).
-Two non-mutually exclusive hypotheses may explain this pattern.
-First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002].
-The northern species *A.
-picea* may induce a greater response to survive the longer winter period.
-Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012].
-This requirement may be under stronger selection in *A.
-picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.
-
-In contrast to cold tolerance, the enhanced upper thermal limit in *A.
-carolinensis* is best explained by the *tolerance* hypothesis.
-High temperatures were associated with significantly fewer upregulated transcripts in *A.
-carolinensis* (Table 1), and a large proportion (28%) of those transcripts upregulated at high temperatures in *A.
-picea* were either downregulated or expressed at negligible levels overall in *A.
-carolinensis*.
-These results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.
-Such mechanisms may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress.
-These differences are in line with expectations that *A.
-carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions.
-Indeed, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
-
-The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts.
-This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013].
-However, in a short-term selection experiment for heat tolerance, Sikkink et al.
-[-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance.
-Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
-Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.
-
-Given the differences in the patterns of thermal responsiveness between species (Fig.
-2), it is worth noting a number of similarities.
-In both species, there were 2 -- 3 times more transcripts upregulated at low than high temperatures (Table 1).
-The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures.
-However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures.
-At these extreme low temperatures we also found few upregulated transcripts (Fig.
-3A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig.
-3).
-A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002].
-The observation that more transcripts were upregulated at low than high temperatures could also be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013].
-This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.
-
-Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.
-Only ~10% of transcripts upregulated in response to temperature were bimodal, and while activation thresholds differed at low temperatures they did not at high temperatures (Fig.
-3).
-This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.
-
-A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach), which revealed quantitative patterns of temperature response not captured in categorical comparisons.
-For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as we discussed above.
-Further, a number of issues have hampered RNA-seq studies to date.
-Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" interpretations [@pavlidis2012].
-Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species.
-Although we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.
-
-Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009].
-Although we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig.
-2, Fig.
-S1).
-Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression -- as opposed to identifying lists of differentially-expressed transcripts -- can provide meaningful insight on the process of evolution.
-
-A number of caveats do apply to our work.
-First, species may differ in gene expression along axes which we have not measured here, especially temporal patterns of gene expression [@sorensen2005], which could be studied in further work.
-Second, the *de novo* transcriptome assembly is highly fragmented, given that all sequenced ant genomes to date have only about 18,000 genes [@gadau2012].
-Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially assembled transcripts.
-A genome assembly, in progress, will help to reduce fragmentation.
-Third, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for model arthropods such as *Drosophila* and *Apis*.
-Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005], and large differences in patterns of gene expression may have only small effects on organismal fitness.
-In particular, functional protein levels cannot be expected to be fully linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability [@feder2005].
-
-Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, whereas organisms from less stressful environments rely on plastically-induced resistance.
-Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].
-Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins [@krebs1997] or by increasing the thermal stability of proteins at the expense of catalytic activity [@arnold2001].
-Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.
-Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.
-
-## Conclusion
-
-In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A.
-carolinensis*.
-At least 9% of their transcriptomes are thermally responsive.
-Our results indicate that these two ant species have different responses to thermal extremes.
-*A.
-picea* responds by increasing expression of transcripts related to metabolism, heat shock proteins and other protective molecules, whereas *A.
-carolinensis* decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance.
-The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species.
-Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.
-
-# Acknowledgements
-
-Support made possible by NSF DEB Award #1136644 and the Vermont Genetics Network through Grant Number 8P20GM103449 from the INBRE Program of the National Institute of General Medical Sciences (NIGMS) of the National Institutes of Health (NIH).
-Its contents are solely the responsibility of the authors and do not necessarily represent the official views of NIGMS or NIH.
-
-## Data Accessibility
-
-Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626.
-The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).
-The reproducible and version-controlled script for data analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
+The reproducible and version-controlled scripts for data analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
 
@@
 
+# Acknowledgements
+
+Support made possible by NSF DEB Award #1136644 and the Vermont Genetics Network through Grant Number 8P20GM103449 from the INBRE Program of the National Institute of General Medical Sciences (NIGMS) of the National Institutes of Health (NIH).
+Its contents are solely the responsibility of the authors and do not necessarily represent the official views of NIGMS or NIH.
+
 # Figure Legends
@@
 Each column indicates the distribution across all response categories in *A.
-picea*, which has a lower CTmin and is therefore better adapated to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CTmin, *A.
+picea*, which has a lower CT~min~ and is therefore better adapated to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CT~min~, *A.
 carolinensis*, relative to the null hypothesis of equal marginal frequencies.
-The dashed boxes highlight cells that would indicate matched responses in the two species, and the color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts.
+The dashed boxes highlight cells that would indicate matched responses in the two species, and the color of each cell (blue = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts.
 The (A) *enhanced response* hypothesis proposes that the increase in cold tolerance in *A.
@@
 Results of marginal frequencies analysis of thermal response categories.
-The color of each cell  (purple = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
+The color of each cell  (blue = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
 The units are number of transcripts.
@@
 picea* on columns).
-The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (purple = excess, orange = deficit).
+The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (blue = excess, orange = deficit).
 The expression types are **Low** transcripts that had greatest expression temperatures < 10°C, **Intermediate** transcripts with greatest expression between 10 and 30°C, **High** transcripts that had greatest expression at temperatures > 31°, **Bimodal** transcripts with increased expression at both high and low temperatures, and **Not Responsive** transcripts that were not thermally responsive in that species.
@@
 **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}
-
-rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")
 

### 2015-11-05 ApTranscriptome_TR.Rmd
revisions

@@
 output:
-  html_document:
-    keep_md: yes
+  word_document: default
   pdf_document:
@@
     toc_depth: 2
-  word_document: default
+  html_document:
+    keep_md: yes
 ---
@@
 (Ap.type.table <- rbind(A22.type.table, Ar.type.table))
+rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")
 
@@
      responsive_transcripts_w_BLAST, sigtable, signif.transcripts, T_on.low.ttest, 
-     T_on.high.ttest, T_off.int.high.ttest, T_off.int.low.ttest, w3, 
+     T_on.high.ttest, T_off.int.high.ttest, T_off.int.low.ttest, w3, tt2,
      file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))

### 2015-11-05 doc/ApTranscriptome_manuscript.Rmd
revisions

@@
 
-The distributions of species and their responses to climate change are in part determined by their thermal tolerances.
+**Background** The distributions of species and their responses to climate change are in part determined by their thermal tolerances.
 However, little is known about how thermal tolerance evolves.
@@
 We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth.
-We found that at least 2% of all genes were thermally responsive.
+
+**Results** We found that at least 2% of all genes were thermally responsive.
 The majority of upregulation was specific to exposure to low temperatures.
@@
 We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.
-These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.
+
+**Conclusions** These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.
 
@@
 About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-The resampling analysis revealed that the polynomial model was prone to overfitting the data.
-The 95th percentile of the resampled data returned a null expectation of `r quantile(D$num_signif, probs = 0.95)` putatively significant transcripts, when we would expect no significant relationship.
-With the true data, there were `r nrow(pre.signif.transcripts)` transcripts below the 5% FDR threshold, of which we retained only the top `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`)% as true positives given the null expectation from the resampling.
+We fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship.
+To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationshp by chance alone.
+We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) true positive transcripts that exceeded null expectation from the resampling analysis.
 Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
@@
 In addition, we also examined the specific patterns of shifts from one expression category to another between species.
-When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, P < 0.001, Fig.
+When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig.
 S1).
-Specifically, expression patterns, which in slope as well as category, differed significantly between species for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts, most of which shifted from one expression category to another.
+Specifically, the two species differed significantly in expression pattern, which captures differences in slope as well as category, for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts.
+Most of these shifted from one to another expression category.
 
@@
 picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`).
-The difference between temperatures was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
+The difference between species was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
 3B).
@@
 2B).
-These transcripts are less likely to be upregulated in any context, consistent with the tolerance hypothesis.
+These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.
 
@@
 
-To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time of day on 12 different days.
+To limit differences in gene expression not related to the experimental treatment, on 12 different days we haphazardly collected three ants from each 2012 colony at the same time of day to minimize variation due to circadian oscillations.
 We measured response to temperature with a one-hour static temperature application, which is ecologically relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012].
@@
 We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length.
-Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supporting Information).
+Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplementary Methods).
 We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly.
@@
 We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses.
-Because preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
-The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).
+Because preliminary examination of the data (Supplementary Methods) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis.
+The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplementary Methods).
 
@@
 
-All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supporting Information).
+All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supplementary Methods).
 
 ## Availability of supporting data
+
+Table S1 contains the results of the gene set enrichment analysis, showing the enriched gene ontology terms for each species in each thermal response category.
+
+The Supplementary Methods contain the detailed information on the analysis.
+The reproducible and version-controlled scripts underlying the analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
 
@@
 
-The reproducible and version-controlled scripts for data analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).
+## Competing interests
 
-## Author Contributions
+The authors declare they have no competing interests.
+
+## Authors' contributions
 
@@
 **Figure 1**.
-Prediction illustrations of expected thermal response patterns in the two species under alternative mechanistic hypotheses of temperature adaptation.
+Illustrations of the expected thermal response patterns in the two species under alternative mechanistic hypotheses of temperature adaptation.
 Although both temperature extremes were investigated in a similar way, for simplicity only the response to low temperatures is illustrated here.
 Each column indicates the distribution across all response categories in *A.
-picea*, which has a lower CT~min~ and is therefore better adapated to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CT~min~, *A.
+picea*, which has a lower CT~min~ and is therefore better adapted to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CT~min~, *A.
 carolinensis*, relative to the null hypothesis of equal marginal frequencies.
@@
 **Figure 2**.
-Results of marginal frequencies analysis of thermal response categories.
+Results of analysis of thermal response patterns in the two species.
 The color of each cell  (blue = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies.
@@
 
-# Tables and Figures
+# Tables
 
-```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A.
+```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally responsive transcripts by expression type for *A.
 carolinensis* and *A.

### 2016-01-10 scripts/run_busco.sh
added BUSCO analysis script

@@ -1,0 +1,30 @@
+# Script to run BUSCO analysis for ApTranscriptome
+# John Stanton-Geddes
+# 7 January 2016 
+
+# Install BUSCO and dependencies (NCBI Blast, HMMER, EMBOSS
+# following directions at 
+# http://busco.ezlab.org/files/README.html
+# Installed BUSCO to the /scripts directory
+
+# Installed Blast and HMMER to ~/software directory
+
+# NCBI: downloaded binaries from http://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/
+# HMMER 3.1b2: downloaded binaries from http://hmmer.janelia.org/
+# Installed EMBOSS using homebrew 
+# https://www.biostars.org/p/107092/
+# brew install homebrew/science/emboss
+
+# Add binaries to PATH
+
+PATH=$PATH:/Users/JSG/software/ncbi-blast-2.3.0+/bin/:/Users/JSG/software/hmmer-3.1b2-macosx-intel/binaries/:/usr/local/Cellar/emboss/6.6.0/bin
+
+# change directory
+cd ../results/trinity-full
+
+# rename FASTA headers for BUSCO input
+awk '/^>/{print ">chromosome" ++i; next}{print}' < Trinity_cap3_uclust_clean.fa > ../../scripts/BUSCO_v1.1b1/short_headers.fa
+
+# run BUSCO
+cd ../../scripts/BUSCO_v1.1b1
+python3 BUSCO_v1.1b1.py -o Aph20160107 -in short_headers.fa -l arthropoda -m trans

### 2016-01-10 scripts/expression_types_figure.R
added script to make cartoon figure of expression types

@@ -1,0 +1,24 @@
+library(ggplot2)
+library(RColorBrewer)
+library(gridExtra)
+
+x <- seq(0, 38.5, by = 3.5)
+yh <- x^2
+yl <- rev(yh)
+yi <- -2.67 * x^2 + 106.67 * x + 80
+yb <- 2.67 * x^2 - 106.67 * x + 1080
+
+lc <- brewer.pal(4, name = "Set1")
+
+d1 <- as.data.frame(cbind(x, yh, yl, yi, yb))
+
+d1long <- reshape(d1, varying = c("yh", "yl", "yi", "yb"), v.names = "Expression", direction = "long", times = c("High", "Low", "Intermediate", "Bimodal"))
+
+str(d1long)
+
+ggplot(d1long, aes(x = x, y = Expression, group = time)) +
+  geom_line() +
+  xlab(expression(Temperature ~degree~C)) +
+  facet_wrap(~ time) +
+  theme(legend.position = "none") +
+  scale_colour_brewer(palette="Set1")

### 2016-01-12 doc/ApTranscriptome_manuscript.Rmd
edits for BMC Genomics

@@
 
-**Results** We found that at least 2% of all genes were thermally responsive.
+**Results** We found that at least 2% of all genes changed expression with temperature.
 The majority of upregulation was specific to exposure to low temperatures.
@@
 
-We use the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A.
-rudis*-species complex, B.
-DeMarco personal communication) and *A.
+In this study, we used the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* [@demarco2015] and *A.
 picea* [@umphrey1996].
 *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
+Ants, and ectotherms in general, have little or no thermal safety margin [@sunday2014] and thus are highly susceptible to climate change [@diamond2011; @kingsolver2013], putting at risk important ecosystem services [@toro2012].
+Growth chamber studies have demonstrated that reproduction of *Aphaenogaster* will be compromised by increased tempreatures [@pelini2012], while field studies [@diamond2012] and simulations [@bewick2014] indicate that ant species persistence will depend on combinations of physiology and species interactions.
 *Aphaenogaster carolinensis* experiences a higher mean annual temperature (MAT) (14.6°C) and less seasonal temperature variation (temperature seasonality = 7,678°) than does *A.
@@
 
-Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adaptations [@gordon2013; @sunday2014].
+Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from an induced response in favor of structural changes [@lockwood2012] or behavioral adaptations [@gordon2013; @sunday2014].
 This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011].
@@
 picea*.
-Of course, these hypotheses are not mutually exclusive.
-By taking a reactionome approach, we can quantify if, and under what conditions, these mechanisms contribute to thermal adaptation.
+
+To summarise, in this project we generated the transcriptomes of two closely-related temperate ant species, and quantified their gene expression across a wide range of thermal conditions.
+We used a reactionome approach to identify all transcripts that changed expression across a thermal gradient each species.
+We then evaluated three non-mutually exclusive hypotheses (enhanced response, tolerance and genetic assimilation) of the evolution of thermal adaptation by comparing differences between the thermal reactionomes of each species.
+Finally, we used gene ontology information to determine which gene products and pathways are involved in thermal adaptation in the two species.
 
@@
 About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae.
-We fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship.
-To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationshp by chance alone.
-We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) true positive transcripts that exceeded null expectation from the resampling analysis.
+We performed a BUSCO analysis [@simao2015] to assess the quality of the transcriptome assembly against the arthropod Benchmarking Universal Single-Copy Orthologs (BUSCOs).
+This analysis revealed that transcriptome is largely complete, as we recovered 1,426 complete single-copy BUSCOs (62%) and an additional 435 fragmented BUSCOs (16%), which is in line with results of Simao et al.
+[-@simao2015] for transcriptomes of other non-model species.
+Moreover, only 8% of the BUSCOs were found to be duplicated in the transcriptome, which indicates that while we assembled the transcriptomes of the two species together, the steps we took to collapse homologs were successful.
+
+We quantified gene expression using the program `Sailfish` [@patro2014], and fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship (Table S1).
+To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationship by chance alone.
+We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) transcripts that exceeded the null expectation from the resampling analysis as true positive transcripts for further analyses.
 Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
@@
 low temperature).
-The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`).
+The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`, `r round(responsive_transcripts_w_BLAST / nrow(annotation.table.responsive), digits = 3)*100`).
 However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.
+
+We used the predicted transcript expression levels to partition transcripts for each species into five expression categories (Fig.
+1) which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other.
 
@@
 Specifically, the two species differed significantly in expression pattern, which captures differences in slope as well as category, for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts.
-Most of these shifted from one to another expression category.
 
 The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig.
-1).
+2).
 The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig.
-1, left).
+2, left).
 In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig.
-1, right).
+2, right).
 We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A.
@@
 picea* (Fig.
-2A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis.
+3A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis.
 In contrast, transcripts upregulated in response to high temperatures in *A.
@@
 carolinensis* (Fig.
-2B), primarily to the **Intermediate** category (Fig.
-2B).
+3B), primarily to the **Intermediate** category (Fig.
+3B).
 These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.
@@
 
-The gene set enrichment analysis revealed a number of gene groups enriched in each expression category.
+The gene set enrichment analysis revealed a number of gene groups enriched in each expression category (Table S2).
 Across both species, there were 9 terms enriched in the **Bimodal** category, including terms involved in stress response (regulation of cellular response to stress, signal transduction by p53 class mediator), cell death (apoptotic signaling pathway) and cellular organization (e.g.
@@
 picea* (`r round(T_on.low.ttest$estimate[2], 1)`°C; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig.
-3A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).
+4A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).
 
@@
 The difference between species was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig.
-3B).
+4B).
 
@@
 carolinensis* (Fig.
-4A).
+5A).
 Similarly, transcripts more upregulated at low temperatures in *A.
@@
 picea* (Fig.
-4B).
+5B).
 In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001).
@@
 carolinensis* (top-row of Fig.
-2B).
+3B).
 These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.
@@
 picea* upregulated responsive transcripts at slightly less extreme temperatures (Fig.
-3A).
+4A).
 Moreover, the transcripts upregulated in *A.
-picea* included about about half (`r round(tt2["Low", "Low"] / colSums(tt2)["Low"], 2)*100`%) those upregulated in *A.
-carolinensis* as well as an additional set of `r round(rowSums(tt2)["Low"] - tt2["Low", "Low"], 2)` transcripts (Table 1), enriched for metabolism, organization and translation processes (Table S1).
+picea* included about half (`r round(tt2["Low", "Low"] / colSums(tt2)["Low"], 2)*100`%) those upregulated in *A.
+carolinensis* as well as an additional set of `r round(rowSums(tt2)["Low"] - tt2["Low", "Low"], 2)` transcripts (Table 1), enriched for metabolism, organization and translation processes (Table S2).
 Two non-mutually exclusive hypotheses may explain this pattern.
@@
 Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair.
-Whether a particular taxon evolves one strategy over another may be related to availability of alternative mechanisms as well as the the intensity, frequency and duration of temperature stress in a given environment.
+Whether a particular taxon evolves one strategy over another may be related to availability of alternative mechanisms as well as the intensity, frequency and duration of temperature stress in a given environment.
 
 Given the differences in the patterns of thermal responsiveness between species (Fig.
-2), it is worth noting a number of similarities.
+3), it is worth noting a number of similarities.
 In both species, there were 2 -- 3 times more transcripts upregulated at low than high temperatures (Table 1).
@@
 At these extremes, we also found few upregulated transcripts (Fig.
-3A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig.
-3).
+4A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig.
+4).
 A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002].
@@
 Only ~10% of transcripts upregulated in response to temperature were bimodal, and for both activation and down-regulation, thresholds differed between species at only one temperature extreme (Fig.
-3).
+4).
 This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.
-
-A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach), which revealed quantitative patterns of temperature response not captured in categorical comparisons.
+A major contribution of this study is the construction of a reactionome for gene expression data.
+Similar approaches have been used in other species [@hodgins-davis2012; @aubin-horth2009], but to our knowledge, none have applied a regression approach to identify a complete list of responsive transcript across an environmental gradient.
+This approach revealed quantitative patterns of temperature response not captured in categorical comparisons.
 For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as discussed above.
@@
 Although we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig.
-2, Fig.
+3, Fig.
 S1).
@@
 These sites are centrally located within each species' geographic range.
+Along the East Coast of the United State, the distribution of *A.
+picea* ranges from central Maine south to northern Pennsylvania, while *A.
+carolinensis* is found from Pennsylvania to the Carolinas.
 Species identity was confirmed with morphological characters (Bernice DeMarco, Michigan State University).
@@
 
-To limit differences in gene expression not related to the experimental treatment, on 12 different days we haphazardly collected three ants from each 2012 colony at the same time of day to minimize variation due to circadian oscillations.
+To limit differences in gene expression not related to the experimental treatment (e.g.
+circadian rhythm), on 12 different days we haphazardly collected three ants from each 2012 colony at the same time of day to minimize variation due to circadian oscillations.
 We measured response to temperature with a one-hour static temperature application, which is ecologically relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012].
@@
 
+We assessed the quality of the transcriptome assembly using the BUSCO program [@simao2015] available from (http://busco.ezlab.org/).
+BUSCO asseses transcriptome completeness by measuring the number of near-universal single-copy orthologs selected from OrthoDB, using the Arthropod database.
+
 To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies transcripts by Gene Ontology (GO) term assignment, enzyme identification and domain identification.
@@
 For each thermally-responsive transcript, we predicted expression levels using the final linear model for each species across the tested thermal range.
-We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other.
+We used the predicted transcript expression levels to partition transcripts for each species into the five *a priori* defined expression categories: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other.
 For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the rearing temperature of 25°C.
@@
 
-Table S1 contains the results of the gene set enrichment analysis, showing the enriched gene ontology terms for each species in each thermal response category.
+Table S1 provides the annotation, *P*-value, r^2^, adjusted *P*-value, and expression type for the thermally-responsive transcripts in each species.
+
+Table S2 provides the results of the gene set enrichment analysis, showing the enriched gene ontology terms for each species in each thermal response category.
 
@@
 
-The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available at the Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).
+The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available at the Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-41, and hf113-42 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).
 
@@
 **Figure 1**.
+Illustration of the patterns against temperature for each of the four expression categories, **Bimodal**, **High**, **Intermediate** and **Low**.
+The fifth category of **Not Responsive** is not shown as these transcripts would be a line with no clear pattern.
+
+**Figure 2**.
 Illustrations of the expected thermal response patterns in the two species under alternative mechanistic hypotheses of temperature adaptation.
@@
 
-**Figure 2**.
+**Figure 3**.
 Results of analysis of thermal response patterns in the two species.
@@
 picea*, consistent with the *enhanced response* hypothesis (Fig.
-1).
+2).
 (B) High temperature extreme: in addition to an excess of matched categories, there is an excess of **High** and **Bimodal** transcripts in *A.
@@
 
-**Figure 3**.
+**Figure 4**.
 Histogram with smooth density estimate of temperature of maximum rate of change in expression for transcripts that have (A) increased expression at **Low** and **High** temperatures and (B) decreased expression at **Low** and **High** temperatures.
@@
 
-**Figure 4**.
+**Figure 5**.
 Scatterplots of log ratios of relative inducibility to relative constitutive expression, defined as expression level at the common rearing temperature (25°C) for (A) **High** transcripts in *A.
@@
 
-![Fig. 1](../results/predictions_Fig1.png)
+![Fig. 1](../results/expression_types_Fig1.png)
 
-![Fig. 2](../results/matched_observations_Fig2.png)
+![Fig. 2](../results/predictions_Fig2.png)
 
-![Fig. 3](../results/crit_temp_regulation_Fig3.png)
+![Fig. 3](../results/matched_observations_Fig3.png)
 
-![Fig. 4](../results/genetic_assimilation_Fig4.png)
+![Fig. 4](../results/crit_temp_regulation_Fig4.png)
+
+![Fig. 5](../results/genetic_assimilation_Fig5.png)
 

### 2016-01-12 scripts/expression_types_figure.R
additional files for submission

@@ -22,3 +22,5 @@ ggplot(d1long, aes(x = x, y = Expression, group = time)) +
   facet_wrap(~ time) +
   theme(legend.position = "none") +
   scale_colour_brewer(palette="Set1")
+
+ggsave(filename = "../results/expression_types_Fig1.png")

### 2016-01-15 ApTranscriptome_TR.Rmd
revisions for BMC Genomics

@@
 The raw Illumina fastq files are available from the NCBI short read archive [link tbd].
-The assembled transcriptome, annotation and expression values are available from the Harvard Forest Archives, HF-113 [http://harvardforest.fas.harvard.edu/data-archive] datasets [hf113-36](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-36-ap-ann-table.csv), [hf113-40](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-40-ap-trans.tar) and [hf113-41](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-gene-exp-quant.tgz).
+The assembled transcriptome, annotation and expression values are available from the Harvard Forest Archives, HF-113 [http://harvardforest.fas.harvard.edu/data-archive] datasets [hf113-38](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-38-ap-ann-table.txt), [hf113-41](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-trans.tar) and [hf113-42](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-42-ap-gene-exp-quant.tgz).
 These processed files are used for this analysis due to the computational demands of re-generating the files, but the exact commands for each of these steps are documented below.

### 2016-01-15 doc/ApTranscriptome_manuscript.Rmd
revisions for BMC Genomics

@@
     keep_md: yes
+  word_document: default
 includes: header.tex
@@
 
-In this study, we used the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* [@demarco2015] and *A.
-picea* [@umphrey1996].
+We used the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* and *A.
+picea* [@umphrey1996; @demarco2015].
 *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012].
-Ants, and ectotherms in general, have little or no thermal safety margin [@sunday2014] and thus are highly susceptible to climate change [@diamond2011; @kingsolver2013], putting at risk important ecosystem services [@toro2012].
+Ants, and ectotherms in general, have little or no thermal safety margin [@sunday2014] and thus are highly susceptible to climate change [@diamond2012; @kingsolver2013], putting at risk important ecosystem services [@toro2012].
 Growth chamber studies have demonstrated that reproduction of *Aphaenogaster* will be compromised by increased tempreatures [@pelini2012], while field studies [@diamond2012] and simulations [@bewick2014] indicate that ant species persistence will depend on combinations of physiology and species interactions.
@@
 To summarise, in this project we generated the transcriptomes of two closely-related temperate ant species, and quantified their gene expression across a wide range of thermal conditions.
-We used a reactionome approach to identify all transcripts that changed expression across a thermal gradient each species.
-We then evaluated three non-mutually exclusive hypotheses (enhanced response, tolerance and genetic assimilation) of the evolution of thermal adaptation by comparing differences between the thermal reactionomes of each species.
+We then evaluated three non-mutually exclusive hypotheses (enhanced response, tolerance and genetic assimilation) of the evolution of thermal adaptation by comparing the number and expression patterns of transcripts between species in response to extreme low and extreme high temperatures.
 Finally, we used gene ontology information to determine which gene products and pathways are involved in thermal adaptation in the two species.
@@
 [-@simao2015] for transcriptomes of other non-model species.
-Moreover, only 8% of the BUSCOs were found to be duplicated in the transcriptome, which indicates that while we assembled the transcriptomes of the two species together, the steps we took to collapse homologs were successful.
+Moreover, only 8% of the BUSCOs were found to be duplicated in the transcriptome, which indicates that the steps (see Methods) we took to collapse homologs in the combined transcriptome of the two species were successful.
 
-We quantified gene expression using the program `Sailfish` [@patro2014], and fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship (Table S1).
+We quantified gene expression using the program `Sailfish` [@patro2014], and fitted polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship (Fig.
+1).
 To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationship by chance alone.
-We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) transcripts that exceeded the null expectation from the resampling analysis as true positive transcripts for further analyses.
+We retained the `r prettyNum(nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95), big.mark = ",")` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) transcripts that exceeded the null expectation from the resampling analysis as true positive transcripts for further analyses (Table S1).
 Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g.
@@
 In addition, we also examined the specific patterns of shifts from one expression category to another between species.
-When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig.
+As transcripts may change expression between species due to neutral drift alone, we used the Stuart-Maxwell test of marginal homogeneity to test if the number of responsive transcripts in each expression category differed between the species when controlling for overall differences in the number of responsive transcripts.
+We found that the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig.
 S1).
@@
 2, right).
-We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A.
+We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by a species (cool temperatures for the warm-climate *A.
 carolinensis*, and warm temperatures for the cool-climate *A.
-picea*) displayed the same response profile in the species that more frequently experiences those conditions.
+picea*) displayed the same response profile in the other species that more frequently experiences those conditions.
 
@@
 carolinensis*.
-At least 2.5% of their transcriptomes are thermally responsive.
+At least 2% of their transcriptomes are thermally responsive.
 Our results indicate that these two ant species have different responses to thermal extremes.
@@
 
-The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available at the Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-41, and hf113-42 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).
+The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available from the LTER data portal, datasets hf113-38, hf113-41, and hf113-42 (http://dx.doi.org/10.6073/pasta/05ea6464df30efa2f1e2c7439366bf47).
 
@@
 Illustration of the patterns against temperature for each of the four expression categories, **Bimodal**, **High**, **Intermediate** and **Low**.
-The fifth category of **Not Responsive** is not shown as these transcripts would be a line with no clear pattern.
+The fifth category of **Not Responsive** is not shown.
 

### 2016-01-15 doc/BMCGenomics_cover_letter.tex
revisions for BMC Genomics

--- 2014-11-26 doc/MolEco_cover_letter.tex
@@
 
-\noindent{Dear Dr. Riesberg,}
+\noindent{Dear Editor,}
 
-We are excited to submit our manuscript \emph{Thermal
-reactionomes reveal adaptive responses to thermal extremes in warm and
-cold-climate ant species} for consideration as an Original Article in
-\emph{Molecular Ecology}. There are
+We are writing to submit our manuscript \emph{Thermal
+reactionomes reveal divergent responses to thermal extremes in warm and
+cold-climate ant species} for consideration as a Research Article. There are
 several aspects of the work that we think make it worthy of publication
-in \emph{Molecular Ecology}.
+in \emph{BMC Genomics}.
 
@@
 
-In this study, we resolve this problem by extending the reaction
+In this study, we address this problem by extending the reaction
 norm approach to the study of gene expression. We define the
@@
 
-\textbf{Suggested reviewers:}
-
-\begin{itemize}
-  \item Dr. Nicholas Teets - expertise with insect genomics and transcriptomics
-  \item Dr. Vanessa Kellerman - expertise with insect population and quantitative genetics
-  \item Dr. Pavlos Pavlidis - critical insight on function annotation and gene ontologies
-  \item Dr. Joel Kingsolver - expertise in insect thermal limits
-  \item Dr. Daniel Barshis - expertise in ecological genomics
-\end{itemize}
-
-\textbf{Suggested editors:}
-
-\begin{itemize}
-  \item Dr. Ary Hoffman - expertise with insect population and quantitative genetics
-  \item Dr. Stephen Palumbi - expertise in ecological genomics
-  \item Dr. Aurelie Bonin - expertise in population genomics
-\end{itemize}
-
 \noindent{Thank you for consideration of our work,} \\
